### 2016-06-01 Chapters/response_reviewers.tex
First meeting with H&H

\documentclass[11pt]{article}
\usepackage{amsmath,amscd,amssymb,amstext,epsfig,textcomp}
\usepackage{lscape}
%\usepackage{multirow}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage[authoryear]{natbib}

\usepackage{color}
\newcommand{\tr}[1]{\textcolor{red}{#1}}

\begin{document}

\title{\bf Response to Reviewers of "Dynamical allocation of cellular resources as an optimal control problem: Novel
insights into microbial growth strategies" (PCOMPBIOL-S-15-02269)}

\author{Nils Giordano$^{\text{}1,2}$, Francis Mairet$^{\text{}3}$, Jean-Luc Gouz\'{e}$^{\text{}3}$, \\ Johannes Geiselmann$^{\text{}1,2,*}$, Hidde de Jong$^{\text{}2,*}$ \\ \\
1. Universit\'e Grenoble Alpes, Laboratoire Interdisciplinaire de Physique (CNRS UMR 5588) \\ 140 rue de la physique BP 87, 38402 Saint Martin d'H\`{e}res  France\\
2. INRIA, Grenoble - Rh\^one-Alpes research centre,\\ 655 avenue de l'Europe, Montbonnot, 38334 Saint Ismier Cedex, France \\
3. Inria, Sophia-Antipolis M\'{e}diterran\'{e}e research centre, 2004 route des Lucioles, BP 93,\\ 06902 Sophia-Antipolis Cedex, France \\
\\
$^{*}$ Corresponding authors with equal contributions: \\ Johannes Geiselmann (Hans.Geiselmann@ujf-grenoble.fr), \\ Hidde de Jong (Hidde.de-Jong@inria.fr)
}

\date{}

\maketitle
%\nocite{*}

\noindent  We would like to thank the reviewers for their comments on our manuscript "Dynamical allocation of cellular resources as an optimal control problem: Novel insights into microbial growth strategies" (PCOMPBIOL-S-15-02269). In this document we describe how our revised manuscript addresses the comments of reviewer 3, which mostly concern the presentation of our work. In summary, we have made the following two major changes:

\begin{enumerate}
\item The second part of the \textit{Results} section has been streamlined and we have tried to clarify the presentation throughout;
\item A \textit{Methods} section has been added to the paper and part of the material in the Supplementary Information has been shifted to this new section.
\end{enumerate}
Below we respond to the reviewer comments in detail and we indicate which changes have been made to the manuscript (major changes in red). The reviewer comments are in italic, and our response in default font.

\section*{Reviewer 3}

\textit{In this paper the authors study a resource allocation problem to identify strategies that microbes use to adapt their proteome in response to nutrient upshifts. The paper addresses a very relevant topic in systems biology with a unique (and very much needed) angle from a dynamic optimization framework. Their main results say basically that optimal regulatory strategies that maximize steady state growth are not necessarily optimal in dynamic environments. As far as I know this is the first study that reveals this property from dynamic resource allocation problems. They go further and propose a number of feedback strategies, based on precursor-sensing, nutrient-sensing, and the imbalance between precursors and ribosomes (akin to the ppGpp system), and compare them in terms of their dynamic responses. This is a very interesting paper and a solid contribution. The results seem correct but I have a number of comments to streamline ideas and homogenize the quality of the different sections of the paper in terms of language, figures and flow of ideas.} \\

\textit{- Some parts of Text S1 and S4 should be included as methods or an appendix in the main text to improve readability. I also found a small typo in S1.10, a missing \^ in the jacobian.} \\

\noindent\textbf{Answer:} We have added a \textit{Methods} section to the paper, with the following subsections:

\begin{enumerate}
\item \textit{Steady-state analysis of model.} This subsection summarizes the major results of the model analysis in Text~S1, notably explicit expressions for the optimal growth rate at steady state referred to in the main text. The details of the derivation of these expression remain in Text~S1. 
\item \textit{Model fitting.} This subsection describes how the self-replicator model was fitted against the data of \cite{Scott2010}. The contents of this subsection were formerly included in Text~S2.
\item \textit{Solution of optimal control problem.} This subsection provides an outline of the way in which the analytical results for the optimal control problem were obtained. The details of the derivation remain in Text~S3. The subsection also describes how the optimal control problem was solved numerically for generating Fig.~4, a text formerly included in Text~S3.
\item \textit{Specification and analysis of control strategies.} This subsection shows that all three control strategies discussed in the paper drive the self-replicator to the optimal steady state after the upshift. The contents were moved here from Text~S4, which has been renamed to \textit{Kinetic model of the ppGpp system in Escherichia coli} to better cover the remaining material.
\end{enumerate}

The typo in S.1.10 was corrected.\\

\textit{- Fig 2 is important but a bit convoluted. I am not sure panel 2B adds too much information and moreover, the references to Fig 2 in the text are a bit confusing, as they go back and forth between panels and it is cumbersome to follow.} \\

\noindent\textbf{Answer:} We have reduced the information overload by splitting Fig.~2 in the original manuscript into two new figures. Fig.~2 in the revised manuscript (corresponding to Fig.~2\text{A-B} in the original manuscript) presents the analysis of the self-replicator model of bacterial growth, whereas Fig.~3 in the revised manuscript (corresponding to Fig.~2\text{C-D} in the original manuscript) shows that the model is consistent with the steady-state data of Scott \textit{et al.} We believe it is important to show the dynamics in the phase plane, but we have improved the legibility of the plot (Fig.~2\textit{A} in the revised manucript). The order of the figures and the figure panels now corresponds to the order in which they are mentioned in the text.\\

\textit{- I suggest to unify the language across the whole manuscript. For example, some sections use a natural language while others a control-theoretic jargon (eg p 10), which I think it’s not the best choice unless concepts are clearly defined. A similar problem happens with several concepts in the manuscript, eg. the authors refer to the ‘environmental variable’ $e_{m}(t)$, but the constant turnover rate $E_m$ as a parameter; another example is the use of ‘intensive’ variables, vs ‘extensive’ ones, without an explanation of what these mean.} \\

\noindent\textbf{Answer:} We have tried to unify the language across the manuscript and avoid jargon when possible. In particular, we have removed some terminology specific to control theory from the main text, such as "singular arc" and "bang arc", and we have summarized the properties of the solution of the optimal control problem more informally in the main text. The notions of "bang-bang-singular" and "Pontryagin maximum principle" have been preserved, as they are central to the paper and have been used before in systems biology and mathematical biology (see the references in the \textit{Discussion}). The \textit{Methods} subsection \textit{Solution of optimal control problem} provides a more technical description and uses the appropriate terminology from control theory.

Whereas $e_m(t)$ is a variable in the general model, it becomes a constant $e_m$ (or $E_M$ in its rescaled form) in the specific upshift scenarios we consider. This has been further emphasized in the text, notably in the sections \textit{Solution of the optimal control problem} and \textit{Simple feedback control strategies}. We use the notion of "intensive" variables only once, but in a context where it is difficult to replace by another term and where it is immediately followed by an intuitive explanation and an appropriate reference (\textit{Introduction}).\\

\textit{- The introduction, first results sections and discussion are excellently written, with clean explanations and model assumptions clearly laid out. However, from p10 the flow is a bit broken and results are not so cleanly presented. I would advise the authors to streamline the text, lay out assumptions clearly (as in the first results section) and improve the figures as well. A figure for the different feedback schemes in p11 would help to clarify what is being sought, while Fig 4 could be removed or converted to an inset, as it contains little information. Another idea is to improve Figs 5 and 6: I am not sure the vector fields in panels A-B speak out the point the authors make in the text. Maybe these panels should be together for better comparison, or the authors could better explain what they mean by ‘performance’ when they compare the different dynamic responses in 5c-d vs 6c-d.} \\

\noindent\textbf{Answer:} We have rewritten the final three \textit{Results} sections to improve the overall flow of the text. In particular, we have added more intuitions and motivations, and we have tried to make the assumptions more explicit. We have also relegated Fig.~4 to a Supplementary figure (Fig.~S1) and added a new Fig.~5 in which the control strategies are graphically superposed on the self-replicator system of Fig.~1. We have preserved Figs~5 and 6 (Figs~6 and 7 in the revised manuscript), but we have better explained in the text how the vector fields in panels \textit{A-B} relate to the temporal response curves in panels \textit{C-D}.  \\

\textit{- Text S3. I think the authors could expand this section to cleanly explain the variable definitions, the derivations and pontryagin’s conditions in more detail. 3 examples: a) the constancy of the hamiltonian appears as a condition only in the middle of the derivation, b ) the definition of ‘overtaking optimality’ is missing (but crucial to the derivation, is this the role of $\lambda_0$ in the hamiltonian? ), c) the footnote 2 about abnormal extremal trajectories needs more explanation. Since this is a supplementary file, the authors can afford to explain things in detail while keeping it accessible to someone not entirely familiar with the concepts from optimal control.} \\

\noindent\textbf{Answer:} Text~S3 has been extended to provide more explanations. In particular, the conservation of the Hamiltonian along extremal trajectories is now explicitly stated earlier in the text, the notion of "overtaking optimality" is explicity defined, and footnote~2 is further developed in the text.\\

\textit{- In Fig 3B, it is unclear why the switching function has constant sign but $\alpha(t)$ switches right after $\hat{t}=2$.} \\

\noindent\textbf{Answer:} In the revised manuscript, this point is further developed at the end of the section \textit{Solution of the optimal control problem}, when discussing the figure (Fig.~4\textit{B} in the revised manuscript). We explain that the optimal solution does cross the switching curve at the moment that $\alpha$ switches from 1 to 0, around $\hat{t}=2$, and remains close to the switching curve until the next switch occurs.   \\

\textit{- In S4.1, it is unclear why ‘we know from Eq. S1.20 that $g(\hat{p})$ is the only control law...'} \\

\noindent\textbf{Answer:} The formulation was misleading: the statement is not "known", but shown here, immediately after Eq.~S4.1, with the help of Eq. S1.20. We changed the text to better bring this out (notice that this section has been moved to the \textit{Methods} section of the main text, as described above, while the proof is now included in Text~S1). \\

\textit{- S4.3, I suggest to include the definitions/notation from Bosdriesz et al explicitly to make it a standalone document.}

\noindent\textbf{Answer:} Done, in Text~S4. \\

\begin{thebibliography}{10}

\bibitem[{Scott {\it et~al\/}(2010)Scott, Gunderson, Mateescu, Zhang \&
  Hwa}]{Scott2010}
Scott M, Gunderson CW, Mateescu EM, Zhang Z, Hwa T (2010) {Interdependence of
  cell growth and gene expression: origins and consequences.} {\it Science\/}
  {\bf 330}: 1099--1102

\end{thebibliography}

\end{document}

### 2016-06-01 Chapters/S1_Table.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}
% Add an S1
\renewcommand{\thefigure}{S2}%

\usepackage{color} 
\newcommand{\tr}[1]{{\color{red} #1}}
\newcommand{\tb}[1]{{\color{blue} #1}}

% Add an S1
\renewcommand{\thetable}{S\arabic{table}}%

\begin{document}
\begin{flushleft}
{\Large
\textbf\newline{\textbf{S1 Table -- Parameter values of self-replicator model\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}

\begin{table}[!h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Parameter & Unit & Literature value & Fitted value \\
\hline
$\gamma$ & $-$ & No value & 1.39\\
\hline
$k_R$ & h$^{-1}$ & 3.6 & 2.23\\
\hline
$e_M$ for M63+glycerol & h$^{-1}$ &  $< 3.6$ & 0.587\\
\hline
$e_M$ for M63+glucose & h$^{-1}$ &  $< 3.6$ & 0.867\\
\hline
$e_M$ for cAA+glycerol & h$^{-1}$ &  $< 3.6$ & 1.07\\
\hline
$e_M$ for cAA+glucose & h$^{-1}$ &  $< 3.6$ & 1.57\\
\hline
$e_M$ for RDM+glycerol & h$^{-1}$ &  $< 3.6$ & 3.48\\
\hline
$e_M$ for RDM+glucose & h$^{-1}$ &  $< 3.6$ & 4.76\\
\hline
$\beta K_R$ & $-$  & 0.003 & Not fitted\\
\hline
\end{tabular}
\caption{\textbf{Parameter values of self-replicator model}
The parameter values in the model were obtained by fitting Eq.~\ref{eq:alpha_mu_optimal_dim} to the data of Scott et al~\cite{scott_interdependence_2010} (Fig.~\ref{fig:model_validation} in the main text), as described in the \textit{Methods} section. They are compared with order-of-magnitude estimates from the literature (\ref{S2_Text}). }
\end{table}

\bibliographystyle{plos2015}
\bibliography{references}

\end{document}

### 2016-06-01 Chapters/S3_Text.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}

%%%%%  AJOUT  FRANCIS %%%%%%%%%%%%%%%%%%%%%%%%%55
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}

\usepackage{amssymb}
\usepackage{mathrsfs}

\usepackage{color} 
\newcommand{\tr}[1]{{#1}}

%%%%%% FIN AJOUT FRANCIS %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Add an S3
\renewcommand{\thetable}{S3.\arabic{table}}%
\renewcommand{\thefigure}{S3.\arabic{figure}}%
\renewcommand{\theequation}{S3.\arabic{equation}}
\renewcommand{\thesection}{S3.\arabic{section}}%

\begin{document}
\begin{flushleft}
{\Large
\textbf\newline{\textbf{S3 Text -- Solution of optimal control problem\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}
\section{Statement of the problem}

We consider the dimensionless system defined by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} in the main text, which are here repeated for clarity:
\begin{equation}{\label{eq:model}}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1+\hat{p})\, \hat{r} \, \frac{\hat{p}}{K+\hat{p}}, \\
\frac{d\hat{r}}{d\hat{t}} &= \hat{r} \frac{\hat{p}}{K + \hat{p}} \, (\alpha (\hat{t}) - \hat{r}).
\end{aligned}
\end{equation}

\tr{As stated in the section \textit{Biomass maximization as an optimal control problem} in the main text, the objective of this study is to maximize the growth rate on an interval $[0,\tau]$ after a nutrient upshift. With Eq.~\ref{S1.eq:supp_growthrate_adim} in \ref{S1_Text}, we have

\begin{equation*}
\hat{\mu}= \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}.
\end{equation*}
In order to avoid boundary effects occurring over finite time intervals, notably the depletion of precursors just before $\tau$, we solve the optimal control problem over an infinite horizon ($\tau \rightarrow \infty$).} Consider the set of admissible controls
\[
\mathcal{U}=\{\alpha:\mathbb{R} \rightarrow [0,1] \ \mid \ \alpha(\cdot) \ \mathrm{measurable}\}.
\]
The optimization problem can then be stated as follows:
\begin{equation}\label{Prob}
\tr{\alpha_{opt} = \arg \max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{+\infty} \hat{r}(\hat{t}) \frac{\hat{p}(\hat{t})}{K + \hat{p}(\hat{t})} d\hat{t} ,} 
\end{equation}
where $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))$ is the unique solution of Eq.~\ref{eq:model} starting at a given point $(\hat{p}_0,\hat{r}_0)\in \Omega \equiv \mathbb{R}^+_* \times (0,1)$ for a given control $\alpha\in \mathcal{U}$.

\tr{Given that the performance index $J(\alpha)$ diverges, we actually consider \textit{overtaking optimality}~\cite{carlson_infinite_1991}.
Consider the performance index of the trajectory $x(\cdot)$ emanating from $x_0$ and generated by $u(\cdot)$ defined for any $T\geq 0$ by
$$
J_T(x_0,u(\cdot))=\int_0^T f_0(x(t),u(t),t)dt.
$$
A trajectory $x^*(\cdot)$ emanating from $x_0$ and generated by $u^*(\cdot)$ is said to be overtaking optimal if for any other trajectory $x(\cdot)$ emanating from $x_0$ and generated by $u(\cdot)$ the following holds
$$
\liminf_{T\rightarrow\infty} \left\lbrace J_T(x_0,u^*(\cdot)) - J_T(x_0,u(\cdot))\right\rbrace\geq 0.
$$
Roughly speaking, a trajectory is overtaking optimal if "the performance index catches up with the performance index of any other trajectory"~\cite{carlson_infinite_1991}.}

\section{Maximum Principle}
Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
Let $H(\hat{p},\hat{r},\lambda_p,\lambda_r,\lambda_0,\alpha)$ be the Hamiltonian of the system, defined by
\[
H(\cdot) \equiv \lambda_p\, E_M\, (1-\hat{r}) - \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}\left[\lambda_p (1+\hat{p}) +\lambda_r\, \hat{r} +\lambda_0\right] + \alpha \, \lambda_r \, \hat{r}\, \frac{\hat{p}}{K+\hat{p}}.
\]
Moreover, let $\alpha$ be an optimal control, and $\hat{x}(\cdot)=(\hat{p}(\cdot),\hat{r}(\cdot))$ the associated trajectory.
Then, there exists $\lambda_0 \leq 0$ and an absolutely continuous map $\lambda=(\lambda_p,\lambda_r):[0,+\infty) \rightarrow \mathbb{R}^2$
such that $(\lambda,\lambda_0)\neq0$, and
\begin{align}
\dot{\lambda}_p\, \tr{=-\frac{\partial H}{\partial \hat p}}&= \hat{r} \, \frac{K}{(K+\hat{p})^2}\, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, ( \hat{r}-\alpha) + \lambda_0\right] + \hat{r}\, \frac{\hat{p}}{K+\hat{p}}\, \lambda_p, \label{eq:adjoint_p}\\
 \dot{\lambda}_r\, \tr{=-\frac{\partial H}{\partial \hat r}}&= \lambda_p\, E_M +\frac{\hat{p}}{K+\hat{p}} \, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, (2 \hat{r}-\alpha) +\lambda_0\right] \label{eq:adjoint_r}.
\end{align}
The maximization condition is given by:
\begin{equation}{\label{eq:PMP}}
\begin{aligned}
&\alpha(\hat{t}) \in \arg\ max_{v \in [0,1]} H(\hat{x}(\hat{t}),\lambda(\hat{t}),\lambda_0,v), \\ &\mathrm{almost}\ \mathrm{everywhere}\ \mathrm{on}\ [0,+\infty).
\end{aligned}
\end{equation}

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}. \tr{The extremal is said to be normal (resp. abnormal) if $\lambda_0<0$ (resp. $\lambda_0=0$). In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.}\\

From Eq.~\ref{eq:PMP}, it follows that the control strategy is given by the sign of the \textit{switching function} $\phi(\cdot) \equiv \lambda_r \, \hat{r}\, \hat{p}/(K+\hat{p})$, that is,
\[
\begin{cases}
\alpha=1 \iff \phi(\cdot) >0,\\
\alpha=0 \iff \phi(\cdot)<0.
\end{cases}
\]

\tr{Finally, given that the system is autonomous, the Hamiltonian is conserved along any extremal trajectory.}

\section{Characterization of singular arcs}

Whenever $\phi$ is vanishing over a time interval, we say that the trajectory is \textit{singular}.
We will now characterize such trajectories.
If $I=[\hat{t}_1,\hat{t}_2]$ is a singular arc, we have $\phi(\hat{t})=\dot{\phi}(\hat{t})=0$, for all $\hat{t}\in[\hat{t}_1,\hat{t}_2]$, that is, $\lambda_r(\hat{t})=0$ and $\dot\lambda_r(\hat{t})=0$.

\tr{For abnormal extremal} \tr{trajectories}, \tr{we get $\lambda_p(\hat{t})=0$, in contradiction with the Maximum Principle, so there is no singular arc.} \tr{An abnormal extremal trajectory is therefore a concatenation of bang arcs.}

\tr{For normal extremal} \tr{trajectories}, using additionally that $H$ is constant along an extremal trajectory, we obtain that $\lambda_p$ is constant along a singular arc. \tr{By combining $\dot \lambda_p=0$ and $\dot \lambda_r=0$, we obtain} $\hat{p}(\hat{t})=\sqrt{E_M\, K}=\hat{p}_{opt}^*$. Using $d\hat{p}/d\hat{t}=0$, we finally get $\hat{r}(\hat{t})=\hat{r}_{opt}^*$.
Thus, the singular arc is the optimal steady state, corresponding to a singular control \tr{$\alpha(\hat{t})=\alpha_{opt}^*$, with $\alpha_{opt}^*$} depending on $E_M$ (\ref{S1_Text}).

A necessary condition of optimality for a singular arc is given by the Kelley condition~\cite{borisov_fullers_2000}.
We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
$$
(-1)^q\frac{\partial}{\partial\alpha}\frac{d^{2q}}{d\hat{t}^{2q}}\phi(\hat{t})<0,
$$
satisfying the Kelley condition necessary for optimality. Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (also called the Fuller's phenomenon, \textit{i.e.}, an arc with an infinite number of switches~\cite{borisov_fullers_2000,marchal_chattering_2013}). \\

\tr{\section{Analysis of the adjoint system}
Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (\tr{Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}}) may be useful to characterize the switches of extremal trajectories.

First, \tr{for the abnormal case, we can easily determine in the phase-plane the possible transitions between the four regions defined by the axes (see Fig.~\ref{fig-adj}). A trajectory can cross at most twice the $\lambda_p$-axis, so}  we conclude that an abnormal extremal cannot have more than two switches. Thus, an abnormal extremal is a concatenation of at most three bang arcs ($\alpha(t)=0$ or $\alpha(t)=1$). When  $\alpha(t)=0$ or $\alpha(t)=1$ for a long time, the growth rate tends to zero. We therefore conclude that  abnormal extremal trajectories are not optimal.

Secondly, for the normal case, after the first switch, a trajectory with two consecutive switches in the regions $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$ is not} \tr{possible, as shown in Fig.~\ref{fig-adj}. Therefore, such a trajectory is not optimal given that it does not fulfill the conditions given by the Maximum Principle.
We conclude that if the optimal trajectory has a concatenation of bang arcs, the switches must alternatingly occur in the regions  $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ and $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.}

\begin{figure}[h]
\centering
\includegraphics[width=0.35\textwidth]{Fig/adj2ab.eps} 
\includegraphics[width=0.6\textwidth]{Fig/adj2.eps} 
\caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis. \tr{ Left: abnormal case. An extremal trajectory cannot have more than two switches. Right: normal case.} $(\lambda_p^s,0)$ corresponds to the singular arc.  After the first switch, an extremal trajectory cannot have two consecutive switches if it stays in the region $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.}
\label{fig-adj}                               
\end{figure}

\section{Optimal trajectories}

From the Maximum Principle, we have shown that the optimal trajectory is a concatenation of bang arcs ($\alpha(t)=0$ or $\alpha(t)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$. Moreover, if the optimal trajectory has a singular arc, it must enter it through a chattering arc (\textit{i.e.}, with an infinite number of switches between $\alpha=0$ and $\alpha=1$).

These elements motivate the supposition that optimal solutions consist in a transient (chattering arc) towards the optimal steady state, after which they remain there (until the next change of environment). The chattering arc can be characterized by a switching curve $\hat{p}\mapsto\varphi(\hat{p})$ which passes through the optimal steady state.
Defining $A_0$ and $A_1$ the regions above and below $\varphi$ in the $(\hat{p},\hat{r})$-plane, respectively, we conjecture that the following feedback control law is optimal: 
%The simplest way to reach the optimal steady-state is thus two arcs, either 0-1 or 1-0 depending on the initial conditions. More precisely, let $\phi_0$ and $\phi_1$ be defined as the the trajectories solution of System \eqref{model} backward in time starting at $(p_{opt}^*,r_{opt}^*)$ with respectively $\alpha=0$ and $\alpha=1$. Given $\phi=\phi_0 \cup \phi_1$, we define $A_0$ and $A_1$ the regions above and below $\phi$ in the plane $(p,r)$.

\begin{equation}\label{control-opt}
\begin{cases}
\alpha(\hat{t})=0 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_0,\\
\alpha(\hat{t})=1 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_1,\\
\alpha(\hat{t})=\alpha_{opt} \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}

\tr{Loosely speaking, the chattering arc corresponds to a spiral composed of bang arcs wrapping around the optimal steady state,} \tr{where the switches alternatingly occur} \tr{in the regions  $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ and $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$, in line with the analysis of the adjoint system.}
This is a first hint that the proposed control strategy is optimal. Moreover, our conjecture is also in line with the \textit{turnpike property}:
Trélat and Zuazua~\cite{trelat_turnpike_2015} have shown that, for quite a generic class of systems, the optimal strategy consists in staying at the optimal steady state (after a short transient).

\tr{As explained in the \textit{Methods} section of the main text, we} numerically solved the optimal control problem by the direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012}. 
It is important to stress that the optimization process was performed without any preliminary assumptions on the characteristics of the optimal trajectory. 
The fact that the numerical solution verifies the Maximum Principle (\textit{i.e.}, the singular arc corresponds to the optimal steady state) and the Kelley condition (\textit{i.e.}, the presence of a chattering arc)
tends to confirm that the control strategy of Eq.~\ref{control-opt} is optimal. As an aside, we note that due to the fact that numerical optimization was performed for a finite horizon, we actually obtained a second chattering arc escaping from the singular arc at the end of the simulation.  This is a classical property of the turnpike strategy: the optimal trajectory leaves the optimal steady state just before the end of the time interval of interest, in our case consuming almost all precursors. This arc was removed from the plot in Fig.~\ref{fig:optimalcontrol}, because it does not occur with an infinite horizon and is therefore a numerical artifact for this study.

\bibliographystyle{unsrt}
\bibliography{references}

\end{document}

### 2016-06-01 Chapters/abstract.tex
First meeting with H&H

\section*{Abstract}
Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.

### 2016-06-01 Chapters/core.tex
First meeting with H&H

\section*{Introduction}

Microorganisms adapt their physiology to changes in nutrient availability in the environment.
This involves changes in the expression of a large number of genes, encoding proteins with a variety of cellular functions, such as transporters for the uptake of nutrients, enzymes for the conversion of nutrients to energy and building blocks for macromolecules, the components of the transcriptional and translational machinery, and transcription factors to preferentially direct RNA polymerase to specific promoters~\cite{schaechter_microbe_2006,keseler_ecocyc_2013}.
Fundamentally, the reorganization of gene expression in response to changes in environmental conditions is a resource allocation problem.
It poses the question how microorganisms redistribute their protein synthesis capacity over different cellular functions when constrained by the changing environment.

The mechanisms responsible for resource allocation in microbial cells are usually assumed to have been optimized through evolution, so as to maximize the offspring of cells in their natural environment.
How this general principle manifests itself on the level of cellular physiology is not straightforward though.
Many studies have reasoned that growth-rate maximization provides a selective advantage to microorganisms, because it allows competitors to be outgrown when resources are scarce.
Others have shown, however, that appropriate optimization criteria will depend on the structure of the environment and the ecosystem, as well as on the molecular properties of metabolic pathways~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuetz_multidimensional_2012,schuster_maximization_2008,schuetz_systematic_2007}.
For instance,  in environments without competition for a shared resource, maximization of growth yield rather than growth rate is expected to provide a selective advantage.
Although what counts as optimal is thus context-dependent, growth and evolution experiments in \textit{Escherichia coli} have shown that in certain conditions bacterial metabolism is indeed geared towards growth-rate maximization~\cite{edwards_silico_2001,ibarra_escherichia_2002,lewis_omic_2010}.

For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-grained models of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
Along these lines, Molenaar \textit{et al.} developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \textit{et al.} derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
Based on a structurally similar model, Maitra and Dill related optimal resource allocation to the basic constants of the metabolic and gene expression machinery, in particular energy efficiency~\cite{maitra_bacterial_2014}.

The assumption of growth-rate maximization may lead to correct predictions in some situations, but ignores the regulatory mechanisms achieving resource allocation and therefore cannot provide a causal explanation of cellular behavior~\cite{kremling_understanding_2015}.
Several studies have used coarse-grained models to understand which control strategies microorganisms employ to achieve (optimal) resource allocation~\cite{bosdriesz_how_2015,scott_emergence_2014,weisse_mechanistic_2015}.
Scott \textit{et al.} have shown that a robust feedforward control strategy, based on the sensing of the amino acid pool size and the corresponding adjustment of the fraction of ribosomes producing ribosomal proteins, allows the ribosomal protein fraction to be maintained close to its optimal value under a variety of growth conditions~\cite{scott_emergence_2014}.
The authors suggest that this control strategy involves the signalling molecule ppGpp, in agreement with conclusions drawn from a recent kinetic model of the regulatory mechanisms achieving optimal adjustment of the ribosomal protein fraction~\cite{bosdriesz_how_2015}.
Wei{\ss}e \textit{et al.} also developed a coarse-grained model of microbial growth based on resource allocation trade-offs~\cite{weisse_mechanistic_2015}. 
Without including specific regulatory interactions, the model accounts for the above-mentioned bacterial growth laws, predicts host-circuit interactions in synthetic biology, and relates gene regulation to the nutrient composition of the medium.

The above studies consider resource allocation at steady state, where all intensive variables describing the growing microbial culture, in particular the concentrations of its molecular components, are constant (see~\cite{fishov_microbial_1995} for a precise definition of steady-state growth and the closely related notions of balanced and exponential growth).
This requires an environment to be stable over a long period of time.
Such conditions can be achieved in the laboratory~\cite{borirak_molecular_2014}, but many microorganisms naturally experience frequently-changing conditions.
For example, \textit{E. coli} can cycle between two distinct habitats, the mammalian intestine and the earth's surface (water, sediment, soil)~\cite{savageau_escherichia_1983}.
The bacteria transit through different microenvironments in the intestinal system, where they encounter different mixes of sugars~\cite{savageau_demand_1998}.
They are even more challenged in the open environment outside the host, with a greatly fluctuating availability of carbon and energy sources and a large variability in temperature, osmolarity, oxygen, and microbial communities~\cite{blount_unexhausted_2015,vanelsas_survival_2011}.

This situation motivates a dynamical perspective on microbial growth and resource allocation~\cite{pavlov_optimal_2013,vandenberg_optimal_1998,waldherr_dynamic_2015,ehrenberg_mediumdependent_2012}. However, fundamental results like the growth laws uncovered for steady-state conditions are still lacking.
In particular, extending the results reviewed above to dynamical conditions raises the following questions:
Are control strategies that maximize steady-state growth also optimal in dynamical environments?
If this is not the case, then which alternative strategies would be optimal for such conditions?
And finally, how do these strategies compare with the regulatory mechanisms that have actually evolved in microorganisms?

The aim of this study is to address the above fundamental questions in a specific dynamical growth scenario, namely a transition between two steady states following an environmental perturbation.
In particular, we consider the upshift of a microbial culture from a medium supporting growth at a low rate to a medium supporting growth at a high rate~\cite{ehrenberg_mediumdependent_2012}.
We develop a coarse-grained model of the cell, inspired by the self-replicator model of Molenaar \textit{et al.}~\cite{molenaar_shifts_2009}, and reformulate our questions in the context of optimal control theory~\cite{stengel_optimal_1994} to identify control schemes maximizing biomass production over an interval of time, the dynamical equivalent of growth-rate maximization.

We show that Pontryagin's Maximum Principle suggests that optimal resource allocation after a growth transition is achieved by a bang-bang-singular control law~\cite{stengel_optimal_1994}, a conjecture confirmed by direct numerical optimization.
This optimal solution provides a gold standard against which possible control strategies of the cell can be compared.
We consider simple strategies that drive the system to the steady state enabling growth at the maximal rate in the new medium, after the upshift.
In a dynamical growth scenario, the strategy sensing the concentration of precursor metabolites emerges as the best candidate, consistent with the analysis of Scott \textit{et al.} that feedforward activation of the rate of synthesis of ribosomal proteins, involving ppGpp-mediated sensing of the amino acid pool~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}, is the key regulatory mechanism for growth control.
It is possible, however, to define a strategy approaching the theoretical optimum even more closely by exploiting information on both the precursor concentration and the abundance of the gene expression machinery.
Interestingly, a thorough analysis of the functioning of the ppGpp system, as described by a kinetic model of the synthesis and degradation of this signalling molecule, suggests similarities between our two-variable control strategy and the regulation of the transcription of ribosomal RNA by ppGpp~\cite{bosdriesz_how_2015}.

The results presented here generalize the analysis of control strategies enabling optimal growth of microorganisms from steady-state to dynamical scenarios.
The control strategies are formulated in the context of a coarse-grained model of resource allocation, based on minimal assumptions, that accounts for empirical growth laws at steady state.
The analysis shows that during growth transitions, control strategies based on information of a single variable are outperformed by systems measuring several variables.
This conclusion agrees with the intuition that, in dynamical environments, there may be an evolutionary pressure towards more elaborate sensory systems.
From a methodological point of view, our study illustrates how optimal control theory can provide novel insights into complex biological phenomena~\cite{iglesias_control_2010}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Results}

\subsection*{Self-replicator model of resource allocation}
\label{sec:model}

Resource allocation in bacteria involves the distribution of cellular resources (precursor metabolites and energy) over processes supporting maintenance and growth~\cite{schaechter_microbe_2006}.
A simple modelling tool for analyzing resource allocation questions in a precise way are so-called self-replicator models.
These models have a long history in various domains of chemistry, biology, physics, and computer science~\cite{sipper_fifty_1998}, and were recently put to use as an analytical tool in systems biology~\cite{molenaar_shifts_2009} (see also~\cite{flamm_minimal_2007}).
We will show that despite their simplicity, which make them tractable for mathematical analysis, self-replicator models are sufficiently expressive to account for empirical observations and make testable predictions.

Bearing in mind that the major constituents of the cell are macromolecules (DNA, RNA, proteins), produced from precursor metabolites, a fundamental resource allocation question is the following: How much of the cellular resources are invested in the making of new macromolecules (gene expression machinery) and how much in performing other functions, in particular producing metabolic enzymes involved in the uptake of nutrients and their conversion to precursor metabolites (metabolic machinery)?
In order to address this question, we consider a self-replicating system composed of the gene expression machinery ($R$) and the metabolic machinery ($M$).
The system, shown schematically in Fig.~\ref{fig:self_replicator}, is thus defined by two macroreactions which are conveniently written as:
\begin{equation}\label{eq:reactions}
\begin{aligned}
S &\overset{V_M}{\longrightarrow} P, \\
P &\overset{V_R}{\longrightarrow} \alpha R + (1-\alpha) M.
\end{aligned}
\end{equation}
The first reaction, catalyzed by $M$, converts external substrates ($S$) into precursor metabolites ($P$).
The second reaction, catalyzed by $R$, converts precursors into macromolecules ($R$ and $M$).
The resource allocation parameter $\alpha \in [0,1]$ defines the proportion of precursor mass used for making gene expression machinery as compared to metabolic machinery.
We will interchangeably use the symbols $M$, $R$, $S$, and $P$ for the components of the replicators themselves and their total mass [g].
We will denote the rates at which the macroreactions occur by $V_R$ and $V_M$ [g~h\textsuperscript{-1}].

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

The self-replicator system in Fig.~\ref{fig:self_replicator} is based on a number of simplifying assumptions.
First, cell division is not explicitly modelled and replication should therefore be interpreted as the growth of (the mass of) a cell population.
This amounts to the assumption that individual cells in a growing populations have the same macromolecular composition.
Second, degradation of the macromolecules is ignored.
In other words, we assume that macromolecules are stable and that their degradation rates are negligible with respect to the rates of other reactions in the system.
Third, we consider only two classes of macromolecules ($R$ and $M$).
In particular, we do not assume that an irreducible mass fraction of the precursors is dedicated to cell maintenance~\cite{scott_interdependence_2010}.
The system could be easily extended to relax the above assumptions, but this would complicate the analysis of the model and obscure the points we want to make.

In what follows, it will be more convenient to describe the quantities in the system as intracellular concentrations rather than as the total mass in the cell population.
To this end, we define the volume $\mathtt{Vol}$ [L] of the cell population as follows:
\begin{equation}
\mathtt{Vol} = \beta \, (M + R),
\label{eq:voldef}
\end{equation}
with $\beta$ a conversion constant [L~g\textsuperscript{-1}] equal to the inverse of the cytoplasmic density.
Dividing each variable $M$, $R$, and $P$ by $\mathtt{Vol}$ yields the concentrations $m$, $r$, and $p$ of metabolic enzymes, ribosomes and other components of the gene expression machinery, and precursor metabolites, respectively [g~L\textsuperscript{-1}].
Henceforth, these variables as well as $\mathtt{Vol}$ and $\alpha$ will be considered functions of time $t$ [h].

The dynamics of the self-replicator in Fig.~\ref{fig:self_replicator} can be described by the following system of ordinary differential equations (see \ref{S1_Text} for the derivation):
\begin{eqnarray}
\frac{dp}{dt} &=& v_M(s,r) - v_R(p,r) \, (1+\beta\, p), \label{eq:pdef}\\
\frac{dr}{dt} &=& v_R(p,r) \, (\alpha(t) - \beta\, r), \label{eq:rdef}
\end{eqnarray}
where $s$ [g~L\textsuperscript{-1}] denote the (extracellular) concentration of substrate.
$v_M(s,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] and $v_R(p,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] denote the precursor synthesis rate and the macromolecule synthesis rate, respectively.
The growth rate $\mu$ [h\textsuperscript{-1}] of the replicator system is defined as the relative increase of the volume, and can be rewritten with Eqs~\ref{eq:pdef}-\ref{eq:rdef} as proportional to the macromolecule synthesis rate (\ref{S1_Text}):
\begin{equation}\label{eq:growthrate}
\mu = \frac{1}{\mathtt{Vol}} \frac{d\mathtt{Vol}}{dt} = \frac{1}{M+R}\frac{d(M+R)}{dt} = \beta\, v_R(p,r).
\end{equation}

The precursor concentration changes through the joint effect of the precursor synthesis rate $v_M(\cdot )$, the macromolecule synthesis rate $v_R(\cdot )$, and the rate of growth dilution ($\beta\, v_R(\cdot ) \,p$).
The change in concentration of ribosomes and other components of the gene expression machinery is the net effect of the ribosome synthesis rate ($\alpha (\cdot )\, v_R (\cdot)$) and the rate of growth dilution ($\beta \, v_R(\cdot) \, r $).
Remark that it is not necessary to add an equation for $m$ because it follows from Eq.~\ref{eq:voldef} that $r + m = 1/\beta$, and therefore $dm/dt = - dr/dt$.

We use Michaelis-Menten kinetics to define the synthesis rate of each reaction:
\begin{align}
v_M(s,r) &= k_M\,m\, \frac{s}{K_M + s} = k_M\,(1/\beta - r)\, \frac{s}{K_M + s},  \label{eq:metaflux} \\
v_R(p,r) &= k_R\,r\, \frac{p}{K_R + p}, \label{eq:machflux}
\end{align}
with rate constants $k_M, k_R$ [h\textsuperscript{-1}] and half-saturation constants $K_M, K_R$ [g~L\textsuperscript{-1}].
Note that the rate of precursor synthesis is proportional to the concentration of the components of the metabolic machinery, while the macromolecule synthesis rate is proportional to the concentration of the components of the gene expression machinery.
These catalytic effects correspond to the dashed arrows in Fig.~\ref{fig:self_replicator}.
The rate constant $k_M$ depends both on the quality of the nutrients in the medium (higher $k_M$ for a richer medium) and on the metabolic efficiency of the macroreaction converting the substrate into precursors (higher $k_M$ for a more efficient reaction).
For convenience, we henceforth assume that the environmental conditions do not change over the time-interval considered, either because $s$ is constant or because $s \gg K_M$, corresponding to a situation in which the substrate is available in excess.
In both cases, $e_M(s) = k_M\, s/(K_M + s)$ is approximately constant, so that we can write
\begin{equation}
\label{eq:metaflux_simplified}
v_M (r) = e_M \, (1/\beta-r).
\end{equation}
The rate constant $k_R$ characterizes the efficiency of the gene expression machinery, depending on the elongation rate of ribosomes, among other things. The ratio $p/K_R$ is an indicator of the saturation of the gene expression machinery by precursors.

The system of Eqs~\ref{eq:pdef}-\ref{eq:rdef} thus has four parameters ($e_M$, $k_R$, $K_R$, $\beta$), one of which characterizes the input from the environment ($e_M$).
The order of magnitude of the parameters can be inferred from data in the literature, as explained in \ref{S2_Text}.
Below we use the following values for the parameters $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1} (\ref{S1_Table}).
However, it should be emphasized that the conclusions of this paper do not depend on the exact quantitative values of these parameters.

An interesting property of the model is that it is built on minimal assumptions, basically the two macroreactions and the definition of the volume as proportional to the total mass of macromolecules.
Like in~\cite{molenaar_shifts_2009,pavlov_optimal_2013,scott_emergence_2014}, these assumptions directly lead to the expression of the growth rate in Eq.~\ref{eq:growthrate}, without additional assumptions.

\subsection*{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
\label{sec:growthlaws}

The nullcline for $r$ is given by $r=0$, $r=\alpha/\beta$, and $p=0$, while the nullcline for $p$ is defined by

\[
r = \frac{e_M}{\beta \left(e_M + k_R\, \frac{p}{K_R+p}\, (1+\beta p)\right)}.
\]
The nullclines define a single stable steady state $(p^*, r^*)$ (Fig.~\ref{fig:model_analysis}\textit{A} and \textit{Methods}).
At this steady state, the growth rate is constant and denoted by $\mu^*$.
The nullcline for $p$ is defined by the environment $e_M$.
The nullcline for $r$, and thus the location of the steady state with the associated growth rate, are given by $\alpha$.
Fig.~\ref{fig:model_analysis}\textit{B} shows the dependency of the steady-state growth rate $\mu^*$ on the resource allocation parameter $\alpha$.
As can be seen, $\mu^*$ is maximal for a specific, unique value of $\alpha$, which we denote $\alpha_{opt}^*$.
That is, the model predicts that there is a single optimal way to divide the precursor flux over the synthesis of the gene expression machinery and the metabolic machinery.
The same result, using a similar model, was obtained by Scott \textit{et al.}~\cite{scott_emergence_2014}.
The self-replicator model is simple enough to derive an algebraic expression for computing $\alpha^*_{opt}$ and the corresponding maximal growth rate $\mu^*_{opt}$ (\textit{Methods} and \ref{S1_Text}), which will simplify analysis of the system in later sections.

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

In order to validate the model, we verified that it can account for data on the macromolecular composition of \textit{E. coli} at steady state~\cite{scott_interdependence_2010}.
When optimizing $\alpha$ for different values of $e_M$ (assuming cells attain maximal growth), the model predicts a relation between $\alpha^*_{opt}$ and $\mu^*_{opt}$ (colored dots and black dashed line in Fig.~\ref{fig:model_validation}\textit{A}) that is quasi-linear for high growth rates.
We compared this prediction with the results of experiments where the relation between the growth rate and the mass ratio of total RNA and protein was determined in different growth media (Fig.~\ref{fig:model_validation}\textit{B}).
In the framework of our model, different media correspond to different values of $e_M$, and different total RNA/protein mass ratios to different values of $\alpha$ (up to a conversion factor), allowing a direct comparison of the model predictions in Fig.~\ref{fig:model_validation}\textit{A} with the data in Fig.~\ref{fig:model_validation}\textit{B} (see \textit{Methods}).
As can be seen, the model is able to account for the observed quasi-linear relation between the growth rate and the total mass ratio of RNA and protein. 
Moreover, for realistic values of $k_R$ and $e_M$, a good quantitative fit is obtained (\textit{Methods} and \ref{S1_Table}).

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig3.eps}
\caption[Self-replicator model accounts for bacterial growth laws]{
\textbf{Self-replicator model accounts for bacterial growth laws.}\\
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

The data from Scott \textit{et al.} also reveal a second apparently linear relation between the growth rate and the total RNA/protein mass ratio.
This relation is obtained when varying, in the same growth medium, the efficiency of protein synthesis by adding different doses of an inhibitor of translation (chloramphenicol)~\cite{scott_interdependence_2010}.
Using the model, we computed $\alpha^*_{opt}$ and $\mu^*_{opt}$, for constant environment $e_M$ and different values of the efficiency of protein synthesis $k_R$ (dashed colored lines in Fig.~\ref{fig:model_validation}\textit{A}).
As can be seen in Fig.~\ref{fig:model_validation}\textit{B}, the model also captures the second linear relation in the data.

We conclude that the self-replicator model is able to reproduce known observations of resource allocation in bacteria, so-called growth laws~\cite{scott_interdependence_2010}.
The model is similar to a model recently proposed by Scott \textit{et al.}~\cite{scott_emergence_2014}.
Contrary to the latter model, the translation rate is not assumed to be constant in the self-replicator model, but rather depends on precursor abundance, as proposed by the same authors in~\cite{klumpp_molecular_2013}.

The above analysis of bacterial growth has two major limitations.
First, the predictions of optimal resource allocation (the value of $\alpha$ leading to the maximal growth rate) hold at steady state, for a constant environment, whereas most bacteria are not expected to encounter such conditions outside the laboratory.
An allocation of resources that is optimal for steady-state growth and constant over time may not be optimal in dynamical growth conditions.
Second, while it predicts which value of $\alpha$ is optimal at steady state, the model says nothing about the strategies that could be used to control resource allocation and set $\alpha$ to its optimal value.
In other words, how could bacterial cells use sensors of changes in their internal state and the environment to optimally adjust $\alpha$?
In what follows, we will address the above two questions, after having given a precise statement of the problem of optimal resource allocation in a dynamical environment in the next section.

\subsection*{Biomass maximization as an optimal control problem}
\label{sec:optimalcontrolproblem}

A self-replicator at steady state accumulates biomass according to $\texttt{Vol}(0)\, e^{\mu^*\, t}$, $t\in [0,\tau]$, when $\mu^*$ is the growth rate at steady state. 
The accumulation of biomass is obviously maximal when the growth rate is maximal ($\mu^* = \mu^*_{opt}$).
In dynamical conditions, the growth rate is not constant and biomass accumulation is described more generally by: 
\begin{equation*}
\frac{d\texttt{Vol}}{dt} = \mu(t) \, \texttt{Vol}.
\end{equation*}
In other words, when integrating over the time interval $[0,\tau]$:
\begin{equation}\label{eq:biomassdyn}
\ln \left(\frac{\texttt{Vol}(\tau)}{\texttt{Vol}(0)}\right) =  \int_{0}^{\tau} \mu(t) \, dt.
\end{equation}
Since the logarithm is an increasing function, maximizing the biomass produced over $[0,\tau]$ requires maximization of the right-hand side of the equation.

In a changing environment, maximization of the integral in Eq.~\ref{eq:biomassdyn} will generally require the optimal value of $\alpha$ to be a function of time instead of a specific constant value.
This dynamical resource allocation problem can be formulated in a more precise way using concepts from optimal control theory~\cite{stengel_optimal_1994}. 
Let $J$ be the objective function  
\begin{equation*}
J(\alpha)= \int_{0}^{\tau} \mu(t) \, dt = \int_{0}^{\tau} \beta \, v_R(p,r) \, dt,
\end{equation*}
where $\alpha:\mathbb{R}^+ \rightarrow [0,1]$ is a time-dependent function. 
The time evolution of $p$ and $r$ is determined by the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}, and $p$ and $r$ thus depend on $e_M$ and $\alpha$. 
Moreover, let $\mathcal{U}=\{\alpha:\mathbb{R}^+ \rightarrow [0,1] \}$ be the set of admissible controls. 
The optimal dynamical control problem then consists in finding the time-varying function $\alpha_{opt}(t)$ that maximizes $J(\alpha)$ over the time-interval $[0,\tau]$:

\begin{equation}
\alpha_{opt} = \arg \underset{\alpha \in \mathcal{U}}{\max} \ J(\alpha).
\label{eq:J}
\end{equation}

In what follows, we will simplify the above problem by considering that the environment changes in a step-wise fashion at $t=0$, but remains constant over the time-interval $[0,\tau]$, that is, $e_M(t)=e_M$.
More specifically, we focus on the case of a nutrient upshift, corresponding to a step-wise increase of $e_M$.
This upshift scenario corresponds to classical experiments in bacterial physiology~\cite{kjeldgaard_kinetics_1961,schaechter_patterns_1961,johnsen_control_1977}, reviewed in~\cite{ehrenberg_mediumdependent_2012}, and is frequently encountered in the life cycle of a microorganism~\cite{schaechter_microbe_2006}.
Notice that more complex environments can be approximated by a sequence of step-wise nutrient upshifts and downshifts.

\subsection*{Solution of the optimal control problem}
\label{sec:optimalcontrolsolution}

Optimal dynamical control problems for two-dimensional nonlinear dynamical systems, like the problem of Eq.~\ref{eq:J}, are generally difficult to solve.
However, we will show that the class of functions to which $\alpha_{opt}$ belongs can be identified, and we will use numerical optimization to identify a particular $\alpha_{opt}$ maximizing $J$.

As a preliminary step, in order to simplify the analysis, the variables in the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef} are made nondimensional, by defining $\hat{t} = k_R\, t$, $\hat{p} = \beta\, p$, and $\hat{r} = \beta\, r$.
This leads to the following ODE system:
\begin{eqnarray}
\frac{d\hat{p}}{d\hat{t}} &=& (1-\hat{r})\, E_M - (1+\hat{p})\, \hat{r} \, \frac{\hat{p}}{K+\hat{p}}, \label{eq:pdefnondim}\\
\frac{d\hat{r}}{d\hat{t}} &=& \hat{r} \frac{\hat{p}}{K + \hat{p}} \, (\alpha (\hat{t}) - \hat{r}), \label{eq:rdefnondim}
\end{eqnarray}
where $K= \beta\, K_R$ and $E_M= e_M / k_R$.
The nondimensional growth rate is given by:
\begin{equation}
\label{eq:munondim}
\hat{\mu} = \frac{\mu}{k_R} = \frac{\hat{p}}{K + \hat{p}} \, \hat{r} .
\end{equation}
Notice that the nondimensionalized system depends on a single parameter $K$, in addition to the constant environment $E_M$, which functions as an input to the system.

Analysis of the nondimensionalized system allows a number of properties of the solution of the optimal control problem of Eq.~\ref{eq:J} to be derived (\textit{Methods} and \ref{S3_Text}).
First, by applying a version of the well-known Pontryagin Maximum Principle \cite{carlson_infinite_1991}, we can prove that the optimal solution is obtained for an alternating sequence of $\alpha(\cdot)=0$ and $\alpha(\cdot)=1$, possibly ending with an intermediate value of $\alpha(\cdot)$, corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the post-upshift environment $E_M$.
Second, if the optimal solution reaches the optimal steady state for the new environment, then it does so after an infinite number of switches of $\alpha (\cdot)$ between 0 and 1.
Third, this switching behavior is characterized by a so-called switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
The switching curve divides the phase plane into two regions, such that $\alpha(\cdot)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (black dashed curve in Fig.~\ref{fig:optimalcontrol}\textit{A}).

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig4.eps}
\caption[Optimal control of the self-replicator during a nutrient upshift]{
{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

In line with these results, we conjecture that the optimal solution consists in a switching transient towards the optimal steady state for the new environment, and remains at this steady state until the next environmental change.
Such a solution is known as a bang-bang-singular solution in the control theory literature~\cite{stengel_optimal_1994}.
Formally, the solution of Eq.~\ref{eq:J} can be described as
\begin{equation}
\label{eq:optcontrol}
\alpha_{opt}(\hat{t})=
\begin{cases}
0, \ \textrm{if} \ \hat{r}(\hat{t}) > \varphi(\hat{p}(\hat{t})),\\
1, \ \textrm{if} \ \hat{r}(\hat{t}) < \varphi(\hat{p}(\hat{t})), \\
\alpha^*_{opt}, \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}
Notice that the optimal solution involves dynamical feedback from the state of the system to the control variable $\alpha(\cdot)$, and is therefore an instance of closed-loop optimization~\cite{stengel_optimal_1994}.

The optimal control problem of Eq.~\ref{eq:J} was also solved numerically by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012} (see \textit{Methods} for details).
A time discretization allows the problem to be transformed into a nonlinear optimization problem solved here by interior point techniques.
The optimal trajectories obtained numerically confirm our conjecture that the optimal control is bang-bang-singular.
An example solution, obtained by numerical optimization is shown in Fig.~\ref{fig:optimalcontrol}.
At time $\hat{t}=0$, $E_M$ jumps from a low to a high value, corresponding to a nutrient upshift (dashed black line in Fig.~\ref{fig:optimalcontrol}B).
The optimal solution $\alpha_{opt}$ consists of a sequence of switches between $\alpha=1$, corresponding to maximal synthesis of the gene expression machinery, and $\alpha=0$, corresponding to maximal synthesis of the metabolic machinery, until $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ is reached.
$\alpha$ is then set to $\alpha^*_{opt}$, the value leading to the maximum growth rate in the new medium (here 0.5, for $E_M=1$).
The sequence of switches of $\alpha$ in Fig.~\ref{fig:optimalcontrol}\textit{B} corresponds to successive crossings of the switching curve in Fig.~\ref{fig:optimalcontrol}\textit{A}.
In particular, the switch just after $\hat{t}=2$ corresponds to the first crossing of the switching curve; the subsequent switches accumulate around the steady state and are therefore difficult to identify in the plot.

What is the biological relevance of the bang-bang-singular solution maximizing growth of the bacterial self-replicator?
In order to answer this question, we will investigate in the next two sections the different ways in which microorganisms could implement or have been shown to implement feedback growth control by sensing the environment and cellular physiology.
Although the idealized solution proposed by optimal control theory will obviously not be found in nature, actual control strategies may produce solutions that are close.
The optimal solution can thus be used as a gold standard, a benchmark for comparing actual control strategies.

\subsection*{Simple feedback control strategies: exploiting information on nutrients or precursors}
\label{sec:steadystate}

The control strategies that microbial cells have evolved to bring resource allocation in line with changes in the environment involve a variety of molecular mechanisms~\cite{chubukov_coordination_2014}.
These mechanisms are responsible for sensing the environment and the physiological state of the cell, as well as for adjusting the expression of genes that encode components of the transcriptional and translational machinery, enzymes, transporters, and proteins with other metabolic functions.

In the framework of the self-replicator model of bacterial growth, control strategies take the form of feedback control laws mapping the value of system variables to a value of the control variable $\alpha(\cdot)$.
In this section, we explore two such strategies, the first exploiting information on the quality and quantity of substrate present in the environment, as reflected in the value of $E_M$, and the second using information on the precursor concentration $\hat{p}$.
The feedback control strategies are graphically displayed in Fig.~\ref{fig:strategies_overview}, as an extension of the self-replicator of Fig.~\ref{fig:self_replicator}.
We pose a number of mathematical constraints on the feedback control strategies considered below.
First, we require the control laws to be functions of the variables of the self-replicator but not involve derivatives or integrals of these variables.
Second, for a constant environment $E_M$, the control strategies must drive the system to a unique stable and non-trivial steady state, enabling a non-zero growth rate.
Third, this steady state must equal the optimal steady state for that environment, given by $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig5.eps}
\caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

The first control strategy is defined by the function $f \colon \mathbb{R}^+ \to [0,1]$, mapping $E_M$ to $\alpha$:
\begin{equation}
\alpha = f(E_M).
\label{eq:f}
\end{equation}
Notice that $\alpha$ is constant because $E_M$ is fixed to the value defining the new environment after the upshift.
What would be an appropriate choice for $f$?
An advantage of the self-replicator model is that the optimal allocation at steady state can be explicitly formulated as a function of $E_M$ (Eq.~\ref{eq:alpha_mu_optimal} in \textit{Methods}, with derivation in \ref{S1_Text}).
This function is the unique function satisfying all of the above criteria (\ref{S1_Text}).
\ref{S1_Fig} plots $f$ and shows that it is conveniently approximated by a Michaelis-Menten function, \textit{i.e.},
\begin{equation}\label{eq:MMApprox}
\alpha(\cdot) = \dfrac{E_M}{E_M + K_{mE}},
\end{equation}
with the dimensionless half-saturation constant $K_{mE}$.
The interest of the approximation is that it demonstrates that the control strategy can be described by a simple and ubiquitous response curve in biochemical kinetics.

As an example of a regulatory system resembling the above control strategy consider the phosphotransferase system responsible for the uptake of glucose, the preferred substrate of \textit{E. coli}~\cite{deutscher_how_2006}.
In the presence of glucose, the EIIA\textsuperscript{Glc} component of the phosphotransferase system is mostly unphosphorylated, since the phosphate groups are used for the conversion of extracellular glucose to intracellular glucose-6-phosphate.
When glucose disappears from the medium, however, the glucose uptake rate decreases and, correspondingly, the phosphorylated fraction of EIIA\textsuperscript{\textrm{Glc}} increases.
The phosphorylation state of EIIA\textsuperscript{\textrm{Glc}} thus provides an indirect read-out of glucose availability.
In response to this signal, a variety of metabolic processes are upregulated or downregulated, notably involving the signalling molecule cAMP which activates the pleiotropic transcription factor Crp~\cite{deutscher_how_2006,gorke_carbon_2008}.

How does the control strategy of Eq.~\ref{eq:f}, which we call a nutrient-only strategy, perform in comparison with the optimal solution derived in the previous section?
That is, how much biomass does this strategy produce compared with the maximal amount of biomass that can theoretically be obtained after a nutrient upshift?
In order to answer these questions, we simulated the response to a sudden upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} controlled by the nutrient-only strategy of Eq.~\ref{eq:f}.
The results are shown in Fig.~\ref{fig:comparison_nutrients_precursor}.
Panel~\textit{A} shows the trajectory of the controlled self-replicator system and panel~\textit{D} plots the evolution of the amount of biomass as a fraction of the amount of biomass produced by the optimal strategy.
While the system does reach the steady state that is optimal for $E_M$, the nutrient-only strategy has poor performance in the transient phase immediately following the nutrient upshift.
As can be seen from the solution trajectory in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{A}, fixing $\alpha$ to the value that enables optimal growth at steady state leads to a huge transient overshoot of the precursor concentration.
The overshoot reveals that resource allocation is initially suboptimal, with too many resources invested in the metabolic machinery at the expense of the gene expression machinery.
This causes a transiently suboptimal growth rate, leading to lower biomass accumulation (Eq.~\ref{eq:biomassdyn}).

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig6.eps}
\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

One way to avoid the transient precursor imbalance observed in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{A} would be to exploit information on the precursor concentration in the control strategy.
The second strategy considered here, which we label a precursor-only strategy, does exactly this: it involves a feedback control law $g \colon \mathbb{R}^+ \to [0,1]$ mapping $\hat{p}$ to $\alpha$:

\begin{equation}
\alpha = g(\hat{p}).
\label{eq:precursorstrategy}
\end{equation}
Since $\hat{p}$ will vary during the upshift experiment, $\alpha$ is not constant, contrary to the nutrient-only strategy above.
In the \textit{Methods} section, we present a function $g$ satisfying the requirements listed in the beginning of this section, in particular that the system converge to a stable steady state ensuring maximal growth in the new environment.
Moreover, we show that any other choice for $g$ leads to lower biomass production.
The function is plotted in \ref{S1_Fig}, and as shown in the same panel, is conveniently approximated by a Hill function with cooperativity coefficient 2:
\begin{equation}\label{eq:HillApprox}
\alpha (\cdot)= \dfrac{\hat{p}^2}{\hat{p}^2 + {K_{mp}}^2},
\end{equation}
where $K_{mp}$ is a dimensionless half-saturation constant.

While converging to the same steady state, this second strategy, which we will refer to as the precursor-only strategy, performs much better than the nutrient-only strategy after an upshift, as shown in Fig.~\ref{fig:comparison_nutrients_precursor}.
We simulated the response to a nutrient upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} with the precursor-only strategy of Eq.~\ref{eq:precursorstrategy}.
The relative biomass increases by 51\% and reaches 94\% of the biomass produced by the optimal control strategy (the theoretical maximum).
The precursor-only strategy notably avoids the inefficient transient accumulation of precursors directly after the nutrient upshift, by alternatingly investing more resources in gene expression (consumption of precursors) and metabolism (production of precursors).
In this respect, the oscillatory time profile of $\alpha$ (Fig.~\ref{fig:comparison_nutrients_precursor}\textit{C}) is somewhat reminiscent of the bang-bang-singular control in the solution of the optimal control problem (Fig.~\ref{fig:optimalcontrol}\textit{B}).

Both strategies, nutrient-only and precursor-only, drive the self-replicator towards the same steady state.
Whereas the two strategies are thus indistinguishable when the analysis is restricted to steady state, the precursor-only strategy is shown to perform much better in a dynamical upshift scenario, in the sense that the biomass produced is much closer to that produced by the optimal strategy.
Several authors have concluded that control strategies based on precursor sensing are key for maintaining optimal growth at steady state.
Scott \textit{et al.} argue that a strategy similar to the precursor-only approach above allows robust control of amino acid supply and demand, resulting in optimal steady-state growth over a range of nutrient conditions~\cite{scott_emergence_2014}.
They associate this strategy with ppGpp-mediated control of the synthesis of ribosomal proteins~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}.
The signalling molecule ppGpp accumulates in response to an increase in the level of uncharged tRNA, when amino acid concentrations in the cell drop. 
This causes ribosomes to "stall" and leads to RelA-mediated conversion of GTP to ppGpp, the molecular details of which are still subject of debate~\cite{english_singlemolecule_2011,hauryliuk_recent_2015}.
Since ppGpp inhibits the transcription of ribosomal RNAs~\cite{dennis_control_2004}, the concentration of the latter decreases, leading to more inactive ribosomal proteins and, through a well-characterized post-transcriptional autoregulatory mechanism, a lower synthesis rate of ribosomal proteins~\cite{keener_regulation_1996,potrykus_pppgpp_2008}.
Our analysis adds to the above study a novel insight: measuring precursors does not only enable resource allocation control to achieve maximal growth at steady state, but is also a good strategy in a dynamical context.

While the precursor-only strategy is thus seen to lead to good results, Fig.~\ref{fig:comparison_nutrients_precursor}\textit{D} shows that there remains room for improvement.
It seems reasonable to expect that control strategies exploiting information of not just a single variable, but several variables simultaneously, could further improve the performance of the self-replicator during a growth transition.

\subsection*{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
\label{sec:strategies}

In the quest for further improvements, a natural starting-point would be to consider the curve defining the optimal steady states $(\hat{p}^*_{opt},\hat{r}^*_{opt})$  for different environments $E_M$.
This curve is defined by a function mapping $\hat{p}^*$ to $\hat{r}^*$, which is actually the same as the function $g$ introduced in the precursor-only strategy (\textit{Methods} and \ref{S1_Fig}), given that at steady state $\hat{r}=\alpha$ (Eq.~\ref{eq:rdefnondim}).
The curve can be seen as representing an optimal balance between precursors and the gene expression machinery, in the sense that the maximal growth rate attainable for a given precursor concentration $\hat{p}$ requires a concentration $\hat{r}$ of ribosomes and other components of the gene expression machinery equal to $g(\hat{p})$. 
If either $\hat{r}>g(\hat{p})$ or $\hat{r}<g(\hat{p})$, the growth rate is suboptimal.

These considerations suggest an intuitive control strategy, namely to avoid an imbalance between $\hat{p}$ and $\hat{r}$ at all times, and remain as close as possible to the curve defined by $g$.
In particular, when the gene expression machinery is more abundant than what is optimal given the available precursors ($\hat{r} > g(\hat{p})$), its synthesis is switched off ($\alpha = 0$).
Conversely, when $\hat{r} < g(\hat{p})$, synthesis of the gene expression machinery is switched on.
This strategy thus tries to restore "as quickly as possible" the optimal balance between precursors $\hat{p}$ and the gene expression machinery $\hat{r}$, giving rise to a so-called on-off control strategy:
\begin{equation}
\label{eq:stratswitch}
\alpha = h(\hat{p},\hat{r}) =
\begin{cases}
0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
\alpha_{opt}^* \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}

As shown in the \textit{Methods} section, the on-off strategy drives the system to a stable steady state ensuring growth at the maximal rate.
Notice that, contrary to the strategies discussed in the previous section, the value of $\alpha$ selected by the on-off strategy depends on both $\hat{p}$ and $\hat{r}$ (Fig.~\ref{fig:strategies_overview}).
It thus uses more information on the state of the system than the nutrient-only and precursor-only strategies.

Fig.~\ref{fig:onoffresults} shows the performance of the on-off strategy after a nutrient upshift, as compared to the precursor-only strategy.
The transition is seen to be nearly perfect, in the sense that 98\% of the optimal biomass is produced by the strategy.
The time course of $\alpha$ in panel \textit{D} is very similar to the optimal time course obtained by numerical optimization, shown in Fig.~\ref{fig:optimalcontrol}\textit{B}, and clearly brings out the bang-bang-singular nature of the solution. 
These results show that a strategy exploiting complete information on the internal state of the self-replicator can lead to near-optimal performance, outcompeting a strategy that uses partial information on the internal state (precursor abundance only).

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig7.eps}
\caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
{
{\bf Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}\newline
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

Are microbial cells equipped with mechanisms implementing a strategy similar to the on-off strategy?
A possible candidate would again be the ppGpp system.
A kinetic model of ppGpp metabolism and the regulation of the synthesis of ribosomal proteins was recently presented by Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}.
The model proved capable of accounting for a range of experimental data, including the steady-state concentration of ppGpp as a function of the growth rate~\cite{bremer_modulation_1996} and the dynamical response of ppGpp to a nutrient upshift or downshift~\cite{murray_control_2003}.
A major conclusion of the model is that the steady-state concentration of ppGpp exhibits a strongly ultrasensitive response to deviations of the ribosomal protein fraction from the optimal ribosomal protein fraction at a given growth rate. 
These deviations from optimality, in turn, lead to a switch-like response of the synthesis rate of ribosomal proteins (Fig.~4 in Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}).

How does this mechanistic model of ppGpp regulation relate to the on-off strategy presented above?
In order to answer this question, we first need to find a correspondence between the variables $p$ and $r$ of our coarse-grained model and the concentrations of molecular species in the kinetic model of Bosdriesz \textit{et al.}
This is rather straightforward to achieve, by equating $p$ to the total amino acid concentration and $r$ to the ribosome concentration.
Second, \ref{S4_Text} shows that by making two simplifying assumptions, ppGpp can be expressed as a function of the total amino acid concentration and the ribosome concentration.
In particular, we assume that concentrations of all individual amino acids are equal, and that the concentrations of charged tRNAs and ppGpp evolve fast relative to the dynamics of the amino acid and ribosome concentrations.
The third step consists in positing an explicit relation between ppGpp and $\alpha$, based on the regulatory action of ppGpp on the transcription of ribosomal RNA~\cite{dennis_control_2004}:

\begin{equation}
\alpha(\cdot) = \frac{K_I}{K_I + \texttt{ppGpp}(\cdot)},
\label{eq:ppGpp_empirical}
\end{equation}
with $K_I$ a Michaelis-Menten inhibition constant [$\mu$mol~L\textsuperscript{-1}] and $\texttt{ppGpp}$ the (time-varying) intracellular concentration of ppGpp [$\mu$mol~L\textsuperscript{-1}].

The response function for ppGpp thus obtained and evaluated for a range of amino acid and ribosome concentrations is represented in Fig.~\ref{fig:ppGppsurface}, and visually compared with the on-off strategy.
As can be seen, the two response surfaces are very similar.
In other words, the ultrasensitive response of the synthesis rate of ribosomal proteins to the suboptimal allocation of cellular resources, derived from a model of the molecular mechanisms involved in the synthesis, degradation, and regulatory action of ppGpp~\cite{bosdriesz_how_2015}, implements a control strategy that is close to the optimal predicted by a control-theoretical analysis of the self-replicator.
While the role of ppGpp in maintaining optimal resource allocation was already pointed out by Scott \textit{et al.} and Bosdriesz \textit{et al.}, the latter studies were restricted to optimizing steady-state growth.
A major insight from the analysis in this section is that this conclusion seems to carry over to dynamical scenarios as well.
Fundamentally, the analysis suggests that the ppGpp system is a likely candidate to fulfill this role because it integrates information on the imbalance between precursor concentration and abundance of the gene expression machinery.

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig8.eps}
\caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
{
{\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

\section*{Discussion}
\label{sec:discussion}

Quantitative growth laws are empirical regularities pointing at fundamental properties of microbial life~\cite{scott_bacterial_2011}.
Recent work has led to the precise theoretical formulation of growth laws and has shown that they can be derived from basic assumptions on the molecular processes responsible for the assimilation of nutrients and their conversion to biomass~\cite{bosdriesz_how_2015,maitra_bacterial_2014,molenaar_shifts_2009,scott_emergence_2014,weisse_mechanistic_2015}.
The growth laws are uniquely defined under the hypothesis that microorganisms allocate resources in such a way as to maximize their growth rate.
Several of the above-mentioned studies have analyzed feedback control strategies on the molecular level enabling cells to achieve optimal resource allocation in a robust manner.
The control strategies exploit information on the physiological state of the cell to adjust the (relative) rate of synthesis of different classes of proteins (ribsomes, metabolic enzymes, $\ldots$).
Whereas the growth laws describe microbial growth at steady state, most microorganisms live in complex, continuously changing environments.
Despite some precursory work~\cite{pavlov_optimal_2013,vandenberg_optimal_1998}, questions about the dynamics of microbial growth remain largely unanswered:
Which resource allocation schemes are optimal in changing environments? Which dynamical control strategies lead to (near-)optimal resource allocation? How do these strategies compare with those actually implemented by microorganisms?

We have addressed the above questions by means of a self-replicator model of microbial growth, which, like other coarse-grained models of bacterial growth~\cite{maitra_bacterial_2014,molenaar_shifts_2009,scott_emergence_2014}, is capable of reproducing the growth laws at steady state (Fig.~\ref{fig:model_validation}).
A first major contribution of our work is to show that, in the case of a dynamical upshift scenario, optimal production of biomass requires a bang-bang-singular resource allocation scheme (Fig.~\ref{fig:optimalcontrol}).
That is, the optimal self-replicator should iteratively allocate all of its resources to the gene expression machinery (bang control input) and the metabolic machinery (another bang control input), until the steady state enabling maximal growth in the post-upshift environment is reached, corresponding to a trade-off in the allocation of resources to the two processes (singular control input).

Bang-bang phenomena are widespread in a variety of life processes.
Applications of optimal control theory to reproductive strategies in insects~\cite{macevicz_modeling_1976}, the development of intestinal crypts~\cite{itzkovitz_optimality_2012}, and the activation of metabolic pathways~\cite{bartl_modelling_2010,oyarzun_sequential_2009} have led to bang-bang or bang-bang-singular strategies.
In optimal control problems, such a solution arises with systems where the differential equations are linear in the control variable (in our case, $\alpha (\cdot)$).
Examples of applications that are close to the problem considered here are the control of gene expression for adaptation to environmental changes~\cite{pavlov_optimal_2013, madar_promoter_2013}, and the allocation of resources between nutrient uptake and growth in microorganisms~\cite{vandenberg_optimal_1998, kiselev_resource_2009}.
Whereas the former applications focus on minimization of response times, the latter also optimize biomass during a growth transition, using a different model, not derived from first principles as in this study.
However, the optimal solution of the corresponding optimal control problem is also bang-bang-singular, thus showing that our conclusions are robust to model variations.

Our second major contribution is the assessment of how different feedback control strategies perform with respect to each other and to the gold standard determined from optimal control theory.
We show that the precursor-only and nutrient-only strategies, both of which drive the self-replicator to the steady state with maximal growth rate in a static environment, perform quite differently in a dynamical upshift scenario (Fig.~\ref{fig:comparison_nutrients_precursor}).
While the precursor-only strategy is better than the nutrient-only strategy in a dynamical environment, it is in turn outperformed by a so-called on-off strategy, which achieves a near-perfect growth transition by exploiting information on the imbalance between the precursor concentration and the abundance of the gene expression machinery (Fig.~\ref{fig:onoffresults}).
The superior performance of the on-off strategy can be intuitively explained by the fact that during a growth transition the two variables are not fully correlated, which means that sensing both instead of either one provides additional information in a dynamical context.

Interestingly, the on-off strategy is based on a feedback control law that very much resembles the response function for ppGpp-mediated regulation of the synthesis of ribosomal RNAs in \textit{E. coli}~\cite{bosdriesz_how_2015}.
The role of ppGpp in controlling microbial growth has been amply documented~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}.
For example, Potrykus \textit{et al.} observed that in cells without ppGpp (ppGpp$^0$ mutants) the RNA/protein mass ratio, a proxy for our resource allocation variable $\alpha$, does not change with the growth rate, which has led these authors to conclude that ppGpp is the major source of growth-rate control in \textit{E. coli}~\cite{potrykus_ppgpp_2011}.
The central importance of ppGpp in the reallocation of gene expression resources in \textit{E. coli} following changes in nutrient availability has also been mapped with higher resolution, using genome-wide transcriptome studies~\cite{traxler_guanosine_2006,traxler_global_2008}.
In nearly all bacterial species examined so far, ppGpp is known to accumulate in response to an increase in the level of uncharged tRNA~\cite{gaca_many_2015}, although the molecular details of ppGpp metabolism and the range of other functions of the alarmone may greatly vary across species~\cite{gaca_many_2015,hauryliuk_recent_2015,liu_diversity_2015}.
While it has thus been well-established that regulation by ppGpp is an evolutionary conserved mechanism of growth control in the bacterial cell, our analysis provides a new perspective by suggesting that ppGpp enables optimal reallocation of resources after a growth transition, dynamically maximizing the accumulation of biomass.

The model on which the above results are based is built from first principles by distinguishing two fundamental cellular processes: metabolism (converting nutrients to precursors) and gene expression (converting precursors to the proteins that make up biomass) (Fig.~\ref{fig:self_replicator}). 
Despite its simplicity, our self-replicator model is capable of reproducing the empirical growth laws and of making testable predictions on the time-course profile of the resource allocation variable $\alpha$ and on the concentrations $p$ and $r$ of components of the gene expression machinery and metabolic machinery, respectively (see Fig.~\ref{fig:ppGppsurface} and below). 
The model can be easily extended with more details on protein synthesis, central carbon and energy metabolism, stress systems, or cell membranes, but this would make the mathematical analysis of the model dynamics and the optimal control problem more complicated. 
Notice, however, that the direct numerical approach for solving the optimal control problem remains applicable, even for more fine-grained models (Fig.~\ref{fig:optimalcontrol}, see also~\cite{waldherr_dynamic_2015}).

The comparison of different control strategies during a classical growth transition should be interpreted carefully, in a qualitative rather than quantitative manner. 
Whereas the differences in performance based on the biomass ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ of the control strategies are robust, the absolute numbers for the biomass ratio will depend on details of the growth experiment chosen and the exact parameter values. 
Another implicit assumption in the analysis of the control strategies is that the costs of their molecular implementation can be neglected.
This is not true in general, since every control strategy requires resources to be diverted towards the synthesis of sensory systems and regulatory proteins, with possibly detrimental effects on growth.
In other words, a control strategy entails a trade-off between the growth burden of regulation and the growth benefit of the improved capability to adapt to changes in the environment~\cite{kalisky_costbenefit_2007,poelwijk_tradeoffs_2011}. 
The analysis of control strategies could be refined by adding a reaction to the self-replicator that models the loss of resources incurred by regulatory strategies.
While in general the growth burden of a control strategy requiring information on several aspects of cellular physiology is expected to be higher, notice that a single regulatory system may be capable of sensing more than one variable.
For example, we show that ppGpp levels in the cell carry information on both the metabolic and the gene expression state (Fig.~\ref{fig:ppGppsurface}), thus integrating several signals in a cost-efficient manner.

The model predictions for the dynamical adaptation of resource allocation after a nutrient upshift suggest several interesting experimental tests.
In particular, the switching profile of the resource allocation variable $\alpha$ is a promising candidate for experimental validation. 
The most straightforward option would be direct measurement of the synthesis rate of ribosomal proteins, using a translational fusion of a fluorescent reporter with a ribosomal protein~\cite{bakshi_superresolution_2012,english_singlemolecule_2011}.
However, a more indirect approach based on the quantification of ppGpp concentrations in the cell or the activity of the ribosomal RNA (rRNA) promoters would also be a possibility.
Interestingly, some data are already available in the literature.
For instance, Gausing has reviewed data on the synthesis of ribosomal proteins after a nutrient upshift, showing that the synthesis rate goes through "a series of rapid changes" resembling oscillations~\cite{gausing_regulation_1980}.
Later work attributed this pattern to regulation on the transcriptional level~\cite{zengel_transcription_1986}.
Friesen \textit{et al.} observed oscillatory patterns in ppGpp concentrations after a nutrient upshift, with an initial response resembling bang control for an upshift to a particularly rich medium~\cite{friesen_synthesis_1975}.
Murray \textit{et al.} also present data on the ppGpp concentration after a nutrient upshift~\cite{murray_control_2003}, but with a lower temporal resolution and no clear oscillatory pattern.
All of the above measurements were carried out on the population level, which means that switching patterns may be obscured by desynchronisation of the individual cells.
More sophisticated experimental set-ups are necessary for the decisive validation of the model predictions, allowing gene expression in single cells to be followed over time in tightly regulated growth conditions~\cite{young_measuring_2011,duncombe_microfluidics_2015}.
In addition, the model could be validated on other dynamical scenarios, for example nutrient downshifts~\cite{molin_control_1977,murray_control_2003}.

Apart from its interest for fundamental science, resource allocation is also a critical question in biotechnology, where there exists an inherent trade-off between the maximization of yield and productivity\cite{venayak_engineering_2015}.
High yield means that most of the substrate is converted to a metabolite, peptide or recombinant protein of interest, but this leads to low productivity if the remaining nutrient influx is insufficient to sustain population growth.
Engineered control of resource allocation may help in establishing the right trade-off, the most profitable balance between yield and productivity, in a biotechnological process.
Such a trade-off could be attained either in steady-state conditions (the incoming nutrient flux is optimally distributed over growth and production) or in dynamical conditions (alternating utilization of the incoming nutrient flux for growth or production)~\cite{dahl_engineering_2013,xu_improving_2014,izard_synthetic_2015}.
When extended with heterologous metabolic pathways, the self-replicator models used in this study would provide an adequate \textit{in-silico} test bed for the rapid screening and comparison of alternative control strategies in bioprocess engineering.

### 2016-06-01 Chapters/core_label.tex
First meeting with H&H

--- 2016-06-01 Chapters/core.tex
@@
 The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
-Along these lines, Molenaar \textit{et al.} developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
+Along these lines, Molenaar \etal developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
 Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
-Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \textit{et al.} derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
+Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \etal derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
 In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
@@
 Several studies have used coarse-grained models to understand which control strategies microorganisms employ to achieve (optimal) resource allocation~\cite{bosdriesz_how_2015,scott_emergence_2014,weisse_mechanistic_2015}.
-Scott \textit{et al.} have shown that a robust feedforward control strategy, based on the sensing of the amino acid pool size and the corresponding adjustment of the fraction of ribosomes producing ribosomal proteins, allows the ribosomal protein fraction to be maintained close to its optimal value under a variety of growth conditions~\cite{scott_emergence_2014}.
+Scott \etal have shown that a robust feedforward control strategy, based on the sensing of the amino acid pool size and the corresponding adjustment of the fraction of ribosomes producing ribosomal proteins, allows the ribosomal protein fraction to be maintained close to its optimal value under a variety of growth conditions~\cite{scott_emergence_2014}.
 The authors suggest that this control strategy involves the signalling molecule ppGpp, in agreement with conclusions drawn from a recent kinetic model of the regulatory mechanisms achieving optimal adjustment of the ribosomal protein fraction~\cite{bosdriesz_how_2015}.
-Wei{\ss}e \textit{et al.} also developed a coarse-grained model of microbial growth based on resource allocation trade-offs~\cite{weisse_mechanistic_2015}. 
+Wei{\ss}e \etal also developed a coarse-grained model of microbial growth based on resource allocation trade-offs~\cite{weisse_mechanistic_2015}. 
 Without including specific regulatory interactions, the model accounts for the above-mentioned bacterial growth laws, predicts host-circuit interactions in synthetic biology, and relates gene regulation to the nutrient composition of the medium.
 
-The above studies consider resource allocation at steady state, where all intensive variables describing the growing microbial culture, in particular the concentrations of its molecular components, are constant (see~\cite{fishov_microbial_1995} for a precise definition of steady-state growth and the closely related notions of balanced and exponential growth).
+The above studies consider resource allocation at steady state, where all intensive variables describing the growing microbial culture, \tr{in particular the concentrations of its molecular components, are constant (see~\cite{fishov_microbial_1995} for a precise definition of steady-state growth and the closely related notions of balanced and exponential growth).}
 This requires an environment to be stable over a long period of time.
@@
 In particular, we consider the upshift of a microbial culture from a medium supporting growth at a low rate to a medium supporting growth at a high rate~\cite{ehrenberg_mediumdependent_2012}.
-We develop a coarse-grained model of the cell, inspired by the self-replicator model of Molenaar \textit{et al.}~\cite{molenaar_shifts_2009}, and reformulate our questions in the context of optimal control theory~\cite{stengel_optimal_1994} to identify control schemes maximizing biomass production over an interval of time, the dynamical equivalent of growth-rate maximization.
+We develop a coarse-grained model of the cell, inspired by the self-replicator model of Molenaar \etal~\cite{molenaar_shifts_2009}, and reformulate our questions in the context of optimal control theory~\cite{stengel_optimal_1994} to identify control schemes maximizing biomass production over an interval of time, the dynamical equivalent of growth-rate maximization.
 
@@
 We consider simple strategies that drive the system to the steady state enabling growth at the maximal rate in the new medium, after the upshift.
-In a dynamical growth scenario, the strategy sensing the concentration of precursor metabolites emerges as the best candidate, consistent with the analysis of Scott \textit{et al.} that feedforward activation of the rate of synthesis of ribosomal proteins, involving ppGpp-mediated sensing of the amino acid pool~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}, is the key regulatory mechanism for growth control.
+In a dynamical growth scenario, the strategy sensing the concentration of precursor metabolites emerges as the best candidate, consistent with the analysis of Scott \etal that feedforward activation of the rate of synthesis of ribosomal proteins, involving ppGpp-mediated sensing of the amino acid pool~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}, is the key regulatory mechanism for growth control.
 It is possible, however, to define a strategy approaching the theoretical optimum even more closely by exploiting information on both the precursor concentration and the abundance of the gene expression machinery.
@@
 
-\begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig1.eps}
+\begin{figure}[h]
+\includegraphics[scale=1]{./Fig/Fig1.eps}
 \caption[Self-replicator model of bacterial growth]{
@@
 
-The dynamics of the self-replicator in Fig.~\ref{fig:self_replicator} can be described by the following system of ordinary differential equations (see \ref{S1_Text} for the derivation):
+The dynamics of the self-replicator in Fig.~\ref{fig:self_replicator} can be described by the following system of ordinary differential equations (see Text~\ref{S1_Text} for the derivation):
 \begin{eqnarray}
@@
 $v_M(s,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] and $v_R(p,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] denote the precursor synthesis rate and the macromolecule synthesis rate, respectively.
-The growth rate $\mu$ [h\textsuperscript{-1}] of the replicator system is defined as the relative increase of the volume, and can be rewritten with Eqs~\ref{eq:pdef}-\ref{eq:rdef} as proportional to the macromolecule synthesis rate (\ref{S1_Text}):
+The growth rate $\mu$ [h\textsuperscript{-1}] of the replicator system is defined as the relative increase of the volume, and can be rewritten with Eqs~\ref{eq:pdef}-\ref{eq:rdef} as proportional to the macromolecule synthesis rate (Text~\ref{S1_Text}):
 \begin{equation}\label{eq:growthrate}
@@
 The system of Eqs~\ref{eq:pdef}-\ref{eq:rdef} thus has four parameters ($e_M$, $k_R$, $K_R$, $\beta$), one of which characterizes the input from the environment ($e_M$).
-The order of magnitude of the parameters can be inferred from data in the literature, as explained in \ref{S2_Text}.
-Below we use the following values for the parameters $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1} (\ref{S1_Table}).
+The order of magnitude of the parameters can be inferred from data in the literature, as explained in Text~\ref{S2_Text}.
+Below we use the following values for the parameters $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1} (Table \ref{S1_Table}).
 However, it should be emphasized that the conclusions of this paper do not depend on the exact quantitative values of these parameters.
@@
 
+%The dynamics of the self-replicator model can be easily analyzed in the phase plane $(p,r)$, assuming constant resource allocation $\alpha(t)=\alpha$ and a constant environment $e_M$.
 The nullcline for $r$ is given by $r=0$, $r=\alpha/\beta$, and $p=0$, while the nullcline for $p$ is defined by
@@
 \]
-The nullclines define a single stable steady state $(p^*, r^*)$ (Fig.~\ref{fig:model_analysis}\textit{A} and \textit{Methods}).
+The nullclines define a single stable steady state $(p^*, r^*)$ \tr{(Fig.~\ref{fig:model_analysis}\textit{A} and \textit{Methods}).}
 At this steady state, the growth rate is constant and denoted by $\mu^*$.
@@
 The nullcline for $r$, and thus the location of the steady state with the associated growth rate, are given by $\alpha$.
-Fig.~\ref{fig:model_analysis}\textit{B} shows the dependency of the steady-state growth rate $\mu^*$ on the resource allocation parameter $\alpha$.
+\tr{Fig.~\ref{fig:model_analysis}\textit{B}} shows the dependency of the steady-state growth rate $\mu^*$ on the resource allocation parameter $\alpha$.
 As can be seen, $\mu^*$ is maximal for a specific, unique value of $\alpha$, which we denote $\alpha_{opt}^*$.
 That is, the model predicts that there is a single optimal way to divide the precursor flux over the synthesis of the gene expression machinery and the metabolic machinery.
-The same result, using a similar model, was obtained by Scott \textit{et al.}~\cite{scott_emergence_2014}.
-The self-replicator model is simple enough to derive an algebraic expression for computing $\alpha^*_{opt}$ and the corresponding maximal growth rate $\mu^*_{opt}$ (\textit{Methods} and \ref{S1_Text}), which will simplify analysis of the system in later sections.
+The same result, using a similar model, was obtained by Scott \etal~\cite{scott_emergence_2014}.
+The self-replicator model is simple enough to derive an algebraic expression for computing $\alpha^*_{opt}$ and the corresponding maximal growth rate $\mu^*_{opt}$ \tr{(\textit{Methods} and Text~\ref{S1_Text}),} which will simplify analysis of the system in later sections.
 
-\begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig2.eps}
-\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
-\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
+\begin{figure}[!ht]
+\includegraphics[scale=1]{./Fig/Fig2.eps}
+\caption[Analysis of self-replicator model of bacterial growth]{\textbf{\tr{Analysis of self-replicator model of bacterial growth.}}\\
+\tr{\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
 The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
-Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
+}
+\tr{Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.}
+\tr{
 \textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
-The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
+The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.}
 }
@@
 
-In order to validate the model, we verified that it can account for data on the macromolecular composition of \textit{E. coli} at steady state~\cite{scott_interdependence_2010}.
-When optimizing $\alpha$ for different values of $e_M$ (assuming cells attain maximal growth), the model predicts a relation between $\alpha^*_{opt}$ and $\mu^*_{opt}$ (colored dots and black dashed line in Fig.~\ref{fig:model_validation}\textit{A}) that is quasi-linear for high growth rates.
-We compared this prediction with the results of experiments where the relation between the growth rate and the mass ratio of total RNA and protein was determined in different growth media (Fig.~\ref{fig:model_validation}\textit{B}).
-In the framework of our model, different media correspond to different values of $e_M$, and different total RNA/protein mass ratios to different values of $\alpha$ (up to a conversion factor), allowing a direct comparison of the model predictions in Fig.~\ref{fig:model_validation}\textit{A} with the data in Fig.~\ref{fig:model_validation}\textit{B} (see \textit{Methods}).
-As can be seen, the model is able to account for the observed quasi-linear relation between the growth rate and the total mass ratio of RNA and protein. 
-Moreover, for realistic values of $k_R$ and $e_M$, a good quantitative fit is obtained (\textit{Methods} and \ref{S1_Table}).
-
-\begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig3.eps}
+\begin{figure}[!ht]
+\includegraphics[scale=1]{./Fig/Fig3.eps}
 \caption[Self-replicator model accounts for bacterial growth laws]{
-\textbf{Self-replicator model accounts for bacterial growth laws.}\\
-\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
+\textbf{\tr{Self-replicator model accounts for bacterial growth laws.}}\\
+\tr{\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
 The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
@@
 }
+}
 \label{fig:model_validation}
@@
 
-The data from Scott \textit{et al.} also reveal a second apparently linear relation between the growth rate and the total RNA/protein mass ratio.
+In order to validate the model, we verified that it can account for data on the macromolecular composition of \textit{E. coli} at steady state~\cite{scott_interdependence_2010}.
+When optimizing $\alpha$ for different values of $e_M$ (assuming cells attain maximal growth), the model predicts a relation between $\alpha^*_{opt}$ and $\mu^*_{opt}$ (colored dots and black dashed line in \tr{Fig.~\ref{fig:model_validation}\textit{A}}) that is quasi-linear for high growth rates.
+We compared this prediction with the results of experiments where the relation between the growth rate and the mass ratio of total RNA and protein was determined in different growth media (\tr{Fig.~\ref{fig:model_validation}\textit{B}}).
+In the framework of our model, different media correspond to different values of $e_M$, and different total RNA/protein mass ratios to different values of $\alpha$ (up to a conversion factor)
+\footnote{The RNA/protein mass ratio is often used as a proxy for the fraction of total protein mass that is ribosomal protein.
+At steady state, it correlates with the fraction of the total protein synthesis flux devoted to the production of ribosomal proteins~\cite{scott_interdependence_2010}, and therefore with $\alpha$.}
+, allowing a direct comparison of \tr{the model predictions in Fig.~\ref{fig:model_validation}\textit{A} with the data in Fig.~\ref{fig:model_validation}\textit{B} (see \textit{Methods}).}
+As can be seen, the model is able to account for the observed quasi-linear relation between the growth rate and the total mass ratio of RNA and protein. 
+\tr{Moreover, for realistic values of $k_R$ and $e_M$, a good quantitative fit is obtained (\textit{Methods} and Table~S1).}
+
+The data from Scott \etal also reveal a second apparently linear relation between the growth rate and the total RNA/protein mass ratio.
 This relation is obtained when varying, in the same growth medium, the efficiency of protein synthesis by adding different doses of an inhibitor of translation (chloramphenicol)~\cite{scott_interdependence_2010}.
@@
 We conclude that the self-replicator model is able to reproduce known observations of resource allocation in bacteria, so-called growth laws~\cite{scott_interdependence_2010}.
-The model is similar to a model recently proposed by Scott \textit{et al.}~\cite{scott_emergence_2014}.
+The model is similar to a model recently proposed by Scott \etal~\cite{scott_emergence_2014}.
 Contrary to the latter model, the translation rate is not assumed to be constant in the self-replicator model, but rather depends on precursor abundance, as proposed by the same authors in~\cite{klumpp_molecular_2013}.
@@
 Optimal dynamical control problems for two-dimensional nonlinear dynamical systems, like the problem of Eq.~\ref{eq:J}, are generally difficult to solve.
-However, we will show that the class of functions to which $\alpha_{opt}$ belongs can be identified, and we will use numerical optimization to identify a particular $\alpha_{opt}$ maximizing $J$.
+However, we will show that the class of functions to which $\alpha_{opt}$ belongs can be identified, and we will use numerical optimization to identify a particular \tr{$\alpha_{opt}$} maximizing $J$.
 
@@
 where $K= \beta\, K_R$ and $E_M= e_M / k_R$.
-The nondimensional growth rate is given by:
+\tr{The nondimensional growth rate is given by:
 \begin{equation}
@@
 \end{equation}
-Notice that the nondimensionalized system depends on a single parameter $K$, in addition to the constant environment $E_M$, which functions as an input to the system.
+Notice that the nondimensionalized system depends on a single parameter $K$, \tr{in addition to the constant environment $E_M$, which functions as an input to the system.}
 
-Analysis of the nondimensionalized system allows a number of properties of the solution of the optimal control problem of Eq.~\ref{eq:J} to be derived (\textit{Methods} and \ref{S3_Text}).
+Analysis of the nondimensionalized system allows a number of properties of the solution of the optimal control problem of Eq.~\ref{eq:J} to be derived (\textit{Methods} and Text~\ref{S3_Text}).
 First, by applying a version of the well-known Pontryagin Maximum Principle \cite{carlson_infinite_1991}, we can prove that the optimal solution is obtained for an alternating sequence of $\alpha(\cdot)=0$ and $\alpha(\cdot)=1$, possibly ending with an intermediate value of $\alpha(\cdot)$, corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the post-upshift environment $E_M$.
@@
 Third, this switching behavior is characterized by a so-called switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
-The switching curve divides the phase plane into two regions, such that $\alpha(\cdot)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (black dashed curve in Fig.~\ref{fig:optimalcontrol}\textit{A}).
+The switching curve divides the phase plane into two regions, such that $\alpha(\cdot)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (black dashed curve in Fig.~\ref{fig:optimalcontrol}\textit{A}).}
 
-\begin{figure}[!h]
-\centering
-%\includegraphics[scale=1]{./Fig/Fig4.eps}
-\caption[Optimal control of the self-replicator during a nutrient upshift]{
-{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
-\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines.
-The optimal trajectory is shown as a solid, red curve.
-The solid, black curve represents the $\hat{p}$-nullcline.
-The dashed, black curve is the switching curve $\varphi(\hat{p})$.
-The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
-\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
-}
-\label{fig:optimalcontrol}
-\end{figure}
-
-In line with these results, we conjecture that the optimal solution consists in a switching transient towards the optimal steady state for the new environment, and remains at this steady state until the next environmental change.
+\tr{In line with these results, we conjecture that the optimal solution consists in a switching transient towards the optimal steady state for the new environment, and remains at this steady state until the next environmental change.
 Such a solution is known as a bang-bang-singular solution in the control theory literature~\cite{stengel_optimal_1994}.
-Formally, the solution of Eq.~\ref{eq:J} can be described as
+Formally, the solution of Eq.~\ref{eq:J} can be described as}
 \begin{equation}
@@
 
-The optimal control problem of Eq.~\ref{eq:J} was also solved numerically by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012} (see \textit{Methods} for details).
+The optimal control problem of Eq.~\ref{eq:J} was also solved numerically by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012} (see \tr{\textit{Methods}} for details).
 A time discretization allows the problem to be transformed into a nonlinear optimization problem solved here by interior point techniques.
-The optimal trajectories obtained numerically confirm our conjecture that the optimal control is bang-bang-singular.
+\tr{The optimal trajectories obtained numerically confirm our conjecture that the optimal control is bang-bang-singular.}
 An example solution, obtained by numerical optimization is shown in Fig.~\ref{fig:optimalcontrol}.
@@
 The optimal solution $\alpha_{opt}$ consists of a sequence of switches between $\alpha=1$, corresponding to maximal synthesis of the gene expression machinery, and $\alpha=0$, corresponding to maximal synthesis of the metabolic machinery, until $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ is reached.
-$\alpha$ is then set to $\alpha^*_{opt}$, the value leading to the maximum growth rate in the new medium (here 0.5, for $E_M=1$).
+\tr{$\alpha$ is then set to $\alpha^*_{opt}$, the value leading to the maximum growth rate in the new medium (here 0.5, for $E_M=1$).
 The sequence of switches of $\alpha$ in Fig.~\ref{fig:optimalcontrol}\textit{B} corresponds to successive crossings of the switching curve in Fig.~\ref{fig:optimalcontrol}\textit{A}.
-In particular, the switch just after $\hat{t}=2$ corresponds to the first crossing of the switching curve; the subsequent switches accumulate around the steady state and are therefore difficult to identify in the plot.
+In particular, the switch just after $\hat{t}=2$ corresponds to the first crossing of the switching curve; the subsequent switches accumulate around the steady state and are therefore difficult to identify in the plot.}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=1]{./Fig/Fig4.eps}
+\caption[Optimal control of the self-replicator during a nutrient upshift]{
+{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
+\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines.
+The optimal trajectory is shown as a solid, red curve.
+The solid, black curve represents the $\hat{p}$-nullcline.
+The dashed, black curve is the switching curve $\varphi(\hat{p})$.
+The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \tr{\textit{Methods}} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
+\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
+}
+\label{fig:optimalcontrol}
+\end{figure}
 
@@
 In the framework of the self-replicator model of bacterial growth, control strategies take the form of feedback control laws mapping the value of system variables to a value of the control variable $\alpha(\cdot)$.
-In this section, we explore two such strategies, the first exploiting information on the quality and quantity of substrate present in the environment, as reflected in the value of $E_M$, and the second using information on the precursor concentration $\hat{p}$.
-The feedback control strategies are graphically displayed in Fig.~\ref{fig:strategies_overview}, as an extension of the self-replicator of Fig.~\ref{fig:self_replicator}.
-We pose a number of mathematical constraints on the feedback control strategies considered below.
+In this section, we explore two such strategies, the first exploiting information \tr{on the quality and quantity of substrate present} in the environment, as reflected in the value of $E_M$, and the second using information on the precursor concentration $\hat{p}$.
+\tr{The feedback control strategies are graphically displayed in Fig.~\ref{fig:strategies_overview}, as an extension of the self-replicator of Fig.~\ref{fig:self_replicator}.}
+\tr{We pose a number of mathematical constraints on the feedback control strategies considered below.
 First, we require the control laws to be functions of the variables of the self-replicator but not involve derivatives or integrals of these variables.
@@
 Third, this steady state must equal the optimal steady state for that environment, given by $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
+}
 
-\begin{figure}[!h]
+\begin{figure}[ht]
 \centering
-%\includegraphics[scale=1]{./Fig/Fig5.eps}
+\includegraphics[scale=1]{./Fig/Fig5.eps}
 \caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
-{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
-The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
+{\bf \tr{Alternative strategies for controlling the self-replicator of bacterial growth.}}\\
+\tr{The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
+}
 }
@@
 \end{equation}
-Notice that $\alpha$ is constant because $E_M$ is fixed to the value defining the new environment after the upshift.
+\tr{Notice that $\alpha$ is constant because $E_M$ is fixed to the value defining the new environment after the upshift.
 What would be an appropriate choice for $f$?
-An advantage of the self-replicator model is that the optimal allocation at steady state can be explicitly formulated as a function of $E_M$ (Eq.~\ref{eq:alpha_mu_optimal} in \textit{Methods}, with derivation in \ref{S1_Text}).
-This function is the unique function satisfying all of the above criteria (\ref{S1_Text}).
-\ref{S1_Fig} plots $f$ and shows that it is conveniently approximated by a Michaelis-Menten function, \textit{i.e.},
+An advantage of the self-replicator model is that the optimal allocation at steady state can be explicitly formulated as a function of $E_M$ (Eq.~\ref{eq:alpha_mu_optimal} in \textit{Methods}, with derivation in Text~\ref{S1_Text}).
+This function is the unique function satisfying all of the above criteria (Text~\ref{S1_Text}).
+Fig.~\ref{S1_Fig}\textit{A} plots $f$ and shows that it is conveniently approximated by a Michaelis-Menten function, \textit{i.e.},}
 \begin{equation}\label{eq:MMApprox}
@@
 
-As an example of a regulatory system resembling the above control strategy consider the phosphotransferase system responsible for the uptake of glucose, the preferred substrate of \textit{E.
+\tr{As an example of a regulatory system resembling the above control strategy consider the phosphotransferase system responsible for the uptake of glucose, the preferred substrate of \textit{E.
 coli}~\cite{deutscher_how_2006}.
@@
 The phosphorylation state of EIIA\textsuperscript{\textrm{Glc}} thus provides an indirect read-out of glucose availability.
-In response to this signal, a variety of metabolic processes are upregulated or downregulated, notably involving the signalling molecule cAMP which activates the pleiotropic transcription factor Crp~\cite{deutscher_how_2006,gorke_carbon_2008}.
+In response to this signal, a variety of metabolic processes are upregulated or downregulated, notably involving the signalling molecule cAMP which activates the pleiotropic transcription factor Crp~\cite{deutscher_how_2006,gorke_carbon_2008}.}
 
-How does the control strategy of Eq.~\ref{eq:f}, which we call a nutrient-only strategy, perform in comparison with the optimal solution derived in the previous section?
+\tr{How does the control strategy of Eq.~\ref{eq:f},} which we call a nutrient-only strategy, perform in comparison with the optimal solution derived in the previous section?
 That is, how much biomass does this strategy produce compared with the maximal amount of biomass that can theoretically be obtained after a nutrient upshift?
-In order to answer these questions, we simulated the response to a sudden upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} controlled by the nutrient-only strategy of Eq.~\ref{eq:f}.
-The results are shown in Fig.~\ref{fig:comparison_nutrients_precursor}.
+\tr{In order to answer these questions, we simulated the response to a sudden upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} controlled by the nutrient-only strategy of Eq.~\ref{eq:f}.
+The results are shown} in Fig.~\ref{fig:comparison_nutrients_precursor}.
 Panel~\textit{A} shows the trajectory of the controlled self-replicator system and panel~\textit{D} plots the evolution of the amount of biomass as a fraction of the amount of biomass produced by the optimal strategy.
 While the system does reach the steady state that is optimal for $E_M$, the nutrient-only strategy has poor performance in the transient phase immediately following the nutrient upshift.
-As can be seen from the solution trajectory in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{A}, fixing $\alpha$ to the value that enables optimal growth at steady state leads to a huge transient overshoot of the precursor concentration.
+As can be seen \tr{from the solution trajectory} in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{A}, fixing $\alpha$ to the value that enables optimal growth at steady state leads to a huge transient overshoot of the precursor concentration.
 The overshoot reveals that resource allocation is initially suboptimal, with too many resources invested in the metabolic machinery at the expense of the gene expression machinery.
-This causes a transiently suboptimal growth rate, leading to lower biomass accumulation (Eq.~\ref{eq:biomassdyn}).
-
-\begin{figure}[!h]
-\centering
-%\includegraphics[scale=1]{./Fig/Fig6.eps}
-\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
-{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
-\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
-The solid, black curve represents the $\hat{p}$-nullcline.
-The dashed, black curve is the $\hat{r}$-nullcline.
-The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
-The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
-While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
-\textit{B:}~As above, but for the precursor-only strategy.
-The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
-The solution trajectory (blue curve) exhibits a lower overshoot.
-\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
-Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
-\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
-In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
-}
-\label{fig:comparison_nutrients_precursor}
-\end{figure}
+\tr{This causes a transiently suboptimal growth rate, leading to lower biomass accumulation (Eq.~\ref{eq:biomassdyn}).}
 
@@
 \end{equation}
-Since $\hat{p}$ will vary during the upshift experiment, $\alpha$ is not constant, contrary to the nutrient-only strategy above.
+Since $\hat{p}$ will vary during the upshift experiment, $\alpha$ is not constant, \tr{contrary to the nutrient-only strategy above.
 In the \textit{Methods} section, we present a function $g$ satisfying the requirements listed in the beginning of this section, in particular that the system converge to a stable steady state ensuring maximal growth in the new environment.
-Moreover, we show that any other choice for $g$ leads to lower biomass production.
-The function is plotted in \ref{S1_Fig}, and as shown in the same panel, is conveniently approximated by a Hill function with cooperativity coefficient 2:
+Moreover, we show that any other choice for $g$ leads to lower biomass production.}
+The function is plotted in Fig.~\ref{S1_Fig}\textit{B}, and as shown in the same panel, is conveniently approximated by a Hill function with cooperativity coefficient 2:
 \begin{equation}\label{eq:HillApprox}
@@
 While converging to the same steady state, this second strategy, which we will refer to as the precursor-only strategy, performs much better than the nutrient-only strategy after an upshift, as shown in Fig.~\ref{fig:comparison_nutrients_precursor}.
-We simulated the response to a nutrient upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} with the precursor-only strategy of Eq.~\ref{eq:precursorstrategy}.
-The relative biomass increases by 51\% and reaches 94\% of the biomass produced by the optimal control strategy (the theoretical maximum).
+\tr{We simulated the response to a nutrient upshift of the self-replicator of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} with the precursor-only strategy of Eq.~\ref{eq:precursorstrategy}.
+The relative biomass increases by} 51\% and reaches 94\% of the biomass produced by the optimal control strategy (the theoretical maximum).
 The precursor-only strategy notably avoids the inefficient transient accumulation of precursors directly after the nutrient upshift, by alternatingly investing more resources in gene expression (consumption of precursors) and metabolism (production of precursors).
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=1]{./Fig/Fig6.eps}
+\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
+{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
+\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy \tr{(green curve).}
+The solid, black curve represents the $\hat{p}$-nullcline.
+The dashed, black curve is the $\hat{r}$-nullcline.
+The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by \tr{Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in Fig.~\ref{S1_Fig}\textit{A}.}
+The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
+\tr{While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.}
+\textit{B:}~As above, but for the precursor-only strategy.
+The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by \tr{Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in Fig.~\ref{S1_Fig}\textit{B}.
+The solution trajectory (blue curve) exhibits a lower overshoot.}
+\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
+\tr{Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).}
+\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
+In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
+}
+\label{fig:comparison_nutrients_precursor}
+\end{figure}
+
 Both strategies, nutrient-only and precursor-only, drive the self-replicator towards the same steady state.
-Whereas the two strategies are thus indistinguishable when the analysis is restricted to steady state, the precursor-only strategy is shown to perform much better in a dynamical upshift scenario, in the sense that the biomass produced is much closer to that produced by the optimal strategy.
+Whereas the two strategies are thus indistinguishable when the analysis is restricted to steady state, the precursor-only strategy is shown to perform much better in a dynamical upshift scenario, \tr{in the sense that the biomass produced is much closer to that produced by the optimal strategy}.
 Several authors have concluded that control strategies based on precursor sensing are key for maintaining optimal growth at steady state.
@@
 
-In the quest for further improvements, a natural starting-point would be to consider the curve defining the optimal steady states $(\hat{p}^*_{opt},\hat{r}^*_{opt})$  for different environments $E_M$.
-This curve is defined by a function mapping $\hat{p}^*$ to $\hat{r}^*$, which is actually the same as the function $g$ introduced in the precursor-only strategy (\textit{Methods} and \ref{S1_Fig}), given that at steady state $\hat{r}=\alpha$ (Eq.~\ref{eq:rdefnondim}).
+\tr{In the quest for further improvements, a} natural starting-point would be to consider the curve defining the optimal steady states $(\hat{p}^*_{opt},\hat{r}^*_{opt})$  for \tr{different environments $E_M$.
+This curve is defined by a function mapping $\hat{p}^*$ to $\hat{r}^*$, which is actually the same as the function $g$ introduced in the precursor-only strategy (\textit{Methods} and Figure~\ref{S1_Fig}), given that at steady state $\hat{r}=\alpha$ (Eq.~\ref{eq:rdefnondim}).
 The curve can be seen as representing an optimal balance between precursors and the gene expression machinery, in the sense that the maximal growth rate attainable for a given precursor concentration $\hat{p}$ requires a concentration $\hat{r}$ of ribosomes and other components of the gene expression machinery equal to $g(\hat{p})$.
-If either $\hat{r}>g(\hat{p})$ or $\hat{r}<g(\hat{p})$, the growth rate is suboptimal.
+If either $\hat{r}>g(\hat{p})$ or $\hat{r}<g(\hat{p})$, the growth rate is suboptimal.}
 
-These considerations suggest an intuitive control strategy, namely to avoid an imbalance between $\hat{p}$ and $\hat{r}$ at all times, and remain as close as possible to the curve defined by $g$.
-In particular, when the gene expression machinery is more abundant than what is optimal given the available precursors ($\hat{r} > g(\hat{p})$), its synthesis is switched off ($\alpha = 0$).
+\tr{These considerations suggest an intuitive control strategy, namely to avoid an imbalance between $\hat{p}$ and $\hat{r}$ at all times, and remain as close as possible to the curve defined by $g$.
+In particular, when} the gene expression machinery is more abundant than what is optimal given the available precursors ($\hat{r} > g(\hat{p})$), its synthesis is switched off ($\alpha = 0$).
 Conversely, when $\hat{r} < g(\hat{p})$, synthesis of the gene expression machinery is switched on.
-This strategy thus tries to restore "as quickly as possible" the optimal balance between precursors $\hat{p}$ and the gene expression machinery $\hat{r}$, giving rise to a so-called on-off control strategy:
+\tr{This strategy thus tries to restore} "as quickly as possible" the optimal balance between precursors $\hat{p}$ and the gene expression machinery $\hat{r}$, giving rise to a so-called on-off control strategy:
 \begin{equation}
@@
 
-As shown in the \textit{Methods} section, the on-off strategy drives the system to a stable steady state ensuring growth at the maximal rate.
-Notice that, contrary to the strategies discussed in the previous section, the value of $\alpha$ selected by the on-off strategy depends on both $\hat{p}$ and $\hat{r}$ (Fig.~\ref{fig:strategies_overview}).
-It thus uses more information on the state of the system than the nutrient-only and precursor-only strategies.
+\tr{As shown in the \textit{Methods} section, the on-off strategy drives the system to a stable steady state ensuring growth at the maximal rate.}
+Notice that, contrary to the strategies discussed in the previous section, the value of $\alpha$ selected by the on-off strategy depends on both $\hat{p}$ and $\hat{r}$ \tr{(Fig.~\ref{fig:strategies_overview}).
+It thus uses more information on the state of the system than the nutrient-only and precursor-only strategies.}
 
@@
 
-\begin{figure}[!h]
+\begin{figure}
 \centering
-%\includegraphics[scale=1]{./Fig/Fig7.eps}
+\includegraphics[scale=1]{./Fig/Fig7.eps}
 \caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
@@
 
-How does this mechanistic model of ppGpp regulation relate to the on-off strategy presented above?
+\tr{How does this mechanistic model of ppGpp regulation relate to the on-off strategy presented above?
 In order to answer this question, we first need to find a correspondence between the variables $p$ and $r$ of our coarse-grained model and the concentrations of molecular species in the kinetic model of Bosdriesz \textit{et al.}
 This is rather straightforward to achieve, by equating $p$ to the total amino acid concentration and $r$ to the ribosome concentration.
-Second, \ref{S4_Text} shows that by making two simplifying assumptions, ppGpp can be expressed as a function of the total amino acid concentration and the ribosome concentration.
+Second, Text~\ref{S4_Text} shows that by making two simplifying assumptions, ppGpp can be expressed as a function of the total amino acid concentration and the ribosome concentration.
 In particular, we assume that concentrations of all individual amino acids are equal, and that the concentrations of charged tRNAs and ppGpp evolve fast relative to the dynamics of the amino acid and ribosome concentrations.
@@
 \end{equation}
-with $K_I$ a Michaelis-Menten inhibition constant [$\mu$mol~L\textsuperscript{-1}] and $\texttt{ppGpp}$ the (time-varying) intracellular concentration of ppGpp [$\mu$mol~L\textsuperscript{-1}].
+with $K_I$ a Michaelis-Menten inhibition constant [$\mu$mol~L\textsuperscript{-1}] and $\texttt{ppGpp}$ the (time-varying) intracellular concentration of ppGpp [$\mu$mol~L\textsuperscript{-1}].}
 
-The response function for ppGpp thus obtained and evaluated for a range of amino acid and ribosome concentrations is represented in Fig.~\ref{fig:ppGppsurface}, and visually compared with the on-off strategy.
+The response function for ppGpp \tr{thus obtained and evaluated for a range of amino acid and ribosome concentrations} is represented in Fig.~\ref{fig:ppGppsurface}, and visually compared with the on-off strategy.
 As can be seen, the two response surfaces are very similar.
 In other words, the ultrasensitive response of the synthesis rate of ribosomal proteins to the suboptimal allocation of cellular resources, derived from a model of the molecular mechanisms involved in the synthesis, degradation, and regulatory action of ppGpp~\cite{bosdriesz_how_2015}, implements a control strategy that is close to the optimal predicted by a control-theoretical analysis of the self-replicator.
-While the role of ppGpp in maintaining optimal resource allocation was already pointed out by Scott \textit{et al.} and Bosdriesz \textit{et al.}, the latter studies were restricted to optimizing steady-state growth.
+While the role of ppGpp in maintaining optimal resource allocation was already pointed out by Scott \etal and Bosdriesz \etal, the latter studies were restricted to optimizing steady-state growth.
 A major insight from the analysis in this section is that this conclusion seems to carry over to dynamical scenarios as well.
@@
 
-\begin{figure}[!h]
+\begin{figure}
 \centering
-%\includegraphics[scale=1]{./Fig/Fig8.eps}
+\includegraphics[scale=1]{./Fig/Fig8.eps}
 \caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
@@
 {\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
-\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
-\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
+\textit{A:}~Response surface of the on-off control strategy, \tr{defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.}
+\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and \tr{the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (Text~\ref{S4_Text}).}
 The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
@@
 A first major contribution of our work is to show that, in the case of a dynamical upshift scenario, optimal production of biomass requires a bang-bang-singular resource allocation scheme (Fig.~\ref{fig:optimalcontrol}).
-That is, the optimal self-replicator should iteratively allocate all of its resources to the gene expression machinery (bang control input) and the metabolic machinery (another bang control input), until the steady state enabling maximal growth in the post-upshift environment is reached, corresponding to a trade-off in the allocation of resources to the two processes (singular control input).
+\tr{That is, the optimal self-replicator should iteratively allocate all of its resources to the gene expression machinery (bang control input) and the metabolic machinery (another bang control input), until the steady state enabling maximal growth in the post-upshift environment is reached, corresponding to a trade-off in the allocation of resources to the two processes (singular control input).}
 

### 2016-06-01 Chapters/introduction.tex
First meeting with H&H

\chapter{Introduction}

\textit{There's an infinity of things that have never been done before, and
most of those things are not worth doing.} -- Jeremy Fox

\section*{Summary}

In this section...

\selectlanguage{french}
\section*{Résumé}

Dans cette section...
\selectlanguage{english}

\section{Context and motivations}

The theory of evolution has deeply changed biology.\\
\textit{Very broad introduction on how we use evolution to actually make sense of biology}

We are still trying to understand the fundamental forces that shape life in all its diversity.\\
\textit{Evolution through natural selection is a verbal theory, and people have been trying to formalize it for 200 years.
It implies working and looking for "the hand of evolution" at several scales, from population to molecular biology.
It is challenging to translate the mechanisms of natural selection (fitness) at every scale, hence the problems to uncover fundamental properties of life.}
%Since its invention, huge efforts have been made to put Darwin's theory into equation, in order to clarify our understanding, to see how it apply at different scales (species, populations, cells, molecules...).

Microorganisms exhibits the most fundamental features of life. 
\textit{They are open systems that take energy from the environment around them, and use it to maintain and reproduce their structures.
They live literally everywhere. They represent a perfect model organism for fundamental questions.}

They also managed to become the thoughest of all the living systems.
\textit{Their diversity allow them to colonize every biotope, but remarkably, a single species can live in broad conditions, growing on different carbon sources or resist to really harmful environments.}

Their capacity to survive and exploit strong environmental fluctuations lay on another ubiquitous process of life as we know it: gene expression.
\textit{Short explanation on DNA, RNA, Protein and the regulations behind, that can be seen as loading "programs" depending on the environment the organism as to cope with.}

Such a mechanism can be seen as a resource allocation problem.
\textit{Limited protein production capacity, so choice have to be made for what need to be produced, natural selection selecting for the best strategies.}

General growth laws have been identified.
\textit{The study of resource allocation led to fundamental laws like ribosome allocation or overflow metabolism, processes that can only be understood through growth rate maximization.}

Biomass production is an important fitness factor of microorganisms.
\textit{Broadly used in a variety of study}

One of the key feature that made these works possible was to study the living system in a well-defined and reproducible state: balanced-growth.
\textit{We describe balanced-growth, what it means for the dynamical system, and how you can reproduce it in a lab.
We also explain why it is so convenient.}

The mechanisms identified exist to cope with environmental fluctuations in the natural habitat of microorganisms, but are being studied in constant environments.
\textit{Now we start to see the problem: there's a caveat in studying a system that is designed to respond to changes, if we focus on unchanging conditions.
And we are currently unable to create such beautiful laws when taking into account dynamical conditions.}

While adopting a dynamical perspective on resource allocation is challenging, it could be fundamental to the life-treat-history of microorganisms.
\textit{It is also a problem if the system we are studying poorly encountered this state during its evolution.
In other words, "maybe bacteria are confused during steady-state".
Hence, there is really a strong need for a mathematical framework to study resource allocation in dynamical conditions, and to know what actually happens during these growth transitions.}

\section{Problem statement}

How do microorganisms adjust their resource allocation during environmental transitions ?
\textit{As we discussed, I tried to be as specific as possible here.}

\section{Related work}

(the work on dynamic, Van der Berg, Alon, Levy, etc... the work on steady-state has normally been described previously. Or, I don't know, maybe I can shortly describe it in the context and extensively describe it here.)

\section{Approach}

We first tackle the problem of creating a mathematical framework for studying environmental transitions.
\textit{Building upon the work cited above, we create a simple self-replicator model that allocate resources between two sections.}

The goal is to identify what would be an ideal transition.
\textit{We all have an intuitive idea of what should be a good transition: it should be fast, robust, etc... But it is unclear how this relates to the fitness factor that shaped the regulatory mechanisms of microorganisms.}

We describe dynamical optimilaty by using optimal control theory.
\textit{Brieve explanation about optimal control theory and how it is used in economic or engineering to optimize or understant dynamical processes.}

Then, after we know what to look for, we try to actually measure how resource allocation occurs in a model organism: \textit{Escherichia coli}.
\textit{We justify the use of coli and introduce the gene expression fluorescent reporter.
We made a strain with fluorescent ribosomes, and studied our experimental framework in different transitions, at the population and the single cell level.
We justify the use of both by the fact that population is easy for "screening", and single cell allow for a better resolution and take care of the "desynchronization" biais that could occur.}

\section{Organisation of the dissertation}

(announcing the plan of the manuscript, which actually follow exactly the approach part, without the lenghty justifications)

### 2016-06-01 Chapters/main_plos.tex
First meeting with H&H

\vspace*{0.35in}
\input{titlepage}

% Please keep the abstract below 300 words
\input{abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\input{author_summary}

\linenumbers

%%%%%% CORE OF THE ARTICLE %%%%%
\input{core}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Methods}

\subsection*{Steady-state analysis of model}
\label{sec:methods_model_analysis}

The nondimensional version of the model, given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, was used for a steady-state analysis of the self-replicator.
Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} were derived from the original model of Eqs~\ref{eq:pdef}-\ref{eq:rdef} by means of the following rescalings:
\begin{equation*}
\hat{p}  = \beta \, p,\;\;\;
\hat{r}  = \beta \, r,\;\;\;
\hat{t}  = k_R \, t,\;\;\;
E_M = e_M/k_R,\;\;\;
K = \beta K_R.
\end{equation*}

As shown in \ref{S1_Text}, for a constant environment $E_M$ and constant resource allocation $\alpha$, the system has two steady states: a trivial unstable steady state $(\hat{p}^*, \hat{r}^*) = (0,1)$, allowing no growth in the absence of precursors, and a steady state with a positive growth rate given by
\begin{equation}
\label{eq:meth_steadystate}
(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
\end{equation}
The two eigenvalues of the Jacobian matrix evaluated at $(\hat{p}^*, \hat{r}^*)$ are negative (\ref{S1_Text}), so that this steady state is stable.

The growth rate at steady state, as a function of $\hat{p}^*$ and $\hat{r}^*$, is given by Eq.~\ref{eq:munondim}, which we repeat here for clarity:
\begin{equation*}
\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
\end{equation*}
Evaluating $d\hat{p}/dt=0$ at $(\hat{p}^*, \hat{r}^*)$ allows $\hat{r}^*$, and therefore $\hat{\mu}^*$, to be written as a function of $\hat{p}^*$ (\ref{S1_Text}).
Accordingly, we can compute $\partial\hat{\mu}^*/\partial\hat{p}^*$ and, when setting this partial derivative to 0, determine the maximum growth rate at steady state $\mu_{opt}^*$ and the optimal resource allocation $\alpha_{opt}^*$ bringing about this maximal growth rate.
As shown in \ref{S1_Text}, $\mu_{opt}^*$ and $\alpha_{opt}^*$ can be written as explicit functions of either the environment $E_M$:
\begin{equation}
\label{eq:alpha_mu_optimal}
\alpha_{opt}^* = \frac{E_M + \sqrt{K\, E_M}}{E_M + 2\sqrt{K\, E_M} + 1}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1},
\end{equation}
or the precursor abundance $\hat{p}_{opt}^*$:
\begin{equation}
\label{eq:alpha_mu_optimal_p}
\alpha_{opt}^* = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
\end{equation}
The above equations were used for the derivation of the control strategies (see below).

\subsection*{Model fitting}
\label{sec:methods_model_fitting}

As can be seen by comparing Figs~\ref{fig:model_validation}\textit{A} and~\ref{fig:model_validation}\textit{B}, growth-rate maximization in the self-replicator model leads to a good qualitative correspondence with the growth laws.
In order to determine if a good quantitative fit of the model with the data from Scott~\textit{et al.} \cite{scott_interdependence_2010} can be obtained, for reasonable parameter values, we estimated $e_M$ and $k_R$ in Eqs~\ref{eq:pdef}-\ref{eq:rdef} from the measured RNA/protein mass ratios.
At steady state, the RNA/protein mass ratio can be interpreted as proportional to $\hat{r}^*$ (and thus $\alpha_{opt}^*$), with an unknown (dimensionless) proportionality constant $\gamma$ (see~\cite{scott_interdependence_2010} for details on the use of the RNA/protein mass ratio as a proxy for the ribosomal protein mass fraction):
\begin{equation}
\hat{r}^* = \alpha_{opt}^* = \gamma \, \frac{\text{RNA mass}}{\text{protein mass}}.
\label{eq:gamma}
\end{equation}
Reformulating Eq.~\ref{eq:alpha_mu_optimal} in terms of the original parameters $e_M$ and $k_R$, which have physical dimensions facilitating the biological interpretation of their values, we obtain a straighforward relation between $e_M$, $k_R$, $K$, $\alpha_{opt}^*$ and $\mu_{opt}^*$:
\begin{equation}
\label{eq:alpha_mu_optimal_dim}
\alpha_{opt}^* = \frac{e_M + \sqrt{K\, e_M\, k_R}}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\mu^*_{opt} = \frac{e_M \, k_R}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}.
\end{equation}

Eqs~\ref{eq:gamma}-\ref{eq:alpha_mu_optimal_dim} were used to estimate values of $k_R$ and $\gamma$, as well as $e_M$ for each of the six growth conditions, from the measurements of the growth rate and the RNA/protein mass ratio. 
The value $K$ was not estimated from the experimental data, but set to a value inferred from the literature (\ref{S1_Text}).
The optimization process was carried out by means of the differential evolution algorithm of Storn and Price~\cite{storn_differential_1997}.
The results are shown in Fig.~\ref{fig:model_validation}\textit{B}, while the estimated parameter values are summarized in \ref{S1_Table}.
The parameter values are in very good agreement with order-of-magnitude values determined from the literature (\ref{S2_Text} and \ref{S1_Table}).

\subsection*{Solution of optimal control problem}

The optimal control problem of Eq.~\ref{eq:J} consists in identifying the function $\alpha_{opt}(t)$ that maximizes the integral of the growth rate $\hat{\mu}$ over an interval $[0, \tau]$.
In order to solve this problem, we first redefined it over an infinite horizon (\textit{i.e.}, $\tau \rightarrow \infty $) in order to avoid boundary effects occurring over finite time intervals, in particular the depletion of precursors just before reaching $\tau$.
With $\mathcal{U}=\{\alpha:\mathbb{R}^+ \rightarrow [0,1] \}$ the set of admissible controls, the full optimization problem for the nondimensionalized system is given by
\begin{equation}\label{eq:meth_prob}
\max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{\infty} \hat{r}(\hat{t}) \, \dfrac{\hat{p}(\hat{t})}{K+\hat{p}(\hat{t})}\, d\hat{t}.
\end{equation}
Since $J(\alpha)$ diverges, we actually consider overtaking optimality:
A solution is overtaking optimal if its performance index catches up with the performance index of any other solution (\cite{carlson_infinite_1991}, see \ref{S3_Text} for details).

Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle \cite{carlson_infinite_1991}, an extension of the well-known Pontryagin Maximum Principle.
Analysis of the Hamiltonian of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and the associated adjoint system shows that the optimal trajectory is a concatenation of bang arcs ($\alpha(\cdot)=0$ or $\alpha(\cdot)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the new environment after the upshift (\ref{S3_Text}).
Moreover, from the Kelley condition~\cite{borisov_fullers_2000}, we can show that if the optimal trajectory has a singular arc, then it must enter this singular arc \textit{via} a chattering arc, \textit{i.e.}, with an infinite number of switches of $\alpha (\cdot)$ between 0 and 1 (\ref{S3_Text}).
The chattering arc is characterized by a switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
The switching curve divides the phase plane into two regions, such that $\alpha(t)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (\ref{S3_Text} and Fig.~\ref{fig:optimalcontrol}).

The above results have led to the conjectured optimal solution of Eq.~\ref{eq:optcontrol}.
In parallel, we numerically solved the problem of Eq.~\ref{eq:meth_prob} by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012}.
A time discretization allows the optimal control problem to be transformed into a nonlinear optimization problem,
solved here by interior point techniques.
A discretization by a Lobatto IIIC formula (6th order) was used with 4000 time steps, and the relative tolerance for the NLP solver was set to $10^{-14}$.
The optimal trajectories thus obtained are composed of a chattering arc followed by a steady state corresponding to the singular arc (Fig.~\ref{fig:optimalcontrol}).
The switching curve $\varphi(\hat{p})$ was computed from numerical simulations with different initial conditions.

\subsection*{Specification and analysis of control strategies}

As described in the \textit{Results} section, we are interested in control strategies satisfying the following conditions:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}
It can be directly verified from the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies (Eqs~\ref{eq:f}, \ref{eq:precursorstrategy}, and \ref{eq:stratswitch}) that they are indeed static functions of the system variables (or the system input, in the case of the nutrient-only strategy). 
Here we show that the other two conditions are also satisfied for all three strategies.

Following Eq.~\ref{eq:f}, the nutrient-only strategy is defined by $\alpha = f(E_M)$, so that $\alpha$ is constant after the upshift.
As shown above and in \ref{S1_Text}, this means that the system controlled by the nutrient-only strategy has a single nontrivial stable steady state (Condition~C2). 
In addition, in this case the optimal steady state is attained for $\alpha^*_{opt}$ defined as in Eq.~\ref{eq:alpha_mu_optimal}, and the following function $f$ therefore guarantees Condition~C3:
\begin{equation}
\label{eq:meth_f}
f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K\, E_M} + 1}.
\end{equation}
In \ref{S1_Text}, it is shown that Eq.~\ref{eq:meth_f} is the only definition of $f$ satisfying all conditions.
\ref{S1_Fig} shows a plot of $f(E_M)$ together with a biologically plausible Michaelis-Menten approximation (Eq.~\ref{eq:MMApprox}).

The full specification of the precursor-only strategy demands an expression for the function $g$ in Eq.~\ref{eq:precursorstrategy}.
Recall that Eq.~\ref{eq:alpha_mu_optimal_p} defines $\alpha_{opt}^*$ in terms of the precursor concentration $\hat{p}_{opt}^*$, which leads us to propose the following function $g$:
\begin{equation}
\label{eq:meth_g}
g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
\end{equation}
As shown in \ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
This can be seen by noting that at steady state, $d\hat{r}/dt=0$ implies $\hat{r}^*=g(\hat{p}^*)$ (Eq.~\ref{eq:rdefnondim}).
In order for the self-replicator to attain a maximal growth rate at steady rate, Eq.~\ref{eq:alpha_mu_optimal_p} needs to be satified, which is the case for the above choice of the function $g$.
Like for $f$, Eq.~\ref{eq:meth_g} is the only choice for $g$ satisfying C1-C3.
\ref{S1_Fig} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).

The on-off control strategy is defined in Eq.~\ref{eq:stratswitch} and repeated below:
\begin{equation}
h(\hat{p}, \hat{r}) = 
\begin{cases}
0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}
This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in Fig.~\ref{fig:onoffresults}\textit{A}.
In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
Moreover, the steady state coincides with the optimal steady state $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ by construction, so that Condition~C3 is satisfied as well.
Fig~\ref{fig:ppGppsurface}\textit{A} shows a plot of $h(\hat{p}, \hat{r})$.

Note that since $h(\cdot)$ is discontinuous, numerical instabilities occur during simulations.
We therefore used the following continuous approximation of this function:
\begin{equation}
\frac{g(\hat{p})^{100}}{g(\hat{p})^{100} + \hat{r}^{100}}, \ \textrm{if} \ \hat{r} \neq g(\hat{p}).
\end{equation}
The approximation causes $\alpha$ to take intermediate values (instead of 0 or 1) just before reaching the optimal steady state in Fig~\ref{fig:onoffresults}\textit{C}.
For numerical simulations of the ODE system, we used the CVODE solver~\cite{cohen_cvode_1996} from SUNDIALS 2.6.2~\cite{hindmarsh_sundials_2005}.

\section*{Supporting Information}

%%%%%% FOR MANUAL REF%%%%%
\makeatletter
\newcommand{\manuallabel}[2]{\def\@currentlabel{#2}\label{#1}}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{S1 Text -- Model derivation and analysis}
\manuallabel{S1_Text}{S1~Text}

\subsection*{S2 Text -- Model parameters}
\manuallabel{S2_Text}{S2~Text}

\subsection*{S3 Text -- Solution of optimal control problem}
\manuallabel{S3_Text}{S3~Text}

\subsection*{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
\manuallabel{S4_Text}{S4~Text}

\subsection*{S1 Table -- Parameter values of self-replicator model}
\manuallabel{S1_Table}{S1~Table}

\subsection*{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth}
\manuallabel{S1_Fig}{S1~Figure}

\section*{Acknowledgments}
The authors thank Eugenio Cinquemani for discussions and comments on the manuscript.

\nolinenumbers

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
%

### 2016-06-01 Chapters/main_plos_label.tex
First meeting with H&H

--- 2016-06-01 Chapters/main_plos.tex
@@
+% Template for PLoS
+% Version 3.1 February 2015
+%
+% To compile to pdf, run:
+% latex plos.template
+% bibtex plos.template
+% latex plos.template
+% latex plos.template
+% dvipdf plos.template
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
+% the final text of your manuscript.
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig." instead of "Figure".
+% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - tabs/spacing/line breaks within cells to alter layout or alignment
+% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
+% - colors, shading, or graphic objects
+% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
+%
+% Please be sure to include all portions of an equation in the math environment.
+%
+% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% fixltx2e package for \textsubscript
+\usepackage{fixltx2e}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% rotating package for sideways tables
+\usepackage{rotating}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%% Personalized preambule %%%%
+% For cross-referencing
+\usepackage{xr}
+\externaldocument[S1.]{supp_S1}
+\externaldocument[S2.]{supp_S2}
+\externaldocument[S3.]{supp_S3}
+\externaldocument[S4.]{supp_S4}
+
+%%%% Language and fonts
+%%%% Math Stuff
+\usepackage{amsfonts}
+\newcommand{\argmax}{\operatornamewithlimits{argmax}}
+\newtheorem{theorem}{Theorem}
+%%%% Chemical Stuff
+\usepackage[version=3]{mhchem}
+%%%% OPTIONAL MACRO DEFINITIONS
+\usepackage{xspace}
+\newcommand*{\coli}{\textit{Escherichia coli}\xspace}
+\newcommand*{\etal}{\textit{et al.}\xspace}
+
+\usepackage{color} 
+\newcommand{\tr}[1]{{\color{red} #1}}
+\usepackage{soul}
+
+\begin{document}
 \vspace*{0.35in}
@@
 %%%%%% CORE OF THE ARTICLE %%%%%
-\input{core}
+\input{core_label}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\tr{
 \section*{Methods}
@@
 
-As shown in \ref{S1_Text}, for a constant environment $E_M$ and constant resource allocation $\alpha$, the system has two steady states: a trivial unstable steady state $(\hat{p}^*, \hat{r}^*) = (0,1)$, allowing no growth in the absence of precursors, and a steady state with a positive growth rate given by
+As shown in Text~\ref{S1_Text}, for a constant environment $E_M$ and constant resource allocation $\alpha$, the system has two steady states: a trivial unstable steady state $(\hat{p}^*, \hat{r}^*) = (0,1)$, allowing no growth in the absence of precursors, and a steady state with a positive growth rate given by
 \begin{equation}
@@
 \end{equation}
-The two eigenvalues of the Jacobian matrix evaluated at $(\hat{p}^*, \hat{r}^*)$ are negative (\ref{S1_Text}), so that this steady state is stable.
+The two eigenvalues of the Jacobian matrix evaluated at $(\hat{p}^*, \hat{r}^*)$ are negative (Text~\ref{S1_Text}), so that this steady state is stable.
 
@@
 \end{equation*}
-Evaluating $d\hat{p}/dt=0$ at $(\hat{p}^*, \hat{r}^*)$ allows $\hat{r}^*$, and therefore $\hat{\mu}^*$, to be written as a function of $\hat{p}^*$ (\ref{S1_Text}).
+Evaluating $d\hat{p}/dt=0$ at $(\hat{p}^*, \hat{r}^*)$ allows $\hat{r}^*$, and therefore $\hat{\mu}^*$, to be written as a function of $\hat{p}^*$ (Text~\ref{S1_Text}).
 Accordingly, we can compute $\partial\hat{\mu}^*/\partial\hat{p}^*$ and, when setting this partial derivative to 0, determine the maximum growth rate at steady state $\mu_{opt}^*$ and the optimal resource allocation $\alpha_{opt}^*$ bringing about this maximal growth rate.
-As shown in \ref{S1_Text}, $\mu_{opt}^*$ and $\alpha_{opt}^*$ can be written as explicit functions of either the environment $E_M$:
+As shown in Text~\ref{S1_Text}, $\mu_{opt}^*$ and $\alpha_{opt}^*$ can be written as explicit functions of either the environment $E_M$:
 \begin{equation}
@@
 In order to determine if a good quantitative fit of the model with the data from Scott~\textit{et al.} \cite{scott_interdependence_2010} can be obtained, for reasonable parameter values, we estimated $e_M$ and $k_R$ in Eqs~\ref{eq:pdef}-\ref{eq:rdef} from the measured RNA/protein mass ratios.
-At steady state, the RNA/protein mass ratio can be interpreted as proportional to $\hat{r}^*$ (and thus $\alpha_{opt}^*$), with an unknown (dimensionless) proportionality constant $\gamma$ (see~\cite{scott_interdependence_2010} for details on the use of the RNA/protein mass ratio as a proxy for the ribosomal protein mass fraction):
+At steady state, the RNA/protein mass ratio can be interpreted as proportional to $\hat{r}^*$ (and thus $\alpha_{opt}^*$), with an unknown (dimensionless) proportionality constant $\gamma$:
 \begin{equation}
@@
 Eqs~\ref{eq:gamma}-\ref{eq:alpha_mu_optimal_dim} were used to estimate values of $k_R$ and $\gamma$, as well as $e_M$ for each of the six growth conditions, from the measurements of the growth rate and the RNA/protein mass ratio.
-The value $K$ was not estimated from the experimental data, but set to a value inferred from the literature (\ref{S1_Text}).
+The value $K$ was not estimated from the experimental data, but set to a value inferred from the literature (Text~\ref{S1_Text}).
 The optimization process was carried out by means of the differential evolution algorithm of Storn and Price~\cite{storn_differential_1997}.
-The results are shown in Fig.~\ref{fig:model_validation}\textit{B}, while the estimated parameter values are summarized in \ref{S1_Table}.
-The parameter values are in very good agreement with order-of-magnitude values determined from the literature (\ref{S2_Text} and \ref{S1_Table}).
+The results are shown in Fig.~\ref{fig:model_validation}\textit{B}, while the estimated parameter values are summarized in Table~\ref{S1_Table}.
+The parameter values are in very good agreement with order-of-magnitude values determined from the literature (Text~\ref{S2_Text} and Table~\ref{S1_Table}).
 
@@
 Since $J(\alpha)$ diverges, we actually consider overtaking optimality:
-A solution is overtaking optimal if its performance index catches up with the performance index of any other solution (\cite{carlson_infinite_1991}, see \ref{S3_Text} for details).
+A solution is overtaking optimal if its performance index catches up with the performance index of any other solution (\cite{carlson_infinite_1991}, see Text~\ref{S3_Text} for details).
 
 Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle \cite{carlson_infinite_1991}, an extension of the well-known Pontryagin Maximum Principle.
-Analysis of the Hamiltonian of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and the associated adjoint system shows that the optimal trajectory is a concatenation of bang arcs ($\alpha(\cdot)=0$ or $\alpha(\cdot)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the new environment after the upshift (\ref{S3_Text}).
-Moreover, from the Kelley condition~\cite{borisov_fullers_2000}, we can show that if the optimal trajectory has a singular arc, then it must enter this singular arc \textit{via} a chattering arc, \textit{i.e.}, with an infinite number of switches of $\alpha (\cdot)$ between 0 and 1 (\ref{S3_Text}).
+Analysis of the Hamiltonian of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and the associated adjoint system shows that the optimal trajectory is a concatenation of bang arcs ($\alpha(\cdot)=0$ or $\alpha(\cdot)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the new environment after the upshift (Text~\ref{S3_Text}).
+Moreover, from the Kelley condition~\cite{borisov_fullers_2000}, we can show that if the optimal trajectory has a singular arc, then it must enter this singular arc \textit{via} a chattering arc, \textit{i.e.}, with an infinite number of switches of $\alpha (\cdot)$ between 0 and 1 (Text~\ref{S3_Text}).
 The chattering arc is characterized by a switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
-The switching curve divides the phase plane into two regions, such that $\alpha(t)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (\ref{S3_Text} and Fig.~\ref{fig:optimalcontrol}).
+The switching curve divides the phase plane into two regions, such that $\alpha(t)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (Text~\ref{S3_Text} and Fig.~\ref{fig:optimalcontrol}).
 
@@
 Following Eq.~\ref{eq:f}, the nutrient-only strategy is defined by $\alpha = f(E_M)$, so that $\alpha$ is constant after the upshift.
-As shown above and in \ref{S1_Text}, this means that the system controlled by the nutrient-only strategy has a single nontrivial stable steady state (Condition~C2). 
+As shown above and in Text~\ref{S1_Text}, this means that the system controlled by the nutrient-only strategy has a single nontrivial stable steady state (Condition~C2). 
 In addition, in this case the optimal steady state is attained for $\alpha^*_{opt}$ defined as in Eq.~\ref{eq:alpha_mu_optimal}, and the following function $f$ therefore guarantees Condition~C3:
@@
 \end{equation}
-In \ref{S1_Text}, it is shown that Eq.~\ref{eq:meth_f} is the only definition of $f$ satisfying all conditions.
-\ref{S1_Fig} shows a plot of $f(E_M)$ together with a biologically plausible Michaelis-Menten approximation (Eq.~\ref{eq:MMApprox}).
+In Text~\ref{S1_Text} it is shown that Eq.~\ref{eq:meth_f} is the only definition of $f$ satisfying all conditions.
+Fig.~\ref{S1_Fig}\textit{A} shows a plot of $f(E_M)$ together with a biologically plausible Michaelis-Menten approximation (Eq.~\ref{eq:MMApprox}).
 
@@
 \end{equation}
-As shown in \ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
+As shown in Text~\ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
 Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
@@
 Like for $f$, Eq.~\ref{eq:meth_g} is the only choice for $g$ satisfying C1-C3.
-\ref{S1_Fig} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).
+Fig.~\ref{S1_Fig}\textit{B} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).
 
@@
 This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in Fig.~\ref{fig:onoffresults}\textit{A}.
-In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
+In Text~\ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
 Moreover, the steady state coincides with the optimal steady state $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ by construction, so that Condition~C3 is satisfied as well.
@@
 
+}
+
 \section*{Supporting Information}
@@
 
-\subsection*{S1 Text -- Model derivation and analysis}
-\manuallabel{S1_Text}{S1~Text}
+\tr{
+\subsection*{Text S1 -- Model derivation and analysis}
+\manuallabel{S1_Text}{S1}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsection*{S2 Text -- Model parameters}
-\manuallabel{S2_Text}{S2~Text}
+\subsection*{Text S2 -- Model parameters}
+\manuallabel{S2_Text}{S2}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsection*{S3 Text -- Solution of optimal control problem}
-\manuallabel{S3_Text}{S3~Text}
+\subsection*{Text S3 -- Solution of optimal control problem}
+\manuallabel{S3_Text}{S3}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsection*{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
-\manuallabel{S4_Text}{S4~Text}
+\subsection*{Text S4 -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
+\manuallabel{S4_Text}{S4}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+\subsection*{Table S1 -- Parameter values of self-replicator model}
+\manuallabel{S1_Table}{S1}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsection*{S1 Table -- Parameter values of self-replicator model}
-\manuallabel{S1_Table}{S1~Table}
+\subsection*{Figure S1 -- Simple control strategies for the self-replicator of bacterial growth}
+\manuallabel{S1_Fig}{S1}
+%{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\subsection*{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth}
-\manuallabel{S1_Fig}{S1~Figure}
+}
 
 \section*{Acknowledgments}
+This work was partially supported by the Investissements d'Avenir Bio-informatique programme under project RESET (ANR-11-BINF-0005), the Inria Project Lab Algae in silico, and the Labex SIGNALIFE (ANR-11-LABX-0028-01).
 The authors thank Eugenio Cinquemani for discussions and comments on the manuscript.
@@
 % 
+\bibliography{references}
+
+\end{document}

### 2016-06-01 Chapters/S1_Figure.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}
% Add an S1
\renewcommand{\thefigure}{S1}%

\usepackage{color} 
\newcommand{\tr}[1]{{#1}}

\begin{document}
\begin{flushleft}
{\Large
\textbf\newline{\textbf{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/plot_strategies.eps}
\caption{
\noindent\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state, that is, the state in which growth occurs at the maximal rate supported by $E_M$.
\tr{The function $f$ is defined by Eq.~\ref{eq:meth_f} in the \textit{Methods} section of the main text.}
The solid, red curve is an approximation of this function by the simple Michaelis-Menten curve of Eq.~\ref{eq:MMApprox}, with $K_{mE} = 1.0$.
\textit{B:} Precursor-only strategy: $\alpha = g(\hat{p})$.
The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state.
\tr{The function $g$ is defined by Eq.~\ref{eq:meth_g} in the \textit{Methods} section of the main text.}
The solid, red curve is an approximation of this function by the simple Hill curve of Eq.~\ref{eq:HillApprox}, with $K_{mp} = 0.06$ and a cooperativity coefficient 2.}
\end{figure}

\end{document}

### 2016-06-01 Chapters/S1_Text.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}
% Add an S1
\renewcommand{\thetable}{S1.\arabic{table}}%
\renewcommand{\thefigure}{S1.\arabic{figure}}%
\renewcommand{\theequation}{S1.\arabic{equation}}
\renewcommand{\thesection}{S1.\arabic{section}}%

\usepackage{color} 
\newcommand{\tr}[1]{{#1}}

\begin{document}

\begin{flushleft}
{\Large
\textbf\newline{\textbf{S1 Text -- Model derivation and analysis\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}

\section{Model formulation}

The time evolution of the total mass of each component of the self-replicator can be written as follows:
\begin{eqnarray}
\frac{dP}{dt} &=& V_M(t) - V_R(t) \nonumber , \\
\frac{dM}{dt} &=& (1-\alpha(t))\, V_R(t) \label{eq:ext_system},\\
\frac{dR}{dt} &=& \alpha(t) \, V_R(t)  \nonumber ,
\end{eqnarray}
where $P$, $M$, $R$ [g] denote the total mass of precursors, metabolic machinery and gene expression machinery, respectively.
$V_M$ [g h$^{-1}$] is the rate of production of precursors by metabolism and $V_R$ [g h$^{-1}$] the rate of utilisation of precursors for gene expression.

Dividing the mass variables by the total time-varying volume $\texttt{Vol}(t)$ of the system, we obtain the concentration variables $p = P/\texttt{Vol}$, $m = M/\texttt{Vol}$, $r = R/\texttt{Vol}$ [g L$^{-1}$].
The dynamics of the concentration variables then follows with Eq.~\ref{eq:ext_system}:
\begin{eqnarray}
\frac{dp}{dt} &=& \frac{V_M(t)}{\texttt{Vol}} - \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt}\, p, \nonumber \\
\frac{dm}{dt} &=& (1-\alpha(t)) \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, m  \label{eq:deriv_int_system},\\ 
\frac{dr}{dt} &=& \alpha(t) \, \frac{V_R(t)}{\texttt{Vol}}  - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, r. \nonumber
\end{eqnarray}

At this point, we define $v_M = V_M/\texttt{Vol}$ and $v_R = V_R/\texttt{Vol}$ [g L$^{-1}$ h$^{-1}$] as the mass fluxes per unit volume.
Moreover, with the definition of the volume in terms of the total protein mass in Eq.~\ref{eq:voldef} of the main text, that is, $\texttt{Vol} = \beta\, (M + R)$, we find that
\begin{equation}
\label{eq:supp_deriv_growthrate}
\frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \frac{\beta}{\texttt{Vol}} \frac{d(M+R)}{dt} = \beta\, \frac{V_R(t)}{\texttt{Vol}} = \beta \, v_R(t).
\end{equation}
This leads to the system
\begin{eqnarray}
\frac{dp}{dt} &=& v_M(t) - v_R(t) \, (1+\beta\, p), \label{eq:supp_pdef}\\
\frac{dr}{dt} &=& v_R(t)  \, (\alpha(t) - \beta\, r) \label{eq:supp_rdef},
\end{eqnarray}
where the equation for $m(t)$ is omitted since by construction $r(t) + m(t) = 1/\beta$ and $dr/dt + dm/dt = 0$.

As stated in the main text, we use Michaelis-Menten kinetics to express $v_M$ and $v_R$ in terms of the system variables:
\begin{eqnarray}
v_M(t) &=& m(t) \, k_M \, \frac{s(t)}{K_M +s(t)} = \left(\frac{1}{\beta} - r(t)\right)\, e_M(t) \nonumber, \\
v_R(t) &=& r(t) \, k_R \, \frac{p(t)}{K_R +p(t)}, \nonumber 
\end{eqnarray}
with rate constants $k_M$, $k_R$ [h$^{-1}$] and half-saturation constants $K_M$, $K_R$ [g L$^{-1}$].
$s(t)$ is an exogenous variable representing the nutrient concentration in the external medium.
We simplify $v_M(t)$ by defining the environmental input $e_M(t) = k_M \, s(t) / (K_M + s(t))$.
\tr{Throughout the paper, as explained in the main text, we assume the environment is constant, \textit{i.e.}, $e_M(t)=e_M$.}

Finally, the growth rate $\mu$ [h$^{-1}$] is defined as the relative increase of the volume of the self-replicator.
From Eq.~\ref{eq:supp_deriv_growthrate}, it follows that:
\begin{equation}
\label{eq:supp_growthrate}
\mu (t) = \frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \beta\, v_R(t).
\end{equation}

\section{Nondimensionalization of the system}

For the sake of simplifying the proofs and derivations below, we define the following nondimensional variables:
\begin{equation*}
\hat{p}  = \beta \, p,\;\;\;
\hat{r}  = \beta \, r,\;\;\;
\hat{t}  = k_R \, t.
\end{equation*}
When injecting these into Eq.~\ref{eq:supp_pdef}, we obtain
\[
\frac{k_R}{\beta} \, \frac{d\hat{p}}{d\hat{t}} = \left( \frac{1}{\beta} - \frac{\hat{r}}{\beta} \right) \, e_M - \frac{\hat{r}}{\beta} \, k_R \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}),
\]
which simplifies to
\[
\frac{d\hat{p}}{d\hat{t}} = ( 1 - \hat{r} ) \, \frac{e_M}{k_R} - \hat{r} \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}).
\]
In a similar manner, we derive the time evolution of the nondimensional $\hat{r}$, and thus obtain the system
\begin{equation}
\label{eq:supp_adim}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
\frac{d\hat{r}}{d\hat{t}} &= (\alpha - \hat{r}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},
\end{aligned}
\end{equation}
with the lumped parameters $E_M = e_M / k_R$ and $K = \beta \, K_R$.
The corresponding nondimensionalized growth rate is given by
\begin{equation}
\label{eq:supp_growthrate_adim}
\hat{\mu} = \frac{\mu}{k_R} = \frac{\hat{p}}{K+\hat{p}} \, \hat{r}.
\end{equation}

\section{Steady-state growth of the self-replicator}
\label{si::growthrate}

If we suppose $E_M > 0$, $K > 0$ and $\alpha \in ]0, 1[$, there is a trivial unstable steady state at $(0, 1)$.
A second steady-state exists for the point in which $\hat{r}^* = \alpha$ and $\hat{p}^*$ is a root of the following polynomial:
\[
\alpha \, \hat{p}^2 + \left(\alpha - (1-\alpha) \, E_M \right) \hat{p} - (1-\alpha)\, E_M\, K.
\]
If we keep the only admissible root for this polynomial (\textit{i.e.}, for which $\hat{p} \geq 0$), the second steady state is given by
\begin{equation}
\label{eq:supp_steadystate}
(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
\end{equation}
We can determine the stability of this steady state by looking at the Jacobian matrix $J$ of the ODE system:
\begin{equation}
\label{eq:supp_jacop}
J = \left(\begin{matrix}
- \frac{\hat{r}}{K + \hat{p}} \left[ \hat{p} + (1+\hat{p})\, \frac{K}{K+\hat{p}}\right] & - E_M - (1+\hat{p})\frac{\hat{p}}{K+\hat{p}}\\
(\alpha - \hat{r})\, \hat{r} \, \frac{K}{(K+\hat{p})^2} & (\alpha - 2\hat{r})\, \frac{\hat{p}}{K+\hat{p}}
\end{matrix}\right) .
\end{equation}
Evaluated at the point $(\hat{p}^*, \hat{r}^*)$, the Jacobian matrix becomes
\[
J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
- \frac{\alpha}{K + \hat{p}^*} \left[ \hat{p}^* + (1+\hat{p}^*)\frac{K}{K+\hat{p}^*}\right] & - E_M - (1+\hat{p}^*)\frac{\hat{p}^*}{K+\hat{p}^*}\\
0 & -\alpha\frac{\hat{p}^*}{K+\hat{p}^*}
\end{matrix}\right).
\]
Since $\hat{p}^*$, $\alpha$, $E_M$, $K$ $>0$, the two eigenvalues are negative and therefore the steady state $(\hat{p}^*, \hat{r}^*)$ is stable (see also the streamlines in Figure~\ref{fig:model_analysis}\textit{A} in the main text).
It means that for fixed environmental conditions $E_M$ and resource allocation $\alpha$, the self-replicator converges towards a steady state in which the concentration variables are constant.

One can now easily derive the steady-state growth rate, denoted $\hat{\mu}^*$.
By substituting Eq.~\ref{eq:supp_growthrate_adim} into the first ODE of the system of Eq.~\ref{eq:supp_adim}, we find at steady state:
\[
\left(\frac{d\hat{p}}{d\hat{t}}\right)_{(\hat{p}^*, \hat{r}^*)} = 0 = (1-\alpha)\, E_M - (1+\hat{p}^*)\, \hat{\mu}^*,
\]
which by means of Eq.~\ref{eq:supp_steadystate} gives the following relation:
\begin{equation}
\label{eq:supp_growthrate_p}
\hat{\mu}^* = \frac{(1-\alpha)\, E_M}{1+\hat{p}^*} = \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
\end{equation}
Finally, we can transform this expression to obtain
\begin{equation}
\label{eq:supp_growthrate_final}
\hat{\mu}^* = \begin{cases}
\frac{(1-\alpha)\, E_M + \alpha - \sqrt{\left[(1-\alpha)\, E_M - \alpha\right]^2 + 4(1-\alpha)\, \alpha\, E_M\, K}}{2(1-K)} &\text{ for }K\neq 1,\\
\frac{\alpha\, (1-\alpha)\, E_M}{\alpha + (1-\alpha)\, E_M} &\text{ for }K = 1.
\end{cases}
\end{equation}
This function of $\alpha$ is plotted in Figure~\ref{fig:model_analysis}\textit{B} in the main text.

\section{Maximization of growth rate at steady state}
\label{si::optimal}

We are interested in the steady state at which growth occurs at the maximum rate.
The growth rate at steady state $\hat{\mu}^*$ is given by
\begin{equation}
\label{eq:supp_growthrate_steadystate}
\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
\end{equation}
From the first ODE of the system of Eq.~\ref{eq:supp_adim}, we have
\begin{equation}
\label{eq:supp_isop}
\hat{r}^* = \frac{E_M}{E_M + \frac{\hat{p}^*}{K + \hat{p}^*} (1+\hat{p}^*)} .
\end{equation}
Substituting Eq.~\ref{eq:supp_isop} into Eq.~\ref{eq:supp_growthrate_steadystate}, we obtain
\begin{equation}
\label{eq:supp_growthrate_steadystate_p}
\hat{\mu}^* = \frac{E_M \, \hat{p}^*}{\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K} .
\end{equation}
The value of $\hat{p}^*$ maximizing $\hat{\mu}^*$ can be determined from
\begin{equation}
\label{eq:supp_growthrate_steadystate_deriv_p}
\frac{\partial\hat{\mu}^*}{\partial\hat{p}^*} = \frac{E_M \, (E_M\, K - \hat{p}^{*2})}{\left(\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K\right)^2},
\end{equation}
by looking at the values of $\hat{p}^*$ for which this derivative equals 0.
It follows that $\hat{\mu}^*$ is maximal for
\begin{equation}
\label{eq:supp_p_optimal}
\hat{p}^* = \hat{p}^*_{opt} = \sqrt{K\, E_M}.
\end{equation}
By substituting $\hat{p}^*_{opt}$ and $\alpha_{opt}^*$ for $\hat{p}^*$ and $\hat{r}^*$, respectively, in Eq.~\ref{eq:supp_isop}, we obtain the resource allocation maximizing the growth rate
\begin{equation}
\label{eq:supp_alpha_optimal}
\alpha_{opt}^* = \frac{E_M + \sqrt{KE_M}}{E_M + 2\sqrt{KE_M} + 1}.
\end{equation}
Finally, injecting this result into Eq.~\ref{eq:supp_growthrate_steadystate} we obtain the optimal steady-state growth rate:
\begin{equation}
\label{eq:supp_growthrate_optimal}
\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1}.
\end{equation}
In addition, by using Eq. \ref{eq:supp_p_optimal}, we can write $\alpha_{opt}^*$ and $\hat{\mu}_{opt}^*$ as a function of $\hat{p}_{opt}^*$ only:
\begin{equation}
\label{eq:supp_alpha_mu_optimal_p}
\alpha_{opt}^* = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
\end{equation}

\tr{
\section{Analysis of the control strategies}
\label{si::control_strategies}

In this section, we derive the main results for the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies.
For each of these, we prove that the Conditions C1, C2 and C3 from the \textit{Methods} section are satisfied, which we repeat here for clarity:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}

\subsection{Nutrient-only strategy}

The nutrient-only strategy is defined by:
\begin{equation}
\label{eq:supp_f}
\alpha = f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K\, E_M} + 1}.
\end{equation}
It drives the system to the optimal steady state by measuring the environment $E_M$.
Note that Condition C1 is satisfied by definition.

By injecting Eq.~\ref{eq:supp_f} into Eq.~\ref{eq:supp_adim}, the ODE system under the control of $f$ becomes:
\begin{equation}
\label{eq:supp_adim_f}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
\frac{d\hat{r}}{d\hat{t}} &= \left(f(E_M) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
\end{aligned}
\end{equation}
Since $E_M$ is constant on the interval of interest (starting right after the upshift), we are in the case of Section~\ref{si::growthrate} (\textit{i.e.}, $\alpha$ constant).
In particular, the system has two steady states: a trivial unstable one at (0, 1) (with zero growth), and a stable one defined by Eq.~\ref{eq:supp_steadystate} (Condition C2).
Since $f(E_M) = \alpha_{opt}^*$, we conclude from the derivations in Section~\ref{si::optimal} that the stable steady state is optimal for every environment $E_M$ (Condition C3).

It is interesting to note that the expression in Eq.~\ref{eq:supp_f} is the only function $f(E_M)$ satisfying C1-C3.
We can prove this statement by contradiction.
Assume a control strategy $c(E_M)$ satisfying C1-C3, and different from $f(E_M)$, \textit{i.e.}, there exists $E_M={E_M}_1$ such that $c({E_M}_1) \neq f({E_M}_1)$.
In this environment, the system reaches a steady state $(\hat{p}_1^*, \hat{r}_1^*)$ with $\hat{r}_1^* = c({E_M}_1) \neq f({E_M}_1)$.
However, by Eq.~\ref{eq:supp_alpha_optimal} the optimal value for $\hat{r}^*$ in this environment is given by $f({E_M}_1)$.
So, the control law $c(E_M)$ does not drive the system to the optimal steady state in this environment, in contradiction with Condition~C3.

\subsection{Precursor-only strategy}

The precursor-only strategy is defined by:
\begin{equation}
\label{eq:supp_g}
\alpha = g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
\end{equation}
Here as well, C1 is satisfied by construction.

The ODE system under the control of $g$ becomes
\begin{equation}
\label{eq:supp_adim_g}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
\frac{d\hat{r}}{d\hat{t}} &= \left(g(\hat{p}) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
\end{aligned}
\end{equation}
The nullcline for $\hat{p}$ remains unchanged and is defined by
\begin{equation}
\label{eq:supp_pnullcline_g}
\frac{d\hat{p}}{dt} = 0 \Leftrightarrow 
\hat{r} = \frac{E_M}{E_M + \frac{\hat{p}}{K+\hat{p}} (1+\hat{p})},
\end{equation}
while the nullcline for $\hat{r}$ is
\begin{equation}
\label{eq:supp_rnullcline_g}
\frac{d\hat{r}}{dt} = 0 \Leftrightarrow \begin{cases}
\hat{p} = 0,\\
\hat{r} = 0,\\
\hat{r} = \frac{\hat{p}}{\hat{p} + \frac{K}{K+\hat{p}} (1+\hat{p})}.
\end{cases}
\end{equation}
Hence, we also have a trivial unstable steady state at (0,1) (with zero growth).
The second steady state is obtained from Eqs~\ref{eq:supp_pnullcline_g}-\ref{eq:supp_rnullcline_g}:
\begin{equation*}
\frac{E_M}{E_M + \frac{\hat{p}^*}{K+\hat{p}^*} (1+\hat{p}^*)} = \frac{\hat{p}^*}{\hat{p}^* + \frac{K}{K+\hat{p}^*} (1+\hat{p}^*)},
\end{equation*}
which we rearrange into
\begin{equation*}
\hat{p}^* E_M + \frac{K}{K+\hat{p}^*} (1+\hat{p}^*) E_M = \hat{p}^* E_M + \frac{\hat{p}^*}{K+\hat{p}^*} (1+\hat{p}^*) \hat{p}^*.
\end{equation*}
This leads to
\begin{equation*}
\hat{p}^* = \sqrt{K E_M},
\end{equation*}
and therefore
\begin{equation*}
\hat{r}^* = g(\hat{p}^*) = \frac{\sqrt{K E_M}}{\sqrt{K E_M} + \frac{K}{K + \sqrt{K E_M}} (1+\sqrt{K E_M})} = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K E_M} + 1}.
\end{equation*}
From Eqs~\ref{eq:supp_p_optimal}-\ref{eq:supp_alpha_optimal}, we recognize the optimal steady state for the environment $E_M$, validating Condition~C3.
We now look for the stability of this (optimal) steady state by deriving the Jacobian of this system:
\begin{equation}
\label{eq:supp_jacobian_g}
J = \left(\begin{matrix}
- \frac{\hat{r}}{K+\hat{p}} \frac{\hat{p}^2 + 2K\hat{p} + K}{\hat{p} + K} & - E_M - \frac{\hat{p}}{K+\hat{p}} (1+\hat{p}) \\
\frac{\hat{r}}{K+\hat{p}} \left[ \frac{K}{K+\hat{p}} (g(\hat{p}) - \hat{r}) + \hat{p}K\frac{\hat{p}^2 + 2\hat{p} + K}{(\hat{p}^2 + 2K\hat{p} + K)^2}\right] & \frac{\hat{p}}{K+\hat{p}} (g(\hat{p}) - 2\hat{r})
\end{matrix}\right) .
\end{equation}
Evaluated at $(\hat{p}^*, \hat{r}^*) = (\sqrt{K E_M}, g(\sqrt{K E_M}))$, the Jacobian becomes
\begin{equation}
\label{eq:supp_jacobian_g_eq}
J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
- \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} & - E_M - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} (1+\sqrt{K E_M}) \\
\frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}}\frac{K E_M + 2 \sqrt{K E_M} + K}{K (E_M + 2 \sqrt{K E_M} + 1)^2} g(\sqrt{K E_M}) & - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} g(\sqrt{K E_M})
\end{matrix}\right).
\end{equation}
Since $K$, $E_M$, and $g(\sqrt{K E_M)} > 0$, it follows immediately that the real part of the eigenvalues of this matrix are both negative.
\footnote{Notice that the eigenvalues $\lambda_1$ and $\lambda_2$ of $J_{(\hat{p}^*, \hat{r}^*)}$ satisfy the inequalities $\text{Tr}(J) = \lambda_1 + \lambda_2 < 0$ and $\det(J) = \lambda_1 \lambda_2 > 0$.}
Hence, the non-trivial steady state is stable, completing the proof of Condition~C2.

Here again, it is interesting to observe that the expression in Eq.~\ref{eq:supp_g} is the only function $g(\hat{p})$ satisfying C1-C3. This can be proven in a similar way as for $f$.

\subsection{On-off strategy}

The on-off strategy is defined by:
\begin{equation}
\alpha = h(\hat{p}, \hat{r}) = 
\begin{cases}
0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}
$h$ is a static function of $\hat{p}$ and $\hat{r}$ (Condition~C1).

As a consequence, the ODE system under the control of $h$ is given by
\begin{equation}
\label{eq:supp_adim_h}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
\frac{d\hat{r}}{d\hat{t}} &= \left(h(\hat{p}, \hat{r}) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
\end{aligned}
\end{equation}
Notice that the system has a discontinuitous right-hand side, due to the fact that $\alpha$ switches between 0 and 1 on $\hat{r}=g(\hat{p})$. 
Fig.~\ref{fig:sliding} shows the dynamics of the system in the phase plane.
Due to the direction of the vector fields relative to $\hat{r}=g(\hat{p})$, a \textit{sliding mode} occurs on the latter curve \cite{filippov_differential_1988}.  
The system is seen to evolve towards a locally asymptotically stable steady state, which is the single non-trivial steady state (Condition~C2). 
This steady state coincides with the intersection of $\hat{r}=g(\hat{p})$ and the $\hat{p}$-nullcline, which is the steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ allowing maximal growth, thus verifying Condition~C3.

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
\caption
{
\tr{
\textbf{Local stability of the on-off strategy.}
The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
The solid, black curve is the $\hat{p}$-nullcline.
The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
}
}
\label{fig:sliding}
\end{figure}

}

\bibliographystyle{plos2015}
\bibliography{references}

\end{document}

### 2016-06-01 Chapters/S2_Text.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}
% Add an S2
\renewcommand{\thetable}{S2.\arabic{table}}%
\renewcommand{\thefigure}{S2.\arabic{figure}}%
\renewcommand{\theequation}{S2.\arabic{equation}}
\renewcommand{\thesection}{S2.\arabic{section}}%

\begin{document}
\begin{flushleft}
{\Large
\textbf\newline{\textbf{S2 Text -- Model parameters\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}

Most of the conclusions of this paper are parameter-independent in the range of physically admissible values.
The exact parameter values used in the simulations aim to represent relevant orders of magnitude.
They were derived from the available literature on fast-growing bacteria (mostly \textit{Escherichia coli}).
This document describes this derivation for each parameter.
In the \textit{Methods} section of the main text, we also describe how some of them were validated by fitting the model to available experimental data (Fig.~\ref{fig:model_validation} in the main text).

The model parameters (Eqs.~\ref{eq:pdef}-\ref{eq:rdef} in the main text) are listed in the table below:
\begin{center}
\begin{tabular}{|c|c|l|}
\hline
Name & Unit & Description \\
\hline
$e_M$ & h$^{-1}$ & Constant characterizing nutrient composition of medium\\
\hline
$k_R$ & h$^{-1}$ & Rate constant of macromolecular synthesis\\
\hline
$K_R$ & g\ L$^{-1}$ & Half-saturation constant of macromolecular synthesis\\
\hline
$\beta$ & L\ g$^{-1}$ & Inverse of the cellular density of macromolecules\\
\hline
$\alpha$ & -- & Resource allocation parameter\\
\hline
\end{tabular}
\end{center}
The values derived below are summarized in \ref{S1_Table}.

\section*{\Large \texorpdfstring{$e_M$}{eM}}

By definition, $e_M$ is the effective turnover of the metabolic macroreaction producing precursors from external substrates, obtained by dividing the reaction rate $v_M$ by the enzyme concentration $m$ (Eq.~\ref{eq:metaflux}).
The unit of $e_M$ is min$^{-1}$, and can be decomposed as follows:
\[
[e_M]  = \frac{\text{[mass of metabolic product]}}{\text{[mass of enzyme M]}\cdot \text{[time]}} = \frac{1}{\text{[time]}}.
\]
Note that $e_M = k_M\, s / (K_M + s)$ where $k_M$ is a rate constant, indicating the maximal rate of conversion of external nutrients to precursor metabolites. 
$e_M$ will thus vary with the concentration $s$ of the external nutrients and the kind of nutrient. 
For example, the precursor mass that can be produced from 1 g of glucose is higher than that produced from 1 g of acetate.

How can we find a typical value for $k_M$, and thus for $e_M$ (both have the same order of magnitude if we suppose that the reaction is not operating far below saturation, that is, $e_M \approx k_M$)?
A reasonable estimate for $k_M$ can be obtained from the turnover numbers of reactions involved in the synthesis of charged tRNA, since the latter are directly consumed by the most abundant part of the gene expression machinery, the ribosomes.

Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glutaminyl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second. 
After conversion to mass units using molar weight from~\cite{freist_glutaminyltrna_1997}, this yields
\[
k_{\textit{cat,GlnRS}} = \frac{3.2 \cdot 147}{64.4 \cdot 10^ 3} \approx 10^{-3} \, \text{g of glutaminyl-tRNA} \cdot \text{g of enzyme}^{-1} \cdot \text{s}^{-1}.
\]
We therefore take
\[
k_M \approx 3.6 \text{ h}^{-1},
\]
and thus obtain an upper bound for $e_M$ in our simulations.

\section*{\Large \texorpdfstring{$k_R$}{kR}}

$k_R$ is the mass rate constant describing the maximal rate of conversion of precursors to macromolecules [h\textsuperscript{-1}].
As for $e_M$, we can decompose this into
\[
[k_R]  = \frac{\text{[mass of macromolecules]}}{\text{[mass of gene expression machinery]}\cdot \text{[time]}} = \frac{1}{\text{[time]}}
\]
To obtain an order of magnitude for the mass of macromolecules, we focus on proteins since they are the most abundant macromolecules in the cell~\cite{ehrenberg_mediumdependent_2012}.
The dimensional analysis of $k_R$ thus becomes:
\[
\begin{aligned}
\left[k_R\right] &= \frac{\text{[protein mass produced]}}{\text{[ribosomal mass]} \cdot \text{[hour]}}\\
&= \frac{\text{[moles of protein]} \cdot \text{[protein molar mass]}}{\text{[moles of ribosome]} \cdot \text{[ribosome molar mass]} \cdot \text{[hour]}}\\
&= \frac{\text{[moles of amino acids]} \cdot \text{[molar mass of amino acids]}}{\text{[moles of ribosome]} \cdot \text{[hour]} \cdot \text{[ribosome molar mass]}}\\
&\approx \frac{\text{[maximal protein elongation rate]} \cdot \text{[molar mass of amino-acids]}}{\text{[ribosome molar mass]}}.
\end{aligned}
\]
The values in the last equality are available from the literature~\cite{ehrenberg_mediumdependent_2012,klumpp_molecular_2013,hachiya_increase_2007,yamamoto_mass_2006}.
We obtain
\[
k_R \approx \frac{10 \cdot 100}{10^6}\cdot 3600 \approx 3.6 \text{ h}^{-1}.
\]
This value is comparable with the translational capacity $k_T$, in $\mu$g of protein per $\mu$g of ribosomal protein per hour, given by Scott \textit{et al.}~\cite{scott_interdependence_2010}:
\[
k_T = \frac{4.5 \text{ }\mu\text{g of protein} \text{ / }\mu\text{g of RNA} \text{ / h }}{0.76 \text{ }\mu\text{g of ribosomal protein} \text{ / } \mu\text{g of RNA}} = 5.9 \text{ h}^{-1}.
\]

\section*{\Large \texorpdfstring{$K_R$}{KR}}

A value for the parameter $K_R$, representing the half-saturation constant of macromolecular synthesis, is more difficult to obtain from the literature.
However, assuming that ribosomes operate close to saturation (80\% over a range of growth rates~\cite{ehrenberg_mediumdependent_2012}), we find that $K_R \approx 0.25\,p$, with p the total amino acid concentration.
The total concentration of amino acids in the cell is around 150~mmol~L\textsuperscript{-1}~\cite{bennett_absolute_2009}, which with a mean molecular weight of 118.9~g~mol\textsuperscript{-1} for amino acids~\cite{hachiya_increase_2007}, yields a mass concentration of 17.8~g~L\textsuperscript{-1}.
These considerations led to the following order of magnitude for $K_R$:
\[
K_R \approx 1\text{ g L}^{-1}.
\]

\section*{\Large \texorpdfstring{$\beta$}{Beta}}

$\beta$ is the inverse of the cellular density of macromolecules, which has been shown constant during balanced growth over a large range of growth rates~\cite{churchward_macromolecular_1982}, and there is some data suggesting that $\beta$ varies little during growth transitions as well~\cite{zhou_carbon_2013}.
From~\cite{zimmerman_estimation_1991,mcguffee_diffusion_2010} we take the following typical value for $\beta$:
\[
\beta \approx \frac{1}{300} \approx 0.003 \text{ L g}^{-1}.
\]

\section*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}

From the values of the parameter in the dimensional model, one can deduce the parameters in the nondimensional model used in the simulations:
\[
E_M = \frac{e_M}{k_R} = \frac{3.6}{3.6} = 1 \;\;\;\;\; , \;\;\;\;\;  K = \beta\, K_R = 3\cdot 10^{-3} \cdot 1 = 0.003.
\]

\bibliographystyle{plos2015}
\bibliography{references}

\end{document}

### 2016-06-01 Chapters/S4_Text.tex
First meeting with H&H

\documentclass[a4paper,12pt]{article}

\input{suppinfo_preambule}
% Add an S4
\renewcommand{\thetable}{S4.\arabic{table}}%
\renewcommand{\thefigure}{S4.\arabic{figure}}%
\renewcommand{\theequation}{S4.\arabic{equation}}
\renewcommand{\thesection}{S4.\arabic{section}}%

\usepackage{color} 
\newcommand{\tr}[1]{{#1}}

\begin{document}
\begin{flushleft}
{\Large
\textbf\newline{\textbf{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}\footnote{Supporting Information of "\input{title}"}}}
}
\newline
\input{authors_affil}
\end{flushleft}

The recently published model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015} provides a synthesis of the currently available knowledge of the ppGpp regulatory system.
\tr{Through the mechanisms of ppGpp production and degradation, it describes regulation of the synthesis of ribosomal RNA.}
We explain below how we use the model to compare the action of the ppGpp system with the on-off control strategy.
The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} \tr{, and is reproduced in Table~\ref{S4.tab:notations} in order to make the text self-contained.}

\tr{The evolution of the cellular concentration of ppGpp is described in~\cite{bosdriesz_how_2015} by}
\begin{equation}
\label{eq:ppGpp}
\frac{d\texttt{ppGpp}}{dt} = v_{RelA}(r_{t,tot}) + v_{spoT} - k_{spoT} \cdot \texttt{ppGpp},
\end{equation}
\tr{where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{S4.tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:} 
\begin{equation}
\label{eq:vrelA}
v_{RelA}(r_{t,tot}) = k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}}{K_{D,RelA} + r_{t,tot}}.
\end{equation}
\tr{The amount of stalled ribosomes is determined by the equilibrium between charged and uncharged tRNA, $t_{ai}$ and $t_{i}$, in the cell:}
\begin{equation}
\label{eq:rttot}
r_{t,tot} = \sum_i r_{ti} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+t_{ai}/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
\tr{which can be rewritten as
\begin{equation}
\label{eq:rttot-f}
r_{t,tot} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
using the assumption that $t_{tot,i}  = t_{ai} + t_i = 0.5 \cdot r$.
$r_i$ denotes the concentration of ribosomes recognizing amino acid  $i$.
Finally, with $r = \sum_i r_i$ the total ribosome concentration and $a_i$ the concentration of amino acid $i$, the dynamics of the charged tRNA concentration is described by
\begin{equation}
\label{eq:ti_dynamic}
\frac{dt_{ai}}{dt} = v_{tai}(a_i, t_{i}) - f_i \cdot v_{ribosome}(t_i,r),
\end{equation}
with $v_{tai}(a_i, t_{i})$ the synthesis rate of charged tRNA, and $f_i \cdot v_{ribosome}(t_i,r)$ their consumption via protein synthesis.
In particular,
\begin{eqnarray}
v_{tai}(a_i, t_{i}) &=& k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\,  K_{Mai} + a_i\,  K_{Mti} + t_i\,  a_i}, \label{eq:vtai}\\
v_{ribosome}(t_i, r) &=& k_{rib} \cdot r \cdot \left(1+ \sum_i \left[ f_i \cdot \left( 1 + \frac{t_i}{\kappa_t} \right) \frac{\kappa_{ta}}{0.5\cdot r - t_i} \right]\right)^{-1}. \label{eq:vrib}
\end{eqnarray}

For comparison with our framework, we need $\texttt{ppGpp}$ as a direct function of the total amino acid concentration $a = \sum_i a_i$ (a proxy for precursors) and total ribosome concentration $r$ (a proxy for gene expression machinery).
To this end, we made two additionnal assumptions}:
\begin{description}
\item[(A1)] All concentrations specific to one type of amino acid $i$ ($a_i$, $t_{ai}$, $t_{i}$, $r_i$) are in the same proportion $f_i = f = 1/20$ with respect to the total concentrations ($a$, $t_{a}$, $t$, $r$).
\item[(A2)] We apply a quasi-steady-state approximation (QSSA) to the dynamics of the concentration of the charged tRNAs ($t_{ai}$) and the concentration of ppGpp ($\texttt{ppGpp}$).
That is, the dynamics of these variables are assumed fast relative to the dynamics of the amino acid concentrations ($a_i$) and the total ribosome concentration ($r$).
\end{description}
\tr{Using (A2), we can rewrite Eq.~\ref{eq:ti_dynamic} as follow:
\begin{equation}\label{eq:vtai_vrib}
v_{tai}(a_i, t_{i}) = f_i \cdot v_{ribosome}(t_i,r),
\end{equation}
which, using (A1) and Eqs~\ref{eq:vtai} and \ref{eq:vrib}, leads to:
\begin{equation}
\label{eq:vtai_vrib_without_sum}
k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\, K_{Mai} + a_i\, K_{Mti} + t_i\, a_i}
= f_i \cdot k_{rib} \cdot r \cdot
  \left(1 + \frac{\kappa_{ta}}{0.5 r - t_i} \cdot \left( 1 + \frac{t_i}{\kappa_t} \right)  \right)^{-1}.
\end{equation}
By rearranging both sides of the equation, $t_i$ can be expressed as a function of $a_i$ and $r$, which yields:}

\begin{equation}
\begin{alignedat}{2}
&A {t_i}^2 + B t_i + C = 0, \;\;\; \text{with}\\
&A = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i\, k_{rib} \,r} \left( \frac{\kappa_{ta}}{\kappa_t} - 1\right) + K_{Mai} + a_i,\\
&B = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i \,k_{rib} \,r} (0.5\,r+\kappa_{ta}) + a_i \,K_{Mti} - 0.5\,r\,(K_{Mai} + a_i),\\
&C = - 0.5 \,r \,a_i \,K_{Mti},
\end{alignedat}
\end{equation}
and therefore
\[
t_i (a_i, r) = \frac{-B \pm \sqrt{B^2 - 4AC}}{2A}.
\]
It is not difficult to show that the only solution on $[0, 0.5\,r]$ is
\begin{equation}
\label{eq:ti_final}
t_i (a_i, r) = \frac{-B + \sqrt{B^2 - 4AC}}{2A}.
\end{equation}
\tr{
From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{S4.eq:rttot-f}:
\begin{equation}
\label{eq:rttot-f_without_sum}
r_{t,tot}(t_i, r) = r \cdot \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
and substituting $t_i$ by the expression of Eq.~\ref{eq:ti_final}.

Finally, we apply (A2) to Eq.~\ref{eq:ppGpp} and obtain the final expression giving the concentration of ppGpp as a function of the total amino acid and ribosome concentrations:
\begin{equation}
\texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
\end{equation}
}
This function is represented in Fig.~\ref{S4.fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.

The plotted surface of the function resembles the inverse of the on-off control strategy in Fig.~\ref{fig:ppGppsurface}, as expected, bearing in mind that ppGpp has an inhibitory effect on the synthesis of ribosomal RNA.
We assumed a Michaelis-Menten inhibition for the regulatory effect of ppGpp on rRNA synthesis, and thus indirectly on the synthesis of ribosomal proteins~\cite{potrykus_pppgpp_2008,keener_regulation_1996}:
\begin{equation}
\alpha (\texttt{ppGpp}) = \frac{K_I}{K_I + \texttt{ppGpp}}.
\end{equation}
The inhibitory constant $K_I$ lies in the dynamical range of variation of $\texttt{ppGpp}$.
In Fig.~\ref{fig:ppGppsurface} in the main text, we took $K_I = 10$~$\mu$M.

\bibliographystyle{plos2015}
\bibliography{references}

\pagebreak
\begin{figure}[!ht]
\centering
\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1.eps}
\caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
{
{\bf ppGpp concentration is a function of total ribosome and amino acid concentrations.}\newline
We assume the dynamics of ppGpp to be fast on the time-scale of changes in the ribosome and amino acid concentrations.
The concentration of ppGpp can thus be expressed as a function of the latter two variables, using the model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}.
Parameters are taken from Table~\ref{tab:notations}.
}
\label{fig:ppGpp}
\end{figure}

\pagebreak

\begin{table}[!h]
\centering
\tr{
\begin{tabular}{cccl}
\hline
\hline
\textbf{Symbol} & \textbf{Value} & \textbf{Unit} & \textbf{Description} \\
\hline
\hline
$a_i$ & -- & $\mu$M & Concentration of aa $i$ (not incorporated in protein)\\

$t_{ai}$ & -- & $\mu$M & Concentration of tRNA charged with aa $i$\\

$t_{i}$ & -- & $\mu$M & Concentration of free tRNA conjugate to aa $i$\\

$t_{tot,i}$ &  $0.5 \cdot r$ & $\mu$M & Total concentration of tRNA conjugate to aa $i$\\

$r_{i}$ & -- & $\mu$M & Total concentration of ribosome with an A-site for aa $i$\\

$r_{ti}$ & -- & $\mu$M & Ribosomes with uncharged tRNA in an A-site for aa $i$\\

$\texttt{ppGpp}$ & -- & $\mu$M & Concentration of ppGpp\\
\hline
$a$ & $\sum_i a_{i}$ & $\mu$M & Total concentration of aa (not incorporated in protein)\\

$t_{a}$ & $\sum_i t_{ai}$ & $\mu$M & Total concentration of tRNA charged with aa\\

$t$ & $\sum_i t_{i}$ & $\mu$M & Total concentration of free tRNA\\

$r_{t,tot}$ & $\sum_i r_{ti}$ & $\mu$M & Total concentration of uncharged tRNA bound to ribosomes\\

$r$ & $\sum_i r_i$ & $\mu$M & Total concentration of ribosomes\\
\hline
$v_{RelA}$ & -- & $\mu$M/s & Rate of RelA-catalyzed ppGpp synthesis\\

$v_{SpoT}$ & $10^{-3}$ & $\mu$M/s & Rate of ppGpp synthesis by SpoT\\

$v_{tai}$ & -- & $\mu$M/s & Rate of amino-acyl tRNA $i$ synthetase\\

$v_{ribosome}$ & -- & $\mu$M/s & Total rate of protein synthesis\\
\hline
$k_{rib}$ & 20 & s\textsuperscript{-1} & $k_{cat}$ of protein elongation\\

$k_{RelA}$ & 75 & s\textsuperscript{-1} & $k_{cat}$ of ppGpp synthesis by RelA\\

$K_{D,RelA}$ & 0.26 & $\mu$M & Michaelis constant of RelA-catalyzed ppGpp production\\

$RelA_{tot}$ & 1/15 & $\mu$M & RelA concentration\\

$k_{SpoT}$ & $\ln (2) / 30$ & s\textsuperscript{-1} & Rate of ppGpp degradation by SpoT\\

$\kappa_t$ & 500 & $\mu$M & Dissociation constant of uncharged tRNA-ribosome complex\\

$\kappa_{ta}$ & 1 & $\mu$M & Dissociation constant of charged tRNA-ribosome complex\\

$k_{Si}$ & 100 & s\textsuperscript{-1} & $k_{cat}$ of aminoacyl-tRNA synthetase\\

$S_{tot,i}$ & 1 & $\mu$M & Total concentration of aminoacyl-tRNA synthetase for aa $i$\\

$K_{Mai}$ & 100 & $\mu$M & Michaelis constant of aa-tRNA synthetase for amino acids\\

$K_{Mti}$ & 1 & $\mu$M & Michaelis constant of aa-tRNA synthetase for uncharged tRNA\\

$f_i$ & 1/20 & -- & Proportion of aa $i$ in proteins (Assumption A1)\\

\end{tabular}
}
\caption{\textbf{Parameters and variables reused from Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}}.} The abbreviation aa denotes amino acids.
\label{tab:notations}
\end{table}

\end{document}

### 2016-06-01 Chapters/author_summary.tex
First meeting with H&H

\section*{Author Summary}
Microbial growth is the process by which cells sustain and reproduce themselves from available matter and energy.
Strategies enabling microorganisms to optimize their growth rate have been extensively studied, but mostly in stable environments.
Here, we build a coarse-grained model of microbial growth and use methods from optimal control theory to determine a resource allocation scheme that would lead to maximal biomass accumulation when the cells are dynamically shifted from one growth medium to another.
We compare this optimal solution with several cellular implementations of growth control, based on the capacity of the cell to sense different physiological variables.
We find that strategies maximizing growth in steady-state conditions perform quite differently in dynamical conditions.  
Moreover, the control strategy with performance close to the theoretical maximum exploits information of more than one physiological variable, suggesting that optimization of microbial growth in dynamical rather than steady environments requires broader sensory capacities.
Interestingly, the ppGpp alarmone system in the enterobacterium \textit{Escherichia coli}, known to play an important role in growth control, has structural similarities with the control strategy approaching the theoretical maximum.
It senses a discrepancy between the concentrations of precursors and ribosomes, and adjusts ribosome synthesis in an on-off fashion.
This suggests that \textit{E. coli} is adapted for environments with intermittent, rapid changes in nutrient availability.

### 2016-06-01 Chapters/authors_affil.tex
First meeting with H&H

% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Nils Giordano \textsuperscript{1,3},
Francis Mairet \textsuperscript{2},
Jean-Luc Gouzé \textsuperscript{2},
Johannes Geiselmann \textsuperscript{1,3,*},
Hidde de Jong \textsuperscript{3,*}
\\
\bigskip
\bf{1} Université Grenoble Alpes, Laboratoire Interdisciplinaire de Physique (CNRS UMR 5588), 140 rue de la Physique BP 87, 38402 Saint Martin d'Hères, France
\\
\bf{2} Inria, Sophia-Antipolis Méditerranée research centre, 2004 route des Lucioles, BP 93, 06902 Sophia-Antipolis Cedex, France
\\
\bf{3} Inria, Grenoble - Rhône-Alpes research centre, 655 avenue de l'Europe, Montbonnot, 38334 Saint Ismier Cedex, France
\\
\bigskip

* Corresponding authors with equal contributions:\\
Hidde.de-Jong@inria.fr, Hans.Geiselmann@ujf-grenoble.fr

### 2016-06-01 Chapters/discussion.tex
First meeting with H&H

\chapter{Discussion}

- in praise of simple models
- an intrinsic link between complexity and environment dynamic
- some perspectives for industrial production

### 2016-06-01 Chapters/preambule.tex
First meeting with H&H

%\usepackage{epic}
\usepackage[english]{babel}
%\usepackage{appendix}
%\usepackage{lipsum}
%\usepackage{natbib}

\usepackage[hypcap]{caption}

%%%%% Math Stuff
\usepackage{amsmath}
\usepackage{amsfonts}
% New operator
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
% Theorems
\usepackage{amsthm}
\newtheorem{prop}{Proposition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Réglage des interlignes
%\usepackage{setspace}
%\onehalfspacing

% For roman numbers
%\renewcommand\thesection{\Roman{section}}

% New colors
%\usepackage{color} % for color stuff…
%\definecolor{lightred}{rgb}{1, 0.3, 0.3}

\usepackage{hyperref}
\hypersetup{
unicode=false, % non-Latin characters in Acrobat’s bookmarks
pdftoolbar=true, % show Acrobat’s toolbar?
pdfmenubar=true, % show Acrobat’s menu?
pdffitwindow=false, % window fit to page when opened
pdfstartview={FitH}, % fits the width of the page to the window
pdftitle={Environment dynamic structures optimality in cellular economics}, % title
pdfauthor={Nils Giordano}, % author
pdfsubject={Machinery optimal control article}, % subject of the document
pdfcreator={pdflatex}, % creator of the document
pdfproducer={LateX}, % producer of the document
pdfkeywords={gene expression machinery} {optimal control} {cell composition}, % list of keywords
pdfnewwindow=true, % links in new window
colorlinks=true, % false: boxed links; true: colored links
linkcolor=blue, % color of internal links (change box color with linkbordercolor)
citecolor=blue, % color of links to bibliography
filecolor=magenta, % color of file links
urlcolor=cyan % color of external links
}

### 2016-06-01 Chapters/suppinfo.tex
First meeting with H&H

\section{Model derivation and analysis}
\label{sec:supp_S1}

\subsection{Model formulation}

The time evolution of the total mass of each component of the self-replicator can be written as follows:
\begin{eqnarray}
\frac{dP}{dt} &=& V_M(t) - V_R(t) \nonumber \\
\frac{dM}{dt} &=& (1-\alpha(t))\, V_R(t) \label{eq:ext_system}\\
\frac{dR}{dt} &=& \alpha(t) \, V_R(t)  \nonumber ,
\end{eqnarray}
where $P$, $M$, $R$ [g] denote the total mass of precursors, metabolic machinery and gene expression machinery, respectively.
$V_M$ [g min$^{-1}$] is the rate of production of precursors by metabolism and $V_R$ [g min$^{-1}$] the rate of utilisation of precursors for gene expression.

Dividing the mass variables by the total time-varying volume $\texttt{Vol}(t)$ of the system, we obtain the concentration variables $p = P/\texttt{Vol}$, $m = M/\texttt{Vol}$, $r = R/\texttt{Vol}$ [g L$^{-1}$].
The dynamics of the concentration variables then follows with Eq.~\ref{eq:ext_system}:
\begin{eqnarray}
\frac{dp}{dt} &=& \frac{V_M(t)}{\texttt{Vol}} - \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt}\, p, \nonumber \\
\frac{dm}{dt} &=& (1-\alpha(t))\cdot \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, m  \label{eq:deriv_int_system}\\ 
\frac{dr}{dt} &=& \alpha(t) \, \frac{V_R(t)}{\texttt{Vol}}  - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, r. \nonumber
\end{eqnarray}

At this point, we define $v_M = V_M/\texttt{Vol}$ and $v_R = V_R/\texttt{Vol}$ [g L$^{-1}$ min$^{-1}$] as the mass fluxes per unit volume.
Moreover, with the definition of the volume in terms of the total protein mass in Eq.~\ref{eq:voldef}, that is, $\texttt{Vol} = \beta\, (M + R)$, we find that:
\begin{equation}
\label{eq:sup_deriv_growthrate}
\frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \frac{\beta}{\texttt{Vol}} \frac{d(M+R)}{dt} = \beta\, \frac{V_R(t)}{\texttt{Vol}} = \beta \, v_R(t).
\end{equation}
This leads to the system:
\begin{eqnarray}
\frac{dp}{dt} &=& v_M(t) - v_R(t) \, (1+\beta\, p), \label{eq:sup_pdef}\\
\frac{dr}{dt} &=& v_R(t)  \, (\alpha(t) - \beta\, r), \label{eq:sup_rdef},
\end{eqnarray}
where the equation for $m(t)$ is omitted since by construction $r(t) + m(t) = 1/\beta$ and $dr/dt + dm/dt = 0$.

As stated in the main text, we use Michaelis-Menten kinetics to express $v_M$ and $v_R$ in terms of the system variables:
\begin{eqnarray}
v_M(t) &=& m(t) \, k_M \, \frac{s(t)}{K_M +s(t)} = \left(\frac{1}{\beta} - r(t)\right)\, e_M(t) \nonumber \\
v_R(t) &=& r(t) \, k_R \, \frac{p(t)}{K_R +p(t)}, \nonumber 
\end{eqnarray}
with rate constants $k_M$, $k_R$ [min$^{-1}$] and half-saturation constants $K_M$, $K_R$ [g L$^{-1}$].
$s(t)$ is an exogenous variable representing the nutrient concentration in the external medium.
We simplify $v_M(t)$ by defining the environmental input $e_M(t) = k_M \, s(t) / (K_M + s(t))$.

Finally, the growth rate $\mu$ [min$^{-1}$] is defined as the relative increase of the volume of the self-replicator.
From Eq.~\ref{eq:sup_deriv_growthrate}), it follows that:
\begin{equation}
\label{eq:sup_growthrate}
\mu (t) = \frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \beta\, v_R(t).
\end{equation}

\subsection{Nondimensionalization of the system}

For the sake of simplifying the proofs and derivations below, we simplify the system by defining the following nondimensional variables:
\begin{equation*}
\hat{p}  = \beta \, p,\;\;\;
\hat{r}  = \beta \, r,\;\;\;
\hat{t}  = k_R \, t.
\end{equation*}
When injecting these into Eq.~\ref{eq:sup_pdef}, we obtain:
\[
\frac{k_R}{\beta} \, \frac{d\hat{p}}{d\hat{t}} = \left( \frac{1}{\beta} - \frac{\hat{r}}{\beta} \right) \, e_M - \frac{\hat{r}}{\beta} \, k_R \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}),
\]
which simplifies to:
\[
\frac{d\hat{p}}{d\hat{t}} = ( 1 - \hat{r} ) \, \frac{e_M}{k_R} - \hat{r} \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}).
\]
In a similar manner, we derive the time evolution of the nondimensional $\hat{r}$, and thus obtain the system
\begin{equation}
\label{eq:sup_adim}
\begin{aligned}
\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
\frac{d\hat{r}}{d\hat{t}} &= (\alpha - \hat{r}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},
\end{aligned}
\end{equation}
with the lumped parameters $E_M = e_M / k_R$ and $K = \beta \, K_R$.
The corresponding nondimensionalized growth rate is given by:
\begin{equation}
\label{eq:sup_growthrate_adim}
\hat{\mu} = \frac{\mu}{k_R} = \frac{\hat{p}}{K+\hat{p}} \, \hat{r}.
\end{equation}

\subsection{Steady-state growth of the self-replicator}
\label{si::growthrate}

If we suppose $E_M > 0$, $K > 0$ and $\alpha \in ]0, 1[$, there is a trivial unstable steady state at the point $(0, 0)$.
A second steady-state exists for the point in which $\hat{r}^* = \alpha$ and $\hat{p}^*$ is a root of the following polynomial:
\[
\alpha \, \hat{p}^2 + \left(\alpha - (1-\alpha) \, E_M \right) \hat{p} - (1-\alpha)\, E_M\, K.
\]
If we keep the only admissible root for this polynomial (\textit{i.e.}, for which $\hat{p} \geq 0$), the second steady state is given by:
\begin{equation}
\label{eq:sup_steadystate}
(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
\end{equation}
We can determine the stability of this steady state by looking at the Jacobian matrix $J$ of the ODE system:
\begin{equation}
\label{eq:sup_jacop}
J = \left(\begin{matrix}
- \frac{\hat{r}}{K + \hat{p}} \left[ \hat{p} + (1+\hat{p})\, \frac{K}{K+\hat{p}}\right] & - E_M - (1+\hat{p})\frac{\hat{p}}{K+\hat{p}}\\
(\alpha - \hat{r})\, r \, \frac{K}{(K+\hat{p})^2} & (\alpha - 2\hat{r})\, \frac{\hat{p}}{K+\hat{p}}
\end{matrix}\right) .
\end{equation}
Evaluated at the point $(\hat{p}^*, \hat{r}^*)$, the Jacobian matrix becomes:
\[
J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
- \frac{\alpha}{K + \hat{p}^*} \left[ \hat{p}^* + (1+\hat{p}^*)\frac{K}{K+\hat{p}^*}\right] & - E_M - (1+\hat{p}^*)\frac{\hat{p}^*}{K+\hat{p}^*}\\
0 & -\alpha\frac{\hat{p}^*}{K+\hat{p}^*}
\end{matrix}\right).
\]
Since $\hat{p}^*$, $\alpha$, $E_M$, $K$ $>0$, the two eigenvalues are negative and therefore the steady state $(\hat{p}^*, \hat{r}^*)$ is stable (see also the streamlines in Figure~\ref{fig:isoclines}\textit{B} in the main text).
It means that for fixed environmental conditions $E_M$ and resource allocation $\alpha$, the self-replicator converges towards a steady state in which all intensive variables are constant, also called balanced growth [Fishov95].

One can now easily derive the steady-state growth rate, denoted $\hat{\mu}^*$.
By substituting Eq.~\ref{eq:sup_growthrate_adim} into the first ODE of the system of Eq.~\ref{eq:sup_adim}, we find at steady state:
\[
\left(\frac{d\hat{p}}{d\hat{t}}\right)_{(\hat{p}^*, \hat{r}^*)} = 0 = (1-\alpha)\, E_M - (1+\hat{p}^*)\, \hat{\mu}^*,
\]
which by means of Eq.~\ref{eq:sup_steadystate} gives the following relation:
\begin{equation}
\label{eq:sup_growthrate_p}
\hat{\mu}^* = \frac{(1-\alpha)\, E_M}{1+\hat{p}^*} = \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
\end{equation}
Finally, we can transform this expression to obtain:
\begin{equation}
\label{eq:sup_growthrate_final}
\hat{\mu}^* = \begin{cases}
\frac{(1-\alpha)\, E_M + \alpha - \sqrt{\left[(1-\alpha)\, E_M - \alpha\right]^2 + 4(1-\alpha)\, \alpha\, E_M\, K}}{2(1-K)} &\text{ for }K\neq 1,\\
\frac{\alpha\, (1-\alpha)\, E_M}{\alpha + (1-\alpha)\, E_M} &\text{ for }K = 1.
\end{cases}
\end{equation}
This function of $\alpha$ is plotted in Figure~\ref{fig:isoclines}\textit{A} in the main text.

\subsection{Maximization of growth rate at steady state}
\label{si::optimal}

We are interested in the steady state at which growth occurs at the maximum rate.
Growth rate at steady state $\hat{\mu}^\star$ is given by:
\begin{equation}
\label{eq:sup_growthrate_steadystate}
\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
\end{equation}
From the first ODE of the system of Eq.~\ref{eq:sup_adim}, we have:
\begin{equation}
\label{eq:sup_isop}
\hat{r}^* = \frac{E_M}{E_M + \frac{\hat{p}^*}{K + \hat{p}^*} (1+\hat{p}^*)} .
\end{equation}
Substituting Eq.~\ref{eq:sup_isop} into Eq.~\ref{eq:sup_growthrate_steadystate}, we obtain:
\begin{equation}
\label{eq:sup_growthrate_steadystate_p}
\hat{\mu}^* = \frac{E_M \, \hat{p}^*}{\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K} .
\end{equation}
The value of $\hat{p}^*$ maximizing $\hat{\mu}^*$ can be determined from
\begin{equation}
\label{eq:sup_growthrate_steadystate_deriv_p}
\frac{\partial\hat{\mu}^*}{\partial\hat{p}^*} = \frac{E_M \, (E_M\, K - \hat{p}^{*2})}{\left(\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K\right)^2},
\end{equation}
by looking at the values of $\hat{p}^*$ for which this derivative exceeds 0.
It follows that $\hat{\mu}^*$ is maximal for:
\begin{equation}
\label{eq:sup_p_optimal}
\hat{p}^* = \hat{p}^*_{opt} = \sqrt{K\, E_M}.
\end{equation}
By substituting $\hat{p}^*_{opt}$ and $\alpha_{opt}$ for $\hat{p}^*$ and $\hat{r}^*$, respectively, in Eq.~\ref{eq:sup_isop}, we obtain the resource allocation maximizing the growth rate:
\begin{equation}
\label{eq:sup_alpha_optimal}
\alpha_{opt} = \frac{E_M + \sqrt{KE_M}}{E_M + 2\sqrt{KE_M} + 1}.
\end{equation}
Finally, injecting this result in Eq. \ref{eq:sup_growthrate_steadystate} we obtain the optimal steady-state growth rate:
\begin{equation}
\label{eq:sup_growthrate_optimal}
\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1}.
\end{equation}
In addition, by using Eq. \ref{eq:sup_p_optimal}, we can write $\alpha_{opt}$ and $\hat{\mu}_{opt}^*$ as a functions of $\hat{p}_{opt}^*$ only:
\begin{equation}
\label{eq:sup_alpha_mu_optimal_p}
\alpha_{opt} = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
\;\;\;\;\;\;\; ; \;\;\;\;\;\;\;
\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
\end{equation}

\section{Model parameters}
\label{sec:supp_S2}

Most of the conclusions of this paper are parameter-independent in the range of physically admissible values.
The parameter values used in the simulations are orders of magnitude based on data reported in the literature for fast-growing bacteria (mostly \textit{Escherichia coli}).
As a consistency test, some of the parameter values where validated by fitting the model to experimental data at steady state (Figure~\ref{fig:isoclines} in the main text).
The parameters of the model of Eqs.~\ref{eq:pdef} and \ref{eq:rdef} in the main text and their values used in the simulations are listed in the table below:

\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Name & Unit & Description \\
\hline
$e_M$ & h$^{-1}$ & Constant characterizing nutrient composition of medium\\
\hline
$k_R$ & h$^{-1}$ & Rate constant of macromolecular synthesis\\
\hline
$K_R$ & g\ L$^{-1}$ & Half-saturation constant of macromolecular synthesis\\
\hline
$\beta$ & L\ g$^{-1}$ & Inverse of the cellular density of macromolecules\\ % & Inverse de la molarit\'e totale & $\frac{118.9}{400} \approx 0.3 $ & \cite{totconc,weightaa}\\
\hline
$\alpha$ & $\emptyset$ & Resource allocation parameter\\
\hline
\end{tabular}
\end{center}

\textbf{COMMENT: still needs to be decided if we speak about synthesis of macromolecules or protein synthesis. }

\subsection{Parameter values used in simulations}

\subsubsection{\texorpdfstring{$e_M$}{eM}}

By definition, $e_M$ is the effective turnover of the metabolic macroreaction producing precursors from external substrates, obtained by dividing the reaction rate $v_M$ by the enzyme concentration $m$ (Eq.~\ref{eq:metaflux}).
The unit of $e_M$ is min$^{-1}$,  decomposed as follows:
\[
[e_M]  = \frac{\text{mass of metabolic product}}{\text{mass of enzyme M}\cdot \text{time}} = \frac{1}{\text{time}}.
\]
Note that $e_M = k_M\, s / (K_M + s)$ where $k_M$ is the turnover number of the reaction, indicating the maximum rate of conversion of external nutrients to precursor metabolites. 
$e_M$ will thus vary with the concentration $s$ of the external nutrients taken up by the cells and the identity of the nutrients. 
For example, the precursor mass that can be produced from 1 g of glucose is higher than that produced from 1 g of acetate.

How can we find a typical value for $k_M$, and thus for $e_M$ (both have the same order of magnitude if we suppose that the reaction is not operating far below saturation, that is, $e_M \approx K_M$)?
We stipulate that in our formalism the concentration of $p$, the concentration of precursors feeding into macromolecular synthesis, is close to the concentration of charged tRNA in the cell, since the latter are directly consumed by the most abundant part of the gene expression machinery, the ribosomes.
A reasonable estimate for $k_M$ could therefore be obtained from the turnover numbers of reactions involved in the synthesis of charged tRNA.
\cite{uter_longrange_2004} provides a typical value for one such reaction, catalyzed by glutaminyl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2 s^{-1}$, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second. 
After conversion to mass units \cite{freist_glutaminyltrna_1997}, this gives:
\[
k_{\textit{cat,GlnRS}} = \frac{3.2 \cdot 147}{64.4 \cdot 10^ 3} \approx 10^{-3} \, \text{g of glutaminyl-tRNA} \cdot \text{g of enzyme}^{-1} \cdot \text{s}^{-1}.
\]
We therefore take
\[
k_M \approx 3.6 \text{ h}^{-1} \geq e_M,
\]
and thus obtain an upper bound for $e_M$.

\subsubsection{\texorpdfstring{$k_R$}{kR}}

$k_R$ is the mass turnover number of the synthesis of macromolecules, in units h$^{-1}$.
As for $e_M$, we can decompose this into
\[
[k_R]  = \frac{\text{g of macromolecules}}{\text{g of gene expression machinery}\cdot \text{time}} = \frac{1}{\text{time}}
\]
We equate the mass of macromolecules the mass of proteins, the most abundant macromolecule in the cell [DennisBremer], and the mass of gene expression machinery with the mass of ribosomes (see above).
The dimensional analysis of $k_R$ thus becomes:
\[
\begin{aligned}
\left[k_R\right] &= \frac{\text{[protein mass produced]}}{\text{[ribosomal mass]} \cdot \text{[hour]}}\\
&= \frac{\text{[moles of protein]} \cdot \text{[protein molar mass]}}{\text{[moles of ribosome]} \cdot \text{[ribosome molar mass]} \cdot \text{[hour]}}\\
&= \frac{\text{[moles of amino acids]} \cdot \text{[molar mass of amino acids]}}{\text{[moles of ribosome]} \cdot \text{[hour]} \cdot \text{[ribosome molar mass]}}\\
&= \frac{\text{[maximal protein elongation rate]} \cdot \text{[molar mass of amino-acids]}}{\text{[ribosome molar mass]}}
\end{aligned}
\]
The values in the last equality are available from the literature\cite{ehrenberg_mediumdependent_2012,klumpp_molecular_2013} [weightaa,massribo].
We obtain:
\[
k_R \approx \frac{10 \cdot 100}{10^6}\cdot 3600 \approx 3.6 \text{ h}^{-1}.
\]
This value is comparable with the translational capacity $k_T$, in $\mu$g of protein per $\mu$g of ribosomal protein per hour, given by Scott \textit{et al.}\cite{scott_interdependence_2010}:
\[
k_T = \frac{4.5 \text{ }\mu\text{g of protein} \text{ / }\mu\text{g of RNA} \text{ / h }}{0.76 \text{ }\mu\text{g of ribosomal protein} \text{ / } \mu\text{g of RNA}} = 5.9 \text{ h}^{-1}
\]

\subsubsection{\texorpdfstring{$K_R$}{KR}}

While a value for the parameter $K_R$, representing the half-saturation constant of macromolecular synthesis, is more difficult to obtain from the literature, it should satisfy a straightforward constraint.
If the activity of the gene expression machinery is to significantly vary over time and between different environments, $p$ should neither be much smaller nor be much larger than $K_M$.
We satisfy this constraint by choosing $K_R$ to equal the total concentration of metabolites in the cell, as an order of magnitude\cite{bennett_absolute_2009}:
\[
K_R \approx 10\text{ g} \cdot \text{L}^{-1}
\]

\textbf{COMMENT: Why don't we choose total concentration of charged tRNAs here? Would be consistent with derivation of $k_R$...}

\subsubsection{\texorpdfstring{$\beta$}{Beta}}

$\beta$ is the inverse of cellular density of macromolecules, which has been shown constant during balanced growth over a large range of growth rates \cite{churchward_macromolecular_1982}, and there is some data suggesting that $\beta$ varies little during growth transitions as well [Zhou13].
From \cite{zimmerman_estimation_1991,mcguffee_diffusion_2010} we take the following typical value for $\beta$:
\[
\beta \approx \frac{1}{300} \approx 3\cdot 10^{-3} \text{ L} \cdot \text{g}^{-1}.
\]

\subsubsection{\texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}

From the values of the parameter in the dimensional model, one can deduce the parameters in the nondimensional model, used in the simulations:
\[
E_M = \frac{e_M}{k_R} = \frac{3.6}{3.6} = 1 \;\;\;\;\; ; \;\;\;\;\;  K = \beta\, K_R = 3\cdot 10^{-3} \cdot 10 = 0.03
\]

\subsection{Validation of parameters by fitting model to steady-state data set}

As a validation of the chosen orders of magnitude for the parameters, we fitted the model to the data on the RNA/protein mass ratio in \textit{Escherichia coli} as a function of the steady-state growth rate [ScottHwa].
These data were used to formulate so-called empirical growth laws.
As can be seen by comparing panels \textit{C} and \textit{D} of Figure~\ref{fig:isoclines} in the main text, growth-rate maximization in the self-replicator model leads to a good qualitative correspondence with the growth laws.
Can our model also quantitatively reproduce the growth laws, for parameter values close to those used in our simulations?

As explained in footnote~1 in the main text, the RNA/protein mass ratio can be interpreted as proportional to $\alpha$, with an unknown proportionality constant $\gamma$: $\alpha = \gamma \cdot \text{RNA/protein mass ratio}$.
We consider Eqs.~\ref{eq:pdef} and \ref{eq:rdef} in the main text at steady state (\textit{i.e.}, $dp/dt = dr/dt = 0$ ), and fit the model to the data using the differential evolution algorithm of Storn and Price [R133].
We first search for the value of $k_R$ that minimizes the squared difference between the data points and the model predictions when maximizing the growth rate for various values of $e_M$ (black dashed line in Figure~\ref{fig:isoclines}\textit{D}).
Second, we search for the six $e_M$ parameters (one per growth medium) that minimize the squared difference between the translational inhibition data for a given medium when maximizing the growth rate for different values of $k_R$ (colored dashed lines in Figure~\ref{fig:isoclines}\textit{D}).
Third, the correction parameter $\gamma$ is optimized on top of that so as to optimize both processes.
\textbf{COMMENT: the description of the last should be made more precise. Are you running an optimization process with all data? Fixing the values of $e_M$ and $k_R$ or using them as initial values only?}
The results of the fitting procedure are shown in Figure~\ref{fig:isoclines}\textit{D}, and discussed in the main text.
The parameter values are summarized in the table below (M63, cAA, and RDM refer to the medium compositions in [Scott10]).
As can be seen, they are in excellent agreement with the values used in the simulations, as determined from the literature in the previous subsection.

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Parameter & Unit & Literature value & Fitted value \\
\hline
$\gamma$ & $-$ & No value & 1.33\\
\hline
$k_R$ & h$^{-1}$ & 3.6 & 2.74\\
\hline
$e_M$ for M63+glycerol & h$^{-1}$ &  $< 3.6$ & 0.607\\
\hline
$e_M$ for M63+glucose & h$^{-1}$ &  $< 3.6$ & 0.912\\
\hline
$e_M$ for cAA+glycerol & h$^{-1}$ &  $< 3.6$ & 1.12\\
\hline
$e_M$ for cAA+glucose & h$^{-1}$ &  $< 3.6$ & 1.68\\
\hline
$e_M$ for RDM+glycerol & h$^{-1}$ &  $< 3.6$ & 3.48\\
\hline
$e_M$ for RDM+glucose & h$^{-1}$ &  $< 3.6$ & 5.06\\
\hline
$\beta K_R$ & $-$  & 0.03 & Not fitted\\
\hline
\end{tabular}
\end{center}

\section{Derivation of optimal control strategy}
\label{sec:supp_S3}

\subsection{Statement of the problem}

We consider the dimensionless system defined by Eqs~\ref{eq:pdefnondim} and \ref{eq:rdefnondim} in the main text, which are here repeated for clarity:
\begin{equation}{\label{eq:model}}
\begin{cases}
\frac{d\hat{p}}{d\hat{t}} = (1-\hat{r})\, E_M - (1+\hat{p})\, \hat{r} \, \frac{\hat{p}}{K+\hat{p}}, \\
\frac{d\hat{r}}{d\hat{t}} = \hat{r} \frac{\hat{p}}{K + \hat{p}} \, (\alpha (\hat{t}) - \hat{r}).
\end{cases}
\end{equation}

The objective of this study is to maximize the growth rate $\hat{\mu}= \hat{r}\, \hat{p}/(K+\hat{p})$ on an infinite time interval.
Consider the set of admissible controls
\[
\mathcal{U}=\{\alpha:\mathbb{R} \rightarrow [0,1] \ \mid \ \alpha(\cdot) \ \mathrm{measurable}\}.
\]
The optimization problem can then be stated as follows:
\begin{equation}\label{Prob}\tag{P}
\max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{+\infty} \hat{r}(\hat{t}) \, \dfrac{\hat{p}(\hat{t})}{K+\hat{p}(\hat{t})}\, d\hat{t},
\end{equation}
where $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))$ is the unique solution of Eq.~\ref{eq:model} starting at a given point $(\hat{p}_0,\hat{r}_0)\in \Omega \equiv \mathbb{R}^+_* \times (0,1)$ for a given control $\alpha\in \mathcal{U}$.
Note that since $J(\alpha)$ diverges, we actually consider \textit{overtaking optimality}: a trajectory is overtaking optimal if the performance index catches up to the performance
index of any other trajectory (see Definition 1.2(ii) in\cite{carlson_infinite_1991} for a rigorous mathematical definition).

\subsection{Maximum Principle}
Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle\cite{carlson_infinite_1991}.
Let $H=H(\hat{p},\hat{r},\lambda_p,\lambda_r,\lambda_0,\alpha)$ be the Hamiltonian of the system defined by:
\[
H \equiv \lambda_p\, E_M\, (1-\hat{r}) - \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}\left[\lambda_p (1+\hat{p}) +\lambda_r\, \hat{r} +\lambda_0\right] + \alpha \, \lambda_r \, \hat{r}\, \frac{\hat{p}}{K+\hat{p}}.
\]
Moreover, let $\alpha$ be an optimal control, and $\hat{x}(\cdot)=(\hat{p}(\cdot),\hat{r}(\cdot))$ the associated trajectory.
Then, there exists $\lambda_0 \leq 0$ and an absolutely continuous map $\lambda=(\lambda_p,\lambda_r):[0,+\infty) \rightarrow \mathbb{R}^2$
such that $(\lambda,\lambda_0)\neq0$, and
\begin{align}{\label{eq:adjoint}}
\dot{\lambda}_p&= \hat{r} \, \frac{K}{(K+\hat{p})^2}\, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, ( \hat{r}-\alpha) + \lambda_0\right] + \hat{r}\, \frac{\hat{p}}{K+\hat{p}}\, \lambda_p,\\
 \dot{\lambda}_r&= \lambda_p\, E_M +\frac{\hat{p}}{K+\hat{p}} \, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, (2 \hat{r}-\alpha) +\lambda_0\right].
\end{align}
The maximization condition is given by:
\begin{equation}{\label{eq:PMP}}
\begin{aligned}
&\alpha(\hat{t}) \in \mathrm{argmax}_{v \in [0,1]} H(\hat{x}(\hat{t}),\lambda(\hat{t}),\lambda_0,v), \\ &\mathrm{a.e.} \ \hat{t}\in [0,+\infty).
\end{aligned}
\end{equation}

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}. 
Next, we take $\lambda_0=-1$.\footnote{One can show that abnormal extremal trajectories (obtained for $\lambda_0=0)$ are not optimal. \textbf{A VERIFIER}}
From Eq.~\ref{eq:PMP}, we have that the control strategy is given by the sign of the \textit{switching function} $\phi \equiv \lambda_r \, \hat{r}\, \hat{p}/(K+\hat{p})$, that is, \textbf{COMMENT: IS THIS EVIDENT?}
\[
\begin{cases}
\alpha=1 \iff \phi>0,\\
\alpha=0 \iff \phi<0.
\end{cases}
\]

\subsection{Characterization of singular arcs}

Whenever $\phi$ is vanishing over a time interval, we say that the trajectory is \textit{singular}.
We will now characterize such trajectories.
If $I=[\hat{t}_1,\hat{t}_2]$ is a singular arc, we have $\phi(\hat{t})=\dot{\phi}(\hat{t})=0$, for all $\hat{t}\in[\hat{t}_1,\hat{t}_2]$, that is, $\lambda_r(\hat{t})=0$ and $\dot\lambda_r(\hat{t})=0$. Using additionally that $H$ is constant along an extremal trajectory, we obtain that $\lambda_p$ is constant along a singular arc. From Eq.~\ref{eq:adjoint}, this gives $\hat{p}(\hat{t})=\sqrt{E_M\, K}=\hat{p}_{opt}^*$. Using $d\hat{p}/d\hat{t}=0$, we finally get $\hat{r}(\hat{t})=\hat{r}_{opt}^*$.
Thus, the singular arc is the optimal steady state, corresponding to a singular control $\alpha(\hat{t})=\alpha_{opt}$, with $\alpha_{opt}$ depending on $E_M$ (Supplementary Text ???).

A necessary condition of optimality for a singular arc is given by the Kelley condition~\cite{borisov_fullers_2000}.
We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
$$
(-1)^q\frac{\partial}{\partial\alpha}\frac{d^{2q}}{d\hat{t}^{2q}}\phi(\hat{t})<0,
$$
satisfying the Kelley condition necessary for optimality. Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (which contains an infinite number of switching points) \cite{marchal_chattering_2013,borisov_fullers_2000}.

\subsection{Optimal trajectories}

From the Maximum Principle, we have shown that the optimal trajectory is a concatenation of bang arcs ($\alpha(t)=0$ or $\alpha(t)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$. Moreover, if the optimal trajectory has a singular arc, it must enter it through a chattering arc (\textit{i.e.}, with an infinite number of switches between $\alpha=0$ and $\alpha=1$).

These elements motivate the supposition that optimal solutions consist in a transient (chattering arc) towards the optimal steady state, after which they remain there (until a change to another environment). The chattering arc can be characterized by a switching curve $\hat{p}\mapsto\varphi(\hat{p})$ which passes through the optimal steady state.
Defining $A_0$ and $A_1$ the regions above and below $\varphi$ in the $(\hat{p},\hat{r})$-plane, respectively, we conjecture that the following feedback control law is optimal: 
%The simplest way to reach the optimal steady-state is thus two arcs, either 0-1 or 1-0 depending on the initial conditions. More precisely, let $\phi_0$ and $\phi_1$ be defined as the the trajectories solution of System \eqref{model} backward in time starting at $(p_{opt}^*,r_{opt}^*)$ with respectively $\alpha=0$ and $\alpha=1$. Given $\phi=\phi_0 \cup \phi_1$, we define $A_0$ and $A_1$ the regions above and below $\phi$ in the plane $(p,r)$.

\begin{equation}\label{control-opt}
\begin{cases}
\alpha(\hat{t})=0 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_0,\\
\alpha(\hat{t})=1 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_1,\\
\alpha(\hat{t})=\alpha_{opt} \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
\end{cases}
\end{equation}

It is possible to show from an analysis of the adjoint system that, after the first switch, a trajectory with two consecutive switches in the regions $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$ is not optimal. \textbf{COMMENT: IS THIS EVIDENT?} This is a first hint that the proposed control strategy is optimal. Moreover, our conjecture is also in line with the \textit{turnpike property}: \cite{trelat_turnpike_2015} have shown that, for a quite generic class of systems, the optimal strategy consists in staying on the optimal steady state (after a short transient).

Finally, we numerically solved the optimal control problem by the direct method using the \texttt{bocop} software\cite{bonnans_bocop_2012}. A time discretization allows the optimal control problem to be transformed into a nonlinear optimization problem,
solved here by interior point techniques.

A discretization by a Lobatto IIIC formula (6th order) was used with 4000 time steps, and the relative tolerance for the NLP solver was set at $10^{-14}$.
The optimal trajectories thus obtained are composed of a chattering arc followed by a steady state corresponding to the singular arc, as shown in Figure~\ref{fig:optimalcontrol} in the main text.
The switching curve $\varphi(\hat{p})$ was computed from numerical simulations with different initial conditions. \textbf{COMMENT: NUMERICAL OPTIMIZATION FOR DIFFERENT ENVIRONMENTS?}

It is important to stress that the optimization process was performed without any preliminary assumptions on the characteristics of the optimal trajectory. 
The fact that the numerical solution verifies the Maximum Principle (\textit{i.e.}, the singular arc corresponds to the optimal steady state) and the Kelley condition (\textit{i.e.}, the presence of a chattering arc)
tends to confirm that the control strategy of Eq.~\ref{control-opt} is optimal. As an aside, we note that due to the fact that numerical optimization was performed for a finite horizon, we actually obtained a second chattering arc escaping from the singular arc at the end of the simulation.  This is a classical property of the turnpike strategy: the optimal trajectory leaves the optimal steady state just before the end of the time interval of interest, in our case consuming almost all precursors. This arc was removed from the plot in Figure~\ref{fig:optimalcontrol}, because it does not occur with an infinite horizon and is therefore an artifact for the purpose of this study.

\section{Feedback control strategies based on the system variables}
\label{sec:supp_S4}

\subsection{Static feedback controls of one variable}

As described in the main text, one can describe the regulatory process of the microorganism by feedback control strategy, which sense variation in the system variables and adapt the resource allocation $\alpha$ accordingly.

First, we assume that the allocation of resource is directly regulated just by one variable, e.g. the nondimensional concentration of precursors $\hat{p}$.
In our settings, this corresponds to consider $\alpha(t)=h(\hat{p}(t))$, where $h:\mathbb{R}^+\rightarrow [0,1]$ is a function to be defined.
In control theory, this function is said to be a static function of $\hat{p}$ (as opposed to, for instance, functions that could depend on the derivative or the integral of $\hat{p}$).
Within this class of functions, we know from Eq.~\ref{eq:sup_alpha_mu_optimal_p} that $g(p)$ is the only control law that guarantees optimal steady-state:
\begin{equation}
\label{eq:supp_g}
g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
\end{equation}
We prove this statement by contradiction. Assume a control law $h(\hat{p})$ different from $g(\hat{p})$, i.e. $\exists \tilde p \mid h(\tilde p)\neq g(\tilde p)$. At steady-state, $\dot r=0$ gives $r=h(\tilde p)$. Thus, we obtain with this control law a steady-state $(\tilde p,h(\tilde p))\neq(\tilde p,\gamma(\tilde p))$, i.e. which is not optimal, in contradiction with our statement.

Similarly, using Eq.~\ref{eq:sup_alpha_optimal} we can show that $\alpha(t)=f(E_M(t))$ such that:
\begin{equation}
\label{eq:supp_f}
f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{KE_M} + 1}
\end{equation}
is the only control law using the environmental parameter $E_M$ that guarantees optimal steady-state.
Figure~\ref{fig:regfunction} represents these function along biologically plausible approximation (hill-type functions).
Note that in both cases, these control laws do not fulfill the conditions given by the Maximum Principle (whenever initial conditions are different from the optimal steady-state).
Thus it is trivial that these control laws are not optimal.
How far from the optimal these strategies performs is represented in Figure~\ref{fig:strategies}.

\subsection{Static feedback controls of several variables: example of internal variables, \texorpdfstring{$\hat{p}$}{p} and \texorpdfstring{$\hat{r}$}{r}}

### 2016-06-01 theory.tex
First meeting with H&H

\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem}

\section*{Summary of chapter \thechapter}
In this section, we present the topic of microorganism growth, in particular the recent focus which is made on resource allocation, and discuss some of the limitations of those studies.
We describe the self-replicator approach, and how we're going to apply it in dynamical conditions.

\selectlanguage{french}
\section*{Résumé du chapître \thechapter : [titre français]}
dans cette section blablabla
\selectlanguage{english}

### 2016-06-06 Chapters/theory.tex
Refactoring paragraph 3 and changing a lot of file names

--- 2016-06-01 Chapters/core.tex
@@
-\section*{Introduction}
+\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies }
+\chaptermark{Dynamical allocation of cell resources}
+
+\selectlanguage{french}
+\section*{Résumé du chapître \thechapter : [titre français]}
+dans cette section blablabla
+\selectlanguage{english}
+
+{\Large
+\textbf\newline{\textbf{}}
+}
+
+\section{Abstract}
+Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
+Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
+While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
+The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
+We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
+We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
+We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
+A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
+Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
+It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
+Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
+
+%%%%%% CORE OF THE ARTICLE %%%%%
+\section{Introduction}
 
@@
 
-\section*{Results}
-
-\subsection*{Self-replicator model of resource allocation}
+\section{Results}
+
+\section{Self-replicator model of resource allocation}
 \label{sec:model}
@@
 \begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig1.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig1.eps}
 \caption[Self-replicator model of bacterial growth]{
@@
 
-\subsection*{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
+\section{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
 \label{sec:growthlaws}
@@
 \begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig2.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
 \caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
@@
 \begin{figure}[!h]
-%\includegraphics[scale=1]{./Fig/Fig3.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
 \caption[Self-replicator model accounts for bacterial growth laws]{
@@
 
-\subsection*{Biomass maximization as an optimal control problem}
+\section{Biomass maximization as an optimal control problem}
 \label{sec:optimalcontrolproblem}
@@
 
-\subsection*{Solution of the optimal control problem}
+\section{Solution of the optimal control problem}
 \label{sec:optimalcontrolsolution}
@@
 \centering
-%\includegraphics[scale=1]{./Fig/Fig4.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
 \caption[Optimal control of the self-replicator during a nutrient upshift]{
@@
 
-\subsection*{Simple feedback control strategies: exploiting information on nutrients or precursors}
+\section{Simple feedback control strategies: exploiting information on nutrients or precursors}
 \label{sec:steadystate}
@@
 \centering
-%\includegraphics[scale=1]{./Fig/Fig5.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
 \caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
@@
 \centering
-%\includegraphics[scale=1]{./Fig/Fig6.eps}
+\includegraphics[height=5cm]{./Fig/Fig6.eps}
 \caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
@@
 
-\subsection*{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
+\section{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
 \label{sec:strategies}
@@
 \centering
-%\includegraphics[scale=1]{./Fig/Fig7.eps}
+\includegraphics[height=5cm]{./Fig/Fig7.eps}
 \caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
@@
 \centering
-%\includegraphics[scale=1]{./Fig/Fig8.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig8.eps}
 \caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
@@
 
-\section*{Discussion}
+\section{Discussion}
 \label{sec:discussion}
@@
 When extended with heterologous metabolic pathways, the self-replicator models used in this study would provide an adequate \textit{in-silico} test bed for the rapid screening and comparison of alternative control strategies in bioprocess engineering.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Methods}
+
+\section{Steady-state analysis of model}
+\label{sec:methods_model_analysis}
+
+The nondimensional version of the model, given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, was used for a steady-state analysis of the self-replicator.
+Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} were derived from the original model of Eqs~\ref{eq:pdef}-\ref{eq:rdef} by means of the following rescalings:
+\begin{equation*}
+\hat{p}  = \beta \, p,\;\;\;
+\hat{r}  = \beta \, r,\;\;\;
+\hat{t}  = k_R \, t,\;\;\;
+E_M = e_M/k_R,\;\;\;
+K = \beta K_R.
+\end{equation*}
+
+As shown in \ref{S1_Text}, for a constant environment $E_M$ and constant resource allocation $\alpha$, the system has two steady states: a trivial unstable steady state $(\hat{p}^*, \hat{r}^*) = (0,1)$, allowing no growth in the absence of precursors, and a steady state with a positive growth rate given by
+\begin{equation}
+\label{eq:meth_steadystate}
+(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
+\end{equation}
+The two eigenvalues of the Jacobian matrix evaluated at $(\hat{p}^*, \hat{r}^*)$ are negative (\ref{S1_Text}), so that this steady state is stable.
+
+The growth rate at steady state, as a function of $\hat{p}^*$ and $\hat{r}^*$, is given by Eq.~\ref{eq:munondim}, which we repeat here for clarity:
+\begin{equation*}
+\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
+\end{equation*}
+Evaluating $d\hat{p}/dt=0$ at $(\hat{p}^*, \hat{r}^*)$ allows $\hat{r}^*$, and therefore $\hat{\mu}^*$, to be written as a function of $\hat{p}^*$ (\ref{S1_Text}).
+Accordingly, we can compute $\partial\hat{\mu}^*/\partial\hat{p}^*$ and, when setting this partial derivative to 0, determine the maximum growth rate at steady state $\mu_{opt}^*$ and the optimal resource allocation $\alpha_{opt}^*$ bringing about this maximal growth rate.
+As shown in \ref{S1_Text}, $\mu_{opt}^*$ and $\alpha_{opt}^*$ can be written as explicit functions of either the environment $E_M$:
+\begin{equation}
+\label{eq:alpha_mu_optimal}
+\alpha_{opt}^* = \frac{E_M + \sqrt{K\, E_M}}{E_M + 2\sqrt{K\, E_M} + 1}
+, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
+\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1},
+\end{equation}
+or the precursor abundance $\hat{p}_{opt}^*$:
+\begin{equation}
+\label{eq:alpha_mu_optimal_p}
+\alpha_{opt}^* = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
+, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
+\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
+\end{equation}
+The above equations were used for the derivation of the control strategies (see below).
+
+\section{Model fitting}
+\label{sec:methods_model_fitting}
+
+As can be seen by comparing Figs~\ref{fig:model_validation}\textit{A} and~\ref{fig:model_validation}\textit{B}, growth-rate maximization in the self-replicator model leads to a good qualitative correspondence with the growth laws.
+In order to determine if a good quantitative fit of the model with the data from Scott~\textit{et al.} \cite{scott_interdependence_2010} can be obtained, for reasonable parameter values, we estimated $e_M$ and $k_R$ in Eqs~\ref{eq:pdef}-\ref{eq:rdef} from the measured RNA/protein mass ratios.
+At steady state, the RNA/protein mass ratio can be interpreted as proportional to $\hat{r}^*$ (and thus $\alpha_{opt}^*$), with an unknown (dimensionless) proportionality constant $\gamma$ (see~\cite{scott_interdependence_2010} for details on the use of the RNA/protein mass ratio as a proxy for the ribosomal protein mass fraction):
+\begin{equation}
+\hat{r}^* = \alpha_{opt}^* = \gamma \, \frac{\text{RNA mass}}{\text{protein mass}}.
+\label{eq:gamma}
+\end{equation}
+Reformulating Eq.~\ref{eq:alpha_mu_optimal} in terms of the original parameters $e_M$ and $k_R$, which have physical dimensions facilitating the biological interpretation of their values, we obtain a straighforward relation between $e_M$, $k_R$, $K$, $\alpha_{opt}^*$ and $\mu_{opt}^*$:
+\begin{equation}
+\label{eq:alpha_mu_optimal_dim}
+\alpha_{opt}^* = \frac{e_M + \sqrt{K\, e_M\, k_R}}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}
+, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
+\mu^*_{opt} = \frac{e_M \, k_R}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}.
+\end{equation}
+
+Eqs~\ref{eq:gamma}-\ref{eq:alpha_mu_optimal_dim} were used to estimate values of $k_R$ and $\gamma$, as well as $e_M$ for each of the six growth conditions, from the measurements of the growth rate and the RNA/protein mass ratio.
+The value $K$ was not estimated from the experimental data, but set to a value inferred from the literature (\ref{S1_Text}).
+The optimization process was carried out by means of the differential evolution algorithm of Storn and Price~\cite{storn_differential_1997}.
+The results are shown in Fig.~\ref{fig:model_validation}\textit{B}, while the estimated parameter values are summarized in \ref{S1_Table}.
+The parameter values are in very good agreement with order-of-magnitude values determined from the literature (\ref{S2_Text} and \ref{S1_Table}).
+
+\section{Solution of optimal control problem}
+
+The optimal control problem of Eq.~\ref{eq:J} consists in identifying the function $\alpha_{opt}(t)$ that maximizes the integral of the growth rate $\hat{\mu}$ over an interval $[0, \tau]$.
+In order to solve this problem, we first redefined it over an infinite horizon (\textit{i.e.}, $\tau \rightarrow \infty $) in order to avoid boundary effects occurring over finite time intervals, in particular the depletion of precursors just before reaching $\tau$.
+With $\mathcal{U}=\{\alpha:\mathbb{R}^+ \rightarrow [0,1] \}$ the set of admissible controls, the full optimization problem for the nondimensionalized system is given by
+\begin{equation}\label{eq:meth_prob}
+\max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{\infty} \hat{r}(\hat{t}) \, \dfrac{\hat{p}(\hat{t})}{K+\hat{p}(\hat{t})}\, d\hat{t}.
+\end{equation}
+Since $J(\alpha)$ diverges, we actually consider overtaking optimality:
+A solution is overtaking optimal if its performance index catches up with the performance index of any other solution (\cite{carlson_infinite_1991}, see \ref{S3_Text} for details).
+
+Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle \cite{carlson_infinite_1991}, an extension of the well-known Pontryagin Maximum Principle.
+Analysis of the Hamiltonian of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and the associated adjoint system shows that the optimal trajectory is a concatenation of bang arcs ($\alpha(\cdot)=0$ or $\alpha(\cdot)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the new environment after the upshift (\ref{S3_Text}).
+Moreover, from the Kelley condition~\cite{borisov_fullers_2000}, we can show that if the optimal trajectory has a singular arc, then it must enter this singular arc \textit{via} a chattering arc, \textit{i.e.}, with an infinite number of switches of $\alpha (\cdot)$ between 0 and 1 (\ref{S3_Text}).
+The chattering arc is characterized by a switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
+The switching curve divides the phase plane into two regions, such that $\alpha(t)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (\ref{S3_Text} and Fig.~\ref{fig:optimalcontrol}).
+
+The above results have led to the conjectured optimal solution of Eq.~\ref{eq:optcontrol}.
+In parallel, we numerically solved the problem of Eq.~\ref{eq:meth_prob} by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012}.
+A time discretization allows the optimal control problem to be transformed into a nonlinear optimization problem,
+solved here by interior point techniques.
+A discretization by a Lobatto IIIC formula (6th order) was used with 4000 time steps, and the relative tolerance for the NLP solver was set to $10^{-14}$.
+The optimal trajectories thus obtained are composed of a chattering arc followed by a steady state corresponding to the singular arc (Fig.~\ref{fig:optimalcontrol}).
+The switching curve $\varphi(\hat{p})$ was computed from numerical simulations with different initial conditions.
+
+\section{Specification and analysis of control strategies}
+
+As described in the \textit{Results} section, we are interested in control strategies satisfying the following conditions:
+\begin{description}
+\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
+\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
+\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
+\end{description}
+It can be directly verified from the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies (Eqs~\ref{eq:f}, \ref{eq:precursorstrategy}, and \ref{eq:stratswitch}) that they are indeed static functions of the system variables (or the system input, in the case of the nutrient-only strategy).
+Here we show that the other two conditions are also satisfied for all three strategies.
+
+Following Eq.~\ref{eq:f}, the nutrient-only strategy is defined by $\alpha = f(E_M)$, so that $\alpha$ is constant after the upshift.
+As shown above and in \ref{S1_Text}, this means that the system controlled by the nutrient-only strategy has a single nontrivial stable steady state (Condition~C2). 
+In addition, in this case the optimal steady state is attained for $\alpha^*_{opt}$ defined as in Eq.~\ref{eq:alpha_mu_optimal}, and the following function $f$ therefore guarantees Condition~C3:
+\begin{equation}
+\label{eq:meth_f}
+f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K\, E_M} + 1}.
+\end{equation}
+In \ref{S1_Text}, it is shown that Eq.~\ref{eq:meth_f} is the only definition of $f$ satisfying all conditions.
+\ref{S1_Fig} shows a plot of $f(E_M)$ together with a biologically plausible Michaelis-Menten approximation (Eq.~\ref{eq:MMApprox}).
+
+The full specification of the precursor-only strategy demands an expression for the function $g$ in Eq.~\ref{eq:precursorstrategy}.
+Recall that Eq.~\ref{eq:alpha_mu_optimal_p} defines $\alpha_{opt}^*$ in terms of the precursor concentration $\hat{p}_{opt}^*$, which leads us to propose the following function $g$:
+\begin{equation}
+\label{eq:meth_g}
+g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
+\end{equation}
+As shown in \ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
+Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
+This can be seen by noting that at steady state, $d\hat{r}/dt=0$ implies $\hat{r}^*=g(\hat{p}^*)$ (Eq.~\ref{eq:rdefnondim}).
+In order for the self-replicator to attain a maximal growth rate at steady rate, Eq.~\ref{eq:alpha_mu_optimal_p} needs to be satified, which is the case for the above choice of the function $g$.
+Like for $f$, Eq.~\ref{eq:meth_g} is the only choice for $g$ satisfying C1-C3.
+\ref{S1_Fig} shows a plot of $g(\hat{p})$ together with a biologically plausible Hill approximation (Eq.~\ref{eq:HillApprox}).
+
+The on-off control strategy is defined in Eq.~\ref{eq:stratswitch} and repeated below:
+\begin{equation}
+h(\hat{p}, \hat{r}) = 
+\begin{cases}
+0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
+1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
+\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
+\end{cases}
+\end{equation}
+This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in Fig.~\ref{fig:onoffresults}\textit{A}.
+In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
+Moreover, the steady state coincides with the optimal steady state $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$ by construction, so that Condition~C3 is satisfied as well.
+Fig~\ref{fig:ppGppsurface}\textit{A} shows a plot of $h(\hat{p}, \hat{r})$.
+
+Note that since $h(\cdot)$ is discontinuous, numerical instabilities occur during simulations.
+We therefore used the following continuous approximation of this function:
+\begin{equation}
+\frac{g(\hat{p})^{100}}{g(\hat{p})^{100} + \hat{r}^{100}}, \ \textrm{if} \ \hat{r} \neq g(\hat{p}).
+\end{equation}
+The approximation causes $\alpha$ to take intermediate values (instead of 0 or 1) just before reaching the optimal steady state in Fig~\ref{fig:onoffresults}\textit{C}.
+For numerical simulations of the ODE system, we used the CVODE solver~\cite{cohen_cvode_1996} from SUNDIALS 2.6.2~\cite{hindmarsh_sundials_2005}.
+
+\clearpage
+
+%%%%%% FOR MANUAL REF%%%%%
+\makeatletter
+\newcommand{\manuallabel}[2]{\def\@currentlabel{#2}\label{#1}}
+\makeatother
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{S1 Text -- Model derivation and analysis}
+\manuallabel{S1_Text}{S1~Text}
+
+\subsubsection{Model formulation}
+
+The time evolution of the total mass of each component of the self-replicator can be written as follows:
+\begin{eqnarray}
+\frac{dP}{dt} &=& V_M(t) - V_R(t) \nonumber , \\
+\frac{dM}{dt} &=& (1-\alpha(t))\, V_R(t) \label{eq:ext_system},\\
+\frac{dR}{dt} &=& \alpha(t) \, V_R(t)  \nonumber ,
+\end{eqnarray}
+where $P$, $M$, $R$ [g] denote the total mass of precursors, metabolic machinery and gene expression machinery, respectively.
+$V_M$ [g h$^{-1}$] is the rate of production of precursors by metabolism and $V_R$ [g h$^{-1}$] the rate of utilisation of precursors for gene expression.
+
+Dividing the mass variables by the total time-varying volume $\texttt{Vol}(t)$ of the system, we obtain the concentration variables $p = P/\texttt{Vol}$, $m = M/\texttt{Vol}$, $r = R/\texttt{Vol}$ [g L$^{-1}$].
+The dynamics of the concentration variables then follows with Eq.~\ref{eq:ext_system}:
+\begin{eqnarray}
+\frac{dp}{dt} &=& \frac{V_M(t)}{\texttt{Vol}} - \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt}\, p, \nonumber \\
+\frac{dm}{dt} &=& (1-\alpha(t)) \frac{V_R(t)}{\texttt{Vol}} - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, m  \label{eq:deriv_int_system},\\ 
+\frac{dr}{dt} &=& \alpha(t) \, \frac{V_R(t)}{\texttt{Vol}}  - \frac{1}{\texttt{Vol}}\frac{d\texttt{Vol}}{dt} \, r. \nonumber
+\end{eqnarray}
+
+At this point, we define $v_M = V_M/\texttt{Vol}$ and $v_R = V_R/\texttt{Vol}$ [g L$^{-1}$ h$^{-1}$] as the mass fluxes per unit volume.
+Moreover, with the definition of the volume in terms of the total protein mass in Eq.~\ref{eq:voldef} of the main text, that is, $\texttt{Vol} = \beta\, (M + R)$, we find that
+\begin{equation}
+\label{eq:supp_deriv_growthrate}
+\frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \frac{\beta}{\texttt{Vol}} \frac{d(M+R)}{dt} = \beta\, \frac{V_R(t)}{\texttt{Vol}} = \beta \, v_R(t).
+\end{equation}
+This leads to the system
+\begin{eqnarray}
+\frac{dp}{dt} &=& v_M(t) - v_R(t) \, (1+\beta\, p), \label{eq:supp_pdef}\\
+\frac{dr}{dt} &=& v_R(t)  \, (\alpha(t) - \beta\, r) \label{eq:supp_rdef},
+\end{eqnarray}
+where the equation for $m(t)$ is omitted since by construction $r(t) + m(t) = 1/\beta$ and $dr/dt + dm/dt = 0$.
+
+As stated in the main text, we use Michaelis-Menten kinetics to express $v_M$ and $v_R$ in terms of the system variables:
+\begin{eqnarray}
+v_M(t) &=& m(t) \, k_M \, \frac{s(t)}{K_M +s(t)} = \left(\frac{1}{\beta} - r(t)\right)\, e_M(t) \nonumber, \\
+v_R(t) &=& r(t) \, k_R \, \frac{p(t)}{K_R +p(t)}, \nonumber 
+\end{eqnarray}
+with rate constants $k_M$, $k_R$ [h$^{-1}$] and half-saturation constants $K_M$, $K_R$ [g L$^{-1}$].
+$s(t)$ is an exogenous variable representing the nutrient concentration in the external medium.
+We simplify $v_M(t)$ by defining the environmental input $e_M(t) = k_M \, s(t) / (K_M + s(t))$.
+Throughout the paper, as explained in the main text, we assume the environment is constant, \textit{i.e.}, $e_M(t)=e_M$.
+
+Finally, the growth rate $\mu$ [h$^{-1}$] is defined as the relative increase of the volume of the self-replicator.
+From Eq.~\ref{eq:supp_deriv_growthrate}, it follows that:
+\begin{equation}
+\label{eq:supp_growthrate}
+\mu (t) = \frac{1}{\texttt{Vol}} \frac{d\texttt{Vol}}{dt} = \beta\, v_R(t).
+\end{equation}
+
+\subsubsection{Nondimensionalization of the system}
+
+For the sake of simplifying the proofs and derivations below, we define the following nondimensional variables:
+\begin{equation*}
+\hat{p}  = \beta \, p,\;\;\;
+\hat{r}  = \beta \, r,\;\;\;
+\hat{t}  = k_R \, t.
+\end{equation*}
+When injecting these into Eq.~\ref{eq:supp_pdef}, we obtain
+\[
+\frac{k_R}{\beta} \, \frac{d\hat{p}}{d\hat{t}} = \left( \frac{1}{\beta} - \frac{\hat{r}}{\beta} \right) \, e_M - \frac{\hat{r}}{\beta} \, k_R \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}),
+\]
+which simplifies to
+\[
+\frac{d\hat{p}}{d\hat{t}} = ( 1 - \hat{r} ) \, \frac{e_M}{k_R} - \hat{r} \, \frac{\hat{p}}{\beta K_R + \hat{p}} \, (1 + \hat{p}).
+\]
+In a similar manner, we derive the time evolution of the nondimensional $\hat{r}$, and thus obtain the system
+\begin{equation}
+\label{eq:supp_adim}
+\begin{aligned}
+\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
+\frac{d\hat{r}}{d\hat{t}} &= (\alpha - \hat{r}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},
+\end{aligned}
+\end{equation}
+with the lumped parameters $E_M = e_M / k_R$ and $K = \beta \, K_R$.
+The corresponding nondimensionalized growth rate is given by
+\begin{equation}
+\label{eq:supp_growthrate_adim}
+\hat{\mu} = \frac{\mu}{k_R} = \frac{\hat{p}}{K+\hat{p}} \, \hat{r}.
+\end{equation}
+
+\subsubsection{Steady-state growth of the self-replicator}
+\label{si::growthrate}
+
+If we suppose $E_M > 0$, $K > 0$ and $\alpha \in ]0, 1[$, there is a trivial unstable steady state at $(0, 1)$.
+A second steady-state exists for the point in which $\hat{r}^* = \alpha$ and $\hat{p}^*$ is a root of the following polynomial:
+\[
+\alpha \, \hat{p}^2 + \left(\alpha - (1-\alpha) \, E_M \right) \hat{p} - (1-\alpha)\, E_M\, K.
+\]
+If we keep the only admissible root for this polynomial (\textit{i.e.}, for which $\hat{p} \geq 0$), the second steady state is given by
+\begin{equation}
+\label{eq:supp_steadystate}
+(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
+\end{equation}
+We can determine the stability of this steady state by looking at the Jacobian matrix $J$ of the ODE system:
+\begin{equation}
+\label{eq:supp_jacop}
+J = \left(\begin{matrix}
+- \frac{\hat{r}}{K + \hat{p}} \left[ \hat{p} + (1+\hat{p})\, \frac{K}{K+\hat{p}}\right] & - E_M - (1+\hat{p})\frac{\hat{p}}{K+\hat{p}}\\
+(\alpha - \hat{r})\, \hat{r} \, \frac{K}{(K+\hat{p})^2} & (\alpha - 2\hat{r})\, \frac{\hat{p}}{K+\hat{p}}
+\end{matrix}\right) .
+\end{equation}
+Evaluated at the point $(\hat{p}^*, \hat{r}^*)$, the Jacobian matrix becomes
+\[
+J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
+- \frac{\alpha}{K + \hat{p}^*} \left[ \hat{p}^* + (1+\hat{p}^*)\frac{K}{K+\hat{p}^*}\right] & - E_M - (1+\hat{p}^*)\frac{\hat{p}^*}{K+\hat{p}^*}\\
+0 & -\alpha\frac{\hat{p}^*}{K+\hat{p}^*}
+\end{matrix}\right).
+\]
+Since $\hat{p}^*$, $\alpha$, $E_M$, $K$ $>0$, the two eigenvalues are negative and therefore the steady state $(\hat{p}^*, \hat{r}^*)$ is stable (see also the streamlines in Figure~\ref{fig:model_analysis}\textit{A} in the main text).
+It means that for fixed environmental conditions $E_M$ and resource allocation $\alpha$, the self-replicator converges towards a steady state in which the concentration variables are constant.
+
+One can now easily derive the steady-state growth rate, denoted $\hat{\mu}^*$.
+By substituting Eq.~\ref{eq:supp_growthrate_adim} into the first ODE of the system of Eq.~\ref{eq:supp_adim}, we find at steady state:
+\[
+\left(\frac{d\hat{p}}{d\hat{t}}\right)_{(\hat{p}^*, \hat{r}^*)} = 0 = (1-\alpha)\, E_M - (1+\hat{p}^*)\, \hat{\mu}^*,
+\]
+which by means of Eq.~\ref{eq:supp_steadystate} gives the following relation:
+\begin{equation}
+\label{eq:supp_growthrate_p}
+\hat{\mu}^* = \frac{(1-\alpha)\, E_M}{1+\hat{p}^*} = \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
+\end{equation}
+Finally, we can transform this expression to obtain
+\begin{equation}
+\label{eq:supp_growthrate_final}
+\hat{\mu}^* = \begin{cases}
+\frac{(1-\alpha)\, E_M + \alpha - \sqrt{\left[(1-\alpha)\, E_M - \alpha\right]^2 + 4(1-\alpha)\, \alpha\, E_M\, K}}{2(1-K)} &\text{ for }K\neq 1,\\
+\frac{\alpha\, (1-\alpha)\, E_M}{\alpha + (1-\alpha)\, E_M} &\text{ for }K = 1.
+\end{cases}
+\end{equation}
+This function of $\alpha$ is plotted in Figure~\ref{fig:model_analysis}\textit{B} in the main text.
+
+\subsubsection{Maximization of growth rate at steady state}
+\label{si::optimal}
+
+We are interested in the steady state at which growth occurs at the maximum rate.
+The growth rate at steady state $\hat{\mu}^*$ is given by
+\begin{equation}
+\label{eq:supp_growthrate_steadystate}
+\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
+\end{equation}
+From the first ODE of the system of Eq.~\ref{eq:supp_adim}, we have
+\begin{equation}
+\label{eq:supp_isop}
+\hat{r}^* = \frac{E_M}{E_M + \frac{\hat{p}^*}{K + \hat{p}^*} (1+\hat{p}^*)} .
+\end{equation}
+Substituting Eq.~\ref{eq:supp_isop} into Eq.~\ref{eq:supp_growthrate_steadystate}, we obtain
+\begin{equation}
+\label{eq:supp_growthrate_steadystate_p}
+\hat{\mu}^* = \frac{E_M \, \hat{p}^*}{\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K} .
+\end{equation}
+The value of $\hat{p}^*$ maximizing $\hat{\mu}^*$ can be determined from
+\begin{equation}
+\label{eq:supp_growthrate_steadystate_deriv_p}
+\frac{\partial\hat{\mu}^*}{\partial\hat{p}^*} = \frac{E_M \, (E_M\, K - \hat{p}^{*2})}{\left(\hat{p}^{* 2} + (E_M + 1)\, \hat{p}^* + E_M\, K\right)^2},
+\end{equation}
+by looking at the values of $\hat{p}^*$ for which this derivative equals 0.
+It follows that $\hat{\mu}^*$ is maximal for
+\begin{equation}
+\label{eq:supp_p_optimal}
+\hat{p}^* = \hat{p}^*_{opt} = \sqrt{K\, E_M}.
+\end{equation}
+By substituting $\hat{p}^*_{opt}$ and $\alpha_{opt}^*$ for $\hat{p}^*$ and $\hat{r}^*$, respectively, in Eq.~\ref{eq:supp_isop}, we obtain the resource allocation maximizing the growth rate
+\begin{equation}
+\label{eq:supp_alpha_optimal}
+\alpha_{opt}^* = \frac{E_M + \sqrt{KE_M}}{E_M + 2\sqrt{KE_M} + 1}.
+\end{equation}
+Finally, injecting this result into Eq.~\ref{eq:supp_growthrate_steadystate} we obtain the optimal steady-state growth rate:
+\begin{equation}
+\label{eq:supp_growthrate_optimal}
+\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1}.
+\end{equation}
+In addition, by using Eq. \ref{eq:supp_p_optimal}, we can write $\alpha_{opt}^*$ and $\hat{\mu}_{opt}^*$ as a function of $\hat{p}_{opt}^*$ only:
+\begin{equation}
+\label{eq:supp_alpha_mu_optimal_p}
+\alpha_{opt}^* = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
+, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
+\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
+\end{equation}
+
+\subsubsection{Analysis of the control strategies}
+\label{si::control_strategies}
+
+In this section, we derive the main results for the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies.
+For each of these, we prove that the Conditions C1, C2 and C3 from the \textit{Methods} section are satisfied, which we repeat here for clarity:
+\begin{description}
+\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
+\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
+\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
+\end{description}
+
+\subsubsection{Nutrient-only strategy}
+
+The nutrient-only strategy is defined by:
+\begin{equation}
+\label{eq:supp_f}
+\alpha = f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K\, E_M} + 1}.
+\end{equation}
+It drives the system to the optimal steady state by measuring the environment $E_M$.
+Note that Condition C1 is satisfied by definition.
+
+By injecting Eq.~\ref{eq:supp_f} into Eq.~\ref{eq:supp_adim}, the ODE system under the control of $f$ becomes:
+\begin{equation}
+\label{eq:supp_adim_f}
+\begin{aligned}
+\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
+\frac{d\hat{r}}{d\hat{t}} &= \left(f(E_M) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
+\end{aligned}
+\end{equation}
+Since $E_M$ is constant on the interval of interest (starting right after the upshift), we are in the case of Section~\ref{si::growthrate} (\textit{i.e.}, $\alpha$ constant).
+In particular, the system has two steady states: a trivial unstable one at (0, 1) (with zero growth), and a stable one defined by Eq.~\ref{eq:supp_steadystate} (Condition C2).
+Since $f(E_M) = \alpha_{opt}^*$, we conclude from the derivations in Section~\ref{si::optimal} that the stable steady state is optimal for every environment $E_M$ (Condition C3).
+
+It is interesting to note that the expression in Eq.~\ref{eq:supp_f} is the only function $f(E_M)$ satisfying C1-C3.
+We can prove this statement by contradiction.
+Assume a control strategy $c(E_M)$ satisfying C1-C3, and different from $f(E_M)$, \textit{i.e.}, there exists $E_M={E_M}_1$ such that $c({E_M}_1) \neq f({E_M}_1)$.
+In this environment, the system reaches a steady state $(\hat{p}_1^*, \hat{r}_1^*)$ with $\hat{r}_1^* = c({E_M}_1) \neq f({E_M}_1)$.
+However, by Eq.~\ref{eq:supp_alpha_optimal} the optimal value for $\hat{r}^*$ in this environment is given by $f({E_M}_1)$.
+So, the control law $c(E_M)$ does not drive the system to the optimal steady state in this environment, in contradiction with Condition~C3.
+
+\subsubsection{Precursor-only strategy}
+
+The precursor-only strategy is defined by:
+\begin{equation}
+\label{eq:supp_g}
+\alpha = g(\hat{p}) = \frac{\hat{p}}{\hat{p} + \frac{K}{K + \hat{p}}(1+\hat{p})}.
+\end{equation}
+Here as well, C1 is satisfied by construction.
+
+The ODE system under the control of $g$ becomes
+\begin{equation}
+\label{eq:supp_adim_g}
+\begin{aligned}
+\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
+\frac{d\hat{r}}{d\hat{t}} &= \left(g(\hat{p}) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
+\end{aligned}
+\end{equation}
+The nullcline for $\hat{p}$ remains unchanged and is defined by
+\begin{equation}
+\label{eq:supp_pnullcline_g}
+\frac{d\hat{p}}{dt} = 0 \Leftrightarrow 
+\hat{r} = \frac{E_M}{E_M + \frac{\hat{p}}{K+\hat{p}} (1+\hat{p})},
+\end{equation}
+while the nullcline for $\hat{r}$ is
+\begin{equation}
+\label{eq:supp_rnullcline_g}
+\frac{d\hat{r}}{dt} = 0 \Leftrightarrow \begin{cases}
+\hat{p} = 0,\\
+\hat{r} = 0,\\
+\hat{r} = \frac{\hat{p}}{\hat{p} + \frac{K}{K+\hat{p}} (1+\hat{p})}.
+\end{cases}
+\end{equation}
+Hence, we also have a trivial unstable steady state at (0,1) (with zero growth).
+The second steady state is obtained from Eqs~\ref{eq:supp_pnullcline_g}-\ref{eq:supp_rnullcline_g}:
+\begin{equation*}
+\frac{E_M}{E_M + \frac{\hat{p}^*}{K+\hat{p}^*} (1+\hat{p}^*)} = \frac{\hat{p}^*}{\hat{p}^* + \frac{K}{K+\hat{p}^*} (1+\hat{p}^*)},
+\end{equation*}
+which we rearrange into
+\begin{equation*}
+\hat{p}^* E_M + \frac{K}{K+\hat{p}^*} (1+\hat{p}^*) E_M = \hat{p}^* E_M + \frac{\hat{p}^*}{K+\hat{p}^*} (1+\hat{p}^*) \hat{p}^*.
+\end{equation*}
+This leads to
+\begin{equation*}
+\hat{p}^* = \sqrt{K E_M},
+\end{equation*}
+and therefore
+\begin{equation*}
+\hat{r}^* = g(\hat{p}^*) = \frac{\sqrt{K E_M}}{\sqrt{K E_M} + \frac{K}{K + \sqrt{K E_M}} (1+\sqrt{K E_M})} = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K E_M} + 1}.
+\end{equation*}
+From Eqs~\ref{eq:supp_p_optimal}-\ref{eq:supp_alpha_optimal}, we recognize the optimal steady state for the environment $E_M$, validating Condition~C3.
+We now look for the stability of this (optimal) steady state by deriving the Jacobian of this system:
+\begin{equation}
+\label{eq:supp_jacobian_g}
+J = \left(\begin{matrix}
+- \frac{\hat{r}}{K+\hat{p}} \frac{\hat{p}^2 + 2K\hat{p} + K}{\hat{p} + K} & - E_M - \frac{\hat{p}}{K+\hat{p}} (1+\hat{p}) \\
+\frac{\hat{r}}{K+\hat{p}} \left[ \frac{K}{K+\hat{p}} (g(\hat{p}) - \hat{r}) + \hat{p}K\frac{\hat{p}^2 + 2\hat{p} + K}{(\hat{p}^2 + 2K\hat{p} + K)^2}\right] & \frac{\hat{p}}{K+\hat{p}} (g(\hat{p}) - 2\hat{r})
+\end{matrix}\right) .
+\end{equation}
+Evaluated at $(\hat{p}^*, \hat{r}^*) = (\sqrt{K E_M}, g(\sqrt{K E_M}))$, the Jacobian becomes
+\begin{equation}
+\label{eq:supp_jacobian_g_eq}
+J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
+- \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} & - E_M - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} (1+\sqrt{K E_M}) \\
+\frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}}\frac{K E_M + 2 \sqrt{K E_M} + K}{K (E_M + 2 \sqrt{K E_M} + 1)^2} g(\sqrt{K E_M}) & - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} g(\sqrt{K E_M})
+\end{matrix}\right).
+\end{equation}
+Since $K$, $E_M$, and $g(\sqrt{K E_M)} > 0$, it follows immediately that the real part of the eigenvalues of this matrix are both negative.
+\footnote{Notice that the eigenvalues $\lambda_1$ and $\lambda_2$ of $J_{(\hat{p}^*, \hat{r}^*)}$ satisfy the inequalities $\text{Tr}(J) = \lambda_1 + \lambda_2 < 0$ and $\det(J) = \lambda_1 \lambda_2 > 0$.}
+Hence, the non-trivial steady state is stable, completing the proof of Condition~C2.
+
+Here again, it is interesting to observe that the expression in Eq.~\ref{eq:supp_g} is the only function $g(\hat{p})$ satisfying C1-C3. This can be proven in a similar way as for $f$.
+
+\subsubsection{On-off strategy}
+
+The on-off strategy is defined by:
+\begin{equation}
+\alpha = h(\hat{p}, \hat{r}) = 
+\begin{cases}
+0, \ \textrm{if} \ \hat{r} > g(\hat{p}),\\
+1, \ \textrm{if} \ \hat{r} < g(\hat{p}), \\
+\alpha_{opt}^*, \ \textrm{if} \ (\hat{p},\hat{r})=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
+\end{cases}
+\end{equation}
+$h$ is a static function of $\hat{p}$ and $\hat{r}$ (Condition~C1).
+
+As a consequence, the ODE system under the control of $h$ is given by
+\begin{equation}
+\label{eq:supp_adim_h}
+\begin{aligned}
+\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1 + \hat{p}) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r},\\
+\frac{d\hat{r}}{d\hat{t}} &= \left(h(\hat{p}, \hat{r}) - \hat{r} \right) \, \frac{\hat{p}}{K + \hat{p}}\, \hat{r}.
+\end{aligned}
+\end{equation}
+Notice that the system has a discontinuitous right-hand side, due to the fact that $\alpha$ switches between 0 and 1 on $\hat{r}=g(\hat{p})$. 
+Fig.~\ref{fig:sliding} shows the dynamics of the system in the phase plane.
+Due to the direction of the vector fields relative to $\hat{r}=g(\hat{p})$, a \textit{sliding mode} occurs on the latter curve \cite{filippov_differential_1988}.  
+The system is seen to evolve towards a locally asymptotically stable steady state, which is the single non-trivial steady state (Condition~C2). 
+This steady state coincides with the intersection of $\hat{r}=g(\hat{p})$ and the $\hat{p}$-nullcline, which is the steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ allowing maximal growth, thus verifying Condition~C3.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
+\caption
+{
+\textbf{Local stability of the on-off strategy.}
+The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
+The solid, black curve is the $\hat{p}$-nullcline.
+The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
+The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
+The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
+}
+\label{fig:sliding}
+\end{figure}
+
+\clearpage
+
+\section{S2 Text -- Model parameters}
+\manuallabel{S2_Text}{S2~Text}
+
+Most of the conclusions of this paper are parameter-independent in the range of physically admissible values.
+The exact parameter values used in the simulations aim to represent relevant orders of magnitude.
+They were derived from the available literature on fast-growing bacteria (mostly \textit{Escherichia coli}).
+This document describes this derivation for each parameter.
+In the \textit{Methods} section of the main text, we also describe how some of them were validated by fitting the model to available experimental data (Fig.~\ref{fig:model_validation} in the main text).
+
+The model parameters (Eqs.~\ref{eq:pdef}-\ref{eq:rdef} in the main text) are listed in the table below:
+\begin{center}
+\begin{tabular}{|c|c|l|}
+\hline
+Name & Unit & Description \\
+\hline
+$e_M$ & h$^{-1}$ & Constant characterizing nutrient composition of medium\\
+\hline
+$k_R$ & h$^{-1}$ & Rate constant of macromolecular synthesis\\
+\hline
+$K_R$ & g\ L$^{-1}$ & Half-saturation constant of macromolecular synthesis\\
+\hline
+$\beta$ & L\ g$^{-1}$ & Inverse of the cellular density of macromolecules\\
+\hline
+$\alpha$ & -- & Resource allocation parameter\\
+\hline
+\end{tabular}
+\end{center}
+The values derived below are summarized in \ref{S1_Table}.
+
+\subsubsection*{\Large \texorpdfstring{$e_M$}{eM}}
+
+By definition, $e_M$ is the effective turnover of the metabolic macroreaction producing precursors from external substrates, obtained by dividing the reaction rate $v_M$ by the enzyme concentration $m$ (Eq.~\ref{eq:metaflux}).
+The unit of $e_M$ is min$^{-1}$, and can be decomposed as follows:
+\[
+[e_M]  = \frac{\text{[mass of metabolic product]}}{\text{[mass of enzyme M]}\cdot \text{[time]}} = \frac{1}{\text{[time]}}.
+\]
+Note that $e_M = k_M\, s / (K_M + s)$ where $k_M$ is a rate constant, indicating the maximal rate of conversion of external nutrients to precursor metabolites. 
+$e_M$ will thus vary with the concentration $s$ of the external nutrients and the kind of nutrient. 
+For example, the precursor mass that can be produced from 1 g of glucose is higher than that produced from 1 g of acetate.
+
+How can we find a typical value for $k_M$, and thus for $e_M$ (both have the same order of magnitude if we suppose that the reaction is not operating far below saturation, that is, $e_M \approx k_M$)?
+A reasonable estimate for $k_M$ can be obtained from the turnover numbers of reactions involved in the synthesis of charged tRNA, since the latter are directly consumed by the most abundant part of the gene expression machinery, the ribosomes.
+
+Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glutaminyl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second.
+After conversion to mass units using molar weight from~\cite{freist_glutaminyltrna_1997}, this yields
+\[
+k_{\textit{cat,GlnRS}} = \frac{3.2 \cdot 147}{64.4 \cdot 10^ 3} \approx 10^{-3} \, \text{g of glutaminyl-tRNA} \cdot \text{g of enzyme}^{-1} \cdot \text{s}^{-1}.
+\]
+We therefore take
+\[
+k_M \approx 3.6 \text{ h}^{-1},
+\]
+and thus obtain an upper bound for $e_M$ in our simulations.
+
+\subsubsection*{\Large \texorpdfstring{$k_R$}{kR}}
+
+$k_R$ is the mass rate constant describing the maximal rate of conversion of precursors to macromolecules [h\textsuperscript{-1}].
+As for $e_M$, we can decompose this into
+\[
+[k_R]  = \frac{\text{[mass of macromolecules]}}{\text{[mass of gene expression machinery]}\cdot \text{[time]}} = \frac{1}{\text{[time]}}
+\]
+To obtain an order of magnitude for the mass of macromolecules, we focus on proteins since they are the most abundant macromolecules in the cell~\cite{ehrenberg_mediumdependent_2012}.
+The dimensional analysis of $k_R$ thus becomes:
+\[
+\begin{aligned}
+\left[k_R\right] &= \frac{\text{[protein mass produced]}}{\text{[ribosomal mass]} \cdot \text{[hour]}}\\
+&= \frac{\text{[moles of protein]} \cdot \text{[protein molar mass]}}{\text{[moles of ribosome]} \cdot \text{[ribosome molar mass]} \cdot \text{[hour]}}\\
+&= \frac{\text{[moles of amino acids]} \cdot \text{[molar mass of amino acids]}}{\text{[moles of ribosome]} \cdot \text{[hour]} \cdot \text{[ribosome molar mass]}}\\
+&\approx \frac{\text{[maximal protein elongation rate]} \cdot \text{[molar mass of amino-acids]}}{\text{[ribosome molar mass]}}.
+\end{aligned}
+\]
+The values in the last equality are available from the literature~\cite{ehrenberg_mediumdependent_2012,klumpp_molecular_2013,hachiya_increase_2007,yamamoto_mass_2006}.
+We obtain
+\[
+k_R \approx \frac{10 \cdot 100}{10^6}\cdot 3600 \approx 3.6 \text{ h}^{-1}.
+\]
+This value is comparable with the translational capacity $k_T$, in $\mu$g of protein per $\mu$g of ribosomal protein per hour, given by Scott \textit{et al.}~\cite{scott_interdependence_2010}:
+\[
+k_T = \frac{4.5 \text{ }\mu\text{g of protein} \text{ / }\mu\text{g of RNA} \text{ / h }}{0.76 \text{ }\mu\text{g of ribosomal protein} \text{ / } \mu\text{g of RNA}} = 5.9 \text{ h}^{-1}.
+\]
+
+\subsubsection*{\Large \texorpdfstring{$K_R$}{KR}}
+
+A value for the parameter $K_R$, representing the half-saturation constant of macromolecular synthesis, is more difficult to obtain from the literature.
+However, assuming that ribosomes operate close to saturation (80\% over a range of growth rates~\cite{ehrenberg_mediumdependent_2012}), we find that $K_R \approx 0.25\,p$, with p the total amino acid concentration.
+The total concentration of amino acids in the cell is around 150~mmol~L\textsuperscript{-1}~\cite{bennett_absolute_2009}, which with a mean molecular weight of 118.9~g~mol\textsuperscript{-1} for amino acids~\cite{hachiya_increase_2007}, yields a mass concentration of 17.8~g~L\textsuperscript{-1}.
+These considerations led to the following order of magnitude for $K_R$:
+\[
+K_R \approx 1\text{ g L}^{-1}.
+\]
+
+\subsubsection*{\Large \texorpdfstring{$\beta$}{Beta}}
+
+$\beta$ is the inverse of the cellular density of macromolecules, which has been shown constant during balanced growth over a large range of growth rates~\cite{churchward_macromolecular_1982}, and there is some data suggesting that $\beta$ varies little during growth transitions as well~\cite{zhou_carbon_2013}.
+From~\cite{zimmerman_estimation_1991,mcguffee_diffusion_2010} we take the following typical value for $\beta$:
+\[
+\beta \approx \frac{1}{300} \approx 0.003 \text{ L g}^{-1}.
+\]
+
+\subsubsection*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}
+
+From the values of the parameter in the dimensional model, one can deduce the parameters in the nondimensional model used in the simulations:
+\[
+E_M = \frac{e_M}{k_R} = \frac{3.6}{3.6} = 1 \;\;\;\;\; , \;\;\;\;\;  K = \beta\, K_R = 3\cdot 10^{-3} \cdot 1 = 0.003.
+\]
+
+\clearpage
+
+\section{S3 Text -- Solution of optimal control problem}
+\manuallabel{S3_Text}{S3~Text}
+
+\subsubsection{Statement of the problem}
+
+We consider the dimensionless system defined by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} in the main text, which are here repeated for clarity:
+\begin{equation}{\label{eq:model}}
+\begin{aligned}
+\frac{d\hat{p}}{d\hat{t}} &= (1-\hat{r})\, E_M - (1+\hat{p})\, \hat{r} \, \frac{\hat{p}}{K+\hat{p}}, \\
+\frac{d\hat{r}}{d\hat{t}} &= \hat{r} \frac{\hat{p}}{K + \hat{p}} \, (\alpha (\hat{t}) - \hat{r}).
+\end{aligned}
+\end{equation}
+
+As stated in the section \textit{Biomass maximization as an optimal control problem} in the main text, the objective of this study is to maximize the growth rate on an interval $[0,\tau]$ after a nutrient upshift.
+With Eq.~\ref{S1.eq:supp_growthrate_adim} in \ref{S1_Text}, we have
+
+\begin{equation*}
+\hat{\mu}= \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}.
+\end{equation*}
+In order to avoid boundary effects occurring over finite time intervals, notably the depletion of precursors just before $\tau$, we solve the optimal control problem over an infinite horizon ($\tau \rightarrow \infty$).
+Consider the set of admissible controls
+\[
+\mathcal{U}=\{\alpha:\mathbb{R} \rightarrow [0,1] \ \mid \ \alpha(\cdot) \ \mathrm{measurable}\}.
+\]
+The optimization problem can then be stated as follows:
+\begin{equation}\label{Prob}
+\alpha_{opt} = \arg \max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{+\infty} \hat{r}(\hat{t}) \frac{\hat{p}(\hat{t})}{K + \hat{p}(\hat{t})} d\hat{t} ,
+\end{equation}
+where $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))$ is the unique solution of Eq.~\ref{eq:model} starting at a given point $(\hat{p}_0,\hat{r}_0)\in \Omega \equiv \mathbb{R}^+_* \times (0,1)$ for a given control $\alpha\in \mathcal{U}$.
+
+Given that the performance index $J(\alpha)$ diverges, we actually consider \textit{overtaking optimality}~\cite{carlson_infinite_1991}.
+Consider the performance index of the trajectory $x(\cdot)$ emanating from $x_0$ and generated by $u(\cdot)$ defined for any $T\geq 0$ by
+$$
+J_T(x_0,u(\cdot))=\int_0^T f_0(x(t),u(t),t)dt.
+$$
+A trajectory $x^*(\cdot)$ emanating from $x_0$ and generated by $u^*(\cdot)$ is said to be overtaking optimal if for any other trajectory $x(\cdot)$ emanating from $x_0$ and generated by $u(\cdot)$ the following holds
+$$
+\liminf_{T\rightarrow\infty} \left\lbrace J_T(x_0,u^*(\cdot)) - J_T(x_0,u(\cdot))\right\rbrace\geq 0.
+$$
+Roughly speaking, a trajectory is overtaking optimal if "the performance index catches up with the performance index of any other trajectory"~\cite{carlson_infinite_1991}.
+
+\subsubsection{Maximum Principle}
+Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
+Let $H(\hat{p},\hat{r},\lambda_p,\lambda_r,\lambda_0,\alpha)$ be the Hamiltonian of the system, defined by
+\[
+H(\cdot) \equiv \lambda_p\, E_M\, (1-\hat{r}) - \hat{r}\, \dfrac{\hat{p}}{K+\hat{p}}\left[\lambda_p (1+\hat{p}) +\lambda_r\, \hat{r} +\lambda_0\right] + \alpha \, \lambda_r \, \hat{r}\, \frac{\hat{p}}{K+\hat{p}}.
+\]
+Moreover, let $\alpha$ be an optimal control, and $\hat{x}(\cdot)=(\hat{p}(\cdot),\hat{r}(\cdot))$ the associated trajectory.
+Then, there exists $\lambda_0 \leq 0$ and an absolutely continuous map $\lambda=(\lambda_p,\lambda_r):[0,+\infty) \rightarrow \mathbb{R}^2$
+such that $(\lambda,\lambda_0)\neq0$, and
+\begin{align}
+\dot{\lambda}_p\, =-\frac{\partial H}{\partial \hat p}&= \hat{r} \, \frac{K}{(K+\hat{p})^2}\, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, ( \hat{r}-\alpha) + \lambda_0\right] + \hat{r}\, \frac{\hat{p}}{K+\hat{p}}\, \lambda_p, \label{eq:adjoint_p}\\
+ \dot{\lambda}_r\, =-\frac{\partial H}{\partial \hat r}&= \lambda_p\, E_M +\frac{\hat{p}}{K+\hat{p}} \, \left[\lambda_p \, (1+\hat{p}) +\lambda_r\, (2 \hat{r}-\alpha) +\lambda_0\right] \label{eq:adjoint_r}.
+\end{align}
+The maximization condition is given by:
+\begin{equation}{\label{eq:PMP}}
+\begin{aligned}
+&\alpha(\hat{t}) \in \arg\ max_{v \in [0,1]} H(\hat{x}(\hat{t}),\lambda(\hat{t}),\lambda_0,v), \\ &\mathrm{almost}\ \mathrm{everywhere}\ \mathrm{on}\ [0,+\infty).
+\end{aligned}
+\end{equation}
+
+An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}.
+The extremal is said to be normal (resp.
+abnormal) if $\lambda_0<0$ (resp.
+$\lambda_0=0$).
+In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.\\
+
+From Eq.~\ref{eq:PMP}, it follows that the control strategy is given by the sign of the \textit{switching function} $\phi(\cdot) \equiv \lambda_r \, \hat{r}\, \hat{p}/(K+\hat{p})$, that is,
+\[
+\begin{cases}
+\alpha=1 \iff \phi(\cdot) >0,\\
+\alpha=0 \iff \phi(\cdot)<0.
+\end{cases}
+\]
+
+Finally, given that the system is autonomous, the Hamiltonian is conserved along any extremal trajectory.
+
+\subsubsection{Characterization of singular arcs}
+
+Whenever $\phi$ is vanishing over a time interval, we say that the trajectory is \textit{singular}.
+We will now characterize such trajectories.
+If $I=[\hat{t}_1,\hat{t}_2]$ is a singular arc, we have $\phi(\hat{t})=\dot{\phi}(\hat{t})=0$, for all $\hat{t}\in[\hat{t}_1,\hat{t}_2]$, that is, $\lambda_r(\hat{t})=0$ and $\dot\lambda_r(\hat{t})=0$.
+
+For abnormal extremal trajectories, we get $\lambda_p(\hat{t})=0$, in contradiction with the Maximum Principle, so there is no singular arc.
+An abnormal extremal trajectory is therefore a concatenation of bang arcs.
+
+For normal extremal trajectories, using additionally that $H$ is constant along an extremal trajectory, we obtain that $\lambda_p$ is constant along a singular arc.
+By combining $\dot \lambda_p=0$ and $\dot \lambda_r=0$, we obtain $\hat{p}(\hat{t})=\sqrt{E_M\, K}=\hat{p}_{opt}^*$.
+Using $d\hat{p}/d\hat{t}=0$, we finally get $\hat{r}(\hat{t})=\hat{r}_{opt}^*$.
+Thus, the singular arc is the optimal steady state, corresponding to a singular control $\alpha(\hat{t})=\alpha_{opt}^*$, with $\alpha_{opt}^*$ depending on $E_M$ (\ref{S1_Text}).
+
+A necessary condition of optimality for a singular arc is given by the Kelley condition~\cite{borisov_fullers_2000}.
+We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
+$$
+(-1)^q\frac{\partial}{\partial\alpha}\frac{d^{2q}}{d\hat{t}^{2q}}\phi(\hat{t})<0,
+$$
+satisfying the Kelley condition necessary for optimality.
+Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (also called the Fuller's phenomenon, \textit{i.e.}, an arc with an infinite number of switches~\cite{borisov_fullers_2000,marchal_chattering_2013}).
+\\
+
+\subsubsection{Analysis of the adjoint system}
+Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}) may be useful to characterize the switches of extremal trajectories.
+
+First, for the abnormal case, we can easily determine in the phase-plane the possible transitions between the four regions defined by the axes (see Fig.~\ref{fig-adj}).
+A trajectory can cross at most twice the $\lambda_p$-axis, so  we conclude that an abnormal extremal cannot have more than two switches.
+Thus, an abnormal extremal is a concatenation of at most three bang arcs ($\alpha(t)=0$ or $\alpha(t)=1$).
+When  $\alpha(t)=0$ or $\alpha(t)=1$ for a long time, the growth rate tends to zero.
+We therefore conclude that  abnormal extremal trajectories are not optimal.
+
+Secondly, for the normal case, after the first switch, a trajectory with two consecutive switches in the regions $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$ is not possible, as shown in Fig.~\ref{fig-adj}.
+Therefore, such a trajectory is not optimal given that it does not fulfill the conditions given by the Maximum Principle.
+We conclude that if the optimal trajectory has a concatenation of bang arcs, the switches must alternatingly occur in the regions  $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ and $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.35\textwidth]{Fig/adj2ab.eps} 
+\includegraphics[width=0.6\textwidth]{Fig/adj2.eps} 
+\caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis.
+Left: abnormal case.
+An extremal trajectory cannot have more than two switches.
+Right: normal case.
+$(\lambda_p^s,0)$ corresponds to the singular arc.
+After the first switch, an extremal trajectory cannot have two consecutive switches if it stays in the region $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.}
+\label{fig-adj}                               
+\end{figure}
+
+\subsubsection{Optimal trajectories}
+
+From the Maximum Principle, we have shown that the optimal trajectory is a concatenation of bang arcs ($\alpha(t)=0$ or $\alpha(t)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
+Moreover, if the optimal trajectory has a singular arc, it must enter it through a chattering arc (\textit{i.e.}, with an infinite number of switches between $\alpha=0$ and $\alpha=1$).
+
+These elements motivate the supposition that optimal solutions consist in a transient (chattering arc) towards the optimal steady state, after which they remain there (until the next change of environment).
+The chattering arc can be characterized by a switching curve $\hat{p}\mapsto\varphi(\hat{p})$ which passes through the optimal steady state.
+Defining $A_0$ and $A_1$ the regions above and below $\varphi$ in the $(\hat{p},\hat{r})$-plane, respectively, we conjecture that the following feedback control law is optimal: 
+%The simplest way to reach the optimal steady-state is thus two arcs, either 0-1 or 1-0 depending on the initial conditions.
+More precisely, let $\phi_0$ and $\phi_1$ be defined as the the trajectories solution of System \eqref{model} backward in time starting at $(p_{opt}^*,r_{opt}^*)$ with respectively $\alpha=0$ and $\alpha=1$.
+Given $\phi=\phi_0 \cup \phi_1$, we define $A_0$ and $A_1$ the regions above and below $\phi$ in the plane $(p,r)$.
+
+\begin{equation}\label{control-opt}
+\begin{cases}
+\alpha(\hat{t})=0 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_0,\\
+\alpha(\hat{t})=1 \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))\in A_1,\\
+\alpha(\hat{t})=\alpha_{opt} \ \textrm{if} \ (\hat{p}(\hat{t}),\hat{r}(\hat{t}))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*).
+\end{cases}
+\end{equation}
+
+Loosely speaking, the chattering arc corresponds to a spiral composed of bang arcs wrapping around the optimal steady state, where the switches alternatingly occur in the regions  $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ and $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$, in line with the analysis of the adjoint system.
+This is a first hint that the proposed control strategy is optimal. Moreover, our conjecture is also in line with the \textit{turnpike property}:
+Trélat and Zuazua~\cite{trelat_turnpike_2015} have shown that, for quite a generic class of systems, the optimal strategy consists in staying at the optimal steady state (after a short transient).
+
+As explained in the \textit{Methods} section of the main text, we numerically solved the optimal control problem by the direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012}. 
+It is important to stress that the optimization process was performed without any preliminary assumptions on the characteristics of the optimal trajectory. 
+The fact that the numerical solution verifies the Maximum Principle (\textit{i.e.}, the singular arc corresponds to the optimal steady state) and the Kelley condition (\textit{i.e.}, the presence of a chattering arc)
+tends to confirm that the control strategy of Eq.~\ref{control-opt} is optimal.
+As an aside, we note that due to the fact that numerical optimization was performed for a finite horizon, we actually obtained a second chattering arc escaping from the singular arc at the end of the simulation.
+This is a classical property of the turnpike strategy: the optimal trajectory leaves the optimal steady state just before the end of the time interval of interest, in our case consuming almost all precursors.
+This arc was removed from the plot in Fig.~\ref{fig:optimalcontrol}, because it does not occur with an infinite horizon and is therefore a numerical artifact for this study.
+
+\clearpage
+
+\section[S4 Text -- Kinetic model of the ppGpp system]{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
+\sectionmark{S4 Text -- Kinetic model of the ppGpp system}
+\sectionmark{S4 Text -- Kinetic model of the ppGpp system}
+
+\manuallabel{S4_Text}{S4~Text}
+
+The recently published model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015} provides a synthesis of the currently available knowledge of the ppGpp regulatory system.
+Through the mechanisms of ppGpp production and degradation, it describes regulation of the synthesis of ribosomal RNA.
+We explain below how we use the model to compare the action of the ppGpp system with the on-off control strategy.
+The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} , and is reproduced in Table~\ref{S4.tab:notations} in order to make the text self-contained.
+
+The evolution of the cellular concentration of ppGpp is described in~\cite{bosdriesz_how_2015} by
+\begin{equation}
+\label{eq:ppGpp}
+\frac{d\texttt{ppGpp}}{dt} = v_{RelA}(r_{t,tot}) + v_{spoT} - k_{spoT} \cdot \texttt{ppGpp},
+\end{equation}
+where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{S4.tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:
+\begin{equation}
+\label{eq:vrelA}
+v_{RelA}(r_{t,tot}) = k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}}{K_{D,RelA} + r_{t,tot}}.
+\end{equation}
+The amount of stalled ribosomes is determined by the equilibrium between charged and uncharged tRNA, $t_{ai}$ and $t_{i}$, in the cell:
+\begin{equation}
+\label{eq:rttot}
+r_{t,tot} = \sum_i r_{ti} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+t_{ai}/\kappa_{ta}+t_i/\kappa_{t}},
+\end{equation}
+which can be rewritten as
+\begin{equation}
+\label{eq:rttot-f}
+r_{t,tot} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
+\end{equation}
+using the assumption that $t_{tot,i}  = t_{ai} + t_i = 0.5 \cdot r$.
+$r_i$ denotes the concentration of ribosomes recognizing amino acid  $i$.
+Finally, with $r = \sum_i r_i$ the total ribosome concentration and $a_i$ the concentration of amino acid $i$, the dynamics of the charged tRNA concentration is described by
+\begin{equation}
+\label{eq:ti_dynamic}
+\frac{dt_{ai}}{dt} = v_{tai}(a_i, t_{i}) - f_i \cdot v_{ribosome}(t_i,r),
+\end{equation}
+with $v_{tai}(a_i, t_{i})$ the synthesis rate of charged tRNA, and $f_i \cdot v_{ribosome}(t_i,r)$ their consumption via protein synthesis.
+In particular,
+\begin{eqnarray}
+v_{tai}(a_i, t_{i}) &=& k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\,  K_{Mai} + a_i\,  K_{Mti} + t_i\,  a_i}, \label{eq:vtai}\\
+v_{ribosome}(t_i, r) &=& k_{rib} \cdot r \cdot \left(1+ \sum_i \left[ f_i \cdot \left( 1 + \frac{t_i}{\kappa_t} \right) \frac{\kappa_{ta}}{0.5\cdot r - t_i} \right]\right)^{-1}. \label{eq:vrib}
+\end{eqnarray}
+
+For comparison with our framework, we need $\texttt{ppGpp}$ as a direct function of the total amino acid concentration $a = \sum_i a_i$ (a proxy for precursors) and total ribosome concentration $r$ (a proxy for gene expression machinery).
+To this end, we made two additionnal assumptions:
+\begin{description}
+\item[(A1)] All concentrations specific to one type of amino acid $i$ ($a_i$, $t_{ai}$, $t_{i}$, $r_i$) are in the same proportion $f_i = f = 1/20$ with respect to the total concentrations ($a$, $t_{a}$, $t$, $r$).
+\item[(A2)] We apply a quasi-steady-state approximation (QSSA) to the dynamics of the concentration of the charged tRNAs ($t_{ai}$) and the concentration of ppGpp ($\texttt{ppGpp}$).
+That is, the dynamics of these variables are assumed fast relative to the dynamics of the amino acid concentrations ($a_i$) and the total ribosome concentration ($r$).
+\end{description}
+Using (A2), we can rewrite Eq.~\ref{eq:ti_dynamic} as follow:
+\begin{equation}\label{eq:vtai_vrib}
+v_{tai}(a_i, t_{i}) = f_i \cdot v_{ribosome}(t_i,r),
+\end{equation}
+which, using (A1) and Eqs~\ref{eq:vtai} and \ref{eq:vrib}, leads to:
+\begin{equation}
+\label{eq:vtai_vrib_without_sum}
+k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\, K_{Mai} + a_i\, K_{Mti} + t_i\, a_i}
+= f_i \cdot k_{rib} \cdot r \cdot
+  \left(1 + \frac{\kappa_{ta}}{0.5 r - t_i} \cdot \left( 1 + \frac{t_i}{\kappa_t} \right)  \right)^{-1}.
+\end{equation}
+By rearranging both sides of the equation, $t_i$ can be expressed as a function of $a_i$ and $r$, which yields:
+
+\begin{equation}
+\begin{alignedat}{2}
+&A {t_i}^2 + B t_i + C = 0, \;\;\; \text{with}\\
+&A = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i\, k_{rib} \,r} \left( \frac{\kappa_{ta}}{\kappa_t} - 1\right) + K_{Mai} + a_i,\\
+&B = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i \,k_{rib} \,r} (0.5\,r+\kappa_{ta}) + a_i \,K_{Mti} - 0.5\,r\,(K_{Mai} + a_i),\\
+&C = - 0.5 \,r \,a_i \,K_{Mti},
+\end{alignedat}
+\end{equation}
+and therefore
+\[
+t_i (a_i, r) = \frac{-B \pm \sqrt{B^2 - 4AC}}{2A}.
+\]
+It is not difficult to show that the only solution on $[0, 0.5\,r]$ is
+\begin{equation}
+\label{eq:ti_final}
+t_i (a_i, r) = \frac{-B + \sqrt{B^2 - 4AC}}{2A}.
+\end{equation}
+From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{S4.eq:rttot-f}:
+\begin{equation}
+\label{eq:rttot-f_without_sum}
+r_{t,tot}(t_i, r) = r \cdot \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
+\end{equation}
+and substituting $t_i$ by the expression of Eq.~\ref{eq:ti_final}.
+
+Finally, we apply (A2) to Eq.~\ref{eq:ppGpp} and obtain the final expression giving the concentration of ppGpp as a function of the total amino acid and ribosome concentrations:
+\begin{equation}
+\texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
+\end{equation}
+This function is represented in Fig.~\ref{S4.fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.
+
+The plotted surface of the function resembles the inverse of the on-off control strategy in Fig.~\ref{fig:ppGppsurface}, as expected, bearing in mind that ppGpp has an inhibitory effect on the synthesis of ribosomal RNA.
+We assumed a Michaelis-Menten inhibition for the regulatory effect of ppGpp on rRNA synthesis, and thus indirectly on the synthesis of ribosomal proteins~\cite{potrykus_pppgpp_2008,keener_regulation_1996}:
+\begin{equation}
+\alpha (\texttt{ppGpp}) = \frac{K_I}{K_I + \texttt{ppGpp}}.
+\end{equation}
+The inhibitory constant $K_I$ lies in the dynamical range of variation of $\texttt{ppGpp}$.
+In Fig.~\ref{fig:ppGppsurface} in the main text, we took $K_I = 10$~$\mu$M.
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1.eps}
+\caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
+{
+{\bf ppGpp concentration is a function of total ribosome and amino acid concentrations.}\newline
+We assume the dynamics of ppGpp to be fast on the time-scale of changes in the ribosome and amino acid concentrations.
+The concentration of ppGpp can thus be expressed as a function of the latter two variables, using the model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}.
+Parameters are taken from Table~\ref{tab:notations}.
+}
+\label{fig:ppGpp}
+\end{figure}
+
+\begin{table}[!h]
+\tiny
+\centering
+\begin{tabular}{cccl}
+\hline
+\hline
+\textbf{Symbol} & \textbf{Value} & \textbf{Unit} & \textbf{Description} \\
+\hline
+\hline
+$a_i$ & -- & $\mu$M & Concentration of aa $i$ (not incorporated in protein)\\
+
+$t_{ai}$ & -- & $\mu$M & Concentration of tRNA charged with aa $i$\\
+
+$t_{i}$ & -- & $\mu$M & Concentration of free tRNA conjugate to aa $i$\\
+
+$t_{tot,i}$ &  $0.5 \cdot r$ & $\mu$M & Total concentration of tRNA conjugate to aa $i$\\
+
+$r_{i}$ & -- & $\mu$M & Total concentration of ribosome with an A-site for aa $i$\\
+
+$r_{ti}$ & -- & $\mu$M & Ribosomes with uncharged tRNA in an A-site for aa $i$\\
+
+$\texttt{ppGpp}$ & -- & $\mu$M & Concentration of ppGpp\\
+\hline
+$a$ & $\sum_i a_{i}$ & $\mu$M & Total concentration of aa (not incorporated in protein)\\
+
+$t_{a}$ & $\sum_i t_{ai}$ & $\mu$M & Total concentration of tRNA charged with aa\\
+
+$t$ & $\sum_i t_{i}$ & $\mu$M & Total concentration of free tRNA\\
+
+$r_{t,tot}$ & $\sum_i r_{ti}$ & $\mu$M & Total concentration of uncharged tRNA bound to ribosomes\\
+
+$r$ & $\sum_i r_i$ & $\mu$M & Total concentration of ribosomes\\
+\hline
+$v_{RelA}$ & -- & $\mu$M/s & Rate of RelA-catalyzed ppGpp synthesis\\
+
+$v_{SpoT}$ & $10^{-3}$ & $\mu$M/s & Rate of ppGpp synthesis by SpoT\\
+
+$v_{tai}$ & -- & $\mu$M/s & Rate of amino-acyl tRNA $i$ synthetase\\
+
+$v_{ribosome}$ & -- & $\mu$M/s & Total rate of protein synthesis\\
+\hline
+$k_{rib}$ & 20 & s\textsuperscript{-1} & $k_{cat}$ of protein elongation\\
+
+$k_{RelA}$ & 75 & s\textsuperscript{-1} & $k_{cat}$ of ppGpp synthesis by RelA\\
+
+$K_{D,RelA}$ & 0.26 & $\mu$M & Michaelis constant of RelA-catalyzed ppGpp production\\
+
+$RelA_{tot}$ & 1/15 & $\mu$M & RelA concentration\\
+
+$k_{SpoT}$ & $\ln (2) / 30$ & s\textsuperscript{-1} & Rate of ppGpp degradation by SpoT\\
+
+$\kappa_t$ & 500 & $\mu$M & Dissociation constant of uncharged tRNA-ribosome complex\\
+
+$\kappa_{ta}$ & 1 & $\mu$M & Dissociation constant of charged tRNA-ribosome complex\\
+
+$k_{Si}$ & 100 & s\textsuperscript{-1} & $k_{cat}$ of aminoacyl-tRNA synthetase\\
+
+$S_{tot,i}$ & 1 & $\mu$M & Total concentration of aminoacyl-tRNA synthetase for aa $i$\\
+
+$K_{Mai}$ & 100 & $\mu$M & Michaelis constant of aa-tRNA synthetase for amino acids\\
+
+$K_{Mti}$ & 1 & $\mu$M & Michaelis constant of aa-tRNA synthetase for uncharged tRNA\\
+
+$f_i$ & 1/20 & -- & Proportion of aa $i$ in proteins (Assumption A1)\\
+
+\end{tabular}
+\caption{\textbf{Parameters and variables reused from Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}}. The abbreviation aa denotes amino acids.}
+\label{tab:notations}
+\end{table}
+
+\clearpage
+
+\section{S1 Table -- Parameter values of self-replicator model}
+\manuallabel{S1_Table}{S1~Table}
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+Parameter & Unit & Literature value & Fitted value \\
+\hline
+$\gamma$ & $-$ & No value & 1.39\\
+\hline
+$k_R$ & h$^{-1}$ & 3.6 & 2.23\\
+\hline
+$e_M$ for M63+glycerol & h$^{-1}$ &  $< 3.6$ & 0.587\\
+\hline
+$e_M$ for M63+glucose & h$^{-1}$ &  $< 3.6$ & 0.867\\
+\hline
+$e_M$ for cAA+glycerol & h$^{-1}$ &  $< 3.6$ & 1.07\\
+\hline
+$e_M$ for cAA+glucose & h$^{-1}$ &  $< 3.6$ & 1.57\\
+\hline
+$e_M$ for RDM+glycerol & h$^{-1}$ &  $< 3.6$ & 3.48\\
+\hline
+$e_M$ for RDM+glucose & h$^{-1}$ &  $< 3.6$ & 4.76\\
+\hline
+$\beta K_R$ & $-$  & 0.003 & Not fitted\\
+\hline
+\end{tabular}
+\caption{\textbf{Parameter values of self-replicator model}
+The parameter values in the model were obtained by fitting Eq.~\ref{eq:alpha_mu_optimal_dim} to the data of Scott et al~\cite{scott_interdependence_2010} (Fig.~\ref{fig:model_validation} in the main text), as described in the \textit{Methods} section.
+They are compared with order-of-magnitude estimates from the literature (\ref{S2_Text}).}
+\end{table}
+
+\clearpage
+
+\section{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth}
+\manuallabel{S1_Fig}{S1~Figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{./Fig/plot_strategies.eps}
+\caption{
+\noindent\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
+The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state, that is, the state in which growth occurs at the maximal rate supported by $E_M$.
+The function $f$ is defined by Eq.~\ref{eq:meth_f} in the \textit{Methods} section of the main text.
+The solid, red curve is an approximation of this function by the simple Michaelis-Menten curve of Eq.~\ref{eq:MMApprox}, with $K_{mE} = 1.0$.
+\textit{B:} Precursor-only strategy: $\alpha = g(\hat{p})$.
+The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state.
+The function $g$ is defined by Eq.~\ref{eq:meth_g} in the \textit{Methods} section of the main text.
+The solid, red curve is an approximation of this function by the simple Hill curve of Eq.~\ref{eq:HillApprox}, with $K_{mp} = 0.06$ and a cooperativity coefficient 2.}
+\end{figure}

### 2016-06-06 titlepage.tex
Refactoring paragraph 3 and changing a lot of file names

% les lignes en bas sont à insérer obligatoirement après \begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%             Commandes Meta-données               %%
%%   à renseigner par les auteurs pour générer      %%
%%     la couverture modèle Univ. Grenoble          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\Sethpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture horizontalement (valeur par défaut est -5.5mm)
%\Setvpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture verticalement (valeur par défaut est -15.5mm)

%\Universite{}    %%optionnel : à décommenter et à renseigenr si vous voulez changer le non d'université
%\Grade{}         %%optionnel : à décommenter et à renseigenr si vous voulez changer le grade
\Specialite{Biologie cellulaire }
\Arrete{7 Ao\^ut 2006}
\Auteur{Nils Giordano}
\Directeur{Johannes Geiselmann}
\CoDirecteur{Hidde de Jong}    %%optionnel : à décommenter et à renseigenr si présence d'un Co-directeur de thèse
\Laboratoire{Laboratoire Interdisciplinaire de Physique}
\EcoleDoctorale{Ecole doctorale Chimie et Science du Vivant}         
\Titre{Growth control in microorganisms: theoretical and experimental study by mean of self-replicator models}
%\Soustitre{}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
\Depot{}

% Commande pour création de nouvelles catégories dans le jury:

%\UGTNewJuryCategory{...NomDeLaCategorie...}{...Definition...}

% Exemple \UGTNewJuryCategory{UGTFamille}{Membre de la famille} que nous ajoutons dans la commande \Jury ci-dessous sous la forme \UGTFamille{Jean Rousseau}{(...titre_et_affiliation...s'il_y_en_a...)}

\Jury{
\UGTPresident{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
%\UGTPresidente{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}

\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% 1er rapporteur
%\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% second rapporteur

\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% 1er examinateur
%\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% second examinateur
%\UGTExaminatrice{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}    %% 3ème examinateur

\UGTDirecteur{M, Johannes Geiselmann}{Professeur Université Joseph Fourier, LIPhy, Grenoble}       %% Directeur de thèse
\UGTCoDirecteur{M, Hidde de Jong}{Directeur de recherche Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a

\UGTInvite{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
}

\MakeUGthesePDG    %% très important pour générer la couverture de thèse

### 2016-07-25 titlepage.tex
Remove useless pages, reorganize some parts

@@
 
-\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% 1er rapporteur
-%\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% second rapporteur
+\UGTRapporteur{M, Matthew Scott}{Associate Professor, Department of Applied Mathematics, University of Waterloo}      %% 1er rapporteur
+\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% second rapporteur
 
-\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% 1er examinateur
+\UGTExaminateur{M, Matthieu Jules}{Chargé de recherche, Inra, Grignon}     %% 1er examinateur
+\UGTExaminateur{M, Olivier Bernard}{Chargé de recherche, Inria, Nice -- Sophia-Antipolis}     %% 1er examinateur
 %\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% second examinateur
-%\UGTExaminatrice{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}    %% 3ème examinateur
+\UGTExaminatrice{Mme, Irina Mihalcescu}{Maître de conférence, Université Grenoble Alpes, LIPhy}    %% 3ème examinateur
 
-\UGTDirecteur{M, Johannes Geiselmann}{Professeur Université Joseph Fourier, LIPhy, Grenoble}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde de Jong}{Directeur de recherche Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
+\UGTDirecteur{M, Johannes Geiselmann}{Professeur, Université Grenoble Alpes, LIPhy}       %% Directeur de thèse
+\UGTCoDirecteur{M, Hidde de Jong}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
 
-\UGTInvite{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
-\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
+\UGTInvite{Jean-Luc Gouzé}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}
+%\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
 }

### 2016-07-25 Chapters/experiments.tex
Remove useless pages, reorganize some parts

\chapter{Monitoring of Gene Expression Machinery during growth transitions}

Missing results as of today:
\begin{itemize}
\item manual segmentation of one well --> mean RFU per pixel per cell (2-3 days in the best case, it could be difficult to export schnitzcells data)
\item semi-manual tracking of cells based on segmentation (hard to evaluate, will depend on schnitzcells performances, not optimistic)
\item estimation of $\alpha$ and $\mu$ over time (2-3 days for calculation, 2-3 days for implementation). I need tracking for $\mu$, maybe not for $\alpha$
\item make another experiments glu-sugar-glu (glycerol?), depends on Irina
\item analyze this experiments (repeat steps above, will probably be faster if we manage to work everything out, but still takes time because it's manual)
\item model modification to take into account the new data (5 days)
\end{itemize}

What could go wrong: everything. In particular:
\begin{itemize}
\item segmentation results could be impossible to export
\item tracking might not work (schnitzcells not designed to handle massive cell disappearance)
\item $\mu$ might be too noisy due to pixel size (note: image interpolation before segmentation might be a solution)
\item what we saw on the rough analysis might just be an artefact (as Hans usually says "When it works the first time, something went horribly wrong.")
\item the second experiment might not work (poor preculture, device plugged, out of focus overnight, medium contamination...)
\end{itemize}

Would everything go right, the chapter/article would be as follow.

\section{Introduction}

\begin{itemize}
\item Understanding growth transitions is important, because so and so.
\item We predicted earlier counter-intuitive bang-bang-singular behaviours for GEM during transition
\item But we're missing experimental data on what really happens in real cells during these transitions (quick review of available data)
\item We fix it by making a strain with fluorescent ribosomes and monitoring ribosome abundance during growth transitions in a microfluidic device.
\end{itemize}

\section{Methods}

\subsection{Strain design}

We used rpsB, GFP, and a chromosomic construction because so and so.

\subsection{Strain construction}

We used Gibson assembly, Kan-pbad-ccdb, etc and obtain the following sequence.

\subsection{Microfluidic device}

Mother machine because needed to change the medium, single cell data because asynchronization etc.

\subsection{Data analysis}

We used this and that to analyze the microscopic images.
We applied the following corrections.

\section{Results}

\subsection{Model calibration}

We want to get $\alpha (t)$ during a transition. We show that $\alpha$ can be obtained if we measure so and so.

\subsection{Bang-bang expression during transitions, as predicted}

One stunning figure that show the RFU / cell / pixels, and the predicted $\alpha$ for the glucose-acetate transition.

\subsection{Discrepancies with the model and model adaptation}

The growth rate stalls after the addition of glucose, which was not originally predicted by the model.
This is because so and so.
This can be taken into account if we add another sector (M1 and M2).
This raises a new interesting question : during the transition amino-acids drop, so the cell should produce ribosomes, but it also need to produce the new enzymes to actually be able to perform metabolism.
We thus have competition between several layers of regulation, which can be added to the model by doing so and so.

\section{Discussion}

More research need to be done.

Could be interesting to use transitions for syn bio purposes.
Could be interesting to validate such results in different microorganisms, with different type of medium (fast env, slow env, etc)

\section{Supporting Information}

\begin{itemize}
\item Strain validation (signal-to-noise ratio, growth rate not hampered in microplate reader)
\item Primers used for gibson assembly
\item Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
\end{itemize}

### 2016-07-25 Chapters/introduction.tex
Remove useless pages, reorganize some parts

@@
 \textit{There's an infinity of things that have never been done before, and
-most of those things are not worth doing.} -- Jeremy Fox
+most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology}, 2016
 
@@
 
-The theory of evolution has deeply changed biology.\\
-\textit{Very broad introduction on how we use evolution to actually make sense of biology}
-
-We are still trying to understand the fundamental forces that shape life in all its diversity.\\
-\textit{Evolution through natural selection is a verbal theory, and people have been trying to formalize it for 200 years.
-It implies working and looking for "the hand of evolution" at several scales, from population to molecular biology.
-It is challenging to translate the mechanisms of natural selection (fitness) at every scale, hence the problems to uncover fundamental properties of life.}
-%Since its invention, huge efforts have been made to put Darwin's theory into equation, in order to clarify our understanding, to see how it apply at different scales (species, populations, cells, molecules...).
+%The theory of evolution has deeply changed biology.\\
+%\textit{Very broad introduction on how we use evolution to actually make sense of biology}
+%
+%We are still trying to understand the fundamental forces that shape life in all its diversity.\\
+%\textit{Evolution through natural selection is a verbal theory, and people have been trying to formalize it for 200 years.
+%It implies working and looking for "the hand of evolution" at several scales, from population to molecular biology.
+%It is challenging to translate the mechanisms of natural selection (fitness) at every scale, hence the problems to uncover fundamental properties of life.}
+%%Since its invention, huge efforts have been made to put Darwin's theory into equation, in order to clarify our understanding, to see how it apply at different scales (species, populations, cells, molecules...).
 
@@
 \textit{They are open systems that take energy from the environment around them, and use it to maintain and reproduce their structures.
-They live literally everywhere. They represent a perfect model organism for fundamental questions.}
+They live literally everywhere, have the same metabolism, etc...
+They represent a perfect model organism for fundamental questions.}
 
 They also managed to become the thoughest of all the living systems.
-\textit{Their diversity allow them to colonize every biotope, but remarkably, a single species can live in broad conditions, growing on different carbon sources or resist to really harmful environments.}
+\textit{Their diversity allow them to colonize every biotope, but remarkably, a single species can live in broad conditions, growing on different carbon sources or resist to really harmful environments (even space!).
+}
 
-Their capacity to survive and exploit strong environmental fluctuations lay on another ubiquitous process of life as we know it: gene expression.
+Their capacity to survive and exploit strong environmental fluctuations lay on another ubiquitous process of life as we know it: self-catalysis and gene expression.
 \textit{Short explanation on DNA, RNA, Protein and the regulations behind, that can be seen as loading "programs" depending on the environment the organism as to cope with.}
@@
 Such a mechanism can be seen as a resource allocation problem.
-\textit{Limited protein production capacity, so choice have to be made for what need to be produced, natural selection selecting for the best strategies.}
+\textit{Because they can build everything from a single carbon source.
+But they have a limited protein production capacity, so choice have to be made for what need to be produced, natural selection selecting for the best strategies.}
 

### 2016-07-25 Chapters/theory.tex
Remove useless pages, reorganize some parts

@@
 
-\section{Steady-state analysis of model}
+\subsection{Steady-state analysis of model}
 \label{sec:methods_model_analysis}
@@
 
-\section{Model fitting}
+\subsection{Model fitting}
 \label{sec:methods_model_fitting}
@@
 
-\section{Solution of optimal control problem}
+\subsection{Solution of optimal control problem}
 
@@
 
-\section{Specification and analysis of control strategies}
+\subsection{Specification and analysis of control strategies}
 
@@
 
-\subsubsection{Model formulation}
+\subsection{Model formulation}
 
@@
 
-\subsubsection{Nondimensionalization of the system}
+\subsection{Nondimensionalization of the system}
 
@@
 
-\subsubsection{Steady-state growth of the self-replicator}
+\subsection{Steady-state growth of the self-replicator}
 \label{si::growthrate}
@@
 
-\subsubsection{Maximization of growth rate at steady state}
+\subsection{Maximization of growth rate at steady state}
 \label{si::optimal}
@@
 
-\subsubsection{Analysis of the control strategies}
+\subsection{Analysis of the control strategies}
 \label{si::control_strategies}
@@
 
-\subsubsection{Precursor-only strategy}
+\subsection{Precursor-only strategy}
 
@@
 
-\subsubsection{On-off strategy}
+\subsection{On-off strategy}
 
@@
 
-\subsubsection*{\Large \texorpdfstring{$e_M$}{eM}}
+\subsection*{\Large \texorpdfstring{$e_M$}{eM}}
 
@@
 
-\subsubsection*{\Large \texorpdfstring{$k_R$}{kR}}
+\subsection*{\Large \texorpdfstring{$k_R$}{kR}}
 
@@
 
-\subsubsection*{\Large \texorpdfstring{$K_R$}{KR}}
+\subsection*{\Large \texorpdfstring{$K_R$}{KR}}
 
@@
 
-\subsubsection*{\Large \texorpdfstring{$\beta$}{Beta}}
+\subsection*{\Large \texorpdfstring{$\beta$}{Beta}}
 
@@
 
-\subsubsection*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}
+\subsection*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}
 
@@
 
-\subsubsection{Statement of the problem}
+\subsection{Statement of the problem}
 
@@
 
-\subsubsection{Maximum Principle}
+\subsection{Maximum Principle}
 Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
@@
 
-\subsubsection{Characterization of singular arcs}
+\subsection{Characterization of singular arcs}
 
@@
 
-\subsubsection{Analysis of the adjoint system}
+\subsection{Analysis of the adjoint system}
 Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}) may be useful to characterize the switches of extremal trajectories.
@@
 
-\subsubsection{Optimal trajectories}
+\subsection{Optimal trajectories}
 

### 2016-07-25 acknowledgements.tex
Remove useless pages, reorganize some parts

\cleardoublepage
\begin{acknowledgements}
(remerciements ici)
\end{acknowledgements}

### 2016-07-25 Chapters/introduction.tex
update intro

@@
 
-%The theory of evolution has deeply changed biology.\\
-%\textit{Very broad introduction on how we use evolution to actually make sense of biology}
-%
-%We are still trying to understand the fundamental forces that shape life in all its diversity.\\
-%\textit{Evolution through natural selection is a verbal theory, and people have been trying to formalize it for 200 years.
-%It implies working and looking for "the hand of evolution" at several scales, from population to molecular biology.
-%It is challenging to translate the mechanisms of natural selection (fitness) at every scale, hence the problems to uncover fundamental properties of life.}
-%%Since its invention, huge efforts have been made to put Darwin's theory into equation, in order to clarify our understanding, to see how it apply at different scales (species, populations, cells, molecules...).
-
-Microorganisms exhibits the most fundamental features of life. 
-\textit{They are open systems that take energy from the environment around them, and use it to maintain and reproduce their structures.
-They live literally everywhere, have the same metabolism, etc...
-They represent a perfect model organism for fundamental questions.}
+Most part of life beauty and diversity is hidden from us in the form of microorganisms.
+These small factories are optimized to grab and use energy from their environment, in the form of organic matter, highly reactive chemicals, or even photons [citation trophic].
+In that sense, microorganisms exhibit the most fundamental features of life.
+They live almost everywhere, sometimes in the strangest possible places [citation space].
+This nature of ubiquitous, extremely optimized living systems makes them perfect to unearth the fundamental forces that shape life as we know it.
 

### 2016-09-05 Chapters/introduction.tex
Context and problem statement first version, sent to Hidde and Hans

@@
 
-\textit{There's an infinity of things that have never been done before, and
-most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology}, 2016
+\textit{There's an infinity of things that have never been done before, and most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology}~\cite{fox_how_2016}
 
@@
 
-In this section...
+(passive or active voice?)
 
@@
 
-\section{Context and motivations}
+\section{Context}
 
-Most part of life beauty and diversity is hidden from us in the form of microorganisms.
-These small factories are optimized to grab and use energy from their environment, in the form of organic matter, highly reactive chemicals, or even photons [citation trophic].
-In that sense, microorganisms exhibit the most fundamental features of life.
-They live almost everywhere, sometimes in the strangest possible places [citation space].
-This nature of ubiquitous, extremely optimized living systems makes them perfect to unearth the fundamental forces that shape life as we know it.
+\subsection{Self-replication is a resource allocation problem}
 
-They also managed to become the thoughest of all the living systems.
-\textit{Their diversity allow them to colonize every biotope, but remarkably, a single species can live in broad conditions, growing on different carbon sources or resist to really harmful environments (even space!).
+At every moment of our lives, we are surrounded by billions of microscopic organisms.
+Sometimes beneficial, sometimes harmful, these small factories grab and use energy from their environment, in the form of organic matter, highly reactive chemicals, or light [citation trophic].
+While most cells in our body are kept in a stable and friendly micro-environment (temperature, oxygenation, nutrient abundance...), microorganisms made the evolutionary choice to constantly adapt their physiology to a strongly variable environment.
+Microbes have been found to evolve almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life [citation deep water bacteria, citation bacteria in space, citation review bacterial diversity].
+Closer to us, the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different sugars through changes in the expression of thousands of genes [find a citation here].
+
+Polymers make the most of the organic molecules that constitute a living cell.
+They are constituted of many repeated similar sub-units (monomers), but not necessarily identical.
+There is about 20 versions of amino-acids that constitute the proteins, and 5 versions of the nucleobases found in the (desoxy-)ribonucleotids that constitute RNA and DNA (the famous A, T, C, G letters and the extra U that is only found in RNA).
+The information about their sequence is stored in the form of genes on the DNA.
+The process by which this information is used in the synthesis of a functional final product (protein or RNA) is called gene expression.
+A cell can produce as much proteins as it carries genes.
+Each gene contains a target sequence that allows the gene expression machinery to recognize it and start the synthesis of its product.
+This sequence is called a promoter.
+Without any regulation, the gene expression machinery would bind on every promoter, producing every possible gene product in every condition.
+That of course does not happen.
+Some promoters are stronger than others, and would in this condition monopolize all the available gene expression machinery.
+Furthermore, there exist gene products that bind the promoter region of genes (transcription factors), stimulating or inhibiting the binding of the gene expression machinery.
+This simple mechanism, among others, allows the regulation of cell composition.
+
+The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyzes the biochemical reactions that allow the cell to perform different functions.
+For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes that convert it into reductive power, nucleotids and amino-acids~\cite{schaechter_microbe_2006}.
+These precursors are then used to produce new macromolecules, closing the loop of self-replication.
+In order to optimize the overall process, the cell must adjust the abundance of every enzyme implicated in this network of biochemical reactions.
+By doing so, the cell actually resolves a resource allocation problem, where a common wealth (precursors) has to be divided between thousand of potential candidates (proteins and other macromolecules).
+
+\subsection{Optimizing biomass production yields a competitive advantage}
+
+Like most life strategies, solving the cell composition problem lays on finding an optimum.
+How to allocate resources in order to optimize some criteria?
+Under Darwin's theory, the optimization criteria of living systems are called fitness factors, and measure how much an organism fit to a given environment.
+They take into account the ability to survive and reproduce based on the adaptation to the environment and the competition with other organisms.
+Overall, they are usually countless and measure the long-term stability of a species in an environment~\cite{dawkins_selfish_1976}.
+
+There is no consensus about what constitutes the main fitness factors of microorganisms.
+Competition for common resources seems to favor the maximization of the offspring, in the principle that a replicator with more descendants will outnumber its competitors on the long term~\cite{dawkins_selfish_1976}.
+It is however unclear how this principle translates into the physiological behavior of microorganisms.
+For instance, metabolic networks seems to either maximize ATP or biomass yield depending on the environmental condition, a result obtained by looking at the predicting power of genome-scale metabolic networks~\cite{schuetz_systematic_2007}.
+But the yield is not an universal measure, because in fluctuating or limited environments, evolution seems to favor bacteria with a high rate of nutrient assimilation, instead of a high yield of transformation even though it would favor the net biomass production of the population~\cite{frank_tradeoff_2010,maclean_tragedy_2008}, a result that has also been confirmed in yeast~\cite{schuster_maximization_2008}.
+Overall, it seems that cell optimality drift in a multidimensional space made of several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
+
+Nevertheless, numerous studies have shown that considering growth rate maximization confers a good predictive power of microorganism physiology.
+Uptake rates of \textit{Escherichia coli} growing on acetate or succinate are correctly predicted by considering its metabolic network operates in a mode that maximizes growth~\cite{edwards_silico_2001}.
+For carbon sources or strains for which the predictions initialy fail, the metabolic network evolves rapidly towards growth maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
+This result reflects that an organism cannot be adapted to all possible environment, but the short-term at which these mutants are obtained ($\sim$ 400 generations) indicate a strong selective pressure for growth in these conditions~\cite{molenaar_shifts_2009}.
+The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not solely the efficiency of the implicated enzymes is affected, but also necessarily the cell composition.
+Overall, when focusing on growth, growth rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
+
+\subsection{Growth laws are universal strategies employed by microorganisms}
+
+One of the conclusion of the preceding part is that the fitness factors that shape growth strategies are rather general, if not universally shared among microorganisms.
+If microorganism physiology is determined by these optimization criteria, we would expect to find universal regularities in the way physiology adapts to the environment.
+Are there fundamental laws that control the physiological parameters of microorganisms, or self-replication as a whole?
+
+The most emblematic effort to formalize microorganism growth by coarse-grained modelling is the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
+This empirical relationship initially states that the growth rate $R$ [\texttt{div.h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol.L\textsuperscript{-1}}], such that
+\begin{equation}
+\label{eq:monod_law}
+R = R_K \frac{C}{C_1 + C}
+\end{equation}
+with $R_K$ and $C_1$ two constants depending on the quality of the nutrient (Fig~\ref{fig:monod_law}).
+It is remarkable that the growth rate, a parameter that depends on the use and production of thousands of proteins, can be so easily predicted by a simple relation.
+What is even more remarkable is that this relation is rather universal.
+Despite some adjustments (the Droop's model being a famous one~\cite{droop_thoughts_1973}), the relation holds for most microorganisms: the growth rate increases with the abundance of the limiting nutrient, up to a point beyond which the microorganism cannot grow any faster~\cite{koch_why_1988}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[height=3cm]{./Fig/Chapter1/monod_law.eps}
+\caption{The Monod's law, reproduced from~\cite{monod_growth_1949}.
+Monod has been a pioneer in the formalization of microbial growth into fundamental coarse-grained relationships.
+This figure shows the growth rate (in number of divisions per hour) of \textit{Escherichia coli} in synthetic mediums containing different concentrations of glucose (a common carbon source).
+As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following a hyperbolic relationship described in Eq~\ref{eq:monod_law}.
+Solid line is drawn with $R_K = 1.35$ div.h\textsuperscript{-1}, and $C_1 = 0.22 \cdot 10^{-4}$ mol.L\textsuperscript{-1}. Temperature 37$^{\circ}$C.
 }
+\label{fig:monod_law}
+\end{figure}
 
-Their capacity to survive and exploit strong environmental fluctuations lay on another ubiquitous process of life as we know it: self-catalysis and gene expression.
-\textit{Short explanation on DNA, RNA, Protein and the regulations behind, that can be seen as loading "programs" depending on the environment the organism as to cope with.}
+Despite the complexity of their regulations, the abundance of every component of the cell has to follow the doubling rate of growth, or is likely to be diluted out over time~\cite{monod_growth_1949}.
+So a lot of physiological parameters (like mass of DNA, RNA and proteins) are a function of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
+These parameters were carefully recorded~\cite{bremer_modulation_1996} and are still use today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
+Recently, the topic was stirred up by the extensive work of Scott \textit{et al}, whom redefined and revived interest in the bacterial growth laws~\cite{scott_bacterial_2011}.
+They focused on the RNA/Protein ratio, a physiological parameter that was long known to vary linearly with the growth rate in microorganisms (Fig~\ref{fig:scott_rnaprot}).
+By analyzing this ratio under different environmental perturbations of the protein synthesis machinery, they built a coarse-grained model based of the proteome resource allocation~\cite{scott_emergence_2014}.
+It helped to show that whatever the details of the molecular implementation (which are sometimes different between organisms), universal growth laws emerges from the underlying principles of robustness and optimization imposed by natural selection~\cite{scott_emergence_2014}.
 
-Such a mechanism can be seen as a resource allocation problem.
-\textit{Because they can build everything from a single carbon source.
-But they have a limited protein production capacity, so choice have to be made for what need to be produced, natural selection selecting for the best strategies.}
+\begin{figure}[!h]
+\centering
+\includegraphics[height=6cm]{./Fig/Chapter1/scott_rnaprot.eps}
+\caption{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).
+For different quality of carbon sources, the RNA/Protein ratio is linearly correlated with the growth rate, a relation that old for many species of microorganisms.
+This ratio is correlated with the fraction of ribosome-affiliated proteins, and so with the relative abundance of ribosomes in the cell.
+}
+\label{fig:scott_rnaprot}
+\end{figure}
 
-General growth laws have been identified.
-\textit{The study of resource allocation led to fundamental laws like ribosome allocation or overflow metabolism, processes that can only be understood through growth rate maximization.}
+The growth strategy of overflow metabolism is another good example where a coarse-grained understanding of growth control was needed.
+It is the process by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability, a strategy that apparently waste energy~\cite{molenaar_shifts_2009} (Fig~\ref{fig:molenaar_overflow}).
+This process is so general that it even wear several names: originally called the Crabtree effect when first observed in \textit{Saccharomyces cerevisiae}~\cite{dijken_kinetics_1993} and later in \textit{E.
+coli}~\cite{vemuri_overflow_2006}, it has also been called the Warburg effect in many types of tumor cells~\cite{mckeehan_glycolysis_1982,hsu_cancer_2008}.
+Molenaar \textit{et al} showed that this phenomenon simply results from the underlying principle of growth rate maximization, in particular the trade-off between the cost of enzyme synthesis and the pathway efficiency~\cite{molenaar_shifts_2009}.
 
-Biomass production is an important fitness factor of microorganisms.
-\textit{Broadly used in a variety of study}
+\begin{figure}[!h]
+\centering
+\includegraphics[height=5cm]{./Fig/Chapter1/molenaar_overflow.eps}
+\caption{The strategy of overflow metabolism (reproduced from~\cite{molenaar_shifts_2009}, CC-BY-NC-SA 3.0).
+When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
+This somehow incoherent behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that the inefficient catabolism is cheaper to produce.
+(A) Simple view of the metabolism, with the two competitive efficient and inefficient pathways.
+(B) The switch curve from efficient to inefficient catabolism when the concentration of nutrient (and so the growth rate) increases.
+}
+\label{fig:molenaar_overflow}
+\end{figure}
 
-One of the key feature that made these works possible was to study the living system in a well-defined and reproducible state: balanced-growth.
-\textit{We describe balanced-growth, what it means for the dynamical system, and how you can reproduce it in a lab.
-We also explain why it is so convenient.}
+\subsection{Static versus dynamical perspective on growth}
 
-The mechanisms identified exist to cope with environmental fluctuations in the natural habitat of microorganisms, but are being studied in constant environments.
-\textit{Now we start to see the problem: there's a caveat in studying a system that is designed to respond to changes, if we focus on unchanging conditions.
-And we are currently unable to create such beautiful laws when taking into account dynamical conditions.}
+Growth rate maximization does a pretty good job at explaining the universal growth strategies described above.
+Since it technically favors the offspring, it seems to line up perfectly with the principles of natural selection.
+However, all the studies cited above apply this principle in a very specific state of growth, one that in fact does not correspond to the one in which these replicators have been selected, as we will argue below.
 
-While adopting a dynamical perspective on resource allocation is challenging, it could be fundamental to the life-treat-history of microorganisms.
-\textit{It is also a problem if the system we are studying poorly encountered this state during its evolution.
-In other words, "maybe bacteria are confused during steady-state".
-Hence, there is really a strong need for a mathematical framework to study resource allocation in dynamical conditions, and to know what actually happens during these growth transitions.}
+Most microbial studies are done at steady-state, where all intensive variables are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
+It is also defined as balanced growth since it requires that the components of the cell "increases by the same factor over a time interval" (hence the term "balanced")~\cite{campbell_synchronization_1957}.
+Experimentally, this state is used as a standard growing condition because it improves reproducibility, since the time-point of sampling becomes irrelevant, and the history of the culture is forgotten~\cite{schaechter_microbe_2006}.
+It can be easily achieve in the laboratory either in continuous culture where the substrate is constantly supplied, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
+This tradition was hold in mathematical modeling because considering the system at steady state reduces away the complexity of the dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells [citation FBA].
+
+\begin{figure}[!h]
+\centering
+\includegraphics[height=5cm]{./Fig/Chapter1/growth_curve.eps}
+\caption{The different states of a typical growth curve are visualized by the biomass accumulation (black thick line) and the nutrient abundance in the medium (green dashed line).
+(A) The lag phase, a variable period of time during which the organism adapt to the new medium.
+It is hardly reproducible and strongly depends on the recent history of the culture.
+(B) The steady-state or balanced-growth phase which is characterized by an exponential growth of biomass.
+It presents constant and reproducible characteristics which made it a standard for microbial growth studies.
+This phase can be extended for thousands of generations by the constant addition of fresh medium (continuous culture).
+As represented, in batch conditions nutrients are quickly depleted, which does not allow to maintain steady-state growth for a long period of time (less than 10 generations).
+(C) The stationary phase, after the depletion of the limited nutrient in the medium.
+Microorganisms usually encounter poor media and spend most of their time in stationary phase.
+In some species, it can be the siege of specific long-term resistance mechanisms like sporulation.
+}
+\label{fig:growth_curve}
+\end{figure}
+
+Balanced growth might well be convenient, it is widely admitted that microorganisms rarely encounter this state in nature since it requires a stable environment over a long period of time~\cite{schaechter_microbe_2006}.
+Microorganisms live in strongly dynamic communities, in medium where the limited nutrient is quickly depleted by the competitors when appearing, the exception being bacteria that live a parasitic or symbiotic life in bigger cells [citation genitallium, deep water worms].
+On an evolutionary perspective, optimizing the steady-state growth rate should not play any role in the replicator persistence.
+Why would in that case microorganisms be optimal for steady-state growth?
+Are we missing something by studying systems that were selected to cope with environmental variations (regulations of cell composition) in stable, unchanging conditions that probably confuse them?
+
+[I think I need more references about the need of dynamic here, it's always tricky to justify this, do we have examples were it actually helped explain/understand/find something?]
 
@@
 
-How do microorganisms adjust their resource allocation during environmental transitions ?
-\textit{As we discussed, I tried to be as specific as possible here.}
+With their beautiful simplicity, growth laws represent an exceptional point of view on microbial life strategies.
+But being established at balanced growth might have drawbacks since it is a growth state totally irrelevant to the life-trait history of microorganisms.
+How would growth laws be affected in a dynamical context?
+Can we still find universal regularities during growth states more relevant to the natural habitat of microorganisms?
+
+Since we lose the inherent advantages of working at steady state, we must confine our problem to keep it tractable.
+We then focus on the fundamentals behind resource allocation (the proteome composition growth law), especially the regulation of the abundance of gene expression machinery during the transition between two different growth media.
+
+On the theoretical side, we assess if there exist a universal optimal way to allocate cell resources during growth transitions.
+On the experimental side, we monitor the abundance of key macromolecular components in order to get an insight that could be compared with the theoretical predictions.
 
@@
 
-(the work on dynamic, Van der Berg, Alon, Levy, etc...
-the work on steady-state has normally been described previously.
-Or, I don't know, maybe I can shortly describe it in the context and extensively describe it here.)
+\subsection{Modeling growth of microorganisms}
+
+What is modeling?
+\textit{Why is it important?
+Why so many efforts to model life?
+Model are based on 4 principles: question, scope, granularity and mathematical framework (quantitative/qualitative, probabilistic/deterministic, non-linear/linear, dynamic/equilibrium, etc).}
+
+(In each of the following part I point out what is relevant to our question)
+
+Constraint-based modeling.
+\textit{FBA and all the derivatives that appeared these years.
+The main successes and the main drawbacks are described, like the importance of the objective function, the problem of taking several layers into account, the steady-state assumption, and the example of overflow metabolism.
+They have a large scope, a detailed granularity but can do so by abstracting away the mathematical difficulties.}
+
+Big kinetic models.
+\textit{Especially the work of Klumpp, Kremling, Sauer, etc, also the whole model of genitallium.
+Biggest problem is the lack of parameter values, and some huge trouble with identifiability.
+They have medium to large scope (increasing), highly detailed granularity and deal with the mathematical complexity via the computational power via the use of huge quantity of data.}
+
+Coarse-grained, proof of principle models.
+\textit{Examples are the work of Scott, Hwa, Teusink, van der Berg.
+They focus on the question, have usually a large scope but a very coarse-grained granularity.
+They can work as intuition pump and actually be more useful into given a new way to see previous data (example of Scott/Hwa and Teusink).
+They can exploit complicated math theories (van der Berg).}
+
+\subsection{Dynamical monitoring of microorganisms}
+
+[Talk about the works in dynamic, and how some questions cannot be answered at steady-state.
+Levy is a good example. Alon is another one. Look also at the last publication of Bremer and Dennis (medium-dependent control).]
+
+(the mother machine etc)
 

### 2016-09-05 Chapters/theory.tex
Context and problem statement first version, sent to Hidde and Hans

@@
 
-\selectlanguage{french}
-\section*{Résumé du chapître \thechapter : [titre français]}
-dans cette section blablabla
-\selectlanguage{english}
-
-{\Large
-\textbf\newline{\textbf{}}
-}
-
-\section{Abstract}
+\textit{Rather than propose a new theory or unearth a new fact, often the most important contribution a scientist can make is to discover a new way of seeing old theories or facts.} -- Richard Dawkins, \textit{The Selfish Gene}, preface to 1989 edition.
+
+\section*{Summary of chapter \thechapter}
 Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
@@
 
+\selectlanguage{french}
+\section*{Résumé du chapître \thechapter : [titre français]}
+dans cette section blablabla
+\selectlanguage{english}
+
 %%%%%% CORE OF THE ARTICLE %%%%%
@@
 
-\section{Self-replicator model of resource allocation}
+\subsection{Self-replicator model of resource allocation}
 \label{sec:model}
@@
 
-\section{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
+\subsection{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
 \label{sec:growthlaws}
@@
 
-\section{Biomass maximization as an optimal control problem}
+\subsection{Biomass maximization as an optimal control problem}
 \label{sec:optimalcontrolproblem}
@@
 
-\section{Solution of the optimal control problem}
+\subsection{Solution of the optimal control problem}
 \label{sec:optimalcontrolsolution}
@@
 
-\section{Simple feedback control strategies: exploiting information on nutrients or precursors}
+\subsection{Simple feedback control strategies: exploiting information on nutrients or precursors}
 \label{sec:steadystate}
@@
 
-\section{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
+\subsection{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
 \label{sec:strategies}

### 2016-09-05 abstract.tex
Context and problem statement first version, sent to Hidde and Hans

\selectlanguage{english}
\begin{abstract}
Growth is one of the most fundamental processes of life, and is highly optimized in microorganisms.
It relies on the transformation of matter and energy from the environment into diverse organic structures.
Decades of studies have shown that despite an extreme microbial diversity, this transformation follows general regularities, forming the so-called microbial growth laws.
But those works focused on steady-state growth, despite this condition being irrelevant to the natural growth condition of microorganisms.

Do these laws still hold when considering a dynamical environment?
What does it mean to be optimal during an environmental transition?

This project aims to lay the foundation of dynamical growth laws, more in adequation with the natural habitats of microorganisms and the selective pressures that shaped their current forms.
We develop an intuitive framework by using a self-replicator model that encapsulates the main features of growth through nutrient assimilation.
We formulate the resource allocation problem that arises as an optimal control problem in which the cell has to allocate its resources between gene expression machinery and metabolism.
By applying the Pontryagin's Maximum Principle, the model predicts the optimal dynamical behavior to be a bang-bang allocation between gene expression machinery and metabolism.
Despite not having been observed yet, we show that such behaviors are conceptually feasible with the known regulations of the enterobacteria \textit{Escherichia coli}, in particular the ppGpp system that regulates ribosomal abundance.

In order to find such patterns in real cells, we make a strain displaying fluorescent ribosomes.
We observe \textit{E. coli} cells in a microfluidic device and monitor their fluorescence during growth transitions.
Using state-of-the-art techniques of parameter estimation, we can reconstruct the state of the cellular resource allocation controler with a high temporal precision.
Even if preliminary, our experimental results seem to confirm the bang-bang singular nature of resource allocation during a growth transition, as predicted by our model.

Our work might play an important role in the formulation of dynamical fundamental growth laws.
It also has application for biotechnological purposes, since microorganisms and in particular \textit{E. coli}, are more and more used in health and energy industries to produce valuable elements.
\end{abstract}

\selectlanguage{french}
\begin{abstract}
La croissance et la reproduction sont des mécanismes fondamentaux du vivant.
Généralement couplés chez les micro-organismes, ces deux processus reposent sur la transformation des ressources de l'environnement (matière et énergie) en nouvelles structures organiques.
Malgré l'extrême diversité des micro-organismes, certaines caractéristiques clés du contrôle de la croissance semblent être largement répandues, au point que des lois de croissance ont pu être érigées.
Ces dernières montrent que des mécanismes encore plus fondamentaux dictent la répartition des ressources dans les différents sous-systèmes de la cellule, généralement dans un souci d'optimisation de son taux de croissance.
Mais ces lois ont été établies uniquement lors d'une croissance à l'état stationnaire, c'est-à-dire lorsque l'environnement et le taux de croissance ne varient pas dans le temps.
Ces conditions, même si elles peuvent être reproduites en laboratoire, n'ont rien à voir avec les conditions de vie dans lesquelles les organismes ont évolué durant des milliards d'années.

Ces lois sont-elles toujours valable en environnement dynamique ?
Comment traduire les concepts d'optimalité dans de tels environnements ?

Le but de ce projet est de poser les bases de nouvelles lois de croissance étendues en environnement dynamique, c'est-à-dire plus proche du milieu réel de vie, et tenant mieux compte des pressions de sélection qui ont donné lieu à ces organismes.
Notre cadre d'étude s'articule autour d'un modèle d'auto-réplicateur qui prend en compte les caractéristiques principales de la croissance par assimilation de nutriments.
De ce cadre émerge un problème d'allocation des ressources entre la machinerie d'expression génique et le métabolisme, que nous formulons comme un problème de contrôle optimal.
Le modèle prédit ainsi via l'application du Principe de Maximisation de Pontryagin que le meilleur comportement possible lors d'une transition est une allocation de type bang-bang, c'est-à-dire lors de laquelle les ressources vont par alternance soit uniquement dans la machinerie, soit uniquement dans le métabolisme.
Malgré l'absence d'observation claire d'une tel phénomène, nous montrons qu'un tel comportement est conceptuellement possible à partir des régulations connues de l'enterobactérie \textit{Escherichia coli}, notamment le système ppGpp qui régule l'abondance des ribosomes.

Afin d'observer de tester ces prédictions, nous avons ingéniéré une souche de \textit{E. coli} arborant des ribosomes fluorescents.
Nous avons observé ces cellules dans un système de microfluidique et enregistré leur fluorescence au cours de transitions de croissance.
En utilisant les dernières méthodes d'identification paramétrique, nous pouvons reconstruire l'état du contrôleur d'allocation des ressources avec une très bonne précision temporelle.
Même s'ils ne sont que préliminaires, ces résultats expérimentaux semblent confirmer la nature bang-bang de l'allocation lors des transitions de croissance chez \textit{E. coli}.

Notre travail peut jouer un rôle clé dans la formalution de nouvelles lois de croissances, étendues en environnement dynamique.
Il a également de nombreuses applications en biotechnologie, dans la mesure où les micro-organismes et plus particulièrement \textit{E. coli} sont très largement utilisés par de nombreuses industries pour produire des molécules valorisables.
\end{abstract}
\selectlanguage{english}

### 2016-09-05 titlepage.tex
Context and problem statement first version, sent to Hidde and Hans

--- 2016-06-06 titlepage.tex
@@
 
-%\Sethpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture horizontalement (valeur par défaut est -5.5mm)
-%\Setvpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture verticalement (valeur par défaut est -15.5mm)
+%\Sethpageshift{-8mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture horizontalement (valeur par défaut est -5.5mm)
+%\Setvpageshift{0mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture verticalement (valeur par défaut est -15.5mm)
 
@@
 %\Grade{}         %%optionnel : à décommenter et à renseigenr si vous voulez changer le grade
-\Specialite{Biologie cellulaire }
+\Specialite{Biologie cellulaire}
 \Arrete{7 Ao\^ut 2006}
-\Auteur{Nils Giordano}
-\Directeur{Johannes Geiselmann}
-\CoDirecteur{Hidde de Jong}    %%optionnel : à décommenter et à renseigenr si présence d'un Co-directeur de thèse
+\Auteur{Nils GIORDANO}
+\Directeur{Johannes GEISELMANN}
+\CoDirecteur{Hidde DE JONG}
 \Laboratoire{Laboratoire Interdisciplinaire de Physique}
 \EcoleDoctorale{Ecole doctorale Chimie et Science du Vivant}         
-\Titre{Growth control in microorganisms: theoretical and experimental study by mean of self-replicator models}
-%\Soustitre{}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
+\Titre{Microbial growth control in shifting environments}
+\Soustitre{Theoretical and experimental study of resource allocation in \textit{Escherichia coli}}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
 \Depot{}
@@
 % Commande pour création de nouvelles catégories dans le jury:
-
 %\UGTNewJuryCategory{...NomDeLaCategorie...}{...Definition...}
-
 % Exemple \UGTNewJuryCategory{UGTFamille}{Membre de la famille} que nous ajoutons dans la commande \Jury ci-dessous sous la forme \UGTFamille{Jean Rousseau}{(...titre_et_affiliation...s'il_y_en_a...)}
@@
 \Jury{
-\UGTPresident{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
+\UGTPresident{M, Olivier BERNARD}{Chargé de recherche, Inria, Nice -- Sophia-Antipolis (à confirmer pour président)}     %% 2ème examinateur
 %\UGTPresidente{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
 
-\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% 1er rapporteur
-%\UGTRapporteur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}      %% second rapporteur
+\UGTRapporteur{M, Matthew SCOTT}{Associate Professor, Department of Applied Mathematics, University of Waterloo (CA)}      %% 1er rapporteur
+\UGTRapporteur{M, Guy-Bart STAN}{Reader, Department of Bioengineering, Imperial College London (U.K.)}      %% second rapporteur
 
-\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% 1er examinateur
-%\UGTExaminateur{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}     %% second examinateur
-%\UGTExaminatrice{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}    %% 3ème examinateur
+\UGTExaminateur{M, Matthieu JULES}{Chargé de recherche, Inra, Grignon}     %% 1er examinateur
+\UGTExaminatrice{Mme, Irina MIHALCESCU}{Maître de conférence, Université Grenoble Alpes, LIPhy}    %% 3ème examinateur
 
-\UGTDirecteur{M, Johannes Geiselmann}{Professeur Université Joseph Fourier, LIPhy, Grenoble}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde de Jong}{Directeur de recherche Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
+\UGTDirecteur{M, Johannes GEISELMANN}{Professeur, Université Grenoble Alpes, LIPhy}       %% Directeur de thèse
+\UGTCoDirecteur{M, Hidde DE JONG}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
 
-\UGTInvite{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
-\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
+\UGTInvite{M, Jean-Luc GOUZÉ}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}
+%\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
 }

### 2016-10-04 Chapters/introduction.tex
Corrections from Hidde on 1.1

--- 2016-07-25 Chapters/introduction.tex
@@
 
-\section{Context and motivations}
+\section{Context}
 
-Most part of life beauty and diversity is hidden from us in the form of microorganisms.
-These small factories are optimized to grab and use energy from their environment, in the form of organic matter, highly reactive chemicals, or even photons [citation trophic].
-In that sense, microorganisms exhibit the most fundamental features of life.
-They live almost everywhere, sometimes in the strangest possible places [citation space].
-This nature of ubiquitous, extremely optimized living systems makes them perfect to unearth the fundamental forces that shape life as we know it.
+\subsection{Self-replication is a resource allocation problem}
 
-They also managed to become the thoughest of all the living systems.
-\textit{Their diversity allow them to colonize every biotope, but remarkably, a single species can live in broad conditions, growing on different carbon sources or resist to really harmful environments (even space!).
+At every moment of our lives, we are surrounded by billions of microscopic organisms.
+Sometimes beneficial, sometimes harmful, these small factories collect and use energy from their environment, in the form of organic matter, highly reactive chemicals, or light [citation trophic].
+While most cells in our body are kept in a stable and friendly micro-environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to a strongly variable environment.
+Microbes are found almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life [citation deep water bacteria, citation bacteria in space, citation review bacterial diversity].
+Closer to us, the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different sugars through changes in the expression of thousands of genes\cite{zimmer_microcosm:_2009}.
+
+Polymers are the most abundant organic molecules that constitute a living cell.
+They consist of many repeated similar sub-units (monomers).
+There are about 20 different amino acids in the proteins, and 5 different nucleobases in the (desoxy-)ribonucleotids that constitute RNA and DNA (the famous A, T, C, G letters and the extra U only found in RNA).
+The information about their sequence is stored in the form of genes on the DNA.
+The process by which this information is used in the synthesis of a functional final product (protein or RNA) is called gene expression.
+Each gene contains a target sequence that allows the gene expression machinery to recognize it and start the synthesis of its product.
+This sequence is called a promoter.
+If every promoter were the same, all genes would be expressed at the same level.
+That of course does not happen.
+In practice, some promoters are stronger than others, and attracts more of the available transcriptional machinery.
+Furthermore, there exist gene products that bind the promoter region of genes (transcription factors), stimulating or inhibiting the promoter affinity, hence the gene expression.
+This simple mechanism, among others, allows the regulation of cell composition.
+
+The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyzes the biochemical reactions allowing the cell to perform different functions.
+For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reductive power, nucleotids and amino acids~\cite{schaechter_microbe_2006}.
+These precursors are then used to produce new macromolecules, closing the loop of self-replication.
+In order to optimize the overall process, the cell must adjust the abundance of every enzyme involved in this network of biochemical reactions.
+By doing so, the cell actually resolves a resource allocation problem, where a common wealth (precursors) has to be distributed over thousands of potential beneficiaries (proteins and other macromolecules).
+
+\subsection{Optimizing biomass production yields a competitive advantage}
+
+What is the optimal allocation of precursors for a living cell?
+Under Darwin's theory, the optimization criteria of living systems are called fitness factors, and measure to which extend an organism fits into a given environment.
+Fitness factors take into account the ability to survive and reproduce based on the adaptation to the environment and the competition with other organisms.
+Whether fitness factors apply at the species, the population, the organism, or even the gene level is still highly debated~\cite{dawkins_selfish_1976}.
+This makes their clear identification highly difficult, even for organisms living an apparently simple life.
+
+For this reason, there is no consensus about what constitute the main fitness factors of microorganisms.
+Competition for common resources seems to favor the maximization of offspring, in the sense that a replicator with more descendants will outnumber its competitors in the long term~\cite{dawkins_selfish_1976}.
+It is unclear, however, how this principle translates into the physiological behavior of microorganisms.
+For instance, metabolic networks seems to either maximize ATP or biomass yield depending on the environmental condition, a result obtained by looking at the predicting power of genome-scale metabolic networks~\cite{schuetz_systematic_2007}.
+But, yield is not an universal fitness factor, because in fluctuating or limited environments, evolution seems to favor bacteria with a high rate of nutrient assimilation, instead of a high yield of transformation even though it would favor the net biomass production of the population~\cite{frank_tradeoff_2010,maclean_tragedy_2008}, a result that has also been confirmed in yeast~\cite{schuster_maximization_2008}.
+Overall, it seems that cell optimality drifts in a multidimensional space and involves making several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
+
+Nevertheless, numerous studies have shown that considering growth rate maximization confers a good predictive power of microorganism physiology.
+Uptake rates of \textit{Escherichia coli} growing on acetate or succinate are correctly predicted by assuming that its metabolic network operates in a mode that maximizes growth~\cite{edwards_silico_2001}.
+This result is not general and can initially fail for other carbon sources or strain~ \cite{ibarra_escherichia_2002}.
+But even in those cases, the metabolic networks evolve rapidly towards growth maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
+The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not solely the efficiency of the enzymes is affected, but also as a consequence the cellular composition.
+Overall, when focusing on growth, growth rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
+
+\subsection{Growth laws are universal strategies employed by microorganisms}
+
+One of the conclusions of the preceding section is that growth rate maximization is a rather general principle of microbial life.
+Would microbial physiology be determined by this criteria, we could expect to find universal regularities in the way physiology adapts to the environment.
+Are there fundamental laws that control the physiological parameters of microorganisms, or self-replication as a whole?
+
+The most emblematic effort to formalize microorganism growth by coarse-grained modelling is the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
+This empirical relationship states that the growth rate $\mu$ [\texttt{div.h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol.L\textsuperscript{-1}}], such that
+\begin{equation}
+\label{eq:monod_law}
+\mu = \mu_K \frac{C}{K_C + C}
+\end{equation}
+with $\mu_K$ and $K_C$ two constants depending on the quality of the nutrient (Fig~\ref{fig:monod_law}).
+It is remarkable that the growth rate, a parameter that depends on the use and production of thousands of proteins, can be so easily predicted by a simple relation.
+What is even more remarkable is that this relation is rather universal.
+Despite some adjustments (Droop model being a well-known example~\cite{droop_thoughts_1973}), the relation holds for most microorganisms: the growth rate increases with the abundance of the limiting nutrient, up to a point beyond which growth cannot be any faster~\cite{koch_why_1988}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[height=3cm]{./Fig/Chapter1/monod_law.eps}
+\caption{Monod law, reproduced from~\cite{monod_growth_1949}.
+Monod has been a pioneer in the formalization of microbial growth into fundamental coarse-grained relationships.
+This figure shows the growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
+As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following a hyperbolic relationship described in Eq~\ref{eq:monod_law}.
+Solid line is drawn with $\mu_K = 1.35$ div.h\textsuperscript{-1}, and $K_C = 0.22 \cdot 10^{-4}$ mol.L\textsuperscript{-1}.
 }
+\label{fig:monod_law}
+\end{figure}
 
-Their capacity to survive and exploit strong environmental fluctuations lay on another ubiquitous process of life as we know it: self-catalysis and gene expression.
-\textit{Short explanation on DNA, RNA, Protein and the regulations behind, that can be seen as loading "programs" depending on the environment the organism as to cope with.}
+The production of every component of the cell is under complex regulations.
+But they are all linked to the growth rate, since the production has to compensate for the dilution caused by growth~\cite{monod_growth_1949}.
+Many physiological parameters (like mass of DNA, RNA and proteins) are then functions of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
+These functions were carefully measured~\cite{bremer_modulation_1996} and are still used today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
+Recently, the topic was brought to the fore by the work of Scott \textit{et al}, whom redefined and revived interest in the bacterial growth laws~\cite{scott_bacterial_2011}.
+They focused on the RNA/protein ratio, a physiological parameter that was long known to vary linearly with the growth rate in microorganisms (Fig~\ref{fig:scott_rnaprot}).
+By analyzing this ratio under different environmental perturbations of the protein synthesis machinery, they built a coarse-grained model describing proteome resource allocation~\cite{scott_emergence_2014}.
+It has allowed to show that, whatever the details of the molecular implementation (differing between organisms), universal growth laws emerges from the underlying principles of robustness and optimization imposed by natural selection~\cite{scott_emergence_2014}.
+% detail this
 
-Such a mechanism can be seen as a resource allocation problem.
-\textit{Because they can build everything from a single carbon source.
-But they have a limited protein production capacity, so choice have to be made for what need to be produced, natural selection selecting for the best strategies.}
+\begin{figure}[!h]
+\centering
+\includegraphics[height=6cm]{./Fig/Chapter1/scott_rnaprot.eps}
+\caption{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).
+For different carbon sources more or less rich, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
+This ratio is correlated with the fraction of ribosome-affiliated proteins, and therefore with the relative abundance of ribosomes in the cell.
+}
+\label{fig:scott_rnaprot}
+\end{figure}
 
-General growth laws have been identified.
-\textit{The study of resource allocation led to fundamental laws like ribosome allocation or overflow metabolism, processes that can only be understood through growth rate maximization.}
+Coarse-grained models have also been helpful to understand the process by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability.
+Since nutrients are not used at their full potential given the environmental conditions, this widespread growth strategy apparently leads to energy waste, ~\cite{molenaar_shifts_2009} (Fig~\ref{fig:molenaar_overflow}).
+Originally called the Crabtree effect when first observed in \textit{Saccharomyces cerevisiae}~\cite{dijken_kinetics_1993} and overflow metabolism when later found in \textit{E.
+coli}~\cite{vemuri_overflow_2006}, it has also been called the Warburg effect in many types of tumor cells~\cite{mckeehan_glycolysis_1982,hsu_cancer_2008}.
+Molenaar \textit{et al} showed that this phenomenon simply results from the underlying principle of growth rate maximization, in particular the trade-off between the cost of enzyme synthesis and the pathway efficiency~\cite{molenaar_shifts_2009}.
+Like traders, bacteria distribute their resources between different beneficiaries, and tend to do so in a way that maximizes returns on investments.
 
-Biomass production is an important fitness factor of microorganisms.
-\textit{Broadly used in a variety of study}
+\begin{figure}[!h]
+\centering
+\includegraphics[height=5cm]{./Fig/Chapter1/molenaar_overflow.eps}
+\caption{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.
+When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
+This paradoxical behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that pathways for inefficient catabolism are shorter and therefore cheaper to produce.
+(A) Simple view of microbial metabolism, with the two competitive efficient and inefficient pathways.
+(B) The switching from efficient to inefficient catabolism when the concentration of nutrients, and hence the growth rate, increases.
+}
+\label{fig:molenaar_overflow}
+\end{figure}
 
-One of the key feature that made these works possible was to study the living system in a well-defined and reproducible state: balanced-growth.
-\textit{We describe balanced-growth, what it means for the dynamical system, and how you can reproduce it in a lab.
-We also explain why it is so convenient.}
+\subsection{Static versus dynamical perspective on growth}
 
-The mechanisms identified exist to cope with environmental fluctuations in the natural habitat of microorganisms, but are being studied in constant environments.
-\textit{Now we start to see the problem: there's a caveat in studying a system that is designed to respond to changes, if we focus on unchanging conditions.
-And we are currently unable to create such beautiful laws when taking into account dynamical conditions.}
+Growth rate maximization does a pretty good job at explaining the universal growth strategies described above.
+By maximizing offspring, it seems to comply perfectly with the principles of natural selection cited earlier.
+However, all the studies cited above apply this growth rate maximization in a very specific growth scenario, one that in fact does not correspond to the environmental conditions in which microorganisms have been selected, as we will argue below.
 
-While adopting a dynamical perspective on resource allocation is challenging, it could be fundamental to the life-treat-history of microorganisms.
-\textit{It is also a problem if the system we are studying poorly encountered this state during its evolution.
-In other words, "maybe bacteria are confused during steady-state".
-Hence, there is really a strong need for a mathematical framework to study resource allocation in dynamical conditions, and to know what actually happens during these growth transitions.}
+Most microbial studies have been done at steady state, where all intensive properties of the cell are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
+It means that the properties independent of the cell volume or mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
+It requires that the components of the cell "increases by the same factor over a time interval", which have motivated the creation of the term balanced growth~\cite{campbell_synchronization_1957}.
+Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that the time-point of sampling becomes irrelevant~\cite{schaechter_microbe_2006}.
+It can be easily achieved laboratory conditions either in continuous culture, where the substrate is continually supplied, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
+This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells [citation FBA].
+
+\begin{figure}[!h]
+\centering
+\includegraphics[height=5cm]{./Fig/Chapter1/growth_curve.eps}
+\caption{The different phases of a typical growth curve are visualized by the accumulation of biomass (black thick line) and the nutrient abundance in the medium (green dashed line).
+(A) The lag phase, a variable period of time during which the organism adapt to the new medium.
+It is hardly reproducible and strongly depends on the recent history of the culture.
+(B) The steady-state (or balanced-growth) phase which is characterized by an exponential growth of biomass.
+It presents constant and reproducible characteristics which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
+This phase can be extended for thousands of generations by the constant addition of fresh medium (continuous culture).
+As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
+(C) The stationary phase, after the depletion of the limited nutrient in the medium.
+In environmental conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase.
+Some species which have evolved long-term resistance mechanisms like sporulation can survive harsh conditions when in stationary phase.
+}
+\label{fig:growth_curve}
+\end{figure}
+
+Although balanced growth is convenient from an experimental and theoretical point of view, it is widely admitted that microorganisms rarely encounter this state in nature since it requires a stable environment over a long period of time~\cite{schaechter_microbe_2006}.
+Microorganisms live in strongly dynamic communities, in medium where the limited nutrient is quickly depleted by the competitors when available\footnote{an exception being bacteria that live a parasitic or symbiotic life in bigger cells [citation Micoplasma genitallium, Giant tube worm]}.
+With that in mind, a microorganism aptness to optimize its steady-state growth rate should not play any role for its persistence.
+Why would in that case microorganisms be optimal for steady-state growth?
+Are we missing something by studying systems that were selected to cope with environmental variations (regulations of cell composition) in stable, unchanging conditions that probably confuse them?
 

### 2016-11-23 Chapters/experiments.tex
Remove commentary on section 3

@@
-\chapter{Monitoring of Gene Expression Machinery during growth transitions}
-
-Missing results as of today:
-\begin{itemize}
-\item manual segmentation of one well --> mean RFU per pixel per cell (2-3 days in the best case, it could be difficult to export schnitzcells data)
-\item semi-manual tracking of cells based on segmentation (hard to evaluate, will depend on schnitzcells performances, not optimistic)
-\item estimation of $\alpha$ and $\mu$ over time (2-3 days for calculation, 2-3 days for implementation). I need tracking for $\mu$, maybe not for $\alpha$
-\item make another experiments glu-sugar-glu (glycerol?), depends on Irina
-\item analyze this experiments (repeat steps above, will probably be faster if we manage to work everything out, but still takes time because it's manual)
-\item model modification to take into account the new data (5 days)
-\end{itemize}
-
-What could go wrong: everything. In particular:
-\begin{itemize}
-\item segmentation results could be impossible to export
-\item tracking might not work (schnitzcells not designed to handle massive cell disappearance)
-\item $\mu$ might be too noisy due to pixel size (note: image interpolation before segmentation might be a solution)
-\item what we saw on the rough analysis might just be an artefact (as Hans usually says "When it works the first time, something went horribly wrong.")
-\item the second experiment might not work (poor preculture, device plugged, out of focus overnight, medium contamination...)
-\end{itemize}
-
-Would everything go right, the chapter/article would be as follow.
+\chapter{Monitoring Gene Expression Machinery abundance during a glucose upshift}
 
@@
 
-\subsection{Data analysis}
+\subsection{Cell segmentation}
+
+We used this and that to analyze the microscopic images.
+We applied the following corrections.
+
+\subsection{Kalman smoothing}
 

### 2016-11-23 Chapters/introduction.tex
Add 3 paragraphs in related work about what is modeling for" "

@@
 
-What is modeling?
-\textit{Why is it important?
-Why so many efforts to model life?
-Model are based on 4 principles: question, scope, granularity and mathematical framework (quantitative/qualitative, probabilistic/deterministic, non-linear/linear, dynamic/equilibrium, etc).}
+My recent experience as a teacher taught me that, as of today, a lot of students still choose to study biology in the hope to find a math-free job in science.
+They fall from the clouds when you tell them modeling is an important part of a biologist's life.
+This misconception usually comes from a bad understanding of what is modeling.
+In its general sense, a model depicts and simplifies reality.
+Maps, sketches, or pictures satisfy this definition, as do graphs, sets of equations, or even the DNA sequences I store in a text-file on my computer.
+Overall, depicting and simplifying reality makes you a modeler, not so much drawing equations on a board.
+But this does not change the fact that mathematics are incredible tools for these purpose~\cite{servedio_not_2014,mcgill_calm_2013}.
 
-(In each of the following part I point out what is relevant to our question)
+What makes mathematical modeling so useful in microbial growth studies?
+Microbial physiology comes out from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation.
+They usually occur on a short time scale as compared with the organism life span.
+They nevertheless have been (and still are) selected by evolutionary processes that occur on long time scales.
+These physical limitations strongly impede a full empirical study of growth processes, and it's sometimes hard to test the most simple of our verbal hypothesis.
+
+Since they are designed to abstract away complexity, mathematical models are great tools to deal with such limitations.
+Overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
+Growth is abstracted into a world of rigor and rules, where logical reasoning can be applied, which allow to test hypothesis~\cite{servedio_not_2014}.
+Predictions can be made, "unpacked" into our real world, and confronted with experimental testing~\cite{mcgill_calm_2013}.
+Such approach has been successfully applied these decades, creating a diversity a modeling framework which are relevant to growth studies.
 

### 2016-11-24 Chapters/introduction.tex
Almost finished 1.3.1, need more details on literature results and some reformulation

@@
 \section{Context}
+\label{sec:context}
 
@@
 Overall, depicting and simplifying reality makes you a modeler, not so much drawing equations on a board.
-But this does not change the fact that mathematics are incredible tools for these purpose~\cite{servedio_not_2014,mcgill_calm_2013}.
+But this does not change the fact that mathematics are incredible tools for these purposes~\cite{servedio_not_2014,mcgill_calm_2013}.
 
@@
 They nevertheless have been (and still are) selected by evolutionary processes that occur on long time scales.
-These physical limitations strongly impede a full empirical study of growth processes, and it's sometimes hard to test the most simple of our verbal hypothesis.
+These physical limitations strongly impede a full empirical study of growth processes, making sometimes hard to test the most simple verbal hypothesis.
 
 Since they are designed to abstract away complexity, mathematical models are great tools to deal with such limitations.
-Overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
-Growth is abstracted into a world of rigor and rules, where logical reasoning can be applied, which allow to test hypothesis~\cite{servedio_not_2014}.
-Predictions can be made, "unpacked" into our real world, and confronted with experimental testing~\cite{mcgill_calm_2013}.
-Such approach has been successfully applied these decades, creating a diversity a modeling framework which are relevant to growth studies.
+The overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
+When working on a model, growth is abstracted into a world of clearing stated rules, where logical reasoning can be applied~\cite{servedio_not_2014}.
+Predictions can be made, "unpacked" into our real world, and confronted with experimental testing.
+In that sense, modeling is at the core of the scientific method~\cite{mcgill_calm_2013}.
+This could explain why mathematical modeling has been extensively applied to microbial growth studies, creating a diversity a modeling framework.
 
-Constraint-based modeling.
-\textit{FBA and all the derivatives that appeared these years.
-The main successes and the main drawbacks are described, like the importance of the objective function, the problem of taking several layers into account, the steady-state assumption, and the example of overflow metabolism.
-They have a large scope, a detailed granularity but can do so by abstracting away the mathematical difficulties.}
+For an increasing number of microorganisms, we are now able to draw quasi-exhaustive maps of their inner metabolic reactions.
+This has been made possible by the advances on genome reconstruction, which are easier and cheaper to accomplish every passing year.
+But making sense of these data proved to be even more challenging.
+Which reactions are important, and in which conditions?
+How will a metabolic network react to perturbations?
+Can we predict its evolution in a given environment?
+When dealing with genome-scale metabolic networks, one is challenged by the number of metabolites and possible reactions (e.g.
+the most recent genome-scale reconstruction of \textit{E.
+coli} metabolism contains "1366 genes, 2251 metabolic reactions, and 1136 unique metabolites"~\cite{orth_comprehensive_2011}).
+More importantly, while progress in genomic made possible such reconstructions, measuring precise \textit{in-vivo} kinetic parameters for all these reactions still represents a technological bottleneck~[citation?].
+Nevertheless, a modeling framework is specifically designed to deal with such limitations: constraint-based modeling.
+This framework abstracts away the unknown kinetics and represents the metabolic maps as a flux map on which physico-chemical constraints can be applied, e.g.
+steady-state, compartmentalization, mass conservation, molecular crowding, and thermodynamic directionality~\cite{ebrahim_cobrapy_2013}.
+Candidates for the flux distribution are obtained by optimizing the system towards an objective function, which usually represents growth rate for microorganisms.
+One can find a lot of derivatives of this framework, but its the Flux Balance Analysis that renewed interest into it [citation review ?].
+\textbf{[Need some key results that were obtained, or this paragraph makes no sense].}
+Note that such a level of abstraction necessarily has its drawbacks, e.g.
+the definition of the objective function, the difficulty to work in dynamic or implement information on the metabolic regulations [citation?].
 
-Big kinetic models.
-\textit{Especially the work of Klumpp, Kremling, Sauer, etc, also the whole model of genitallium.
+The ever-growing available data on metabolic networks can also be used to make large kinetic models.
+For the reasons we stated above, we expect these models to be inherently limited in scope.
+It means that the modeler has to be extremely careful since he or she injects its knowledge about what might be important in the system, for the question he or she tries to solve.
+They are however more flexible, in the sense that one can easily make interplay different level of regulations.
+\textit{[Present the work of Klumpp, Kremling, Sauer, whole model of genitallium, etc]
+}\textit{Especially the work of Klumpp, Kremling, Sauer, etc, also the whole model of genitallium.
 Biggest problem is the lack of parameter values, and some huge trouble with identifiability.
@@
 
-Coarse-grained, proof of principle models.
-\textit{Examples are the work of Scott, Hwa, Teusink, van der Berg.
-They focus on the question, have usually a large scope but a very coarse-grained granularity.
-They can work as intuition pump and actually be more useful into given a new way to see previous data (example of Scott/Hwa and Teusink).
-They can exploit complicated math theories (van der Berg).}
+Another approach has shown successful in understanding some aspects of growth behaviors of microorganisms.
+By applying a high level of abstraction on how the cell functions, it becomes easy to compare data about different species and to uncover general evolutionary principles that are independent of the specific implementation in each organism.
+Best examples of this approach are the empirical growth laws of ribosomal abundance~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}, or the proof-of-concept explanation of overflow metabolism~\cite{molenaar_shifts_2009} that were described in section~\ref{sec:context} of this introduction.
+It is also the preferred approach when one wants to broaden its access to mathematical tools~\cite{vandenberg_optimal_1998}, most of them unhelpful with the high dimensionality of biological models.
+As we will describe in section~\ref{sec:approach}, this represents the modeling strategy that was used in this thesis.
 
@@
 \section{Approach}
+\label{sec:approach}
 

### 2016-11-28 Chapters/introduction.tex
Very rough draft of relevant work experimental part

@@
 
-\textit{There's an infinity of things that have never been done before, and most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology}~\cite{fox_how_2016}
+\textit{There's an infinity of things that have never been done before, and most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology (2016)}~\cite{fox_how_2016}
 
@@
 
-(passive or active voice?)
+This chapter presents... (approx. 1 page)
 
@@
 
-Dans cette section...
+Dans cette section... (env. 1 page)
 \selectlanguage{english}
@@
 The overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
-When working on a model, growth is abstracted into a world of clearing stated rules, where logical reasoning can be applied~\cite{servedio_not_2014}.
+When working on a model, growth is abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014}.
 Predictions can be made, "unpacked" into our real world, and confronted with experimental testing.
@@
 Best examples of this approach are the empirical growth laws of ribosomal abundance~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}, or the proof-of-concept explanation of overflow metabolism~\cite{molenaar_shifts_2009} that were described in section~\ref{sec:context} of this introduction.
-It is also the preferred approach when one wants to broaden its access to mathematical tools~\cite{vandenberg_optimal_1998}, most of them unhelpful with the high dimensionality of biological models.
+It is also the preferred approach when one wants to broaden its access to mathematical tools~\cite{vandenberg_optimal_1998}, most of them being unhelpful with the high dimensionality of biological models.
 As we will describe in section~\ref{sec:approach}, this represents the modeling strategy that was used in this thesis.
 
-\subsection{Dynamical monitoring of microorganisms}
+\subsection{Measuring growth of microorganisms}
 
-[Talk about the works in dynamic, and how some questions cannot be answered at steady-state.
-Levy is a good example. Alon is another one. Look also at the last publication of Bremer and Dennis (medium-dependent control).]
+Models have to be tested with data.
+Measuring growth = measuring biomass accumulation, over time, and get relevant information on the cell inner state.
+We have a lot of data on this.
 
-(the mother machine etc)
+Population measurement is easy to implement, and allow to measure metabolites.
+But small unsynchronized effects can be hidden since we mostly work with the mean.
+Despite most of the knowledge we have on microorganism growth are from population measurements, like the RNA/DNA/etc ratios, the growth laws, etc.
+Extensive work of Bremmer and Dennis.
+Big discoveries of the last decades that rely on population measurements ? (cite 2 or 3)
+
+Single-cell measurements lay on technologies like microscopy, fluorescence, FACS...
+More interesting since we can use fluorescence to not just look at the cell but also measure interesting parameters (like gene expression) at the single-cell level.
+Information on the localisation of DNA, ribosomes [Bakshi].
+Variability in physiology, persistors, antibiotics resistance, etc [citations]
+
+Both approaches can be performed either in batch or continuous conditions.
+Batch conditions is the most straight forward in lab: you put sugar in a flask, they grow until nutrient is depleted.
+Implemented in flash or petri-microscopy.
+Resist to contamination.
+It is close to natural conditions, and industrial conditions.
+Has even been used to study long-term evolution of E. coli.
+But the system is poorly control, mostly because the history of the cells play a huge role in the waking process.
+
+Continuous culture consists into constantly providing nutrients to a culture, and maintaining it in a steady-state for a long time.
+Implemented in bioreactor [citation] or microfluidic devices [mother machine citation].
+Sensible to contaminations so hard to maintained (that's why industry ignore it so far).
+Does not correspond to natural conditions but the system is fully controlled: perturbations can be applied and a lot of parameters can be monitored.
+Possibility to let the cell long-enough in steady state to "forget" strain history.
+Most suitable if one need to have full control on its system.
+
+Finally, most of the work cited above are based on steady-state studies.
+But there are also work in dynamic.
+Levy showed that... in yeast.
+Alon in Coli.
+After decades working at steady-state, Bremmer and Dennis argued we should measure things during transitions.
+de Jong / Gerosa showed that the gene expression machinery has its own dynamic that should be taken into account when measuring gene expression.
+The evolutionary relevance of growth transitions has been emphasized in numerous others studies [citations lag phase, last generation before stationary, ...].
 
@@
 
-We describe dynamical optimilaty by using optimal control theory.
+We describe dynamical optimality by using optimal control theory.
 \textit{Brieve explanation about optimal control theory and how it is used in economic or engineering to optimize or understant dynamical processes.}

### 2016-11-29 Chapters/introduction.tex
Uncomplete version of intro sent to Hidde/Hans, add rem_intro.md that describe what is missing.

@@
 These physical limitations strongly impede a full empirical study of growth processes, making sometimes hard to test the most simple verbal hypothesis.
-
 Since they are designed to abstract away complexity, mathematical models are great tools to deal with such limitations.
@@
 
-Models have to be tested with data.
+\textbf{Either to build models or test their predictions, data on microbial growth have to be acquired.}
 Measuring growth = measuring biomass accumulation, over time, and get relevant information on the cell inner state.
@@
 
-Population measurement is easy to implement, and allow to measure metabolites.
+\textbf{Measuring population-wide parameters is convenient and provides robust data.}
+Population measurement is easy to implement, and some things can only be measured at the population level (so far).
 But small unsynchronized effects can be hidden since we mostly work with the mean.
@@
 
+\textbf{Single-cell measurements give access to data and behaviors that are hidden at the population level.}
 Single-cell measurements lay on technologies like microscopy, fluorescence, FACS...
@@
 
+\textbf{Batch growing conditions are convenient and close to the natural conditions of bacteria, but they raise several problems.}
 Both approaches can be performed either in batch or continuous conditions.
@@
 
+\textbf{Continous cultures allow to precisely control the growing conditions.}
 Continuous culture consists into constantly providing nutrients to a culture, and maintaining it in a steady-state for a long time.
@@
 
+\textbf{In all these practical conditions, one can choose to either measure cells at steady-state or in dynamic.}
 Finally, most of the work cited above are based on steady-state studies.
@@
 
+All of this is described in chapter 2.
+
 Then, after we know what to look for, we try to actually measure how resource allocation occurs in a model organism: \textit{Escherichia coli}.
@@
 
-\section{Organisation of the dissertation}
-
-(announcing the plan of the manuscript, which actually follow exactly the approach part, without the lenghty justifications)
+This, is described in chapter 3.

### 2016-11-29 Chapters/experiments.tex
Add Bacterial strain construction methods

@@
 
-\section{Methods}
+\section{Material and Methods}
 
-\subsection{Strain design}
+\subsection*{Bacterial strain construction}
 
-We used rpsB, GFP, and a chromosomic construction because so and so.
+To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2} [citation Alon] to the C-terminus of rpsB (the gene coding the ribosome subunit S2).
+This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.
+coli}~\cite{bakshi_superresolution_2012}.
 
-\subsection{Strain construction}
+The transcription factor tsf is under the control of the same promoter than rpsB, and is located after the C-terminus end.
+In order to ensure that tsf expression was not affected by our translation fusion of rpsB, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
+This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
 
-We used Gibson assembly, Kan-pbad-ccdb, etc and obtain the following sequence.
+A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus rpsB, and to the 3'-end of gfpmut2.
+The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
+This cassette is referred below as kan-P\textsubscript{BAD}-ccdB.
+
+Another DNA fragment containing an ATG-less gfpmut2~\cite{zaslaver_comprehensive_2006} was amplified with long primers annealing respectively to the C-terminus region of rpsB (just before the stop codon), and the 5'-end of the kan-P\textsubscript{BAD}-ccdB cassette.
+The first primer also contained a 18-bp (base pair) linker that was selected according to the method section of~\cite{bakshi_superresolution_2012}.
+
+Both fragments (the gfpmut2 reporter and the kan-P\textsubscript{BAD}-ccdB cassette) were assembled using Gibson assembly~[citation].
+Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
+A final product of 2683 bp was obtained:
+\begin{itemize}
+\item *(50 bp) the C-terminus of rpsB without the STOP codon 
+\item (18 bp) a linker
+\item (714 bp) the gfpmut2 sequence without the initial ATG
+\item (1851 bp) the kan-P\textsubscript{BAD}-ccdB
+\item *(50 bp) the region directly after rpsB in \textit{E. coli}
+\end{itemize}
+Regions labeled with * are supposed to anneal with the \textit{E. coli} chromosome.
+The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
+
+This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid providing Chloramphenicol resistance (lambda-red recombinaison).
+A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
+A new 100-bp fragment containing 50 bp of the 3'-end of gfpmut2 and 50 bp of the region just after rpsB was electroporated into this strain (sequence available in Supporting Information) to get rid of the kan-P\textsubscript{BAD}-ccdB cassette.
+An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
+Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
+The region after rpsB was sequenced and determined to be as anticipated (full sequence available in Supporting Information).
+
+In parallel, the same protocol was used to construct mCherry and cfp variants of the same strain.
+However, we were only able to recombine and obtain the gfp and mCherry versions for reasons that were not investigated.
+The full sequence of the final rpsB-mCherry strain is available in Supporting Information.
+
+The rpsB-gfp and rpsB-mCherry strains were caracterised on different media using a Tecan microplate reader.
+They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
+However, the rpsB-mCherry strain was shown to exhibit strange fluorescence dynamics, especially during growth transitions (entering stationary phase), which made it unsuitable for our study.
+Our effort were then concentrated on the rpsB-gfp strain.
+Data as well as information on the matter are available in Supporting Information.
 

### 2016-12-07 Chapters/experiments.tex
Add all the figures and draw a map of chapter 3. Version sent to Hidde and Hans.

--- 2016-11-23 Chapters/experiments.tex
@@
-\chapter{Monitoring Gene Expression Machinery abundance during a glucose upshift}
+\chapter{Measuring the abundance of Gene Expression Machinery during an acetate-glucose upshift}
+\chaptermark{Gene Expr. Machinery during an upshift}
+\label{chap:experiments}
+
+\textit{"If reality does not fit the concept, too bad for reality."} -- Georg Wilhelm Friedrich Hegel
 
@@
 
+\textbf{Most of the studies on the growth of microorganisms are done at steady state.}
+It is a reasonable and logical choice since working in dynamic represent a burden.
+Experiments are reproducible, and we can draw simple theory that help us to understand the inherent complexity of growth.
+But while this ideal environment is easily achievable, microorganisms are rarely found in such a condition.
+Changing our paradigm might uncover interesting laws, and raise interesting questions.
+What exactly can one see when looking into growth transitions?
+
+\textbf{The principles that define the regulatory processes in dynamic appear to be different than the ones that apply at steady state.}
+We showed in the previous chapter that good regulatory strategies have to be able to perform extremely rapid and intense variations of the gene activity.
+This rely on the fact that a bang-bang-singular production of gene expression machinery settles the upper-bound for biomass production during a nutrient upshift or downshift.
+Such strategies are only implementable if the regulatory system focus on the internal state of the cell.
+But not every information is equivalent, and performing a near-perfect transition might require to measure several variables, a concept totally missing from steady-state considerations.
+Interestingly, we uncovered how the ppGpp regulatory system of \textit{E.
+coli} might be a cheap way for the cell to gain access to enough information, and implement such an "on-off" strategy to regulate the synthesis of ribosomes.
+Overall, the theory teaches us that the actual complexity of regulatory systems is only beneficial in a dynamical context.
+Certainly, we can expect data to agree with the theory, and \textit{E. coli} to regulate its ribosome synthesis in an on-off manner...
+
+\textbf{The reality actually highlights the problem with adopting a dynamical perspective on growth.}
+There are far more information on ribosome abundance at steady state than during growth transitions.
+It is also widely admitted that the latter are hard to control and rely too much on hidden variables about the strain history which make them harder to reproduce~[citation ?].
+Though, as we discussed in the previous chapter, some data are actually available and seem to confirm that during growth transitions, the  synthesis rate of ribomoses can oscillate~\cite{gausing_regulation_1980,zengel_transcription_1986}, and that the ppGpp regulatory system can go through rapid changes~\cite{friesen_synthesis_1975,murray_control_2003}.
+But we cannot decisively validate or disprove the model predictions since these measurements were carried out on the population level, at which it is inherently hard to identify switching patterns.
+What we need are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
+Would the bang-bang strategy be observed, it could confirm the importance of optimizing transitions for the organism.
+Would it be disproved, that could raise even more interesting questions.
+Growth rate and biomass might not necessarily be what natural selection care for, or other unexpected limitations might prevent the cell to behave optimally.
+\textit{[Remarks : might be interesting to actually show the figures of the cited publications just after this paragraph]}
+
+\textbf{To tackle the issue, we took the challenge of measuring ribosomal abundance of \textit{E. coli} at the single-cell level during a controlled upshift.}
+We made a strain displaying fluorescent ribosomes, thus allowing in-vivo measurement of their abundance in dynamical conditions.
+We placed it in a microfluidic device allowing the long-term imaging of individual cells while they are presented different growing medium, and focused on the classic upshift from acetate to glucose.
+We then used Kalman smoothing to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift.
+While not allowing a decisive validation of the expected behavior, this experiment opens interesting questions and is a step towards a better understanding of the gene expression machinery regulation during growth transitions.
+
+\section{Material and Methods}
+
+\subsection{Bacterial strain construction}
+
+To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} to the C-terminus of \textit{rpsB} (the gene coding the ribosome subunit S2).
+This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.~coli}~\cite{bakshi_superresolution_2012}.
+
+The transcription factor \textit{tsf} is under the control of the same promoter than \textit{rpsB}, and is located after the C-terminus.
+In order to ensure that \textit{tsf} expression was not affected by our translation fusion on \textit{rpsB}, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
+This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
+
+A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
+The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
+This cassette is referred below as \textit{kan-P\textsubscript{BAD}-ccdB}.
+
+Another DNA fragment containing \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} without ATG was amplified with long primers annealing respectively to the C-terminus of rpsB (just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+The first primer also contained a 18-bp (base pair) linker that was chosen according to the method section of~\cite{bakshi_superresolution_2012}.
+
+Both fragments (the \textit{gfpmut2} reporter and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette) were assembled using Gibson assembly~\cite{gibson_enzymatic_2009}.
+Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
+A final product of 2683 bp was obtained:
 \begin{itemize}
-\item Understanding growth transitions is important, because so and so.
-\item We predicted earlier counter-intuitive bang-bang-singular behaviours for GEM during transition
-\item But we're missing experimental data on what really happens in real cells during these transitions (quick review of available data)
-\item We fix it by making a strain with fluorescent ribosomes and monitoring ribosome abundance during growth transitions in a microfluidic device.
+\item *(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
+\item (18 bp) a linker (see Supporting Information)
+\item (714 bp) the \textit{gfpmut2} sequence without the initial ATG
+\item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette
+\item *(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli}
 \end{itemize}
+Regions labeled with * are expected to anneal with the \textit{E.~coli} chromosome.
+The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
 
-\section{Methods}
+This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid with chloramphenicol resistance (lambda-red recombinaison).
+A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
+A new 100-bp fragment containing 50 bp of the end of \textit{gfpmut2} and 50 bp of the region just after \textit{rpsB} was electroporated into this strain (sequence available in Supporting Information) to get rid of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
+Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
+The region after \textit{rpsB} was verified by sequencing (full sequence available in Supporting Information).
 
-\subsection{Strain design}
+In parallel, the same protocol was used to construct \textit{mCherry} and \textit{cfp} variants of the same strain.
+However, only the \textit{gfpmut2} and \textit{mCherry} versions were successfully obtained for reasons that were not investigated.
+The full sequence of the final \textit{rpsB-mCherry} strain is available in Supporting Information.
 
-We used rpsB, GFP, and a chromosomic construction because so and so.
-
-\subsection{Strain construction}
-
-We used Gibson assembly, Kan-pbad-ccdb, etc and obtain the following sequence.
+The \textit{rpsB-gfpmut2} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
+They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
+However, the \textit{rpsB-mCherry} strain exhibited strange fluorescence dynamics, especially during growth transitions, which made it unsuitable for our study.
+Our effort were then concentrated on the \textit{rpsB-gfp} strain.
+Data as well as information on the matter are available in Supporting Information.
 
@@
 
-Mother machine because needed to change the medium, single cell data because asynchronization etc.
+\textit{Irina gave me the relevant information about the microscope, the pump, the making of the microfluidic device, the auto-focus algorithm, etc.
+Just need to rephrase everything.
+We took 6 fields each containing 15 wells.}
 
@@
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+\textit{We used OpenCV, scikit-image, and numpy in Python 3.5.
+The raw 16-bit images given by the Olympus microscope were converted in 16-bit Tif format.
+They were aligned using cross-correlation to correct for microscope drift.
+Each well was cropped into an individual picture.
+Only a single cell was segmented because we had trouble using available software with our experimental set-up (big pixels, reflection problem, fluorescence and phase contrast not aligned, not enough time...).
+As a preliminary analysis, the segmentation was done manually by focusing on the poles of the deepest cell.
+We used a rectangular mask around the cell with width of X pixels.
+After applying the mask, we corrected the background by removing the mean of the noise (evaluated with a closed shutter, see supporting information).
+We then computed the mean RFU in the mask (for ribosome concentration) and the lenght of the bacteria (for growth rate).
+}
 
@@
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+\textit{We used Pandas and numpy to manipulate the data, and pykalman to perform the smoothing in Python 3.5.
+Kalman smoother rely on this and that.
+The following problem was considered :
+[equations]
+We used a double integral of the white noise because so and so.
+}
 
@@
 
-\subsection{Model calibration}
+\textbf{In Chapter~\ref{chap:theory}, we developed a model of resource allocation during growth transitions.}
+By applying optimal control theory, we predicted that in order to maximize biomass production, the cell should produce its gene expression machinery in a bang-bang manner.
+The full model with dimensional parameters is reproduced below for clarity, with dotted variables representing their time derivative:
+\begin{eqnarray}
+\dot{p}(t) &=& e_M(t)\cdot (1/\beta - r(t)) - \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (1+\beta\, p(t)), \label{eq:pdef-exp}\\
+\dot{r}(t) &=& \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (\alpha(t) - \beta\, r(t)). \label{eq:rdef-exp}
+\end{eqnarray}
+In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
+$e_M(t)$ [min\textsuperscript{-1}] is the richness of the environment.
+$p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
+$r(t)$ [g.L\textsuperscript{-1}] is the gene expression machinery concentration inside the cell.
+$\alpha (t)$ [$\emptyset$] is the part of resources that goes to the gene expression machinery.
+$k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
+$K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
+$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules (assumed constant).
+The goal of this chapter is to measure in vivo $\alpha (t)$ during a growth transition, and check how it compares with the gold standard established in Chapter~\ref{chap:theory}.
 
-We want to get $\alpha (t)$ during a transition. We show that $\alpha$ can be obtained if we measure so and so.
+\subsection{Experimental design and data acquisition}
 
-\subsection{Bang-bang expression during transitions, as predicted}
+\textbf{How does one get $\alpha (t)$ in real cells?}
+There is no trivial way to directly measure $\alpha (t)$ in an experiment, mostly because of its abstract nature.
+$\alpha (t)$ can thus only be correctly estimated if one knows the value of every single term in the equations in which it appears.
+$\alpha (t)$ exists only in Eq.~\ref{eq:rdef-exp}, which simplify the burden.
+Since $\alpha (t)$ is already isolated in this equation, one can easily see that 3 terms need to be estimated for $\alpha (t)$ to be identifiable: $\beta r(t)$, $\dot{r}(t)$, and $\frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t)$.
+This also works if one can estimate the 6 individual components $r(t)$, $\dot{r}(t)$, $p(t)$, $k_R$, $K_R$, $\beta$.
+Even if we assume that the constant parameters are known, or at least easier to estimate independently, using individual components would require to co-estimate the gene expression machinery concentration ($r(t)$, $\dot{r}(t)$) and the precursor concentration $p(t)$.
 
-One stunning figure that show the RFU / cell / pixels, and the predicted $\alpha$ for the glucose-acetate transition.
+\textbf{Measuring the abundance of gene expression is achievable if one focus on a proxy that represents the overall abundance.}
+\textit{
+Ribosomes are the most abundant part of gene expression machinery and are usually considered as its main component.
+It has been shown that the RNA polymerase is not the limiting factor for GEM activity~[citation].
+Measuring ribosomes has already been done, for instance by using the total RNA of the cell, radioactive markers, or fluorescent reporter genes.
+Measuring precursor abundance is more challenging, because no clear proxy exist that represents all the precursors in the cell.
+Furthermore, in-vivo and single-cell measurements are totally out of reach~(citations citations)}
 
-\subsection{Discrepancies with the model and model adaptation}
+\textbf{This problem can be overcome by using a simple transformation.}
+Taking into account that by construction the growth rate  $\mu(t)$ is given by:
+\[
+	\mu (t) = \beta \frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t),
+\]
+we can rewrite Eq.~\ref{eq:rdef-exp} as follows
+\[
+\dot{r}(t) = \mu (t) \, \left(\frac{\alpha(t)}{\beta} - r(t) \right).
+\]
+The problem of estimating $\alpha(t)$ is thus equivalent to the same problem where $r(t)$, $\dot{r}(t)$, $\mu (t)$ and $\beta$ have to be co-estimated.
+If we are satisfied with estimating $\alpha (t)$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (t) / \beta$ identifiable.
 
-The growth rate stalls after the addition of glucose, which was not originally predicted by the model.
-This is because so and so.
-This can be taken into account if we add another sector (M1 and M2).
-This raises a new interesting question : during the transition amino-acids drop, so the cell should produce ribosomes, but it also need to produce the new enzymes to actually be able to perform metabolism.
-We thus have competition between several layers of regulation, which can be added to the model by doing so and so.
+\textbf{There are several possible strategies to measure ribosome concentration and the growth rate, and their choice will shape the study limitations.}
+\textit{
+We need short sample time (dynamics in 120 min during simulation, with abrupt changes) and single-cell measurements.
+Indeed the individual dynamic could be hidden (bang-bang) if cells are not synchronous, because it goes up then down.
+We also need to control the growth conditions.
+So we use fluorescence, microscopy and microfluidics.
+The growth rate can be directly estimated by looking at the cell size.
+The ribosome abundance can be evaluated by using a strain with GFP-labeled ribosomes.
+}
+
+\textbf{What kind of growth conditions?}
+\textit{
+We need to study growth transitions.
+Stationary phase is ill-controlled, with other processes not included in the model occuring, we must stay away from it.
+We thus go from steady-state to steady-state.
+Large change in $\mu$ might be necessary so we choose M9 minimal medium with acetate to glucose.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/experiment_schema.eps}
+\caption{The experiment. Caption to write.}
+\label{fig:experiment_schema}
+\end{figure}
+
+\textbf{Note that several issues occurred during the experiment.
+}\textit{
+Out of focus during the night.
+Cells die at the end of the experiment for an unknown reason, see discussion.
+We use the exploitable part of the time scale, and get rid of out of focus images.
+}
+
+\textbf{The image analysis was done by focusing on the deepest cell in each well.
+}\textit{Clicking at the pole on each image.
+Defining a rectangle.
+Figure about how we defined the rectangle.
+The RFU, the length and the size must appear. Can be grouped with the last figure.
+Correction background, closed shutter.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/data_acquisition}
+\caption{Result of the data acquisition. Caption to write.}
+\label{fig:data_acquisition}
+\end{figure}
+
+\subsection{Data analysis using Kalman smoother}
+
+\textbf{We have the bacteria size and RFU, but we want growth rate and gene expression machinery abundance.}
+\textit{
+They are several ways to obtain this information.
+We will use Kalman smoothing.
+}
+
+\textit{
+Kalman smoothing is a method that allow to do so and so, based on this principle, bayesian, blablabla.
+}
+
+\textit{
+We state the problem as follows.
+Gamma blablabla.
+mu and alpha can be co-estimated, but we decided to first estimate mu than alpha because so and so.
+}
+
+\textit{
+We generate synthetic data based on noise estimation (see supp info).
+Can the kalman smoother estimate synthetic data for bang-bang?
+Here figure with the synthetic data and the estimate.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/synthetic_upshift}
+\caption{Synthetic data based on the model in Chapter 2. Caption to write.}
+\label{fig:synthetic_upshift}
+\end{figure}
+
+\subsection{Estimation of the relevant parameters of the cell}
+
+\subsubsection*{Growth rate estimation}
+
+\textit{
+Mu estimation is done by looking at the length, since bacteria can be modeled as cylinder.
+We use such technique to identify discontinuities.
+We use kalman smoother with such parameters, selected by trial and error.
+There is a better way to do that (cross-validation) as is discussed in the discussion section.
+We obtain these results.
+Here a figure with the growth rate and estimation for one cell, and all the other cells in light background behind.
+We comment the result.
+As we can see there is a lot of variability, with some cells having a growth rate that goes to zero.
+This probably reflects uncoherent behaviors, like cells dying before the end of the experiment.
+Can we classify cells depending on their behavior?
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/growth_rate_estimation}
+\caption{Result of the growth rate estimation. Caption to write.}
+\label{fig:growth_rate_estimation}
+\end{figure}
+
+\textit{
+We identified 3 type of cells : normal, pausing, dying.
+We list them (see supplementary information for details on individual cells).
+We put a figure with an example for each type of cell, and the other in light gray in the background.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_cells}
+\caption{Subtypes of cells. Caption to write. Cutting the x axis might be a good idea because a third of the x-axis of each graph are "missing data".}
+\label{fig:subcat_cells}
+\end{figure}
+
+\textit{
+We focus on normal cells.
+Here figure with the median, 5 percent and 95 percent curves (or 25 and 75) for the growth rate.
+We comment the result for normal cells.
+The growth rate increase after addition of glucose, that's neat.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_median}
+\caption{Subtypes of cells, median. Caption to write.}
+\label{fig:subcat_median}
+\end{figure}
+
+\subsubsection*{Estimation of gene activity}
+
+\textit{
+We use the mean RFU.
+No need to identify discontinuities here since the signal is continuous.
+We use kalman smoother on real cells with the same parameters.
+Here figure with a typical cell.
+We see strange fluctuations.
+Not sure if they are real...
+Comment comment.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity}
+\caption{Gene activity. Caption to write.}
+\label{fig:gene_activity}
+\end{figure}
+
+\textit{
+If we look at all the cells, we can plot the median and quartiles.
+The first bang seems to be reproducible and synchronous.
+That could represent what we expect, but be careful. See discussion.
+}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity_median}
+\caption{Summary statistics for normal cells. Caption to write.}
+\label{fig:gene_activity_median}
+\end{figure}
 

### 2016-12-09 Chapters/theory.tex
Add some ref and label
Minor correction for subtitles
Add space for french summary everywhere. Comment out chapter 2 abstract because chapter 3 has no abstract.

@@
 \chaptermark{Dynamical allocation of cell resources}
+\label{chap:theory}
 
@@
 
-\section*{Summary of chapter \thechapter}
-Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
-Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
-While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
-The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
-We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
-We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
-We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
-A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
-Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
-It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
-Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
+%\section*{Summary of chapter \thechapter}
+%Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
+%Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
+%While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
+%The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
+%We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
+%We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
+%We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
+%A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
+%Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
+%It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
+%Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
 
 \selectlanguage{french}
-\section*{Résumé du chapître \thechapter : [titre français]}
+\section*{Résumé en français du chapître \thechapter : [titre français]}
 dans cette section blablabla
+
 \selectlanguage{english}
@@
 
-\subsection{Precursor-only strategy}
+\subsubsection{Precursor-only strategy}
 
@@
 
-\subsection{On-off strategy}
+\subsubsection{On-off strategy}
 

### 2016-12-17 Chapters/theory.tex
Finishing chapter 3, sent to Hidde and Hans

@@
 
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{./Fig/Fig1.eps}
-\caption[Self-replicator model of bacterial growth]{
-{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=0.9]{./Fig/Fig1.eps}
+\caption{
+\textbf{Self-replicator model of bacterial growth.}
+External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
 The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \includegraphics[width=\textwidth]{./Fig/Fig2.eps}
-\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
+\caption{\textbf{Analysis of self-replicator model of bacterial growth.}
 \textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \includegraphics[width=\textwidth]{./Fig/Fig3.eps}
-\caption[Self-replicator model accounts for bacterial growth laws]{
-\textbf{Self-replicator model accounts for bacterial growth laws.}\\
+\caption{
+\textbf{Self-replicator model accounts for bacterial growth laws.}
 \textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
@@
 
-A self-replicator at steady state accumulates biomass according to $\texttt{Vol}(0)\, e^{\mu^*\, t}$, $t\in [0,\tau]$, when $\mu^*$ is the growth rate at steady state. 
+A self-replicator at steady state accumulates biomass according to \\ $\texttt{Vol}(0)\, e^{\mu^*\, t}$, $t\in [0,\tau]$, when $\mu^*$ is the growth rate at steady state. 
 The accumulation of biomass is obviously maximal when the growth rate is maximal ($\mu^* = \mu^*_{opt}$).
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
 \includegraphics[width=\textwidth]{./Fig/Fig4.eps}
-\caption[Optimal control of the self-replicator during a nutrient upshift]{
-{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
+\caption{
+\textbf{Optimal control of the self-replicator during a nutrient upshift.}
 \textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines.
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
 \includegraphics[width=\textwidth]{./Fig/Fig5.eps}
-\caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
-{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
+\caption{
+\textbf{Alternative strategies for controlling the self-replicator of bacterial growth.}
 The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \centering
-\includegraphics[height=5cm]{./Fig/Fig6.eps}
-\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
-{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
+\includegraphics[scale=1]{./Fig/Fig6.eps}
+\caption{
+\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
 \textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \centering
-\includegraphics[height=5cm]{./Fig/Fig7.eps}
-\caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
-{
-{\bf Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}\newline
+\includegraphics[scale=1]{./Fig/Fig7.eps}
+\caption{
+\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
 \textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
-\includegraphics[width=\textwidth]{./Fig/Fig8.eps}
-\caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
-{
-{\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
+\includegraphics[scale=1]{./Fig/Fig8.eps}
+\caption{
+\textbf{ppGpp regulation implements an on-off control strategy of resource allocation.}
 \textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
@@
 
-\section{S1 Text -- Model derivation and analysis}
+\section{Supporting information for Chapter~2}
+
+\subsection{S1 Text -- Model derivation and analysis}
 \manuallabel{S1_Text}{S1~Text}
 
-\subsection{Model formulation}
+\subsubsection{Model formulation}
 
@@
 
-\subsection{Nondimensionalization of the system}
+\subsubsection{Nondimensionalization of the system}
 
@@
 
-\subsection{Steady-state growth of the self-replicator}
+\subsubsection{Steady-state growth of the self-replicator}
 \label{si::growthrate}
@@
 \label{eq:supp_growthrate_p}
-\hat{\mu}^* = \frac{(1-\alpha)\, E_M}{1+\hat{p}^*} = \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
+\begin{split}
+\hat{\mu}^* &= \frac{(1-\alpha)\, E_M}{1+\hat{p}^*}\\
+ &= \frac{2\alpha(1-\alpha)\, E_M}{(1-\alpha)\, E_M + \alpha + \sqrt{\left[(1-\alpha)\, E_M -\alpha\right]^2 + 4\alpha(1-\alpha)\, E_M\, K}}.
+\end{split}
 \end{equation}
@@
 
-\subsection{Maximization of growth rate at steady state}
+\subsubsection{Maximization of growth rate at steady state}
 \label{si::optimal}
@@
 
-\subsection{Analysis of the control strategies}
+\subsubsection{Analysis of the control strategies}
 \label{si::control_strategies}
@@
 
-\subsubsection{Nutrient-only strategy}
+\paragraph{Nutrient-only strategy}
 
@@
 
-\subsubsection{Precursor-only strategy}
+\paragraph{Precursor-only strategy}
 
@@
 \end{equation}
-Evaluated at $(\hat{p}^*, \hat{r}^*) = (\sqrt{K E_M}, g(\sqrt{K E_M}))$, the Jacobian becomes
+Evaluated at $(\hat{p}^*, \hat{r}^*) = (\sqrt{K E_M}, g(\sqrt{K E_M}))$, the Jacobian $J_{(\hat{p}^*, \hat{r}^*)}$ becomes
 \begin{equation}
 \label{eq:supp_jacobian_g_eq}
-J_{(\hat{p}^*, \hat{r}^*)} = \left(\begin{matrix}
+\left(\begin{matrix}
 - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} & - E_M - \frac{\sqrt{E_M}}{\sqrt{K}+\sqrt{E_M}} (1+\sqrt{K E_M}) \\
@@
 
-\subsubsection{On-off strategy}
+\paragraph{On-off strategy}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[tb]
 \centering
@@
 
-\section{S2 Text -- Model parameters}
+\subsection{S2 Text -- Model parameters}
 \manuallabel{S2_Text}{S2~Text}
@@
 
-\subsection*{\Large \texorpdfstring{$e_M$}{eM}}
+\subsubsection{\Large \texorpdfstring{$e_M$}{eM}}
 
@@
 
-\subsection*{\Large \texorpdfstring{$k_R$}{kR}}
+\subsubsection{\Large \texorpdfstring{$k_R$}{kR}}
 
@@
 
-\subsection*{\Large \texorpdfstring{$K_R$}{KR}}
+\subsubsection{\Large \texorpdfstring{$K_R$}{KR}}
 
@@
 
-\subsection*{\Large \texorpdfstring{$\beta$}{Beta}}
+\subsubsection{\Large \texorpdfstring{$\beta$}{Beta}}
 
@@
 
-\subsection*{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}
+\subsubsection{\Large \texorpdfstring{$E_M$}{EM} and \texorpdfstring{$K$}{K}}
 
@@
 
-\section{S3 Text -- Solution of optimal control problem}
+\subsection{S3 Text -- Solution of optimal control problem}
 \manuallabel{S3_Text}{S3~Text}
 
-\subsection{Statement of the problem}
+\subsubsection{Statement of the problem}
 
@@
 
-\subsection{Maximum Principle}
+\subsubsection{Maximum Principle}
+
 Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle~\cite{carlson_infinite_1991}.
@@
 
-\subsection{Characterization of singular arcs}
+\subsubsection{Characterization of singular arcs}
 
@@
 
-\subsection{Analysis of the adjoint system}
+\subsubsection{Analysis of the adjoint system}
+
 Recalling that a switch corresponds to a change of sign of $\lambda_r$, the analysis of the adjoint system (Eqs~\ref{eq:adjoint_p}-\ref{eq:adjoint_r}) may be useful to characterize the switches of extremal trajectories.
@@
 
-\begin{figure}[h]
+\begin{figure}[tb]
 \centering
-\includegraphics[width=0.35\textwidth]{Fig/adj2ab.eps} 
-\includegraphics[width=0.6\textwidth]{Fig/adj2.eps} 
+\includegraphics[height=4cm]{Fig/adj2ab.eps} 
+\includegraphics[height=4cm]{Fig/adj2.eps} 
 \caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis.
@@
 
-\subsection{Optimal trajectories}
+\subsubsection{Optimal trajectories}
 
@@
 
-\section[S4 Text -- Kinetic model of the ppGpp system]{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
-\sectionmark{S4 Text -- Kinetic model of the ppGpp system}
-\sectionmark{S4 Text -- Kinetic model of the ppGpp system}
-
+\subsection[S4 Text -- Kinetic model of the ppGpp system]{S4 Text -- Kinetic model of the ppGpp system in \textit{Escherichia coli}}
+\sectionmark{Kinetic model of the ppGpp system}
 \manuallabel{S4_Text}{S4~Text}
@@
 \begin{equation}
+\footnotesize
 \texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
@@
 
-\begin{figure}[!ht]
+\begin{figure}[tb]~Text
 \centering
@@
 
-\begin{table}[!h]
-\tiny
+\begin{table}[tb]
+\scriptsize
 \centering
@@
 
-\section{S1 Table -- Parameter values of self-replicator model}
+\subsection{S1 Table -- Parameter values of self-replicator model}
 \manuallabel{S1_Table}{S1~Table}
 
-\begin{table}[!h]
+\begin{table}[h]
 \centering
@@
 
-\section{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth}
+\subsection{S1 Figure -- Simple control strategies for the self-replicator of bacterial growth}
 \manuallabel{S1_Fig}{S1~Figure}
 
-\begin{figure}[!h]
+\begin{figure}[h]
 \centering

### 2016-12-17 Chapters/experiments.tex
Move Methods after the discussion
Add space for french summary everywhere. Comment out chapter 2 abstract because chapter 3 has no abstract.
Finishing chapter 3, sent to Hidde and Hans

--- 2016-11-23 Chapters/experiments.tex
@@
-\chapter{Monitoring Gene Expression Machinery abundance during a glucose upshift}
+\chapter{Dynamics of the gene expression machinery during an acetate-glucose upshift}
+\chaptermark{Gene expr. machinery during an upshift}
+\label{chap:experiments}
+
+%\textit{"If reality does not fit the concept, too bad for reality."} -- Georg Wilhelm Friedrich Hegel
+
+\selectlanguage{french}
+\section*{Résumé en français du Chapître \thechapter : Dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose.}
+
+Résumé d'environ 6-8 paragraphes (2 pages).
+\selectlanguage{english}
 
@@
 
+Most studies on the growth of microorganisms are done at steady state.
+It is a reasonable and logical choice given the drawbacks of working in dynamic.
+At steady state, microorganism behaviors are reproducible, and which helps uncover simple theories to apprehend their inherent complexity.
+But while this ideal environment is easily achievable at the bench, microorganisms naturally spend very little time in such a perfect condition.
+This motivated in chapter~\ref{chap:theory} the construction of a new theoretical framework to study growth laws during growth transitions.
+
+The principles that define the regulatory processes in dynamic appear to be different than the ones that apply at steady state.
+We showed that efficient regulatory strategies have to be able to perform extremely abrupt and intense variations of the gene activity.
+This relies on the fact that bang-bang synthesis of gene expression machinery settles the upper-bound for biomass production during an environmental change.
+Such strategies are only implementable if the regulatory system focus on the internal state of the cell.
+But not all information is equivalent, and performing a near-perfect transition might require to measure several variables, a concept totally missing from steady-state considerations.
+Interestingly, we uncovered how the ppGpp regulatory system of \textit{E.
+coli} might be a cheap way for the cell to access just enough information, and to implement a \textit{on-off} strategy to regulate the synthesis of ribosomes.
+Overall, it taught us that the actual complexity of regulatory systems is only beneficial in a dynamical context.
+But elucidating what is the best behavior, and showing that \textit{E. coli} actually implement it are different matters...
+
+Reality actually highlights the problem with adopting a dynamical perspective on growth.
+There are far more information on ribosome abundance at steady state than during growth transitions.
+It is widely admitted that the latter are hard to control and rely too much on hidden variables about the strain history~[citation ? Bremmer ?].
+Though, as we discussed in section~\ref{sec:discussion}, some data are available and seem to confirm that during growth transitions, the  synthesis rate of ribomoses can oscillate~\cite{gausing_regulation_1980,zengel_transcription_1986}, and that the ppGpp regulatory system can go through rapid changes~\cite{friesen_synthesis_1975,murray_control_2003}.
+But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
+What we need are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
+Would the \textit{on-off} strategy be observed, it could confirm the importance of optimizing transitions for the organism.
+Would it be disproved, that could raise even more interesting questions: growth rate and biomass might not necessarily be what natural selection care for, or other unexpected limitations might prevent the cell to behave optimally.
+
+In Chapter~\ref{chap:theory}, we developed a model of resource allocation during growth transitions.
+By applying optimal control theory, we predicted that in order to maximize biomass production, the cell should produce its gene expression machinery in a bang-bang manner.
+The full model with dimensional parameters is reproduced below, with dotted variables representing their time derivative:
+\begin{eqnarray}
+\dot{p}(t) &=& e_M(t)\cdot (1/\beta - r(t)) - \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (1+\beta\, p(t)), \label{eq:pdef-exp}\\
+\dot{r}(t) &=& \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (\alpha(t) - \beta\, r(t)). \label{eq:rdef-exp}
+\end{eqnarray}
+In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
+$e_M(t)$ [min\textsuperscript{-1}] is the richness of the environment.
+$p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
+$r(t)$ [g.L\textsuperscript{-1}] is the gene expression machinery concentration inside the cell.
+$\alpha (t)$ [$\emptyset$] is the mass ratio of resources that goes to the gene expression machinery.
+$k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
+$K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
+$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules (assumed constant).
+The goal of this chapter is to measure in vivo $\alpha (\cdot)$ during a growth transition, and check how it compares with the gold standard established in chapter~\ref{chap:theory}.
+
+To this end, we took the challenge of measuring ribosomal abundance of \textit{E. coli} at the single-cell level during a controlled upshift.
+We made a strain displaying fluorescent ribosomes, thus allowing in-vivo measurement of their abundance in dynamical conditions.
+We placed it in a microfluidic device allowing the long-term imaging of individual cells while they are presented different growing media, and focused on the classic upshift from acetate to glucose.
+We then used Kalman smoothing to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift.
+While not allowing a decisive validation of the expected behavior, this experiment opens interesting questions and is a step towards a better understanding of the gene expression machinery regulation during growth transitions.
+
+\section{Results}
+
+\subsection{Experimental design}
+\label{sec:exp_design}
+
+How does one measure resource allocation in real cells?
+There is no direct way to measure $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
+To correctly reconstruct it, one has to know the value of every single term in the equations in which it appears.
+In the full dynamical system Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only pops up in Eq.~\ref{eq:rdef-exp}, and is already isolated in this equation, which simplifies the burden.
+We can identify $\alpha (t_k)$ for each time $t_k$ where the following 3 terms are known: $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$.
+Such a consideration remains valid if one can estimate the 6 individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$.
+Even if we assume that the constant parameters are known, or at least easier to estimate independently, reconstructing the dynamics of $\alpha$ would require to co-estimate the gene expression machinery concentration ($r$, $\dot{r}$) and the precursor concentration $p$.
+
+Are such measurements acceptable?
+Measuring the abundance of gene expression is achievable if one focus on a proxy that represents the overall abundance.
+Ribosomes are widely known to be the most abundant part of the gene expression machinery, which allows to consider it as its main component~\cite{scott_interdependence_2010, scott_bacterial_2011,scott_emergence_2014}.
+Their abundance can be estimated for instance by using the total RNA of the cell~\cite{scott_interdependence_2010}, radioactive markers\cite{gausing_regulation_1980,zengel_transcription_1986}, or fluorescent labeling~\cite{bakshi_superresolution_2012}.
+Estimating precursor abundance is more challenging.
+There is no clear proxy that represents all the precursors in the cell, so using common methods would be excessively time expensive~[citation classic methods to estimate precursor ?].
+While absolute quantification of all the internal metabolites has been achieved in \textit{E.
+coli}~\cite{bennett_absolute_2009}, this method do not comply with our need of real-time single-cell quantification.
+
+This problem can be overcome by using a simple transformation.
+Taking into account that by construction the growth rate  $\mu(t)$ is given by:
+\[
+	\mu (t) = \beta \frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t),
+\]
+we can rewrite Eq.~\ref{eq:rdef-exp} as follows
+\begin{equation}
+\label{eq:dot_r}
+\dot{r}(t) = \mu (t) \, \left(\frac{\alpha(t)}{\beta} - r(t) \right).
+\end{equation}
+The problem of estimating $\alpha(\cdot)$ is thus equivalent to the same problem where $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$ have to be co-estimated.
+If we are satisfied with estimating $\alpha$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (\cdot) / \beta$ identifiable.
+
+These new measurements are more acceptable given the requirements of the study.
+We want to be able to identify potential abrupt changes of the gene expression.
+Such switching patterns are easily hidden at the population level if cells are not synchronous, which is generally the case.
+A short sample time and single-cell measurements are thus necessary, which motivated the use of fluorescent reporters like the green fluorescent protein (gfp).
+We constructed a strain displaying a gfp-tagged S2 ribosomal subunit (the gene name being \textit{rpsB}).
+The \textit{rpsB-gfp} strain produces fluorescent ribosomes, making them quantifiable while minimizing the burden for the cell (see Material and Methods~\ref{sec:methods_strain}).
+By observing such a strain through a microscope during a growth transition, we could theoretically estimate the ribosome concentration by measuring the cell fluorescence, and reconstruct the growth rate by measuring the cell size.
+
+The growth conditions might also play a role in the limitations of the study.
+In natural conditions, the most straightforward growth transitions, or at least the one at which most of the selection is expected to occur, is the transition from the stationary phase in nutrient-deprived medium to any medium sustaining growth.
+There are good reasons though for which this particular growth transition is ignored from most studies.
+The transition from stationary to exponential phase is dramatically ill-controlled, in the sense that a lot of hard-to-control parameters seem to play a huge role in it.
+For instance, this is the case of the duration of the nutritive stress~[citation to find], the nutrient encountered during the last exponential phase~[citation to find], or the previous accumulation of wastes in the medium~[citation to find].
+This somehow results in a phenomenon called the lag phase~[citation lag phase review to find], the mechanisms of which are missing from our model.
+Furthermore, most microorganisms dramatically change their physiology in stationary-phase, either sporulating or organizing into biofilms~[citations subtilis sporulation and coli biofilms to find], which goes largely beyond the scope of this study.
+For this reason, we decided to focus on a transition from a slow exponential growth to high exponential growth, in particular the transition from acetate to glucose.
+By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively to ensure cells have \textit{forgotten} the previous stress phase, and have enough time to reach the new steady state.
+The mothermachine, a microfluidic device that was designed to sustain exponential growth for long period of time, seemed like a good choice for this purpose.
+It allows to maintain a constant flow of fresh medium around the cells, and make fast transitions possibles by switching the medium that is fed to the cells (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
+The complete experimental scheme is summarized in Fig.~\ref{fig:experiment_schema}.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/experiment_schema.eps}
+\caption{
+\textbf{Scheme of the upshift experiment.}
+The goal is to measure the fluorescence and growth rate of cells during an acetate-to-glucose upshift.
+All the media used are M9 minimal medium supplemented with 0.2\% of the indicated carbon source, in this case acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
+The \textit{rpsB-gfp} strain was precultivated for 2.5 days on 0.2\% acetate in batch condition (shaken flask).
+The day of the experiment, the cells were injected into the mothermachine and fed by a constant flow of fresh 0.2\% acetate medium (see Material and Methods~\ref{sec:microflu}).
+Fluorescence and phase contrast images were taken every 5 minutes.
+Twenty hours later, the feeding media was switched to 0.2\% glucose (known to theoretically sustain a 5-time-higher growth rate) and was maintained for twenty more hours while imaging carried on.
+Time 0 corresponds to the exact moment when cells start to be in contact with the glucose medium.
+}
+\label{fig:experiment_schema}
+\end{figure}
+
+\subsection{Data acquisition}
+
+It is important to evoke most of the issues that occurred while carrying on the experiment.
+At the time, a serious issue was starting to affect our only microscope: the motorization along the Z-axis was randomly stalling until an operator manually intervenes.
+This deactivates the autofocus and the Z-compensation for the tilt of the device when moving along the XY plane.
+In other words, series of images are out of focus and thus not exploitable if the operator was not around, which can happen a lot in a 40-hour-long experiment.
+For the experiment presented here, we timed the operations so that the operator can be around as long as possible when it counts, \textit{i.e.} close to the transition at time 0.
+This is the reason why data points between -720 and -150 minutes are not available (corresponding to the steady state on acetate).
+Another issue, more specific to the presented occurence of the experiment, was that bacteria started to massively die several hours after the transition.
+The reason is unknown, the most plausible hypothesis being a phage contamination.
+\footnote{Note that the operator was present when the dying started, which excludes the implication of lab gnomes.}
+Data analysis was thus stopped at 550 minutes, when roughly 3/4 of the cells were still growing.
+Note that for these reasons, the results of this preliminary experiment must be taken with caution.
+The necessary improvements are extensively discussed in section~\ref{sec:chap3_discussion}.
+
+Image analysis was performed on the remaining shots.
+While image analysis softwares overflow the literature~[citations schitcells + others], some of them being specifically designed for the mother machine~[molysso + Wang et al], technical limitations did not allow to use any of them.
+For instance, the segmentation was systematically impeded by the poor resolution of the camera.
+The phase contrast images were not directly exploitable since we noticed they were not aligned with the fluorescence images.
+Fluorescence was also organized in hot spots (see Fig.~\ref{fig:data_acquisition}) and their intensity vary widely during the experiment, which was expected since ribosomes are extremely localized~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}.
+It, however, dramatically impeded the automatic segmentation on the fluorescence images.
+In section~\ref{sec:chap3_discussion}, we give insights into how we plan to overtake these limitations.
+
+The analysis presented in this study is thus preliminary.
+First, we focused solely on the deepest cell of each well.
+It avoided the tracking step, and ensured that the cell was visible from the beginning to the end of the experiment.
+Secondly, the segmentation was done by manually selecting two pixels at the poles of the cell of interest.
+These pixels were used to create a rectangular mask around the cell (see Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
+After background correction (see Material and Methods~\ref{sec:cell_segmentation}), the RFU concentration was evaluated in this region of interest, as well as the length of the cell, defined by the distance in pixels between the poles (see Fig.~\ref{fig:data_acquisition}).
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/data_acquisition}
+\caption{
+\textbf{Results of data acquisition.}
+We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
+From this, only 68 wells were available for analysis (the others were either empty, out of frame for some period of time, or plugged).
+For all wells, data points are missing in the time interval [-720,-150] because the camera was out of focus.
+We stopped the time series analysis at 550 minutes when visually 3/4 of the bacteria were still growing.
+The whole population was wiped out in the next few hours for an unknown reason.
+The Raw image is the last image analyzed for the highlighted well.
+The left bacteria on this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
+A 6-pixel-wide rectangular mask was computed for each image, giving the segmented image on the right.
+Fluorescence levels are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335), and were corrected for camera background, but not autofluorescence background (see Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
+The RFU/pixel/cell was computed by taking the ratio between the sum of the pixels in the mask, and the total number of pixels in the mask.
+The bacteria length is the distance in pixels between the two poles (red line on the raw and segmented images).
+The thick lines highlights the time series for the cell visible in the top images.
+The dashed black line corresponds to the switch time from acetate to glucose.
+}
+\label{fig:data_acquisition}
+\end{figure}
+
+From this preliminary analysis, we obtained time series for the concentration of fluorescence (RFU/pixel) and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
+The RFU/pixel appears constant on the acetate phase (before 0) and immediately increases when the medium is switched to glucose.
+Bacterial length shows that the division time abruptly decreases after the nutrient upshift, corresponding to a higher growth rate.
+Unfortunately, while the steady state on acetate was reached before the upshift, it is obvious that the time series stop before a new steady state on glucose is reached, at least for the fluorescence.
+Nevertheless, the time series are not interrupted around the transition.
+What do we observe if we estimate the growth rate and $\alpha$ in this region?
+
+\subsection{Data analysis using Kalman smoother}
+\label{sec:res_kalman}
+
+As we presented in section~\ref{sec:exp_design}, our goal is to reconstruct the signal $\alpha (\cdot)$ during the transition.
+By modeling \textit{E. coli} as a cylinder, the length can be assumed directly proportional to the volume of the cell.
+If we assume a negligible background of autofluorescence (see Material and Methods~\ref{sec:cell_segmentation}), the fluorescence concentration in the cell can be assumed proportional to the ribosome concentration.
+More precisely, based on the data available, we have the following measurement model at each time-point $t_k$:
+\begin{eqnarray}
+L(t_k) &= \lambda \cdot V(t_k) + \epsilon_k, \label{eq:mes_L}\\
+F(t_k) &= \gamma \cdot r(t_k) + \eta_k, \label{eq:mes_F}
+\end{eqnarray}
+where $L(t_k)$ (resp.
+$F(t_k)$) is the length (resp.
+RFU/pixel) measured at time $t_k$ (visible on Fig.~\ref{fig:data_acquisition}), $V(t_k)$ (resp.
+$r(t_k)$) is the actual volume (resp.
+ribosome concentration) at time $t_k$, ($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) are measurement noises.
+As we saw in section~\ref{sec:exp_design}, reconstructing $\alpha (\cdot)$ requires information on $\mu (\cdot)$, $r(\cdot)$ and $\dot{r}(\cdot)$.
+From Eq.~\ref{eq:mes_F}, we could obtain $r(\cdot)$ and $\dot{r}(\cdot)$ by smoothing interpolation and differentiation.
+The same can be said for Eq.~\ref{eq:mes_L} and $\mu (\cdot)$ since it is defined by:
+\[
+\dot{V}(t) = \mu (t) \cdot V(t).
+\]
+However, smoothing interpolation is particularly sensible to the boundary conditions.
+Since each division in the time series is a boundary, the smoothing interpolation of this kind of data is particularly unstable.
+For this reason, single-cell growth rate is usually estimated by fitting an exponential term on the volume between each division (or in particular a linear term on the logarithm of the volume)~[citation Wang et al et Izard et al].
+While it is suitable when bacteria are at steady state, it is not applicable in our case where we expect growth-rate variations between divisions.
+Other techniques are less sensible to this kind of problem~\cite{zulkower_robust_2015}, and we decided to focus on the use of Kalman smoother~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+
+Kalman filtering~\cite{kalman_new_1960} is a bayesian algorithm that uses a series of noisy measurements to predict the state of a dynamical system.
+It is well known for its everyday technological applications (guidance, tracking, control, ...) which requires the real-time estimation of hidden variables in a dynamical system based on previous measurements.
+Kalman smoothing is an extension of Kalman filtering that uses information about the previous but also the next measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+In that sense, it uses all the available information to estimate the hidden variables of a dynamical system.
+For our problem, at each time step $t_k$, our hidden variables are the resource allocation $\alpha (t_k)$ and the growth rate $\mu (t_k)$, and the available information is the N measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$ that were taken during the whole experiment.
+
+We state the problem as follows.
+We have the following dynamical system:
+\begin{eqnarray}
+\dot{r}(t) &=& \mu (t) \cdot \frac{\alpha(t)}{\beta} - \mu (t) \cdot r(t),\\
+\dot{V}(t) &=& \mu (t) \cdot V(t),
+\end{eqnarray}
+with the following measurement model:
+\begin{eqnarray}
+L(t_k) &= \lambda \cdot V(t_k) + \epsilon_k,\\
+F(t_k) &= \gamma \cdot r(t_k) + \eta_k,
+\end{eqnarray}
+where each variable has already been defined in Eqs~\ref{eq:dot_r}, \ref{eq:mes_L} and \ref{eq:mes_F}.
+While $\beta$ can be obtained from the literature (see chapter~\ref{chap:theory}), $\lambda$ and $\gamma$ are unknown variables that need to be estimated along with $\alpha$ and $\mu$.
+Since we are satisfied with a qualitative reconstruction of $\alpha (\cdot)$, we can simplify the system by defining $r_\gamma = \gamma \cdot r$ and $V_\lambda = \lambda \cdot V$.
+The dynamical system is thus rewritten:
+\begin{eqnarray}
+\dot{r_\gamma}(t) &=& \mu (t) \cdot\frac{\gamma \cdot \alpha(t)}{\beta} - \mu (t) \cdot r_\gamma (t),\label{eq:r_prob}\\
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
+\end{eqnarray}
+with the new measurement model:
+\begin{eqnarray}
+F(t_k) &= r_\gamma (t_k) + \eta_k,\label{eq:F_prob}\\
+L(t_k) &= V_\lambda(t_k) + \epsilon_k.\label{eq:L_prob}
+\end{eqnarray}
+The problem becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ through measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+Note that this problem is not linear, while Kalman filtering was initially introduced for linear systems~\cite{kalman_new_1960,kailath_linear_2000}.
+For this reason, we used a non-linear extension of Kalman smoothing called Unscented Kalman Smoothing~\cite{julier_new_1997,jazwinski_stochastic_2007}.
+Details about its exact implementation are available in Material and Methods~\ref{sec:meth_kalman}.
+While it is feasible to co-reconstruct the signals $\mu (\cdot)$ and $\gamma \alpha (\cdot) / \beta$ using Kalman smoothing, it is not necessary and could make the reconstruction unstable.
+We thus decided to proceed in two steps: first reconstruct $\mu$ from the measurements of $L$, than inject this result in the reconstruction of $\gamma \alpha / \beta$ from the measurements of $F$.
+
+\subsection{Estimation of growth rate and gene expression}
+
+\subsubsection*{Growth rate estimation}
+
+As presented in the previous section, we want to reconstruct the growth rate $\mu$ defined by Eq.~\ref{eq:V_prob}, using measurements of the length $L$ defined by Eq.~\ref{eq:L_prob}.
+Note that Kalman smoothing is a procedure that returns the expected mean and covariance of the state of a dynamical system, given some measurements.
+Thus, reconstructing $\mu$ requires its explicit integration in the variables of the dynamical system.
+In some sense, this tells the system which constraints to apply on the variations of $\mu$, and represents a regularization method.
+Here, we used the following Bayesian regularization approach.
+In the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process.
+The laws of this process play the role of a prior on the expected regularity of $\mu$ (the more noisy the process, the less constrained the variations of $\mu$).
+In particular, we let $\mu$ be the double integral of a Gaussian white noise $w$:
+\[
+\dot{\mu} = v(t) \;\;\;\; \dot{v} = w(t),
+\]
+where $v$ is an intermediate variable, and $w(t) \sim \mathcal{N}(0, \theta)$.
+The full problem becomes:
+\begin{eqnarray}
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
+\dot{\mu}(t) &=& v(t),\label{eq:full_mu_prob}\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray}
+with the measurement model:
+\begin{equation}
+L(t_k) = V_\lambda(t_k) + \epsilon_k.
+\end{equation}
+The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable.
+It is also equivalent to other methods like Tikhonov regularization~[citation to take from Val pub], which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
+
+Our prior is thus entirely contained in the parameters of the Kalman procedure:
 \begin{itemize}
-\item Understanding growth transitions is important, because so and so.
-\item We predicted earlier counter-intuitive bang-bang-singular behaviours for GEM during transition
-\item But we're missing experimental data on what really happens in real cells during these transitions (quick review of available data)
-\item We fix it by making a strain with fluorescent ribosomes and monitoring ribosome abundance during growth transitions in a microfluidic device.
+\item the mean of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the covariance of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$,
+\item the observation variance for $L$.
 \end{itemize}
+The means and covariances at the initial state represent the knowledge we have on the initial parameters.
+For instance, if we know exactly the value of $V_\lambda$, $\mu$ and $v$, we can fix the means and use very small covariances, but this "certainty" can be adjusted at will.
+In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because there are not the result of a stochastic process.
+However, the transition variance $\theta^2$ of the derivative of $v$ is crucial: it represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
+The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu$.
+Finally, the observation variance for $L$ is the expected measurement noise.
+All these parameters represent the prior that we use to reconstruct the growth rate $\mu$.
+As we will see below, we widely use these properties to overcome the difficulties raised by the discontinuities introduced by cell divisions.
 
-\section{Methods}
+Since we track bacteria for a long time, the observation of their length $L$ is inherently discontinuous, due to symmetrical divisions of the \textit{E. coli} cells (Fig.~\ref{fig:data_acquisition}).
+Thus, the problem cannot be solved for the whole time series at once, but has to be divided into smaller continuous regions between two divisions.
+We simply localize the divisions by looking at the time points at which the empirical derivative of the length is below an arbitrary threshold, and slice the curve into the appropriate number of continuous time series.
+From here, the most straightforward method would be to treat each portion independently.
+However, we loose an important information by doing so: neighbor portions of the curve are likely to have similar growth rate.
+This comes from the fact that we can expect a daughter cell to start growing with a growth rate similar to its mother cell.
+In other word, while not necessary continuous, the growth rate at the boundaries of a time region can be expected to depend strongly on the growth rate of neighbor regions.
+This can be easily taken into account in the Kalman smoothing procedure by exploiting the priors about the initial mean and covariance of the system variables.
+We simply let the initial state mean of the growth rate be the last estimated value in the mother cell, while defining a reasonable variance around this value to allow flexibility.
+When no data about the mother cell is available (at the beginning of the experiment), we simply let the algorithm choose freely the initial growth rate by defining a large variance around the expected growth rate on acetate.
+The results of this procedure, along with the exact parameters used for the Kalman smoothing, are reported in Fig.~\ref{fig:growth_rate_estimation} and Material and Methods~\ref{sec:meth_kalman}.
 
-\subsection{Strain design}
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/growth_rate_estimation}
+\caption{
+\textbf{Growth rate estimation based on the bacterial length using Kalman smoothing.}
+Grey lines represent estimation of states of the system by the unscented Kalman smoothing procedure for the 68 cells.
+The reconstruction for the bacterial length and the growth rate are presented in the top and bottom graphs, respectively.
+For readability, the solid red lines highlight the result for a particular cell, while black crosses in the top graph are the data points for this cell.
+The dashed grey lines represent the switch time from acetate to glucose.
+As prior for the algorithm, we used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
+Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
+At the beginning of the experiment, were no mother cell is available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
+All the cross-covariances are set to 0 because the system variables are independent by construction.
+The parameters cited above were chosen through trials and errors, and are as a consequence largely optimizable.
+Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+}
+\label{fig:growth_rate_estimation}
+\end{figure}
 
-We used rpsB, GFP, and a chromosomic construction because so and so.
+On the mean, the results presented in~Fig.~\ref{fig:growth_rate_estimation} show a roughly constant growth rate on acetate, followed by a quick increase around the upshift.
+Contrary to what was visible on the fluorescence, a steady state for the growth rate seems to have been reached before the end of the experiment.
+When taken individually, the results are more fuzzy.
+A significant part of the cells stops growing before the end of the experiment.
+This was expected since all the cells die from an unknown cause at the end of the time series, and the interval of analysis was visually chosen so that roughly 3/4 of the cells are still growing.
+Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a hour-long pause around 200 minutes after the upshift, then a growth rate recovery until the end of the time series (see Supporting Information XX).
+For now, we decided to classify the cells in three categories: the dying cells (N=12) for which the growth rate drop to zero after the upshift; the pausing cells (N=11) for which the growth rate drop to zero after the upshift then recover, and finally the normal cells (N=45) which do not exhibit any of the above behaviors.
+More information about each category is available in Supporting Information XX.
+In what follows, we focus on the 45 normal cells (Fig.~\ref{fig:growth_rate_estimation_median}), which present a globally coherent behavior until the end of the interval of analysis.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/growth_rate_estimation_median}
+\caption{
+\textbf{Growth rate estimation for the normal cells only.}
+The top graph is the same as the bottom graph of Fig.~\ref{fig:growth_rate_estimation}, except we removed the dying and pausing cells.
+The bottom graph gives the 25\% (lowest gray line), 50\% (solid black line) and 75\% (highest gray line) quartiles, computed at each time step.
+The gray area represents the interquartile range.
+}
+\label{fig:growth_rate_estimation_median}
+\end{figure}
+
+\subsubsection*{Estimation of gene activity}
+
+We use a similar Bayesian regularization approach than for the growth rate estimation.
+We note $u(t) = \gamma \alpha (t) / \beta$ the input that we want to reconstruct.
+The full problem for the estimation of $u$ is:
+\begin{eqnarray}
+\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
+\dot{u}(t) &=& v(t),\label{eq:full_u_prob}\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray}
+with the measurement model
+\begin{equation*}
+F(t_k) = r_\gamma (t_k) + \eta_k,\\
+\end{equation*}
+where $\hat{\mu}$ is the estimation of $\mu$ from the previous section.
+Contrary to the growth rate problem, this system is linear.
+Furthermore, we do not have to deal with discontinuities of the measurement since $F$ is continuous from the mother to the daughter cells (Fig.~\ref{fig:data_acquisition}).
+
+To be comparable with what we predicted in chapter~\ref{chap:theory}, we expect $u(\cdot)$ to exhibit bang-bang variations after the upshift.
+By definition, it is a very stiff signal that could be complicated to reconstruct depending on the parameters of the regularization method.
+In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of chapter~\ref{chap:theory}.
+This was not possible for the growth rate problem since cell division is absent from the model.
+After estimating the noise on the RFU/pixel measurements (see Supporting Information XX), we simulated an upshift from acetate to glucose (the parameters were chosen based on the resulting growth rate), and used a value for $\gamma$ that generated a fluorescence concentration in the range of the observations.
+We then tried to estimate $\gamma \alpha / \beta$ based these synthetic data, choosing the prior parameters through trials and errors.
+There are probably not optimal, and we discuss possible improvements in section~\ref{sec:chap3_discussion}.
+The results along with the prior parameters are presented in Fig.~\ref{fig:synthetic_upshift}.
+Like any smooth methods, the algorithm has trouble reconstructing stiff variations of the system variables.
+The oscillatory nature of gene expression is however fully captured, which indicate that the algorithm should be able to conclude on the existence of the \textit{on-off} strategy.
+Some methods that could improve this result are also discussed in section~\ref{sec:chap3_discussion}.
+In what follows, we use the same prior parameters on the "normal cell" subset of the real data, in order to see if we can get the same oscillatory pattern.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=0.8]{./Fig/synthetic_upshift}
+\caption{
+\textbf{Performance of the unscented Kalman smoother on synthetic data simulating an upshift.}
+\textit{(A)}~Synthetic data simulating an upshift with and without white noise, along the result of the Kalman smoothing procedure.
+The synthetic data were generated by simulating the model presented in Eqs~\ref{eq:pdef}-\ref{eq:rdef} with the on-off regulatory strategy (Eq.~\ref{eq:stratswitch} and Fig.~\ref{fig:onoffresults}).
+Parameters for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L.g\textsuperscript{-1}, $K_R = 1$~g.L\textsuperscript{-1}.
+The simulated $r(t)$ was multiplied by a factor $\gamma = 0.02$~L.g\textsuperscript{-1}.RFU\textsuperscript{-1} in order to get a synthetic $F$ in the same range of the observations (dashed black curve).
+Additive noise level was estimated from the real data and injected in $F$ (see Supporting Information XX).
+Here again, the parameters were chosen through trials and errors, and are reported in the Material and Methods~\ref{sec:meth_kalman}.
+They are largely optimizable, and possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+\textit{(B)}~Estimation of $\gamma \alpha / \beta$ based on the data in (A).
+As described in chapter~\ref{chap:theory}, $\alpha (t)$ performs a bang-bang-singular response during the upshift (dashed black line).
+While the Kalman smoothing is not able to capture the stiff variations of $\gamma \alpha / \beta$, it fairly reproduces the oscillatory nature of the gene expression (red solid line).
+\textit{(C)}~The growth rate during the upshift.
+This information is used as an input in the procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+}
+\label{fig:synthetic_upshift}
+\end{figure}
+
+Results of the reconstruction of $u = \gamma \alpha / \beta$ are presented in Fig.~\ref{fig:gene_activity}.
+Large oscillations are indeed visible on the reconstruction.
+These oscillations were already visible on the RFU/pixel data (Fig.~\ref{fig:gene_activity}A).
+The region between -150 and 0 min appears to be the most stable, as we could expect for a growth steady state.
+On the contrary, the reconstructed $\gamma \alpha / \beta$ seems particularly unstable at the beginning and the end of the experiment.
+Overall, the oscillations observed must be taken with extreme caution.
+The problem of reconstructing an on-off signal is more challenging than we initially thought, for the simple reason that the signal we expect to reconstruct is exactly the kind of artifacts a poorly calibrated regularization method would generate.
+Data about the pre-upshift and post-upshift steady state are more crucial than expected, and the experimental problem we described earlier truly impedes our capacity to conclude on the regulatory strategy.
+We extensively discuss this in section~\ref{sec:chap3_discussion}, and give numerous indications about the future improvements that will be implement.
+Finally, Fig.~\ref{fig:gene_activity_median} gives an idea of the distribution of the estimations at each time step.
+While most oscillations cancels out when we consider global statistics about the population, the first oscillation just after the upshift is conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}B).
+This is rather encouraging and prompts for future investigations.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity}
+\caption{
+\textbf{Gene activity estimation based on the RFU/pixel measurements using Kalman smoothing.}
+\textit{(A-B)} Grey lines represent estimation of state of the system by the unscented Kalman smoothing procedure for the 45 normal cells.
+For readability, the solid red lines highlight the result for a particular cell, while black crosses in the top graph are the data points for this cell.
+The dashed grey line represents the switch time from acetate to glucose.
+The reconstruction for the RFU/pixel measurements and the gene activity $\gamma \alpha / \beta$ are presented in the (A) and (B), respectively.
+The prior parameters used for the algorithm are exactly the same than in Fig.~\ref{fig:synthetic_upshift} and are reported in Material and Methods~\ref{sec:meth_kalman}.
+}
+\label{fig:gene_activity}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity_median}
+\caption{
+\textbf{Distribution of the estimation results presented in Fig.~\ref{fig:gene_activity}.}
+Interestingly, while most oscillations in $\gamma \alpha / \beta$ are canceled out at the population level, the first one after the upshift is conserved.
+Each graph gives the 25\% (lowest gray line), 50\% (solid black line) and 75\% (highest gray line) quartiles, computed at each time step for the signals reconstructed in Fig.~\ref{fig:gene_activity}.
+The gray area represents the interquartile range.
+}
+\label{fig:gene_activity_median}
+\end{figure}
+
+\section{Discussion}
+\label{sec:chap3_discussion}
+
+As we showed in chapter~\ref{chap:theory}, adopting a dynamical perspective might prove useful to unveil and understand the regulatory strategies employed by microorganisms.
+The criteria of biomass optimization has proved useful to uncover and explain the steady-state empirical growth laws.
+Interestingly, when the same criteria are applied in growth transitions, they suggest that microorganisms should converge toward a bang-bang control of gene expression.
+For the control of ribosome synthesis, this would result into implementing an \textit{on-off} regulatory system, either only producing ribosomes, or not producing them at all, during the adaptation to the new growth medium.
+The widespread ppGpp system, known for its role in the control of ribosome synthesis, is theoretically able to implement such a regulatory system.
+But we lack experimental data to show that it does so, mostly because of the difficulties raised by working in growth transitions.
+Can we measure the ribosomal synthesis rate at the single-cell level during well-controlled growth transitions?
+Would it exhibit the \textit{on-off} strategy that was shown to be close to optimal in chapter~\ref{chap:theory}?
+
+In this chapter, the above questions have been addressed by developing an experimental framework which allows the relative ribosome concentration to be estimated at the single-cell level in real time on \textit{E.
+coli}.
+Our main contribution consists into bringing together several experimental breakthroughs of the recent years.
+On one hand, inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we reconstructed a \textit{E. coli} strain displaying fluorescent, but still fully effective, ribosomes.
+In their paper, Bakshi \textit{et al} essentially used it to localize ribosomes in the cytoplasm through superresolution microscopy.
+While improving their design (see Material and Methods~\ref{sec:methods_strain}), we used it for the purpose of quantifying the ribosomal abundance.
+On the other hand, we used the mother machine~\cite{wang_robust_2010} developed by Wang \textit{et al} to observe this strain growing in continuous culture through real-time microscopy.
+This microfluidic device was originally developed to study long-term steady-state growth of \textit{E. coli}.
+Here, we used it to observe growth transitions in a well controlled framework, \textit{i.e.} from one steady-state growth to another steady-state growth.
+By changing the medium that was injected into the device, we were able to observe \textit{E. coli} transitioning from growth on acetate to growth on glucose.
+Overall, through the combination of fluorescent labeling of a core component of the gene expression machinery and the use of a microfluidic device, we opened the way to the experimental study of growth laws in a dynamical context.
+
+We applied this experimental set-up to the reconstruction of the resources allocation strategy employed by \textit{E. coli} during an environmental upshift.
+Obviously, the goal was to eventually observe the oscillations predicted in chapter~\ref{chap:theory}.
+We applied a signal processing algorithm known as Kalman smoothing and showed how it can be implemented to reconstruct the inner state of a cell from microscopy data.
+Similar to hidden markov chains, Kalman filtering and smoothing are powerful techniques allowing the reconstruction of hidden signals from noisy measurements, and have already numerous everyday applications, though not so much in biology.
+We showed in this chapter that the Bayesian logic behind Kalman smoothing can be exploited to inject in the growth rate reconstruction, a probabilistic prior that define the relation between the mother and its daughter cells.
+We also applied it for the reconstruction of $\alpha$, the abstract resource allocation parameter defined in chapter~\ref{chap:theory}.
+Our work on synthetic data showed that despite being challenged by the high stiffness of the signal, the algorithm was able to fully recover the oscillatory pattern that we expect to occur \textit{in vivo}.
+Unfortunately, its application on our dataset was not so successful: while oscillations can be reproduced, the signal is not strong enough to distinguish whether the oscillations are real or chimerical.
+We propose below several improvements that will certainly be addressed in future work.
+
+Several experimental problems dramatically impeded the analysis of the data.
+Sufficiently long steady-state measurements before and after the transition of interest are needed to provide a strong estimation of the intensity and the nature of the experimental noise.
+They are also critical to calibrate the parameters of the regularization method used in the signal reconstruction.
+Despite what was initially planned, we did not get long enough steady-state measurements far from the transition.
+Several causes are to blame.
+First, the random stall of the motorization along the Z-axis of the microscope, which do not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
+Secondly, the massive death of bacteria starting after the transition on glucose.
+Future work will focus on solving these issues.
+
+The death of bacteria is particularly worrying.
+During the construction of this strain, most people in the lab experienced contamination by an aggressive bacteriophage.
+Unfortunately, the strain used in this study was not spared, and our frozen stocks were later tested positive for phage-induced lysis.
+Interestingly, the lysis seems to depend strongly on the environmental conditions.
+To our disarray, the massive death observed in this experiment suggests that a strong environmental upshift could trigger the lytic cycle of the phage.
+While impeding any possible long-term measurement, the death of the bacteria is not our only worry.
+Phages are machines that are extremely well optimized to divert cell resources to their own end, especially the activity of the gene expression machinery.
+That could dramatically perturb the cell regulation of resource allocation, and so the conclusions of this study.
+For this reason, future work will start by the reconstruction of a clean phage-free strain from scratch.
+
+The data analysis technique used is also largely optimizable.
+The results reported here are largely preliminary.
+We segmented the images using the most powerful and ubiquitous segmentation algorithm available: the human brain.
+In other word, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a mask around the cell of interest.
+This is far from satisfying.
+Unfortunately, it seems that the first rule of image analysis is that every application is more or less unique.
+Every algorithm has to make assumptions about the object it is trying to recognize.
+For bacteria imaging, it could be the curvature at the poles, the size of the pixels, the homogeneity of the background, the homogeneity of the inner fluorescence...
+While some parameters can be tweaked, some assumptions are always hard-coded in the algorithm and reflect the philosophy used to address the identification problem.
+Since this preliminary analysis was made, we started a collaboration with the authors of FluoBacTracker~[citation, ask Hugues if a publication exist] that focuses on adapting their software to our set-up in a near future.
+It should provide a robust, automatic way to identify and track the cells, enhancing reproducibility of the analysis.
+
+The signal reconstruction algorithm that was later applied to these data is also largely optimizable.
+Like we described in the previous section, the algorithm is an instance of Bayesian reconstruction.
+Several parameters describing the expected input define a probabilistic prior.
+Preliminary, we chose these parameters through trials and errors, or by calibrating the system on synthetic data when it was possible.
+A better technique would be to choose these parameters by \textit{generalized cross-validation} (GCV)~\cite{golub_generalized_1979}, a procedure that maximizes the predictive power of the reconstructed signal.
+More generally, parameters could be optimized on a subset of the cells, then tested against the remaining subset.
+It would ensure that the algorithm does not overfit the behavior of a few cells.
+While the initial conditions did not seem to play a huge role in the final shape of the signal, the choice of the smoothing factor $\theta^2$ appeared to be critical, and would strongly benefit from such an implementation.
+
+We also did not fully exploited the power of Kalman smoothing.
+Smoothing methods have trouble estimating abrupt transitions, and our implementation has the same issue.
+As we can see in Fig.~\ref{fig:synthetic_upshift}, the reconstructed signal for $\gamma \alpha / \beta$ increases before the shift actually happen.
+This artifact also appears for the real data (Fig.~\ref{fig:gene_activity_median}), were the signal start to change dozens of minutes before the shift.
+We know exactly when bacteria start to be in contact with the new medium, but this information is never used as prior for the reconstruction.
+Using this information is however possible with Kalman smoothing.
+By implementing time-varying parameters for the regularization method, we can strongly penalize the variations of the reconstructed signal at steady state (when nothing is expected to happen) and release the constraints as soon as the medium is changed (when we expect most of the stuff to happen).
+Such a method has already been successfully applied to correctly estimate changes in growth rate during sugar exhaustion~[citation in preparation Eugenio].
+
+We are convinced the improvements cited above would allow to robustly conclude on the existence of the on-off strategy, and are not excessively hard to implement.
+Other more challenging developments are planed and would probably benefit to the framework as a whole.
+For instance, we did not discuss the autofluorescence of the bacteria in this chapter.
+It was assumed negligible, mostly because its proper estimation in a dynamical, single-cell context is complicated (see Material and Methods~\ref{sec:cell_segmentation}).
+Nevertheless, a simple trick could solve the issue: merging wild-type and fluorescent bacteria when inseminating the device.
+The goal would be to have a significant part of the channels occupied by wild-type cells, while still being able to observe in parallel the strain of interest.
+That way, we could have an exact estimation of the level of autofluorescence of the cells, and robustly correct it for the channels of interest.
+Note that this would reduce the amount of exploitable channels, hence would probably require several adjustments.
+
+Something is also displeasing in the current implement of the Kalman smoothing method.
+As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
+This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
+We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
+There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
+
+\section{Material and Methods}
+
+\subsection{Bacterial strain construction}
+\label{sec:methods_strain}
+
+To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} to the C-terminus of \textit{rpsB} (the gene coding the ribosome subunit S2).
+This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.~coli}~\cite{bakshi_superresolution_2012}.
+
+The transcription factor \textit{tsf} is under the control of the same promoter than \textit{rpsB}, and is located after the C-terminus.
+In order to ensure that \textit{tsf} expression was not affected by our translation fusion on \textit{rpsB}, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
+This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
+
+A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
+The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
+This cassette is referred below as \textit{kan-P\textsubscript{BAD}-ccdB}.
+
+Another DNA fragment containing \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} without ATG was amplified with long primers annealing respectively to the C-terminus of \ref{chap:theory}rpsB (just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+The first primer also contained a 18-bp (base pair) linker that was chosen according to the method section of~\cite{bakshi_superresolution_2012}.
+
+Both fragments (the \textit{gfpmut2} reporter and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette) were assembled using Gibson assembly~\cite{gibson_enzymatic_2009}.
+Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
+A final product of 2683 bp was obtained:
+\begin{itemize}
+\item *(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
+\item (18 bp) a linker (see Supporting Information)
+\item (714 bp) the \textit{gfpmut2} sequence without the initial ATG
+\item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette
+\item *(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli}
+\end{itemize}
+Regions labeled with * are expected to anneal with the \textit{E.~coli} chromosome.
+The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
+
+This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid with chloramphenicol resistance (lambda-red recombinaison).
+A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
+A new 100-bp fragment containing 50 bp of the end of \textit{gfpmut2} and 50 bp of the region just after \textit{rpsB} was electroporated into this strain (sequence available in Supporting Information) to get rid of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
+Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
+The region after \textit{rpsB} was verified by sequencing (full sequence available in Supporting Information).
+
+In parallel, the same protocol was used to construct \textit{mCherry} and \textit{cfp} variants of the same strain.
+However, only the \textit{gfpmut2} and \textit{mCherry} versions were successfully obtained for reasons that were not investigated.
+The full sequence of the final \textit{rpsB-mCherry} strain is available in Supporting Information.
+
+The \textit{rpsB-gfpmut2} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
+They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
+However, the \textit{rpsB-mCherry} strain exhibited strange fluorescence dynamics, especially during growth transitions, which made it unsuitable for our study.
+Our effort were then concentrated on the \textit{rpsB-gfp} strain.
+Data as well as information on the matter are available in Supporting Information.
+
+\subsection{Growth conditions}
+\label{sec:growth_condition}
+
+Sterilization was performed by autoclaving for instruments and filtration for solutions (0.2 $\mu$m).
+During cloning, bacteria were grown in 20-mL flasks filled with LB or spread on Petri dish with LA.
+In all the experiments, we used M9 minimal medium~[citation to find] supplemented with trace elements, thiamine and a carbon source.
+The full recipe is reproduced below.
+Numbers in squared brackets are characteristics of the stock solution.
+All stock solutions were stored at room temperature, except for FeSO$_4$~[30 g.L\textsuperscript{-1}] and Thiamine~[10 g.L\textsuperscript{-1}] that were stored respectively at -20$^\circ$C and 4$^\circ$C.
+\begin{center}
+\begin{tabular}{|l r|}
+  \hline
+  \multicolumn{2}{|c|}{\textbf{M9 medium (100 mL)}}\\
+  \hline
+  CaCl2 [1 mol.L\textsuperscript{-1}] & 10 $\mu$l\\
+  MgSO$_4$ [1 mol.L\textsuperscript{-1}] & 200 $\mu$l\\
+   5x Salts & 20 mL \\
+  Traces elements & 90 $\mu$l \\
+  FeSO$_4$ [30 g.L\textsuperscript{-1}] & 10 $\mu$l \\
+  Thiamine [10 g.L\textsuperscript{-1}] & 50 $\mu$l \\
+  Carbon source & at will \\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+  \hline
+\end{tabular}
+\begin{tabular}{|l r|}
+  \hline
+  \multicolumn{2}{|c|}{\textbf{Traces elements (0.9 mL)}}\\
+  \hline
+  H$_2$0 [18.2 M$\Omega$.cm] & 200$\mu$l\\
+  Na$_2$EDTA 2H$_2$O [150 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  ZnSO$_4$ 7H$_2$O [45 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CoCl$_2$ 6H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  MnCl$_2$ 4H$_2$O [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  H$_3$BO$_3$ [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  Na$_2$MoO$_4$ 2H$_2$O [4 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CuSO$_4$ 5H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  \hline
+\end{tabular}
+\begin{tabular}{|l r|}
+    \hline
+  \multicolumn{2}{|c|}{\textbf{5x Salts (100 mL)}}\\
+  \hline
+  Na$_2$HPO$_4$2H$_2$0 & 4.25 g \\
+  KH$_2$HPO$_4$ & 1.5 mg \\
+  NaCl & 0.25 g\\
+  NH$_4$Cl& 0.5 g\\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+  \hline
+\end{tabular}
+\end{center}
+
+A large quantity of M9 was prepared several days before the experiment, and stored at 4$^\circ$C.
+It contained all the necessary components except FeSO$_4$, thiamine, and carbon sources that were added the day the pre-culture were launched.
+Except stated otherwise, the growth temperature was 37$^\circ$C.
+M9 Acetate contains 0.2\% Acetate (in mass of C$_2$H$_3$O$_2$ per mass of solution), and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per mass of solution).
+
+At J-4, the glycerol stock containing the \textit{rpsB-gfp} strain was spread out on a Petri dish of LA, and incubated at 37$^\circ$C.
+At J-3, an isolated colony was inseminated in 20 mL of M9 Acetate, and incubated in a cotton-plugged sterile flask (Fig.~\ref{fig:experiment_schema}).
+Time of insemination was calculated to obtain a 0.3-0.4 cm\textsuperscript{-1} OD when starting the experiment, which correspond to a mid-exponential phase.
+
+At J+0, the preculture was concentrated by centrifugation and back suspended into 5 mL of M9 Acetate supplemented with 50 mg.mL\textsuperscript{-1} BSA (passivation	buffer) for injection into the microfluidic device.
+Channels were populated by diffusion until most of them were filled with cells ($\sim$1 hour).
+Data acquisition started after a constant medium flow was successfully obtained ($\sim$1 hour, depending on the quality of the device).
+
+\subsection{Microfluidic device}
+\label{sec:microflu}
+
+We used the mothermachine~\cite{wang_robust_2010}.
+It consists of a series of growth wells (or channels), oriented at a 90$^\circ$ angle to a large central channel through which growth medium is passed at a constant flow.
+The wideness of the wells is selected as to constrain the growth direction of the bacteria.
+This design ensures that at least one cell per well (the deepest one) is preserved during the whole experiment, while the others incrementally escape into the central channel as divisions occur.
+For the fabrication of the devices, we thoroughly followed the step described in the supporting information of~\cite{wang_robust_2010}.
+We maintained a stock of chemically treated devices at room temperature (day 2 in workflow summary~\cite{wang_robust_2010}).
+The day of the experiment, a single device was plasma cleaned, bond to a glass coverslip, and injected with bacteria (see section above).
+
+The device was connected using 0.023" inner diameter polyethylene tubes to a waste and a sterile bottle containing 200 mL of growth medium, which was enough for several days of acquisition.
+A microfluidic pump (Elvesys) containing an output flow sensor module was plugged to the medium bottle, and was applied a pressure up to 2 bars.
+The output flow was set to 50 $\mu$L.min$^{-1}$.
+
+For the imaging, the device was placed on a motorized inverted microscope (Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^\circ$C.
+In this setup, fluorescence illumination is provided by a mercury lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma, \#49005 TR/DsRED ET).
+The exposure time to the light source is externally controlled by mechanical shutters (Uniblitz-VS35).
+Images are acquired with a 16-bit gray level CCDcamera cooled to -80$^\circ$C (Roper Scientific, Princeton Instruments PHOTOMAX 512) controlled by a custom-made software using Visual Basic and the Type libraries of the Winview software (Princeton Instruments).
+Every 5 minutes, autofocus was numerically performed by maximizing the contrast of a region of interest, and a serie of acquisitions was made.
+A total of 6 fields, each containing 15 wells was observed during the whole experiment.
+
+\subsection{Cell segmentation}
+\label{sec:cell_segmentation}
+
+Raw data are in the form of 512x512-pixel 16-bit SPE images (proprietary format produced by the Princeton Instruments camera).
+They were converted in 16-bit Tif images using the \textit{SPE} plugin of ImageJ~\cite{goto_open_2005}.
+The rest of the analysis was performed using Python 3.5.2.
+In particular, we used OpenCV~3.1.0-dev, Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
+
+The offset between consecutive images of a given field was evaluated using cross-correlation in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see Scikit-Image documentation at \cite{skimage_cross-correlation}).
+All the images for a field were then aligned on the first acquisition by simple affine translation.
+
+The wells were isolated in individual images through a combination of manual pixel selection and automatic segmentation.
+In particular, we selected the entrance of the two wells at the border of the image, and used this information along with the regularity of the microfluidic device to compute a mask that allowed to crop each well in its own image of size 100x21 pixels.
+These sub-images were labeled $\left\{W0, W1, ..., W14\right\}$ depending on the position of the well on the original image, from left to right.
+
+In this study, we performed a preliminary analysis on each well that focused on the deepest cell.
+For each image, we manually selected two pixels representing the position of the poles of the cell of interest.
+The distance between the two selected pixels was directly used as the bacteria length $L$ in pixels.
+They were also used to compute a rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
+We then computed the RFU/pixel in this cell mask by summing each pixel and dividing by the mask size.
+
+The camera noise and background was evaluated by taking a picture with a closed shutter at the end of the experiment.
+Pixels in this picture were found to be Gaussian distributed with a mean of 1101.0 and a standard deviation of 10.788.
+Correction for camera background was thus applied by removing 1101 to the computed RFU/pixel.
+
+The background for autofluorescence of the bacteria and the medium was supposed negligible and was thus not corrected.
+Indeed, the same device was used to image bacteria in stationary phase that do not produce any fluorescent protein.
+They were entirely indistinguishable from the camera noise.
+Not that this does not ensure that the autofluorescence is also negligible when the bacteria are actively growing.
+Furthermore, its value could change with the medium of interest, and even be different at steady state and during growth transitions, which does not allow to estimate it independently.
+In the section~\ref{sec:chap3_discussion}, we discuss possible improvements of the experimental set-up that would allow to dynamically co-estimate the autofluorescence with the ribosome abundance in a single experiment.
+
+\subsection{Kalman smoothing}
+\label{sec:meth_kalman}
+
+The data about the length and the RFU/pixel of each bacteria were analyzed using Python~3.5.2.
+In particular, we manipulated the data using Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
+
+Historical details about the Kalman smoothing procedure are reported in section~\ref{sec:res_kalman} of the main text.
+We used the Additive Unscented Kalman Filter implementation of the pykalman python module.
+This class is reported to be more stable and computationally efficient with non-linear problems featuring additive noises.
+
+As described in section~\ref{sec:res_kalman}, the reconstruction of the growth rate was done on continuous portions of the length, \textit{i.e.} between two division events.
+The full problem, as reported in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity is:
+\begin{eqnarray*}
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
+\dot{\mu}(t) &=& v(t),\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray*}
+with the measurement model:
+\begin{equation*}
+L(t_k) = V_\lambda(t_k) + \epsilon_k.
+\end{equation*}
+The parameters used as prior for the reconstruction of $\mu$ are reported below.
+We used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
+Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
+At the beginning of the experiment, were no mother cell are available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
+All the cross-covariances are set to 0 because the system variables are independent by construction.
+
+For the reconstruction of gene activity, we only had to cope with the large gap in data acquisition between -720 and -150 min.
+Reconstruction was then done independently on the continuous sections before and after this gap.
+The full problem for the estimation of $u = \gamma \alpha / \beta$, as reported in Eq.~\ref{eq:full_u_prob}, is reproduced here for clarity:
+\begin{eqnarray*}
+\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
+\dot{u}(t) &=& v(t),\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray*}
+with the measurement model
+\begin{equation*}
+F(t_k) = r_\gamma (t_k) + \eta_k,\\
+\end{equation*}
+where $\hat{\mu}$ is the estimation of $\mu$ above.
+In both case, we used as prior an observation variance of 800.4 RFU\textsuperscript{2} for $F$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\gamma \alpha / \beta$) is fixed at $10^2$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
+The initial state means used are 600~RFU for $r_\gamma$ and 1000~RFU for $\gamma \alpha / \beta$.
+The variances associated with these means are purposely large and fixed to 10\textsuperscript{6}~RFU\textsuperscript{2} for $r_\gamma$ and $\gamma \alpha / \beta$.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2}, imposing a null second derivative for the reconstructed signal $\gamma \alpha / \beta$.
+Here again, all the cross-covariances are set to 0.
+
+All the parameters cited in this section were chosen through trials and errors, and are as a consequence largely optimizable.
+Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+
+\section{Supporting Information for Chapter~3}
+
+(Plan not complete)
 
@@
 
-We used Gibson assembly, Kan-pbad-ccdb, etc and obtain the following sequence.
+Primers used for gibson assembly
 
-\subsection{Microfluidic device}
+\subsection{Strain validation using growth in microplate reader}
 
-Mother machine because needed to change the medium, single cell data because asynchronization etc.
+(signal-to-noise ratio, growth rate not hampered in microplate reader)
+Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
 
-\subsection{Cell segmentation}
+\subsection{Microscopy analysis}
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+Show the complete analysis for 5 cells.
 
-\subsection{Kalman smoothing}
+\subsubsection*{Category of cells}
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_cells}
+\caption{Subtypes of cells. Caption to write. Cutting the x axis might be a good idea because a third of the x-axis of each graph are "missing data".}
+\label{fig:subcat_cells}
+\end{figure}
 
-\section{Results}
-
-\subsection{Model calibration}
-
-We want to get $\alpha (t)$ during a transition. We show that $\alpha$ can be obtained if we measure so and so.
-
-\subsection{Bang-bang expression during transitions, as predicted}
-
-One stunning figure that show the RFU / cell / pixels, and the predicted $\alpha$ for the glucose-acetate transition.
-
-\subsection{Discrepancies with the model and model adaptation}
-
-The growth rate stalls after the addition of glucose, which was not originally predicted by the model.
-This is because so and so.
-This can be taken into account if we add another sector (M1 and M2).
-This raises a new interesting question : during the transition amino-acids drop, so the cell should produce ribosomes, but it also need to produce the new enzymes to actually be able to perform metabolism.
-We thus have competition between several layers of regulation, which can be added to the model by doing so and so.
-
-\section{Discussion}
-
-More research need to be done.
-
-Could be interesting to use transitions for syn bio purposes.
-Could be interesting to validate such results in different microorganisms, with different type of medium (fast env, slow env, etc)
-
-\section{Supporting Information}
-
-\begin{itemize}
-\item Strain validation (signal-to-noise ratio, growth rate not hampered in microplate reader)
-\item Primers used for gibson assembly
-\item Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
-\end{itemize}
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_median}
+\caption{Subtypes of cells, median. Caption to write.}
+\label{fig:subcat_median}
+\end{figure}

### 2016-12-17 Chapters/introduction.tex
Add space for french summary everywhere. Comment out chapter 2 abstract because chapter 3 has no abstract.
Finishing chapter 3, sent to Hidde and Hans

@@
 
-\section*{Summary}
+\selectlanguage{french}
+\section*{Résumé en français}
 
-This chapter presents... (approx. 1 page)
-
-\selectlanguage{french}
-\section*{Résumé}
-
-Dans cette section... (env. 1 page)
+Dans cette section... (env. 2 pages)
 \selectlanguage{english}
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
-\includegraphics[height=3cm]{./Fig/Chapter1/monod_law.eps}
-\caption{Monod law, reproduced from~\cite{monod_growth_1949}.
+\includegraphics[height=5cm]{./Fig/Chapter1/monod_law.eps}
+\caption{
+\textbf{Monod law, reproduced from~\cite{monod_growth_1949}.}
 Monod has been a pioneer in the formalization of microbial growth into fundamental coarse-grained relationships.
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
-\includegraphics[height=6cm]{./Fig/Chapter1/scott_rnaprot.eps}
-\caption{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).
+\includegraphics[height=7cm]{./Fig/Chapter1/scott_rnaprot.eps}
+\caption{
+\textbf{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).}
 For different carbon sources more or less rich, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
-\includegraphics[height=5cm]{./Fig/Chapter1/molenaar_overflow.eps}
-\caption{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.
+\includegraphics[width=\textwidth]{./Fig/Chapter1/molenaar_overflow.eps}
+\caption{
+\textbf{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.}
 When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
@@
 
-\begin{figure}[!h]
+\begin{figure}[tb]
 \centering
 \includegraphics[height=5cm]{./Fig/Chapter1/growth_curve.eps}
-\caption{The different phases of a typical growth curve are visualized by the accumulation of biomass (black thick line) and the nutrient abundance in the medium (green dashed line).
+\caption{
+\textbf{The different phases of a typical growth curve are visualized by the accumulation of biomass (black thick line) and the nutrient abundance in the medium (green dashed line).}
 (A) The lag phase, a variable period of time during which the organism adapt to the new medium.

### 2016-12-19 titlepage.tex
Change order in jury

@@
 \Jury{
-\UGTPresident{M, Olivier BERNARD}{Chargé de recherche, Inria, Nice -- Sophia-Antipolis (à confirmer pour président)}     %% 2ème examinateur
+%\UGTPresident{M, Olivier BERNARD}{Chargé de recherche, Inria, Nice -- Sophia-Antipolis (à confirmer pour président)}     %% 2ème examinateur
 %\UGTPresidente{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
@@
 
+\UGTExaminateur{M, Olivier BERNARD}{Chargé de recherche, Inria, Nice -- Sophia-Antipolis}     %% 2ème examinateur
 \UGTExaminateur{M, Matthieu JULES}{Chargé de recherche, Inra, Grignon}     %% 1er examinateur
@@
 
-\UGTDirecteur{M, Johannes GEISELMANN}{Professeur, Université Grenoble Alpes, LIPhy}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde DE JONG}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
 
@@
 %\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
+
+\UGTDirecteur{M, Johannes GEISELMANN}{Professeur, Université Grenoble Alpes, LIPhy}       %% Directeur de thèse
+\UGTCoDirecteur{M, Hidde DE JONG}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
 }

### 2016-12-21 Chapters/discussion.tex
Add space for french summary everywhere. Comment out chapter 2 abstract because chapter 3 has no abstract.
First final draft of discussion

@@
 
-- in praise of simple models
-- an intrinsic link between complexity and environment dynamic
-- some perspectives for industrial production
+\textit{"Apples fall onto the Earth because natural selection eliminated apples falling towards the sky."} -- Source Unknown
+
+\selectlanguage{french}
+\section*{Résumé en français}
+
+Dans cette section... (env. 1 page)
+\selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
+Physics has yet to be reduced to a formula that will fit on a piece of clothing~\cite{falk_universe_2005}.
+Biology is even more behind, but the extensive work on the mechanisms of evolution are helping to close this gap, the Price equation being the closer we have of such an elegant formula~\cite{frank_natural_2012}.
+But evolution by itself is not sufficient to define life.
+Even the less restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
+Self-sustainment appears to be as fundamental as evolution, and rely on the transformation of matter and energy from the environment, in other word, growth.
+But we are far from being as mathematically advance as in evolution on this question.
+
+Fundamental growth laws have been established by focusing on the life of microorganisms.
+They show that, regardless of the molecular implementations, microorganisms tend to follow the same empirical regularities when grown at steady state in different kind of environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
+In particular, they change their inner composition, and do so by adjusting the allocation of resources from the environment, in a way that appears to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
+Growth laws are a strong support for the theory of a modular organization of microorganisms~\cite{scott_emergence_2014,hartwell_molecular_1999,arkin_fast_2006,guido_bottom-up_2006}.
+They represent a big step towards the general understanding of the process of growth.
+They are, however, established at steady state, a state in which most microorganisms spend very little~[citation].
+Why would microorganisms be optimal for a state they rarely encounter?
+Would known growth laws be specific cases of more general laws that implement the dynamics of the environment?
+
+Our goal in this manuscript was to establish a theoretical and experimental framework that would help to extend growth laws in a dynamical context.
+We focused on the growth law of resource allocation, describing how the ribosome abundance seems to adapt in different media as to maximize growth.
+Applying the same criteria, what is the best way to allocate resources during a growth transition between two different environments?
+In chapter~\ref{chap:theory}, we used a simple self-replicator model of resource allocation, and showed that the steady state is insensitive to the precise nature of the regulatory schemes: optimization of growth rate can be achieved by measuring either the environment or the cell variables.
+Only the optimization of biomass production over a growth transition does truly require information about the inner state of the growing cell.
+The steady-state growth laws suggested this information to be implemented into a precise control of the intensity of gene expression.
+Interestingly, far from the steady state, we observe exactly the contrary.
+The cell would rather take advantage of a strongly non-linear system that achieves on-off switches of the expression level.
+Overall, when the biomass produced is at play, we observed a divergence between steady state and dynamical considerations of optimality, which does not mean that a single system cannot satisfy both.
+
+The model developed here is an instance of proof-of-concept model~\cite{servedio_not_2014}.
+It does not necessarily aim at predicting or controlling the behavior of microorganisms.
+By using a simple and abstract representation, it provides a convenient and tractable way to evaluate the implications of dynamical optimality, in particular their divergences with steady-state optimality.
+Is a dynamical perspective on growth laws necessary?
+Do regulatory mechanisms need to differ in a dynamical context?
+Those are the verbal hypothesis that were tested by this model.
+To our surprise, it additionally provided a new way of seeing the regulatory systems controlling the ribosomal abundance in most bacteria.
+
+Some readers might be skeptic about the simplicity of the model.
+Given the profusion of knowledge and data available on biological systems, one is often tempted to introduce every known details about the molecular implementation of a system of interest.
+One of the main drawbacks of such big models are that we can quickly lose track of their underlying hypothesis.
+When modeled in details, biological system also tend to become untractable, prohibiting the use of mathematical tools that are quickly overwhelmed by the size of a dynamical system (optimal control being one of them).
+Besides, simple, abstract models make it easy for the modeler to state the critical, exploratory, and logistical assumptions that are made~\cite{servedio_not_2014}.
+For instance, an exploratory assumption of our model was to consider that the proposed strategies had to be also optimal at steady state.
+This was used as a necessary step to reduce the space of possible solutions, and make the comparison between different strategies time independent.
+Relaxing this assumption would require to explore richer environments than the simple upshift, but could probably raise interesting considerations~\cite{geisel_constitutive_2011,lopez-maury_tuning_2008,lambert_memory_2014,kussell_phenotypic_2005}.
+
+But the assumptions that need most to be discussed are the critical ones, i.e. the ones that would invalidate the conclusions drawn if falsified.
+In our case, it is the assumption that biomass production is the main fitness factor in stable or fluctuating environments.
+Through the manuscript, we did not really challenge this hypothesis, but only provided arguments about why it is so accepted.
+While it is clear that microorganisms can be cultivated, and even naturally found in conditions where maximizing growth rate ensure their persistence, we have no direct proof that biomass production is what allowed them to naturally persist on evolutionary timescales.
+In addition, as illustrated by the quote at the beginning of this chapter, one must be excessively careful when explaining the behavior of living system solely through natural selection considerations.
+Such systems are embedded in the physical world, and evolutionary arguments are not always relevant to explain an empirically observed behavior.
+
+The second part of this manuscript was dedicated to setting up an experimental benchmark to study growth transitions.
+It was applied to the question of dynamical resource allocation to emphasize the limitations that affect the study of transitions, and to propose possible workarounds.
+Fluorescent reporter proteins were used to cope with the need for \textit{in vivo} measurements with a high time resolution.
+Advances in microfluidics were exploited to realize steady-state to steady-state transitions from one medium to another, in order to temper the effect of strain history and enhance reproducibility.
+Finally, the signal of interest was reconstructed through the use of Kalman smoothing, a powerful signal processing algorithm that has many features that make it suitable for microscopy time series.
+Overall, the set-up still need many improvements, though most of them are experimental peculiarities that are just here to remind us that lab work is excessively harsh and merciless.
+
+Chapter~\ref{chap:experiments} must not be taken as a test of the model in chapter~\ref{chap:theory}, but rather as a test of its assumptions (see box~2 in~\cite{servedio_not_2014}).
+We apply the experimental set-up to empirically test the predictions of bang-bang expression during a growth transition.
+This prediction followed directly from the critical assumption that microorganisms maximize their biomass production at steady state and over growth transitions.
+While we were not able to conclude in this manuscript, we can project ourselves in the two possible scenarios that will arise from the future improvements of the experimental set-up.
+
+The oscillatory nature of resource allocation during a growth transition could be confirmed.
+This would be a strong argument in favor of an adaptation of microorganisms to changing environments, hence to the existence of more general dynamical growth laws.
+As was performed to establish the steady-state growth laws, we would need measurements in broader environmental conditions.
+The most straightforward would be to test additional upshift and downshift schemes (others sugars, amino-acids supplementation, azote instead of carbon starvation...).
+The experimental set-up could also be slightly modified to enrich the conditions controlled for.
+For instance, the strain could be modified to display an additional fluorescent protein that would report the abundance of a key enzyme of the metabolism.
+Along with the ribosomal reporter, we could control for the expression in anti-phase of the metabolism and the gene expression machinery during the oscillations.
+
+The other scenario, \textit{i.e.} the absence of oscillations, would definitively falsify the hypothesis of the model.
+Generally in that case, the problem becomes to identify which critical assumptions are not met in nature.
+We also cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might be in fact critical for the predictions of the model.
+Though unlikely, this could be the case of the expressions used for the macroreactions in the model (Eqs~\ref{eq:metaflux}-\ref{eq:machflux}), the definition of the volume as invariably proportional to the total mass of macromolecules (Eq.~\ref{eq:voldef}), or the fact that degradation is assumed negligible with respect to the rates of other reactions in the system.
+More probably, the assumption regarding the cost of the regulatory system might prove more critical than initially thought.
+By only comparing the regulatory schemes \textit{via} their benefits, we implicitly made the assumption that their cost are negligible, or at least comparable in regards of the difference in benefits.
+The costs of regulatory schemes are actually difficult to model, as organisms always prove more and more inventive when it is about minimizing it~[citation Alon modeling cost, citation free regulation somewhere in my mailbox].
+All variables in the system are not equivalently costly to measure, and as we showed with the ppGpp system, their combinations do not make their cost necessarily additive.
+Nevertheless, relaxation of this hypothesis would definitively need to be explored before concluding that microorganisms do not optimize their biomass production during transitions.
+
+In any case, exploring the cost of the regulatory mechanisms might uncover a fundamental link between the complexity of the regulatory schemes and the dynamics of the environment.
+As we illustrated in chapter~\ref{chap:theory}, complex regulatory schemes only prove beneficial in a dynamical context, far from the steady-state growth.
+By projection, one could expect the benefits of complexity to saturate at a level that depends on how much \textit{dynamic} the environment is, a definition that as yet to be established.
+On the contrary, the cost of complexity would probably not saturate, since acquiring more information is expected to require the production of more signaling molecules, diverting an ever growing part of the cell resources.
+Overall, we could expect these two functions to cross at a point that would represent an evolutionary bottleneck, beyond which the costs outclass the benefits (see~\cite{short_flows_2006} for another example of such a bottleneck).
+Interestingly, the position of the bottleneck would be environment-dependent, establishing a dynamic law that would link the complexity of the regulations with the dynamics of the environment in which the organism evolved.
+In other words, while the current growth laws consider how the physiology of the cell vary with the nutrient quality of the medium, we could add another dimension that represents the actual availability of this nutrient in the natural environment...

### 2016-12-21 titlepage.tex
Some changes on the title page

--- 2016-09-05 titlepage.tex
@@
 
+\UGTExaminateur{M, Olivier BERNARD}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}     %% 2ème examinateur
 \UGTExaminateur{M, Matthieu JULES}{Chargé de recherche, Inra, Grignon}     %% 1er examinateur
-\UGTExaminatrice{Mme, Irina MIHALCESCU}{Maître de conférence, Université Grenoble Alpes, LIPhy}    %% 3ème examinateur
+\UGTExaminatrice{Mme, Irina MIHALCESCU}{Professeur, Université Grenoble Alpes, LIPhy}    %% 3ème examinateur
+
+\noindent\hfil\rule{4cm}{0.1pt}\hfil
+
+\vspace{0.3cm}
 
@@
 \UGTCoDirecteur{M, Hidde DE JONG}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
-
 \UGTInvite{M, Jean-Luc GOUZÉ}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}

### 2016-12-22 Chapters/introduction.tex
Rework section 1.1 of intro using Hidde's comments

@@
 
-Dans cette section... (env. 2 pages)
+Dans cette section... (env. 1-2 pages)
 \selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
 
@@
 At every moment of our lives, we are surrounded by billions of microscopic organisms.
-Sometimes beneficial, sometimes harmful, these small factories collect and use energy from their environment, in the form of organic matter, highly reactive chemicals, or light [citation trophic].
-While most cells in our body are kept in a stable and friendly micro-environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to a strongly variable environment.
-Microbes are found almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life [citation deep water bacteria, citation bacteria in space, citation review bacterial diversity].
-Closer to us, the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different sugars through changes in the expression of thousands of genes\cite{zimmer_microcosm:_2009}.
+They sustain themselves by collecting energy from their environment, in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
+While most of our cells are kept in a stable and friendly internal environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to strongly variable environments.
+Not without a certain success, because microbes have been found to survive almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life~\cite{rothschild_life_2001,nicholson_transcriptomic_2012,madigan_biology_2006,schaechter_microbe_2006}.
+Such ubiquity translates a strong potential for adaptation.
+Even the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes\cite{zimmer_microcosm:_2009}.
 
-Polymers are the most abundant organic molecules that constitute a living cell.
-They consist of many repeated similar sub-units (monomers).
-There are about 20 different amino acids in the proteins, and 5 different nucleobases in the (desoxy-)ribonucleotids that constitute RNA and DNA (the famous A, T, C, G letters and the extra U only found in RNA).
-The information about their sequence is stored in the form of genes on the DNA.
-The process by which this information is used in the synthesis of a functional final product (protein or RNA) is called gene expression.
-Each gene contains a target sequence that allows the gene expression machinery to recognize it and start the synthesis of its product.
-This sequence is called a promoter.
-If every promoter were the same, all genes would be expressed at the same level.
-That of course does not happen.
-In practice, some promoters are stronger than others, and attracts more of the available transcriptional machinery.
-Furthermore, there exist gene products that bind the promoter region of genes (transcription factors), stimulating or inhibiting the promoter affinity, hence the gene expression.
-This simple mechanism, among others, allows the regulation of cell composition.
+Microorganisms, along with all living cells, are mainly constituted of polymers (DNA, RNA, Proteins) made of many repeated sub-units (nucleotides, amino acids).
+The information about their sequences is stored in the form of genes on the DNA.
+Gene expression is the process by which this information is used in the synthesis of a functional final product (protein or RNA).
+The gene expression machinery (itself constituted of RNA and proteins) synthesizes the gene products after binding to gene promoters, which are target sequences on the DNA.
+In a given cell, they are roughly as many different target sequences as genes, which ensures that not all products are synthesized at the same rate.
+Furthermore, some of these products specifically bind some promoters and stimulates or represses the promoter activity, hence, the gene expression.
+They are called transcription factors and allow, among other mechanisms, the regulation of cell composition.
 
 The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyzes the biochemical reactions allowing the cell to perform different functions.
-For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reductive power, nucleotids and amino acids~\cite{schaechter_microbe_2006}.
+For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reductive power, nucleotides and amino acids~\cite{schaechter_microbe_2006}.
 These precursors are then used to produce new macromolecules, closing the loop of self-replication.
-In order to optimize the overall process, the cell must adjust the abundance of every enzyme involved in this network of biochemical reactions.
+In order to optimize the overall process in different environmental conditions, the cell must adjust the abundance of every enzyme involved in this network of biochemical reactions.
 By doing so, the cell actually resolves a resource allocation problem, where a common wealth (precursors) has to be distributed over thousands of potential beneficiaries (proteins and other macromolecules).
@@
 
-What is the optimal allocation of precursors for a living cell?
-Under Darwin's theory, the optimization criteria of living systems are called fitness factors, and measure to which extend an organism fits into a given environment.
-Fitness factors take into account the ability to survive and reproduce based on the adaptation to the environment and the competition with other organisms.
-Whether fitness factors apply at the species, the population, the organism, or even the gene level is still highly debated~\cite{dawkins_selfish_1976}.
-This makes their clear identification highly difficult, even for organisms living an apparently simple life.
+Which resource allocations are optimal for a living cell?
+Systems capable of Darwinian evolution sustain themselves on the long term by \textit{fitting} to their environment.
+Living cells with the best \textit{fitness} are more likely to survive, replicate, and through long-term competition, eliminate their siblings.
+The optimization criteria that quantitatively measure \textit{fitness} are called \textit{fitness factors}.
+Whether they apply at the species, the population, the organism, or even the gene level has long been debated~\cite{dawkins_selfish_1976}.
+An optimal resource allocation would be a gene expression scheme that maximizes the fitness factors.
+But since fitness factors have to take into account the ability to survive, reproduce, and compete with other organisms in a given (sometimes changing) environment, they are often context-dependent and highly difficult to identify.
 
@@
 Competition for common resources seems to favor the maximization of offspring, in the sense that a replicator with more descendants will outnumber its competitors in the long term~\cite{dawkins_selfish_1976}.
-It is unclear, however, how this principle translates into the physiological behavior of microorganisms.
-For instance, metabolic networks seems to either maximize ATP or biomass yield depending on the environmental condition, a result obtained by looking at the predicting power of genome-scale metabolic networks~\cite{schuetz_systematic_2007}.
-But, yield is not an universal fitness factor, because in fluctuating or limited environments, evolution seems to favor bacteria with a high rate of nutrient assimilation, instead of a high yield of transformation even though it would favor the net biomass production of the population~\cite{frank_tradeoff_2010,maclean_tragedy_2008}, a result that has also been confirmed in yeast~\cite{schuster_maximization_2008}.
+It is unclear, however, how this general principle translates into their physiological behaviors.
+For instance, the use of genome-scale metabolic network models have shown that the metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental condition~\cite{schuetz_systematic_2007}.
+But we can also found different strategies when the biomass production is at play.
+Bacteria can either optimize their production yield (taking its time to make the most of the nutrient) or their production rate (growing quickly while wasting part of the nutrient potential), the later being favored in fluctuating or poor environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008}, a result that has also been confirmed in yeast~\cite{schuster_maximization_2008}.
 Overall, it seems that cell optimality drifts in a multidimensional space and involves making several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
@@
 Nevertheless, numerous studies have shown that considering growth rate maximization confers a good predictive power of microorganism physiology.
-Uptake rates of \textit{Escherichia coli} growing on acetate or succinate are correctly predicted by assuming that its metabolic network operates in a mode that maximizes growth~\cite{edwards_silico_2001}.
-This result is not general and can initially fail for other carbon sources or strain~ \cite{ibarra_escherichia_2002}.
-But even in those cases, the metabolic networks evolve rapidly towards growth maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
-The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not solely the efficiency of the enzymes is affected, but also as a consequence the cellular composition.
-Overall, when focusing on growth, growth rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
+When cultivated in minimal medium with acetate or succinate, \textit{Escherichia coli} uptake rates are correctly predicted by assuming that the metabolic network operates in a mode that maximizes growth rate~\cite{edwards_silico_2001}.
+This result is not general though, and depends on the carbon source and strain used~\cite{ibarra_escherichia_2002}.
+But even in those cases, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
+The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not only the efficiency of the enzymes is affected, but also their abundance and thus the cellular composition.
+Overall, with the currently available knowledge and its limitations (discussed in section~\ref{chap:discussion}), growth-rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
 
@@
 
-One of the conclusions of the preceding section is that growth rate maximization is a rather general principle of microbial life.
-Would microbial physiology be determined by this criteria, we could expect to find universal regularities in the way physiology adapts to the environment.
-Are there fundamental laws that control the physiological parameters of microorganisms, or self-replication as a whole?
-
-The most emblematic effort to formalize microorganism growth by coarse-grained modelling is the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
+One of the first attempt to formalize microorganism physiology into elegant, fundamental laws was the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
 This empirical relationship states that the growth rate $\mu$ [\texttt{div.h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol.L\textsuperscript{-1}}], such that
@@
 What is even more remarkable is that this relation is rather universal.
-Despite some adjustments (Droop model being a well-known example~\cite{droop_thoughts_1973}), the relation holds for most microorganisms: the growth rate increases with the abundance of the limiting nutrient, up to a point beyond which growth cannot be any faster~\cite{koch_why_1988}.
+Despite some adjustments (the Droop model being a well-known example~\cite{droop_thoughts_1973}), the relation holds for most microorganisms: the growth rate increases with the abundance of the limiting nutrient, up to a point beyond which they cannot grow any faster~\cite{koch_why_1988}.
 
@@
 \textbf{Monod law, reproduced from~\cite{monod_growth_1949}.}
-Monod has been a pioneer in the formalization of microbial growth into fundamental coarse-grained relationships.
-This figure shows the growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
-As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following a hyperbolic relationship described in Eq~\ref{eq:monod_law}.
+Monod has been a pioneer in the formalization of microbial growth into fundamental, coarse-grained relationships.
+This figure displays the growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
+As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following the hyperbolic relationship described in Eq~\ref{eq:monod_law}.
 Solid line is drawn with $\mu_K = 1.35$ div.h\textsuperscript{-1}, and $K_C = 0.22 \cdot 10^{-4}$ mol.L\textsuperscript{-1}.
@@
 
-The production of every component of the cell is under complex regulations.
-But they are all linked to the growth rate, since the production has to compensate for the dilution caused by growth~\cite{monod_growth_1949}.
-Many physiological parameters (like mass of DNA, RNA and proteins) are then functions of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
-These functions were carefully measured~\cite{bremer_modulation_1996} and are still used today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
-Recently, the topic was brought to the fore by the work of Scott \textit{et al}, whom redefined and revived interest in the bacterial growth laws~\cite{scott_bacterial_2011}.
-They focused on the RNA/protein ratio, a physiological parameter that was long known to vary linearly with the growth rate in microorganisms (Fig~\ref{fig:scott_rnaprot}).
-By analyzing this ratio under different environmental perturbations of the protein synthesis machinery, they built a coarse-grained model describing proteome resource allocation~\cite{scott_emergence_2014}.
-It has allowed to show that, whatever the details of the molecular implementation (differing between organisms), universal growth laws emerges from the underlying principles of robustness and optimization imposed by natural selection~\cite{scott_emergence_2014}.
-% detail this
+The production of every component of the cell is controlled by a complex network of regulatory interactions.
+But an obvious constraint is that at steady state, the synthesis rate of all individual components must be proportional to the growth rate in order to compensate for growth dilution~\cite{monod_growth_1949}.
+Many physiological parameters (like mass of DNA, RNA and proteins) are thus functions of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
+These so-called growth laws were carefully measured~\cite{bremer_modulation_1996} and are still used today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
+Recently, the topic was revitalized through the work of Scott \textit{et al}~\cite{scott_bacterial_2011}.
+They focused on the ribosome concentration, a physiological parameter that was long known to vary linearly with the growth rate in microorganisms (Fig~\ref{fig:scott_rnaprot}).
+By measuring it under different environmental perturbations of the protein synthesis machinery, they built a coarse-grained model describing proteome resource allocation~\cite{scott_emergence_2014}.
+It has allowed to show that, whatever the details of the molecular implementation (differing between organisms), this growth law emerges from the underlying principles of robustness and optimization imposed by natural selection, especially growth-rate maximization in every conditions~\cite{scott_emergence_2014}.
 
@@
 \textbf{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).}
-For different carbon sources more or less rich, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
+For a variety of carbon nutrients and their corresponding steady-state growth rate, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
 This ratio is correlated with the fraction of ribosome-affiliated proteins, and therefore with the relative abundance of ribosomes in the cell.
@@
 
-Coarse-grained models have also been helpful to understand the process by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability.
-Since nutrients are not used at their full potential given the environmental conditions, this widespread growth strategy apparently leads to energy waste, ~\cite{molenaar_shifts_2009} (Fig~\ref{fig:molenaar_overflow}).
-Originally called the Crabtree effect when first observed in \textit{Saccharomyces cerevisiae}~\cite{dijken_kinetics_1993} and overflow metabolism when later found in \textit{E.
-coli}~\cite{vemuri_overflow_2006}, it has also been called the Warburg effect in many types of tumor cells~\cite{mckeehan_glycolysis_1982,hsu_cancer_2008}.
-Molenaar \textit{et al} showed that this phenomenon simply results from the underlying principle of growth rate maximization, in particular the trade-off between the cost of enzyme synthesis and the pathway efficiency~\cite{molenaar_shifts_2009}.
-Like traders, bacteria distribute their resources between different beneficiaries, and tend to do so in a way that maximizes returns on investments.
-
-\begin{figure}[tb]
-\centering
-\includegraphics[width=\textwidth]{./Fig/Chapter1/molenaar_overflow.eps}
-\caption{
-\textbf{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.}
-When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
-This paradoxical behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that pathways for inefficient catabolism are shorter and therefore cheaper to produce.
-(A) Simple view of microbial metabolism, with the two competitive efficient and inefficient pathways.
-(B) The switching from efficient to inefficient catabolism when the concentration of nutrients, and hence the growth rate, increases.
-}
-\label{fig:molenaar_overflow}
-\end{figure}
-
 \subsection{Static versus dynamical perspective on growth}
 
-Growth rate maximization does a pretty good job at explaining the universal growth strategies described above.
-By maximizing offspring, it seems to comply perfectly with the principles of natural selection cited earlier.
-However, all the studies cited above apply this growth rate maximization in a very specific growth scenario, one that in fact does not correspond to the environmental conditions in which microorganisms have been selected, as we will argue below.
-
-Most microbial studies have been done at steady state, where all intensive properties of the cell are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
+The growth laws cited above apply at steady state, where all intensive properties of the cell are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
 It means that the properties independent of the cell volume or mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
 It requires that the components of the cell "increases by the same factor over a time interval", which have motivated the creation of the term balanced growth~\cite{campbell_synchronization_1957}.
-Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that the time-point of sampling becomes irrelevant~\cite{schaechter_microbe_2006}.
-It can be easily achieved laboratory conditions either in continuous culture, where the substrate is continually supplied, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
-This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells [citation FBA].
+Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that samplings are no longer time-sensitive~\cite{schaechter_microbe_2006}.
+It can be easily achieved in the laboratory either in continuous culture, where the substrate is continually supplied, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
+This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells~\cite{orth_what_2010}.
 
-\begin{figure}[tb]
+\begin{figure}[p]
 \centering
-\includegraphics[height=5cm]{./Fig/Chapter1/growth_curve.eps}
+\includegraphics[height=6cm]{./Fig/Chapter1/growth_curve.eps}
 \caption{
-\textbf{The different phases of a typical growth curve are visualized by the accumulation of biomass (black thick line) and the nutrient abundance in the medium (green dashed line).}
-(A) The lag phase, a variable period of time during which the organism adapt to the new medium.
-It is hardly reproducible and strongly depends on the recent history of the culture.
-(B) The steady-state (or balanced-growth) phase which is characterized by an exponential growth of biomass.
+\textbf{The different phases of a typical growth curve}.
+During a typical batch growth scenario, biomass accumulates (black thick line) whereas nutrients are consumed until depletion (green dashed line)~\cite{schaechter_microbe_2006}.
+(A) The lag phase, a variable period of time during which the organism adapt to the new medium~\cite{swinnen_predictive_2004}.
+It is hardly reproducible and can depends on the pre-culturing history of the strain~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, the magnitude and the rate of the change between the past and present environments~\cite{mcmeekin_predictive_2002}, and other hard-to-control environmental conditions~\cite{cheroutre-vialette_application_2002}.
+(B) The steady-state (or balanced-growth) phase which is characterized by an exponential production of biomass.
 It presents constant and reproducible characteristics which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
-This phase can be extended for thousands of generations by the constant addition of fresh medium (continuous culture).
+This phase can be extended for hundreds of generations by the constant addition of fresh medium (continuous culture)~\cite{wang_robust_2010}.
 As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
-(C) The stationary phase, after the depletion of the limited nutrient in the medium.
-In environmental conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase.
-Some species which have evolved long-term resistance mechanisms like sporulation can survive harsh conditions when in stationary phase.
+(C) The stationary phase, after the depletion of the limited nutrient in the medium~\cite{chubukov_environmental_2014,schaechter_microbe_2006}.
+In environmental conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+Some species which have evolved long-term resistance mechanisms, like sporulation~\cite{stragier_molecular_1996} or cannibalism~\cite{gonzalez-pastor_cannibalism:_2011}, can survive particularly long stationary phases.
 }
@@
 
-Although balanced growth is convenient from an experimental and theoretical point of view, it is widely admitted that microorganisms rarely encounter this state in nature since it requires a stable environment over a long period of time~\cite{schaechter_microbe_2006}.
-Microorganisms live in strongly dynamic communities, in medium where the limited nutrient is quickly depleted by the competitors when available\footnote{an exception being bacteria that live a parasitic or symbiotic life in bigger cells [citation Micoplasma genitallium, Giant tube worm]}.
-With that in mind, a microorganism aptness to optimize its steady-state growth rate should not play any role for its persistence.
-Why would in that case microorganisms be optimal for steady-state growth?
-Are we missing something by studying systems that were selected to cope with environmental variations (regulations of cell composition) in stable, unchanging conditions that probably confuse them?
+Although balanced growth is convenient from an experimental and theoretical point of view, it is widely admitted that microorganisms rarely encounter this state in nature~\cite{schaechter_microbe_2006}.
+Steady-state growth requires stable conditions over a long period of time, but microorganisms live in
+environments where the key elements, such as carbon, nitrogen or phosphorus, are quickly depleted by the competitors as soon as they become available~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+Why would microorganisms be optimal for a state they have barely encountered during their evolution?
+Are we missing something by studying in stable, unchanging conditions, systems that were in fact selected to cope with environmental variations?
 
@@
 
+Coarse-grained models have also been helpful to understand the process by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability.
+Since nutrients are not used at their full potential given the environmental conditions, this widespread growth strategy apparently leads to energy waste, ~\cite{molenaar_shifts_2009} (Fig~\ref{fig:molenaar_overflow}).
+Originally called the Crabtree effect when first observed in \textit{Saccharomyces cerevisiae}~\cite{dijken_kinetics_1993} and overflow metabolism when later found in \textit{E.
+coli}~\cite{vemuri_overflow_2006}, it has also been called the Warburg effect in many types of tumor cells~\cite{mckeehan_glycolysis_1982,hsu_cancer_2008}.
+Molenaar \textit{et al} showed that this phenomenon simply results from the underlying principle of growth rate maximization, in particular the trade-off between the cost of enzyme synthesis and the pathway efficiency~\cite{molenaar_shifts_2009}.
+Like traders, bacteria distribute their resources between different beneficiaries, and tend to do so in a way that maximizes returns on investments.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=\textwidth]{./Fig/Chapter1/molenaar_overflow.eps}
+\caption{
+\textbf{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.}
+When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
+This paradoxical behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that pathways for inefficient catabolism are shorter and therefore cheaper to produce.
+(A) Simple view of microbial metabolism, with the two competing efficient and inefficient pathways.
+(B) The switching from efficient to inefficient catabolism when the concentration of nutrients, and hence the growth rate, increases.
+}
+\label{fig:molenaar_overflow}
+\end{figure}
+
 \subsection{Measuring growth of microorganisms}

### 2016-12-22 Chapters/introduction.tex
Introduction 1.2 and first part of 1.3

@@
 During a typical batch growth scenario, biomass accumulates (black thick line) whereas nutrients are consumed until depletion (green dashed line)~\cite{schaechter_microbe_2006}.
-(A) The lag phase, a variable period of time during which the organism adapt to the new medium~\cite{swinnen_predictive_2004}.
-It is hardly reproducible and can depends on the pre-culturing history of the strain~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, the magnitude and the rate of the change between the past and present environments~\cite{mcmeekin_predictive_2002}, and other hard-to-control environmental conditions~\cite{cheroutre-vialette_application_2002}.
-(B) The steady-state (or balanced-growth) phase which is characterized by an exponential production of biomass.
-It presents constant and reproducible characteristics which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
-This phase can be extended for hundreds of generations by the constant addition of fresh medium (continuous culture)~\cite{wang_robust_2010}.
+(A) The lag phase is a variable period of time during which the organism adapt to the new medium~\cite{swinnen_predictive_2004}.
+It is hard to study experimentaly and is known to be affected by the pre-culturing history of the strain~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, the magnitude and the rate of the change between the past and present environments~\cite{mcmeekin_predictive_2002}, and other hard-to-control environmental conditions~\cite{cheroutre-vialette_application_2002}.
+(B) The steady-state (or balanced-growth) phase is characterized by an exponential production of biomass.
+Its characteristics are time-invariant and reproducible, which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
+This phase can be extended for hundreds of generations in continuous cultures, by the constant addition of fresh medium~\cite{wang_robust_2010}.
 As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
-(C) The stationary phase, after the depletion of the limited nutrient in the medium~\cite{chubukov_environmental_2014,schaechter_microbe_2006}.
-In environmental conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+(C) The stationary phase occurs after the depletion of the limiting nutrients in the medium~\cite{chubukov_environmental_2014,schaechter_microbe_2006}.
+In natural conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
 Some species which have evolved long-term resistance mechanisms, like sporulation~\cite{stragier_molecular_1996} or cannibalism~\cite{gonzalez-pastor_cannibalism:_2011}, can survive particularly long stationary phases.
@@
 Steady-state growth requires stable conditions over a long period of time, but microorganisms live in
-environments where the key elements, such as carbon, nitrogen or phosphorus, are quickly depleted by the competitors as soon as they become available~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+environments where the key elements, such as carbon, nitrogen or phosphorus, are quickly depleted by the competitors as soon as they become available~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, a consideration that also holds for the lab strains on which growth laws are studied~\cite{savageau_escherichia_1983,savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
 Why would microorganisms be optimal for a state they have barely encountered during their evolution?
@@
 
-With their beautiful simplicity, growth laws represent an exceptional point of view on microbial life strategies.
-But being established at balanced growth might have drawbacks since it is a growth state totally irrelevant to the life-trait history of microorganisms.
-How would growth laws be affected in a dynamical context?
-Can we still find universal regularities during growth states more relevant to the natural habitat of microorganisms?
+Microbial growth is essentially a resource allocation problem that can be summarized in simple, fundamental growth laws.
+Though established empirically, those laws are elegantly explained if we consider that microorganisms behave in way that optimizes their biomass production.
+However, these laws describe growth in stable environments, but it is hard to imagine how natural selection could have made microorganisms optimal for conditions they rarely encounter.
+This motivates the study of growth laws through a dynamical perspective, that is, in a changing environment, and leads us to the following problem statement: \textbf{Which strategies microorganisms follow to dynamically reallocate their resources after a change in the environment?}
+The development of this question prompts the following investigations.
 
-Since we lose the inherent advantages of working at steady state, we must confine our problem to keep it tractable.
-We then focus on the fundamentals behind resource allocation (the proteome composition growth law), especially the regulation of the abundance of gene expression machinery during the transition between two different growth media.
+What are the best strategies of resource allocation if, like for the steady-state growth laws, the biomass production is kept as the only optimization criteria?
+Through answering this question, we could assess to which extend the features of dynamical optimization differ from the steady-state optimization.
+In other words, it would test our verbal hypothesis that motivated a dynamical perspective on growth: the fact that we expect evolutionary pressure to apply differently between dynamic and stable conditions.
+This requires to translate the problem in the mathematical world, through the development of a proof-of-concept model of resource allocation in bacterial cells, and the theoretical determination of the optimal dynamical allocation of resources following a change in the environment.
 
-On the theoretical side, we assess if there exist a universal optimal way to allocate cell resources during growth transitions.
-On the experimental side, we monitor the abundance of key macromolecular components in order to get an insight that could be compared with the theoretical predictions.
+How could one measure empirically the resource allocation during a growth transition?
+The model will help us to establish actual resource allocation strategies that optimize biomass production during an environmental change.
+However, it will not be able to tell us if microorganism actually care to optimize such a criterion.
+We need though the possibility to measure biomass accumulation and changes in resource allocation when a shift is made in the environment.
+This would require a way to dynamically track the abundance of key macromolecules in living cells, while establishing an experimental set-up that temper the inherent variability of dynamical experimental studies of microbial growth.
 
@@
 
-My recent experience as a teacher taught me that, as of today, a lot of students still choose to study biology in the hope to find a math-free job in science.
-They fall from the clouds when you tell them modeling is an important part of a biologist's life.
-This misconception usually comes from a bad understanding of what is modeling.
-In its general sense, a model depicts and simplifies reality.
-Maps, sketches, or pictures satisfy this definition, as do graphs, sets of equations, or even the DNA sequences I store in a text-file on my computer.
-Overall, depicting and simplifying reality makes you a modeler, not so much drawing equations on a board.
-But this does not change the fact that mathematics are incredible tools for these purposes~\cite{servedio_not_2014,mcgill_calm_2013}.
+My recent experience as a teacher taught me that, as of today, a lot of students in biology do not like mathematics very much.
+Since mathematics is the main language of modeling, this has often made it difficult for me to convince them that modeling is of key importance for biologists.
+However, while mathematics is the language, it is not the essence of modeling.
+Generally speaking, a model depicts and simplifies reality.
+Maps, sketches, or pictures satisfy this definition, as do graphs, sets of equations, or even the DNA sequences stored in a text file on a computer.
+In other words, depicting and simplifying reality makes you a modeler, not so much drawing equations on a board.
+But this does not change the fact that mathematics are powerful tools for formulating and analyzing models~\cite{servedio_not_2014,mcgill_calm_2013}.
 
 What makes mathematical modeling so useful in microbial growth studies?
-Microbial physiology comes out from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation.
-They usually occur on a short time scale as compared with the organism life span.
-They nevertheless have been (and still are) selected by evolutionary processes that occur on long time scales.
-These physical limitations strongly impede a full empirical study of growth processes, making sometimes hard to test the most simple verbal hypothesis.
-Since they are designed to abstract away complexity, mathematical models are great tools to deal with such limitations.
+Evolutionary biology is a good instance of field that has been resorting for a long time on mathematical modeling~\cite{servedio_not_2014}.
+The large time and population scales at which evolutionary processes occur truly impede most of the experimental work, despite some attempts on organisms with a short generation time~\cite{elena_evolution_2003}.
+Besides, microbiology is a field particularly prone to experimental study.
+A growth curve as presented in Fig.~\ref{fig:growth_curve} can be made in a day, so the limitations of evolutionary biology should not apply to microbial growth studies.
+Why care to mimic reality on a computer when we can just directly bend it to our will~\cite{hillis_why_1993}?
+The last decades have strongly challenged this viewpoint: as knowledge and data on microbes accumulated, the need for a systemic understanding of biology was more and more pressing~\cite{alon_introduction_2006,kremling_systems_2013}.
+
+Microbial physiology results from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation~\cite{schaechter_microbe_2006}.
+These reactions occurs on short time scales, and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations.
+As a consequence, even the simplest verbal hypothesis can resist direct empirical testing.
+But abstracting away complexity is the purpose of mathematical models.
 The overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
-When working on a model, growth is abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014}.
+When working on a model, the microbe behaviors are abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014,mcgill_calm_2013}.
 Predictions can be made, "unpacked" into our real world, and confronted with experimental testing.
-In that sense, modeling is at the core of the scientific method~\cite{mcgill_calm_2013}.
-This could explain why mathematical modeling has been extensively applied to microbial growth studies, creating a diversity a modeling framework.
 
-For an increasing number of microorganisms, we are now able to draw quasi-exhaustive maps of their inner metabolic reactions.
-This has been made possible by the advances on genome reconstruction, which are easier and cheaper to accomplish every passing year.
-But making sense of these data proved to be even more challenging.
-Which reactions are important, and in which conditions?
-How will a metabolic network react to perturbations?
-Can we predict its evolution in a given environment?
-When dealing with genome-scale metabolic networks, one is challenged by the number of metabolites and possible reactions (e.g.
-the most recent genome-scale reconstruction of \textit{E.
-coli} metabolism contains "1366 genes, 2251 metabolic reactions, and 1136 unique metabolites"~\cite{orth_comprehensive_2011}).
-More importantly, while progress in genomic made possible such reconstructions, measuring precise \textit{in-vivo} kinetic parameters for all these reactions still represents a technological bottleneck~[citation?].
-Nevertheless, a modeling framework is specifically designed to deal with such limitations: constraint-based modeling.
-This framework abstracts away the unknown kinetics and represents the metabolic maps as a flux map on which physico-chemical constraints can be applied, e.g.
-steady-state, compartmentalization, mass conservation, molecular crowding, and thermodynamic directionality~\cite{ebrahim_cobrapy_2013}.
-Candidates for the flux distribution are obtained by optimizing the system towards an objective function, which usually represents growth rate for microorganisms.
-One can find a lot of derivatives of this framework, but its the Flux Balance Analysis that renewed interest into it [citation review ?].
-\textbf{[Need some key results that were obtained, or this paragraph makes no sense].}
-Note that such a level of abstraction necessarily has its drawbacks, e.g.
-the definition of the objective function, the difficulty to work in dynamic or implement information on the metabolic regulations [citation?].
+Modeling was particularly helpful to unveil how metabolic networks operate.
+For an increasing number of microorganisms, we are now able to draw quasi-exhaustive maps of their metabolic reactions.
+For instance, the most recent genome-scale reconstruction of \textit{E. coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
+Such a number of variables might seem overwhelming, and making sense of these knowledge is rarely straightforward.
+Which reactions are important, and in which environmental conditions?
+Can we predict how perturbations will affect a given metabolic network?
+Do fundamental regularities exist between different metabolic networks, from different species?
 
-The ever-growing available data on metabolic networks can also be used to make large kinetic models.
-For the reasons we stated above, we expect these models to be inherently limited in scope.
-It means that the modeler has to be extremely careful since he or she injects its knowledge about what might be important in the system, for the question he or she tries to solve.
-They are however more flexible, in the sense that one can easily make interplay different level of regulations.
-\textit{[Present the work of Klumpp, Kremling, Sauer, whole model of genitallium, etc]
-}\textit{Especially the work of Klumpp, Kremling, Sauer, etc, also the whole model of genitallium.
-Biggest problem is the lack of parameter values, and some huge trouble with identifiability.
-They have medium to large scope (increasing), highly detailed granularity and deal with the mathematical complexity via the computational power via the use of huge quantity of data.}
+This apparent complexity does not impede a mathematical studies of metabolic networks.
+Constraint-based modeling is a framework that abstracts away the unknown kinetics and represents the metabolic reactions by steady-state fluxes to which physico-chemical constraints can be applied, e.g.
+compartmentalization, mass conservation, molecular crowding, and thermodynamic directionality~\cite{ebrahim_cobrapy_2013}.
+In other words, through constraints imposed to the metabolic networks, it provides a space of admissible flux distributions, eliminating reactions that are unlikely to occur in the environmental condition of interest.
+This space is often further reduced by selecting the flux distributions optimizing a specified objective function (e.g., the growth rate of the cell).
+Such an approach, initially developed through flux balance analysis (FBA)~\cite{orth_comprehensive_2011,palsson_systems_2011}, has been shown to correctly predict many behaviors of the metabolic network of \textit{E.
+coli}~\cite{varma_stoichiometric_1994,edwards_silico_2001}.
+It also proved capable of predicting its long-term recovery through genetic evolution after a gene deletion~\cite{fong_metabolic_2004}.
 
-Another approach has shown successful in understanding some aspects of growth behaviors of microorganisms.
-By applying a high level of abstraction on how the cell functions, it becomes easy to compare data about different species and to uncover general evolutionary principles that are independent of the specific implementation in each organism.
-Best examples of this approach are the empirical growth laws of ribosomal abundance~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}, or the proof-of-concept explanation of overflow metabolism~\cite{molenaar_shifts_2009} that were described in section~\ref{sec:context} of this introduction.
-It is also the preferred approach when one wants to broaden its access to mathematical tools~\cite{vandenberg_optimal_1998}, most of them being unhelpful with the high dimensionality of biological models.
-As we will describe in section~\ref{sec:approach}, this represents the modeling strategy that was used in this thesis.
+More direct modeling approaches, in the form of large kinetic models, have also been helpful to understand growth-related processes.
+Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} showed how metabolic fluxes are sensed at different locations of the central carbon metabolism, and integrated in a global cell response by the coupling of enzymatic and transcriptional regulations~\cite{kotte_bacterial_2010}.
+Such models are often sufficiently detailed to be used as \textit{in silico} testbed to investigate and uncover specific molecular mechanisms~\cite{peskov_kinetic_2012}.
+The most emblematic instance of that kind of approach is probably the recent whole-cell model of \textit{Mycoplasma genitalium}~\cite{karr_whole-cell_2012}.
+By aggregating all the available knowledge from more than 900 publications, Karr \textit{et al} constructed a dynamical model accounting for all the annotated gene functions of this pathogenic organism, and successfully used it to investigate unobserved molecular mechanisms and guide novel experimental analysis.
+While such approach is not yet applicable to all kind of organisms, it genuinely demonstrates how the increase in computational power could one day make at hand \textit{in silico} experimentation.
 
-Coarse-grained models have also been helpful to understand the process by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability.
-Since nutrients are not used at their full potential given the environmental conditions, this widespread growth strategy apparently leads to energy waste, ~\cite{molenaar_shifts_2009} (Fig~\ref{fig:molenaar_overflow}).
-Originally called the Crabtree effect when first observed in \textit{Saccharomyces cerevisiae}~\cite{dijken_kinetics_1993} and overflow metabolism when later found in \textit{E.
-coli}~\cite{vemuri_overflow_2006}, it has also been called the Warburg effect in many types of tumor cells~\cite{mckeehan_glycolysis_1982,hsu_cancer_2008}.
-Molenaar \textit{et al} showed that this phenomenon simply results from the underlying principle of growth rate maximization, in particular the trade-off between the cost of enzyme synthesis and the pathway efficiency~\cite{molenaar_shifts_2009}.
-Like traders, bacteria distribute their resources between different beneficiaries, and tend to do so in a way that maximizes returns on investments.
+But while it represents their strength, the granularity of the models cited above is also their main weakness.
+Despite this apparently rich available knowledge, precise measurements of \textit{in-vivo} kinetic parameters still represents a technological bottleneck~\cite{park_metabolite_2016,bennett_absolute_2009,buscher_cross-platform_2009}.
+Constraint-based modeling overcomes this limitation by using a mathematical approach that does not require any knowledge about the reaction kinetics.
+The drawback is that they highly depend on the constraints used to reduce the space of admissible solutions, e.g.
+the choice of the objective function that could sometimes be tricky~\cite{kauffman_advances_2003}.
+Kinetic models however are particularly sensible to missing parameters.
+While the so-called free parameters are often collectively fitted to available data~\cite{jaqaman_linking_2006,mendes_non-linear_1998}, this is known to produce large parameter uncertainty~\cite{cho_experimental_2003,brodersen_characterization_1987,rodriguez-fernandez_hybrid_2006}, which can significantly affect the predictions of the model~\cite{gutenkunst_universally_2007,ingram_network_2006,mayo_plasticity_2006}.
 
-\begin{figure}[tb]
-\centering
-\includegraphics[width=\textwidth]{./Fig/Chapter1/molenaar_overflow.eps}
-\caption{
-\textbf{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.}
-When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
-This paradoxical behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that pathways for inefficient catabolism are shorter and therefore cheaper to produce.
-(A) Simple view of microbial metabolism, with the two competing efficient and inefficient pathways.
-(B) The switching from efficient to inefficient catabolism when the concentration of nutrients, and hence the growth rate, increases.
-}
-\label{fig:molenaar_overflow}
-\end{figure}
+Such limitations often motivate the construction of simpler, coarse-grained models.
+The approach is dramatically different: instead of aggregating the available knowledge into a highly detailed model, the modeler is concerned about carefully filtering this information to keep the model as simple as possible.
+They generally consider a high level of abstraction of the cell functions, and are particularly valuable when looking for universal laws or principles~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}.
+They can take the form of minimal core models that focus on a given aspect of growth, while still abstracting away the molecular details (see \textit{e.g.}, ~\cite{spiesser_size_2012}).
+Such models can also be used as proof-of-concept to submit verbal hypothesis to the logic and rigor of mathematical reasoning~\cite{servedio_not_2014}.
+For instance, a simple proof-of-concept model was used to uncover principles that lead to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
+This paradoxal and widely spread mechanism~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the trade-off between the efficiency of a pathway and the cost of synthesis of its enzymes.
+Overall, these models have several advantages for the purpose of our study: they clearly state and do not lose track of their assumptions, and are sufficiently tractable to allow access to complex mathematical tools (as will be required for dynamical optimization).
+
+%It is also the preferred approach when one wants to broaden its access to mathematical tools~\cite{vandenberg_optimal_1998}, most of them being unhelpful with the high dimensionality of biological models.
+%As we will describe in section~\ref{sec:approach}, this represents the modeling strategy that was used in this thesis.
+
+%\begin{figure}[tb]
+%\centering
+%\includegraphics[width=\textwidth]{./Fig/Chapter1/molenaar_overflow.eps}
+%\caption{
+%\textbf{The strategy of overflow metabolism, reproduced from~\cite{molenaar_shifts_2009}.}
+%When increasing the concentration of substrate, the metabolism of microorganisms switches toward the use of inefficient pathways.
+%This paradoxical behavior does in fact maximize the growth rate if we take into account the costs and benefits of enzyme synthesis, in particular the fact that pathways for inefficient catabolism are shorter and therefore cheaper to produce.
+%(A) Simple view of microbial metabolism, with the two competing efficient and inefficient pathways.
+%(B) The switching from efficient to inefficient catabolism when the concentration of nutrients, and hence the growth rate, increases.
+%}
+%\label{fig:molenaar_overflow}
+%\end{figure}
 

### 2016-12-23 Chapters/introduction.tex
Finished introduction

@@
 They sustain themselves by collecting energy from their environment, in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
-While most of our cells are kept in a stable and friendly internal environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to strongly variable environments.
-Not without a certain success, because microbes have been found to survive almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life~\cite{rothschild_life_2001,nicholson_transcriptomic_2012,madigan_biology_2006,schaechter_microbe_2006}.
+While most of our cells are kept in a stable and friendly internal environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to strongly variable conditions~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013,
+savageau_escherichia_1983,savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
+Not without success, because microbes have been found to survive almost everywhere, sometimes in the strangest places that were initially thought unsuitable for life~\cite{rothschild_life_2001,nicholson_transcriptomic_2012,madigan_biology_2006,schaechter_microbe_2006}.
 Such ubiquity translates a strong potential for adaptation.
-Even the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes\cite{zimmer_microcosm:_2009}.
+Even the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes~\cite{zimmer_microcosm:_2009}.
 
@@
 Furthermore, some of these products specifically bind some promoters and stimulates or represses the promoter activity, hence, the gene expression.
-They are called transcription factors and allow, among other mechanisms, the regulation of cell composition.
+They are called transcription factors and allow, in addition to other mechanisms, the regulation of cell composition.
 
@@
 Which resource allocations are optimal for a living cell?
-Systems capable of Darwinian evolution sustain themselves on the long term by \textit{fitting} to their environment.
+Systems capable of Darwinian evolution sustain themselves on the long term by \textit{fitting} to their environment~\cite{dawkins_selfish_1976}.
 Living cells with the best \textit{fitness} are more likely to survive, replicate, and through long-term competition, eliminate their siblings.
@@
 
-For this reason, there is no consensus about what constitute the main fitness factors of microorganisms.
+For this reason, there is no consensus about what constitutes the main fitness factors of microorganisms.
 Competition for common resources seems to favor the maximization of offspring, in the sense that a replicator with more descendants will outnumber its competitors in the long term~\cite{dawkins_selfish_1976}.
 It is unclear, however, how this general principle translates into their physiological behaviors.
-For instance, the use of genome-scale metabolic network models have shown that the metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental condition~\cite{schuetz_systematic_2007}.
+For instance, the use of genome-scale metabolic network models have shown that the metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental conditions~\cite{schuetz_systematic_2007}.
 But we can also found different strategies when the biomass production is at play.
-Bacteria can either optimize their production yield (taking its time to make the most of the nutrient) or their production rate (growing quickly while wasting part of the nutrient potential), the later being favored in fluctuating or poor environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008}, a result that has also been confirmed in yeast~\cite{schuster_maximization_2008}.
+Microbes can either optimize their production yield (taking time to make the most of the nutrient) or their production rate (growing quickly while wasting part of the nutrient potential), the later being favored in fluctuating or poor environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuster_maximization_2008}.
 Overall, it seems that cell optimality drifts in a multidimensional space and involves making several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
@@
 Monod has been a pioneer in the formalization of microbial growth into fundamental, coarse-grained relationships.
-This figure displays the growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
+This figure displays the steady-state growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
 As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following the hyperbolic relationship described in Eq~\ref{eq:monod_law}.
@@
 The growth laws cited above apply at steady state, where all intensive properties of the cell are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
-It means that the properties independent of the cell volume or mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
+It means that the properties that are independent of the cell volume or the cell mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
 It requires that the components of the cell "increases by the same factor over a time interval", which have motivated the creation of the term balanced growth~\cite{campbell_synchronization_1957}.
 Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that samplings are no longer time-sensitive~\cite{schaechter_microbe_2006}.
-It can be easily achieved in the laboratory either in continuous culture, where the substrate is continually supplied, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
+It can be easily achieved in the laboratory either in continuous culture, where the substrate is continually supplied~\cite{borirak_molecular_2014}, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
 This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells~\cite{orth_what_2010}.
@@
 (B) The steady-state (or balanced-growth) phase is characterized by an exponential production of biomass.
-Its characteristics are time-invariant and reproducible, which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
-This phase can be extended for hundreds of generations in continuous cultures, by the constant addition of fresh medium~\cite{wang_robust_2010}.
+Its characteristics are time-invariant and quite robust accross conditions, which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
+This phase can be extended for hundreds of generations in continuous cultures, by the constant renewable of the medium~\cite{borirak_molecular_2014,herbert_continuous_1956,wang_robust_2010}.
 As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
@@
 We need though the possibility to measure biomass accumulation and changes in resource allocation when a shift is made in the environment.
-This would require a way to dynamically track the abundance of key macromolecules in living cells, while establishing an experimental set-up that temper the inherent variability of dynamical experimental studies of microbial growth.
+This would require to dynamically track the abundance of key macromolecules in living cells, while establishing an experimental set-up that temper the inherent variability of dynamical experimental studies of microbial growth.
 
@@
 For instance, the most recent genome-scale reconstruction of \textit{E. coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
-Such a number of variables might seem overwhelming, and making sense of these knowledge is rarely straightforward.
+Such a number of variables might seem overwhelming, and making sense of such knowledge is rarely straightforward.
 Which reactions are important, and in which environmental conditions?
@@
 
-This apparent complexity does not impede a mathematical studies of metabolic networks.
+This apparent complexity does not impede a mathematical study of metabolic networks.
 Constraint-based modeling is a framework that abstracts away the unknown kinetics and represents the metabolic reactions by steady-state fluxes to which physico-chemical constraints can be applied, e.g.
@@
 More direct modeling approaches, in the form of large kinetic models, have also been helpful to understand growth-related processes.
-Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} showed how metabolic fluxes are sensed at different locations of the central carbon metabolism, and integrated in a global cell response by the coupling of enzymatic and transcriptional regulations~\cite{kotte_bacterial_2010}.
+Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} showed how metabolic fluxes are sensed at different locations of the central carbon metabolism, and are integrated in a global cell response by the coupling of enzymatic and transcriptional regulations~\cite{kotte_bacterial_2010}.
 Such models are often sufficiently detailed to be used as \textit{in silico} testbed to investigate and uncover specific molecular mechanisms~\cite{peskov_kinetic_2012}.
@@
 Such models can also be used as proof-of-concept to submit verbal hypothesis to the logic and rigor of mathematical reasoning~\cite{servedio_not_2014}.
-For instance, a simple proof-of-concept model was used to uncover principles that lead to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
+For instance, a simple proof-of-concept model was used to uncover the principles leading to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
 This paradoxal and widely spread mechanism~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the trade-off between the efficiency of a pathway and the cost of synthesis of its enzymes.
@@
 
-\textbf{Either to build models or test their predictions, data on microbial growth have to be acquired.}
-Measuring growth = measuring biomass accumulation, over time, and get relevant information on the cell inner state.
-We have a lot of data on this.
+Either to formulate hypothesis or test predictions, data on microbial growth need to be acquired.
+The most straightforward method is to work at the population level.
+When working with microorganisms, a clonal population of genetically identical cells can easily be obtained by inseminating a single colony into a growth medium~\cite{schaechter_microbe_2006}.
+Over time, this culture can be subject to many kind of measurements: for instance, biomass can be estimated  directly through measurements of the dry weight of samples~\cite{monod_growth_1949}, or indirectly through measurements of transmitted or scattered light across the culture~\cite{volkmer_condition-dependent_2011}, allowing to construct growth curves as presented in Fig.~\ref{fig:growth_curve}.
+Other population-wide parameters like the macromolecular composition~\cite{scott_interdependence_2010,scott_bacterial_2011} or the concentration of metabolites~\cite{bennett_absolute_2009} can also be evaluated.
+They represent averaged estimates from billions of genetically identical cells, and are thus usually robust and reliable measurements.
+In fact, most of \textit{E.~coli} parameters that are currently used in today's growth models have been measured at the population level, notably through the emblematic work of Bremer and Dennis~\cite{churchward_macromolecular_1982,bremer_modulation_1996,bremer_free_2003,bremer_feedback_2008}.
 
-\textbf{Measuring population-wide parameters is convenient and provides robust data.}
-Population measurement is easy to implement, and some things can only be measured at the population level (so far).
-But small unsynchronized effects can be hidden since we mostly work with the mean.
-Despite most of the knowledge we have on microorganism growth are from population measurements, like the RNA/DNA/etc ratios, the growth laws, etc.
-Extensive work of Bremmer and Dennis.
-Big discoveries of the last decades that rely on population measurements ? (cite 2 or 3)
+Nevertheless, some questions are not solvable at the population level~\cite{davey_flow_1996,bakshi_superresolution_2012,balaban_bacterial_2004}.
+This can be the case of the structure of the inner cell~\cite{bakshi_superresolution_2012}.
+Bakshi \textit{et al} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerazes and DNA in living \textit{E.
+coli} cells~\cite{bakshi_superresolution_2012}.
+They showed that, despite what was previously thought, most of the translation occur far from the DNA, on free mRNA that migrate to ribosome-rich regions~\cite{bakshi_superresolution_2012}.
+But single-cell variability has also been shown to be exploited by the microbial cells themselves~\cite{balaban_bacterial_2004,booth_stress_2002,sumner_phenotypic_2002}.
+Through growth and imaging of single cells in a microfluidic device, Balaban \textit{et al} showed that a clonal bacterial population can profit from a preexisting heterogeneity to persist when challenged by the presence of an antibiotic, a short-term mechanism that does not involve any genetic mutation~\cite{balaban_bacterial_2004}.
+On the long-term, heterogeneity has also been shown to help subpopulations to resist letal stresses in nature, letting them enough time to adapt and conquer new ecological niches~\cite{booth_stress_2002,sumner_phenotypic_2002}.
+In fact, cell heterogeneity seems so crucial for the fitness of microorganisms that diversity-generating mechanisms have been identified~\cite{true_yeast_2000,fraser_noise_2004,raser_control_2004}.
 
-\textbf{Single-cell measurements give access to data and behaviors that are hidden at the population level.}
-Single-cell measurements lay on technologies like microscopy, fluorescence, FACS...
-More interesting since we can use fluorescence to not just look at the cell but also measure interesting parameters (like gene expression) at the single-cell level.
-Information on the localisation of DNA, ribosomes [Bakshi].
-Variability in physiology, persistors, antibiotics resistance, etc [citations]
+In addition to the level of measurement, the cultivation method plays a significant role in the kind of question that an experiment can answer.
+For a long time, batch growing condition has been the preferred choice of most microbiologists~[citation?].
+As illustrated in Fig.~\ref{fig:growth_curve}, the organism is inseminated in a closed-system and grows until the nutrient is depleted.
+In a sense, such a condition is close to what microbes encounter in nature, where key nutrients are only available for a short time and quickly depleted~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+The golden age of batch culturing was attained in the last decade with the democratization of culture in microplates.
+In parallel, dozens to hundreds of growth experiments can be performed each in less than one milliliter, while the optical density or the fluorescence of each culture is easily automatically monitored.
+This has proven extremely helpful for screening purposes and the establishment of standard libraries of gene labeling and modification~\cite{baba_construction_2006,zaslaver_comprehensive_2006}.
+Parallel growing has notably been exploited to construct roughly 4,000 single-cell knock-out mutants of \textit{Escherichia coli}, forming the Keio collection~\cite{baba_construction_2006}.
+It was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples had been distributed worldwide~\cite{yamamoto_update_2009}).
+Another approach was to build a library that allow for the quantitative estimation of gene expression.
+Here again, microplates helped building a massive library of transcription fusion of the green fluorescent protein to each of about 2,000 different promoters in E.~coli~\cite{zaslaver_comprehensive_2006}.
+Interestingly, microplate batch growing conditions are also used when applying the say constructions in the inference and analysis of gene regulatory networks~\cite{gerosa_dissecting_2013,berthoumieux_shared_2013,keren_promoters_2013,
+ronen_assigning_2002,stefan_inference_2015}.
+But the later use is hazardous, because the main drawback of the batch condition is that the culturing system is poorly controlled.
+Changes of key chemical parameters (\textit{e.g.} pH, pO\textsubscript{2}) have been shown to occur during the whole growth curve~\cite{bekker_changes_2007}.
+At high density, metabolic by-products can accumulate in the medium and impede growth before nutrients are exhausted~\cite{ackerman_accumulation_1974,lenski_chemical_2002,luli_comparison_1990}.
+This could be dealt with by focusing on the beginning of the culture, but this part has been shown to strongly depend on the preculture history~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, and one is never sure when an internal steady state has been reached~\cite{myers_culture_1944}.
 
-\textbf{Batch growing conditions are convenient and close to the natural conditions of bacteria, but they raise several problems.}
-Both approaches can be performed either in batch or continuous conditions.
-Batch conditions is the most straight forward in lab: you put sugar in a flask, they grow until nutrient is depleted.
-Implemented in flash or petri-microscopy.
-Resist to contamination.
-It is close to natural conditions, and industrial conditions.
-Has even been used to study long-term evolution of E. coli.
-But the system is poorly control, mostly because the history of the cells play a huge role in the waking process.
+For this reason, we early observed development of continuous cultures for microbial studies~\cite{myers_culture_1944,novick_experiments_1950,herbert_continuous_1956}.
+Through permanent renewal of the growth medium, continuous cultures allow for maintenance of a chemically well controlled environment, allowing for the acquisition of reproducible and reliable data~\cite{borirak_molecular_2014,hoskisson_continuous_2005}.
+This showed particularly valuable for comparative omic analysis, for instance the analysis of changes in protein expressions~\cite{kolkman_comparative_2005} or genome-wide transcriptomic changes~\cite{boer_genome-wide_2003} that occur when \textit{Saccharomyces cerevisiae} is grown in different media.
+Continuous culture are however more difficult to set-up~\cite{novick_experiments_1950,borirak_molecular_2014}.
+They involve large quantity of medium and are, by design, expected to be more prone to contamination~\cite{novick_experiments_1950}.
+But the recent advances in microfluidic culturing has improved the former point~\cite{wang_robust_2010,balaban_bacterial_2004}.
+For instance, the mother machine~\cite{wang_robust_2010} was developed to perform long-term growth of microorganisms while allowing for single-cell tracking of a large population.
+It showed that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
+Overall, such a device is able to set up a robust continuous culture while consuming less than 100~mL of fresh medium per day.
 
-\textbf{Continous cultures allow to precisely control the growing conditions.}
-Continuous culture consists into constantly providing nutrients to a culture, and maintaining it in a steady-state for a long time.
-Implemented in bioreactor [citation] or microfluidic devices [mother machine citation].
-Sensible to contaminations so hard to maintained (that's why industry ignore it so far).
-Does not correspond to natural conditions but the system is fully controlled: perturbations can be applied and a lot of parameters can be monitored.
-Possibility to let the cell long-enough in steady state to "forget" strain history.
-Most suitable if one need to have full control on its system.
-
-\textbf{In all these practical conditions, one can choose to either measure cells at steady-state or in dynamic.}
-Finally, most of the work cited above are based on steady-state studies.
-But there are also work in dynamic.
-Levy showed that... in yeast.
-Alon in Coli.
-After decades working at steady-state, Bremmer and Dennis argued we should measure things during transitions.
-de Jong / Gerosa showed that the gene expression machinery has its own dynamic that should be taken into account when measuring gene expression.
-The evolutionary relevance of growth transitions has been emphasized in numerous others studies [citations lag phase, last generation before stationary, ...].
+Finally, all these conditions can be applied to study microbial growth through dynamical scenarii~\cite{levy_strategy_2007,levy_coordination_2009,madar_promoter_2013,ehrenberg_mediumdependent_2012}.
+For instance, Levy \textit{et al}~\cite{levy_strategy_2007,levy_coordination_2009} measured the transcriptomic reorganization occuring after an environmental perturbation in yeasts grown at steady state in a continuous culture.
+They observed that the transcription regulation is affected before the growth rate of the organism, suggesting a significant role of feed-forward sensing from the environment.
+Madar \textit{et al}~\cite{madar_promoter_2013} analyzed the transcriptional reorganization occuring in \textit{E.~coli} during the lag phase through the coupling of culture in microplates and single-cell measurements \textit{via} flow cytometry of time samples.
+They showed that \textit{bottleneck enzymes} were produced early in the lag phase, before the cells actually switch to the production of ribosomes and general metabolic enzymes.
+As in here, the studies cited above investigated questions that could only be solved in a dynamical context.
+There is still however no consensus about the growing conditions that are the more suitable for dynamical studies.
+By taking the best of the works presented above, a part of our study will focus on establishing such experimental conditions.
 
@@
 
-We first tackle the problem of creating a mathematical framework for studying environmental transitions.
-\textit{Building upon the work cited above, we create a simple self-replicator model that allocate resources between two sections.}
+Thorough this study, we focus on a specific dynamical growth scenario, in particular the case of a nutrient upshift~\cite{ehrenberg_mediumdependent_2012,kjeldgaard_kinetics_1961,schaechter_patterns_1961,johnsen_control_1977}.
+Contrary to steady state growth, nutrient upshifts are frequently encountered in the life cycle of a microorganism~\cite{schaechter_microbe_2006,mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+If combined with downshifts, they are also a good starting point if one wants to approximate more complex environments.
+However, note that in nature, nutrient upshifts generally start from stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, which involve complex mechanisms that impede most theoretical and experimental studies~\cite{stragier_molecular_1996,gonzalez-pastor_cannibalism:_2011,ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
+For this reason, we confined the problem to the study of resource allocation along a so called steady-state-to-steady-state transition, which could be seen as investigating what happens when the cells move along the growth law presented in Fig.~\ref{fig:scott_rnaprot}.
 
-The goal is to identify what would be an ideal transition.
-\textit{We all have an intuitive idea of what should be a good transition: it should be fast, robust, etc...
-But it is unclear how this relates to the fitness factor that shaped the regulatory mechanisms of microorganisms.}
+We start by developing in chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation which evaluates how biomass production can be maximized during such a growth transition.
+The model is an instance of self-replicator model~\cite{molenaar_shifts_2009}, but only focus on the allocation between two main sectors in the microbial cell: the metabolism that uptakes and converts nutrients in precursor metabolites, and the gene expression machinery that performs the macromolecular synthesis from these precursors.
+Though being kept as simple as possible, we show how this model is sufficient to account for the known growth laws of resource allocation at steady state~\cite{scott_interdependence_2010,scott_bacterial_2011}.
 
-We describe dynamical optimality by using optimal control theory.
-\textit{Brieve explanation about optimal control theory and how it is used in economic or engineering to optimize or understant dynamical processes.}
+Using this model, we pose the problem of dynamical optimization as an optimal control problem~\cite{stengel_optimal_1994}.
+Like for steady-state optimization, the cell objective is kept as being the biomass production, but we look for the dynamical resource allocation that maximizes it over an upshift scenario.
+Through the combination of Pontryagin Maximum Principle~\cite{carlson_infinite_1991} and numerical optimization~\cite{bonnans_bocop_2012}, we identify a solution involving bang-bang control~\cite{stengel_optimal_1994}.
+The mathematical gold standard for resources allocation is thus a scenario in which resources are always fully allocated to a single sector in a switchable manner.
 
-All of this is described in chapter 2.
+This theoretical solution is then used to explore possible regulatory strategies.
+We identify the system variables that need to be sensed in order to optimize resource allocation.
+We show that regulatory schemes that are equivalent at steady state significantly differ during an upshift scenario.
+In particular, feedback from the inner state of the cell is an information extremely more valuable than sensing directly the environmental variations (contrary to what is usually assumed, see e.g.
+\cite{levy_coordination_2009}).
+Furthermore, by just sensing at the same time the precursor and the gene expression machinery concentrations, the system is theoretically able to perform a near-perfect transition.
+We investigate if such a strategy could be implemented in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} to be a likely candidate.
+Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should allocate their resources towards the gene expression machinery in an on-off manner.
 
-Then, after we know what to look for, we try to actually measure how resource allocation occurs in a model organism: \textit{Escherichia coli}.
-\textit{We justify the use of coli and introduce the gene expression fluorescent reporter.
-We made a strain with fluorescent ribosomes, and studied our experimental framework in different transitions, at the population and the single cell level.
-We justify the use of both by the fact that population is easy for "screening", and single cell allow for a better resolution and take care of the "desynchronization" biais that could occur.}
+In chapter~\ref{chap:experiments}, we address the challenging problem of verifying this prediction in \textit{Escherichia coli}.
+It requires the development of a set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
+The time scales identified in chapter~\ref{chap:theory} motivates the use of \textit{in vivo}, high-frequency, single cell measurements of the gene expression machinery concentration.
+Inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we construct a strain that displays a gfp-tagged S2 ribosomal subunit.
+We place this strain in a microfluidic device called the mother machine~\cite{wang_robust_2010}.
+While it was originally developed for studies of long-term steady-state growth, we actually use it to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source).
 
-This, is described in chapter 3.
+The set-up is applied to the reconstruction of the resource allocation control during the upshift scenario.
+A preliminary analysis of the microscopy acquisition provided time series of the bacterial length and the single-cell concentration of fluorescence.
+We show that these measurements are sufficient for the intended signal reconstructions, and use Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007} to reconstruct the growth rate and the resource allocation over time.
+By exploiting the Bayesian logic behind this well-known technique (though rarely encountered in biology), we show it to be particularly suitable for microscopic data analysis and the reconstruction of complex gene expression signals.
+While the data acquired did not allow to conclude on the tested prediction, we present several realistic improvements that should allow to conclude in the near future.

### 2016-12-23 Chapters/discussion.tex
Finish Discussion

@@
 \chapter{Discussion}
+\label{chap:discussion}
 
-\textit{"Apples fall onto the Earth because natural selection eliminated apples falling towards the sky."} -- Source Unknown
+\textit{"Apples fall onto the Earth because natural selection eliminated apples falling towards the sky."} -- @tomroud~\cite{tomroud_tom_2016}, original source unknown.
 
@@
 Physics has yet to be reduced to a formula that will fit on a piece of clothing~\cite{falk_universe_2005}.
-Biology is even more behind, but the extensive work on the mechanisms of evolution are helping to close this gap, the Price equation being the closer we have of such an elegant formula~\cite{frank_natural_2012}.
+Biology is even more behind, but the extensive work on the mechanisms of evolution are helping to close this gap (the Price equation being the closer we have of such an elegant formula~\cite{frank_natural_2012}).
 But evolution by itself is not sufficient to define life.
 Even the less restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
-Self-sustainment appears to be as fundamental as evolution, and rely on the transformation of matter and energy from the environment, in other word, growth.
-But we are far from being as mathematically advance as in evolution on this question.
+Self-sustainment appears to be as fundamental as evolution, and rely on the transformation of matter and energy from the environment, in other words, on growth.
+But we are far from being as mathematically advanced on this question as we are on evolution.
 
-Fundamental growth laws have been established by focusing on the life of microorganisms.
+Nevertheless, fundamental growth laws have been established by focusing on microorganisms.
 They show that, regardless of the molecular implementations, microorganisms tend to follow the same empirical regularities when grown at steady state in different kind of environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
-In particular, they change their inner composition, and do so by adjusting the allocation of resources from the environment, in a way that appears to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
+In particular, they change their inner composition, and do so by adjusting how they allocate resources from their environment in a way that appears to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
 Growth laws are a strong support for the theory of a modular organization of microorganisms~\cite{scott_emergence_2014,hartwell_molecular_1999,arkin_fast_2006,guido_bottom-up_2006}.
 They represent a big step towards the general understanding of the process of growth.
-They are, however, established at steady state, a state in which most microorganisms spend very little~[citation].
+They are, however, established at steady state, a state in which most microorganisms spend very little time~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,
+hobbie_microbes_2013,savageau_escherichia_1983,
+savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
 Why would microorganisms be optimal for a state they rarely encounter?
@@
 It does not necessarily aim at predicting or controlling the behavior of microorganisms.
-By using a simple and abstract representation, it provides a convenient and tractable way to evaluate the implications of dynamical optimality, in particular their divergences with steady-state optimality.
+By using a simple and abstract representation, it provides a convenient and tractable way to evaluate the implications of dynamical optimality, in particular its divergence from steady-state optimality.
 Is a dynamical perspective on growth laws necessary?
@@
 One of the main drawbacks of such big models are that we can quickly lose track of their underlying hypothesis.
-When modeled in details, biological system also tend to become untractable, prohibiting the use of mathematical tools that are quickly overwhelmed by the size of a dynamical system (optimal control being one of them).
+When modeled in details, biological systems also tend to become untractable, prohibiting the use of mathematical tools that are quickly overwhelmed by the size of a dynamical system (optimal control being one of them).
 Besides, simple, abstract models make it easy for the modeler to state the critical, exploratory, and logistical assumptions that are made~\cite{servedio_not_2014}.
@@
 
-But the assumptions that need most to be discussed are the critical ones, i.e. the ones that would invalidate the conclusions drawn if falsified.
+But the assumptions that need most to be discussed are the critical ones, \textit{i.e.} the ones that would invalidate the conclusions drawn if falsified.
 In our case, it is the assumption that biomass production is the main fitness factor in stable or fluctuating environments.
 Through the manuscript, we did not really challenge this hypothesis, but only provided arguments about why it is so accepted.
-While it is clear that microorganisms can be cultivated, and even naturally found in conditions where maximizing growth rate ensure their persistence, we have no direct proof that biomass production is what allowed them to naturally persist on evolutionary timescales.
-In addition, as illustrated by the quote at the beginning of this chapter, one must be excessively careful when explaining the behavior of living system solely through natural selection considerations.
+These arguments however are mostly based on research made on lab strains in lab conditions.
+While it is clear that microorganisms can be cultivated, and even naturally found in conditions where maximizing growth rate ensure their persistence~\cite{edwards_silico_2001,ibarra_escherichia_2002,lewis_omic_2010,molenaar_shifts_2009}, we have no direct proof that biomass production is what allowed them to naturally persist on evolutionary time scales.
+In addition, as illustrated by the quote at the beginning of this chapter, one must be excessively careful when explaining the behavior of living systems solely through natural selection considerations.
 Such systems are embedded in the physical world, and evolutionary arguments are not always relevant to explain an empirically observed behavior.
@@
 It was applied to the question of dynamical resource allocation to emphasize the limitations that affect the study of transitions, and to propose possible workarounds.
-Fluorescent reporter proteins were used to cope with the need for \textit{in vivo} measurements with a high time resolution.
-Advances in microfluidics were exploited to realize steady-state to steady-state transitions from one medium to another, in order to temper the effect of strain history and enhance reproducibility.
-Finally, the signal of interest was reconstructed through the use of Kalman smoothing, a powerful signal processing algorithm that has many features that make it suitable for microscopy time series.
-Overall, the set-up still need many improvements, though most of them are experimental peculiarities that are just here to remind us that lab work is excessively harsh and merciless.
+Fluorescent reporter proteins were used to cope with the need for \textit{in vivo} measurements with a high-frequency resolution.
+Advances in microfluidics were exploited to perform steady-state-to-steady-state transitions from one medium to another, in order to temper the effect of pre-culturing history~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}.
+Finally, the signal of interest was reconstructed through the use of Kalman smoothing, a powerful signal processing algorithm that has, as we showed, many features that make it suitable for microscopy time series~\cite{kailath_linear_2000,jazwinski_stochastic_2007,kalman_new_1960}.
+Overall, the set-up still needs many improvements, though most of them are experimental peculiarities that are just here to remind us that lab work is excessively harsh and merciless.
 
@@
 This prediction followed directly from the critical assumption that microorganisms maximize their biomass production at steady state and over growth transitions.
-While we were not able to conclude in this manuscript, we can project ourselves in the two possible scenarios that will arise from the future improvements of the experimental set-up.
+While we were not able to conclude in this manuscript, we can project ourselves in the two possible scenarii that will arise from the future improvements of the experimental set-up.
 
@@
 As was performed to establish the steady-state growth laws, we would need measurements in broader environmental conditions.
-The most straightforward would be to test additional upshift and downshift schemes (others sugars, amino-acids supplementation, azote instead of carbon starvation...).
+The most straightforward would be to test additional upshift and downshift schemes (others sugars, amino-acids supplementation, azote instead of carbon limitation...).
 The experimental set-up could also be slightly modified to enrich the conditions controlled for.
@@
 By only comparing the regulatory schemes \textit{via} their benefits, we implicitly made the assumption that their cost are negligible, or at least comparable in regards of the difference in benefits.
-The costs of regulatory schemes are actually difficult to model, as organisms always prove more and more inventive when it is about minimizing it~[citation Alon modeling cost, citation free regulation somewhere in my mailbox].
+The costs of regulatory schemes are actually difficult to model~\cite{shachrai_cost_2010,dong_gratuitous_1995,dekel_environmental_2005}.
 All variables in the system are not equivalently costly to measure, and as we showed with the ppGpp system, their combinations do not make their cost necessarily additive.
@@
 Overall, we could expect these two functions to cross at a point that would represent an evolutionary bottleneck, beyond which the costs outclass the benefits (see~\cite{short_flows_2006} for another example of such a bottleneck).
-Interestingly, the position of the bottleneck would be environment-dependent, establishing a dynamic law that would link the complexity of the regulations with the dynamics of the environment in which the organism evolved.
-In other words, while the current growth laws consider how the physiology of the cell vary with the nutrient quality of the medium, we could add another dimension that represents the actual availability of this nutrient in the natural environment...
+Interestingly, the position of the bottleneck would be environment-dependent, establishing a dynamic law that would link the complexity of the regulations with the dynamics of the environment in which the organism has evolved.
+In other words, while the current growth laws consider how the physiology of the cell vary with the nutrient quality of the medium, we could add another dimension that represents the actual availability of this nutrient in the natural environment.

### 2016-12-23 Chapters/theory.tex
minor correction in theory.tex. Version of intro and discussion sent to Hidde and Hans

@@
 
-\textit{Rather than propose a new theory or unearth a new fact, often the most important contribution a scientist can make is to discover a new way of seeing old theories or facts.} -- Richard Dawkins, \textit{The Selfish Gene}, preface to 1989 edition.
+\textit{Rather than propose a new theory or unearth a new fact, often the most important contribution a scientist can make is to discover a new way of seeing old theories or facts.} -- Richard Dawkins, \textit{The Selfish Gene}~\cite{dawkins_selfish_1976}, preface to 1989 edition.
 
@@
 Analysis of the nondimensionalized system allows a number of properties of the solution of the optimal control problem of Eq.~\ref{eq:J} to be derived (\textit{Methods} and \ref{S3_Text}).
-First, by applying a version of the well-known Pontryagin Maximum Principle \cite{carlson_infinite_1991}, we can prove that the optimal solution is obtained for an alternating sequence of $\alpha(\cdot)=0$ and $\alpha(\cdot)=1$, possibly ending with an intermediate value of $\alpha(\cdot)$, corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the post-upshift environment $E_M$.
+First, by applying a version of the well-known Pontryagin Maximum Principlek~\cite{carlson_infinite_1991}, we can prove that the optimal solution is obtained for an alternating sequence of $\alpha(\cdot)=0$ and $\alpha(\cdot)=1$, possibly ending with an intermediate value of $\alpha(\cdot)$, corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the post-upshift environment $E_M$.
 Second, if the optimal solution reaches the optimal steady state for the new environment, then it does so after an infinite number of switches of $\alpha (\cdot)$ between 0 and 1.

### 2016-12-27 Chapters/experiments.tex
Hidde's correction of chapter3

--- 2016-11-29 Chapters/experiments.tex
@@
 
+Most studies of the growth of microorganisms have been done at steady state.
+This is a reasonable and logical choice given that at steady state, the behavior of microorganisms is reproducible, which helps uncovering simple theories to apprehend their inherent complexity.
+But while this condition can be easily achieved in the laboratory, microorganisms naturally spend very little time in steay state.
+This motivated the construction of a new theoretical framework in Chapter~\ref{chap:theory} to study growth laws during growth transitions.
+
+The principles underlying regulatory processes in dynamical conditions appear to be different from those applying at steady state.
+We showed that regulatory strategies approaching the theoretical optimal, given by bang-bang control of gene expression machinery, have to be able to effect abrupt and strong variations in the activity of a gene.
+It was seen that for implementing such strategies the regulatory system needs to be capable of sensing the internal state of the cell and not just the environment.
+A near-optimal transition also requires information of several different variables, whereas for maintaining a steady state with maximal biomass accumulation a single variable turned out to be sufficient in our theoretical framework.
+Interestingly, we showed that the ppGpp regulatory system of \textit{E.
+coli} fulfills the requirement of such a regulatory strategy, by implementing an on-off strategy for regulating the synthesis of ribosomes.
+However, although this provides circumstantial evidence that bacteria control resource allocation in a manner consistent with theoretical optimality, experimental data are necessary to decide if an on-off strategy is at work in \textit{E.
+coli} during growth transitions.
+
+Unfortunately, far more information is available on ribosome abundance at steady state than during growth transitions [refs].
+The main reason for this bias on steady-state conditions is that, from an experimental point of view, growth transitions are hard to control and may depend on the history of the culture~[citation ?
+Bremmer ?].
+However, as we discussed in Section~\ref{sec:discussion}, the few studies reported that have been reported seem to be consistent with our predictions, in the sense that they indicate that during growth transitions, the synthesis rate of ribomoses oscillates~\cite{gausing_regulation_1980,zengel_transcription_1986} and the ppGpp concentration manifests a rapid succession of increases and decreases~\cite{friesen_synthesis_1975,murray_control_2003}.
+But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
+What is needed for the verification of our predictions are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
+While the observation of an on-off strategy would show that simple models of the type introduced in Chapter~\ref{chap:theory} are instrumental in gaining a better understainding of microbial physiology, its refutation would also raise interesting questions.
+If maximization of growth rate and biomass accumulation have been retained by natural selection in \textit{E.
+coli}, which factors would prevent the cell from behaving optimally from a theoretical point of view?
+
+The aim of this chapter is to measure \textit{in vivo}, during a growth transition, the resource allocation profile $\alpha (\cdot)$ and compare its dynamics with the gold standard established in Chapter~\ref{chap:theory}.
+To this end, we have measured ribosomal abundance of \textit{E. coli} at the single-cell level during a nutrient upshift.
+More precisely, we constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing the \textit{in-vivo} monitoring of (changes in) the abundance of ribosomes.
+In collaboration with Irina Mihalcescu of the Laboratoire Interdisciplinaire de Physique, we cultivated this \textit{E.
+coli} strain in a microfluidic device, allowing the long-term imaging of individual cells in well-controlled conditions, notably involving a classic upshift experiment from growth in minimal medium with acetate to minimal medium with glucose.
+We then developed a Kalman smoothing method, in collaboration with Eugenio Cinquemain of Inria Grenoble - Rh\^{o}ne-Alpes, to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift, which allows $\alpha (\cdot)$ to be reconstructed.
+While the preliminary results presented here do not allow a decisive validation of the expected behavior, among other things due to the experimental difficulties that were encountered and image analysis challenges, we believe they are promising as a first step towards the better understanding of global resource allocation during growth transitions.
+
+For clarity, the model of resource allocation during growth transitions developed in Chapter~\ref{chap:theory} is reproduced here, where dotted variables represent time derivatives:
+\begin{eqnarray}
+\dot{p}(t) &=& e_M(t)\cdot (1/\beta - r(t)) - \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (1+\beta\, p(t)), \label{eq:pdef-exp}\\
+\dot{r}(t) &=& \frac{k_R \cdot p(t)}{K_R + p(t)}\cdot r(t) \, (\alpha(t) - \beta\, r(t)). \label{eq:rdef-exp}
+\end{eqnarray}
+In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
+$e_M(t)$ [min\textsuperscript{-1}] is an indicator of the richness of the environment.
+$p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
+$r(t)$ [g.L\textsuperscript{-1}] is the concentration of gene expression machinery inside the cell.
+$\alpha (t)$ [$\emptyset$] is the fraction of resources allocated to the synthesis of gene expression machinery.
+$k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
+$K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
+$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules, assumed to be constant.
+
+\section{Results}
+
+\subsection{Experimental design}
+\label{sec:exp_design}
+
+How can one measure the resource allocation variable in \textit{E. coli} cells?
+There is no direct way to quantify $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
+To correctly reconstruct it, one has to know the value of every single term in the equations in which it appears.
+In the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only appears in Eq.~\ref{eq:rdef-exp}.
+We can therefore identify $\alpha (t_k)$ for each time $t_k$ when the terms $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$ are known, or equivalently, the six individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$ they are composed of.
+Even when assuming that the parameters $k_R$, $K_R$, and $\beta$ are known, or at least easy to estimate independently, reconstructing the dynamics of $\alpha$ would require estimation of the (change in) concentration of the gene expression machinery ($r$, $\dot{r}$) and the precursor concentration $p$.
+
+Are such measurements feasible in practice?
+The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques \cite{HwaMSBProteomics}.
+However, only the use of fluorescent proteins complies with our need for real-time single-cell quantification.
+Obtaining an estimate of precursor abundance is even more challenging, since there is no clear proxy for the totality of precursors in the cell and, while absolute quantification of all internal metabolites has been achieved in \textit{E.
+coli}~\cite{bennett_absolute_2009}, this method is also not suitable for the estimation of the precursor concentration dynamics in individual cells.
+
+The problem of quantifying $\alpha$ can be simplified by applying a transformation of the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}.
+Taking into account that by construction the growth rate $\mu(t)$ is given by
+\[
+	\mu (t) = \beta \frac{k_R \cdot p(t)}{K_R + p(t)} \cdot r(t),
+\]
+we can rewrite Eq.~\ref{eq:rdef-exp} as follows
+\begin{equation}
+\label{eq:dot_r}
+\dot{r}(t) = \mu (t) \, \left(\frac{\alpha(t)}{\beta} - r(t) \right).
+\end{equation}
+The problem of estimating $\alpha(\cdot)$ is thus equivalent to the problem of estimating $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$.
+If we are satisfied with estimating $\alpha$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (\cdot) / \beta$ identifiable.
+
+Given that fluorescent reporters of ribosomal proteins can provide information on both $r(\cdot)$ and $\dot{r}(\cdot)$ \cite{ValentinBioinformatics}, the above reformulation of the problem is promising for the purposes of our study.
+Moreover, since the predicted bang-bang profiles in gene activity are easily hidden at the population level if cells are not synchronous, which is generally the case, expression of the ribosomal genes need to be monitored at high sampling density on the single-cell level, which is also possible with fluorescent reporters.
+We therefore constructed an \textit{E. coli} strain in which the S2 ribosomal subunit, encoded by the gene \textit{rpsB}, has been tagged with a green fluorescent protein (GFP).
+The strain with the chromosomal \textit{rpsB-gfp} fusion, inspired by previous work \cite{refBakshi}, produces fluorescent ribosomes that are quantifiable \textit{in vivo} in single cells while not affecting the growth rate (see Material and Methods~\ref{sec:methods_strain}).
+The monitoring of single cells of this strain, by time-lapse fluorescence miscroscopy, thus makes it possible to estimate $r(\cdot)$ and $\dot{r}(\cdot)$ and reconstruct $\mu(t)$.
+
+The growth conditions also need to be carefully chosen for this study.
+In natural environments, for many bacteria one of the most frequently encountered growth transitions, or at least the transition on which selection is expected to operate, is the transition from stationary phase in a nutrient-deprived medium [see reference in recent PNAS paper by Balaban] to (exponential) growth after the renewed availability of nutrients.
+The transition from stationary to exponential phase is difficult to study though as many hard-to-control parameters play a role, including the duration of the nutrient stress, the composition of the medium before growth arrest, and the accumulation of waste products in the medium \cite{reviewBremerDennisBiochimie}.
+As a consequence, growth resumption only occurs after a lag phase of variable duration~[citation lag phase review to find], which is not included in the model.
+For this reason, we decided to focus on a transition from exponential growth on a poor carbon source (acetate) to exponential growth on a rich carbon source (glucose) in a well-defined minimal medium.
+By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively, to ensure that the cells have no memory of their physiological state before the transition and have enough time to reach the new steady state after the transition.
+The use of the mother machine, a microfluidic device that was designed to sustain exponential growth for long periods of time, is therefore a good choice for this purpose.
+It allows to maintain a constant flow of fresh medium and to effect fast transitions by switching the medium (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
+The complete experimental plan is summarized in Fig.~\ref{fig:experiment_schema}.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/experiment_schema.eps}
+\caption{
+\textbf{Schematic outline of the upshift experiment.}
+The goal is to measure the fluorescence and length/area of \textit{E. coli} \textit{rpsB-gfp} cells during an acetate-to-glucose upshift.
+We use M9 minimal medium supplemented with 0.2\% of acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
+A preculture was started from glycerol stock for 2.5 days on 0.2\% acetate in batch condition (shake flask).
+The day of the experiment, the cells were injected into the mothermachine and fed by a constant flow of fresh 0.2\% acetate medium (see Material and Methods~\ref{sec:microflu}).
+Fluorescence and phase contrast images were taken every 5 minutes.
+After 20~h, the feeding media was switched to 0.2\% glucose and maintained for 20~h while continuing image acquisition.
+Time 0 corresponds to the moment of the acetate-glucose upshift.
+}
+\label{fig:experiment_schema}
+\end{figure}
+
+\subsection{Data acquisition}
+\label{sec:data_acquisition}
+
+An experiment following the plan of Fig.~\ref{fig:experiment_schema} was carried out, over a period of 40 h, but encountered a number of difficulties.
+First, the motor displacing the microscope along the Z-axis intermittently stalled, thus deactivating the autofocus and the Z-compensation for the tilt of the [microfluidics???] device in the XY plane.
+As a consequence, until the experimenter manually intervened, many of the images acquired were out of focus.
+This is the reason why data points between -720 and -150 min are not exploitable, but fortunately the experiment was planned in such a way that when it really mattered, notably around the growth transition, someone was present to monitor the microscope.
+Second, at around 7~h after the nutrient upshift, the bacteria started to die for reasons that are unknown, the most plausible hypothesis being a phage contamination.
+Data analysis was therefore limited to 550 min after the upshift, when roughly 3/4 of the cells were still growing.
+For these and other reasons, the experiment will have to be repeated (Section~\ref{sec:chap3_discussion}), but there was no time left for this in the framework of my PhD thesis.
+
+The acquired data consisted of phase contrast and fluorescence images of 60 wells in total, in four different fields.
+While quite a few image analysis programs have been reported in the literature~[citations schitcells + others], and some of these specifically address the mother machine design of the microfluidics device~[molysso + Wang et al], they all presented limitations when applied to our data.
+For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidics device.
+[LACK OF ALIGNMENT IS NOT VERY STRONG ARGUMENT, UNLESS IT CANNOT BE REPAIRED]
+Moreover, the fluorescence density was concentrated in hot spots (Fig.~\ref{fig:data_acquisition}) and its intensity varied during the experiment. 
+While this was expected, since ribosomes are localized in the cell poles~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}, it complicated automatic segmentation on the fluorescence images.
+
+For these reasons, the image analysis carried out for this chapter has been much simplified. 
+First, we have focused on the cell at the bottom of each well, since this avoids the tracking of individual cells across several generations and ensures that the descendance of this cell can be followed throughout the experiment.
+Second, segmentation was done by manually selecting two pixels, one at each pole of the cell.
+These pixels were used to create a rectangle surrounding the cell (Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
+After background correction (Material and Methods~\ref{sec:cell_segmentation}), the fluorescence intensity in units RFU was evaluated for each pixel in the rectangle, as well as the length of the cell, defined by the distance in pixels between the poles (Fig.~\ref{fig:data_acquisition}).
+The fluorescence density for the entire cell [RFU/pixel/cell] was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
+Although we are well aware that the above procedure can be improved on many counts (Section~\ref{sec:chap3_discussion}), we nevertheless believe that it provides a reasonable approximation of the quantities of interest and a valid starting-point for the estimation of the growth rate and the resource allocation profile in the remainder of this chapter.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/data_acquisition}
+\caption{
+\textbf{Results of data acquisition.}
+We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
+From the 90 wells, 68 were suitable for further analysis (the others being empty, out of frame for some period of time, or plugged).
+For all wells, data points are missing in the interval between -720 and -150 min because the camera was out of focus.
+We stopped the analysis at 550 min, when about 3/4 of the bacteria were still growing, before the entire population died within a few hours for an unknown reason.
+The image labelled "Raw image" is the last image analyzed for the highlighted well.
+The bacterium on the left of this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
+A 6-pixel-wide rectangular mask was computed for each image, resulting in the "Segmented image" on the right.
+Fluorescence intensities are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335) and were corrected for camera background, but not autofluorescence background (Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
+The fluorescence intensity of the cell, expressed in units RFU/pixel/cell, was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
+The cell length is the distance in pixels between the two poles (red line on the raw and segmented images).
+The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labelled "X3Y2, W2".
+The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
+}
+\label{fig:data_acquisition}
+\end{figure}
+
+In total, we obtained time courses of fluorescence density [RFU/pixel/cell] and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
+The fluorescence density appears constant during growth on acetate (before time 0) and immediately increases when the carbon source is switched to glucose.
+Inspection of the cellular length shows that the division frequency abruptly increases after the nutrient upshift, corresponding to a higher growth rate.
+Unfortunately, while steady-state growth on acetate was reached before the upshift, the fluorescence density profile suggests that the experiment did not continue long enough to ensure that a new steady state on glucose was reached.
+However, the data before and after the nutrient upshift are exploitable.
+What do we observe if we estimate the growth rate and the resource allocation profile from these data?
+
+\subsection{Data analysis using Kalman smoothing}
+\label{sec:res_kalman}
+
+As we presented in section~\ref{sec:exp_design}, our goal is to reconstruct the signal $\alpha (\cdot)$ during a growth transition.
+By modeling \textit{E. coli} as a cylinder, the volume can be assumed directly proportional to the length of the cell.
+If we assume a negligible background of autofluorescence (see Material and Methods~\ref{sec:cell_segmentation}), the fluorescence concentration in the cell can be assumed proportional to the ribosome concentration.
+More precisely, based on the available data, we have the following measurement model at each time-point $t_k$, $0\leq k \leq N-1$:
+\begin{eqnarray}
+L(t_k) &= \lambda \cdot V(t_k) + \epsilon_k, \label{eq:mes_L}\\
+F(t_k) &= \gamma \cdot r(t_k) + \eta_k, \label{eq:mes_F}
+\end{eqnarray}
+where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in units RFU/pixel measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$) the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
+($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) measurement noise.
+[YOU MAY NEED TO SPECIFY THE ASSUMPTIONS ON THE NOISE DISTRIBUTION (0 MEAN, NORMAL?)]
+
+As explained in Section~\ref{sec:exp_design}, reconstructing $\alpha (\cdot)$ requires information on $\mu (\cdot)$, $r(\cdot)$ and $\dot{r}(\cdot)$.
+From Eq.~\ref{eq:mes_F}, we can obtain $r(\cdot)$ and $\dot{r}(\cdot)$ by smoothing interpolation and differentiation.
+Similarly, $\mu (\cdot)$ can be derived from Eq.~\ref{eq:mes_L} as it is defined by
+\[
+\dot{V}(t) = \mu (t) \cdot V(t).
+\]
+However, smoothing interpolation is particularly sensible to the boundary conditions.
+Since each division in the time series generates a new boundary condition, smoothing interpolation of our data is expected to be little robust.
+For this reason, the growth rate of single cells is usually estimated by fitting an exponential function to the volume data between each cell~[citation Wang et al et Izard et al].
+While this is suitable when bacteria are at steady state, it is not applicable during a growth transition, where we expect the growth rate to vary between successive divisions.
+Other techniques are less sensible to the above problem~\cite{zulkower_robust_2015}, and we decided to use Kalman smoothing for our purpose~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+
+Kalman filtering~\cite{kalman_new_1960} is a Bayesian algorithm using a series of noisy measurements to predict the state of a dynamical system.
+It has been extensively used in engineering applications (guidance, tracking, control, ...) requiring the real-time estimation of hidden variables in a dynamical system from present and past measurements.
+Kalman smoothing is an extension of Kalman filtering using information about past and present but also future measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+The advantage of Kalman filtering with respect to Kalman smoothing is that it uses all available information to estimate the hidden variables of a dynamical system, in applications where a real-time response is not required.
+For our problem, at each time step $t_k$, our hidden variables are the resource allocation variable $\alpha (t_k)$ and the growth rate $\mu (t_k)$, while the available information consists of the $N$ measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$ that were taken during the experiment.
+
+The Kalman filtering problem can now be formulated as follows.
+The state of a growing bacterial cell is described by the dynamical system
+\begin{eqnarray}
+\dot{r}(t) &=& \mu (t) \cdot \frac{\alpha(t)}{\beta} - \mu (t) \cdot r(t),\\
+\dot{V}(t) &=& \mu (t) \cdot V(t),
+\end{eqnarray}
+with the following measurement model:
+\begin{eqnarray}
+L(t_k) &= \lambda \cdot V(t_k) + \epsilon_k,\\
+F(t_k) &= \gamma \cdot r(t_k) + \eta_k,
+\end{eqnarray}
+where the variables are as defined as for Eqs~\ref{eq:dot_r}, \ref{eq:mes_L} and \ref{eq:mes_F}.
+While $\beta$ can be obtained from the literature (Chapter~\ref{chap:theory}), $\lambda$ and $\gamma$ are unknown parameters that need to be estimated along with $\alpha (\cdot)$ and $\mu (\cdot)$.
+Since we are satisfied with a qualitative reconstruction of $\alpha (\cdot)$, we can simplify the system by defining $r_\gamma = \gamma \cdot r$ and $V_\lambda = \lambda \cdot V$.
+The dynamical system is consequently rewritten as
+\begin{eqnarray}
+\dot{r}_\gamma(t) &=& \mu (t) \cdot\frac{\gamma \cdot \alpha(t)}{\beta} - \mu (t) \cdot r_\gamma (t),\label{eq:r_prob}\\
+\dot{V}_\lambda}t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
+\end{eqnarray}
+with the new measurement model:
+\begin{eqnarray}
+F(t_k) &= r_\gamma (t_k) + \eta_k,\label{eq:F_prob}\\
+L(t_k) &= V_\lambda(t_k) + \epsilon_k.\label{eq:L_prob}
+\end{eqnarray}
+The Kalman filtering problem thus becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ from measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+
+Note that the above problem is not linear, whereas Kalman filtering was initially introduced for linear systems~\cite{kalman_new_1960,kailath_linear_2000}.
+For this reason, we use a nonlinear extension of Kalman smoothing called unscented Kalman smoothing~\cite{julier_new_1997,jazwinski_stochastic_2007}.
+Details about its exact implementation are available in Material and Methods~\ref{sec:meth_kalman}.
+While it is feasible to simultaneously reconstruct the signals $\mu (\cdot)$ and $\gamma \alpha (\cdot) / \beta$ using Kalman smoothing, it is not necessary in our case and might lead to unstable results.
+We therefore decided to proceed in two steps: first reconstruct $\mu(\cdot)$ from the measurements $L(\cdot)$, than inject this result into the reconstruction of $\gamma \alpha(\cdot) / \beta$ from the measurements $F(\cdot)$.
+
+\subsection{Estimation of growth rate and resource allocation profile}
+
+\subsubsection*{Growth rate estimation}
+
+As presented in the previous section, we want to reconstruct the growth rate $\mu$ defined by Eq.~\ref{eq:V_prob}, using measurements of the length $L$ defined by Eq.~\ref{eq:L_prob}.
+Note that Kalman smoothing is a procedure that returns the expected mean and covariance of the state of a dynamical system, given the  measurements.
+Therefore, reconstructing $\mu$ requires it to be explicitly included as a state variable of the dynamical system [I GUESS THIS IS WHAT YOU MEAN?].
+This provides constraints on the variation of $\mu$ that have the effect of a regularization.
+In particular, in the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process.
+The laws describing this process play the role of a prior on the expected regularity of $\mu$ (the more noisy the process, the less constrained the variations of $\mu$).
+In particular, we define $\mu$ as the double integral of a Gaussian white noise $w$:
+\[
+\dot{\mu} = v(t), \;\;\;\; \dot{v} = w(t),
+\]
+where $v$ is an intermediate variable and $w(t) \sim \mathcal{N}(0, \theta)$. [WHAT IS $\theta$?]
+The resulting system then becomes
+\begin{eqnarray}
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
+\dot{\mu}(t) &=& v(t),\label{eq:full_mu_prob}\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray}
+with the measurement model
+\begin{equation}
+L(t_k) = V_\lambda(t_k) + \epsilon_k.
+\end{equation}
+The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable.
+It is also equivalent to other methods like Tikhonov regularization~[citation to take from Val pub], which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
+
+Our prior is thus entirely contained in the parameters of the Kalman procedure:
 \begin{itemize}
-\item Understanding growth transitions is important, because so and so.
-\item We predicted earlier counter-intuitive bang-bang-singular behaviours for GEM during transition
-\item But we're missing experimental data on what really happens in real cells during these transitions (quick review of available data)
-\item We fix it by making a strain with fluorescent ribosomes and monitoring ribosome abundance during growth transitions in a microfluidic device.
+\item the mean of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the covariance of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$,
+\item the observation variance for $L$.
 \end{itemize}
+[IF YOU DESCRIBE THE PROCEDURE IN SO MUCH DETAIL, THEN YOU NEED TO EXPLICITLY INTRODUCE THE MEANS/(CO)VARIANCES.
+THESE PARAGRAPHS ARE ONLY COMPREHENSIBLE FOR THE SPECIALISTS, BUT IN THAT CASE NEED TO BE MATHEMATICALLY PRECISE.]
+
+The means and covariances at the initial state represent the knowledge we have of the initial value of the variables. [YOU NEED TO EXTEND THE MODEL WITH THE INITIAL VALUES]
+For instance, if we know exactly the initial value of $V_\lambda$, $\mu$ and $v$, we can fix the means and use very small covariances.
+In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because they are not the result of a stochastic process.
+However, the transition variance $\theta^2$ of the derivative of $v$ is crucial: it represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
+The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu$.
+Finally, the observation variance for $L$ is the expected measurement noise.
+Together,  these parameters represent the priors that we use to reconstruct the growth rate $\mu$.
+As we will see below, we widely use these properties to overcome the difficulties introduced by the discontinuities following cell divisions.
+
+The observation of the length $L$ of growing bacterial cells is inherently discontinuous, due to the division of cells at regular time-points (Fig.~\ref{fig:data_acquisition}).
+The estimation problem can therefore not be solved for the experiment as a whole, but only for the time-intervals between two successive division events, when the cell length is expected to change continuously.
+We localize the division events by identifying the time points at which the empirical derivative of the length is below an arbitrary threshold, and correspondingly slice the total duration of the experiment into time-intervals with continuous change of cell length.
+The simplest way to proceed from there on would be to treat each portion independently.
+However, by doing this we would lose important information, since the growth rates of a mother and its daughter cells are expected to be similar.
+In other word, while not necessary continuous, the growth rate of new-born cells will depend strongly on the growth rate of their mother just before division.
+This can be easily taken into account in the Kalman smoothing procedure by exploiting the priors about the initial mean and covariance of the system variables.
+In particular, we set the mean of the initial growth rate of a daughter cell equal to the final estimated growth rate of the mother cell, while defining a reasonable variance of the growth rate to allow for uncertainty.
+Right after the upshift, when the above assumption may not hold, we set the mean of the initial growth rate to the estimated growth rate right before the upshift, but define a large variance.
+The results of this Kalman smoothing procedure, along with the exact parameters used, are reported in Fig.~\ref{fig:growth_rate_estimation} and Material and Methods~\ref{sec:meth_kalman}.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/growth_rate_estimation}
+\caption{
+\textbf{Growth-rate estimation using Kalman smoothing based on measurement of the length of bacteria growing in a microfluidics device.}
+Gray lines represent the estimation of the time-varying length (upper plot) and growth rate (lower plot) of 68 cells by the unscented Kalman smoothing procedure.
+The solid red lines highlight the result for one particular cell, located at the bottom of the well labelled "X3Y2, W2", while black crosses in the top graph are the data points for this cell.
+The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
+As prior for the algorithm, we used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
+The transition variance $\theta^2$ (\textit{i.e.}, the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6}.
+Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean growth rate equal to the last estimated value for $\mu$ before cell division, and to half the last estimated value for $V_\gamma$, bearing in mind that \textit{E.
+coli} cells divide symmetrically.
+At the start of the experiment, when no mother cell is available to provide initial estimates, the above values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
+The initial mean of $v$ is set equal to 0, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
+All the cross-covariances are set to 0 because the system variables are independent by construction.
+}
+\label{fig:growth_rate_estimation}
+\end{figure}
+
+On average, the results presented in~Fig.~\ref{fig:growth_rate_estimation} show a roughly constant growth rate on acetate, followed by a quick increase after the upshift.
+Contrary to what was visible in the plot with the fluorescence data, a steady state for the growth rate seems to have been reached before the end of the experiment.
+When considering the individual cells, the results are more difficult to interpret.
+A significant part of the cells stopped growing before the end of the experiment, probably due to the fact that all cells die from an unknown cause around the end of the experiment. 
+As a consequence, the analysis has been limited to a time-interval after the upshift in which roughly 3/4 of the cells are still growing.
+Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a 1~h-long pause around 200 min after the upshift, followed by a recovery of the growth rate until the end of the experiment (see Supporting Information XX).
+In order to focus the analysis, we decided to classify the cells in three categories: dying cells (N=12) for which the growth rate drops to zero after the upshift, pausing cells (N=11) for which the growth rate drops to zero after the upshift and then recovers, and finally so-called normal cells (N=45) which do not exhibit any of the above behaviors.
+More information about each category is available in Supporting Information XX.
+In what follows, we focus on the 45 normal cells (Fig.~\ref{fig:growth_rate_estimation_median}) which show a globally coherent behavior in the time window considered here.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/growth_rate_estimation_median}
+\caption{
+\textbf{Growth-rate estimation using Kalman smoothing for the normal cells only.}
+The top graph is the same as the bottom graph of Fig.~\ref{fig:growth_rate_estimation}, except that the dying and pausing cells were removed.
+The bottom graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step.
+The gray area represents the interquartile range.
+}
+\label{fig:growth_rate_estimation_median}
+\end{figure}
+
+\subsubsection*{Estimation of resource allocation profile}
+
+In order to estimate the time-varying allocation of resources after the upshift, we use a similar Bayesian regularization approach as for the estimation of the growth rate.
+We note $u(t) = \gamma \alpha (t) / \beta$ the resource allocation input that we wish to reconstruct.
+The full model for the estimation of $u$ is given by
+\begin{eqnarray}
+\dot{r}_\gamma(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
+\dot{u}(t) &=& v(t),\label{eq:full_u_prob}\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray}
+with the measurement model
+\begin{equation*}
+F(t_k) = r_\gamma (t_k) + \eta_k,\\
+\end{equation*}
+where $\hat{\mu}$ is the estimation of $\mu$ obtained in the previous section.
+Contrary to the model used for the estimation of the growth rate, the system of Eq.~\ref{eq:full_u_prob} is linear.
+Furthermore, we do not have to deal with discontinuities in the measurements, since the fluorescence density $F$ in the cells is continuous between mother and daughter cells (Fig.~\ref{fig:data_acquisition}).
+
+Given the analysis in Chapter~\ref{chap:theory}, we expect $u(\cdot)$ to exhibit bang-bang variations after the upshift.
+This is a discontinuous signal that could be complicated to reconstruct for certain choices of the parameters of the regularization method.
+In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of Chapter~\ref{chap:theory}.
+After estimating the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (Supporting Information XX), choosing parameters that allow the model to reproduce the observed growth rates and fluorescence densities, we simulated an upshift from acetate to glucose.
+We then tried to estimate $\gamma \alpha(\cdot) / \beta$ from these synthetic data, selecting the \textit{a priori} values of the parameters in the Kalman smoothing procedure by trial and error.
+[THE USE OF PARAMETERS IS AMBIGUOUS HERE: YOU PROBABLY DO NOT REFER TO MODEL PARAMETERS AS IN THE PREVIOUS PHRASE, BUT TO THE PARAMETERS OF THE KALMAN PROCEDURE]
+This approach may not have been optimal for our purpose and possible improvements are discussed in Section~\ref{sec:chap3_discussion}.
+The results along with the parameter values [BOTH TYPES] are shown in Fig.~\ref{fig:synthetic_upshift}.
+As expected, like any smoothing method, the algorithm has difficulty in reconstructing stiff variations in the state variables of Eq.~\ref{eq:full_u_prob}.
+The switching profile of the resource allocation input is relatively well captured though, which indicates that the algorithm is in principle capable of reconstructing an on-off strategy (further improvements are discussed in Section~\ref{sec:chap3_discussion}).
+In what follows, we use the same values for the parameters of the Kalman smoothing procedure for the normal cells identified in the previous section, to test the occurrence of an on-off switching profile in our data.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=0.8]{./Fig/synthetic_upshift}
+\caption{
+\textbf{Performance of the Kalman smoothing procedure on synthetic data simulating an acetate-glucose upshift.}
+\textit{(A)}~Synthetic data simulating an upshift from minimal medium with acetate to minimal medium with glucose, with and without additive white noise, as well as the results obtained by Kalman smoothing.
+The synthetic data were generated by simulating the model presented in Eqs~\ref{eq:pdef}-\ref{eq:rdef} with the on-off regulatory strategy (Eq.~\ref{eq:stratswitch} and Fig.~\ref{fig:onoffresults}).
+The model parameters used for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L g\textsuperscript{-1}, $K_R = 1$~g L\textsuperscript{-1}.
+The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~L g\textsuperscript{-1} RFU\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
+The noise level was estimated from the data (Supporting Information XX) and added to $F$.
+The choice of the parameters of the Kalman smoothing procedure is discussed in the Material and Methods~\ref{sec:meth_kalman}.
+\textit{(B)}~Estimation of the resource allocation profile $\gamma \alpha / \beta$ based on the data in \textit{(A)}.
+Following Chapter~\ref{chap:theory}, $\alpha (t)$ displays a bang-bang-singular profile during the upshift (dashed black curve).
+While the Kalman smoother is not able to capture the discontinuous variations in $\gamma \alpha / \beta$, it qualitatively reproduces the input quite well (red solid curve).
+\textit{(C)}~The predicted growth rate during the upshift experiment.
+This information is used as an input in the smoothing procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+}
+\label{fig:synthetic_upshift}
+\end{figure}
+
+The results of the reconstruction of the resource allocation profile $u = \gamma \alpha (\cdot) / \beta$ are presented in Fig.~\ref{fig:gene_activity}.
+Within the interval between -150 and 0 min, resource allocation remains more or less stable, as expected for steady-state growth on acetate.
+On the contrary, the reconstructed resource allocation profile seems particularly unstable at the beginning and the end of the experiment.
+While oscillations do occur after the upshift, these need to be taken with much care.
+The problem of reconstructing an on-off strategy is more challenging than we initially thought, for the simple reason that the expected signal is similar to the kind of artifacts a poorly calibrated regularization method would generate.
+Data about the pre-upshift and post-upshift steady state are crucial for the calibration of the method and, as explained in Section~\ref{sec:data_acquisition}, the experiment had to be interrupted before a steady state on glucose was attained.
+We extensively discuss this and other problems in Section~\ref{sec:chap3_discussion} and give directions for future improvements.
+
+Nevertheless, when not focusing on the individual cells but looking at the statistics of the entire data set, some interesting patterns emerge. 
+Fig.~\ref{fig:gene_activity_median} shows the median of the time-varying growth rate and resource allocation profile as well as the 25\%-75\% interquartile range. 
+The use of these statistics gives a more robust view on the population level of the response of the cells to a nutrient upshift. 
+When changing the carbon source from acetate to glucose, the growth rate increases to a value of around 0.011 min$^{-1}$, consistent with growth rates reported in the literature for the \textit{E.
+coli} strain used here \cite{Volkmer/Heinemann,Izard}, before decreasing when the first cells start to die (Fig.~\ref{fig:gene_activity_median}\textit{C}).
+In addition, the data show one period of an oscillation in the first 3~h after the upshift, conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}\textit{B}).
+The heatmap in Fig~\ref{fig:gene_activity_heatmap} reveals that almost all of the normal cells show this oscillatory feature.
+Moreover, the first peak is seen to be even more pronounced on the level of the individual cells, reflecting the fact that in the different cells it occurs at different times after the upshift and is therefore dampened out at the population level.
+While the results of the microfluidics experiment presented here do not allow to confirm the occurrence of an on-off strategy for resource allocation after a nutrient upshift, the preliminary data are encouraging and prompt further investigation.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity}
+\caption{
+\textbf{Estimation of the resource allocation profile using Kalman smoothing based on the fluorescence density measurements and the estimated growth rates from Fig.~\ref{fig:growth_rate_estimation}.}
+\textit{(A-B)} Gray lines represent the estimation of the fluorescence density (RFU/pixel) $F(\cdot)$ (in A) and the resource allocation profile $\gamma \alpha / \beta$ (in B) by the Kalman smoothing procedure for the 45 normal cells.
+The solid red curves highlight the result for one particular cell, , located at the bottom of the well labelled "X3Y2, W2", while black crosses in the top graph are the data points for this cell.
+[WHY THIS CELL AND NOT ONE OF THE CELLS YOU SENT US A WHILE AGO?
+THIS CELL HAS A PARTICULARLY IRREGULAR PROFILE...]
+The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
+The prior values for the parameters of the smoothing algorithm are exactly the same as those used for Fig.~\ref{fig:synthetic_upshift} and are reported in the Material and Methods~\ref{sec:meth_kalman}.
+}
+\label{fig:gene_activity}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/gene_activity_median}
+\caption{
+\textbf{Robust statistics for the estimation results presented in Fig.~\ref{fig:gene_activity}.}
+Each graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step for the signals reconstructed in Fig.~\ref{fig:gene_activity}.
+The gray area represents the interquartile range.
+Interestingly, while most oscillations in the resource allocation profile $\gamma \alpha (\cdot) / \beta$ cancel out at the population level, the first peak after the upshift is conserved.
+}
+\label{fig:gene_activity_median}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/}
+\caption{
+\textbf{Global overview of all estimated resource allocation profiles presented in Fig.~\ref{fig:gene_activity}.}
+The individual $\gamma \alpha (\cdot) / \beta$  curves have been normalized with respect to their maximum in the interval $[0, 400]$ and then ordered with respect to their maximum in the interval $[0, 100]$.
+
+}
+\label{fig:gene_activity_heatmap}
+\end{figure}
+
+\section{Discussion}
+\label{sec:chap3_discussion}
+
+As we showed in Chapter~\ref{chap:theory}, adopting a dynamical perspective might prove useful to unveil and understand the regulatory strategies employed by microorganisms.
+The criterion of biomass maximization has allowed to account for steady-state empirical growth laws.
+Interestingly, when the same criterion is applied to growth transitions, it predicts that microorganisms allocates their resources through bang-bang control of gene expression.
+For the control of ribosome synthesis, the implementation of this optimal control scheme results in an on-off strategy, where a bacterial cell is either producing only ribosomes or not producing them at all during the adaptation to a new growth medium.
+The ppGpp system, ubiquitous in bacteria and playing an important role in the control of ribosome synthesis in \textit{E. coli}, satisfies the requirements posed by the on-off strategy.
+However, we currently lack experimental data to show that it actually functions in this manner, among other things because of the difficulties to produce well-controlled growth transitions on the single-cell level in the laboratory.
+Can we set up appropriate experiments to measure the expression of ribosomes during a growth transitions?
+Will the bacterial cells exhibit the on-off strategy that was shown to be close to optimal in Chapter~\ref{chap:theory}?
+
+In this chapter, the above questions have been addressed by developing an experimental framework which allows the ribosome concentration to be quantified in individual \textit{E.
+coli} cells in real time.
+Our main contribution consists in bringing together several experimental techniques developed in recent years.
+First, inspired by the work of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, we constructed an \textit{E. coli} strain with fluorescent ribosomes.
+In their paper, Bakshi \textit{et al} essentially used this fluorescent marker to localize ribosomes in the cytoplasm through superresolution microscopy.
+While improving their design (Material and Methods~\ref{sec:methods_strain}), we used our strain for the purpose of quantifying ribosomal abundance in single cells.
+Second, we employed the mother machine developed by Wang \textit{et al.}~\cite{wang_robust_2010}, originally developed to study long-term steady-state growth of \textit{E.
+coli}, to observe individual \textit{E.
+coli} cells growing in continuous culture by means of time-lapse fluorescence microscopy.
+Here, the mother machine was used it to establish well-controlled growth transitions, \textit{i.e.} from steady-state growth in minimal medium with acetate to steady-state growth in minimal medium with glucose.
+By changing the medium source of the mother machine, we were able to observe how \textit{E. coli} cells adapt their growth rate and the expression of ribosomes.
+Overall, through the combination of fluorescent labeling of a core component of the gene expression machinery and the use of a microfluidic device, we have opened the way to experimentally study growth laws in a dynamical context.
+
+We applied this experimental set-up to the reconstruction of the resources allocation strategy employed by \textit{E.
+coli} during a nutrient upshift, with the aim to observe the bang-bang profile predicted in Chapter~\ref{chap:theory}.
+We developed a Kalman smoothing algorithm adapted to our question and showed how it can be implemented to reconstruct the internal state of a cell from time-lapse microscopy images.
+Similar to hidden Markov chains, Kalman filtering and smoothing are powerful techniques for the reconstruction of unobserved signals from noisy measurements, and have had numerous applications in a variety of domains, though less in biology than in other fields.
+We showed in this chapter that the Bayesian framework of Kalman smoothing can be exploited for growth rate reconstruction from a probabilistic prior defining the relation between mother and daughter cells.
+We also developed another varian of Kalman smoothing for the reconstruction of the resource allocation variable $\alpha (\cdot)$ that is at the heart of the models of Chapter~\ref{chap:theory}.
+Our work on synthetic data showed that, despite the challenge posed by the reconstruction of a discontinuous signal, the algorithm was capable of recovering an oscillatory pattern close to the expected on-off profile.
+Unfortunately, its application to the real data did not allow us an unambiguous conclusion to be drawn: while oscillations do occur, and a first peak between 0 and 120 min after the upshift is visiable in almost all of the cells, it turned out to be far from straightforward to distinguish a real oscillatory pattern from artifacts of the data analysis and signal reconstruction methods.
+Below we propose several improvements that will be addressed in future work.
+
+Several experimental problems complicated the analysis of the data.
+Sufficiently long steady-state measurements before and after the transition of interest are needed to provide a reliable estimation of the intensity and the nature of the experimental noise.
+They are also critical to calibrate the parameters of the regularization method used for signal reconstruction.
+Despite what was initially planned, we did not completely reach a new steady state after the transition.
+Several causes are to blame.
+First, the random stall of the motorization along the Z-axis of the microscope, which do not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
+Second, the massive death of bacteria starting after the transition on glucose.
+Future work will focus on solving these issues.
+
+[I DID NOT TOUCH THE PARARAPH THAT FOLLOWS]
+The death of bacteria is particularly worrying.
+During the construction of this strain, most people in the lab experienced contamination by an aggressive bacteriophage.
+Unfortunately, the strain used in this study was not spared, and our frozen stocks were later tested positive for phage-induced lysis.
+Interestingly, the lysis seems to depend strongly on the environmental conditions.
+To our disarray, the massive death observed in this experiment suggests that a strong environmental upshift could trigger the lytic cycle of the phage.
+While impeding any possible long-term measurement, the death of the bacteria is not our only worry.
+Phages are machines that are extremely well optimized to divert cell resources to their own end, especially the activity of the gene expression machinery.
+That could dramatically perturb the cell regulation of resource allocation, and so the conclusions of this study.
+For this reason, future work will start by the reconstruction of a clean phage-free strain from scratch.
+
+It is also possible to optimize the image analysis techniques used in this study.
+We segmented the images using the most powerful and ubiquitous segmentation algorithm available: the human brain.
+In other word, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a rectangle around the cell of interest.
+While this is far from satisfying, it may turn out to be difficult to improve upon this, as it seems that the first rule of image analysis is that every application is more or less unique.
+Every algorithm has to make assumptions about the object it is trying to recognize.
+For imaging of microbial cells, it could be the curvature at the poles, the size of the pixels, the homogeneity of the background, the homogeneity of intracellular fluorescence, ...
+While some parameters can be tweaked, some assumptions are always hard-coded in the algorithm and reflect the philosophy used to address the identification problem.
+Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~[fluobactracker.inrialpes.fr] that focuses on adapting their software to our set-up in a near future.
+It should provide a robust, automated way to identify and track the cells, enhancing the reproducibility of the analysis and increasing the number of cells available fro computing population statistics.
+
+The Kalman filtering algorithm that was later applied to these data can also be improved in many respects.
+As described in the Results section, the algorithm is an instance of Bayesian inference and several parameters describing the expected input define a probabilistic prior.
+In the work reported here, we chose these parameters through trial and error, or by calibrating the system on synthetic data whenever possible.
+A better technique would be to choose these parameters by \textit{generalized cross-validation} (GCV)~\cite{golub_generalized_1979}, a procedure maximizing the predictive power of the reconstructed signal by reducing overfitting.
+The use of GCV in our framework amounts to choosing the parameters of the Kalman smoother by optimizing predictions on a subset of the cells, then testing the performance on the remaining cells.
+While the initial conditions did not seem to play a huge role in the final shape of the signal, the choice of the smoothing factor $\theta^2$ is critical and may strongly benefit from the proposed GCV extension.
+
+The Kalman smoothing algorithm itself can also be further improved, notably by addressing the difficulty of estimating abrupt transitions.
+As can be seen in Fig.~\ref{fig:synthetic_upshift}, the resource allocation profile $\gamma \alpha (\cdot) / \beta$ reconstructed from the synthetic data increases before the nutrient shift and this artifact also occurs when using the measured fluorescent densities (Fig.~\ref{fig:gene_activity_median}), the change in resource allocation preceding the nutrient upshift by several dozens of minutes.
+While we know exactly when the change of medium occurs, this information is currently not used as a prior for the reconstruction.
+The Kalman smoother can be improved by implementing time-varying parameters for the regularization method, so as to strongly penalize variations of the reconstructed signal at steady state and to release this constraint following the change in medium.
+
+The improvements of the approach proposed in this section, which would help in reaching a conclusion about the existence of the on-off strategy, are not very complicated to realize.
+Another improvement consists in taking into account the autofluorescence of the bacteria, which we assumed to be negligible, mostly because its proper estimation in a dynamical, single-cell context is complicated (Material and Methods~\ref{sec:cell_segmentation}).
+A simple trick, however, could help in quantifying autofluorescence: mixing wild-type bacteria with bacteria having fluorescent ribosomes when loading the device.
+This would allow some channels to be occupied by wild-type cells, in addition to the  channels in which the strain of interest is growing, and thus obtain a reliable estimation of the level of autofluorescence of the cells.
+Notice though that the propertion between the two types of bacteria need to be carefully adjusted in order to preserve enough cells for the estimation of the resource allocation profile.
+
+[I WOULD SKIP THIS FINAL PARAGRAPH OR REDUCE IT TO ONE PHRASE AND ADD IT TO THE DISCUSSION IN THE PARARAPH STARTING WITH "The Kalman smoothing algorithm..."]
+Something is also displeasing in the current implement of the Kalman smoothing method.
+As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
+This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
+We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
+There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
 
@@
 
-\subsection*{Bacterial strain construction}
+\subsection{Bacterial strain construction}
+\label{sec:methods_strain}
 
-To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2} [citation Alon] to the C-terminus of rpsB (the gene coding the ribosome subunit S2).
-This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.
-coli}~\cite{bakshi_superresolution_2012}.
+To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} to the C-terminus of \textit{rpsB} (the gene coding the ribosome subunit S2).
+This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.~coli}~\cite{bakshi_superresolution_2012}.
 
-The transcription factor tsf is under the control of the same promoter than rpsB, and is located after the C-terminus end.
-In order to ensure that tsf expression was not affected by our translation fusion of rpsB, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
+The transcription factor \textit{tsf} is under the control of the same promoter than \textit{rpsB}, and is located after the C-terminus.
+In order to ensure that \textit{tsf} expression was not affected by our translation fusion on \textit{rpsB}, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
 This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
 
-A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus rpsB, and to the 3'-end of gfpmut2.
+A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
 The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
-This cassette is referred below as kan-P\textsubscript{BAD}-ccdB.
+This cassette is referred below as \textit{kan-P\textsubscript{BAD}-ccdB}.
 
-Another DNA fragment containing an ATG-less gfpmut2~\cite{zaslaver_comprehensive_2006} was amplified with long primers annealing respectively to the C-terminus region of rpsB (just before the stop codon), and the 5'-end of the kan-P\textsubscript{BAD}-ccdB cassette.
-The first primer also contained a 18-bp (base pair) linker that was selected according to the method section of~\cite{bakshi_superresolution_2012}.
+Another DNA fragment containing \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} without ATG was amplified with long primers annealing respectively to the C-terminus of \ref{chap:theory}rpsB (just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+The first primer also contained a 18-bp (base pair) linker that was chosen according to the method section of~\cite{bakshi_superresolution_2012}.
 
-Both fragments (the gfpmut2 reporter and the kan-P\textsubscript{BAD}-ccdB cassette) were assembled using Gibson assembly~[citation].
+Both fragments (the \textit{gfpmut2} reporter and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette) were assembled using Gibson assembly~\cite{gibson_enzymatic_2009}.
 Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
@@
 \begin{itemize}
-\item *(50 bp) the C-terminus of rpsB without the STOP codon 
-\item (18 bp) a linker
-\item (714 bp) the gfpmut2 sequence without the initial ATG
-\item (1851 bp) the kan-P\textsubscript{BAD}-ccdB
-\item *(50 bp) the region directly after rpsB in \textit{E. coli}
+\item *(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
+\item (18 bp) a linker (see Supporting Information)
+\item (714 bp) the \textit{gfpmut2} sequence without the initial ATG
+\item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette
+\item *(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli}
 \end{itemize}
-Regions labeled with * are supposed to anneal with the \textit{E. coli} chromosome.
+Regions labeled with * are expected to anneal with the \textit{E.~coli} chromosome.
 The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
 
-This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid providing Chloramphenicol resistance (lambda-red recombinaison).
+This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid with chloramphenicol resistance (lambda-red recombinaison).
 A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
-A new 100-bp fragment containing 50 bp of the 3'-end of gfpmut2 and 50 bp of the region just after rpsB was electroporated into this strain (sequence available in Supporting Information) to get rid of the kan-P\textsubscript{BAD}-ccdB cassette.
+A new 100-bp fragment containing 50 bp of the end of \textit{gfpmut2} and 50 bp of the region just after \textit{rpsB} was electroporated into this strain (sequence available in Supporting Information) to get rid of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
 An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
 Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
-The region after rpsB was sequenced and determined to be as anticipated (full sequence available in Supporting Information).
+The region after \textit{rpsB} was verified by sequencing (full sequence available in Supporting Information).
 
-In parallel, the same protocol was used to construct mCherry and cfp variants of the same strain.
-However, we were only able to recombine and obtain the gfp and mCherry versions for reasons that were not investigated.
-The full sequence of the final rpsB-mCherry strain is available in Supporting Information.
+In parallel, the same protocol was used to construct \textit{mCherry} and \textit{cfp} variants of the same strain.
+However, only the \textit{gfpmut2} and \textit{mCherry} versions were successfully obtained for reasons that were not investigated.
+The full sequence of the final \textit{rpsB-mCherry} strain is available in Supporting Information.
 
-The rpsB-gfp and rpsB-mCherry strains were caracterised on different media using a Tecan microplate reader.
+The \textit{rpsB-gfpmut2} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
 They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
-However, the rpsB-mCherry strain was shown to exhibit strange fluorescence dynamics, especially during growth transitions (entering stationary phase), which made it unsuitable for our study.
-Our effort were then concentrated on the rpsB-gfp strain.
+However, the \textit{rpsB-mCherry} strain exhibited strange fluorescence dynamics, especially during growth transitions, which made it unsuitable for our study.
+Our effort were then concentrated on the \textit{rpsB-gfp} strain.
 Data as well as information on the matter are available in Supporting Information.
 
+\subsection{Growth conditions}
+\label{sec:growth_condition}
+
+Sterilization was performed by autoclaving for instruments and filtration for solutions (0.2 $\mu$m).
+During cloning, bacteria were grown in 20-mL flasks filled with LB or spread on Petri dish with LA.
+In all the experiments, we used M9 minimal medium~[citation to find] supplemented with trace elements, thiamine and a carbon source.
+The full recipe is reproduced below.
+Numbers in squared brackets are characteristics of the stock solution.
+All stock solutions were stored at room temperature, except for FeSO$_4$~[30 g.L\textsuperscript{-1}] and Thiamine~[10 g.L\textsuperscript{-1}] that were stored respectively at -20$^\circ$C and 4$^\circ$C.
+\begin{center}
+\begin{tabular}{|l r|}
+  \hline
+  \multicolumn{2}{|c|}{\textbf{M9 medium (100 mL)}}\\
+  \hline
+  CaCl2 [1 mol.L\textsuperscript{-1}] & 10 $\mu$l\\
+  MgSO$_4$ [1 mol.L\textsuperscript{-1}] & 200 $\mu$l\\
+   5x Salts & 20 mL \\
+  Traces elements & 90 $\mu$l \\
+  FeSO$_4$ [30 g.L\textsuperscript{-1}] & 10 $\mu$l \\
+  Thiamine [10 g.L\textsuperscript{-1}] & 50 $\mu$l \\
+  Carbon source & at will \\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+  \hline
+\end{tabular}
+\begin{tabular}{|l r|}
+  \hline
+  \multicolumn{2}{|c|}{\textbf{Traces elements (0.9 mL)}}\\
+  \hline
+  H$_2$0 [18.2 M$\Omega$.cm] & 200$\mu$l\\
+  Na$_2$EDTA 2H$_2$O [150 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  ZnSO$_4$ 7H$_2$O [45 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CoCl$_2$ 6H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  MnCl$_2$ 4H$_2$O [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  H$_3$BO$_3$ [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  Na$_2$MoO$_4$ 2H$_2$O [4 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CuSO$_4$ 5H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  \hline
+\end{tabular}
+\begin{tabular}{|l r|}
+    \hline
+  \multicolumn{2}{|c|}{\textbf{5x Salts (100 mL)}}\\
+  \hline
+  Na$_2$HPO$_4$2H$_2$0 & 4.25 g \\
+  KH$_2$HPO$_4$ & 1.5 mg \\
+  NaCl & 0.25 g\\
+  NH$_4$Cl& 0.5 g\\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+  \hline
+\end{tabular}
+\end{center}
+
+A large quantity of M9 was prepared several days before the experiment, and stored at 4$^\circ$C.
+It contained all the necessary components except FeSO$_4$, thiamine, and carbon sources that were added the day the pre-culture were launched.
+Except stated otherwise, the growth temperature was 37$^\circ$C.
+M9 Acetate contains 0.2\% Acetate (in mass of C$_2$H$_3$O$_2$ per mass of solution), and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per mass of solution).
+
+At J-4, the glycerol stock containing the \textit{rpsB-gfp} strain was spread out on a Petri dish of LA, and incubated at 37$^\circ$C.
+At J-3, an isolated colony was inseminated in 20 mL of M9 Acetate, and incubated in a cotton-plugged sterile flask (Fig.~\ref{fig:experiment_schema}).
+Time of insemination was calculated to obtain a 0.3-0.4 cm\textsuperscript{-1} OD when starting the experiment, which correspond to a mid-exponential phase.
+
+At J+0, the preculture was concentrated by centrifugation and back suspended into 5 mL of M9 Acetate supplemented with 50 mg.mL\textsuperscript{-1} BSA (passivation	buffer) for injection into the microfluidic device.
+Channels were populated by diffusion until most of them were filled with cells ($\sim$1 hour).
+Data acquisition started after a constant medium flow was successfully obtained ($\sim$1 hour, depending on the quality of the device).
+
 \subsection{Microfluidic device}
+\label{sec:microflu}
 
-Mother machine because needed to change the medium, single cell data because asynchronization etc.
+We used the mothermachine~\cite{wang_robust_2010}.
+It consists of a series of growth wells (or channels), oriented at a 90$^\circ$ angle to a large central channel through which growth medium is passed at a constant flow.
+The wideness of the wells is selected as to constrain the growth direction of the bacteria.
+This design ensures that at least one cell per well (the deepest one) is preserved during the whole experiment, while the others incrementally escape into the central channel as divisions occur.
+For the fabrication of the devices, we thoroughly followed the step described in the supporting information of~\cite{wang_robust_2010}.
+We maintained a stock of chemically treated devices at room temperature (day 2 in workflow summary~\cite{wang_robust_2010}).
+The day of the experiment, a single device was plasma cleaned, bond to a glass coverslip, and injected with bacteria (see section above).
+
+The device was connected using 0.023" inner diameter polyethylene tubes to a waste and a sterile bottle containing 200 mL of growth medium, which was enough for several days of acquisition.
+A microfluidic pump (Elvesys) containing an output flow sensor module was plugged to the medium bottle, and was applied a pressure up to 2 bars.
+The output flow was set to 50 $\mu$L.min$^{-1}$.
+
+For the imaging, the device was placed on a motorized inverted microscope (Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^\circ$C.
+In this setup, fluorescence illumination is provided by a mercury lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma, \#49005 TR/DsRED ET).
+The exposure time to the light source is externally controlled by mechanical shutters (Uniblitz-VS35).
+Images are acquired with a 16-bit gray level CCDcamera cooled to -80$^\circ$C (Roper Scientific, Princeton Instruments PHOTOMAX 512) controlled by a custom-made software using Visual Basic and the Type libraries of the Winview software (Princeton Instruments).
+Every 5 minutes, autofocus was numerically performed by maximizing the contrast of a region of interest, and a serie of acquisitions was made.
+A total of 6 fields, each containing 15 wells was observed during the whole experiment.
 
 \subsection{Cell segmentation}
+\label{sec:cell_segmentation}
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+Raw data are in the form of 512x512-pixel 16-bit SPE images (proprietary format produced by the Princeton Instruments camera).
+They were converted in 16-bit Tif images using the \textit{SPE} plugin of ImageJ~\cite{goto_open_2005}.
+The rest of the analysis was performed using Python 3.5.2.
+In particular, we used OpenCV~3.1.0-dev, Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
+
+The offset between consecutive images of a given field was evaluated using cross-correlation in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see Scikit-Image documentation at \cite{skimage_cross-correlation}).
+All the images for a field were then aligned on the first acquisition by simple affine translation.
+
+The wells were isolated in individual images through a combination of manual pixel selection and automatic segmentation.
+In particular, we selected the entrance of the two wells at the border of the image, and used this information along with the regularity of the microfluidic device to compute a mask that allowed to crop each well in its own image of size 100x21 pixels.
+These sub-images were labeled $\left\{W0, W1, ..., W14\right\}$ depending on the position of the well on the original image, from left to right.
+
+In this study, we performed a preliminary analysis on each well that focused on the deepest cell.
+For each image, we manually selected two pixels representing the position of the poles of the cell of interest.
+The distance between the two selected pixels was directly used as the bacteria length $L$ in pixels.
+They were also used to compute a rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
+We then computed the RFU/pixel in this cell mask by summing each pixel and dividing by the mask size.
+
+The camera noise and background was evaluated by taking a picture with a closed shutter at the end of the experiment.
+Pixels in this picture were found to be Gaussian distributed with a mean of 1101.0 and a standard deviation of 10.788.
+Correction for camera background was thus applied by removing 1101 to the computed RFU/pixel.
+
+The background for autofluorescence of the bacteria and the medium was supposed negligible and was thus not corrected.
+Indeed, the same device was used to image bacteria in stationary phase that do not produce any fluorescent protein.
+They were entirely indistinguishable from the camera noise.
+Not that this does not ensure that the autofluorescence is also negligible when the bacteria are actively growing.
+Furthermore, its value could change with the medium of interest, and even be different at steady state and during growth transitions, which does not allow to estimate it independently.
+In the section~\ref{sec:chap3_discussion}, we discuss possible improvements of the experimental set-up that would allow to dynamically co-estimate the autofluorescence with the ribosome abundance in a single experiment.
 
 \subsection{Kalman smoothing}
+\label{sec:meth_kalman}
 
-We used this and that to analyze the microscopic images.
-We applied the following corrections.
+The data about the length and the RFU/pixel of each bacteria were analyzed using Python~3.5.2.
+In particular, we manipulated the data using Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
 
-\section{Results}
+Historical details about the Kalman smoothing procedure are reported in section~\ref{sec:res_kalman} of the main text.
+We used the Additive Unscented Kalman Filter implementation of the pykalman python module.
+This class is reported to be more stable and computationally efficient with non-linear problems featuring additive noises.
 
-\subsection{Model calibration}
+As described in section~\ref{sec:res_kalman}, the reconstruction of the growth rate was done on continuous portions of the length, \textit{i.e.} between two division events.
+The full problem, as reported in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity is:
+\begin{eqnarray*}
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
+\dot{\mu}(t) &=& v(t),\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray*}
+with the measurement model:
+\begin{equation*}
+L(t_k) = V_\lambda(t_k) + \epsilon_k.
+\end{equation*}
+The parameters used as prior for the reconstruction of $\mu$ are reported below.
+We used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
+Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
+At the beginning of the experiment, were no mother cell are available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
+All the cross-covariances are set to 0 because the system variables are independent by construction.
 
-We want to get $\alpha (t)$ during a transition. We show that $\alpha$ can be obtained if we measure so and so.
+For the reconstruction of gene activity, we only had to cope with the large gap in data acquisition between -720 and -150 min.
+Reconstruction was then done independently on the continuous sections before and after this gap.
+The full problem for the estimation of $u = \gamma \alpha / \beta$, as reported in Eq.~\ref{eq:full_u_prob}, is reproduced here for clarity:
+\begin{eqnarray*}
+\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
+\dot{u}(t) &=& v(t),\\
+\dot{v}(t) &=& w(t),\nonumber
+\end{eqnarray*}
+with the measurement model
+\begin{equation*}
+F(t_k) = r_\gamma (t_k) + \eta_k,\\
+\end{equation*}
+where $\hat{\mu}$ is the estimation of $\mu$ above.
+In both case, we used as prior an observation variance of 800.4 RFU\textsuperscript{2} for $F$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\gamma \alpha / \beta$) is fixed at $10^2$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
+The initial state means used are 600~RFU for $r_\gamma$ and 1000~RFU for $\gamma \alpha / \beta$.
+The variances associated with these means are purposely large and fixed to 10\textsuperscript{6}~RFU\textsuperscript{2} for $r_\gamma$ and $\gamma \alpha / \beta$.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2}, imposing a null second derivative for the reconstructed signal $\gamma \alpha / \beta$.
+Here again, all the cross-covariances are set to 0.
 
-\subsection{Bang-bang expression during transitions, as predicted}
+All the parameters cited in this section were chosen through trials and errors, and are as a consequence largely optimizable.
+Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
 
-One stunning figure that show the RFU / cell / pixels, and the predicted $\alpha$ for the glucose-acetate transition.
+\section{Supporting Information for Chapter~3}
 
-\subsection{Discrepancies with the model and model adaptation}
+(Plan not complete)
 
-The growth rate stalls after the addition of glucose, which was not originally predicted by the model.
-This is because so and so.
-This can be taken into account if we add another sector (M1 and M2).
-This raises a new interesting question : during the transition amino-acids drop, so the cell should produce ribosomes, but it also need to produce the new enzymes to actually be able to perform metabolism.
-We thus have competition between several layers of regulation, which can be added to the model by doing so and so.
+\subsection{Strain construction}
 
-\section{Discussion}
+Primers used for gibson assembly
 
-More research need to be done.
+\subsection{Strain validation using growth in microplate reader}
 
-Could be interesting to use transitions for syn bio purposes.
-Could be interesting to validate such results in different microorganisms, with different type of medium (fast env, slow env, etc)
+(signal-to-noise ratio, growth rate not hampered in microplate reader)
+Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
 
-\section{Supporting Information}
+\subsection{Microscopy analysis}
 
-\begin{itemize}
-\item Strain validation (signal-to-noise ratio, growth rate not hampered in microplate reader)
-\item Primers used for gibson assembly
-\item Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
-\end{itemize}
+Show the complete analysis for 5 cells.
+
+\subsubsection*{Category of cells}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_cells}
+\caption{Subtypes of cells. Caption to write. Cutting the x axis might be a good idea because a third of the x-axis of each graph are "missing data".}
+\label{fig:subcat_cells}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/subcat_median}
+\caption{Subtypes of cells, median. Caption to write.}
+\label{fig:subcat_median}
+\end{figure}

### 2016-12-27 Chapters/experiments.tex
Review of Hidde's comment of Chap 3

@@
 This is a reasonable and logical choice given that at steady state, the behavior of microorganisms is reproducible, which helps uncovering simple theories to apprehend their inherent complexity.
-But while this condition can be easily achieved in the laboratory, microorganisms naturally spend very little time in steay state.
+But while this condition can be easily achieved in the laboratory, microorganisms naturally spend very little time in steady state.
 This motivated the construction of a new theoretical framework in Chapter~\ref{chap:theory} to study growth laws during growth transitions.
@@
 
-Unfortunately, far more information is available on ribosome abundance at steady state than during growth transitions [refs].
-The main reason for this bias on steady-state conditions is that, from an experimental point of view, growth transitions are hard to control and may depend on the history of the culture~[citation ?
-Bremmer ?].
-However, as we discussed in Section~\ref{sec:discussion}, the few studies reported that have been reported seem to be consistent with our predictions, in the sense that they indicate that during growth transitions, the synthesis rate of ribomoses oscillates~\cite{gausing_regulation_1980,zengel_transcription_1986} and the ppGpp concentration manifests a rapid succession of increases and decreases~\cite{friesen_synthesis_1975,murray_control_2003}.
+Unfortunately, far more information is available on ribosome abundance at steady state than during growth transitions~\cite{scott_interdependence_2010,gausing_regulation_1980}.
+The main reason for this bias on steady-state conditions is that, from an experimental point of view, growth transitions are hard to control and may depend on the history of the culture~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
+However, as we discussed in Section~\ref{sec:discussion}, the few studies that have been reported seem to be consistent with our predictions, in the sense that they indicate that during growth transitions, the synthesis rate of ribomoses oscillates~\cite{gausing_regulation_1980,zengel_transcription_1986} and the ppGpp concentration manifests a rapid succession of increases and decreases~\cite{friesen_synthesis_1975,murray_control_2003}.
 But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
 What is needed for the verification of our predictions are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
-While the observation of an on-off strategy would show that simple models of the type introduced in Chapter~\ref{chap:theory} are instrumental in gaining a better understainding of microbial physiology, its refutation would also raise interesting questions.
+While the observation of an on-off strategy would show that simple models of the type introduced in Chapter~\ref{chap:theory} are instrumental in gaining a better understanding of microbial physiology, its refutation would also raise interesting questions.
 If maximization of growth rate and biomass accumulation have been retained by natural selection in \textit{E.
@@
 coli} strain in a microfluidic device, allowing the long-term imaging of individual cells in well-controlled conditions, notably involving a classic upshift experiment from growth in minimal medium with acetate to minimal medium with glucose.
-We then developed a Kalman smoothing method, in collaboration with Eugenio Cinquemain of Inria Grenoble - Rh\^{o}ne-Alpes, to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift, which allows $\alpha (\cdot)$ to be reconstructed.
-While the preliminary results presented here do not allow a decisive validation of the expected behavior, among other things due to the experimental difficulties that were encountered and image analysis challenges, we believe they are promising as a first step towards the better understanding of global resource allocation during growth transitions.
+We then developed a Kalman smoothing method, in collaboration with Eugenio Cinquemani of Inria Grenoble -- Rh\^{o}ne-Alpes, to reconstruct $\alpha (\cdot)$  from the estimates of the variations of the growth rate and the relative ribosomal synthesis rate.
+While the preliminary results presented here do not allow a decisive validation of the expected behavior, among other things due to the difficulties that were encountered during the experiments and the image analysis, we believe they are promising as a first step towards the better understanding of global resource allocation during growth transitions.
 
@@
 We can therefore identify $\alpha (t_k)$ for each time $t_k$ when the terms $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$ are known, or equivalently, the six individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$ they are composed of.
-Even when assuming that the parameters $k_R$, $K_R$, and $\beta$ are known, or at least easy to estimate independently, reconstructing the dynamics of $\alpha$ would require estimation of the (change in) concentration of the gene expression machinery ($r$, $\dot{r}$) and the precursor concentration $p$.
+Even when assuming that the parameters $k_R$, $K_R$, and $\beta$ are known, or at least easy to estimate independently, reconstructing the dynamics of $\alpha$ would require estimation of the (changes in) concentration of the gene expression machinery ($r$, $\dot{r}$) and the precursor concentration $p$.
 
 Are such measurements feasible in practice?
-The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques \cite{HwaMSBProteomics}.
+The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, or using radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques~\cite{hui_quantitative_2015}.
 However, only the use of fluorescent proteins complies with our need for real-time single-cell quantification.
 Obtaining an estimate of precursor abundance is even more challenging, since there is no clear proxy for the totality of precursors in the cell and, while absolute quantification of all internal metabolites has been achieved in \textit{E.
-coli}~\cite{bennett_absolute_2009}, this method is also not suitable for the estimation of the precursor concentration dynamics in individual cells.
+coli}~\cite{bennett_absolute_2009}, this method is also not suitable for a dynamical estimation in individual cells.
 
-The problem of quantifying $\alpha$ can be simplified by applying a transformation of the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}.
+The problem of quantifying $\alpha (\cdot)$ can be simplified by applying a transformation of the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}.
 Taking into account that by construction the growth rate $\mu(t)$ is given by
@@
 
-Given that fluorescent reporters of ribosomal proteins can provide information on both $r(\cdot)$ and $\dot{r}(\cdot)$ \cite{ValentinBioinformatics}, the above reformulation of the problem is promising for the purposes of our study.
+Given that fluorescent reporters of ribosomal proteins can provide information on both $r(\cdot)$ and $\dot{r}(\cdot)$ \cite{zulkower_robust_2015}, the above reformulation of the problem is promising for the purposes of our study.
 Moreover, since the predicted bang-bang profiles in gene activity are easily hidden at the population level if cells are not synchronous, which is generally the case, expression of the ribosomal genes need to be monitored at high sampling density on the single-cell level, which is also possible with fluorescent reporters.
-We therefore constructed an \textit{E. coli} strain in which the S2 ribosomal subunit, encoded by the gene \textit{rpsB}, has been tagged with a green fluorescent protein (GFP).
-The strain with the chromosomal \textit{rpsB-gfp} fusion, inspired by previous work \cite{refBakshi}, produces fluorescent ribosomes that are quantifiable \textit{in vivo} in single cells while not affecting the growth rate (see Material and Methods~\ref{sec:methods_strain}).
-The monitoring of single cells of this strain, by time-lapse fluorescence miscroscopy, thus makes it possible to estimate $r(\cdot)$ and $\dot{r}(\cdot)$ and reconstruct $\mu(t)$.
+Inspired by previous work~\cite{bakshi_superresolution_2012}, we therefore constructed an \textit{E.
+coli} strain in which the S2 ribosomal subunit, encoded by the gene \textit{rpsB}, has been tagged with a green fluorescent protein (GFP).
+The strain with the chromosomal \textit{rpsB-gfp} fusion produces fluorescent ribosomes that are quantifiable \textit{in vivo} in single cells while not affecting the growth rate (see Material and Methods~\ref{sec:methods_strain} and~\cite{bakshi_superresolution_2012}).
+Monitoring single cells of this strain, by time-lapse fluorescence microscopy, thus makes it possible to estimate $r(\cdot)$ and $\dot{r}(\cdot)$ and reconstruct $\mu(\cdot)$.
 
 The growth conditions also need to be carefully chosen for this study.
-In natural environments, for many bacteria one of the most frequently encountered growth transitions, or at least the transition on which selection is expected to operate, is the transition from stationary phase in a nutrient-deprived medium [see reference in recent PNAS paper by Balaban] to (exponential) growth after the renewed availability of nutrients.
-The transition from stationary to exponential phase is difficult to study though as many hard-to-control parameters play a role, including the duration of the nutrient stress, the composition of the medium before growth arrest, and the accumulation of waste products in the medium \cite{reviewBremerDennisBiochimie}.
-As a consequence, growth resumption only occurs after a lag phase of variable duration~[citation lag phase review to find], which is not included in the model.
+In natural environments, for many bacteria one of the most frequently encountered growth transitions, or at least the transition on which selection is expected to operate, is the transition from stationary phase in a nutrient-deprived medium~\cite{gefen_direct_2014} to (exponential) growth after the renewed availability of nutrients.
+The transition from stationary to exponential phase is difficult to study though as many hard-to-control parameters play a role, including the duration of the nutrient stress, the composition of the medium before growth arrest, and the accumulation of waste products in the medium~\cite{ehrenberg_mediumdependent_2012}.
+As a consequence, growth resumption only occurs after a lag phase of variable duration~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}, which is not included in the model.
 For this reason, we decided to focus on a transition from exponential growth on a poor carbon source (acetate) to exponential growth on a rich carbon source (glucose) in a well-defined minimal medium.
@@
 
-An experiment following the plan of Fig.~\ref{fig:experiment_schema} was carried out, over a period of 40 h, but encountered a number of difficulties.
-First, the motor displacing the microscope along the Z-axis intermittently stalled, thus deactivating the autofocus and the Z-compensation for the tilt of the [microfluidics???] device in the XY plane.
+An experiment following the plan of Fig.~\ref{fig:experiment_schema} was carried out, over a period of 40~h, but encountered a number of difficulties.
+First, the motor displacing the microscope along the Z-axis intermittently stalled, thus deactivating the autofocus and the Z-compensation for the tilt of the microfluidic device in the XY plane.
 As a consequence, until the experimenter manually intervened, many of the images acquired were out of focus.
 This is the reason why data points between -720 and -150 min are not exploitable, but fortunately the experiment was planned in such a way that when it really mattered, notably around the growth transition, someone was present to monitor the microscope.
-Second, at around 7~h after the nutrient upshift, the bacteria started to die for reasons that are unknown, the most plausible hypothesis being a phage contamination.
+Second, at around 7~h after the nutrient upshift, the bacteria started to die for reasons that are unknown, the most plausible hypothesis being a phage contamination (see Discussion~\ref{sec:chap3_discussion}).
 Data analysis was therefore limited to 550 min after the upshift, when roughly 3/4 of the cells were still growing.
-For these and other reasons, the experiment will have to be repeated (Section~\ref{sec:chap3_discussion}), but there was no time left for this in the framework of my PhD thesis.
+For these and other reasons (see Section~\ref{sec:chap3_discussion}), the experiment will have to be repeated, but there was no time left for this in the framework of my PhD thesis.
 
 The acquired data consisted of phase contrast and fluorescence images of 60 wells in total, in four different fields.
-While quite a few image analysis programs have been reported in the literature~[citations schitcells + others], and some of these specifically address the mother machine design of the microfluidics device~[molysso + Wang et al], they all presented limitations when applied to our data.
-For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidics device.
-[LACK OF ALIGNMENT IS NOT VERY STRONG ARGUMENT, UNLESS IT CANNOT BE REPAIRED]
+While quite a few image analysis programs have been reported in the literature~\cite{klein_tlm-tracker:_2012,wang_image_2010,young_measuring_2012,paintdakhi_oufti:_2016}, and some of these specifically address the mother machine design of the microfluidic device~\cite{wang_robust_2010,sachs_image-based_2016}, they all presented limitations when applied to our data.
+For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidic device.
 Moreover, the fluorescence density was concentrated in hot spots (Fig.~\ref{fig:data_acquisition}) and its intensity varied during the experiment. 
@@
 The fluorescence intensity of the cell, expressed in units RFU/pixel/cell, was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
-The cell length is the distance in pixels between the two poles (red line on the raw and segmented images).
+The cell length is the distance in pixels between the two poles (red line).
 The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labelled "X3Y2, W2".
@@
 \end{eqnarray}
-where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in units RFU/pixel measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$) the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
-($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) measurement noise.
-[YOU MAY NEED TO SPECIFY THE ASSUMPTIONS ON THE NOISE DISTRIBUTION (0 MEAN, NORMAL?)]
+where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in units RFU/pixel measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$ the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
+($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) uncorrelated sequence of measurement noise assumed normally distributed with mean 0.
 
@@
 Since each division in the time series generates a new boundary condition, smoothing interpolation of our data is expected to be little robust.
-For this reason, the growth rate of single cells is usually estimated by fitting an exponential function to the volume data between each cell~[citation Wang et al et Izard et al].
+For this reason, the growth rate of single cells is usually estimated by fitting an exponential function to the volume data between each cell~\cite{wang_robust_2010,izard_synthetic_2015}.
 While this is suitable when bacteria are at steady state, it is not applicable during a growth transition, where we expect the growth rate to vary between successive divisions.
@@
 \end{eqnarray}
-with the following measurement model:
+with initial conditions $r(0) = r_0$, $V(0) = V_0$, and the following measurement model:
 \begin{eqnarray}
@@
 \dot{r}_\gamma(t) &=& \mu (t) \cdot\frac{\gamma \cdot \alpha(t)}{\beta} - \mu (t) \cdot r_\gamma (t),\label{eq:r_prob}\\
-\dot{V}_\lambda}t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
+\dot{V}_\lambda (t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
 \end{eqnarray}
-with the new measurement model:
+with initial conditions $r_\gamma (0) = \gamma r_0$, $V_\lambda (0) = \lambda V_0$, and the new measurement model:
 \begin{eqnarray}
@@
 \end{eqnarray}
-The Kalman filtering problem thus becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ from measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+The Kalman filtering problem thus becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ from measurements $\left\{ F(t_0), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), ..., L(t_{N-1}) \right\}$.
 
@@
 Note that Kalman smoothing is a procedure that returns the expected mean and covariance of the state of a dynamical system, given the  measurements.
-Therefore, reconstructing $\mu$ requires it to be explicitly included as a state variable of the dynamical system [I GUESS THIS IS WHAT YOU MEAN?].
+Therefore, reconstructing $\mu$ requires it to be explicitly included as a state variable of the dynamical system.
 This provides constraints on the variation of $\mu$ that have the effect of a regularization.
@@
 \]
-where $v$ is an intermediate variable and $w(t) \sim \mathcal{N}(0, \theta)$. [WHAT IS $\theta$?]
+where $v$ is an intermediate variable and $w(t)$ is assumed normally distributed with mean 0 and standard deviation $\theta$: $w(t) \sim \mathcal{N}(0, \theta)$.
 The resulting system then becomes
@@
 \begin{equation}
-L(t_k) = V_\lambda(t_k) + \epsilon_k.
+L(t_k) = V_\lambda(t_k) + \epsilon_k,
 \end{equation}
+and the initial conditions
+\[
+V_\lambda (0) = V_{\lambda,0}, \;\;\;\; \mu (0) = \mu_0 , \;\;\;\; v(0) = v_0,
+\]
+that may themselves be Gaussian random variables with some given statistics.
 The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable.
-It is also equivalent to other methods like Tikhonov regularization~[citation to take from Val pub], which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
+It is also equivalent to other methods like Tikhonov regularization~\cite{tikhonov1977solutions,tikhonov1963solution,de_nicolao_nonparametric_1997}, which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
 
@@
 \begin{itemize}
-\item the mean of the initial state for $V_\lambda$, $\mu$ and $v$,
-\item the covariance of the initial state for $V_\lambda$, $\mu$ and $v$,
-\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$,
-\item the observation variance for $L$.
+\item the mean of the initial state $(V_{\lambda,0}, \mu_0, v_0)$,
+\item the covariance of the initial state $(V_{\lambda,0}, \mu_0, v_0)$,
+\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$ (see below),
+\item the variance of the observation noise $\epsilon_k$.
 \end{itemize}
-[IF YOU DESCRIBE THE PROCEDURE IN SO MUCH DETAIL, THEN YOU NEED TO EXPLICITLY INTRODUCE THE MEANS/(CO)VARIANCES.
-THESE PARAGRAPHS ARE ONLY COMPREHENSIBLE FOR THE SPECIALISTS, BUT IN THAT CASE NEED TO BE MATHEMATICALLY PRECISE.]
-
-The means and covariances at the initial state represent the knowledge we have of the initial value of the variables. [YOU NEED TO EXTEND THE MODEL WITH THE INITIAL VALUES]
-For instance, if we know exactly the initial value of $V_\lambda$, $\mu$ and $v$, we can fix the means and use very small covariances.
-In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because they are not the result of a stochastic process.
-However, the transition variance $\theta^2$ of the derivative of $v$ is crucial: it represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
-The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu$.
-Finally, the observation variance for $L$ is the expected measurement noise.
-Together,  these parameters represent the priors that we use to reconstruct the growth rate $\mu$.
+The mean and covariance of the initial state $(V_{\lambda,0}, \mu_0, v_0)$ represent the knowledge we have of the initial values of the variables.
+For instance, if we know exactly the initial value of the growth rate $\mu$ (\textit{via} independent measurements or literature data), we can use it as mean for $\mu_0$ along with a very small variance.
+This will constraint the signal reconstruction by penalizing deviations from this value at $t=0$.
+On the contrary, if we are very uncertain of the value for $V_{\lambda,0}$ (because cells are not synchronized, and can be in any state when the data acquisition starts), we can use a large variance for this initial condition.
+In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because they are not the result of a stochastic process (the equations that define their derivatives are fully deterministic).
+However, by definition $v$ is the result of a stochastic process of mean 0 and transition variance $\theta^2$.
+This value is crucial and represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
+The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu (\cdot)$.
+Finally, the variance of the observation noise $\epsilon_k$ is simply the expected measurement noise.
+Together, these parameters represents the \textit{prior} information we have on the initial conditions, the smoothness of the reconstructed signal, and the precision of our measurements, in order to reconstruct the growth rate $\mu (\cdot)$.
 As we will see below, we widely use these properties to overcome the difficulties introduced by the discontinuities following cell divisions.
@@
 The estimation problem can therefore not be solved for the experiment as a whole, but only for the time-intervals between two successive division events, when the cell length is expected to change continuously.
-We localize the division events by identifying the time points at which the empirical derivative of the length is below an arbitrary threshold, and correspondingly slice the total duration of the experiment into time-intervals with continuous change of cell length.
+We localize the division events by identifying the time points at which the empirical derivative of the length is below an arbitrary threshold, and correspondingly slice the total duration of the experiment into time-intervals with continuous changes of cell length.
 The simplest way to proceed from there on would be to treat each portion independently.
@@
 In other word, while not necessary continuous, the growth rate of new-born cells will depend strongly on the growth rate of their mother just before division.
-This can be easily taken into account in the Kalman smoothing procedure by exploiting the priors about the initial mean and covariance of the system variables.
+This can be easily taken into account in the Kalman smoothing procedure by exploiting the prior information about the initial mean and covariance of the system variables.
 In particular, we set the mean of the initial growth rate of a daughter cell equal to the final estimated growth rate of the mother cell, while defining a reasonable variance of the growth rate to allow for uncertainty.
-Right after the upshift, when the above assumption may not hold, we set the mean of the initial growth rate to the estimated growth rate right before the upshift, but define a large variance.
 The results of this Kalman smoothing procedure, along with the exact parameters used, are reported in Fig.~\ref{fig:growth_rate_estimation} and Material and Methods~\ref{sec:meth_kalman}.
@@
 \caption{
-\textbf{Growth-rate estimation using Kalman smoothing based on measurement of the length of bacteria growing in a microfluidics device.}
+\textbf{Growth-rate estimation using Kalman smoothing based on measurement of the length of bacteria growing in a microfluidic device.}
 Gray lines represent the estimation of the time-varying length (upper plot) and growth rate (lower plot) of 68 cells by the unscented Kalman smoothing procedure.
@@
 We note $u(t) = \gamma \alpha (t) / \beta$ the resource allocation input that we wish to reconstruct.
-The full model for the estimation of $u$ is given by
+With $\hat{\mu}$ the estimation of $\mu$ obtained in the previous section, the full model for the reconstruction of $u (\cdot)$ is given by
 \begin{eqnarray}
@@
 with the measurement model
-\begin{equation*}
+\begin{equation}
 F(t_k) = r_\gamma (t_k) + \eta_k,\\
-\end{equation*}
-where $\hat{\mu}$ is the estimation of $\mu$ obtained in the previous section.
+\end{equation}
+and the initial conditions
+\[
+r_\gamma (0) = r_{\gamma,0}, \;\;\;\; u (0) = u_0 , \;\;\;\; v(0) = v_0,
+\]
+that may themselves be Gaussian random variables with some given statistics.
 Contrary to the model used for the estimation of the growth rate, the system of Eq.~\ref{eq:full_u_prob} is linear.
@@
 In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of Chapter~\ref{chap:theory}.
-After estimating the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (Supporting Information XX), choosing parameters that allow the model to reproduce the observed growth rates and fluorescence densities, we simulated an upshift from acetate to glucose.
-We then tried to estimate $\gamma \alpha(\cdot) / \beta$ from these synthetic data, selecting the \textit{a priori} values of the parameters in the Kalman smoothing procedure by trial and error.
-[THE USE OF PARAMETERS IS AMBIGUOUS HERE: YOU PROBABLY DO NOT REFER TO MODEL PARAMETERS AS IN THE PREVIOUS PHRASE, BUT TO THE PARAMETERS OF THE KALMAN PROCEDURE]
+We estimated the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (Supporting Information XX) and chose parameters that allow the model to reproduce the observed growth rates and fluorescence densities.
+We simulated an upshift from acetate to glucose and tried to estimate $\gamma \alpha(\cdot) / \beta$ from these synthetic data, selecting the \textit{prior} in the Kalman smoothing procedure by trial and error.
 This approach may not have been optimal for our purpose and possible improvements are discussed in Section~\ref{sec:chap3_discussion}.
-The results along with the parameter values [BOTH TYPES] are shown in Fig.~\ref{fig:synthetic_upshift}.
+The results along with the \textit{prior} used are reported in Fig.~\ref{fig:synthetic_upshift}.
 As expected, like any smoothing method, the algorithm has difficulty in reconstructing stiff variations in the state variables of Eq.~\ref{eq:full_u_prob}.
 The switching profile of the resource allocation input is relatively well captured though, which indicates that the algorithm is in principle capable of reconstructing an on-off strategy (further improvements are discussed in Section~\ref{sec:chap3_discussion}).
-In what follows, we use the same values for the parameters of the Kalman smoothing procedure for the normal cells identified in the previous section, to test the occurrence of an on-off switching profile in our data.
+In what follows, we use the same \textit{prior} of the Kalman smoothing procedure for the normal cells identified in the previous section, to test the occurrence of an on-off switching profile in our data.
 
@@
 The synthetic data were generated by simulating the model presented in Eqs~\ref{eq:pdef}-\ref{eq:rdef} with the on-off regulatory strategy (Eq.~\ref{eq:stratswitch} and Fig.~\ref{fig:onoffresults}).
-The model parameters used for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L g\textsuperscript{-1}, $K_R = 1$~g L\textsuperscript{-1}.
-The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~L g\textsuperscript{-1} RFU\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
+The model parameters used for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $K_R = 1$~g~L\textsuperscript{-1}.
+The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~L~g\textsuperscript{-1}~RFU\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
 The noise level was estimated from the data (Supporting Information XX) and added to $F$.
@@
 
-Nevertheless, when not focusing on the individual cells but looking at the statistics of the entire data set, some interesting patterns emerge. 
-Fig.~\ref{fig:gene_activity_median} shows the median of the time-varying growth rate and resource allocation profile as well as the 25\%-75\% interquartile range. 
-The use of these statistics gives a more robust view on the population level of the response of the cells to a nutrient upshift. 
-When changing the carbon source from acetate to glucose, the growth rate increases to a value of around 0.011 min$^{-1}$, consistent with growth rates reported in the literature for the \textit{E.
-coli} strain used here \cite{Volkmer/Heinemann,Izard}, before decreasing when the first cells start to die (Fig.~\ref{fig:gene_activity_median}\textit{C}).
-In addition, the data show one period of an oscillation in the first 3~h after the upshift, conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}\textit{B}).
-The heatmap in Fig~\ref{fig:gene_activity_heatmap} reveals that almost all of the normal cells show this oscillatory feature.
-Moreover, the first peak is seen to be even more pronounced on the level of the individual cells, reflecting the fact that in the different cells it occurs at different times after the upshift and is therefore dampened out at the population level.
-While the results of the microfluidics experiment presented here do not allow to confirm the occurrence of an on-off strategy for resource allocation after a nutrient upshift, the preliminary data are encouraging and prompt further investigation.
-
 \begin{figure}[p]
@@
 The solid red curves highlight the result for one particular cell, , located at the bottom of the well labelled "X3Y2, W2", while black crosses in the top graph are the data points for this cell.
-[WHY THIS CELL AND NOT ONE OF THE CELLS YOU SENT US A WHILE AGO?
-THIS CELL HAS A PARTICULARLY IRREGULAR PROFILE...]
 The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
@@
 \end{figure}
+
+Nevertheless, when not focusing on the individual cells but looking at the statistics of the entire data set, some interesting patterns emerge. 
+Fig.~\ref{fig:gene_activity_median} shows the median of the time-varying growth rate and resource allocation profile as well as the 25\%-75\% interquartile range. 
+The use of these statistics gives a more robust view on the population level of the response of the cells to a nutrient upshift. 
+When changing the carbon source from acetate to glucose, the growth rate increases to a value of around 0.011 min$^{-1}$, consistent with growth rates reported in the literature for the \textit{E.
+coli} strain used here~\cite{volkmer_condition-dependent_2011,izard_synthetic_2015}, before decreasing when the first cells start to die (Fig.~\ref{fig:gene_activity_median}\textit{C}).
+In addition, the data show one period of an oscillation in the first 3~h after the upshift, conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}\textit{B}).
+The heatmap in Fig.~\ref{fig:gene_activity_heatmap} reveals that almost all of the normal cells show this oscillatory feature.
+Moreover, the first peak is seen to be even more pronounced on the level of the individual cells, reflecting the fact that in the different cells it occurs at different times after the upshift and is therefore dampened out at the population level.
+While the results of the microfluidic experiment presented here do not allow to confirm the occurrence of an on-off strategy for resource allocation after a nutrient upshift, the preliminary data are encouraging and prompt further investigation.
 
@@
 
-\begin{figure}[p]
+\begin{figure}[tb]
 \centering
-\includegraphics[scale=1]{./Fig/}
+%\includegraphics[scale=1]{./Fig/}
 \caption{
@@
 The individual $\gamma \alpha (\cdot) / \beta$  curves have been normalized with respect to their maximum in the interval $[0, 400]$ and then ordered with respect to their maximum in the interval $[0, 100]$.
-
 }
@@
 coli} cells growing in continuous culture by means of time-lapse fluorescence microscopy.
-Here, the mother machine was used it to establish well-controlled growth transitions, \textit{i.e.} from steady-state growth in minimal medium with acetate to steady-state growth in minimal medium with glucose.
+Here, the mother machine was used to establish well-controlled growth transitions, \textit{i.e.} from steady-state growth in minimal medium with acetate to steady-state growth in minimal medium with glucose.
 By changing the medium source of the mother machine, we were able to observe how \textit{E. coli} cells adapt their growth rate and the expression of ribosomes.
@@
 coli} during a nutrient upshift, with the aim to observe the bang-bang profile predicted in Chapter~\ref{chap:theory}.
-We developed a Kalman smoothing algorithm adapted to our question and showed how it can be implemented to reconstruct the internal state of a cell from time-lapse microscopy images.
+We developed a Kalman smoothing procedure adapted to our question and showed how it can be implemented to reconstruct the internal state of a cell from time-lapse microscopy images.
 Similar to hidden Markov chains, Kalman filtering and smoothing are powerful techniques for the reconstruction of unobserved signals from noisy measurements, and have had numerous applications in a variety of domains, though less in biology than in other fields.
 We showed in this chapter that the Bayesian framework of Kalman smoothing can be exploited for growth rate reconstruction from a probabilistic prior defining the relation between mother and daughter cells.
-We also developed another varian of Kalman smoothing for the reconstruction of the resource allocation variable $\alpha (\cdot)$ that is at the heart of the models of Chapter~\ref{chap:theory}.
+We also developed another variant of Kalman smoothing for the reconstruction of the resource allocation variable $\alpha (\cdot)$ that is at the heart of the models of Chapter~\ref{chap:theory}.
 Our work on synthetic data showed that, despite the challenge posed by the reconstruction of a discontinuous signal, the algorithm was capable of recovering an oscillatory pattern close to the expected on-off profile.
-Unfortunately, its application to the real data did not allow us an unambiguous conclusion to be drawn: while oscillations do occur, and a first peak between 0 and 120 min after the upshift is visiable in almost all of the cells, it turned out to be far from straightforward to distinguish a real oscillatory pattern from artifacts of the data analysis and signal reconstruction methods.
+Unfortunately, its application to the real data did not allow an unambiguous conclusion to be drawn: while oscillations do occur, and a first peak between 0 and 120 min after the upshift is visible in almost all of the cells, it turned out to be far from straightforward to distinguish a real oscillatory pattern from artifacts of the data analysis and signal reconstruction methods.
 Below we propose several improvements that will be addressed in future work.
@@
 Several causes are to blame.
-First, the random stall of the motorization along the Z-axis of the microscope, which do not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
+First, the random stall of the motorization along the Z-axis of the microscope, which does not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
 Second, the massive death of bacteria starting after the transition on glucose.
@@
 
-[I DID NOT TOUCH THE PARARAPH THAT FOLLOWS]
 The death of bacteria is particularly worrying.
@@
 We segmented the images using the most powerful and ubiquitous segmentation algorithm available: the human brain.
-In other word, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a rectangle around the cell of interest.
+In other words, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a rectangle around the cell of interest.
 While this is far from satisfying, it may turn out to be difficult to improve upon this, as it seems that the first rule of image analysis is that every application is more or less unique.
@@
 While some parameters can be tweaked, some assumptions are always hard-coded in the algorithm and reflect the philosophy used to address the identification problem.
-Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~[fluobactracker.inrialpes.fr] that focuses on adapting their software to our set-up in a near future.
-It should provide a robust, automated way to identify and track the cells, enhancing the reproducibility of the analysis and increasing the number of cells available fro computing population statistics.
+Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~\cite{fluobactracker} that focuses on adapting their software to our set-up in a near future.
+It should provide a robust, automated way to identify and track the cells, enhancing the reproducibility of the analysis and increasing the number of cells available for computing population statistics.
 
@@
 
-[I WOULD SKIP THIS FINAL PARAGRAPH OR REDUCE IT TO ONE PHRASE AND ADD IT TO THE DISCUSSION IN THE PARARAPH STARTING WITH "The Kalman smoothing algorithm..."]
-Something is also displeasing in the current implement of the Kalman smoothing method.
-As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
-This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
-We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
-There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
+%[I WOULD SKIP THIS FINAL PARAGRAPH OR REDUCE IT TO ONE PHRASE AND ADD IT TO THE DISCUSSION IN THE PARARAPH STARTING WITH "The Kalman smoothing algorithm..."]
+%Something is also displeasing in the current implement of the Kalman smoothing method.
+%As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
+%This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
+%We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
+%There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
 

### 2016-12-28 Chapters/experiments.tex
Change titlepage (new format) and write Suppinfo S5 about strains

@@
 \centering
-\includegraphics[scale=1]{./Fig/experiment_schema.eps}
+\includegraphics[scale=1]{./Fig/experiment_schema}
 \caption{
@@
 
-(Plan not complete)
+\subsection{S5 Text -- Strain construction}
+\manuallabel{S5_Text}{S5~Text}
 
-\subsection{Strain construction}
+\subsubsection{Exhaustive list of the primers used...}
 
-Primers used for gibson assembly
+\paragraph{...for the \textit{gfpmut2} amplification.}
 
-\subsection{Strain validation using growth in microplate reader}
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/gfpmut2_L.png} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGAGTAAAGGAGAAGAACTTTTCACTG}
+(Length: 93)
+\item[] \includegraphics[scale=1]{Fig/gfpmut2_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTATAGTTCATCCATGCC} (Length: 44)
+\end{itemize}
+
+\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
+The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+
+\paragraph{...for the \textit{mCherry} amplification}
+
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/mCherry_L.png} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGACTAGCAAAAGATCCAAGGG}
+(Length: 88)
+\item[] \includegraphics[scale=1]{Fig/mCherry_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTACAGCTCATCCATG} (Length: 42)
+\end{itemize}
+
+\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
+The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
+
+\paragraph{...for the cassette amplification}
+
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAACTATACAAATAATTATATTCCCCAGAACATCAGG}
+(Length: 42, used to assemble with gfp)
+\item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAGCTGTACAAATAATTATATTCCCCAGAACATCAG}
+(Length: 41, used to assemble with mCherry)
+\item[] \includegraphics[scale=1]{Fig/cassette_R} \seqsplit{GAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTAGAAGAACTCGTCAAGAAGG}
+(Length: 72, used for both)
+\end{itemize}
+
+\textbf{Rem.} The 20 first bases of the first forward primer are supposed to anneal with the end of the \textit{gfpmut2} gene.
+The 20 first bases of the second forward primer are supposed to anneal with the end of the \textit{mCherry} gene.
+The 50 last bases of the reverse primer are supposed to anneal with the region after \textit{rpsB} on the \textit{E.~coli} BW25113 chromosome.
+
+\paragraph{...for the whole insert amplification}
+
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/overlap_L} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGG}
+(Length: 24)
+\item[] \includegraphics[scale=1]{Fig/overlap_R} \seqsplit{GAGCTTGCCGCCTTTCTGCA} (Length: 20)
+\end{itemize}
+
+\paragraph{...for the cassette elimination}
+
+\begin{itemize}
+\item[\textbf{\textit{rpsB-gfpmut2}}] \seqsplit{TTGTAACAGCTGCTGGGATTACACATGGCATGGATGAACTATACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
+(Length: 100)
+\item[\textbf{\textit{rpsB-mCherry}}] \seqsplit{GCGCGGAGGGTCGTCATTCTACCGGTGGCATGGATGAGCTGTACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
+(Length: 100)
+\end{itemize}
+
+\paragraph{...for the sequencing of the final strain}
+
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/sequencing_L} \seqsplit{CGTCTGAAAGACCTGGAAAC}
+(Length: 20)
+\item[] \includegraphics[scale=1]{Fig/sequencing_R} \seqsplit{AAACGTGTACTACCTGGTCTATAAGG} (Length: 26)
+\end{itemize}
+
+\subsubsection{Full sequence of the inserts}
+
+The following pages contain the full sequences of the \textit{gfpmut2-cassette} and \textit{mCherry-cassette} constructions, inserted on the \textit{E.~coli} BW25113 chromosome at the region after the \textit{rpsB} gene.
+The primers annealing positions are indicated using the same color code as in the description of the primers above.
+The primers reported in the original publication of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012} are indicated for information.
+
+\includepdf[pages={1-7}]{./Fig/gfp-mut2-cassette-after-rpsB-gene-sequence.pdf}
+\includepdf[pages={1-6}]{./Fig/mcherry-cassette-after-rpsB-gene-sequence.pdf}
+
+\subsubsection{Sequencing of the final strains}
+
+The strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}.
+Regions at the end of the \textit{rpsB} gene were amplified, and sent for sequencing.
+Alignments of the results showed they exhibit the expected sequences.
+The raw results of the sequencing are reproduced below.
+
+\paragraph{Final sequences of the \textit{rpsB-gfpmut2} strain (region after \textit{rpsB})}
+
+\begin{small}
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/sequencing_L} \seqsplit{AGAAAGAAGCGCTGTATGCGCACTCGTGAGCTGGAGAAACTGGAAAACAGCCTGGGCGGTATCAAAGACATGGGCGGTCTGCCGGACGCTCTGTTTGTAATCGATGCTGACCACGAACACATTGCTATCAAAGAAGCAAACAACCTGGGTATTCCGGTATTTGCTATCGTTGATACCAACTCTGATCCGGACGGTGTTGACTTCGTTATCCCGGGTAACGACGACGCAATCCGTGCTGTGACCCTGTACCTGGGCGCTGTTGCTGCAACCGTACGTGAAGGCCGTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGCGTAAAGGAGAAGAACTTTTCACTGGAGTTGTTCCAATTCTTGTTGAATTAGATGGTGATGTTAATGGGCACAAATTTTCTGTCAGTGGAGAGGGTGAAGGTGATGCAACATA}
+\item[] \includegraphics[scale=1]{Fig/sequencing_R} \seqsplit{CGCCGCAGATGCGTTATCTTCGCTCGCTCATCCCGGTCACTTACTGATGTAAGCTCCCGGGAATTCTCGAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTATTTGTATAGTTCATCCATGCCATGTGTAATCCCAGCAGCTGTTACAAACTCAAGAAGGACCATGTGGTCTCTCTTTTCGTTGGGATCTTTCGAAAGGGCAGATTGTGTGGACAGGTAATGGTTGTCTGGTAAAAGGACAGGGCCATCGCCAATTGGAGTATTTTGTTGATAATGGTCTGCTAGTTGAACGCTTCCATCTTCAATGTTGTGTCTAATTTTGAAGTTAACTTTGATTCCATTCTTTTGTTTGTCTGCCATGATGTATACATTGTGTGAGTTATAGTTGTATTCCAATTTGTGTCCAAGAATGTTTCCATCTTCTTTAAAATCAATACCTTTTAACTCGATTCTATTAACAAGGGTATCACCTTCAAACTTGACTTCAGCACGTGTCTTGTAGTTCCCGTCATCTTTGAAAAATATAGTTCTTTCCTGTACATAAACCTTCGGGCATGGCACTCTTGAAAAAGTCATGCTGTTTCATATGATCTGGGTATCTCGCAAAGCATTGAAGACCATACGCGAAAAGTAGTGACAAGTGTTGGCCATGGAACAGGTAGTTTTCCAGTAGTGCAAATAAATTTAAGGGTAAAGTTTTCCGTATGTTGCATCACCTTCACCCTCTCCACTGACAGAAAAATTTGTGCCCATTTAACATCACCATCTAATTCAACAAGAATTGGAAACAACTCCAGTGAAAGT}
+\end{itemize}
+\end{small}
+
+\newpage
+\paragraph{Final sequences of the \textit{rpsB-mCherry} strain (region after \textit{rpsB})}
+
+\begin{small}
+\begin{itemize}
+\item[] \includegraphics[scale=1]{Fig/sequencing_L}
+\seqsplit{TTTCGACAGCTGACCAAGAAGAAGCGCTGATGCGCACTCGTGAGCTGGAGAAACTGGAAA
+ACAGCCTGGGCGGTATCAAAGACATGGGCGGTCTGCCGGACGCTCTGTTTGTAATCGATG
+CTGACCACGAACACATTGCTATCAAAGAAGCAAACAACCTGGGTATTCCGGTATTTGCTA
+TCGTTGATACCAACTCTGATCCGGACGGTGTTGACTTCGTTATCCCGGGTAACGACGACG
+CAATCCGTGCTGTGACCCTGTACCTGGGCGCTGTTGCTGCAACCGTACGTGAAGGCCGTT
+CTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGAC
+AGGAGACTAGCAAAAGATCCAAGGGCGAGGAGGATAACATGGCTATCATTAAAGAGTTCA
+TGCGCTTCAAAGTTCACATGGAGGGTTCTGTTAACGGTCACGAGTTCGAGATCGAAGGCG
+AAGGCGAGGGCCGTCCGTATGAAGGCACCCAGACCGCCAAACTGAAAGTGACTAAAGGCG
+GCCCGCTGCCTTTTGCGTGGGACATCCTGAGCCCGCAATTTATGTACGGTTCTAAAGCGT
+ATGTTAAACACCCAGCGGATATCCCGGACTATCTGAAGCTGTCTTTTCCGGAAGGTTTCA
+AGTGGGAACGCGTAATGAATTTTGAAGATGGTGGTGTCGTGACCGTCACTCAGGACTCCT
+CCCTGCAAGATGGCGAGTTCATCTATAAAGTTAAACTGCGTGGTACTAATTTTCCATCTG
+ATGGCCCGGTGATGCAGAAAAAGACGATGGGTTGGGAGGCGTCTAGCGAACGCATGTATC
+CGGAAGATGGTGCGCTGAAAGGCGAAATTAAACAGCGCCTGAAACTGAAAGATGGCGG}
+\item[] \includegraphics[scale=1]{Fig/sequencing_R}
+\seqsplit{TTCGCGCCGCAGATGCGTTATCTTCGCTCGCTCATCCCGGTCACTTACTGATGTAAGCTC
+CCGGGAATTCTCGAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAA
+GCTTATTTGTACAGCTCATCCATGCCACCGGTAGAATGACGACCCTCCGCGCGCTCATAT
+TGCTCTACGATCGTATAATCTTCATTATGAGAGGTGATGTCCAGTTTAATATTCACATTG
+TACGCGCCAGGCAGCTGCACAGGTTTCTTGGCTTTGTACGTGGTTTTCACTTCAGCGTCA
+TAATGGCCGCCATCTTTCAGTTTCAGGCGCTGTTTAATTTCGCCTTTCAGCGCACCATCT
+TCCGGATACATGCGTTCGCTAGACGCCTCCCAACCCATCGTCTTTTTCTGCATCACCGGG
+CCATCAGATGGAAAATTAGTACCACGCAGTTTAACTTTATAGATGAACTCGCCATCTTGC
+AGGGAGGAGTCCTGAGTGACGGTCACGACACCACCATCTTCAAAATTCATTACGCGTTCC
+CACTTGAAACCTTCCGGAAAAGACAGCTTCAGATAGTCCGGGATATCCGCTGGGTGTTTA
+ACATACGCTTTAGAACCGTACATAAATTGCGGGCTCAGGATGTCCCACGCAAAAGGCAGC
+GGGCCGCCTTTAGTCACTTTCAGTTTGGCGGTCTGGGTGCCTTCATACGGACGGCCCTCG
+CCTTCGCCTTCGATCTCGAACTCGTGACCGTTAACAGAACCCTCCATGTGAACTTTGAAG
+CGCATGAACTCTTTAATGATAGCCATGTTATCCTCCTCGCCCTTGGAT} 
+\end{itemize}
+\end{small}
+
+\subsection{S6 Text -- Strain validation}
+\manuallabel{S6_Text}{S6~Text}
 
@@
 
-\subsection{Microscopy analysis}
+\subsection{S7 Text -- Microscopy analysis}
+\manuallabel{S7_Text}{S7~Text}
 
-Show the complete analysis for 5 cells.
-
-\subsubsection*{Category of cells}
+\subsubsection{Category of cells}
 
@@
 \end{figure}
+
+\subsubsection{Noise estimation}
+
+Blabla
+
+\subsubsection{Complete analysis for 5 cells}
+
+blabla

### 2016-12-28 titlepage.tex
Change titlepage (new format) and write Suppinfo S5 about strains

@@
 
-%\Sethpageshift{-8mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture horizontalement (valeur par défaut est -5.5mm)
-%\Setvpageshift{0mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture verticalement (valeur par défaut est -15.5mm)
+%\Sethpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture horizontalement (valeur par défaut est -5.5mm)
+%\Setvpageshift{???mm}   %%optionnel : à décommenter si besoin pour ajout d'espace afin de center la couvérture verticalement (valeur par défaut est -15.5mm)
 
@@
 \Specialite{Biologie cellulaire}
-\Arrete{7 Ao\^ut 2006}
+\Arrete{25 mai 2016} % ATTENTION, A VERIFIER CHAQUE ANNEE
 \Auteur{Nils GIORDANO}
@@
 \Soustitre{Theoretical and experimental study of resource allocation in \textit{Escherichia coli}}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
-\Depot{}
+\Depot{ (fin Mars)} %% DATE DE LA SOUTENANCE
 
@@
 
-\UGTRapporteur{M, Matthew SCOTT}{Associate Professor, Department of Applied Mathematics, University of Waterloo (CA)}      %% 1er rapporteur
-\UGTRapporteur{M, Guy-Bart STAN}{Reader, Department of Bioengineering, Imperial College London (U.K.)}      %% second rapporteur
+\UGTRapporteur{M, Matthew SCOTT}{Assoc. Prof.}{University of Waterloo}      %% 1er rapporteur
+\UGTRapporteur{M, Guy-Bart STAN}{Reader}{Imperial College London}      %% second rapporteur
 
-\UGTExaminateur{M, Olivier BERNARD}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}     %% 2ème examinateur
-\UGTExaminateur{M, Matthieu JULES}{Chargé de recherche, Inra, Grignon}     %% 1er examinateur
-\UGTExaminatrice{Mme, Irina MIHALCESCU}{Professeur, Université Grenoble Alpes, LIPhy}    %% 3ème examinateur
+\UGTExaminateur{M, Olivier BERNARD}{DR}{Inria, Nice}     %% 2ème examinateur
+\UGTExaminateur{M, Matthieu JULES}{CR}{Inra, Grignon}     %% 1er examinateur
+\UGTExaminatrice{Mme, Irina MIHALCESCU}{Prof.}{Univ. Grenoble Alpes}    %% 3ème examinateur
 
-\noindent\hfil\rule{4cm}{0.1pt}\hfil
+%\noindent\hfil\rule{4cm}{0.1pt}\hfil
+%
+%\vspace{0.3cm}
 
-\vspace{0.3cm}
-
-\UGTDirecteur{M, Johannes GEISELMANN}{Professeur, Université Grenoble Alpes, LIPhy}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde DE JONG}{Directeur de recherche, Inria, Grenoble -- Rhône-Alpes}     %% Co-Directeur de thèse s'il y en a
-\UGTInvite{M, Jean-Luc GOUZÉ}{Directeur de recherche, Inria, Nice -- Sophia-Antipolis}
+\UGTDirecteur{M, Johannes GEISELMANN}{Prof.}{Univ. Grenoble Alpes}       %% Directeur de thèse
+\UGTCoDirecteur{M, Hidde DE JONG}{DR}{Inria, Grenoble}     %% Co-Directeur de thèse s'il y en a
+\UGTInvite{M, Jean-Luc GOUZÉ}{DR}{Inria, Nice}
 %\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}
@@
 
-\MakeUGthesePDG    %% très important pour générer la couverture de thèse
+\MakeUGthesePDG    %% très important pour générer la couvérture de thèse

### 2016-12-28 Chapters/introduction.tex
Change titlepage (new format) and write Suppinfo S5 about strains

@@
 \centering
-\includegraphics[height=5cm]{./Fig/Chapter1/monod_law.eps}
+\includegraphics[height=5cm]{./Fig/Chapter1/monod_law}
 \caption{
@@
 \centering
-\includegraphics[height=7cm]{./Fig/Chapter1/scott_rnaprot.eps}
+\includegraphics[height=7cm]{./Fig/Chapter1/scott_rnaprot}
 \caption{
@@
 \centering
-\includegraphics[height=6cm]{./Fig/Chapter1/growth_curve.eps}
+\includegraphics[height=6cm]{./Fig/Chapter1/growth_curve}
 \caption{

### 2016-12-29 Chapters/experiments.tex
Correct some figures and add the suppinfo of chapter 3 except the part about the results obtained in microplate reader

@@
 \textbf{Results of data acquisition.}
-We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
+We imaged 6 fields (X[1-3]Y[1-2]) each containing 15 wells (W[0-14]) for $\sim$40~h.
 From the 90 wells, 68 were suitable for further analysis (the others being empty, out of frame for some period of time, or plugged).
-For all wells, data points are missing in the interval between -720 and -150 min because the camera was out of focus.
-We stopped the analysis at 550 min, when about 3/4 of the bacteria were still growing, before the entire population died within a few hours for an unknown reason.
-The image labelled "Raw image" is the last image analyzed for the highlighted well.
+For all wells, data points are missing in the interval between [-720,-150] because the camera was out of focus.
+We stopped the analysis at 550~min, when about 3/4 of the bacteria were still growing, before the entire population died within a few hours for an unknown reason.
+The image labeled "Raw image" is the last image analyzed for the highlighted well.
 The bacterium on the left of this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
 A 6-pixel-wide rectangular mask was computed for each image, resulting in the "Segmented image" on the right.
-Fluorescence intensities are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335) and were corrected for camera background, but not autofluorescence background (Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
+Fluorescence intensities are expressed in Relative Fluorescence Units (RFU) on a 16-bit image and were corrected for camera background, but not autofluorescence background (Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
 The fluorescence intensity of the cell, expressed in units RFU/pixel/cell, was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
 The cell length is the distance in pixels between the two poles (red line).
-The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labelled "X3Y2, W2".
+The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labeled "X3Y2, W2".
 The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
@@
 \subsection{Estimation of growth rate and resource allocation profile}
+\label{sec:estim_gr_ra}
 
@@
 As a consequence, the analysis has been limited to a time-interval after the upshift in which roughly 3/4 of the cells are still growing.
-Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a 1~h-long pause around 200 min after the upshift, followed by a recovery of the growth rate until the end of the experiment (see Supporting Information XX).
-In order to focus the analysis, we decided to classify the cells in three categories: dying cells (N=12) for which the growth rate drops to zero after the upshift, pausing cells (N=11) for which the growth rate drops to zero after the upshift and then recovers, and finally so-called normal cells (N=45) which do not exhibit any of the above behaviors.
-More information about each category is available in Supporting Information XX.
+Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a 1~h-long pause around 200 min after the upshift, followed by a recovery of the growth rate until the end of the experiment (see~\ref{S2_Figure}  and \ref{S3_Figure}).
+In order to focus the analysis, we decided to classify the cells in three categories: dying cells (N=12) for which the growth rate drops to zero after the upshift, pausing cells (N=11) for which the growth rate drops to zero after the upshift and then recovers, and finally so-called normal cells (N=45) which do not exhibit any of the above behaviors (see~\ref{S2_Figure} and \ref{S3_Figure})
 In what follows, we focus on the 45 normal cells (Fig.~\ref{fig:growth_rate_estimation_median}) which show a globally coherent behavior in the time window considered here.
@@
 In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of Chapter~\ref{chap:theory}.
-We estimated the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (Supporting Information XX) and chose parameters that allow the model to reproduce the observed growth rates and fluorescence densities.
+We estimated the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (\ref{S7_Text}) and chose parameters that allow the model to reproduce the observed growth rates and fluorescence densities.
 We simulated an upshift from acetate to glucose and tried to estimate $\gamma \alpha(\cdot) / \beta$ from these synthetic data, selecting the \textit{prior} in the Kalman smoothing procedure by trial and error.
@@
 \centering
-\includegraphics[scale=0.8]{./Fig/synthetic_upshift}
+\includegraphics[scale=1]{./Fig/synthetic_upshift}
 \caption{
 \textbf{Performance of the Kalman smoothing procedure on synthetic data simulating an acetate-glucose upshift.}
-\textit{(A)}~Synthetic data simulating an upshift from minimal medium with acetate to minimal medium with glucose, with and without additive white noise, as well as the results obtained by Kalman smoothing.
+\textit{(A)}~Synthetic data simulating an upshift from acetate to glucose, with and without additive white noise, as well as the results obtained by Kalman smoothing.
 The synthetic data were generated by simulating the model presented in Eqs~\ref{eq:pdef}-\ref{eq:rdef} with the on-off regulatory strategy (Eq.~\ref{eq:stratswitch} and Fig.~\ref{fig:onoffresults}).
 The model parameters used for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $K_R = 1$~g~L\textsuperscript{-1}.
-The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~L~g\textsuperscript{-1}~RFU\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
-The noise level was estimated from the data (Supporting Information XX) and added to $F$.
+The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~RFU~L~g\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
+The noise level was estimated from the data (\ref{S7_Text}) and added to $F$.
 The choice of the parameters of the Kalman smoothing procedure is discussed in the Material and Methods~\ref{sec:meth_kalman}.
@@
 \textit{(C)}~The predicted growth rate during the upshift experiment.
-This information is used as an input in the smoothing procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+This information is used as an input in the smoothing procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), ..., L(t_{N-1}) \right\}$.
 }
@@
 \centering
-%\includegraphics[scale=1]{./Fig/}
+\includegraphics[scale=1]{./Fig/heatmap}
 \caption{
 \textbf{Global overview of all estimated resource allocation profiles presented in Fig.~\ref{fig:gene_activity}.}
-The individual $\gamma \alpha (\cdot) / \beta$  curves have been normalized with respect to their maximum in the interval $[0, 400]$ and then ordered with respect to their maximum in the interval $[0, 100]$.
+The individual $\gamma \alpha (\cdot) / \beta$  curves have been normalized with respect to their maximum in the interval $[0, 400]$ and then ordered with respect to their maximum in the interval $[0, 120]$.
 }
@@
 
-\subsubsection{Full sequence of the inserts}
+\subsubsection{Full annotated sequences of the inserts}
 
@@
 Regions at the end of the \textit{rpsB} gene were amplified, and sent for sequencing.
-Alignments of the results showed they exhibit the expected sequences.
+Alignments of the results showed they exhibit the expected sequences (except for a couple of single-base mutations).
 The raw results of the sequencing are reproduced below.
@@
 
+\clearpage
 \subsection{S6 Text -- Strain validation}
@@
 
-\subsection{S7 Text -- Microscopy analysis}
+\clearpage
+\subsection{S7 Text -- Noise estimation in the microscopy experiment}
 \manuallabel{S7_Text}{S7~Text}
 
-\subsubsection{Category of cells}
-
-\begin{figure}[p]
+We estimated the noise in the RFU/pixel/cell by using the data points just before upshift (Fig.~\ref{fig:noise_poi}).
+Since they have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
+For this reason, ribosome concentration, and so fluorescence concentration, are expected constant.
+Fig.~\ref{fig:noise_poi} represents the distribution of points in this region for a particular cell, that can be approximated by a Gaussian distribution.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/noise_poi}
+\caption{
+\textbf{Point of interest for the noise estimation.}
+\textit{(Top graph)} RFU/pixel/cell data points for the cell at the bottom of the well labeled X3Y2,W2.
+Noise estimation was performed on the points just before the upshift (red points), were the bacteria is assumed to grow at steady state.
+\textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (CDF, in gray) of the points highlighted in red on the top graph.
+They are visually compared with CDF (solid green line) and PDF (dashed green line) of a Gaussian fit with mean 548.88 and standard deviation 17.316.
+}
+\label{fig:noise_poi}
+\end{figure}
+
+Interestingly, as can be seen in Fig.~\ref{fig:noise_distrib_mean_std}, the mean and standard deviation of the distribution vary from cell to cell in a slightly correlated manner (Pearson R\textsuperscript{2}: $0.3213$, p-value: $4.923\cdot10^{-5}$).
+We did not investigate if this heterogeneity between cells is the result of true biological variations or a bias due to the microfluidic device.
+By looking at the points of Fig.~\ref{fig:noise_distrib_mean_std}, there are however no strong correlation of the mean and standard deviation with the acquisition field (XY), or the position of the well on the image (W).
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/noise_distrib_mean_std}
+\caption{
+\textbf{Distribution of the means and standard deviations in the region of interest for the normal cells.}
+The empirical mean and standard deviation were evaluated in the region before the upshift (red points on Fig.~\ref{fig:noise_poi}) for each of the 45 normal cells.
+They appear to be slightly correlated (Pearson R\textsuperscript{2}: $0.3213$, p-value: $4.923\cdot10^{-5}$) which could indicate a multiplicative instead of an additive noise.
+}
+\label{fig:noise_distrib_mean_std}
+\end{figure}
+
+Noise characteristics were computed by normalizing and aggregating data for the 45 normal cells presented in Fig.~\ref{fig:noise_distrib_mean_std}.
+Two scenarii were considered.
+First, the noise model can be considered additive, which means we have the measurement model presented in Section~\ref{sec:estim_gr_ra}:
+\begin{equation}
+F(t_k) = \gamma r (t_k) + \eta_k,\\
+\end{equation}
+were $F(t_k)$ is the fluorescence measured at the time step $t_k$, $r (t_k)$ is the true ribosomal concentration at this time step, $\gamma$ is an unknown factor, and $\eta_k$ is the measurement noise.
+In this scenario, for a time independent ribosome concentration $r$ (steady state), the ribosome concentration is constant and equal to its mean on the interval of interest (noted $m_k(\cdot)$).
+Given that
+\begin{equation}
+m_k (F) = m_k (\gamma r) + m_k(\eta_k) = \gamma m_k (r),\\
+\end{equation}
+the noise residues are given by
+\begin{equation}
+\eta_k = F(t_k) - m_k (F). \label{eq:noise_add}
+\end{equation}
+In other words, we can normalize the data points of each cell by removing their mean in order to aggregate the noise residues and get an estimation of the noise level (Fig.~\ref{fig:noise_estim_add_mul}, top panel).
+
+Secondly, the noise model can be considered multiplicative.
+The noise model is than different than presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
+\begin{equation}
+F(t_k) = \gamma r (t_k) \cdot (1 + \lambda_k),
+\end{equation}
+were $\gamma r (t_k) \cdot \lambda_k$ is the measurement noise, proportional to the value measured.
+In that scenario, with the same notations as above, the noise $\lambda_k$ is given by
+\begin{equation}
+\lambda_k = \frac{F(t_k)}{m_k (F)} - 1, \label{eq:noise_mul}
+\end{equation}
+which means we can normalize data points from different cells by dividing by the mean and removing 1 (Fig.~\ref{fig:noise_estim_add_mul}, bottom panel).
+
+Both models were considered and are represented in Fig.~\ref{fig:noise_estim_add_mul}.
+Despite the correlation identified in Fig.~\ref{fig:noise_distrib_mean_std}, the means of the residues in both models are very close to zero, and the distributions are well approximated by a Gaussian distribution.
+For this reason, we decided in the main analysis to stick with an additive noise, the main reason being that the implementation used for the Kalman smoothing procedure are faster and more stable with an additive noise model.
+We thus assumed an additive white Gaussian noise with mean 0 and standard deviation 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
+As discussed in Section~\ref{sec:chap3_discussion}, the only way to decisively choose a noise model over the other would have been to obtain two different steady-state growths for the each cell.
+Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future instances of the experiment.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/noise_estim_add_mul}
+\caption{
+\textbf{Distribution of the noise residues after normalization and aggregation for the 45 normal cells.}
+Two noise models were considered for the normalization: either the temporal mean $m_k (\cdot)$ was removed from the measurements independently for each cell (additive noise model in Eq.~\ref{eq:noise_add}), or $1$ was removed from the ratio of the measurements with the temporal mean $m_k (\cdot)$ (multiplicative noise model in Eq.~\ref{eq:noise_mul}).
+For the additive noise model, the fitted Gaussian distribution as a mean of $1.256\cdot 10^{-15}$ and a standard deviation of 28.29.
+For the multiplicative noise model, the fitted Gaussian distribution as a mean of $2.666\cdot 10^{-18}$ and a standard deviation of 0.05383.
+}
+\label{fig:noise_estim_add_mul}
+\end{figure}
+
+\clearpage
+\subsection{S8 Text -- Examples of complete analysis for 5 normal cells}
+\manuallabel{S8_Text}{S8~Text}
+
+To help the reader get a better idea on the analysis, we display below the complete reconstructions for 5 cells.
+In Figs~\ref{fig:cell1}-\ref{fig:cell5}, the image on the left is the last image analyzed for the corresponding well.
+The bacteria either at the top or the bottom (depending on the orientation in the device) was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
+A 6-pixel-wide rectangular mask was computed for each image, resulting in the masked image on the right.
+In each of the graphs, the black and green points represents data points, while the red solid lines are the results of the signal reconstruction \textit{via} the Kalman smoothing procedure, the parameters of which are described in Material and Methods~\ref{sec:meth_kalman} and Figs~\ref{fig:growth_rate_estimation} and \ref{fig:gene_activity}.
+Specific comments on each cell are given in the caption of Figs~\ref{fig:cell1}-\ref{fig:cell5}.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/cell1}
+\caption{
+\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y1, W11".}
+(Figure description available in the introduction of \ref{S8_Text}.)\newline
+Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seems to have stabilized before the upshift.
+Oscillations in the RFU/pixel/cell signal are clearly visible, which results in oscillations in the resource allocation signal reconstruction, even though the smoothing factor seems a little too high for this particular cell.
+}
+\label{fig:cell1}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/cell2}
+\caption{
+\textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y1, W7".}
+(Figure description available in the introduction of \ref{S8_Text}.)\newline
+Here, the resource is unstable at the beginning of the experiment despite the apparent regularity of the RFU/pixel/cell.
+This could indicate a smoothing factor that is too low for this particular cell.
+}
+\label{fig:cell2}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/cell3}
+\caption{
+\textbf{Complete analysis for the cell located at the bottom of the well labelled "X3Y2, W5".}
+(Figure description available in the introduction of \ref{S8_Text}.)\newline
+We see for this cell that the reconstruction of the growth rate on the acetate medium (before 0) is affected by the huge gap in the acquisition, the increase before -150~min clearly being an artifact.
+However, because the Kalman smoothing procedure we used allows for flexibility between the mother and the daughter cell, we quickly recover a more realistic growth rate before the upshift.
+Despite an instability at the beginning of the experiment, the resource allocation reconstruction exhibits a remarkable stability during growth on the acetate medium, followed by oscillations after the upshift on glucose (also visible directly on the RFU/pixel/cell data).
+}
+\label{fig:cell3}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/cell4}
+\caption{
+\textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y2, W3".}
+(Figure description available in the introduction of \ref{S8_Text}.)\newline
+The reconstruction on this cell is a good instance of what we aimed to obtain: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
+The resource allocation is stable on a acetate (indicating a steady state) than starts to oscillate after the upshift on glucose.
+}
+\label{fig:cell4}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/cell5}
+\caption{
+\textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y2, W6".}
+(Figure description available in the introduction of \ref{S8_Text}.)\newline
+The results on this cell are particularly terrible: we do not seem to get a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating.
+The oscillations after the upshift are thus very hard to interpret.
+}
+\label{fig:cell5}
+\end{figure}
+
+\clearpage
+\subsection{S2 Figure -- Cell categories identified in the microscopy analysis}
+\manuallabel{S2_Figure}{S2~Fig}
+
+\begin{figure}[h]
 \centering
 \includegraphics[scale=1]{./Fig/subcat_cells}
-\caption{Subtypes of cells. Caption to write. Cutting the x axis might be a good idea because a third of the x-axis of each graph are "missing data".}
+\caption{
+\textbf{Cell categories identified in the microscopy analysis.}
+As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: the dying cells (lowest graphs) stop growing, pausing cells (middle graph) present growth rate oscillations, and normal cells (highest graph) does not present any of the above}
 \label{fig:subcat_cells}
@@
 
-\begin{figure}[p]
+\clearpage
+\subsection{S3 Figure -- Robust statistics for the cell categories}
+\manuallabel{S3_Figure}{S3~Fig}
+
+\begin{figure}[h]
 \centering
 \includegraphics[scale=1]{./Fig/subcat_median}
-\caption{Subtypes of cells, median. Caption to write.}
+\caption{
+\textbf{Robust statistics for the cell categories identified in the microscopy analysis.}
+Each graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step for the growth rate presented in~\ref{S2_Figure}.
+The gray area represents the interquartile range.
+}
 \label{fig:subcat_median}
 \end{figure}
-
-\subsubsection{Noise estimation}
-
-Blabla
-
-\subsubsection{Complete analysis for 5 cells}
-
-blabla

### 2016-12-30 Chapters/experiments.tex
Add strain validation figure

@@
 
-\subsection{S5 Text -- Strain construction}
+\subsection{S5 Text -- DNA sequences used for the strain construction}
 \manuallabel{S5_Text}{S5~Text}
@@
 \clearpage
-\subsection{S6 Text -- Strain validation}
+\subsection{S6 Text -- Strain validation in batch growing conditions}
 \manuallabel{S6_Text}{S6~Text}
@@
 Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
+
+Two type of strains were obtained
+
+Before being used in the microscopy experiments, the strain characteristics were validated by
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/suppinfo_caract}
+\caption{
+\textbf{Growth curves in M9~0.2\% glucose for the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
+Growth and measurements were performed in a 96-well microplate placed in a Tecan infinite 200 pro reader thermostated at 37$^\circ$C.
+We monitored 4 wells inseminated with the WT strain, the \textit{rpsB-gfp} strain, and the \textit{rpsB-mCherry} strain, for a total of 12 wells each containing 150~$\mu$L of growth medium.
+Curves were shifted in time to correct for pipetting variability in the insemination process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
+\textit{(A)}~The light absorbance as measured by the Tecan microplate reader, corrected for background by removing well containing clean M9.
+The superposition of the curves indicates similar growth rates between the WT and the modified strains.
+\textit{(B)}~The fluorescence measured in the wells, expressed in Relative Fluorescence Units (RFU).
+Green fluorescence (485nm excitation, 535nm reception) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
+Red fluorescence (560nm excitation, 635nm reception) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
+As can be seen, fluorescence levels of the modified strains (solid lines) are far above the autofluorescence measured on the WT strain.
+\textit{(C)}~The ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
+Autofluorescence background was corrected by removing the autofluorescence measured on the WT wells.
+We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something that is not visible on the \textit{rpsB-gfp} strain.
+Before each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
+}
+\label{fig:suppinfo_caract}
+\end{figure}
 

### 2016-12-30 Chapters/experiments.tex
Hans modification of Material and Methods

--- 2016-12-28 Chapters/experiments.tex
@@
 
-Most studies of the growth of microorganisms have been done at steady state.
-This is a reasonable and logical choice given that at steady state, the behavior of microorganisms is reproducible, which helps uncovering simple theories to apprehend their inherent complexity.
-But while this condition can be easily achieved in the laboratory, microorganisms naturally spend very little time in steady state.
-This motivated the construction of a new theoretical framework in Chapter~\ref{chap:theory} to study growth laws during growth transitions.
+Most studies on the growth of microorganisms are done at steady state.
+It is a reasonable and logical choice given the drawbacks of working in dynamic.
+At steady state, microorganism behaviors are reproducible, and which helps uncover simple theories to apprehend their inherent complexity.
+But while this ideal environment is easily achievable at the bench, microorganisms naturally spend very little time in such a perfect condition.
+This motivated in chapter~\ref{chap:theory} the construction of a new theoretical framework to study growth laws during growth transitions.
 
-The principles underlying regulatory processes in dynamical conditions appear to be different from those applying at steady state.
-We showed that regulatory strategies approaching the theoretical optimal, given by bang-bang control of gene expression machinery, have to be able to effect abrupt and strong variations in the activity of a gene.
-It was seen that for implementing such strategies the regulatory system needs to be capable of sensing the internal state of the cell and not just the environment.
-A near-optimal transition also requires information of several different variables, whereas for maintaining a steady state with maximal biomass accumulation a single variable turned out to be sufficient in our theoretical framework.
-Interestingly, we showed that the ppGpp regulatory system of \textit{E.
-coli} fulfills the requirement of such a regulatory strategy, by implementing an on-off strategy for regulating the synthesis of ribosomes.
-However, although this provides circumstantial evidence that bacteria control resource allocation in a manner consistent with theoretical optimality, experimental data are necessary to decide if an on-off strategy is at work in \textit{E.
-coli} during growth transitions.
+The principles that define the regulatory processes in dynamic appear to be different than the ones that apply at steady state.
+We showed that efficient regulatory strategies have to be able to perform extremely abrupt and intense variations of the gene activity.
+This relies on the fact that bang-bang synthesis of gene expression machinery settles the upper-bound for biomass production during an environmental change.
+Such strategies are only implementable if the regulatory system focus on the internal state of the cell.
+But not all information is equivalent, and performing a near-perfect transition might require to measure several variables, a concept totally missing from steady-state considerations.
+Interestingly, we uncovered how the ppGpp regulatory system of \textit{E.
+coli} might be a cheap way for the cell to access just enough information, and to implement a \textit{on-off} strategy to regulate the synthesis of ribosomes.
+Overall, it taught us that the actual complexity of regulatory systems is only beneficial in a dynamical context.
+But elucidating what is the best behavior, and showing that \textit{E. coli} actually implement it are different matters...
 
-Unfortunately, far more information is available on ribosome abundance at steady state than during growth transitions~\cite{scott_interdependence_2010,gausing_regulation_1980}.
-The main reason for this bias on steady-state conditions is that, from an experimental point of view, growth transitions are hard to control and may depend on the history of the culture~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
-However, as we discussed in Section~\ref{sec:discussion}, the few studies that have been reported seem to be consistent with our predictions, in the sense that they indicate that during growth transitions, the synthesis rate of ribomoses oscillates~\cite{gausing_regulation_1980,zengel_transcription_1986} and the ppGpp concentration manifests a rapid succession of increases and decreases~\cite{friesen_synthesis_1975,murray_control_2003}.
+Reality actually highlights the problem with adopting a dynamical perspective on growth.
+There are far more information on ribosome abundance at steady state than during growth transitions.
+It is widely admitted that the latter are hard to control and rely too much on hidden variables about the strain history~[citation ? Bremmer ?].
+Though, as we discussed in section~\ref{sec:discussion}, some data are available and seem to confirm that during growth transitions, the  synthesis rate of ribomoses can oscillate~\cite{gausing_regulation_1980,zengel_transcription_1986}, and that the ppGpp regulatory system can go through rapid changes~\cite{friesen_synthesis_1975,murray_control_2003}.
 But we cannot decisively validate or disprove the model predictions from measurements that were carried out at the population level, where it is inherently hard to identify switching patterns.
-What is needed for the verification of our predictions are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
-While the observation of an on-off strategy would show that simple models of the type introduced in Chapter~\ref{chap:theory} are instrumental in gaining a better understanding of microbial physiology, its refutation would also raise interesting questions.
-If maximization of growth rate and biomass accumulation have been retained by natural selection in \textit{E.
-coli}, which factors would prevent the cell from behaving optimally from a theoretical point of view?
+What we need are dynamical single-cell measurements of the ribosome concentration during well-controlled growth transitions.
+Would the \textit{on-off} strategy be observed, it could confirm the importance of optimizing transitions for the organism.
+Would it be disproved, that could raise even more interesting questions: growth rate and biomass might not necessarily be what natural selection care for, or other unexpected limitations might prevent the cell to behave optimally.
 
-The aim of this chapter is to measure \textit{in vivo}, during a growth transition, the resource allocation profile $\alpha (\cdot)$ and compare its dynamics with the gold standard established in Chapter~\ref{chap:theory}.
-To this end, we have measured ribosomal abundance of \textit{E. coli} at the single-cell level during a nutrient upshift.
-More precisely, we constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing the \textit{in-vivo} monitoring of (changes in) the abundance of ribosomes.
-In collaboration with Irina Mihalcescu of the Laboratoire Interdisciplinaire de Physique, we cultivated this \textit{E.
-coli} strain in a microfluidic device, allowing the long-term imaging of individual cells in well-controlled conditions, notably involving a classic upshift experiment from growth in minimal medium with acetate to minimal medium with glucose.
-We then developed a Kalman smoothing method, in collaboration with Eugenio Cinquemani of Inria Grenoble -- Rh\^{o}ne-Alpes, to reconstruct $\alpha (\cdot)$  from the estimates of the variations of the growth rate and the relative ribosomal synthesis rate.
-While the preliminary results presented here do not allow a decisive validation of the expected behavior, among other things due to the difficulties that were encountered during the experiments and the image analysis, we believe they are promising as a first step towards the better understanding of global resource allocation during growth transitions.
-
-For clarity, the model of resource allocation during growth transitions developed in Chapter~\ref{chap:theory} is reproduced here, where dotted variables represent time derivatives:
+In Chapter~\ref{chap:theory}, we developed a model of resource allocation during growth transitions.
+By applying optimal control theory, we predicted that in order to maximize biomass production, the cell should produce its gene expression machinery in a bang-bang manner.
+The full model with dimensional parameters is reproduced below, with dotted variables representing their time derivative:
 \begin{eqnarray}
@@
 In this form, it contains 4 variables ($e_M(t)$, $p(t)$, $r(t)$, $\alpha (t)$) and 3 parameters ($\beta$, $k_R$, $K_R$).
-$e_M(t)$ [min\textsuperscript{-1}] is an indicator of the richness of the environment.
+$e_M(t)$ [min\textsuperscript{-1}] is the richness of the environment.
 $p(t)$ [g.L\textsuperscript{-1}] is the precursor concentration inside the cell.
-$r(t)$ [g.L\textsuperscript{-1}] is the concentration of gene expression machinery inside the cell.
-$\alpha (t)$ [$\emptyset$] is the fraction of resources allocated to the synthesis of gene expression machinery.
+$r(t)$ [g.L\textsuperscript{-1}] is the gene expression machinery concentration inside the cell.
+$\alpha (t)$ [$\emptyset$] is the mass ratio of resources that goes to the gene expression machinery.
 $k_R$ [min\textsuperscript{-1}] is the rate constant of macromolecular synthesis.
 $K_R$ [g.L\textsuperscript{-1}] is the half-saturation constant of macromolecular synthesis.
-$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules, assumed to be constant.
+$\beta$ [L.g\textsuperscript{-1}] is the inverse of the cellular density of macromolecules (assumed constant).
+The goal of this chapter is to measure in vivo $\alpha (\cdot)$ during a growth transition, and check how it compares with the gold standard established in chapter~\ref{chap:theory}.
+
+To this end, we took the challenge of measuring ribosomal abundance of \textit{E. coli} at the single-cell level during a controlled upshift.
+We made a strain displaying fluorescent ribosomes, thus allowing in-vivo measurement of their abundance in dynamical conditions.
+We placed it in a microfluidic device allowing the long-term imaging of individual cells while they are presented different growing media, and focused on the classic upshift from acetate to glucose.
+We then used Kalman smoothing to estimate the variations of the growth rate and the relative ribosomal synthesis rate during this upshift.
+While not allowing a decisive validation of the expected behavior, this experiment opens interesting questions and is a step towards a better understanding of the gene expression machinery regulation during growth transitions.
 
@@
 
-How can one measure the resource allocation variable in \textit{E. coli} cells?
-There is no direct way to quantify $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
+How does one measure resource allocation in real cells?
+There is no direct way to measure $\alpha (\cdot)$ in an experiment, mostly because of its abstract nature.
 To correctly reconstruct it, one has to know the value of every single term in the equations in which it appears.
-In the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only appears in Eq.~\ref{eq:rdef-exp}.
-We can therefore identify $\alpha (t_k)$ for each time $t_k$ when the terms $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$ are known, or equivalently, the six individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$ they are composed of.
-Even when assuming that the parameters $k_R$, $K_R$, and $\beta$ are known, or at least easy to estimate independently, reconstructing the dynamics of $\alpha$ would require estimation of the (changes in) concentration of the gene expression machinery ($r$, $\dot{r}$) and the precursor concentration $p$.
+In the full dynamical system Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}, $\alpha$ only pops up in Eq.~\ref{eq:rdef-exp}, and is already isolated in this equation, which simplifies the burden.
+We can identify $\alpha (t_k)$ for each time $t_k$ where the following 3 terms are known: $\beta r(t_k)$, $\dot{r}(t_k)$, and $\frac{k_R \cdot p(t_k)}{K_R + p(t_k)} \cdot r(t_k)$.
+Such a consideration remains valid if one can estimate the 6 individual components $r(t_k)$, $\dot{r}(t_k)$, $p(t_k)$, $k_R$, $K_R$, $\beta$.
+Even if we assume that the constant parameters are known, or at least easier to estimate independently, reconstructing the dynamics of $\alpha$ would require to co-estimate the gene expression machinery concentration ($r$, $\dot{r}$) and the precursor concentration $p$.
 
-Are such measurements feasible in practice?
-The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, or using radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques~\cite{hui_quantitative_2015}.
-However, only the use of fluorescent proteins complies with our need for real-time single-cell quantification.
-Obtaining an estimate of precursor abundance is even more challenging, since there is no clear proxy for the totality of precursors in the cell and, while absolute quantification of all internal metabolites has been achieved in \textit{E.
-coli}~\cite{bennett_absolute_2009}, this method is also not suitable for a dynamical estimation in individual cells.
+Are such measurements acceptable?
+Measuring the abundance of gene expression is achievable if one focus on a proxy that represents the overall abundance.
+Ribosomes are widely known to be the most abundant part of the gene expression machinery, which allows to consider it as its main component~\cite{scott_interdependence_2010, scott_bacterial_2011,scott_emergence_2014}.
+Their abundance can be estimated for instance by using the total RNA of the cell~\cite{scott_interdependence_2010}, radioactive markers\cite{gausing_regulation_1980,zengel_transcription_1986}, or fluorescent labeling~\cite{bakshi_superresolution_2012}.
+Estimating precursor abundance is more challenging.
+There is no clear proxy that represents all the precursors in the cell, so using common methods would be excessively time expensive~[citation classic methods to estimate precursor ?].
+While absolute quantification of all the internal metabolites has been achieved in \textit{E.
+coli}~\cite{bennett_absolute_2009}, this method do not comply with our need of real-time single-cell quantification.
 
-The problem of quantifying $\alpha (\cdot)$ can be simplified by applying a transformation of the model of Eqs~\ref{eq:pdef-exp}-\ref{eq:rdef-exp}.
-Taking into account that by construction the growth rate $\mu(t)$ is given by
+This problem can be overcome by using a simple transformation.
+Taking into account that by construction the growth rate  $\mu(t)$ is given by:
 \[
@@
 \end{equation}
-The problem of estimating $\alpha(\cdot)$ is thus equivalent to the problem of estimating $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$.
+The problem of estimating $\alpha(\cdot)$ is thus equivalent to the same problem where $r(\cdot)$, $\dot{r}(\cdot)$, $\mu (\cdot)$ and $\beta$ have to be co-estimated.
 If we are satisfied with estimating $\alpha$ up to a constant factor $\beta$, dynamically measuring the ribosome concentration and the growth rate would make $\alpha (\cdot) / \beta$ identifiable.
 
-Given that fluorescent reporters of ribosomal proteins can provide information on both $r(\cdot)$ and $\dot{r}(\cdot)$ \cite{zulkower_robust_2015}, the above reformulation of the problem is promising for the purposes of our study.
-Moreover, since the predicted bang-bang profiles in gene activity are easily hidden at the population level if cells are not synchronous, which is generally the case, expression of the ribosomal genes need to be monitored at high sampling density on the single-cell level, which is also possible with fluorescent reporters.
-Inspired by previous work~\cite{bakshi_superresolution_2012}, we therefore constructed an \textit{E.
-coli} strain in which the S2 ribosomal subunit, encoded by the gene \textit{rpsB}, has been tagged with a green fluorescent protein (GFP).
-The strain with the chromosomal \textit{rpsB-gfp} fusion produces fluorescent ribosomes that are quantifiable \textit{in vivo} in single cells while not affecting the growth rate (see Material and Methods~\ref{sec:methods_strain} and~\cite{bakshi_superresolution_2012}).
-Monitoring single cells of this strain, by time-lapse fluorescence microscopy, thus makes it possible to estimate $r(\cdot)$ and $\dot{r}(\cdot)$ and reconstruct $\mu(\cdot)$.
+These new measurements are more acceptable given the requirements of the study.
+We want to be able to identify potential abrupt changes of the gene expression.
+Such switching patterns are easily hidden at the population level if cells are not synchronous, which is generally the case.
+A short sample time and single-cell measurements are thus necessary, which motivated the use of fluorescent reporters like the green fluorescent protein (gfp).
+We constructed a strain displaying a gfp-tagged S2 ribosomal subunit (the gene name being \textit{rpsB}).
+The \textit{rpsB-gfp} strain produces fluorescent ribosomes, making them quantifiable while minimizing the burden for the cell (see Material and Methods~\ref{sec:methods_strain}).
+By observing such a strain through a microscope during a growth transition, we could theoretically estimate the ribosome concentration by measuring the cell fluorescence, and reconstruct the growth rate by measuring the cell size.
 
-The growth conditions also need to be carefully chosen for this study.
-In natural environments, for many bacteria one of the most frequently encountered growth transitions, or at least the transition on which selection is expected to operate, is the transition from stationary phase in a nutrient-deprived medium~\cite{gefen_direct_2014} to (exponential) growth after the renewed availability of nutrients.
-The transition from stationary to exponential phase is difficult to study though as many hard-to-control parameters play a role, including the duration of the nutrient stress, the composition of the medium before growth arrest, and the accumulation of waste products in the medium~\cite{ehrenberg_mediumdependent_2012}.
-As a consequence, growth resumption only occurs after a lag phase of variable duration~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}, which is not included in the model.
-For this reason, we decided to focus on a transition from exponential growth on a poor carbon source (acetate) to exponential growth on a rich carbon source (glucose) in a well-defined minimal medium.
-By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively, to ensure that the cells have no memory of their physiological state before the transition and have enough time to reach the new steady state after the transition.
-The use of the mother machine, a microfluidic device that was designed to sustain exponential growth for long periods of time, is therefore a good choice for this purpose.
-It allows to maintain a constant flow of fresh medium and to effect fast transitions by switching the medium (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
-The complete experimental plan is summarized in Fig.~\ref{fig:experiment_schema}.
+The growth conditions might also play a role in the limitations of the study.
+In natural conditions, the most straightforward growth transitions, or at least the one at which most of the selection is expected to occur, is the transition from the stationary phase in nutrient-deprived medium to any medium sustaining growth.
+There are good reasons though for which this particular growth transition is ignored from most studies.
+The transition from stationary to exponential phase is dramatically ill-controlled, in the sense that a lot of hard-to-control parameters seem to play a huge role in it.
+For instance, this is the case of the duration of the nutritive stress~[citation to find], the nutrient encountered during the last exponential phase~[citation to find], or the previous accumulation of wastes in the medium~[citation to find].
+This somehow results in a phenomenon called the lag phase~[citation lag phase review to find], the mechanisms of which are missing from our model.
+Furthermore, most microorganisms dramatically change their physiology in stationary-phase, either sporulating or organizing into biofilms~[citations subtilis sporulation and coli biofilms to find], which goes largely beyond the scope of this study.
+For this reason, we decided to focus on a transition from a slow exponential growth to high exponential growth, in particular the transition from acetate to glucose.
+By construction, this steady-state-to-steady-state transition requires a long acquisition time before and after the shift, respectively to ensure cells have \textit{forgotten} the previous stress phase, and have enough time to reach the new steady state.
+The mothermachine, a microfluidic device that was designed to sustain exponential growth for long period of time, seemed like a good choice for this purpose.
+It allows to maintain a constant flow of fresh medium around the cells, and make fast transitions possibles by switching the medium that is fed to the cells (see Material and Methods~\ref{sec:growth_condition} and \ref{sec:microflu}).
+The complete experimental scheme is summarized in Fig.~\ref{fig:experiment_schema}.
 
@@
 \centering
-\includegraphics[scale=1]{./Fig/experiment_schema}
+\includegraphics[scale=1]{./Fig/experiment_schema.eps}
 \caption{
-\textbf{Schematic outline of the upshift experiment.}
-The goal is to measure the fluorescence and length/area of \textit{E. coli} \textit{rpsB-gfp} cells during an acetate-to-glucose upshift.
-We use M9 minimal medium supplemented with 0.2\% of acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
-A preculture was started from glycerol stock for 2.5 days on 0.2\% acetate in batch condition (shake flask).
+\textbf{Scheme of the upshift experiment.}
+The goal is to measure the fluorescence and growth rate of cells during an acetate-to-glucose upshift.
+All the media used are M9 minimal medium supplemented with 0.2\% of the indicated carbon source, in this case acetate or glucose (see Material and Methods~\ref{sec:growth_condition}).
+The \textit{rpsB-gfp} strain was precultivated for 2.5 days on 0.2\% acetate in batch condition (shaken flask).
 The day of the experiment, the cells were injected into the mothermachine and fed by a constant flow of fresh 0.2\% acetate medium (see Material and Methods~\ref{sec:microflu}).
 Fluorescence and phase contrast images were taken every 5 minutes.
-After 20~h, the feeding media was switched to 0.2\% glucose and maintained for 20~h while continuing image acquisition.
-Time 0 corresponds to the moment of the acetate-glucose upshift.
+Twenty hours later, the feeding media was switched to 0.2\% glucose (known to theoretically sustain a 5-time-higher growth rate) and was maintained for twenty more hours while imaging carried on.
+Time 0 corresponds to the exact moment when cells start to be in contact with the glucose medium.
 }
@@
 \subsection{Data acquisition}
-\label{sec:data_acquisition}
 
-An experiment following the plan of Fig.~\ref{fig:experiment_schema} was carried out, over a period of 40~h, but encountered a number of difficulties.
-First, the motor displacing the microscope along the Z-axis intermittently stalled, thus deactivating the autofocus and the Z-compensation for the tilt of the microfluidic device in the XY plane.
-As a consequence, until the experimenter manually intervened, many of the images acquired were out of focus.
-This is the reason why data points between -720 and -150 min are not exploitable, but fortunately the experiment was planned in such a way that when it really mattered, notably around the growth transition, someone was present to monitor the microscope.
-Second, at around 7~h after the nutrient upshift, the bacteria started to die for reasons that are unknown, the most plausible hypothesis being a phage contamination (see Discussion~\ref{sec:chap3_discussion}).
-Data analysis was therefore limited to 550 min after the upshift, when roughly 3/4 of the cells were still growing.
-For these and other reasons (see Section~\ref{sec:chap3_discussion}), the experiment will have to be repeated, but there was no time left for this in the framework of my PhD thesis.
+It is important to evoke most of the issues that occurred while carrying on the experiment.
+At the time, a serious issue was starting to affect our only microscope: the motorization along the Z-axis was randomly stalling until an operator manually intervenes.
+This deactivates the autofocus and the Z-compensation for the tilt of the device when moving along the XY plane.
+In other words, series of images are out of focus and thus not exploitable if the operator was not around, which can happen a lot in a 40-hour-long experiment.
+For the experiment presented here, we timed the operations so that the operator can be around as long as possible when it counts, \textit{i.e.} close to the transition at time 0.
+This is the reason why data points between -720 and -150 minutes are not available (corresponding to the steady state on acetate).
+Another issue, more specific to the presented occurence of the experiment, was that bacteria started to massively die several hours after the transition.
+The reason is unknown, the most plausible hypothesis being a phage contamination.
+\footnote{Note that the operator was present when the dying started, which excludes the implication of lab gnomes.}
+Data analysis was thus stopped at 550 minutes, when roughly 3/4 of the cells were still growing.
+Note that for these reasons, the results of this preliminary experiment must be taken with caution.
+The necessary improvements are extensively discussed in section~\ref{sec:chap3_discussion}.
 
-The acquired data consisted of phase contrast and fluorescence images of 60 wells in total, in four different fields.
-While quite a few image analysis programs have been reported in the literature~\cite{klein_tlm-tracker:_2012,wang_image_2010,young_measuring_2012,paintdakhi_oufti:_2016}, and some of these specifically address the mother machine design of the microfluidic device~\cite{wang_robust_2010,sachs_image-based_2016}, they all presented limitations when applied to our data.
-For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidic device.
-Moreover, the fluorescence density was concentrated in hot spots (Fig.~\ref{fig:data_acquisition}) and its intensity varied during the experiment. 
-While this was expected, since ribosomes are localized in the cell poles~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}, it complicated automatic segmentation on the fluorescence images.
+Image analysis was performed on the remaining shots.
+While image analysis softwares overflow the literature~[citations schitcells + others], some of them being specifically designed for the mother machine~[molysso + Wang et al], technical limitations did not allow to use any of them.
+For instance, the segmentation was systematically impeded by the poor resolution of the camera.
+The phase contrast images were not directly exploitable since we noticed they were not aligned with the fluorescence images.
+Fluorescence was also organized in hot spots (see Fig.~\ref{fig:data_acquisition}) and their intensity vary widely during the experiment, which was expected since ribosomes are extremely localized~\cite{bakshi_superresolution_2012} and ribosomal abundance is known to vary with the growth rate~\cite{scott_interdependence_2010}.
+It, however, dramatically impeded the automatic segmentation on the fluorescence images.
+In section~\ref{sec:chap3_discussion}, we give insights into how we plan to overtake these limitations.
 
-For these reasons, the image analysis carried out for this chapter has been much simplified. 
-First, we have focused on the cell at the bottom of each well, since this avoids the tracking of individual cells across several generations and ensures that the descendance of this cell can be followed throughout the experiment.
-Second, segmentation was done by manually selecting two pixels, one at each pole of the cell.
-These pixels were used to create a rectangle surrounding the cell (Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
-After background correction (Material and Methods~\ref{sec:cell_segmentation}), the fluorescence intensity in units RFU was evaluated for each pixel in the rectangle, as well as the length of the cell, defined by the distance in pixels between the poles (Fig.~\ref{fig:data_acquisition}).
-The fluorescence density for the entire cell [RFU/pixel/cell] was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
-Although we are well aware that the above procedure can be improved on many counts (Section~\ref{sec:chap3_discussion}), we nevertheless believe that it provides a reasonable approximation of the quantities of interest and a valid starting-point for the estimation of the growth rate and the resource allocation profile in the remainder of this chapter.
+The analysis presented in this study is thus preliminary.
+First, we focused solely on the deepest cell of each well.
+It avoided the tracking step, and ensured that the cell was visible from the beginning to the end of the experiment.
+Secondly, the segmentation was done by manually selecting two pixels at the poles of the cell of interest.
+These pixels were used to create a rectangular mask around the cell (see Fig.~\ref{fig:data_acquisition} and Material and Methods~\ref{sec:cell_segmentation}).
+After background correction (see Material and Methods~\ref{sec:cell_segmentation}), the RFU concentration was evaluated in this region of interest, as well as the length of the cell, defined by the distance in pixels between the poles (see Fig.~\ref{fig:data_acquisition}).
 
@@
 We imaged 6 fields (X1Y1, X2Y1, X3Y1, X1Y2, X2Y2, X3Y2) each containing 15 wells (W0 to W14) for $\sim$40 hours.
-From the 90 wells, 68 were suitable for further analysis (the others being empty, out of frame for some period of time, or plugged).
-For all wells, data points are missing in the interval between -720 and -150 min because the camera was out of focus.
-We stopped the analysis at 550 min, when about 3/4 of the bacteria were still growing, before the entire population died within a few hours for an unknown reason.
-The image labelled "Raw image" is the last image analyzed for the highlighted well.
-The bacterium on the left of this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
-A 6-pixel-wide rectangular mask was computed for each image, resulting in the "Segmented image" on the right.
-Fluorescence intensities are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335) and were corrected for camera background, but not autofluorescence background (Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
-The fluorescence intensity of the cell, expressed in units RFU/pixel/cell, was computed by dividing the sum of the fluorescence intensities of the pixels in the rectangle by the total number of pixels in the rectangle.
-The cell length is the distance in pixels between the two poles (red line).
-The thick lines in green and black highlight the time-varying length and fluorescence density for the cell visible in the top images, labelled "X3Y2, W2".
-The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
+From this, only 68 wells were available for analysis (the others were either empty, out of frame for some period of time, or plugged).
+For all wells, data points are missing in the time interval [-720,-150] because the camera was out of focus.
+We stopped the time series analysis at 550 minutes when visually 3/4 of the bacteria were still growing.
+The whole population was wiped out in the next few hours for an unknown reason.
+The Raw image is the last image analyzed for the highlighted well.
+The left bacteria on this image was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
+A 6-pixel-wide rectangular mask was computed for each image, giving the segmented image on the right.
+Fluorescence levels are expressed in Relative Fluorescence Units (RFU) on a 16-bit image (values from 0 to 65335), and were corrected for camera background, but not autofluorescence background (see Material and Methods~\ref{sec:cell_segmentation} and Discussion~\ref{sec:chap3_discussion}).
+The RFU/pixel/cell was computed by taking the ratio between the sum of the pixels in the mask, and the total number of pixels in the mask.
+The bacteria length is the distance in pixels between the two poles (red line on the raw and segmented images).
+The thick lines highlights the time series for the cell visible in the top images.
+The dashed black line corresponds to the switch time from acetate to glucose.
 }
@@
 
-In total, we obtained time courses of fluorescence density [RFU/pixel/cell] and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
-The fluorescence density appears constant during growth on acetate (before time 0) and immediately increases when the carbon source is switched to glucose.
-Inspection of the cellular length shows that the division frequency abruptly increases after the nutrient upshift, corresponding to a higher growth rate.
-Unfortunately, while steady-state growth on acetate was reached before the upshift, the fluorescence density profile suggests that the experiment did not continue long enough to ensure that a new steady state on glucose was reached.
-However, the data before and after the nutrient upshift are exploitable.
-What do we observe if we estimate the growth rate and the resource allocation profile from these data?
+From this preliminary analysis, we obtained time series for the concentration of fluorescence (RFU/pixel) and the length for 68 bacterial cells (Fig.~\ref{fig:data_acquisition}).
+The RFU/pixel appears constant on the acetate phase (before 0) and immediately increases when the medium is switched to glucose.
+Bacterial length shows that the division time abruptly decreases after the nutrient upshift, corresponding to a higher growth rate.
+Unfortunately, while the steady state on acetate was reached before the upshift, it is obvious that the time series stop before a new steady state on glucose is reached, at least for the fluorescence.
+Nevertheless, the time series are not interrupted around the transition.
+What do we observe if we estimate the growth rate and $\alpha$ in this region?
 
-\subsection{Data analysis using Kalman smoothing}
+\subsection{Data analysis using Kalman smoother}
 \label{sec:res_kalman}
 
-As we presented in section~\ref{sec:exp_design}, our goal is to reconstruct the signal $\alpha (\cdot)$ during a growth transition.
-By modeling \textit{E. coli} as a cylinder, the volume can be assumed directly proportional to the length of the cell.
+As we presented in section~\ref{sec:exp_design}, our goal is to reconstruct the signal $\alpha (\cdot)$ during the transition.
+By modeling \textit{E. coli} as a cylinder, the length can be assumed directly proportional to the volume of the cell.
 If we assume a negligible background of autofluorescence (see Material and Methods~\ref{sec:cell_segmentation}), the fluorescence concentration in the cell can be assumed proportional to the ribosome concentration.
-More precisely, based on the available data, we have the following measurement model at each time-point $t_k$, $0\leq k \leq N-1$:
+More precisely, based on the data available, we have the following measurement model at each time-point $t_k$:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in units RFU/pixel measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$ the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
-($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) uncorrelated sequence of measurement noise assumed normally distributed with mean 0.
-
-As explained in Section~\ref{sec:exp_design}, reconstructing $\alpha (\cdot)$ requires information on $\mu (\cdot)$, $r(\cdot)$ and $\dot{r}(\cdot)$.
-From Eq.~\ref{eq:mes_F}, we can obtain $r(\cdot)$ and $\dot{r}(\cdot)$ by smoothing interpolation and differentiation.
-Similarly, $\mu (\cdot)$ can be derived from Eq.~\ref{eq:mes_L} as it is defined by
+where $L(t_k)$ (resp.
+$F(t_k)$) is the length (resp.
+RFU/pixel) measured at time $t_k$ (visible on Fig.~\ref{fig:data_acquisition}), $V(t_k)$ (resp.
+$r(t_k)$) is the actual volume (resp.
+ribosome concentration) at time $t_k$, ($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) are measurement noises.
+As we saw in section~\ref{sec:exp_design}, reconstructing $\alpha (\cdot)$ requires information on $\mu (\cdot)$, $r(\cdot)$ and $\dot{r}(\cdot)$.
+From Eq.~\ref{eq:mes_F}, we could obtain $r(\cdot)$ and $\dot{r}(\cdot)$ by smoothing interpolation and differentiation.
+The same can be said for Eq.~\ref{eq:mes_L} and $\mu (\cdot)$ since it is defined by:
 \[
@@
 However, smoothing interpolation is particularly sensible to the boundary conditions.
-Since each division in the time series generates a new boundary condition, smoothing interpolation of our data is expected to be little robust.
-For this reason, the growth rate of single cells is usually estimated by fitting an exponential function to the volume data between each cell~\cite{wang_robust_2010,izard_synthetic_2015}.
-While this is suitable when bacteria are at steady state, it is not applicable during a growth transition, where we expect the growth rate to vary between successive divisions.
-Other techniques are less sensible to the above problem~\cite{zulkower_robust_2015}, and we decided to use Kalman smoothing for our purpose~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+Since each division in the time series is a boundary, the smoothing interpolation of this kind of data is particularly unstable.
+For this reason, single-cell growth rate is usually estimated by fitting an exponential term on the volume between each division (or in particular a linear term on the logarithm of the volume)~[citation Wang et al et Izard et al].
+While it is suitable when bacteria are at steady state, it is not applicable in our case where we expect growth-rate variations between divisions.
+Other techniques are less sensible to this kind of problem~\cite{zulkower_robust_2015}, and we decided to focus on the use of Kalman smoother~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
 
-Kalman filtering~\cite{kalman_new_1960} is a Bayesian algorithm using a series of noisy measurements to predict the state of a dynamical system.
-It has been extensively used in engineering applications (guidance, tracking, control, ...) requiring the real-time estimation of hidden variables in a dynamical system from present and past measurements.
-Kalman smoothing is an extension of Kalman filtering using information about past and present but also future measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
-The advantage of Kalman filtering with respect to Kalman smoothing is that it uses all available information to estimate the hidden variables of a dynamical system, in applications where a real-time response is not required.
-For our problem, at each time step $t_k$, our hidden variables are the resource allocation variable $\alpha (t_k)$ and the growth rate $\mu (t_k)$, while the available information consists of the $N$ measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$ that were taken during the experiment.
+Kalman filtering~\cite{kalman_new_1960} is a bayesian algorithm that uses a series of noisy measurements to predict the state of a dynamical system.
+It is well known for its everyday technological applications (guidance, tracking, control, ...) which requires the real-time estimation of hidden variables in a dynamical system based on previous measurements.
+Kalman smoothing is an extension of Kalman filtering that uses information about the previous but also the next measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+In that sense, it uses all the available information to estimate the hidden variables of a dynamical system.
+For our problem, at each time step $t_k$, our hidden variables are the resource allocation $\alpha (t_k)$ and the growth rate $\mu (t_k)$, and the available information is the N measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$ that were taken during the whole experiment.
 
-The Kalman filtering problem can now be formulated as follows.
-The state of a growing bacterial cell is described by the dynamical system
+We state the problem as follows.
+We have the following dynamical system:
 \begin{eqnarray}
@@
 \end{eqnarray}
-with initial conditions $r(0) = r_0$, $V(0) = V_0$, and the following measurement model:
+with the following measurement model:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where the variables are as defined as for Eqs~\ref{eq:dot_r}, \ref{eq:mes_L} and \ref{eq:mes_F}.
-While $\beta$ can be obtained from the literature (Chapter~\ref{chap:theory}), $\lambda$ and $\gamma$ are unknown parameters that need to be estimated along with $\alpha (\cdot)$ and $\mu (\cdot)$.
+where each variable has already been defined in Eqs~\ref{eq:dot_r}, \ref{eq:mes_L} and \ref{eq:mes_F}.
+While $\beta$ can be obtained from the literature (see chapter~\ref{chap:theory}), $\lambda$ and $\gamma$ are unknown variables that need to be estimated along with $\alpha$ and $\mu$.
 Since we are satisfied with a qualitative reconstruction of $\alpha (\cdot)$, we can simplify the system by defining $r_\gamma = \gamma \cdot r$ and $V_\lambda = \lambda \cdot V$.
-The dynamical system is consequently rewritten as
+The dynamical system is thus rewritten:
 \begin{eqnarray}
-\dot{r}_\gamma(t) &=& \mu (t) \cdot\frac{\gamma \cdot \alpha(t)}{\beta} - \mu (t) \cdot r_\gamma (t),\label{eq:r_prob}\\
-\dot{V}_\lambda (t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
+\dot{r_\gamma}(t) &=& \mu (t) \cdot\frac{\gamma \cdot \alpha(t)}{\beta} - \mu (t) \cdot r_\gamma (t),\label{eq:r_prob}\\
+\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\label{eq:V_prob}
 \end{eqnarray}
-with initial conditions $r_\gamma (0) = \gamma r_0$, $V_\lambda (0) = \lambda V_0$, and the new measurement model:
+with the new measurement model:
 \begin{eqnarray}
@@
 \end{eqnarray}
-The Kalman filtering problem thus becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ from measurements $\left\{ F(t_0), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), ..., L(t_{N-1}) \right\}$.
+The problem becomes the reconstruction of $\gamma \alpha (\cdot) / \beta$ and $\mu (\cdot)$ through measurements $\left\{ F(t_0), F(t_1), ..., F(t_{N-1}) \right\}$ and $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+Note that this problem is not linear, while Kalman filtering was initially introduced for linear systems~\cite{kalman_new_1960,kailath_linear_2000}.
+For this reason, we used a non-linear extension of Kalman smoothing called Unscented Kalman Smoothing~\cite{julier_new_1997,jazwinski_stochastic_2007}.
+Details about its exact implementation are available in Material and Methods~\ref{sec:meth_kalman}.
+While it is feasible to co-reconstruct the signals $\mu (\cdot)$ and $\gamma \alpha (\cdot) / \beta$ using Kalman smoothing, it is not necessary and could make the reconstruction unstable.
+We thus decided to proceed in two steps: first reconstruct $\mu$ from the measurements of $L$, than inject this result in the reconstruction of $\gamma \alpha / \beta$ from the measurements of $F$.
 
-Note that the above problem is not linear, whereas Kalman filtering was initially introduced for linear systems~\cite{kalman_new_1960,kailath_linear_2000}.
-For this reason, we use a nonlinear extension of Kalman smoothing called unscented Kalman smoothing~\cite{julier_new_1997,jazwinski_stochastic_2007}.
-Details about its exact implementation are available in Material and Methods~\ref{sec:meth_kalman}.
-While it is feasible to simultaneously reconstruct the signals $\mu (\cdot)$ and $\gamma \alpha (\cdot) / \beta$ using Kalman smoothing, it is not necessary in our case and might lead to unstable results.
-We therefore decided to proceed in two steps: first reconstruct $\mu(\cdot)$ from the measurements $L(\cdot)$, than inject this result into the reconstruction of $\gamma \alpha(\cdot) / \beta$ from the measurements $F(\cdot)$.
-
-\subsection{Estimation of growth rate and resource allocation profile}
+\subsection{Estimation of growth rate and gene expression}
 
@@
 As presented in the previous section, we want to reconstruct the growth rate $\mu$ defined by Eq.~\ref{eq:V_prob}, using measurements of the length $L$ defined by Eq.~\ref{eq:L_prob}.
-Note that Kalman smoothing is a procedure that returns the expected mean and covariance of the state of a dynamical system, given the  measurements.
-Therefore, reconstructing $\mu$ requires it to be explicitly included as a state variable of the dynamical system.
-This provides constraints on the variation of $\mu$ that have the effect of a regularization.
-In particular, in the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process.
-The laws describing this process play the role of a prior on the expected regularity of $\mu$ (the more noisy the process, the less constrained the variations of $\mu$).
-In particular, we define $\mu$ as the double integral of a Gaussian white noise $w$:
+Note that Kalman smoothing is a procedure that returns the expected mean and covariance of the state of a dynamical system, given some measurements.
+Thus, reconstructing $\mu$ requires its explicit integration in the variables of the dynamical system.
+In some sense, this tells the system which constraints to apply on the variations of $\mu$, and represents a regularization method.
+Here, we used the following Bayesian regularization approach.
+In the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process.
+The laws of this process play the role of a prior on the expected regularity of $\mu$ (the more noisy the process, the less constrained the variations of $\mu$).
+In particular, we let $\mu$ be the double integral of a Gaussian white noise $w$:
 \[
-\dot{\mu} = v(t), \;\;\;\; \dot{v} = w(t),
+\dot{\mu} = v(t) \;\;\;\; \dot{v} = w(t),
 \]
-where $v$ is an intermediate variable and $w(t)$ is assumed normally distributed with mean 0 and standard deviation $\theta$: $w(t) \sim \mathcal{N}(0, \theta)$.
-The resulting system then becomes
+where $v$ is an intermediate variable, and $w(t) \sim \mathcal{N}(0, \theta)$.
+The full problem becomes:
 \begin{eqnarray}
@@
 \end{eqnarray}
-with the measurement model
+with the measurement model:
 \begin{equation}
-L(t_k) = V_\lambda(t_k) + \epsilon_k,
+L(t_k) = V_\lambda(t_k) + \epsilon_k.
 \end{equation}
-and the initial conditions
-\[
-V_\lambda (0) = V_{\lambda,0}, \;\;\;\; \mu (0) = \mu_0 , \;\;\;\; v(0) = v_0,
-\]
-that may themselves be Gaussian random variables with some given statistics.
 The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable.
-It is also equivalent to other methods like Tikhonov regularization~\cite{tikhonov1977solutions,tikhonov1963solution,de_nicolao_nonparametric_1997}, which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
+It is also equivalent to other methods like Tikhonov regularization~[citation to take from Val pub], which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.
 
@@
 \begin{itemize}
-\item the mean of the initial state $(V_{\lambda,0}, \mu_0, v_0)$,
-\item the covariance of the initial state $(V_{\lambda,0}, \mu_0, v_0)$,
-\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$ (see below),
-\item the variance of the observation noise $\epsilon_k$.
+\item the mean of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the covariance of the initial state for $V_\lambda$, $\mu$ and $v$,
+\item the transition covariance for the derivatives of $V_\lambda$, $\mu$ and $v$,
+\item the observation variance for $L$.
 \end{itemize}
-The mean and covariance of the initial state $(V_{\lambda,0}, \mu_0, v_0)$ represent the knowledge we have of the initial values of the variables.
-For instance, if we know exactly the initial value of the growth rate $\mu$ (\textit{via} independent measurements or literature data), we can use it as mean for $\mu_0$ along with a very small variance.
-This will constraint the signal reconstruction by penalizing deviations from this value at $t=0$.
-On the contrary, if we are very uncertain of the value for $V_{\lambda,0}$ (because cells are not synchronized, and can be in any state when the data acquisition starts), we can use a large variance for this initial condition.
-In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because they are not the result of a stochastic process (the equations that define their derivatives are fully deterministic).
-However, by definition $v$ is the result of a stochastic process of mean 0 and transition variance $\theta^2$.
-This value is crucial and represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
-The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu (\cdot)$.
-Finally, the variance of the observation noise $\epsilon_k$ is simply the expected measurement noise.
-Together, these parameters represents the \textit{prior} information we have on the initial conditions, the smoothness of the reconstructed signal, and the precision of our measurements, in order to reconstruct the growth rate $\mu (\cdot)$.
-As we will see below, we widely use these properties to overcome the difficulties introduced by the discontinuities following cell divisions.
+The means and covariances at the initial state represent the knowledge we have on the initial parameters.
+For instance, if we know exactly the value of $V_\lambda$, $\mu$ and $v$, we can fix the means and use very small covariances, but this "certainty" can be adjusted at will.
+In our framework, there is no transition covariance for $V_\lambda$ and $\mu$, because there are not the result of a stochastic process.
+However, the transition variance $\theta^2$ of the derivative of $v$ is crucial: it represents the intensity of the white noise $w$ that serves as prior for the regularization of $\mu$.
+The smaller its value, the more the variations of $v$ are penalized, hence the smoother the reconstructed $\mu$.
+Finally, the observation variance for $L$ is the expected measurement noise.
+All these parameters represent the prior that we use to reconstruct the growth rate $\mu$.
+As we will see below, we widely use these properties to overcome the difficulties raised by the discontinuities introduced by cell divisions.
 
-The observation of the length $L$ of growing bacterial cells is inherently discontinuous, due to the division of cells at regular time-points (Fig.~\ref{fig:data_acquisition}).
-The estimation problem can therefore not be solved for the experiment as a whole, but only for the time-intervals between two successive division events, when the cell length is expected to change continuously.
-We localize the division events by identifying the time points at which the empirical derivative of the length is below an arbitrary threshold, and correspondingly slice the total duration of the experiment into time-intervals with continuous changes of cell length.
-The simplest way to proceed from there on would be to treat each portion independently.
-However, by doing this we would lose important information, since the growth rates of a mother and its daughter cells are expected to be similar.
-In other word, while not necessary continuous, the growth rate of new-born cells will depend strongly on the growth rate of their mother just before division.
-This can be easily taken into account in the Kalman smoothing procedure by exploiting the prior information about the initial mean and covariance of the system variables.
-In particular, we set the mean of the initial growth rate of a daughter cell equal to the final estimated growth rate of the mother cell, while defining a reasonable variance of the growth rate to allow for uncertainty.
-The results of this Kalman smoothing procedure, along with the exact parameters used, are reported in Fig.~\ref{fig:growth_rate_estimation} and Material and Methods~\ref{sec:meth_kalman}.
+Since we track bacteria for a long time, the observation of their length $L$ is inherently discontinuous, due to symmetrical divisions of the \textit{E. coli} cells (Fig.~\ref{fig:data_acquisition}).
+Thus, the problem cannot be solved for the whole time series at once, but has to be divided into smaller continuous regions between two divisions.
+We simply localize the divisions by looking at the time points at which the empirical derivative of the length is below an arbitrary threshold, and slice the curve into the appropriate number of continuous time series.
+From here, the most straightforward method would be to treat each portion independently.
+However, we loose an important information by doing so: neighbor portions of the curve are likely to have similar growth rate.
+This comes from the fact that we can expect a daughter cell to start growing with a growth rate similar to its mother cell.
+In other word, while not necessary continuous, the growth rate at the boundaries of a time region can be expected to depend strongly on the growth rate of neighbor regions.
+This can be easily taken into account in the Kalman smoothing procedure by exploiting the priors about the initial mean and covariance of the system variables.
+We simply let the initial state mean of the growth rate be the last estimated value in the mother cell, while defining a reasonable variance around this value to allow flexibility.
+When no data about the mother cell is available (at the beginning of the experiment), we simply let the algorithm choose freely the initial growth rate by defining a large variance around the expected growth rate on acetate.
+The results of this procedure, along with the exact parameters used for the Kalman smoothing, are reported in Fig.~\ref{fig:growth_rate_estimation} and Material and Methods~\ref{sec:meth_kalman}.
 
@@
 \caption{
-\textbf{Growth-rate estimation using Kalman smoothing based on measurement of the length of bacteria growing in a microfluidic device.}
-Gray lines represent the estimation of the time-varying length (upper plot) and growth rate (lower plot) of 68 cells by the unscented Kalman smoothing procedure.
-The solid red lines highlight the result for one particular cell, located at the bottom of the well labelled "X3Y2, W2", while black crosses in the top graph are the data points for this cell.
-The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
+\textbf{Growth rate estimation based on the bacterial length using Kalman smoothing.}
+Grey lines represent estimation of states of the system by the unscented Kalman smoothing procedure for the 68 cells.
+The reconstruction for the bacterial length and the growth rate are presented in the top and bottom graphs, respectively.
+For readability, the solid red lines highlight the result for a particular cell, while black crosses in the top graph are the data points for this cell.
+The dashed grey lines represent the switch time from acetate to glucose.
 As prior for the algorithm, we used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
-The transition variance $\theta^2$ (\textit{i.e.}, the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6}.
-Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean growth rate equal to the last estimated value for $\mu$ before cell division, and to half the last estimated value for $V_\gamma$, bearing in mind that \textit{E.
-coli} cells divide symmetrically.
-At the start of the experiment, when no mother cell is available to provide initial estimates, the above values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
+Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
+At the beginning of the experiment, were no mother cell is available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
 The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
-The initial mean of $v$ is set equal to 0, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
+$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
 All the cross-covariances are set to 0 because the system variables are independent by construction.
+The parameters cited above were chosen through trials and errors, and are as a consequence largely optimizable.
+Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
 }
@@
 
-On average, the results presented in~Fig.~\ref{fig:growth_rate_estimation} show a roughly constant growth rate on acetate, followed by a quick increase after the upshift.
-Contrary to what was visible in the plot with the fluorescence data, a steady state for the growth rate seems to have been reached before the end of the experiment.
-When considering the individual cells, the results are more difficult to interpret.
-A significant part of the cells stopped growing before the end of the experiment, probably due to the fact that all cells die from an unknown cause around the end of the experiment. 
-As a consequence, the analysis has been limited to a time-interval after the upshift in which roughly 3/4 of the cells are still growing.
-Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a 1~h-long pause around 200 min after the upshift, followed by a recovery of the growth rate until the end of the experiment (see Supporting Information XX).
-In order to focus the analysis, we decided to classify the cells in three categories: dying cells (N=12) for which the growth rate drops to zero after the upshift, pausing cells (N=11) for which the growth rate drops to zero after the upshift and then recovers, and finally so-called normal cells (N=45) which do not exhibit any of the above behaviors.
+On the mean, the results presented in~Fig.~\ref{fig:growth_rate_estimation} show a roughly constant growth rate on acetate, followed by a quick increase around the upshift.
+Contrary to what was visible on the fluorescence, a steady state for the growth rate seems to have been reached before the end of the experiment.
+When taken individually, the results are more fuzzy.
+A significant part of the cells stops growing before the end of the experiment.
+This was expected since all the cells die from an unknown cause at the end of the time series, and the interval of analysis was visually chosen so that roughly 3/4 of the cells are still growing.
+Interestingly, 1/6 of the cells exhibit growth rate oscillations, with a hour-long pause around 200 minutes after the upshift, then a growth rate recovery until the end of the time series (see Supporting Information XX).
+For now, we decided to classify the cells in three categories: the dying cells (N=12) for which the growth rate drop to zero after the upshift; the pausing cells (N=11) for which the growth rate drop to zero after the upshift then recover, and finally the normal cells (N=45) which do not exhibit any of the above behaviors.
 More information about each category is available in Supporting Information XX.
-In what follows, we focus on the 45 normal cells (Fig.~\ref{fig:growth_rate_estimation_median}) which show a globally coherent behavior in the time window considered here.
+In what follows, we focus on the 45 normal cells (Fig.~\ref{fig:growth_rate_estimation_median}), which present a globally coherent behavior until the end of the interval of analysis.
 
@@
 \caption{
-\textbf{Growth-rate estimation using Kalman smoothing for the normal cells only.}
-The top graph is the same as the bottom graph of Fig.~\ref{fig:growth_rate_estimation}, except that the dying and pausing cells were removed.
-The bottom graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step.
+\textbf{Growth rate estimation for the normal cells only.}
+The top graph is the same as the bottom graph of Fig.~\ref{fig:growth_rate_estimation}, except we removed the dying and pausing cells.
+The bottom graph gives the 25\% (lowest gray line), 50\% (solid black line) and 75\% (highest gray line) quartiles, computed at each time step.
 The gray area represents the interquartile range.
@@
 
-\subsubsection*{Estimation of resource allocation profile}
+\subsubsection*{Estimation of gene activity}
 
-In order to estimate the time-varying allocation of resources after the upshift, we use a similar Bayesian regularization approach as for the estimation of the growth rate.
-We note $u(t) = \gamma \alpha (t) / \beta$ the resource allocation input that we wish to reconstruct.
-With $\hat{\mu}$ the estimation of $\mu$ obtained in the previous section, the full model for the reconstruction of $u (\cdot)$ is given by
+We use a similar Bayesian regularization approach than for the growth rate estimation.
+We note $u(t) = \gamma \alpha (t) / \beta$ the input that we want to reconstruct.
+The full problem for the estimation of $u$ is:
 \begin{eqnarray}
-\dot{r}_\gamma(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
+\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
 \dot{u}(t) &=& v(t),\label{eq:full_u_prob}\\
@@
 with the measurement model
-\begin{equation}
+\begin{equation*}
 F(t_k) = r_\gamma (t_k) + \eta_k,\\
-\end{equation}
-and the initial conditions
-\[
-r_\gamma (0) = r_{\gamma,0}, \;\;\;\; u (0) = u_0 , \;\;\;\; v(0) = v_0,
-\]
-that may themselves be Gaussian random variables with some given statistics.
-Contrary to the model used for the estimation of the growth rate, the system of Eq.~\ref{eq:full_u_prob} is linear.
-Furthermore, we do not have to deal with discontinuities in the measurements, since the fluorescence density $F$ in the cells is continuous between mother and daughter cells (Fig.~\ref{fig:data_acquisition}).
+\end{equation*}
+where $\hat{\mu}$ is the estimation of $\mu$ from the previous section.
+Contrary to the growth rate problem, this system is linear.
+Furthermore, we do not have to deal with discontinuities of the measurement since $F$ is continuous from the mother to the daughter cells (Fig.~\ref{fig:data_acquisition}).
 
-Given the analysis in Chapter~\ref{chap:theory}, we expect $u(\cdot)$ to exhibit bang-bang variations after the upshift.
-This is a discontinuous signal that could be complicated to reconstruct for certain choices of the parameters of the regularization method.
-In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of Chapter~\ref{chap:theory}.
-We estimated the noise characteristics of $F$ using the RFU/pixel measurements from Fig.~\ref{fig:data_acquisition} (Supporting Information XX) and chose parameters that allow the model to reproduce the observed growth rates and fluorescence densities.
-We simulated an upshift from acetate to glucose and tried to estimate $\gamma \alpha(\cdot) / \beta$ from these synthetic data, selecting the \textit{prior} in the Kalman smoothing procedure by trial and error.
-This approach may not have been optimal for our purpose and possible improvements are discussed in Section~\ref{sec:chap3_discussion}.
-The results along with the \textit{prior} used are reported in Fig.~\ref{fig:synthetic_upshift}.
-As expected, like any smoothing method, the algorithm has difficulty in reconstructing stiff variations in the state variables of Eq.~\ref{eq:full_u_prob}.
-The switching profile of the resource allocation input is relatively well captured though, which indicates that the algorithm is in principle capable of reconstructing an on-off strategy (further improvements are discussed in Section~\ref{sec:chap3_discussion}).
-In what follows, we use the same \textit{prior} of the Kalman smoothing procedure for the normal cells identified in the previous section, to test the occurrence of an on-off switching profile in our data.
+To be comparable with what we predicted in chapter~\ref{chap:theory}, we expect $u(\cdot)$ to exhibit bang-bang variations after the upshift.
+By definition, it is a very stiff signal that could be complicated to reconstruct depending on the parameters of the regularization method.
+In order to parametrize the Kalman smoothing algorithm, we generated synthetic data by simulating the model of chapter~\ref{chap:theory}.
+This was not possible for the growth rate problem since cell division is absent from the model.
+After estimating the noise on the RFU/pixel measurements (see Supporting Information XX), we simulated an upshift from acetate to glucose (the parameters were chosen based on the resulting growth rate), and used a value for $\gamma$ that generated a fluorescence concentration in the range of the observations.
+We then tried to estimate $\gamma \alpha / \beta$ based these synthetic data, choosing the prior parameters through trials and errors.
+There are probably not optimal, and we discuss possible improvements in section~\ref{sec:chap3_discussion}.
+The results along with the prior parameters are presented in Fig.~\ref{fig:synthetic_upshift}.
+Like any smooth methods, the algorithm has trouble reconstructing stiff variations of the system variables.
+The oscillatory nature of gene expression is however fully captured, which indicate that the algorithm should be able to conclude on the existence of the \textit{on-off} strategy.
+Some methods that could improve this result are also discussed in section~\ref{sec:chap3_discussion}.
+In what follows, we use the same prior parameters on the "normal cell" subset of the real data, in order to see if we can get the same oscillatory pattern.
 
@@
 \caption{
-\textbf{Performance of the Kalman smoothing procedure on synthetic data simulating an acetate-glucose upshift.}
-\textit{(A)}~Synthetic data simulating an upshift from minimal medium with acetate to minimal medium with glucose, with and without additive white noise, as well as the results obtained by Kalman smoothing.
+\textbf{Performance of the unscented Kalman smoother on synthetic data simulating an upshift.}
+\textit{(A)}~Synthetic data simulating an upshift with and without white noise, along the result of the Kalman smoothing procedure.
 The synthetic data were generated by simulating the model presented in Eqs~\ref{eq:pdef}-\ref{eq:rdef} with the on-off regulatory strategy (Eq.~\ref{eq:stratswitch} and Fig.~\ref{fig:onoffresults}).
-The model parameters used for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $K_R = 1$~g~L\textsuperscript{-1}.
-The predicted $r(t)$ profile was multiplied by a factor $\gamma = 0.02$~L~g\textsuperscript{-1}~RFU\textsuperscript{-1} in order to obtain the corresponding fluorescence intensity profile $F$ (dashed black curve).
-The noise level was estimated from the data (Supporting Information XX) and added to $F$.
-The choice of the parameters of the Kalman smoothing procedure is discussed in the Material and Methods~\ref{sec:meth_kalman}.
-\textit{(B)}~Estimation of the resource allocation profile $\gamma \alpha / \beta$ based on the data in \textit{(A)}.
-Following Chapter~\ref{chap:theory}, $\alpha (t)$ displays a bang-bang-singular profile during the upshift (dashed black curve).
-While the Kalman smoother is not able to capture the discontinuous variations in $\gamma \alpha / \beta$, it qualitatively reproduces the input quite well (red solid curve).
-\textit{(C)}~The predicted growth rate during the upshift experiment.
-This information is used as an input in the smoothing procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
+Parameters for the simulation are $e_{M,\texttt{Ace}} = 0.18$~h\textsuperscript{-1}, $e_{M,\texttt{Glu}} = 0.9$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $\beta = 0.003$~L.g\textsuperscript{-1}, $K_R = 1$~g.L\textsuperscript{-1}.
+The simulated $r(t)$ was multiplied by a factor $\gamma = 0.02$~L.g\textsuperscript{-1}.RFU\textsuperscript{-1} in order to get a synthetic $F$ in the same range of the observations (dashed black curve).
+Additive noise level was estimated from the real data and injected in $F$ (see Supporting Information XX).
+Here again, the parameters were chosen through trials and errors, and are reported in the Material and Methods~\ref{sec:meth_kalman}.
+They are largely optimizable, and possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+\textit{(B)}~Estimation of $\gamma \alpha / \beta$ based on the data in (A).
+As described in chapter~\ref{chap:theory}, $\alpha (t)$ performs a bang-bang-singular response during the upshift (dashed black line).
+While the Kalman smoothing is not able to capture the stiff variations of $\gamma \alpha / \beta$, it fairly reproduces the oscillatory nature of the gene expression (red solid line).
+\textit{(C)}~The growth rate during the upshift.
+This information is used as an input in the procedure, since it is supposed to have been independently estimated from the measurements $\left\{L(t_0), L(t_1), ..., L(t_{N-1}) \right\}$.
 }
@@
 
-The results of the reconstruction of the resource allocation profile $u = \gamma \alpha (\cdot) / \beta$ are presented in Fig.~\ref{fig:gene_activity}.
-Within the interval between -150 and 0 min, resource allocation remains more or less stable, as expected for steady-state growth on acetate.
-On the contrary, the reconstructed resource allocation profile seems particularly unstable at the beginning and the end of the experiment.
-While oscillations do occur after the upshift, these need to be taken with much care.
-The problem of reconstructing an on-off strategy is more challenging than we initially thought, for the simple reason that the expected signal is similar to the kind of artifacts a poorly calibrated regularization method would generate.
-Data about the pre-upshift and post-upshift steady state are crucial for the calibration of the method and, as explained in Section~\ref{sec:data_acquisition}, the experiment had to be interrupted before a steady state on glucose was attained.
-We extensively discuss this and other problems in Section~\ref{sec:chap3_discussion} and give directions for future improvements.
+Results of the reconstruction of $u = \gamma \alpha / \beta$ are presented in Fig.~\ref{fig:gene_activity}.
+Large oscillations are indeed visible on the reconstruction.
+These oscillations were already visible on the RFU/pixel data (Fig.~\ref{fig:gene_activity}A).
+The region between -150 and 0 min appears to be the most stable, as we could expect for a growth steady state.
+On the contrary, the reconstructed $\gamma \alpha / \beta$ seems particularly unstable at the beginning and the end of the experiment.
+Overall, the oscillations observed must be taken with extreme caution.
+The problem of reconstructing an on-off signal is more challenging than we initially thought, for the simple reason that the signal we expect to reconstruct is exactly the kind of artifacts a poorly calibrated regularization method would generate.
+Data about the pre-upshift and post-upshift steady state are more crucial than expected, and the experimental problem we described earlier truly impedes our capacity to conclude on the regulatory strategy.
+We extensively discuss this in section~\ref{sec:chap3_discussion}, and give numerous indications about the future improvements that will be implement.
+Finally, Fig.~\ref{fig:gene_activity_median} gives an idea of the distribution of the estimations at each time step.
+While most oscillations cancels out when we consider global statistics about the population, the first oscillation just after the upshift is conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}B).
+This is rather encouraging and prompts for future investigations.
 
@@
 \caption{
-\textbf{Estimation of the resource allocation profile using Kalman smoothing based on the fluorescence density measurements and the estimated growth rates from Fig.~\ref{fig:growth_rate_estimation}.}
-\textit{(A-B)} Gray lines represent the estimation of the fluorescence density (RFU/pixel) $F(\cdot)$ (in A) and the resource allocation profile $\gamma \alpha / \beta$ (in B) by the Kalman smoothing procedure for the 45 normal cells.
-The solid red curves highlight the result for one particular cell, , located at the bottom of the well labelled "X3Y2, W2", while black crosses in the top graph are the data points for this cell.
-The vertical dashed lines represent the time of the upshift from growth on acetate to growth on glucose.
-The prior values for the parameters of the smoothing algorithm are exactly the same as those used for Fig.~\ref{fig:synthetic_upshift} and are reported in the Material and Methods~\ref{sec:meth_kalman}.
+\textbf{Gene activity estimation based on the RFU/pixel measurements using Kalman smoothing.}
+\textit{(A-B)} Grey lines represent estimation of state of the system by the unscented Kalman smoothing procedure for the 45 normal cells.
+For readability, the solid red lines highlight the result for a particular cell, while black crosses in the top graph are the data points for this cell.
+The dashed grey line represents the switch time from acetate to glucose.
+The reconstruction for the RFU/pixel measurements and the gene activity $\gamma \alpha / \beta$ are presented in the (A) and (B), respectively.
+The prior parameters used for the algorithm are exactly the same than in Fig.~\ref{fig:synthetic_upshift} and are reported in Material and Methods~\ref{sec:meth_kalman}.
 }
@@
 \end{figure}
-
-Nevertheless, when not focusing on the individual cells but looking at the statistics of the entire data set, some interesting patterns emerge. 
-Fig.~\ref{fig:gene_activity_median} shows the median of the time-varying growth rate and resource allocation profile as well as the 25\%-75\% interquartile range. 
-The use of these statistics gives a more robust view on the population level of the response of the cells to a nutrient upshift. 
-When changing the carbon source from acetate to glucose, the growth rate increases to a value of around 0.011 min$^{-1}$, consistent with growth rates reported in the literature for the \textit{E.
-coli} strain used here~\cite{volkmer_condition-dependent_2011,izard_synthetic_2015}, before decreasing when the first cells start to die (Fig.~\ref{fig:gene_activity_median}\textit{C}).
-In addition, the data show one period of an oscillation in the first 3~h after the upshift, conserved in each of the 25\%, 50\% and 75\% quartiles (Fig.~\ref{fig:gene_activity_median}\textit{B}).
-The heatmap in Fig.~\ref{fig:gene_activity_heatmap} reveals that almost all of the normal cells show this oscillatory feature.
-Moreover, the first peak is seen to be even more pronounced on the level of the individual cells, reflecting the fact that in the different cells it occurs at different times after the upshift and is therefore dampened out at the population level.
-While the results of the microfluidic experiment presented here do not allow to confirm the occurrence of an on-off strategy for resource allocation after a nutrient upshift, the preliminary data are encouraging and prompt further investigation.
 
@@
 \caption{
-\textbf{Robust statistics for the estimation results presented in Fig.~\ref{fig:gene_activity}.}
-Each graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step for the signals reconstructed in Fig.~\ref{fig:gene_activity}.
+\textbf{Distribution of the estimation results presented in Fig.~\ref{fig:gene_activity}.}
+Interestingly, while most oscillations in $\gamma \alpha / \beta$ are canceled out at the population level, the first one after the upshift is conserved.
+Each graph gives the 25\% (lowest gray line), 50\% (solid black line) and 75\% (highest gray line) quartiles, computed at each time step for the signals reconstructed in Fig.~\ref{fig:gene_activity}.
 The gray area represents the interquartile range.
-Interestingly, while most oscillations in the resource allocation profile $\gamma \alpha (\cdot) / \beta$ cancel out at the population level, the first peak after the upshift is conserved.
 }
 \label{fig:gene_activity_median}
-\end{figure}
-
-\begin{figure}[tb]
-\centering
-%\includegraphics[scale=1]{./Fig/}
-\caption{
-\textbf{Global overview of all estimated resource allocation profiles presented in Fig.~\ref{fig:gene_activity}.}
-The individual $\gamma \alpha (\cdot) / \beta$  curves have been normalized with respect to their maximum in the interval $[0, 400]$ and then ordered with respect to their maximum in the interval $[0, 100]$.
-}
-\label{fig:gene_activity_heatmap}
 \end{figure}
@@
 
-As we showed in Chapter~\ref{chap:theory}, adopting a dynamical perspective might prove useful to unveil and understand the regulatory strategies employed by microorganisms.
-The criterion of biomass maximization has allowed to account for steady-state empirical growth laws.
-Interestingly, when the same criterion is applied to growth transitions, it predicts that microorganisms allocates their resources through bang-bang control of gene expression.
-For the control of ribosome synthesis, the implementation of this optimal control scheme results in an on-off strategy, where a bacterial cell is either producing only ribosomes or not producing them at all during the adaptation to a new growth medium.
-The ppGpp system, ubiquitous in bacteria and playing an important role in the control of ribosome synthesis in \textit{E. coli}, satisfies the requirements posed by the on-off strategy.
-However, we currently lack experimental data to show that it actually functions in this manner, among other things because of the difficulties to produce well-controlled growth transitions on the single-cell level in the laboratory.
-Can we set up appropriate experiments to measure the expression of ribosomes during a growth transitions?
-Will the bacterial cells exhibit the on-off strategy that was shown to be close to optimal in Chapter~\ref{chap:theory}?
+As we showed in chapter~\ref{chap:theory}, adopting a dynamical perspective might prove useful to unveil and understand the regulatory strategies employed by microorganisms.
+The criteria of biomass optimization has proved useful to uncover and explain the steady-state empirical growth laws.
+Interestingly, when the same criteria are applied in growth transitions, they suggest that microorganisms should converge toward a bang-bang control of gene expression.
+For the control of ribosome synthesis, this would result into implementing an \textit{on-off} regulatory system, either only producing ribosomes, or not producing them at all, during the adaptation to the new growth medium.
+The widespread ppGpp system, known for its role in the control of ribosome synthesis, is theoretically able to implement such a regulatory system.
+But we lack experimental data to show that it does so, mostly because of the difficulties raised by working in growth transitions.
+Can we measure the ribosomal synthesis rate at the single-cell level during well-controlled growth transitions?
+Would it exhibit the \textit{on-off} strategy that was shown to be close to optimal in chapter~\ref{chap:theory}?
 
-In this chapter, the above questions have been addressed by developing an experimental framework which allows the ribosome concentration to be quantified in individual \textit{E.
-coli} cells in real time.
-Our main contribution consists in bringing together several experimental techniques developed in recent years.
-First, inspired by the work of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, we constructed an \textit{E. coli} strain with fluorescent ribosomes.
-In their paper, Bakshi \textit{et al} essentially used this fluorescent marker to localize ribosomes in the cytoplasm through superresolution microscopy.
-While improving their design (Material and Methods~\ref{sec:methods_strain}), we used our strain for the purpose of quantifying ribosomal abundance in single cells.
-Second, we employed the mother machine developed by Wang \textit{et al.}~\cite{wang_robust_2010}, originally developed to study long-term steady-state growth of \textit{E.
-coli}, to observe individual \textit{E.
-coli} cells growing in continuous culture by means of time-lapse fluorescence microscopy.
-Here, the mother machine was used to establish well-controlled growth transitions, \textit{i.e.} from steady-state growth in minimal medium with acetate to steady-state growth in minimal medium with glucose.
-By changing the medium source of the mother machine, we were able to observe how \textit{E. coli} cells adapt their growth rate and the expression of ribosomes.
-Overall, through the combination of fluorescent labeling of a core component of the gene expression machinery and the use of a microfluidic device, we have opened the way to experimentally study growth laws in a dynamical context.
+In this chapter, the above questions have been addressed by developing an experimental framework which allows the relative ribosome concentration to be estimated at the single-cell level in real time on \textit{E.
+coli}.
+Our main contribution consists into bringing together several experimental breakthroughs of the recent years.
+On one hand, inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we reconstructed a \textit{E. coli} strain displaying fluorescent, but still fully effective, ribosomes.
+In their paper, Bakshi \textit{et al} essentially used it to localize ribosomes in the cytoplasm through superresolution microscopy.
+While improving their design (see Material and Methods~\ref{sec:methods_strain}), we used it for the purpose of quantifying the ribosomal abundance.
+On the other hand, we used the mother machine~\cite{wang_robust_2010} developed by Wang \textit{et al} to observe this strain growing in continuous culture through real-time microscopy.
+This microfluidic device was originally developed to study long-term steady-state growth of \textit{E. coli}.
+Here, we used it to observe growth transitions in a well controlled framework, \textit{i.e.} from one steady-state growth to another steady-state growth.
+By changing the medium that was injected into the device, we were able to observe \textit{E. coli} transitioning from growth on acetate to growth on glucose.
+Overall, through the combination of fluorescent labeling of a core component of the gene expression machinery and the use of a microfluidic device, we opened the way to the experimental study of growth laws in a dynamical context.
 
-We applied this experimental set-up to the reconstruction of the resources allocation strategy employed by \textit{E.
-coli} during a nutrient upshift, with the aim to observe the bang-bang profile predicted in Chapter~\ref{chap:theory}.
-We developed a Kalman smoothing procedure adapted to our question and showed how it can be implemented to reconstruct the internal state of a cell from time-lapse microscopy images.
-Similar to hidden Markov chains, Kalman filtering and smoothing are powerful techniques for the reconstruction of unobserved signals from noisy measurements, and have had numerous applications in a variety of domains, though less in biology than in other fields.
-We showed in this chapter that the Bayesian framework of Kalman smoothing can be exploited for growth rate reconstruction from a probabilistic prior defining the relation between mother and daughter cells.
-We also developed another variant of Kalman smoothing for the reconstruction of the resource allocation variable $\alpha (\cdot)$ that is at the heart of the models of Chapter~\ref{chap:theory}.
-Our work on synthetic data showed that, despite the challenge posed by the reconstruction of a discontinuous signal, the algorithm was capable of recovering an oscillatory pattern close to the expected on-off profile.
-Unfortunately, its application to the real data did not allow an unambiguous conclusion to be drawn: while oscillations do occur, and a first peak between 0 and 120 min after the upshift is visible in almost all of the cells, it turned out to be far from straightforward to distinguish a real oscillatory pattern from artifacts of the data analysis and signal reconstruction methods.
-Below we propose several improvements that will be addressed in future work.
+We applied this experimental set-up to the reconstruction of the resources allocation strategy employed by \textit{E. coli} during an environmental upshift.
+Obviously, the goal was to eventually observe the oscillations predicted in chapter~\ref{chap:theory}.
+We applied a signal processing algorithm known as Kalman smoothing and showed how it can be implemented to reconstruct the inner state of a cell from microscopy data.
+Similar to hidden markov chains, Kalman filtering and smoothing are powerful techniques allowing the reconstruction of hidden signals from noisy measurements, and have already numerous everyday applications, though not so much in biology.
+We showed in this chapter that the Bayesian logic behind Kalman smoothing can be exploited to inject in the growth rate reconstruction, a probabilistic prior that define the relation between the mother and its daughter cells.
+We also applied it for the reconstruction of $\alpha$, the abstract resource allocation parameter defined in chapter~\ref{chap:theory}.
+Our work on synthetic data showed that despite being challenged by the high stiffness of the signal, the algorithm was able to fully recover the oscillatory pattern that we expect to occur \textit{in vivo}.
+Unfortunately, its application on our dataset was not so successful: while oscillations can be reproduced, the signal is not strong enough to distinguish whether the oscillations are real or chimerical.
+We propose below several improvements that will certainly be addressed in future work.
 
-Several experimental problems complicated the analysis of the data.
-Sufficiently long steady-state measurements before and after the transition of interest are needed to provide a reliable estimation of the intensity and the nature of the experimental noise.
-They are also critical to calibrate the parameters of the regularization method used for signal reconstruction.
-Despite what was initially planned, we did not completely reach a new steady state after the transition.
+Several experimental problems dramatically impeded the analysis of the data.
+Sufficiently long steady-state measurements before and after the transition of interest are needed to provide a strong estimation of the intensity and the nature of the experimental noise.
+They are also critical to calibrate the parameters of the regularization method used in the signal reconstruction.
+Despite what was initially planned, we did not get long enough steady-state measurements far from the transition.
 Several causes are to blame.
-First, the random stall of the motorization along the Z-axis of the microscope, which does not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
-Second, the massive death of bacteria starting after the transition on glucose.
+First, the random stall of the motorization along the Z-axis of the microscope, which do not allow overnight measurements and occurred during the slow growth on acetate in the presented experiment.
+Secondly, the massive death of bacteria starting after the transition on glucose.
 Future work will focus on solving these issues.
@@
 
-It is also possible to optimize the image analysis techniques used in this study.
+The data analysis technique used is also largely optimizable.
+The results reported here are largely preliminary.
 We segmented the images using the most powerful and ubiquitous segmentation algorithm available: the human brain.
-In other words, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a rectangle around the cell of interest.
-While this is far from satisfying, it may turn out to be difficult to improve upon this, as it seems that the first rule of image analysis is that every application is more or less unique.
+In other word, we manually selected the pixels representing the poles of a single bacteria per well on the fluorescence images, and used this information to arbitrarily define a mask around the cell of interest.
+This is far from satisfying.
+Unfortunately, it seems that the first rule of image analysis is that every application is more or less unique.
 Every algorithm has to make assumptions about the object it is trying to recognize.
-For imaging of microbial cells, it could be the curvature at the poles, the size of the pixels, the homogeneity of the background, the homogeneity of intracellular fluorescence, ...
+For bacteria imaging, it could be the curvature at the poles, the size of the pixels, the homogeneity of the background, the homogeneity of the inner fluorescence...
 While some parameters can be tweaked, some assumptions are always hard-coded in the algorithm and reflect the philosophy used to address the identification problem.
-Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~\cite{fluobactracker} that focuses on adapting their software to our set-up in a near future.
-It should provide a robust, automated way to identify and track the cells, enhancing the reproducibility of the analysis and increasing the number of cells available for computing population statistics.
+Since this preliminary analysis was made, we started a collaboration with the authors of FluoBacTracker~[citation, ask Hugues if a publication exist] that focuses on adapting their software to our set-up in a near future.
+It should provide a robust, automatic way to identify and track the cells, enhancing reproducibility of the analysis.
 
-The Kalman filtering algorithm that was later applied to these data can also be improved in many respects.
-As described in the Results section, the algorithm is an instance of Bayesian inference and several parameters describing the expected input define a probabilistic prior.
-In the work reported here, we chose these parameters through trial and error, or by calibrating the system on synthetic data whenever possible.
-A better technique would be to choose these parameters by \textit{generalized cross-validation} (GCV)~\cite{golub_generalized_1979}, a procedure maximizing the predictive power of the reconstructed signal by reducing overfitting.
-The use of GCV in our framework amounts to choosing the parameters of the Kalman smoother by optimizing predictions on a subset of the cells, then testing the performance on the remaining cells.
-While the initial conditions did not seem to play a huge role in the final shape of the signal, the choice of the smoothing factor $\theta^2$ is critical and may strongly benefit from the proposed GCV extension.
+The signal reconstruction algorithm that was later applied to these data is also largely optimizable.
+Like we described in the previous section, the algorithm is an instance of Bayesian reconstruction.
+Several parameters describing the expected input define a probabilistic prior.
+Preliminary, we chose these parameters through trials and errors, or by calibrating the system on synthetic data when it was possible.
+A better technique would be to choose these parameters by \textit{generalized cross-validation} (GCV)~\cite{golub_generalized_1979}, a procedure that maximizes the predictive power of the reconstructed signal.
+More generally, parameters could be optimized on a subset of the cells, then tested against the remaining subset.
+It would ensure that the algorithm does not overfit the behavior of a few cells.
+While the initial conditions did not seem to play a huge role in the final shape of the signal, the choice of the smoothing factor $\theta^2$ appeared to be critical, and would strongly benefit from such an implementation.
 
-The Kalman smoothing algorithm itself can also be further improved, notably by addressing the difficulty of estimating abrupt transitions.
-As can be seen in Fig.~\ref{fig:synthetic_upshift}, the resource allocation profile $\gamma \alpha (\cdot) / \beta$ reconstructed from the synthetic data increases before the nutrient shift and this artifact also occurs when using the measured fluorescent densities (Fig.~\ref{fig:gene_activity_median}), the change in resource allocation preceding the nutrient upshift by several dozens of minutes.
-While we know exactly when the change of medium occurs, this information is currently not used as a prior for the reconstruction.
-The Kalman smoother can be improved by implementing time-varying parameters for the regularization method, so as to strongly penalize variations of the reconstructed signal at steady state and to release this constraint following the change in medium.
+We also did not fully exploited the power of Kalman smoothing.
+Smoothing methods have trouble estimating abrupt transitions, and our implementation has the same issue.
+As we can see in Fig.~\ref{fig:synthetic_upshift}, the reconstructed signal for $\gamma \alpha / \beta$ increases before the shift actually happen.
+This artifact also appears for the real data (Fig.~\ref{fig:gene_activity_median}), were the signal start to change dozens of minutes before the shift.
+We know exactly when bacteria start to be in contact with the new medium, but this information is never used as prior for the reconstruction.
+Using this information is however possible with Kalman smoothing.
+By implementing time-varying parameters for the regularization method, we can strongly penalize the variations of the reconstructed signal at steady state (when nothing is expected to happen) and release the constraints as soon as the medium is changed (when we expect most of the stuff to happen).
+Such a method has already been successfully applied to correctly estimate changes in growth rate during sugar exhaustion~[citation in preparation Eugenio].
 
-The improvements of the approach proposed in this section, which would help in reaching a conclusion about the existence of the on-off strategy, are not very complicated to realize.
-Another improvement consists in taking into account the autofluorescence of the bacteria, which we assumed to be negligible, mostly because its proper estimation in a dynamical, single-cell context is complicated (Material and Methods~\ref{sec:cell_segmentation}).
-A simple trick, however, could help in quantifying autofluorescence: mixing wild-type bacteria with bacteria having fluorescent ribosomes when loading the device.
-This would allow some channels to be occupied by wild-type cells, in addition to the  channels in which the strain of interest is growing, and thus obtain a reliable estimation of the level of autofluorescence of the cells.
-Notice though that the propertion between the two types of bacteria need to be carefully adjusted in order to preserve enough cells for the estimation of the resource allocation profile.
+We are convinced the improvements cited above would allow to robustly conclude on the existence of the on-off strategy, and are not excessively hard to implement.
+Other more challenging developments are planed and would probably benefit to the framework as a whole.
+For instance, we did not discuss the autofluorescence of the bacteria in this chapter.
+It was assumed negligible, mostly because its proper estimation in a dynamical, single-cell context is complicated (see Material and Methods~\ref{sec:cell_segmentation}).
+Nevertheless, a simple trick could solve the issue: merging wild-type and fluorescent bacteria when inseminating the device.
+The goal would be to have a significant part of the channels occupied by wild-type cells, while still being able to observe in parallel the strain of interest.
+That way, we could have an exact estimation of the level of autofluorescence of the cells, and robustly correct it for the channels of interest.
+Note that this would reduce the amount of exploitable channels, hence would probably require several adjustments.
 
-%[I WOULD SKIP THIS FINAL PARAGRAPH OR REDUCE IT TO ONE PHRASE AND ADD IT TO THE DISCUSSION IN THE PARARAPH STARTING WITH "The Kalman smoothing algorithm..."]
-%Something is also displeasing in the current implement of the Kalman smoothing method.
-%As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
-%This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
-%We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
-%There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
+Something is also displeasing in the current implement of the Kalman smoothing method.
+As described in the main text, the Kalman smoothing procedure not only use previous measuments to estimate the state of the system, but also the next available measurements.
+This is a feature we totally lose in the estimation of the growth rate, because the discontinuities introduced by the cell divisions impede a global analysis of the time series.
+We partially solved the issue by using information about the mother cell at the beginning of the growth rate reconstruction of the daughter cells, but we did not find a way to implement it backward.
+There is probably a better way to deal with cell division in the analysis of microscopy data, and we are actively working on the matter.
 
@@
 \subsection{Bacterial strain construction}
+
 \label{sec:methods_strain}
 
-To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} to the C-terminus of \textit{rpsB} (the gene coding the ribosome subunit S2).
-This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.~coli}~\cite{bakshi_superresolution_2012}.
+In order to quantify the concentration of ribosomes in the cell in
+real time, we constructed a strain  containing a translational fusion
+of \textit{gfp\_mut2}~\cite{zaslaver_comprehensive_2006} to the
+C-terminus of \textit{rpsB}, the gene coding for the ribosome subunit
+S2. The design was inspired by the work of Bakshi \textit{et al}.,
+who used a similar construction to measure the spatial distribution
+of ribosomes in living \textit{E.~coli}~cells \cite{bakshi_superresolution_2012}.
+However, contrary to Bakshi \textit{et al.}, and in order not to create
+any interference with normal gene expression in the modified cells,
+we decided to keep the intergenic region between \textit{rpsB} and
+the downstream gene, \textit{tsf}, unchanged. Since the transcription
+factor \textit{tsf} is under the control of the same promoter as \textit{rpsB},
+leaving a selection marker downstream of \textit{rpsB} would interfere
+with the proper expression of \textit{tsf}. Our strategy comprises
+two steps: (i)~creation of the translational fusion by selecting
+an antibiotic resistance marker and (ii)~removal of the selection
+marker. The strategy is outlined in Figure~\ref{fig:Cloning}.
 
-The transcription factor \textit{tsf} is under the control of the same promoter than \textit{rpsB}, and is located after the C-terminus.
-In order to ensure that \textit{tsf} expression was not affected by our translation fusion on \textit{rpsB}, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
-This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
+\begin{figure}
+\includegraphics[width=13cm]{/home/geiselmann/Persons/Nils/Thesis/Man/Fig/NilsCloning}\caption{\label{fig:Cloning}Construction of the gfp-tagged ribosome subunit.
+The translational fusion of \textit{gfp\_mut2} with \textit{rpsB}
+was constructed on the chromosome of \textit{E.~coli} (top line)
+by homologous recombination of \textit{gfp\_mut2} followed by a selection
+``cassette'' (second line). A small linker was inserted between
+\textit{rpsB} and \textit{gfp\_mut2} in order to minimize interference
+of the fluorescent protein with ribosome functioning. The selection
+cassette consists of a positive selection marker, the gene coding
+for the resistance to the antibiotic kanamycine, and a negative selection
+marker, the gene coding for the toxin CcdB. The latter is transcribed
+from the $p_{BAD}$ promoter, which is only activated in the presence
+of arabinose in the culture medium. Homologous recombination is indicated
+by the grey shaded lines. The resulting recombination product (third
+line) contains the desired fusion protein followed by the selection
+cassette on the chromosome of the bacterium. A second homologous recombination,
+using an oligonucleotide complementary to the end of \textit{gfp\_mut2}
+and the beginning the original region downstream of \textit{rpsB}
+removes the selection cassette. The resulting strain (line four) carries
+the translational fusion of \textit{gfp\_mut2} to \textit{rpsB} without
+any other modification of the chromosome.}
 
-A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
-The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
-This cassette is referred below as \textit{kan-P\textsubscript{BAD}-ccdB}.
+\selectlanguage{english}%
+\selectlanguage{french}%
+\end{figure}
 
-Another DNA fragment containing \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} without ATG was amplified with long primers annealing respectively to the C-terminus of \ref{chap:theory}rpsB (just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-The first primer also contained a 18-bp (base pair) linker that was chosen according to the method section of~\cite{bakshi_superresolution_2012}.
+The DNA fragment containing the selection cassette was amplified using
+two long primers annealing, respectively, to the region just before
+the STOP codon of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
+The selection cassette contains an antiobiotic resistance gene, kanamycin
+(positive selection), and a gene coding for the CcdB toxin under the
+control of the P\textsubscript{BAD} promoter (negative selection
+in presence of arabinose). This cassette is referred below as \textit{kan-p\textsubscript{BAD}-ccdB}.
 
-Both fragments (the \textit{gfpmut2} reporter and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette) were assembled using Gibson assembly~\cite{gibson_enzymatic_2009}.
-Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
-A final product of 2683 bp was obtained:
+Another DNA fragment containing \textit{gfp\_mut2}~\cite{zaslaver_comprehensive_2006}
+without the ATG start codon was amplified using long primers annealing,
+respectively, to the C-terminus of \ref{chap:theory}\textit{rpsB}
+(just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB}
+cassette. The first primer also contained a 18-bp (base pair) linker,
+inserting the same six amino-acids that Bakshi had used for their
+construction ~\cite{bakshi_superresolution_2012}.
+
+The two fragments, the \textit{gfp\_mut2} reporter including the linker
+and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette, were assembled
+using Gibson assembly~\cite{gibson_enzymatic_2009} using a commercial
+Gibson Assembly mix (New England Biolabs). A final product of 2683
+bp was obtained: 
 \begin{itemize}
-\item *(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
+\item {*}(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
 \item (18 bp) a linker (see Supporting Information) 
@@
 \item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette 
-\item *(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli}
+\item {*}(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli} 
 \end{itemize}
-Regions labeled with * are expected to anneal with the \textit{E.~coli} chromosome.
-The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
+Regions labeled with {*} anneal to the \textit{E.~coli} chromosome.
+The complete sequence of this fragment, as well as the sequence of
+all the primers,are listed in the Supporting Information of this chapter.
 
-This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid with chloramphenicol resistance (lambda-red recombinaison).
-A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
-A new 100-bp fragment containing 50 bp of the end of \textit{gfpmut2} and 50 bp of the region just after \textit{rpsB} was electroporated into this strain (sequence available in Supporting Information) to get rid of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
-Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
-The region after \textit{rpsB} was verified by sequencing (full sequence available in Supporting Information).
+This fragment was electroporated into the wildtype strain, BW25113,
+containing the pSIM5 plasmid for lambda-red recombinaison \cite{sharan2009recombineering}.
+A kanamycin-resistant colony was selected on LA-glucose medium and
+verified to show green fluorescence in a Tecan microplate reader.
+A 100-bp oligonucleotide, containing 50 bp of the end of \textit{gfpmut2}
+and 50 bp of the region just downstream of \textit{rpsB} was electroporated
+into this strain (sequence available in Supporting Information). The
+homologous recombination removes the \textit{kan-P\textsubscript{BAD}-ccdB}
+cassette. A colony was selected on an LA-arabinose plate. The clone
+was tested for kanamycin-sensitivity and  green fluorescence. Finally,
+the strain was grown overnight at 42$^{\circ}$C to get rid of the
+pSIM5 plasmid, containing a temperature-sensitive origin of replication.
+A chloramphenicol-sensitive colony was chosen andthe region after
+\textit{rpsB} was verified by sequencing (full sequence available
+in Supporting Information).
 
-In parallel, the same protocol was used to construct \textit{mCherry} and \textit{cfp} variants of the same strain.
-However, only the \textit{gfpmut2} and \textit{mCherry} versions were successfully obtained for reasons that were not investigated.
-The full sequence of the final \textit{rpsB-mCherry} strain is available in Supporting Information.
+In parallel, the same protocol was used to construct \textit{mCherry}
+and \textit{cfp} variants of the same strain. However, only the \textit{gfp\_mut2}
+and \textit{mCherry} versions were successfully obtained. The full
+sequence of the final \textit{rpsB-mCherry} strain is available in
+Supporting Information.
 
-The \textit{rpsB-gfpmut2} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
-They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
-However, the \textit{rpsB-mCherry} strain exhibited strange fluorescence dynamics, especially during growth transitions, which made it unsuitable for our study.
-Our effort were then concentrated on the \textit{rpsB-gfp} strain.
-Data as well as information on the matter are available in Supporting Information.
+The \textit{rpsB-gfp\_mut2} and \textit{rpsB-mCherry} strains were
+characterized on different media using a Tecan microplate reader.
+They possess a wild-type growth rate and sufficient fluorescence to
+allow quantification. However, the \textit{rpsB-mCherry} strain exhibited
+strange fluorescence dynamics, especially during growth transitions,
+which made it unsuitable for our study. We therefore concentrated
+our efforts on the \textit{rpsB-gfp} strain. Detaied data about the
+characterization of the strain are available in Supporting Information.
 
 \subsection{Growth conditions}
+
 \label{sec:growth_condition}
 
-Sterilization was performed by autoclaving for instruments and filtration for solutions (0.2 $\mu$m).
-During cloning, bacteria were grown in 20-mL flasks filled with LB or spread on Petri dish with LA.
-In all the experiments, we used M9 minimal medium~[citation to find] supplemented with trace elements, thiamine and a carbon source.
-The full recipe is reproduced below.
+All experiments were carried out in sterilized media. Sterilization
+was performed by autoclaving for instruments and filtration for solutions
+(0.2 $\mu$m). During cloning, bacteria were grown in 20-mL Erlenmeyer
+flasks filled with LB or spread on Petri dishes with LA. In all the
+experiments, we used M9 minimal medium supplemented with trace elements,
+thiamine and a carbon source. The full recipe is reproduced below.
 Numbers in squared brackets are characteristics of the stock solution.
-All stock solutions were stored at room temperature, except for FeSO$_4$~[30 g.L\textsuperscript{-1}] and Thiamine~[10 g.L\textsuperscript{-1}] that were stored respectively at -20$^\circ$C and 4$^\circ$C.
+All stock solutions were stored at room temperature, except for FeSO$_{4}$~{[}30
+g.L\textsuperscript{-1}{]} and Thiamine~{[}10 g.L\textsuperscript{-1}{]},
+which were stored respectively at -20$^{\circ}$C and 4$^{\circ}$C. 
 \begin{center}
@@
 \hline 
-  \multicolumn{2}{|c|}{\textbf{M9 medium (100 mL)}}\\
+\multicolumn{2}{|c}{\textbf{M9 medium (100 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
 \hline 
-  CaCl2 [1 mol.L\textsuperscript{-1}] & 10 $\mu$l\\
-  MgSO$_4$ [1 mol.L\textsuperscript{-1}] & 200 $\mu$l\\
-   5x Salts & 20 mL \\
-  Traces elements & 90 $\mu$l \\
-  FeSO$_4$ [30 g.L\textsuperscript{-1}] & 10 $\mu$l \\
-  Thiamine [10 g.L\textsuperscript{-1}] & 50 $\mu$l \\
-  Carbon source & at will \\
-  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+CaCl2 {[}1 mol.L\textsuperscript{-1}{]}  & 10 $\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+MgSO$_{4}$ {[}1 mol.L\textsuperscript{-1}{]}  & 200 $\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+5x Salts  & 20 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Traces elements  & 90 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+FeSO$_{4}$ {[}30 g.L\textsuperscript{-1}{]}  & 10 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Thiamine {[}10 g.L\textsuperscript{-1}{]}  & 50 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Carbon source  & at will  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+\end{tabular}%
+\begin{tabular}{|l r|}
+\hline 
+\multicolumn{2}{|c}{\textbf{Traces elements (0.9 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & 200$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Na$_{2}$EDTA 2H$_{2}$O {[}150 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+ZnSO$_{4}$ 7H$_{2}$O {[}45 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+CoCl$_{2}$ 6H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+MnCl$_{2}$ 4H$_{2}$O {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{3}$BO$_{3}$ {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Na$_{2}$MoO$_{4}$ 2H$_{2}$O {[}4 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+CuSO$_{4}$ 5H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+\end{tabular}%
+\begin{tabular}{|l r|}
+\hline 
+\multicolumn{2}{|c}{\textbf{5x Salts (100 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+Na$_{2}$HPO$_{4}$2H$_{2}$0  & 4.25 g  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+KH$_{2}$HPO$_{4}$  & 1.5 mg  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+NaCl  & 0.25 g & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+NH$_{4}$Cl & 0.5 g & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
 \hline 
 \end{tabular}
-\begin{tabular}{|l r|}
-  \hline
-  \multicolumn{2}{|c|}{\textbf{Traces elements (0.9 mL)}}\\
-  \hline
-  H$_2$0 [18.2 M$\Omega$.cm] & 200$\mu$l\\
-  Na$_2$EDTA 2H$_2$O [150 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  ZnSO$_4$ 7H$_2$O [45 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  CoCl$_2$ 6H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  MnCl$_2$ 4H$_2$O [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  H$_3$BO$_3$ [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  Na$_2$MoO$_4$ 2H$_2$O [4 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  CuSO$_4$ 5H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  \hline
-\end{tabular}
-\begin{tabular}{|l r|}
-    \hline
-  \multicolumn{2}{|c|}{\textbf{5x Salts (100 mL)}}\\
-  \hline
-  Na$_2$HPO$_4$2H$_2$0 & 4.25 g \\
-  KH$_2$HPO$_4$ & 1.5 mg \\
-  NaCl & 0.25 g\\
-  NH$_4$Cl& 0.5 g\\
-  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
-  \hline
-\end{tabular}
-\end{center}
+\par\end{center}
 
-A large quantity of M9 was prepared several days before the experiment, and stored at 4$^\circ$C.
-It contained all the necessary components except FeSO$_4$, thiamine, and carbon sources that were added the day the pre-culture were launched.
-Except stated otherwise, the growth temperature was 37$^\circ$C.
-M9 Acetate contains 0.2\% Acetate (in mass of C$_2$H$_3$O$_2$ per mass of solution), and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per mass of solution).
+A large quantity of M9 was prepared several days before the experiment,
+and stored at 4$^{\circ}$C. It contained all the necessary components
+except FeSO$_{4}$, thiamine, and carbon sources, which were added
+just before inocculation of the pre-culture. Except where stated otherwise,
+the growth temperature was 37$^{\circ}$C. M9 Acetate contains 0.2\%
+Acetate (in mass of C$_{2}$H$_{3}$O$_{2}$ per mass of solution),
+and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per
+mass of solution).
 
-At J-4, the glycerol stock containing the \textit{rpsB-gfp} strain was spread out on a Petri dish of LA, and incubated at 37$^\circ$C.
-At J-3, an isolated colony was inseminated in 20 mL of M9 Acetate, and incubated in a cotton-plugged sterile flask (Fig.~\ref{fig:experiment_schema}).
-Time of insemination was calculated to obtain a 0.3-0.4 cm\textsuperscript{-1} OD when starting the experiment, which correspond to a mid-exponential phase.
+Four days before the measurements, the glycerol stock containing the
+\textit{rpsB-gfp} strain was spread on a Petri dish of LA, and incubated
+at 37$^{\circ}$C. On the following day, an isolated colony was was
+used to inoculate a cotton-plugged, sterile flask containing 20 mL
+of M9 acetate (Fig.~\ref{fig:experiment_schema}). The time of inoculation
+was calculated such that the culture obtains an OD of 0.3-0.4 OD at
+the beginning of the experiment. This optical density corresponds
+to a culture in mid-exponential growth phase.
 
-At J+0, the preculture was concentrated by centrifugation and back suspended into 5 mL of M9 Acetate supplemented with 50 mg.mL\textsuperscript{-1} BSA (passivation	buffer) for injection into the microfluidic device.
-Channels were populated by diffusion until most of them were filled with cells ($\sim$1 hour).
-Data acquisition started after a constant medium flow was successfully obtained ($\sim$1 hour, depending on the quality of the device).
+At the day of the experiment, the preculture was concentrated by centrifugation
+and re-suspended in 5 mL of M9 acetate, supplemented with 50 mg.mL\textsuperscript{-1}
+BSA (passivation buffer) for injection into the microfluidic device.
+Channels were filled by diffusion until most of them contained cells
+($\sim$1 hour). Data acquisition started after a constant medium
+flow was successfully obtained ($\sim$1 hour, depending on the quality
+of the microfluidics device).
 
 \subsection{Microfluidic device}
+
 \label{sec:microflu}
 
-We used the mothermachine~\cite{wang_robust_2010}.
-It consists of a series of growth wells (or channels), oriented at a 90$^\circ$ angle to a large central channel through which growth medium is passed at a constant flow.
-The wideness of the wells is selected as to constrain the growth direction of the bacteria.
-This design ensures that at least one cell per well (the deepest one) is preserved during the whole experiment, while the others incrementally escape into the central channel as divisions occur.
-For the fabrication of the devices, we thoroughly followed the step described in the supporting information of~\cite{wang_robust_2010}.
-We maintained a stock of chemically treated devices at room temperature (day 2 in workflow summary~\cite{wang_robust_2010}).
-The day of the experiment, a single device was plasma cleaned, bond to a glass coverslip, and injected with bacteria (see section above).
+In order to observe the bacteria over a long time, we used the so-called
+mothermachine~\cite{wang_robust_2010}. It consists of a series of
+wells (or channels), oriented at a 90$^{\circ}$ angle to a large,
+central channel through which growth medium is passed at a constant
+flow rate. The width of the wells, a little over a $\mu m$, constrains
+the bacteria to grow ``in a line''. This design ensures that at
+least one cell per well (the one at the bottom of the channel) remains
+in the device during the entire experiment, while the others incrementally
+escape into the central channel as divisions occur. For the fabrication
+of the devices, we thoroughly followed the step described in the supporting
+information of~\cite{wang_robust_2010}. We maintained a stock of
+chemically treated devices at room temperature (day 2 in workflow
+summary~\cite{wang_robust_2010}). The day of the experiment, a single
+device was plasma cleaned, bond to a glass coverslip, and injected
+with bacteria (see section above).
 
-The device was connected using 0.023" inner diameter polyethylene tubes to a waste and a sterile bottle containing 200 mL of growth medium, which was enough for several days of acquisition.
-A microfluidic pump (Elvesys) containing an output flow sensor module was plugged to the medium bottle, and was applied a pressure up to 2 bars.
-The output flow was set to 50 $\mu$L.min$^{-1}$.
+The device was connected using 0.023\textquotedbl{} inner diameter
+polyethylene tubes to a waste and a sterile bottle containing 200
+mL of growth medium, which was enough for several days of acquisition.
+A microfluidics pump (Elvesys) containing an output flow sensor module
+was plugged to the medium bottle and a pressure up to 2~bars was
+applied to the bottle. The output flow was set to 50 $\mu$L.min$^{-1}$.
 
-For the imaging, the device was placed on a motorized inverted microscope (Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^\circ$C.
-In this setup, fluorescence illumination is provided by a mercury lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma, \#49005 TR/DsRED ET).
-The exposure time to the light source is externally controlled by mechanical shutters (Uniblitz-VS35).
-Images are acquired with a 16-bit gray level CCDcamera cooled to -80$^\circ$C (Roper Scientific, Princeton Instruments PHOTOMAX 512) controlled by a custom-made software using Visual Basic and the Type libraries of the Winview software (Princeton Instruments).
-Every 5 minutes, autofocus was numerically performed by maximizing the contrast of a region of interest, and a serie of acquisitions was made.
-A total of 6 fields, each containing 15 wells was observed during the whole experiment.
+For imaging, the device was placed on a motorized inverted microscope
+(Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss
+PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^{\circ}$C.
+In this setup, fluorescence illumination is provided by a mercury
+lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass
+excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma,
+\#49005 TR/DsRED ET). The exposure time is externally controlled by
+mechanical shutters (Uniblitz-VS35). Images were acquired with a 16-bit
+gray level CCDcamera cooled to -80$^{\circ}$C (Roper Scientific,
+Princeton Instruments PHOTOMAX 512) controlled by a custom-made software
+using Visual Basic and the Type libraries of the Winview software
+(Princeton Instruments). Every 5 minutes, autofocus was numerically
+performed by maximizing the contrast of a region of interest, and
+a series of acquisitions were made. A total of 6 fields, each containing
+15 wells was observed during the entire experiment.
 
 \subsection{Cell segmentation}
+
 \label{sec:cell_segmentation}
 
-Raw data are in the form of 512x512-pixel 16-bit SPE images (proprietary format produced by the Princeton Instruments camera).
-They were converted in 16-bit Tif images using the \textit{SPE} plugin of ImageJ~\cite{goto_open_2005}.
-The rest of the analysis was performed using Python 3.5.2.
-In particular, we used OpenCV~3.1.0-dev, Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
+The raw data from the camera are in the form of 512x512-pixel 16-bit
+SPE images (proprietary format produced by the Princeton Instruments
+camera). They were converted into 16-bit Tif images using the \textit{SPE}
+plugin of ImageJ~\cite{goto_open_2005}. The rest of the analysis
+was performed using Python 3.5.2. In particular, we used OpenCV~3.1.0-dev,
+Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
 
-The offset between consecutive images of a given field was evaluated using cross-correlation in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see Scikit-Image documentation at \cite{skimage_cross-correlation}).
-All the images for a field were then aligned on the first acquisition by simple affine translation.
+In order to register consecutive image of a time series, we calculated
+the offset between consecutive images of a given field using cross-correlation
+in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see
+Scikit-Image documentation at \cite{skimage_cross-correlation}).
+All images for a field were thereby aligned to the first acquisition
+by a simple translation.
 
-The wells were isolated in individual images through a combination of manual pixel selection and automatic segmentation.
-In particular, we selected the entrance of the two wells at the border of the image, and used this information along with the regularity of the microfluidic device to compute a mask that allowed to crop each well in its own image of size 100x21 pixels.
-These sub-images were labeled $\left\{W0, W1, ..., W14\right\}$ depending on the position of the well on the original image, from left to right.
+In order to simplify the image treatment, we extracted individual
+wells from the images by a combination of manual pixel selection and
+automatic segmentation. In particular, we selected the entrance of
+the two wells at the border of the image, and used this information
+along with the regularity of the microfluidic device to compute a
+mask that allowed to extract each well into an image of size 100x21
+pixels. These sub-images were labeled $\left\{ W0,W1,...,W14\right\} $
+depending on the position of the well in the original image, from
+left to right.
 
-In this study, we performed a preliminary analysis on each well that focused on the deepest cell.
-For each image, we manually selected two pixels representing the position of the poles of the cell of interest.
-The distance between the two selected pixels was directly used as the bacteria length $L$ in pixels.
-They were also used to compute a rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
-We then computed the RFU/pixel in this cell mask by summing each pixel and dividing by the mask size.
+In this study, we performed a preliminary analysis of each well, focusing
+on the cell at the bottom of the channel. For each image, we manually
+selected the poles of the cell of interest. The distance between the
+two selected pixels was directly used as the length $L$ of the bacterium
+in pixels. The two selected positions were also used to compute a
+rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
+We then computed the RFU/pixel in this cell mask by summing each pixel
+and dividing by the mask size.
 
-The camera noise and background was evaluated by taking a picture with a closed shutter at the end of the experiment.
-Pixels in this picture were found to be Gaussian distributed with a mean of 1101.0 and a standard deviation of 10.788.
-Correction for camera background was thus applied by removing 1101 to the computed RFU/pixel.
+The camera noise and background were evaluated by taking a picture
+with a closed shutter at the end of the experiment. Pixels in this
+picture were found to be Gaussian distributed with a mean of 1101.0
+and a standard deviation of 10.788. A correction for the camera background
+was thus applied by removing 1101 from the computed RFU/pixel.
 
-The background for autofluorescence of the bacteria and the medium was supposed negligible and was thus not corrected.
-Indeed, the same device was used to image bacteria in stationary phase that do not produce any fluorescent protein.
-They were entirely indistinguishable from the camera noise.
-Not that this does not ensure that the autofluorescence is also negligible when the bacteria are actively growing.
-Furthermore, its value could change with the medium of interest, and even be different at steady state and during growth transitions, which does not allow to estimate it independently.
-In the section~\ref{sec:chap3_discussion}, we discuss possible improvements of the experimental set-up that would allow to dynamically co-estimate the autofluorescence with the ribosome abundance in a single experiment.
+The autofluorescence of the bacteria and the background fluorescence
+of the medium were supposed to be negligible and were thus not corrected.
+As a control, the same device was used to image bacteria in stationary
+phase that do not produce any fluorescent protein. They were indistinguishable
+from the camera noise. Of course, using this control we can not affirm
+that the autofluorescence is also negligible when the bacteria are
+actively growing. Furthermore, the autofluorescencecould change in
+different media of interest, and may even be different in steady state
+and during growth transitions. An independent estimate of the autofluorescence
+is therefore difficult to obtain. In the section~\ref{sec:chap3_discussion},
+we discuss possible improvements of the experimental setup that would
+allow to dynamically co-estimate the autofluorescence with the ribosome
+abundance in a single experiment.
 
 \subsection{Kalman smoothing}
+
 \label{sec:meth_kalman}
 
-The data about the length and the RFU/pixel of each bacteria were analyzed using Python~3.5.2.
-In particular, we manipulated the data using Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
+The data of the length and the RFU/pixel of each bacteria were analyzed
+using Python~3.5.2. In particular, we manipulated the data using
+Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
 
-Historical details about the Kalman smoothing procedure are reported in section~\ref{sec:res_kalman} of the main text.
-We used the Additive Unscented Kalman Filter implementation of the pykalman python module.
-This class is reported to be more stable and computationally efficient with non-linear problems featuring additive noises.
+Historical details about the Kalman smoothing procedure are reported
+in section~\ref{sec:res_kalman} of the main text. We used the Additive
+Unscented Kalman Filter implementation of the pykalman python module.
+This class is reported to be more stable and computationally efficient
+with non-linear problems containing additive noise.
 
-As described in section~\ref{sec:res_kalman}, the reconstruction of the growth rate was done on continuous portions of the length, \textit{i.e.} between two division events.
-The full problem, as reported in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity is:
+As described in section~\ref{sec:res_kalman}, the reconstruction
+of the growth rate was done on continuous portions of the length,
+\textit{i.e.} between cell divisions. The full problem, as described
+in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity: 
 \begin{eqnarray*}
-\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
-\dot{\mu}(t) &=& v(t),\\
-\dot{v}(t) &=& w(t),\nonumber
+\dot{V_{\lambda}}(t) & = & \mu(t)\cdot V_{\lambda}(t),\\
+\dot{\mu}(t) & = & v(t),\\
+\dot{v}(t) & = & w(t),
 \end{eqnarray*}
 with the measurement model: 
-\begin{equation*}
-L(t_k) = V_\lambda(t_k) + \epsilon_k.
-\end{equation*}
-The parameters used as prior for the reconstruction of $\mu$ are reported below.
-We used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
-The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
-Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
-At the beginning of the experiment, were no mother cell are available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
-The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
-$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
-All the cross-covariances are set to 0 because the system variables are independent by construction.
+\[
+L(t_{k})=V_{\lambda}(t_{k})+\epsilon_{k}.
+\]
+The parameters used as priors for the reconstruction of $\mu$ are
+described below. We used an observation variance of 9 pixels\textsuperscript{2}
+for the length $L$. The transition variance $\theta^{2}$ (\textit{a.k.a.}
+the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6}
+for the entire time series. Inheritance between mother and daughter
+cells is taken into account by systematically choosing an initial
+mean equal to the last estimated value for $\mu$, and to half the
+last estimated value for $V_{\gamma}$ (as expected for a completely
+symmetrical division). At the beginning of the experiment, when no
+mother cells are available, these values were fixed to 15~pixels
+for $V_{\gamma}$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2}
+and $10^{-4}$~min\textsuperscript{-2}, respectively, for $V_{\gamma}$
+and $\mu$. The initial mean of $v$ is always taken as null, with
+an initial variance of $10^{-8}$~min\textsuperscript{-6}. All the
+cross-covariances are set to 0 because the system variables are independent
+by construction.
 
-For the reconstruction of gene activity, we only had to cope with the large gap in data acquisition between -720 and -150 min.
-Reconstruction was then done independently on the continuous sections before and after this gap.
-The full problem for the estimation of $u = \gamma \alpha / \beta$, as reported in Eq.~\ref{eq:full_u_prob}, is reproduced here for clarity:
+For the reconstruction of gene activity, we had to cope with the large
+gap in data acquisition between -720 and -150 min. Reconstruction
+was then done independently on the continuous sections before and
+after this gap. The full problem for the estimation of $u=\gamma\alpha/\beta$,
+as described in Eq.~\ref{eq:full_u_prob}, is reproduced here for
+clarity: 
 \begin{eqnarray*}
-\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
-\dot{u}(t) &=& v(t),\\
-\dot{v}(t) &=& w(t),\nonumber
+\dot{r_{\gamma}}(t) & = & \hat{\mu}(t)\cdot u(t)-\hat{\mu}(t)\cdot r_{\gamma}(t),\\
+\dot{u}(t) & = & v(t),\\
+\dot{v}(t) & = & w(t),
 \end{eqnarray*}
 with the measurement model 
-\begin{equation*}
-F(t_k) = r_\gamma (t_k) + \eta_k,\\
-\end{equation*}
-where $\hat{\mu}$ is the estimation of $\mu$ above.
-In both case, we used as prior an observation variance of 800.4 RFU\textsuperscript{2} for $F$.
-The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\gamma \alpha / \beta$) is fixed at $10^2$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
-The initial state means used are 600~RFU for $r_\gamma$ and 1000~RFU for $\gamma \alpha / \beta$.
-The variances associated with these means are purposely large and fixed to 10\textsuperscript{6}~RFU\textsuperscript{2} for $r_\gamma$ and $\gamma \alpha / \beta$.
-$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2}, imposing a null second derivative for the reconstructed signal $\gamma \alpha / \beta$.
+\[
+F(t_{k})=r_{\gamma}(t_{k})+\eta_{k},
+\]
+where $\hat{\mu}$ is the estimate of $\mu$ above. In both case,
+we used as prior an observation variance of 800.4 RFU\textsuperscript{2}
+for $F$. The transition variance $\theta^{2}$ (\textit{a.k.a.} the
+smoothing factor for $\gamma\alpha/\beta$) is fixed at $10^{2}$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
+The initial state means used are 600~RFU for $r_{\gamma}$ and 1000~RFU
+for $\gamma\alpha/\beta$. The variances associated with these means
+are purposely large and fixed at 10\textsuperscript{6}~RFU\textsuperscript{2}
+for $r_{\gamma}$ and $\gamma\alpha/\beta$. The initial mean of $v$
+is always taken as null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2},
+imposing a null second derivative for the reconstructed signal $\gamma\alpha/\beta$.
 Here again, all the cross-covariances are set to 0.
 
-All the parameters cited in this section were chosen through trials and errors, and are as a consequence largely optimizable.
-Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+All the parameters cited in this section were chosen through trial
+and error, and aretherefore certainly optimizable. Possible improvements
+are discussed in section~\ref{sec:chap3_discussion}.
 
@@
 
-\subsection{S5 Text -- Strain construction}
-\manuallabel{S5_Text}{S5~Text}
+(Plan not complete)
 
-\subsubsection{Exhaustive list of the primers used...}
+\subsection{Strain construction}
 
-\paragraph{...for the \textit{gfpmut2} amplification.}
+Primers used for gibson assembly
 
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/gfpmut2_L.png} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGAGTAAAGGAGAAGAACTTTTCACTG}
-(Length: 93)
-\item[] \includegraphics[scale=1]{Fig/gfpmut2_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTATAGTTCATCCATGCC} (Length: 44)
-\end{itemize}
-
-\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
-The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-
-\paragraph{...for the \textit{mCherry} amplification}
-
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/mCherry_L.png} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGACTAGCAAAAGATCCAAGGG}
-(Length: 88)
-\item[] \includegraphics[scale=1]{Fig/mCherry_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTACAGCTCATCCATG} (Length: 42)
-\end{itemize}
-
-\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
-The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-
-\paragraph{...for the cassette amplification}
-
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAACTATACAAATAATTATATTCCCCAGAACATCAGG}
-(Length: 42, used to assemble with gfp)
-\item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAGCTGTACAAATAATTATATTCCCCAGAACATCAG}
-(Length: 41, used to assemble with mCherry)
-\item[] \includegraphics[scale=1]{Fig/cassette_R} \seqsplit{GAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTAGAAGAACTCGTCAAGAAGG}
-(Length: 72, used for both)
-\end{itemize}
-
-\textbf{Rem.} The 20 first bases of the first forward primer are supposed to anneal with the end of the \textit{gfpmut2} gene.
-The 20 first bases of the second forward primer are supposed to anneal with the end of the \textit{mCherry} gene.
-The 50 last bases of the reverse primer are supposed to anneal with the region after \textit{rpsB} on the \textit{E.~coli} BW25113 chromosome.
-
-\paragraph{...for the whole insert amplification}
-
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/overlap_L} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGG}
-(Length: 24)
-\item[] \includegraphics[scale=1]{Fig/overlap_R} \seqsplit{GAGCTTGCCGCCTTTCTGCA} (Length: 20)
-\end{itemize}
-
-\paragraph{...for the cassette elimination}
-
-\begin{itemize}
-\item[\textbf{\textit{rpsB-gfpmut2}}] \seqsplit{TTGTAACAGCTGCTGGGATTACACATGGCATGGATGAACTATACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
-(Length: 100)
-\item[\textbf{\textit{rpsB-mCherry}}] \seqsplit{GCGCGGAGGGTCGTCATTCTACCGGTGGCATGGATGAGCTGTACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
-(Length: 100)
-\end{itemize}
-
-\paragraph{...for the sequencing of the final strain}
-
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/sequencing_L} \seqsplit{CGTCTGAAAGACCTGGAAAC}
-(Length: 20)
-\item[] \includegraphics[scale=1]{Fig/sequencing_R} \seqsplit{AAACGTGTACTACCTGGTCTATAAGG} (Length: 26)
-\end{itemize}
-
-\subsubsection{Full sequence of the inserts}
-
-The following pages contain the full sequences of the \textit{gfpmut2-cassette} and \textit{mCherry-cassette} constructions, inserted on the \textit{E.~coli} BW25113 chromosome at the region after the \textit{rpsB} gene.
-The primers annealing positions are indicated using the same color code as in the description of the primers above.
-The primers reported in the original publication of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012} are indicated for information.
-
-\includepdf[pages={1-7}]{./Fig/gfp-mut2-cassette-after-rpsB-gene-sequence.pdf}
-\includepdf[pages={1-6}]{./Fig/mcherry-cassette-after-rpsB-gene-sequence.pdf}
-
-\subsubsection{Sequencing of the final strains}
-
-The strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}.
-Regions at the end of the \textit{rpsB} gene were amplified, and sent for sequencing.
-Alignments of the results showed they exhibit the expected sequences.
-The raw results of the sequencing are reproduced below.
-
-\paragraph{Final sequences of the \textit{rpsB-gfpmut2} strain (region after \textit{rpsB})}
-
-\begin{small}
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/sequencing_L} \seqsplit{AGAAAGAAGCGCTGTATGCGCACTCGTGAGCTGGAGAAACTGGAAAACAGCCTGGGCGGTATCAAAGACATGGGCGGTCTGCCGGACGCTCTGTTTGTAATCGATGCTGACCACGAACACATTGCTATCAAAGAAGCAAACAACCTGGGTATTCCGGTATTTGCTATCGTTGATACCAACTCTGATCCGGACGGTGTTGACTTCGTTATCCCGGGTAACGACGACGCAATCCGTGCTGTGACCCTGTACCTGGGCGCTGTTGCTGCAACCGTACGTGAAGGCCGTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGCGTAAAGGAGAAGAACTTTTCACTGGAGTTGTTCCAATTCTTGTTGAATTAGATGGTGATGTTAATGGGCACAAATTTTCTGTCAGTGGAGAGGGTGAAGGTGATGCAACATA}
-\item[] \includegraphics[scale=1]{Fig/sequencing_R} \seqsplit{CGCCGCAGATGCGTTATCTTCGCTCGCTCATCCCGGTCACTTACTGATGTAAGCTCCCGGGAATTCTCGAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTATTTGTATAGTTCATCCATGCCATGTGTAATCCCAGCAGCTGTTACAAACTCAAGAAGGACCATGTGGTCTCTCTTTTCGTTGGGATCTTTCGAAAGGGCAGATTGTGTGGACAGGTAATGGTTGTCTGGTAAAAGGACAGGGCCATCGCCAATTGGAGTATTTTGTTGATAATGGTCTGCTAGTTGAACGCTTCCATCTTCAATGTTGTGTCTAATTTTGAAGTTAACTTTGATTCCATTCTTTTGTTTGTCTGCCATGATGTATACATTGTGTGAGTTATAGTTGTATTCCAATTTGTGTCCAAGAATGTTTCCATCTTCTTTAAAATCAATACCTTTTAACTCGATTCTATTAACAAGGGTATCACCTTCAAACTTGACTTCAGCACGTGTCTTGTAGTTCCCGTCATCTTTGAAAAATATAGTTCTTTCCTGTACATAAACCTTCGGGCATGGCACTCTTGAAAAAGTCATGCTGTTTCATATGATCTGGGTATCTCGCAAAGCATTGAAGACCATACGCGAAAAGTAGTGACAAGTGTTGGCCATGGAACAGGTAGTTTTCCAGTAGTGCAAATAAATTTAAGGGTAAAGTTTTCCGTATGTTGCATCACCTTCACCCTCTCCACTGACAGAAAAATTTGTGCCCATTTAACATCACCATCTAATTCAACAAGAATTGGAAACAACTCCAGTGAAAGT}
-\end{itemize}
-\end{small}
-
-\newpage
-\paragraph{Final sequences of the \textit{rpsB-mCherry} strain (region after \textit{rpsB})}
-
-\begin{small}
-\begin{itemize}
-\item[] \includegraphics[scale=1]{Fig/sequencing_L}
-\seqsplit{TTTCGACAGCTGACCAAGAAGAAGCGCTGATGCGCACTCGTGAGCTGGAGAAACTGGAAA
-ACAGCCTGGGCGGTATCAAAGACATGGGCGGTCTGCCGGACGCTCTGTTTGTAATCGATG
-CTGACCACGAACACATTGCTATCAAAGAAGCAAACAACCTGGGTATTCCGGTATTTGCTA
-TCGTTGATACCAACTCTGATCCGGACGGTGTTGACTTCGTTATCCCGGGTAACGACGACG
-CAATCCGTGCTGTGACCCTGTACCTGGGCGCTGTTGCTGCAACCGTACGTGAAGGCCGTT
-CTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGAC
-AGGAGACTAGCAAAAGATCCAAGGGCGAGGAGGATAACATGGCTATCATTAAAGAGTTCA
-TGCGCTTCAAAGTTCACATGGAGGGTTCTGTTAACGGTCACGAGTTCGAGATCGAAGGCG
-AAGGCGAGGGCCGTCCGTATGAAGGCACCCAGACCGCCAAACTGAAAGTGACTAAAGGCG
-GCCCGCTGCCTTTTGCGTGGGACATCCTGAGCCCGCAATTTATGTACGGTTCTAAAGCGT
-ATGTTAAACACCCAGCGGATATCCCGGACTATCTGAAGCTGTCTTTTCCGGAAGGTTTCA
-AGTGGGAACGCGTAATGAATTTTGAAGATGGTGGTGTCGTGACCGTCACTCAGGACTCCT
-CCCTGCAAGATGGCGAGTTCATCTATAAAGTTAAACTGCGTGGTACTAATTTTCCATCTG
-ATGGCCCGGTGATGCAGAAAAAGACGATGGGTTGGGAGGCGTCTAGCGAACGCATGTATC
-CGGAAGATGGTGCGCTGAAAGGCGAAATTAAACAGCGCCTGAAACTGAAAGATGGCGG}
-\item[] \includegraphics[scale=1]{Fig/sequencing_R}
-\seqsplit{TTCGCGCCGCAGATGCGTTATCTTCGCTCGCTCATCCCGGTCACTTACTGATGTAAGCTC
-CCGGGAATTCTCGAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAA
-GCTTATTTGTACAGCTCATCCATGCCACCGGTAGAATGACGACCCTCCGCGCGCTCATAT
-TGCTCTACGATCGTATAATCTTCATTATGAGAGGTGATGTCCAGTTTAATATTCACATTG
-TACGCGCCAGGCAGCTGCACAGGTTTCTTGGCTTTGTACGTGGTTTTCACTTCAGCGTCA
-TAATGGCCGCCATCTTTCAGTTTCAGGCGCTGTTTAATTTCGCCTTTCAGCGCACCATCT
-TCCGGATACATGCGTTCGCTAGACGCCTCCCAACCCATCGTCTTTTTCTGCATCACCGGG
-CCATCAGATGGAAAATTAGTACCACGCAGTTTAACTTTATAGATGAACTCGCCATCTTGC
-AGGGAGGAGTCCTGAGTGACGGTCACGACACCACCATCTTCAAAATTCATTACGCGTTCC
-CACTTGAAACCTTCCGGAAAAGACAGCTTCAGATAGTCCGGGATATCCGCTGGGTGTTTA
-ACATACGCTTTAGAACCGTACATAAATTGCGGGCTCAGGATGTCCCACGCAAAAGGCAGC
-GGGCCGCCTTTAGTCACTTTCAGTTTGGCGGTCTGGGTGCCTTCATACGGACGGCCCTCG
-CCTTCGCCTTCGATCTCGAACTCGTGACCGTTAACAGAACCCTCCATGTGAACTTTGAAG
-CGCATGAACTCTTTAATGATAGCCATGTTATCCTCCTCGCCCTTGGAT} 
-\end{itemize}
-\end{small}
-
-\subsection{S6 Text -- Strain validation}
-\manuallabel{S6_Text}{S6~Text}
+\subsection{Strain validation using growth in microplate reader}
 
@@
 
-\subsection{S7 Text -- Microscopy analysis}
-\manuallabel{S7_Text}{S7~Text}
+\subsection{Microscopy analysis}
 
-\subsubsection{Category of cells}
+Show the complete analysis for 5 cells.
+
+\subsubsection*{Category of cells}
 
@@
 \end{figure}
-
-\subsubsection{Noise estimation}
-
-Blabla
-
-\subsubsection{Complete analysis for 5 cells}
-
-blabla

### 2016-12-30 Chapters/experiments.tex

--- 2016-12-30 Chapters/experiments.tex
@@
 \subsection{Bacterial strain construction}
+
 \label{sec:methods_strain}
 
-To achieve dynamical quantification of ribosome abundance, we designed and constructed a strain containing a translational fusion of \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} to the C-terminus of \textit{rpsB} (the gene coding the ribosome subunit S2).
-This design was inspired by the work of Bakshi \textit{et al}, who used a similar construction to assess the quantitative spatial distribution of ribosomes in living \textit{E.~coli}~\cite{bakshi_superresolution_2012}.
-
-The transcription factor \textit{tsf} is under the control of the same promoter than \textit{rpsB}, and is located after the C-terminus.
-In order to ensure that \textit{tsf} expression was not affected by our translation fusion on \textit{rpsB}, we used a double-selection procedure to get rid of every resistance gene that the construction might introduce.
-This is something that was not done in the original construction from Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}
-
-A DNA fragment containing a double selection cassette was amplified with two long primers annealing respectively to the region just after the STOP codon at the C-terminus of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
-The double-selection cassette contained a resistance gene to Kanamycin (positive selection) and a gene coding for the CcdB toxin under the control of the P\textsubscript{BAD} promoter (negative selection in presence of Arabinose).
-This cassette is referred below as \textit{kan-P\textsubscript{BAD}-ccdB}.
-
-Another DNA fragment containing \textit{gfpmut2}~\cite{zaslaver_comprehensive_2006} without ATG was amplified with long primers annealing respectively to the C-terminus of \ref{chap:theory}rpsB (just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-The first primer also contained a 18-bp (base pair) linker that was chosen according to the method section of~\cite{bakshi_superresolution_2012}.
-
-Both fragments (the \textit{gfpmut2} reporter and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette) were assembled using Gibson assembly~\cite{gibson_enzymatic_2009}.
-Both PCR products were quantified using NanoDrop and mixed in equimolar proportions with a commercial Gibson Assembly Master mix.
-A final product of 2683 bp was obtained:
+In order to quantify the concentration of ribosomes in the cell in
+real time, we constructed a strain  containing a translational fusion
+of \textit{gfp\_mut2}~\cite{zaslaver_comprehensive_2006} to the
+C-terminus of \textit{rpsB}, the gene coding for the ribosome subunit
+S2. The design was inspired by the work of Bakshi \textit{et al}.,
+who used a similar construction to measure the spatial distribution
+of ribosomes in living \textit{E.~coli}~cells \cite{bakshi_superresolution_2012}.
+However, contrary to Bakshi \textit{et al.}, and in order not to create
+any interference with normal gene expression in the modified cells,
+we decided to keep the intergenic region between \textit{rpsB} and
+the downstream gene, \textit{tsf}, unchanged. Since the transcription
+factor \textit{tsf} is under the control of the same promoter as \textit{rpsB},
+leaving a selection marker downstream of \textit{rpsB} would interfere
+with the proper expression of \textit{tsf}. Our strategy comprises
+two steps: (i)~creation of the translational fusion by selecting
+an antibiotic resistance marker and (ii)~removal of the selection
+marker. The strategy is outlined in Figure~\ref{fig:Cloning}.
+
+\begin{figure}
+\includegraphics[width=13cm]{/home/geiselmann/Persons/Nils/Thesis/Man/Fig/NilsCloning}\caption{\label{fig:Cloning}Construction of the gfp-tagged ribosome subunit.
+The translational fusion of \textit{gfp\_mut2} with \textit{rpsB}
+was constructed on the chromosome of \textit{E.~coli} (top line)
+by homologous recombination of \textit{gfp\_mut2} followed by a selection
+``cassette'' (second line). A small linker was inserted between
+\textit{rpsB} and \textit{gfp\_mut2} in order to minimize interference
+of the fluorescent protein with ribosome functioning. The selection
+cassette consists of a positive selection marker, the gene coding
+for the resistance to the antibiotic kanamycine, and a negative selection
+marker, the gene coding for the toxin CcdB. The latter is transcribed
+from the $p_{BAD}$ promoter, which is only activated in the presence
+of arabinose in the culture medium. Homologous recombination is indicated
+by the grey shaded lines. The resulting recombination product (third
+line) contains the desired fusion protein followed by the selection
+cassette on the chromosome of the bacterium. A second homologous recombination,
+using an oligonucleotide complementary to the end of \textit{gfp\_mut2}
+and the beginning the original region downstream of \textit{rpsB}
+removes the selection cassette. The resulting strain (line four) carries
+the translational fusion of \textit{gfp\_mut2} to \textit{rpsB} without
+any other modification of the chromosome.}
+
+\selectlanguage{english}%
+\selectlanguage{french}%
+\end{figure}
+
+The DNA fragment containing the selection cassette was amplified using
+two long primers annealing, respectively, to the region just before
+the STOP codon of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
+The selection cassette contains an antiobiotic resistance gene, kanamycin
+(positive selection), and a gene coding for the CcdB toxin under the
+control of the P\textsubscript{BAD} promoter (negative selection
+in presence of arabinose). This cassette is referred below as \textit{kan-p\textsubscript{BAD}-ccdB}.
+
+Another DNA fragment containing \textit{gfp\_mut2}~\cite{zaslaver_comprehensive_2006}
+without the ATG start codon was amplified using long primers annealing,
+respectively, to the C-terminus of \ref{chap:theory}\textit{rpsB}
+(just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB}
+cassette. The first primer also contained a 18-bp (base pair) linker,
+inserting the same six amino-acids that Bakshi had used for their
+construction ~\cite{bakshi_superresolution_2012}.
+
+The two fragments, the \textit{gfp\_mut2} reporter including the linker
+and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette, were assembled
+using Gibson assembly~\cite{gibson_enzymatic_2009} using a commercial
+Gibson Assembly mix (New England Biolabs). A final product of 2683
+bp was obtained: 
 \begin{itemize}
-\item *(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
+\item {*}(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
 \item (18 bp) a linker (see Supporting Information) 
@@
 \item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette 
-\item *(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli}
+\item {*}(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli} 
 \end{itemize}
-Regions labeled with * are expected to anneal with the \textit{E.~coli} chromosome.
-The complete sequence of this fragment, as well as all the primers used are available in the Supporting Information of this chapter.
-
-This fragment was electroporated in a BW25113 background strain containing the pSIM5 plasmid with chloramphenicol resistance (lambda-red recombinaison).
-A kanamycin-resistant colony was selected and verified to exhibit green fluorescence in a Tecan microplate reader.
-A new 100-bp fragment containing 50 bp of the end of \textit{gfpmut2} and 50 bp of the region just after \textit{rpsB} was electroporated into this strain (sequence available in Supporting Information) to get rid of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
-An arabinose-resistant colony was selected and verified to be kanamycin-sensitive and to still exhibit green fluorescence.
-Finally, the strain was grown overnight at 42$^\circ$C to get rid of the pSIM5 plasmid and a chloramphenicol-sensitive colony was selected.
-The region after \textit{rpsB} was verified by sequencing (full sequence available in Supporting Information).
-
-In parallel, the same protocol was used to construct \textit{mCherry} and \textit{cfp} variants of the same strain.
-However, only the \textit{gfpmut2} and \textit{mCherry} versions were successfully obtained for reasons that were not investigated.
-The full sequence of the final \textit{rpsB-mCherry} strain is available in Supporting Information.
-
-The \textit{rpsB-gfpmut2} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
-They were showed to exhibit a wild-type growth rate and sufficient fluorescence level to allow quantification.
-However, the \textit{rpsB-mCherry} strain exhibited strange fluorescence dynamics, especially during growth transitions, which made it unsuitable for our study.
-Our effort were then concentrated on the \textit{rpsB-gfp} strain.
-Data as well as information on the matter are available in Supporting Information.
+Regions labeled with {*} anneal to the \textit{E.~coli} chromosome.
+The complete sequence of this fragment, as well as the sequence of
+all the primers,are listed in the Supporting Information of this chapter.
+
+This fragment was electroporated into the wildtype strain, BW25113,
+containing the pSIM5 plasmid for lambda-red recombinaison \cite{sharan2009recombineering}.
+A kanamycin-resistant colony was selected on LA-glucose medium and
+verified to show green fluorescence in a Tecan microplate reader.
+A 100-bp oligonucleotide, containing 50 bp of the end of \textit{gfpmut2}
+and 50 bp of the region just downstream of \textit{rpsB} was electroporated
+into this strain (sequence available in Supporting Information). The
+homologous recombination removes the \textit{kan-P\textsubscript{BAD}-ccdB}
+cassette. A colony was selected on an LA-arabinose plate. The clone
+was tested for kanamycin-sensitivity and  green fluorescence. Finally,
+the strain was grown overnight at 42$^{\circ}$C to get rid of the
+pSIM5 plasmid, containing a temperature-sensitive origin of replication.
+A chloramphenicol-sensitive colony was chosen andthe region after
+\textit{rpsB} was verified by sequencing (full sequence available
+in Supporting Information).
+
+In parallel, the same protocol was used to construct \textit{mCherry}
+and \textit{cfp} variants of the same strain. However, only the \textit{gfp\_mut2}
+and \textit{mCherry} versions were successfully obtained. The full
+sequence of the final \textit{rpsB-mCherry} strain is available in
+Supporting Information.
+
+The \textit{rpsB-gfp\_mut2} and \textit{rpsB-mCherry} strains were
+characterized on different media using a Tecan microplate reader.
+They possess a wild-type growth rate and sufficient fluorescence to
+allow quantification. However, the \textit{rpsB-mCherry} strain exhibited
+strange fluorescence dynamics, especially during growth transitions,
+which made it unsuitable for our study. We therefore concentrated
+our efforts on the \textit{rpsB-gfp} strain. Detaied data about the
+characterization of the strain are available in Supporting Information.
 
 \subsection{Growth conditions}
+
 \label{sec:growth_condition}
 
-Sterilization was performed by autoclaving for instruments and filtration for solutions (0.2 $\mu$m).
-During cloning, bacteria were grown in 20-mL flasks filled with LB or spread on Petri dish with LA.
-In all the experiments, we used M9 minimal medium~[citation to find] supplemented with trace elements, thiamine and a carbon source.
-The full recipe is reproduced below.
+All experiments were carried out in sterilized media. Sterilization
+was performed by autoclaving for instruments and filtration for solutions
+(0.2 $\mu$m). During cloning, bacteria were grown in 20-mL Erlenmeyer
+flasks filled with LB or spread on Petri dishes with LA. In all the
+experiments, we used M9 minimal medium supplemented with trace elements,
+thiamine and a carbon source. The full recipe is reproduced below.
 Numbers in squared brackets are characteristics of the stock solution.
-All stock solutions were stored at room temperature, except for FeSO$_4$~[30 g.L\textsuperscript{-1}] and Thiamine~[10 g.L\textsuperscript{-1}] that were stored respectively at -20$^\circ$C and 4$^\circ$C.
+All stock solutions were stored at room temperature, except for FeSO$_{4}$~{[}30
+g.L\textsuperscript{-1}{]} and Thiamine~{[}10 g.L\textsuperscript{-1}{]},
+which were stored respectively at -20$^{\circ}$C and 4$^{\circ}$C. 
 \begin{center}
@@
 \hline 
-  \multicolumn{2}{|c|}{\textbf{M9 medium (100 mL)}}\\
+\multicolumn{2}{|c}{\textbf{M9 medium (100 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
 \hline 
-  CaCl2 [1 mol.L\textsuperscript{-1}] & 10 $\mu$l\\
-  MgSO$_4$ [1 mol.L\textsuperscript{-1}] & 200 $\mu$l\\
-   5x Salts & 20 mL \\
-  Traces elements & 90 $\mu$l \\
-  FeSO$_4$ [30 g.L\textsuperscript{-1}] & 10 $\mu$l \\
-  Thiamine [10 g.L\textsuperscript{-1}] & 50 $\mu$l \\
-  Carbon source & at will \\
-  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+CaCl2 {[}1 mol.L\textsuperscript{-1}{]}  & 10 $\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+MgSO$_{4}$ {[}1 mol.L\textsuperscript{-1}{]}  & 200 $\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+5x Salts  & 20 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Traces elements  & 90 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+FeSO$_{4}$ {[}30 g.L\textsuperscript{-1}{]}  & 10 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Thiamine {[}10 g.L\textsuperscript{-1}{]}  & 50 $\mu$l  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Carbon source  & at will  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+\end{tabular}%
+\begin{tabular}{|l r|}
+\hline 
+\multicolumn{2}{|c}{\textbf{Traces elements (0.9 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & 200$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Na$_{2}$EDTA 2H$_{2}$O {[}150 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+ZnSO$_{4}$ 7H$_{2}$O {[}45 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+CoCl$_{2}$ 6H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+MnCl$_{2}$ 4H$_{2}$O {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{3}$BO$_{3}$ {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+Na$_{2}$MoO$_{4}$ 2H$_{2}$O {[}4 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+CuSO$_{4}$ 5H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+\end{tabular}%
+\begin{tabular}{|l r|}
+\hline 
+\multicolumn{2}{|c}{\textbf{5x Salts (100 mL)}} & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+\hline 
+Na$_{2}$HPO$_{4}$2H$_{2}$0  & 4.25 g  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+KH$_{2}$HPO$_{4}$  & 1.5 mg  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+NaCl  & 0.25 g & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+NH$_{4}$Cl & 0.5 g & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
+H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
+\selectlanguage{french}%
+\tabularnewline
 \hline 
 \end{tabular}
-\begin{tabular}{|l r|}
-  \hline
-  \multicolumn{2}{|c|}{\textbf{Traces elements (0.9 mL)}}\\
-  \hline
-  H$_2$0 [18.2 M$\Omega$.cm] & 200$\mu$l\\
-  Na$_2$EDTA 2H$_2$O [150 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  ZnSO$_4$ 7H$_2$O [45 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  CoCl$_2$ 6H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  MnCl$_2$ 4H$_2$O [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  H$_3$BO$_3$ [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  Na$_2$MoO$_4$ 2H$_2$O [4 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  CuSO$_4$ 5H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
-  \hline
-\end{tabular}
-\begin{tabular}{|l r|}
-    \hline
-  \multicolumn{2}{|c|}{\textbf{5x Salts (100 mL)}}\\
-  \hline
-  Na$_2$HPO$_4$2H$_2$0 & 4.25 g \\
-  KH$_2$HPO$_4$ & 1.5 mg \\
-  NaCl & 0.25 g\\
-  NH$_4$Cl& 0.5 g\\
-  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
-  \hline
-\end{tabular}
-\end{center}
-
-A large quantity of M9 was prepared several days before the experiment, and stored at 4$^\circ$C.
-It contained all the necessary components except FeSO$_4$, thiamine, and carbon sources that were added the day the pre-culture were launched.
-Except stated otherwise, the growth temperature was 37$^\circ$C.
-M9 Acetate contains 0.2\% Acetate (in mass of C$_2$H$_3$O$_2$ per mass of solution), and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per mass of solution).
-
-At J-4, the glycerol stock containing the \textit{rpsB-gfp} strain was spread out on a Petri dish of LA, and incubated at 37$^\circ$C.
-At J-3, an isolated colony was inseminated in 20 mL of M9 Acetate, and incubated in a cotton-plugged sterile flask (Fig.~\ref{fig:experiment_schema}).
-Time of insemination was calculated to obtain a 0.3-0.4 cm\textsuperscript{-1} OD when starting the experiment, which correspond to a mid-exponential phase.
-
-At J+0, the preculture was concentrated by centrifugation and back suspended into 5 mL of M9 Acetate supplemented with 50 mg.mL\textsuperscript{-1} BSA (passivation	buffer) for injection into the microfluidic device.
-Channels were populated by diffusion until most of them were filled with cells ($\sim$1 hour).
-Data acquisition started after a constant medium flow was successfully obtained ($\sim$1 hour, depending on the quality of the device).
+\par\end{center}
+
+A large quantity of M9 was prepared several days before the experiment,
+and stored at 4$^{\circ}$C. It contained all the necessary components
+except FeSO$_{4}$, thiamine, and carbon sources, which were added
+just before inocculation of the pre-culture. Except where stated otherwise,
+the growth temperature was 37$^{\circ}$C. M9 Acetate contains 0.2\%
+Acetate (in mass of C$_{2}$H$_{3}$O$_{2}$ per mass of solution),
+and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per
+mass of solution).
+
+Four days before the measurements, the glycerol stock containing the
+\textit{rpsB-gfp} strain was spread on a Petri dish of LA, and incubated
+at 37$^{\circ}$C. On the following day, an isolated colony was was
+used to inoculate a cotton-plugged, sterile flask containing 20 mL
+of M9 acetate (Fig.~\ref{fig:experiment_schema}). The time of inoculation
+was calculated such that the culture obtains an OD of 0.3-0.4 OD at
+the beginning of the experiment. This optical density corresponds
+to a culture in mid-exponential growth phase.
+
+At the day of the experiment, the preculture was concentrated by centrifugation
+and re-suspended in 5 mL of M9 acetate, supplemented with 50 mg.mL\textsuperscript{-1}
+BSA (passivation buffer) for injection into the microfluidic device.
+Channels were filled by diffusion until most of them contained cells
+($\sim$1 hour). Data acquisition started after a constant medium
+flow was successfully obtained ($\sim$1 hour, depending on the quality
+of the microfluidics device).
 
 \subsection{Microfluidic device}
+
 \label{sec:microflu}
 
-We used the mothermachine~\cite{wang_robust_2010}.
-It consists of a series of growth wells (or channels), oriented at a 90$^\circ$ angle to a large central channel through which growth medium is passed at a constant flow.
-The wideness of the wells is selected as to constrain the growth direction of the bacteria.
-This design ensures that at least one cell per well (the deepest one) is preserved during the whole experiment, while the others incrementally escape into the central channel as divisions occur.
-For the fabrication of the devices, we thoroughly followed the step described in the supporting information of~\cite{wang_robust_2010}.
-We maintained a stock of chemically treated devices at room temperature (day 2 in workflow summary~\cite{wang_robust_2010}).
-The day of the experiment, a single device was plasma cleaned, bond to a glass coverslip, and injected with bacteria (see section above).
-
-The device was connected using 0.023" inner diameter polyethylene tubes to a waste and a sterile bottle containing 200 mL of growth medium, which was enough for several days of acquisition.
-A microfluidic pump (Elvesys) containing an output flow sensor module was plugged to the medium bottle, and was applied a pressure up to 2 bars.
-The output flow was set to 50 $\mu$L.min$^{-1}$.
-
-For the imaging, the device was placed on a motorized inverted microscope (Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^\circ$C.
-In this setup, fluorescence illumination is provided by a mercury lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma, \#49005 TR/DsRED ET).
-The exposure time to the light source is externally controlled by mechanical shutters (Uniblitz-VS35).
-Images are acquired with a 16-bit gray level CCDcamera cooled to -80$^\circ$C (Roper Scientific, Princeton Instruments PHOTOMAX 512) controlled by a custom-made software using Visual Basic and the Type libraries of the Winview software (Princeton Instruments).
-Every 5 minutes, autofocus was numerically performed by maximizing the contrast of a region of interest, and a serie of acquisitions was made.
-A total of 6 fields, each containing 15 wells was observed during the whole experiment.
+In order to observe the bacteria over a long time, we used the so-called
+mothermachine~\cite{wang_robust_2010}. It consists of a series of
+wells (or channels), oriented at a 90$^{\circ}$ angle to a large,
+central channel through which growth medium is passed at a constant
+flow rate. The width of the wells, a little over a $\mu m$, constrains
+the bacteria to grow ``in a line''. This design ensures that at
+least one cell per well (the one at the bottom of the channel) remains
+in the device during the entire experiment, while the others incrementally
+escape into the central channel as divisions occur. For the fabrication
+of the devices, we thoroughly followed the step described in the supporting
+information of~\cite{wang_robust_2010}. We maintained a stock of
+chemically treated devices at room temperature (day 2 in workflow
+summary~\cite{wang_robust_2010}). The day of the experiment, a single
+device was plasma cleaned, bond to a glass coverslip, and injected
+with bacteria (see section above).
+
+The device was connected using 0.023\textquotedbl{} inner diameter
+polyethylene tubes to a waste and a sterile bottle containing 200
+mL of growth medium, which was enough for several days of acquisition.
+A microfluidics pump (Elvesys) containing an output flow sensor module
+was plugged to the medium bottle and a pressure up to 2~bars was
+applied to the bottle. The output flow was set to 50 $\mu$L.min$^{-1}$.
+
+For imaging, the device was placed on a motorized inverted microscope
+(Zeiss Axiovert 200M) with a phase contrast objective lens (Zeiss
+PlanNeofluar, Ph3 100x/1.3), placed in a thermostated box at 37$^{\circ}$C.
+In this setup, fluorescence illumination is provided by a mercury
+lamp (Osram, 1xHBO 103X/2) and visualization is performed with narrow-bandpass
+excitation and emission filters (Chroma, \#49002 ET-GFP and Chroma,
+\#49005 TR/DsRED ET). The exposure time is externally controlled by
+mechanical shutters (Uniblitz-VS35). Images were acquired with a 16-bit
+gray level CCDcamera cooled to -80$^{\circ}$C (Roper Scientific,
+Princeton Instruments PHOTOMAX 512) controlled by a custom-made software
+using Visual Basic and the Type libraries of the Winview software
+(Princeton Instruments). Every 5 minutes, autofocus was numerically
+performed by maximizing the contrast of a region of interest, and
+a series of acquisitions were made. A total of 6 fields, each containing
+15 wells was observed during the entire experiment.
 
 \subsection{Cell segmentation}
+
 \label{sec:cell_segmentation}
 
-Raw data are in the form of 512x512-pixel 16-bit SPE images (proprietary format produced by the Princeton Instruments camera).
-They were converted in 16-bit Tif images using the \textit{SPE} plugin of ImageJ~\cite{goto_open_2005}.
-The rest of the analysis was performed using Python 3.5.2.
-In particular, we used OpenCV~3.1.0-dev, Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
-
-The offset between consecutive images of a given field was evaluated using cross-correlation in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see Scikit-Image documentation at \cite{skimage_cross-correlation}).
-All the images for a field were then aligned on the first acquisition by simple affine translation.
-
-The wells were isolated in individual images through a combination of manual pixel selection and automatic segmentation.
-In particular, we selected the entrance of the two wells at the border of the image, and used this information along with the regularity of the microfluidic device to compute a mask that allowed to crop each well in its own image of size 100x21 pixels.
-These sub-images were labeled $\left\{W0, W1, ..., W14\right\}$ depending on the position of the well on the original image, from left to right.
-
-In this study, we performed a preliminary analysis on each well that focused on the deepest cell.
-For each image, we manually selected two pixels representing the position of the poles of the cell of interest.
-The distance between the two selected pixels was directly used as the bacteria length $L$ in pixels.
-They were also used to compute a rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
-We then computed the RFU/pixel in this cell mask by summing each pixel and dividing by the mask size.
-
-The camera noise and background was evaluated by taking a picture with a closed shutter at the end of the experiment.
-Pixels in this picture were found to be Gaussian distributed with a mean of 1101.0 and a standard deviation of 10.788.
-Correction for camera background was thus applied by removing 1101 to the computed RFU/pixel.
-
-The background for autofluorescence of the bacteria and the medium was supposed negligible and was thus not corrected.
-Indeed, the same device was used to image bacteria in stationary phase that do not produce any fluorescent protein.
-They were entirely indistinguishable from the camera noise.
-Not that this does not ensure that the autofluorescence is also negligible when the bacteria are actively growing.
-Furthermore, its value could change with the medium of interest, and even be different at steady state and during growth transitions, which does not allow to estimate it independently.
-In the section~\ref{sec:chap3_discussion}, we discuss possible improvements of the experimental set-up that would allow to dynamically co-estimate the autofluorescence with the ribosome abundance in a single experiment.
+The raw data from the camera are in the form of 512x512-pixel 16-bit
+SPE images (proprietary format produced by the Princeton Instruments
+camera). They were converted into 16-bit Tif images using the \textit{SPE}
+plugin of ImageJ~\cite{goto_open_2005}. The rest of the analysis
+was performed using Python 3.5.2. In particular, we used OpenCV~3.1.0-dev,
+Scikit-Image~0.12.3, Numpy~1.11.2 to manipulate the images.
+
+In order to register consecutive image of a time series, we calculated
+the offset between consecutive images of a given field using cross-correlation
+in the Fourier space~\cite{guizar-sicairos_efficient_2008} (see
+Scikit-Image documentation at \cite{skimage_cross-correlation}).
+All images for a field were thereby aligned to the first acquisition
+by a simple translation.
+
+In order to simplify the image treatment, we extracted individual
+wells from the images by a combination of manual pixel selection and
+automatic segmentation. In particular, we selected the entrance of
+the two wells at the border of the image, and used this information
+along with the regularity of the microfluidic device to compute a
+mask that allowed to extract each well into an image of size 100x21
+pixels. These sub-images were labeled $\left\{ W0,W1,...,W14\right\} $
+depending on the position of the well in the original image, from
+left to right.
+
+In this study, we performed a preliminary analysis of each well, focusing
+on the cell at the bottom of the channel. For each image, we manually
+selected the poles of the cell of interest. The distance between the
+two selected pixels was directly used as the length $L$ of the bacterium
+in pixels. The two selected positions were also used to compute a
+rectangular 6-pixel-wide mask around the cell (see Fig.~\ref{fig:data_acquisition}).
+We then computed the RFU/pixel in this cell mask by summing each pixel
+and dividing by the mask size.
+
+The camera noise and background were evaluated by taking a picture
+with a closed shutter at the end of the experiment. Pixels in this
+picture were found to be Gaussian distributed with a mean of 1101.0
+and a standard deviation of 10.788. A correction for the camera background
+was thus applied by removing 1101 from the computed RFU/pixel.
+
+The autofluorescence of the bacteria and the background fluorescence
+of the medium were supposed to be negligible and were thus not corrected.
+As a control, the same device was used to image bacteria in stationary
+phase that do not produce any fluorescent protein. They were indistinguishable
+from the camera noise. Of course, using this control we can not affirm
+that the autofluorescence is also negligible when the bacteria are
+actively growing. Furthermore, the autofluorescencecould change in
+different media of interest, and may even be different in steady state
+and during growth transitions. An independent estimate of the autofluorescence
+is therefore difficult to obtain. In the section~\ref{sec:chap3_discussion},
+we discuss possible improvements of the experimental setup that would
+allow to dynamically co-estimate the autofluorescence with the ribosome
+abundance in a single experiment.
 
 \subsection{Kalman smoothing}
+
 \label{sec:meth_kalman}
 
-The data about the length and the RFU/pixel of each bacteria were analyzed using Python~3.5.2.
-In particular, we manipulated the data using Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
-
-Historical details about the Kalman smoothing procedure are reported in section~\ref{sec:res_kalman} of the main text.
-We used the Additive Unscented Kalman Filter implementation of the pykalman python module.
-This class is reported to be more stable and computationally efficient with non-linear problems featuring additive noises.
-
-As described in section~\ref{sec:res_kalman}, the reconstruction of the growth rate was done on continuous portions of the length, \textit{i.e.} between two division events.
-The full problem, as reported in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity is:
+The data of the length and the RFU/pixel of each bacteria were analyzed
+using Python~3.5.2. In particular, we manipulated the data using
+Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
+
+Historical details about the Kalman smoothing procedure are reported
+in section~\ref{sec:res_kalman} of the main text. We used the Additive
+Unscented Kalman Filter implementation of the pykalman python module.
+This class is reported to be more stable and computationally efficient
+with non-linear problems containing additive noise.
+
+As described in section~\ref{sec:res_kalman}, the reconstruction
+of the growth rate was done on continuous portions of the length,
+\textit{i.e.} between cell divisions. The full problem, as described
+in Eq.~\ref{eq:full_mu_prob}, is reproduced below for clarity: 
 \begin{eqnarray*}
-\dot{V_\lambda}(t) &=& \mu (t) \cdot V_\lambda (t),\nonumber\\
-\dot{\mu}(t) &=& v(t),\\
-\dot{v}(t) &=& w(t),\nonumber
+\dot{V_{\lambda}}(t) & = & \mu(t)\cdot V_{\lambda}(t),\\
+\dot{\mu}(t) & = & v(t),\\
+\dot{v}(t) & = & w(t),
 \end{eqnarray*}
 with the measurement model: 
-\begin{equation*}
-L(t_k) = V_\lambda(t_k) + \epsilon_k.
-\end{equation*}
-The parameters used as prior for the reconstruction of $\mu$ are reported below.
-We used an observation variance of 9 pixels\textsuperscript{2} for the length $L$.
-The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6} during the whole time series.
-Inheritance between mother and daughter cells is taken into account by systematically choosing an initial mean equal to the last estimated value for $\mu$, and to half the last estimated value for $V_\gamma$ (modeling the symmetrical division).
-At the beginning of the experiment, were no mother cell are available, these values were fixed at 15~pixels for $V_\gamma$ and 0.004~min\textsuperscript{-1} for $\mu$.
-The variances associated with these means are 16~pixels\textsuperscript{2} and $10^{-4}$~min\textsuperscript{-2}, respectively for $V_\gamma$ and $\mu$.
-$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~min\textsuperscript{-6}.
-All the cross-covariances are set to 0 because the system variables are independent by construction.
-
-For the reconstruction of gene activity, we only had to cope with the large gap in data acquisition between -720 and -150 min.
-Reconstruction was then done independently on the continuous sections before and after this gap.
-The full problem for the estimation of $u = \gamma \alpha / \beta$, as reported in Eq.~\ref{eq:full_u_prob}, is reproduced here for clarity:
+\[
+L(t_{k})=V_{\lambda}(t_{k})+\epsilon_{k}.
+\]
+The parameters used as priors for the reconstruction of $\mu$ are
+described below. We used an observation variance of 9 pixels\textsuperscript{2}
+for the length $L$. The transition variance $\theta^{2}$ (\textit{a.k.a.}
+the smoothing factor for $\mu$) is fixed at $10^{-8}$~min\textsuperscript{-6}
+for the entire time series. Inheritance between mother and daughter
+cells is taken into account by systematically choosing an initial
+mean equal to the last estimated value for $\mu$, and to half the
+last estimated value for $V_{\gamma}$ (as expected for a completely
+symmetrical division). At the beginning of the experiment, when no
+mother cells are available, these values were fixed to 15~pixels
+for $V_{\gamma}$ and 0.004~min\textsuperscript{-1} for $\mu$.
+The variances associated with these means are 16~pixels\textsuperscript{2}
+and $10^{-4}$~min\textsuperscript{-2}, respectively, for $V_{\gamma}$
+and $\mu$. The initial mean of $v$ is always taken as null, with
+an initial variance of $10^{-8}$~min\textsuperscript{-6}. All the
+cross-covariances are set to 0 because the system variables are independent
+by construction.
+
+For the reconstruction of gene activity, we had to cope with the large
+gap in data acquisition between -720 and -150 min. Reconstruction
+was then done independently on the continuous sections before and
+after this gap. The full problem for the estimation of $u=\gamma\alpha/\beta$,
+as described in Eq.~\ref{eq:full_u_prob}, is reproduced here for
+clarity: 
 \begin{eqnarray*}
-\dot{r_\gamma}(t) &=& \hat{\mu} (t) \cdot u (t) - \hat{\mu} (t) \cdot r_\gamma (t), \nonumber\\
-\dot{u}(t) &=& v(t),\\
-\dot{v}(t) &=& w(t),\nonumber
+\dot{r_{\gamma}}(t) & = & \hat{\mu}(t)\cdot u(t)-\hat{\mu}(t)\cdot r_{\gamma}(t),\\
+\dot{u}(t) & = & v(t),\\
+\dot{v}(t) & = & w(t),
 \end{eqnarray*}
 with the measurement model 
-\begin{equation*}
-F(t_k) = r_\gamma (t_k) + \eta_k,\\
-\end{equation*}
-where $\hat{\mu}$ is the estimation of $\mu$ above.
-In both case, we used as prior an observation variance of 800.4 RFU\textsuperscript{2} for $F$.
-The transition variance $\theta^2$ (\textit{a.k.a.} the smoothing factor for $\gamma \alpha / \beta$) is fixed at $10^2$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
-The initial state means used are 600~RFU for $r_\gamma$ and 1000~RFU for $\gamma \alpha / \beta$.
-The variances associated with these means are purposely large and fixed to 10\textsuperscript{6}~RFU\textsuperscript{2} for $r_\gamma$ and $\gamma \alpha / \beta$.
-$v$ initial mean is always taken null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2}, imposing a null second derivative for the reconstructed signal $\gamma \alpha / \beta$.
+\[
+F(t_{k})=r_{\gamma}(t_{k})+\eta_{k},
+\]
+where $\hat{\mu}$ is the estimate of $\mu$ above. In both case,
+we used as prior an observation variance of 800.4 RFU\textsuperscript{2}
+for $F$. The transition variance $\theta^{2}$ (\textit{a.k.a.} the
+smoothing factor for $\gamma\alpha/\beta$) is fixed at $10^{2}$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
+The initial state means used are 600~RFU for $r_{\gamma}$ and 1000~RFU
+for $\gamma\alpha/\beta$. The variances associated with these means
+are purposely large and fixed at 10\textsuperscript{6}~RFU\textsuperscript{2}
+for $r_{\gamma}$ and $\gamma\alpha/\beta$. The initial mean of $v$
+is always taken as null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2},
+imposing a null second derivative for the reconstructed signal $\gamma\alpha/\beta$.
 Here again, all the cross-covariances are set to 0.
 
-All the parameters cited in this section were chosen through trials and errors, and are as a consequence largely optimizable.
-Possible improvements are discussed in section~\ref{sec:chap3_discussion}.
+All the parameters cited in this section were chosen through trial
+and error, and aretherefore certainly optimizable. Possible improvements
+are discussed in section~\ref{sec:chap3_discussion}.
 

### 2016-12-30 Chapters/experiments.tex
First full draft of chap3 suppinfo

@@
 
-We applied this experimental set-up to the reconstruction of the resources allocation strategy employed by \textit{E.
+We applied this experimental setup to the reconstruction of the resources allocation strategy employed by \textit{E.
 coli} during a nutrient upshift, with the aim to observe the bang-bang profile predicted in Chapter~\ref{chap:theory}.
@@
 While some parameters can be tweaked, some assumptions are always hard-coded in the algorithm and reflect the philosophy used to address the identification problem.
-Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~\cite{fluobactracker} that focuses on adapting their software to our set-up in a near future.
+Since the preliminary analysis reported here was completed, we started a collaboration with the authors of FluoBacTracker~\cite{fluobactracker} that focuses on adapting their software to our setup in a near future.
 It should provide a robust, automated way to identify and track the cells, enhancing the reproducibility of the analysis and increasing the number of cells available for computing population statistics.
@@
 
-\begin{figure}
-\includegraphics[width=13cm]{/home/geiselmann/Persons/Nils/Thesis/Man/Fig/NilsCloning}\caption{\label{fig:Cloning}Construction of the gfp-tagged ribosome subunit.
+\begin{figure}[p]
+\centering
+\includegraphics[scale=1]{./Fig/meth_cloning}
+\caption{
+\textbf{Construction of the gfp-tagged ribosome subunit.}
 The translational fusion of \textit{gfp\_mut2} with \textit{rpsB}
@@
 any other modification of the chromosome.}
-
-\selectlanguage{english}%
-\selectlanguage{french}%
+\label{fig:Cloning}
 \end{figure}
@@
 two long primers annealing, respectively, to the region just before
-the STOP codon of \textit{rpsB}, and to the beginning of \textit{gfpmut2}.
+the STOP codon of \textit{rpsB}, and to the beginning of \textit{gfp\_mut2}.
 The selection cassette contains an antiobiotic resistance gene, kanamycin
@@
 without the ATG start codon was amplified using long primers annealing,
-respectively, to the C-terminus of \ref{chap:theory}\textit{rpsB}
-(just before the STOP codon), and the end of the \textit{kan-P\textsubscript{BAD}-ccdB}
+respectively, to the C-terminus of \textit{rpsB}
+(just before the STOP codon), and the end of the \textit{kan-p\textsubscript{BAD}-ccdB}
 cassette. The first primer also contained a 18-bp (base pair) linker,
-inserting the same six amino-acids that Bakshi had used for their
+inserting the same six amino-acids that Bakshi \textit{et al.} have used for their
 construction ~\cite{bakshi_superresolution_2012}.
@@
 The two fragments, the \textit{gfp\_mut2} reporter including the linker
-and the \textit{kan-P\textsubscript{BAD}-ccdB} cassette, were assembled
-using Gibson assembly~\cite{gibson_enzymatic_2009} using a commercial
+and the \textit{kan-p\textsubscript{BAD}-ccdB} cassette, were assembled
+by Gibson assembly~\cite{gibson_enzymatic_2009} using a commercial
 Gibson Assembly mix (New England Biolabs). A final product of 2683
@@
 \item {*}(50 bp) the C-terminus of \textit{rpsB} without the STOP codon 
-\item (18 bp) a linker (see Supporting Information) 
+\item (18 bp) a linker
 \item (714 bp) the \textit{gfpmut2} sequence without the initial ATG 
-\item (1851 bp) the \textit{kan-P\textsubscript{BAD}-ccdB} cassette 
+\item (1851 bp) the \textit{kan-p\textsubscript{BAD}-ccdB} cassette 
 \item {*}(50 bp) the region directly after \textit{rpsB} in \textit{E.~coli} 
@@
 Regions labeled with {*} anneal to the \textit{E.~coli} chromosome.
-The complete sequence of this fragment, as well as the sequence of
-all the primers,are listed in the Supporting Information of this chapter.
+The complete sequence of this fragment, as well as the sequences of
+all the primers, are listed in \ref{S5_Text}.
 
 This fragment was electroporated into the wildtype strain, BW25113,
-containing the pSIM5 plasmid for lambda-red recombinaison \cite{sharan2009recombineering}.
+containing the pSIM5 plasmid for lambda-red recombinaison~\cite{sharan_recombineering_2009}.
 A kanamycin-resistant colony was selected on LA-glucose medium and
-verified to show green fluorescence in a Tecan microplate reader.
-A 100-bp oligonucleotide, containing 50 bp of the end of \textit{gfpmut2}
+verified to show green fluorescence (485~nm excitation, 535~nm emission) in a microplate reader (Tecan Infinite 200 pro).
+A 100-bp oligonucleotide, containing 50 bp of the end of \textit{gfp\_mut2}
 and 50 bp of the region just downstream of \textit{rpsB} was electroporated
-into this strain (sequence available in Supporting Information). The
-homologous recombination removes the \textit{kan-P\textsubscript{BAD}-ccdB}
+into this strain (sequence available in \ref{S5_Text}). The
+homologous recombination removes the \textit{kan-p\textsubscript{BAD}-ccdB}
 cassette. A colony was selected on an LA-arabinose plate. The clone
@@
 pSIM5 plasmid, containing a temperature-sensitive origin of replication.
-A chloramphenicol-sensitive colony was chosen andthe region after
-\textit{rpsB} was verified by sequencing (full sequence available
-in Supporting Information).
+A chloramphenicol-sensitive colony was chosen and the region after
+\textit{rpsB} was verified by sequencing (full sequence obtained available in \ref{S5_Text}).
 
@@
 and \textit{mCherry} versions were successfully obtained. The full
-sequence of the final \textit{rpsB-mCherry} strain is available in
-Supporting Information.
-
-The \textit{rpsB-gfp\_mut2} and \textit{rpsB-mCherry} strains were
-characterized on different media using a Tecan microplate reader.
+sequence of the final \textit{rpsB-mCherry} strain is available in \ref{S5_Text}.
+
+The \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains were
+characterized on different media using a Tecan microplate reader (\ref{S6_Text}, in particular Fig.~\ref{fig:suppinfo_caract} for growth on glucose).
 They possess a wild-type growth rate and sufficient fluorescence to
@@
 which made it unsuitable for our study. We therefore concentrated
-our efforts on the \textit{rpsB-gfp} strain. Detaied data about the
-characterization of the strain are available in Supporting Information.
+our efforts on the \textit{rpsB-gfp} strain (see \ref{S6_Text} for more details).
 
@@
 which were stored respectively at -20$^{\circ}$C and 4$^{\circ}$C.
+
 \begin{center}
@@
   \hline
-\multicolumn{2}{|c}{\textbf{M9 medium (100 mL)}} & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-\hline 
-CaCl2 {[}1 mol.L\textsuperscript{-1}{]}  & 10 $\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-MgSO$_{4}$ {[}1 mol.L\textsuperscript{-1}{]}  & 200 $\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-5x Salts  & 20 mL  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-Traces elements  & 90 $\mu$l  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-FeSO$_{4}$ {[}30 g.L\textsuperscript{-1}{]}  & 10 $\mu$l  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-Thiamine {[}10 g.L\textsuperscript{-1}{]}  & 50 $\mu$l  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-Carbon source  & at will  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-\hline 
-\end{tabular}%
+  \multicolumn{2}{|c|}{\textbf{M9 medium (100 mL)}}\\
+  \hline
+  CaCl2 [1 mol.L\textsuperscript{-1}] & 10 $\mu$l\\
+  MgSO$_4$ [1 mol.L\textsuperscript{-1}] & 200 $\mu$l\\
+   5x Salts & 20 mL \\
+  Traces elements & 90 $\mu$l \\
+  FeSO$_4$ [30 g.L\textsuperscript{-1}] & 10 $\mu$l \\
+  Thiamine [10 g.L\textsuperscript{-1}] & 50 $\mu$l \\
+  Carbon source & at will \\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
+  \hline
+\end{tabular}
 \begin{tabular}{|l r|}
   \hline
-\multicolumn{2}{|c}{\textbf{Traces elements (0.9 mL)}} & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-\hline 
-H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & 200$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-Na$_{2}$EDTA 2H$_{2}$O {[}150 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-ZnSO$_{4}$ 7H$_{2}$O {[}45 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-CoCl$_{2}$ 6H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-MnCl$_{2}$ 4H$_{2}$O {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-H$_{3}$BO$_{3}$ {[}10 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-Na$_{2}$MoO$_{4}$ 2H$_{2}$O {[}4 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-CuSO$_{4}$ 5H$_{2}$O {[}3 g.L\textsuperscript{-1}{]}  & 100$\mu$l & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-\hline 
-\end{tabular}%
+  \multicolumn{2}{|c|}{\textbf{Traces elements (0.9 mL)}}\\
+  \hline
+  H$_2$0 [18.2 M$\Omega$.cm] & 200$\mu$l\\
+  Na$_2$EDTA 2H$_2$O [150 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  ZnSO$_4$ 7H$_2$O [45 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CoCl$_2$ 6H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  MnCl$_2$ 4H$_2$O [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  H$_3$BO$_3$ [10 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  Na$_2$MoO$_4$ 2H$_2$O [4 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  CuSO$_4$ 5H$_2$O [3 g.L\textsuperscript{-1}] & 100$\mu$l\\
+  \hline
+\end{tabular}
 \begin{tabular}{|l r|}
     \hline
-\multicolumn{2}{|c}{\textbf{5x Salts (100 mL)}} & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-\hline 
-Na$_{2}$HPO$_{4}$2H$_{2}$0  & 4.25 g  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-KH$_{2}$HPO$_{4}$  & 1.5 mg  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-NaCl  & 0.25 g & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-NH$_{4}$Cl & 0.5 g & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
-H$_{2}$0 {[}18.2 M$\Omega$.cm{]}  & to 100 mL  & \selectlanguage{english}%
-\selectlanguage{french}%
-\tabularnewline
+  \multicolumn{2}{|c|}{\textbf{5x Salts (100 mL)}}\\
+  \hline
+  Na$_2$HPO$_4$2H$_2$0 & 4.25 g \\
+  KH$_2$HPO$_4$ & 1.5 mg \\
+  NaCl & 0.25 g\\
+  NH$_4$Cl& 0.5 g\\
+  H$_2$0 [18.2 M$\Omega$.cm] & to 100 mL \\
   \hline
 \end{tabular}
-\par\end{center}
+\end{center}
 
@@
 except FeSO$_{4}$, thiamine, and carbon sources, which were added
-just before inocculation of the pre-culture. Except where stated otherwise,
+just before inoculation of the pre-culture. Except where stated otherwise,
 the growth temperature was 37$^{\circ}$C. M9 Acetate contains 0.2\%
-Acetate (in mass of C$_{2}$H$_{3}$O$_{2}$ per mass of solution),
-and M9 Glucose contains 0.2\% Glucose (in mass of D-(+)-Glucose per
+acetate (in mass of C$_{2}$H$_{3}$O$_{2}$ per mass of solution),
+and M9 Glucose contains 0.2\% glucose (in mass of D-(+)-Glucose per
 mass of solution).
@@
 \textit{rpsB-gfp} strain was spread on a Petri dish of LA, and incubated
-at 37$^{\circ}$C. On the following day, an isolated colony was was
+at 37$^{\circ}$C. On the following day, an isolated colony was
 used to inoculate a cotton-plugged, sterile flask containing 20 mL
@@
 
-At the day of the experiment, the preculture was concentrated by centrifugation
-and re-suspended in 5 mL of M9 acetate, supplemented with 50 mg.mL\textsuperscript{-1}
+At the day of the experiment, the pre-culture was concentrated by centrifugation
+and re-suspended in 5 mL of M9 acetate, supplemented with 50~mg.mL\textsuperscript{-1}
 BSA (passivation buffer) for injection into the microfluidic device.
@@
 mL of growth medium, which was enough for several days of acquisition.
-A microfluidics pump (Elvesys) containing an output flow sensor module
+A microfluidic pump (Elvesys) containing an output flow sensor module
 was plugged to the medium bottle and a pressure up to 2~bars was
@@
 
-In order to register consecutive image of a time series, we calculated
+In order to correct for the drift of the device, we calculated
 the offset between consecutive images of a given field using cross-correlation
@@
 that the autofluorescence is also negligible when the bacteria are
-actively growing. Furthermore, the autofluorescencecould change in
+actively growing. Furthermore, the autofluorescence could change in
 different media of interest, and may even be different in steady state
@@
 using Python~3.5.2. In particular, we manipulated the data using
-Pandas~0.19.1, pykalman~0.9.5 and the curves submodule of wellfare~0.1.1.
+Pandas~0.19.1, Pykalman~0.9.5 and the Curves submodule of Wellfare~0.1.1.
 
@@
 in section~\ref{sec:res_kalman} of the main text. We used the Additive
-Unscented Kalman Filter implementation of the pykalman python module.
+Unscented Kalman Filter implementation of the Pykalman python module.
 This class is reported to be more stable and computationally efficient
@@
 \[
-L(t_{k})=V_{\lambda}(t_{k})+\epsilon_{k}.
+L(t_{k})=V_{\lambda}(t_{k})+\epsilon_{k},
+\]
+and the initial conditions
+\[
+V_\lambda (0) = V_{\lambda,0}, \;\;\;\; \mu (0) = \mu_0 , \;\;\;\; v(0) = v_0.
 \]
@@
 cells is taken into account by systematically choosing an initial
-mean equal to the last estimated value for $\mu$, and to half the
+mean for $\mu_0$ equal to the last estimated value for $\mu$, and for $V_{\gamma,0}$ to half the
 last estimated value for $V_{\gamma}$ (as expected for a completely
@@
 The variances associated with these means are 16~pixels\textsuperscript{2}
-and $10^{-4}$~min\textsuperscript{-2}, respectively, for $V_{\gamma}$
-and $\mu$. The initial mean of $v$ is always taken as null, with
+and $10^{-4}$~min\textsuperscript{-2}, respectively, for $V_{\gamma,0}$
+and $\mu_0$. The initial mean of $v$ is always taken as null, with
 an initial variance of $10^{-8}$~min\textsuperscript{-6}. All the
@@
 \]
+and the initial conditions
+\[
+r_\gamma (0) = r_{\gamma,0}, \;\;\;\; u (0) = u_0 , \;\;\;\; v(0) = v_0,
+\]
 where $\hat{\mu}$ is the estimate of $\mu$ above. In both case,
@@
 smoothing factor for $\gamma\alpha/\beta$) is fixed at $10^{2}$~RFU\textsuperscript{2}.min\textsuperscript{-4}.
-The initial state means used are 600~RFU for $r_{\gamma}$ and 1000~RFU
-for $\gamma\alpha/\beta$. The variances associated with these means
+The initial state means used are 600~RFU for $r_{\gamma,0}$ and 1000~RFU
+for $u_0$. The variances associated with these means
 are purposely large and fixed at 10\textsuperscript{6}~RFU\textsuperscript{2}
-for $r_{\gamma}$ and $\gamma\alpha/\beta$. The initial mean of $v$
+for $r_{\gamma,0}$ and $u_0$. The initial mean for $v_0$
 is always taken as null, with an initial variance of $10^{-8}$~RFU.min\textsuperscript{-2},
@@
 All the parameters cited in this section were chosen through trial
-and error, and aretherefore certainly optimizable. Possible improvements
-are discussed in section~\ref{sec:chap3_discussion}.
+and error, and are therefore certainly optimizable. Possible improvements
+are discussed in Section~\ref{sec:chap3_discussion}.
 
@@
 
-The following pages contain the full sequences of the \textit{gfpmut2-cassette} and \textit{mCherry-cassette} constructions, inserted on the \textit{E.~coli} BW25113 chromosome at the region after the \textit{rpsB} gene.
+The following pages contain the full sequences of the \textit{gfpmut2-cassette} and \textit{mCherry-cassette} constructions, inserted on the \textit{E.~coli} BW25113 chromosome at the region after the \textit{rpsB} gene (see Fig.~\ref{fig:Cloning}).
 The primers annealing positions are indicated using the same color code as in the description of the primers above.
-The primers reported in the original publication of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012} are indicated for information.
+The primers reported in the original publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.
 
@@
 The strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}.
-Regions at the end of the \textit{rpsB} gene were amplified, and sent for sequencing.
-Alignments of the results showed they exhibit the expected sequences (except for a couple of single-base mutations).
+The downstream region of the \textit{rpsB} gene was amplified, and sent for sequencing.
+Alignments of the results showed they exhibit the expected sequences, except for a couple of single-base mutations that do not impede the protein fluorescence.
 The raw results of the sequencing are reproduced below.
 
-\paragraph{Final sequences of the \textit{rpsB-gfpmut2} strain (region after \textit{rpsB})}
+\paragraph{Final sequences of the \textit{rpsB-gfp} strain (downstream region of \textit{rpsB})}
 
@@
 \newpage
-\paragraph{Final sequences of the \textit{rpsB-mCherry} strain (region after \textit{rpsB})}
+\paragraph{Final sequences of the \textit{rpsB-mCherry} strain (downstream region of \textit{rpsB})}
 
@@
 
-(signal-to-noise ratio, growth rate not hampered in microplate reader)
-Difference between mCherry and GFP (mCherry not good because so and so, GFP was chosen)
-
-Two type of strains were obtained
-
-Before being used in the microscopy experiments, the strain characteristics were validated by
+The \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains were characterized on different media using a Tecan microplate reader.
+Strains were grown in M9 minimal medium at 37$^\circ$C in 96-well microplates.
+The absorbance and fluorescence were measured approximately every minute for up to 24~h.
+In a single experiment, this allowed to construct up to 96 growth curves like presented in Fig.~\ref{fig:growth_curve}.
+
+In Fig.~\ref{fig:suppinfo_caract}, we can see the results obtained on M9 supplemented with 0.2\% glucose.
+As for the construction of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, our strains possess a wild-type growth rate (Fig.~\ref{fig:suppinfo_caract}A), indicating that the gfp-tagging or mCherry-tagging of the ribosomal S2 subunit does not impede the functioning of the ribosome.
+It is known that \textit{E.~coli} cultures exhibit a high autofluorescence in the GFP bandwidth~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
+Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the signal level for mCherry was even less a problem.
+Overall, both strains possess enough fluorescence to allow quantification.
+
+In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance.
+As a first approximation, this ratio can be used as a proxy for the fluorescence concentration in the cells, therefore for the ribosome concentration.
+The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty to obtain robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
+The RFU/Abs ratio is stable in the interval [400,600] for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, and can be interpreted as exponential steady-state growth.
+Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problem preventing the use of batch conditions to obtain robust data about growth transitions.
+
+Results for \textit{rpsB-gfp} and \textit{rpsB-mCherry} start to differ when glucose is exhausted and the cells enter stationary phase.
+As would be expected, the fluorescence of the \textit{rpsB-gfp} strain starts to slowly decrease as GFP proteins are degraded or photobleached.
+On the contrary, the fluorescence of the \textit{rpsB-mCherry} strain quickly triples when entering stationary phase, at a rate that is way higher than the physical limit imposed by the protein synthesis rate~\cite{scott_interdependence_2010}.
 
@@
 Growth and measurements were performed in a 96-well microplate placed in a Tecan infinite 200 pro reader thermostated at 37$^\circ$C.
-We monitored 4 wells inseminated with the WT strain, the \textit{rpsB-gfp} strain, and the \textit{rpsB-mCherry} strain, for a total of 12 wells each containing 150~$\mu$L of growth medium.
-Curves were shifted in time to correct for pipetting variability in the insemination process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
+We monitored 4 wells inoculated with the WT strain, the \textit{rpsB-gfp} strain, and the \textit{rpsB-mCherry} strain, for a total of 12 wells each containing 150~$\mu$L of growth medium.
+Curves were shifted in time to correct for pipetting variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
 \textit{(A)}~The light absorbance as measured by the Tecan microplate reader, corrected for background by removing well containing clean M9.
@@
 \end{figure}
+
+In order to explain these strange dynamics, we set-up an experiment to evaluate the degradation and maturation rates of GFP and mCherry.
+The procedure was to abruptly stop the protein synthesis rate in mid-exponential phase, and observe the evolution of the fluorescence in a condition with no synthesis of new reporter proteins.
+The estimation was performed using the following model:
+\begin{eqnarray}
+\frac{dP_m}{dt} &=& K \cdot P - (\mu+D) \cdot P_m, \\
+\frac{dP}{dt} &=& f(t) - (\mu+D+K) P,
+\end{eqnarray}
+where $P_m$ is the concentration of mature (fluorescent) proteins, $P$ is the concentration of immature (non-fluorescent) proteins, $K$ is the maturation rate, $D$ is the degradation rate (assumed identical for mature and immature proteins), $f(t)$ the protein synthesis rate, and $\mu$ the growth rate.
+By stopping all protein synthesis in the bacteria, we obtain $f(t) = \mu = 0$, hence the model becomes:
+\begin{eqnarray}
+\frac{dP_m}{dt} &=& K \cdot P - D \cdot P_m, \\
+\frac{dP}{dt} &=& - (D+K) P.
+\end{eqnarray}
+This system of differential equations can be analytically solved into:
+\[
+P_m(t) = \left(P_m(0) + P(0) \left( 1 - e^{-Kt} \right) \right) e^{-Dt}
+\]
+with $P_m(0)$ and $P(0)$ the initial concentration of mature proteins and immature proteins, respectively.
+By normalizing by $P_m(0)$ and taking the log of both sides, we can rewrite it as
+\begin{equation}
+\log \frac{P_m(t)}{P_m(0)} = - Dt + \log \left(1 + \frac{P(0)}{P_m(0)} \left( 1 - e^{-Kt} \right) \right).
+\label{eq:matur_deg}
+\end{equation}
+Since growth is stopped, the bacterial volume is constant and then $\frac{P_m(t)}{P_m(0)}$ can be easily measured by directly taking the fluorescence normalized by its initial value, while the maturation rate $K$, the degradation rate $D$, and the initial fraction of immature proteins $\frac{P(0)}{P_m(0)}$ are free parameters that can be fitted on fluorescence time series.
+The curves used for the estimation are represented in Fig.~\ref{fig:suppinfo_matur} and the results are reported in Tab.~\ref{tab:suppinfo_matur}.
+As expected, both reporter proteins are stable and exhibit a long half-life in exponential phase (>24~h).
+The maturation rate of GFP was too fast to obtain numerical value.
+However, the maturation rate of mCherry is rather slow, with a half-maturation time in the order of 30~min (time needed to mature 50\% of a pool of mCherry).
+Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value is not sufficient to explain the increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[scale=1]{./Fig/suppinfo_matur}
+\caption{
+\textbf{Estimation of the maturation and degradation rates of the reporter proteins in the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
+The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except a massive dose of Chloramphenicol was added at t=0 (vertical dashed grey line).
+The fluorescence was normalized by the value at t=0 and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
+Parameters of the best fit (solid black line) are reported in Tab.~\ref{tab:suppinfo_matur}.
+Data are the result of the aggregation of 9 and 8 independent growth curves for \textit{rpsB-gfp} and \textit{rpsB-mCherry}, respectively.
+}
+\label{fig:suppinfo_matur}
+\end{figure}
+
+\begin{table}[tb]
+\begin{tabular}{|c|c|c|}
+\hline
+& \textbf{GFP mut2} & \textbf{mCherry}\\
+\hline
+\textbf{$P(0) / P_m(0)$} & 0.05635 $\pm$ 0.001202  & 0.7349 $\pm$ 0.0027362\\
+\hline
+\textbf{$D$~[min\textsuperscript{-1}]} & 0.0002064 $\pm$ 2.066e-06 & 0.0003027 $\pm$ 2.645e-06\\
+\hline
+\textbf{half-life~[min]}& 3325 - 3392 & 2269 - 2309\\
+\hline
+\textbf{$K$~[min\textsuperscript{-1}]} & $\emptyset$ & 0.02254 $\pm$ 0.0003\\
+\hline
+\textbf{half-maturation~[min]} & $\emptyset$  &  30.27 - 31.16\\
+\hline
+\end{tabular}
+\caption{\textbf{Fitted parameters for the degradation and maturation of GFP mut2 and mCherry, according to the model in Eq.~\ref{eq:matur_deg} and data in Fig.~\ref{fig:suppinfo_matur}}}
+\label{tab:suppinfo_matur}
+\end{table}
+
+The fluorescence of GFP and mCherry is known to be affected by intracellular physiological changes like pH or pO\textsubscript{2}~\cite{doherty_stage-specific_2010}.
+In particular, mCherry has been reported to be extremely sensible to the presence of oxygen for its maturation~\cite{doherty_stage-specific_2010}.
+We however reproduced the following results in conditions were the oxygen is not limiting.
+In particular, the abrupt increase of mCherry fluorescence is conserved when the stationary phase is attained at low bacterial density, in a larger volume (20-mL flask), and on several type of carbon sources (acetate, xylose, glycerol, maltose).
+We have so far no explanation for this strange dynamic, and therefore concentrated our effort on the \textit{rpsB-gfp} strain in the microfluidic experiments.
 

### 2017-01-02 Chapters/experiments.tex
Finalize chap 3 suppinfo, finalize abstracts, remove eps files, version sent to Hidde and Hans

@@
 The DNA fragment containing the selection cassette was amplified using
-two long primers annealing, respectively, to the region just before
-the STOP codon of \textit{rpsB}, and to the beginning of \textit{gfp\_mut2}.
-The selection cassette contains an antiobiotic resistance gene, kanamycin
+two long primers annealing, respectively, downstream of \textit{rpsB} (starting just after
+the STOP codon), and to the end of \textit{gfp\_mut2}.
+The selection cassette contains an antibiotic resistance gene, kanamycin
 (positive selection), and a gene coding for the CcdB toxin under the
-control of the P\textsubscript{BAD} promoter (negative selection
+control of the \textit{p\textsubscript{BAD}} promoter (negative selection
 in presence of arabinose). This cassette is referred below as \textit{kan-p\textsubscript{BAD}-ccdB}.
@@
 
+\clearpage
 \section{Supporting Information for Chapter~3}
@@
 
-\subsubsection{Exhaustive list of the primers used...}
-
-\paragraph{...for the \textit{gfpmut2} amplification.}
+\subsubsection{Exhaustive list of the primers used}
+
+\paragraph{...for the \textit{gfp\_mut2} amplification.}
 
@@
 \item[] \includegraphics[scale=1]{Fig/gfpmut2_L.png} \seqsplit{GTTCTCAGGATCTGGCTTCCCAGGCGGAAGAAAGCTTCGTAGAAGCTGAGCAGGAAAGGCGACAGGAGAGTAAAGGAGAAGAACTTTTCACTG}
-(Length: 93)
-\item[] \includegraphics[scale=1]{Fig/gfpmut2_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTATAGTTCATCCATGCC} (Length: 44)
+(Length: 93)\\
+\textit{The 50 first bp anneals with the C-terminus of \textit{rpsB} (just before the STOP codon).
+The 18 next bp code for the linker (see Material and Methods~\ref{sec:methods_strain}).}
+\item[] \includegraphics[scale=1]{Fig/gfpmut2_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTATAGTTCATCCATGCC} (Length: 44)\\
+\textit{The 20 last bp anneals with the end of the \textit{kan-p\textsubscript{BAD}-ccdB} cassette.}
 \end{itemize}
-
-\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
-The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
 
@@
 (Length: 88)
-\item[] \includegraphics[scale=1]{Fig/mCherry_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTACAGCTCATCCATG} (Length: 42)
+\item[] \includegraphics[scale=1]{Fig/mCherry_R.png} \seqsplit{TGATGTTCTGGGGAATATAATTATTTGTACAGCTCATCCATG} (Length: 42)\\
+\textit{Same design as for gfp\_mut2, except it amplifies mCherry.}
 \end{itemize}
-
-\textbf{Rem.} The 50 first bases of the forward primer are supposed to anneal with the end of the \textit{rpsB} gene.
-The 20 last bases of the reverse primer are supposed to anneal with the ccdB-end of the \textit{kan-P\textsubscript{BAD}-ccdB} cassette.
 
@@
 \item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAACTATACAAATAATTATATTCCCCAGAACATCAGG}
-(Length: 42, used to assemble with gfp)
+(Length: 42, used to assemble with gfp)\\
+\textit{The 20 first bp anneals with the end of gfp\_mut2.}
 \item[] \includegraphics[scale=1]{Fig/cassette_L} \seqsplit{TGGATGAGCTGTACAAATAATTATATTCCCCAGAACATCAG}
-(Length: 41, used to assemble with mCherry)
+(Length: 41, used to assemble with mCherry)\\
+\textit{The 20 first bp anneals with the end of mCherry.}
 \item[] \includegraphics[scale=1]{Fig/cassette_R} \seqsplit{GAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTAGAAGAACTCGTCAAGAAGG}
-(Length: 72, used for both)
+(Length: 72, used for both)\\
+\textit{The 50 last bp anneals downstream rpsB (just before the STOP codon).}
 \end{itemize}
-
-\textbf{Rem.} The 20 first bases of the first forward primer are supposed to anneal with the end of the \textit{gfpmut2} gene.
-The 20 first bases of the second forward primer are supposed to anneal with the end of the \textit{mCherry} gene.
-The 50 last bases of the reverse primer are supposed to anneal with the region after \textit{rpsB} on the \textit{E.~coli} BW25113 chromosome.
 
@@
 \begin{itemize}
-\item[\textbf{\textit{rpsB-gfpmut2}}] \seqsplit{TTGTAACAGCTGCTGGGATTACACATGGCATGGATGAACTATACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
+\item[\textbf{\textit{rpsB-gfp}}] \seqsplit{TTGTAACAGCTGCTGGGATTACACATGGCATGGATGAACTATACAAATAAGCTTGATAACTCCCCCAAAATAGTTCGAGTTGCAGAAAGGCGGCAAGCTC}
 (Length: 100)
@@
 
-The following pages contain the full sequences of the \textit{gfpmut2-cassette} and \textit{mCherry-cassette} constructions, inserted on the \textit{E.~coli} BW25113 chromosome at the region after the \textit{rpsB} gene (see Fig.~\ref{fig:Cloning}).
+The following pages contain the full annotated sequences of the region downstream \textit{rpsB} after the insertion of \textit{gfp\_mut2-cassette} and \textit{mCherry-cassette}, respectively (third line in Fig.~\ref{fig:Cloning}).
 The primers annealing positions are indicated using the same color code as in the description of the primers above.
-The primers reported in the original publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.
-
-\includepdf[pages={1-7}]{./Fig/gfp-mut2-cassette-after-rpsB-gene-sequence.pdf}
-\includepdf[pages={1-6}]{./Fig/mcherry-cassette-after-rpsB-gene-sequence.pdf}
+The primers reported in the publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.
+
+\includepdf[pages={1-7}]{./Fig/gfp_mut2-cassette-after-rpsB-E-coli-sequence.pdf}
+\includepdf[pages={1-7}]{./Fig/mCherry-cassette-after-rpsB-E-coli-sequence.pdf}
 
@@
 
-The strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}.
-The downstream region of the \textit{rpsB} gene was amplified, and sent for sequencing.
-Alignments of the results showed they exhibit the expected sequences, except for a couple of single-base mutations that do not impede the protein fluorescence.
+After the strains were made using the protocol described in Material and Methods~\ref{sec:methods_strain}, the downstream region of the \textit{rpsB} gene was amplified and sent for sequencing.
+Alignments of the results showed they exhibit the expected sequences, except for a couple of single-base mutations.
 The raw results of the sequencing are reproduced below.
@@
 The absorbance and fluorescence were measured approximately every minute for up to 24~h.
-In a single experiment, this allowed to construct up to 96 growth curves like presented in Fig.~\ref{fig:growth_curve}.
-
-In Fig.~\ref{fig:suppinfo_caract}, we can see the results obtained on M9 supplemented with 0.2\% glucose.
+In a single experiment, this allowed to generate up to 96 growth curves like the one schematized in Fig.~\ref{fig:growth_curve}.
+
+In Fig.~\ref{fig:suppinfo_caract}, we show growth curves obtained on M9 supplemented with 0.2\% glucose.
 As for the construction of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, our strains possess a wild-type growth rate (Fig.~\ref{fig:suppinfo_caract}A), indicating that the gfp-tagging or mCherry-tagging of the ribosomal S2 subunit does not impede the functioning of the ribosome.
 It is known that \textit{E.~coli} cultures exhibit a high autofluorescence in the GFP bandwidth~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
-Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the signal level for mCherry was even less a problem.
-Overall, both strains possess enough fluorescence to allow quantification.
-
-In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance.
+Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the corrected signal level for mCherry is even more stronger.
+Overall, both strains possess enough fluorescence to allow quantification at the population level, all the more so at the cell level because most part of the autofluorescence is concentrated in the medium outside the cell~\cite{mihalcescu_green_2015}.
+
+In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance (corrected means that we substracted the autofluorescence of the wild-type strain).
 As a first approximation, this ratio can be used as a proxy for the fluorescence concentration in the cells, therefore for the ribosome concentration.
 The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty to obtain robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
-The RFU/Abs ratio is stable in the interval [400,600] for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, and can be interpreted as exponential steady-state growth.
+The RFU/Abs ratio is stable in the interval [400,600]~min for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, and can be interpreted as exponential steady-state growth.
 Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problem preventing the use of batch conditions to obtain robust data about growth transitions.
+Indeed, most of our attempt to perform robust growth transitions in this region were unsuccessful.
 
@@
 Curves were shifted in time to correct for pipetting variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
-\textit{(A)}~The light absorbance as measured by the Tecan microplate reader, corrected for background by removing well containing clean M9.
+\textit{(A)}~The light absorbance as measured by the Tecan microplate reader, corrected for background by removing M9 absorbance.
 The superposition of the curves indicates similar growth rates between the WT and the modified strains.
 \textit{(B)}~The fluorescence measured in the wells, expressed in Relative Fluorescence Units (RFU).
-Green fluorescence (485nm excitation, 535nm reception) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
-Red fluorescence (560nm excitation, 635nm reception) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
-As can be seen, fluorescence levels of the modified strains (solid lines) are far above the autofluorescence measured on the WT strain.
+Green fluorescence (485~nm excitation, 535~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
+Red fluorescence (560~nm excitation, 635~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
+As can be seen, fluorescence levels of the modified strains are far above the autofluorescence measured on the WT strain.
 \textit{(C)}~The ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
 Autofluorescence background was corrected by removing the autofluorescence measured on the WT wells.
-We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something that is not visible on the \textit{rpsB-gfp} strain.
+We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something does not occur on the \textit{rpsB-gfp} strain.
 Before each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
@@
 
-In order to explain these strange dynamics, we set-up an experiment to evaluate the degradation and maturation rates of GFP and mCherry.
-The procedure was to abruptly stop the protein synthesis rate in mid-exponential phase, and observe the evolution of the fluorescence in a condition with no synthesis of new reporter proteins.
+In order to explain these strange dynamics, we set-up an experiment to evaluate the degradation and maturation rates of GFP and mCherry in our strains.
+The procedure was to instantly stop all protein synthesis in mid-exponential phase using an antibiotic, then to observe the evolution of the fluorescence in a condition with no synthesis of new reporter proteins.
 The estimation was performed using the following model:
@@
 \]
-with $P_m(0)$ and $P(0)$ the initial concentration of mature proteins and immature proteins, respectively.
-By normalizing by $P_m(0)$ and taking the log of both sides, we can rewrite it as
+with $P_m(0)$ and $P(0)$ the initial concentrations of mature proteins and immature proteins, respectively.
+By dividing by $P_m(0)$ and taking the log of both sides, we can rewrite it as
 \begin{equation}
@@
 However, the maturation rate of mCherry is rather slow, with a half-maturation time in the order of 30~min (time needed to mature 50\% of a pool of mCherry).
-Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value is not sufficient to explain the increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.
+Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value are not sufficient to explain the rapid increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.
 
@@
 \textbf{Estimation of the maturation and degradation rates of the reporter proteins in the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
-The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except a massive dose of Chloramphenicol was added at t=0 (vertical dashed grey line).
-The fluorescence was normalized by the value at t=0 and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
+The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except that a massive dose of Chloramphenicol was added in mid-exponential phase ($t=0$, vertical dashed grey line).
+The fluorescence was normalized by the value at $t=0$ and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
 Parameters of the best fit (solid black line) are reported in Tab.~\ref{tab:suppinfo_matur}.
@@
 
-We estimated the noise in the RFU/pixel/cell by using the data points just before upshift (Fig.~\ref{fig:noise_poi}).
+We estimated the noise in the RFU/pixel/cell measurements by using the data points just before the upshift (Fig.~\ref{fig:noise_poi}).
 Since they have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
 For this reason, ribosome concentration, and so fluorescence concentration, are expected constant.
-Fig.~\ref{fig:noise_poi} represents the distribution of points in this region for a particular cell, that can be approximated by a Gaussian distribution.
+Fig.~\ref{fig:noise_poi} shows the distribution of points in this region, that can be approximated by a Gaussian distribution, for a particular cell.
 
@@
 \caption{
-\textbf{Point of interest for the noise estimation.}
+\textbf{Points of interest for the noise estimation.}\newline
 \textit{(Top graph)} RFU/pixel/cell data points for the cell at the bottom of the well labeled X3Y2,W2.
 Noise estimation was performed on the points just before the upshift (red points), were the bacteria is assumed to grow at steady state.
-\textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (CDF, in gray) of the points highlighted in red on the top graph.
+\textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (PDF, in gray) of the points highlighted in red on the top graph.
 They are visually compared with CDF (solid green line) and PDF (dashed green line) of a Gaussian fit with mean 548.88 and standard deviation 17.316.
@@
 were $F(t_k)$ is the fluorescence measured at the time step $t_k$, $r (t_k)$ is the true ribosomal concentration at this time step, $\gamma$ is an unknown factor, and $\eta_k$ is the measurement noise.
-In this scenario, for a time independent ribosome concentration $r$ (steady state), the ribosome concentration is constant and equal to its mean on the interval of interest (noted $m_k(\cdot)$).
+In this scenario, for a time independent ribosome concentration $r$ (\textit{i.e.} at steady state), the ribosome concentration is equal to its mean on the interval of interest (noted $m_k(\cdot)$).
 Given that
@@
 
-Secondly, the noise model can be considered multiplicative.
+Alternatively, the noise model can be considered multiplicative.
 The noise model is than different than presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
@@
 For this reason, we decided in the main analysis to stick with an additive noise, the main reason being that the implementation used for the Kalman smoothing procedure are faster and more stable with an additive noise model.
-We thus assumed an additive white Gaussian noise with mean 0 and standard deviation 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
-As discussed in Section~\ref{sec:chap3_discussion}, the only way to decisively choose a noise model over the other would have been to obtain two different steady-state growths for the each cell.
+We assumed an additive white Gaussian noise with mean 0 and standard deviation 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
+As discussed in Section~\ref{sec:chap3_discussion}, the only way to decisively choose a noise model over the other would have been to obtain two different steady-state growths for each cell.
 Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future instances of the experiment.
@@
 
-To help the reader get a better idea on the analysis, we display below the complete reconstructions for 5 cells.
+In order to get a more visual idea of the results of the analysis, we display below the complete signal reconstructions for 5 cells.
 In Figs~\ref{fig:cell1}-\ref{fig:cell5}, the image on the left is the last image analyzed for the corresponding well.
@@
 A 6-pixel-wide rectangular mask was computed for each image, resulting in the masked image on the right.
-In each of the graphs, the black and green points represents data points, while the red solid lines are the results of the signal reconstruction \textit{via} the Kalman smoothing procedure, the parameters of which are described in Material and Methods~\ref{sec:meth_kalman} and Figs~\ref{fig:growth_rate_estimation} and \ref{fig:gene_activity}.
+In each of the graphs, the black and green points represent data points, while the red solid lines are the results of the signal reconstruction \textit{via} the Kalman smoothing procedure.
+The parameters of the Kalman smoothing are described in Material and Methods~\ref{sec:meth_kalman} and Figs~\ref{fig:growth_rate_estimation} and \ref{fig:gene_activity}.
 Specific comments on each cell are given in the caption of Figs~\ref{fig:cell1}-\ref{fig:cell5}.
@@
 Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seems to have stabilized before the upshift.
-Oscillations in the RFU/pixel/cell signal are clearly visible, which results in oscillations in the resource allocation signal reconstruction, even though the smoothing factor seems a little too high for this particular cell.
+Oscillations in the RFU/pixel/cell signal are clearly visible, and result in oscillations in the resource allocation signal reconstruction.
 }
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-Here, the resource is unstable at the beginning of the experiment despite the apparent regularity of the RFU/pixel/cell.
+Here, the resource allocation is unstable at the beginning of the experiment despite the apparent regularity of the RFU/pixel/cell.
 This could indicate a smoothing factor that is too low for this particular cell.
@@
 We see for this cell that the reconstruction of the growth rate on the acetate medium (before 0) is affected by the huge gap in the acquisition, the increase before -150~min clearly being an artifact.
-However, because the Kalman smoothing procedure we used allows for flexibility between the mother and the daughter cell, we quickly recover a more realistic growth rate before the upshift.
+However, because the Kalman smoothing procedure we used allows for flexibility between the mother and the daughter cells, we quickly recover a more realistic growth rate before the upshift.
 Despite an instability at the beginning of the experiment, the resource allocation reconstruction exhibits a remarkable stability during growth on the acetate medium, followed by oscillations after the upshift on glucose (also visible directly on the RFU/pixel/cell data).
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-The reconstruction on this cell is a good instance of what we aimed to obtain: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
-The resource allocation is stable on a acetate (indicating a steady state) than starts to oscillate after the upshift on glucose.
+The reconstruction on this cell is a good instance of what was expected: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
+The resource allocation is stable on acetate (indicating a steady state) than starts to oscillate after the upshift on glucose, even though the smoothing factor seems a little too high for this particular cell (some oscillations are poorly predicted after the upshift, in particular in the interval [200,400]~min).
 }
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-The results on this cell are particularly terrible: we do not seem to get a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating.
-The oscillations after the upshift are thus very hard to interpret.
+The results on this cell are particularly terrible: we do not seem to get a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating before the upshift.
+It is thus impossible to conclude if the oscillations observed after the upshift are real or due to overfitting.
+This cell highlights why robust steady states before and after the transition are important to conclude on the signal reconstruction.
 }
@@
 
-\begin{figure}[h]
+\begin{figure}[!h]
 \centering
@@
 \textbf{Cell categories identified in the microscopy analysis.}
-As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: the dying cells (lowest graphs) stop growing, pausing cells (middle graph) present growth rate oscillations, and normal cells (highest graph) does not present any of the above}
+As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: dying cells stop growing (lowest graphs) , pausing cells present growth rate oscillations (middle graph), and normal cells does not present any of the above (highest graph).}
 \label{fig:subcat_cells}
@@
 \textbf{Robust statistics for the cell categories identified in the microscopy analysis.}
-Each graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step for the growth rate presented in~\ref{S2_Figure}.
+Each graph shows the 25\% (lower gray curve), 50\% (solid black curve) and 75\% (upper gray curve) quartiles, computed at each time step for the growth rates presented in~\ref{S2_Figure}.
 The gray area represents the interquartile range.

### 2017-01-02 abstract.tex
Finalize chap 3 suppinfo, finalize abstracts, remove eps files, version sent to Hidde and Hans

@@
 \selectlanguage{english}
+%%% MAX 4000 CHARACTERS %%%%
 \begin{abstract}
-Growth is one of the most fundamental processes of life, and is highly optimized in microorganisms.
-It relies on the transformation of matter and energy from the environment into diverse organic structures.
-Decades of studies have shown that despite an extreme microbial diversity, this transformation follows general regularities, forming the so-called microbial growth laws.
-But those works focused on steady-state growth, despite this condition being irrelevant to the natural growth condition of microorganisms.
+Growth is one of the most fundamental processes of life.
+In microbial cells, it relies on the transformation of matter and energy from the environment into diverse organic structures.
+Interestingly when comparing growth in different environments, microbial physiology exhibits general growth laws that relate the macromolecular composition of the cell to the growth rate.
+These laws are widespread and conserved in different microbial species, suggesting a fundamental principle of design.
+Indeed, recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation, and explained by the principles of natural selection.
+However, while these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
+The aim of this thesis is to extend the theoretical and experimental studies of microbial growth strategies to dynamical environments.
 
-Do these laws still hold when considering a dynamical environment?
-What does it mean to be optimal during an environmental transition?
+Using a self-replicator model, we develop an intuitive framework that encapsulates the main features of growth through nutrient assimilation.
+We formulate dynamical growth maximization as an optimal control problem that the microbial cell must solve to allocate its resources between gene expression machinery and metabolism.
+We find a general solution using Pontryagin's Maximum Principle, and compare it with different possible implementations of growth control in bacterial cells.
+It shows that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
+A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
+Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
+It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
 
-This project aims to lay the foundation of dynamical growth laws, more in adequation with the natural habitats of microorganisms and the selective pressures that shaped their current forms.
-We develop an intuitive framework by using a self-replicator model that encapsulates the main features of growth through nutrient assimilation.
-We formulate the resource allocation problem that arises as an optimal control problem in which the cell has to allocate its resources between gene expression machinery and metabolism.
-By applying the Pontryagin's Maximum Principle, the model predicts the optimal dynamical behavior to be a bang-bang allocation between gene expression machinery and metabolism.
-Despite not having been observed yet, we show that such behaviors are conceptually feasible with the known regulations of the enterobacteria \textit{Escherichia coli}, in particular the ppGpp system that regulates ribosomal abundance.
+While suggested by old literature data, this switch-like ribosome synthesis after a growth transition has never been experimentally confirmed.
+For this reason, we measure ribosomal abundance of \textit{Escherichia coli} at the single-cell level during a nutrient upshift.
+More precisely, we have constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing \textit{in-vivo} monitoring of the abundance of ribosomes.
+We monitor this strain in a microfluidic device designed for long-term imaging of individual cells in a continuous culture, and use it to simulate a nutrient upshift by changing the input medium.
+We develop a Kalman smoothing method that is able from these data to reconstruct how resources are allocated towards the synthesis of ribosomes.
+The results obtained are preliminary and do not allow a decisive validation, but they suggest the presence of oscillatory patterns after an upshift that could be reminiscent of the expected behavior.
 
-In order to find such patterns in real cells, we make a strain displaying fluorescent ribosomes.
-We observe \textit{E. coli} cells in a microfluidic device and monitor their fluorescence during growth transitions.
-Using state-of-the-art techniques of parameter estimation, we can reconstruct the state of the cellular resource allocation controler with a high temporal precision.
-Even if preliminary, our experimental results seem to confirm the bang-bang singular nature of resource allocation during a growth transition, as predicted by our model.
-
-Our work might play an important role in the formulation of dynamical fundamental growth laws.
-It also has application for biotechnological purposes, since microorganisms and in particular \textit{E. coli}, are more and more used in health and energy industries to produce valuable elements.
+Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
+All over, they demonstrate that when applied in dynamical instead of steady-state conditions, the same principles can generate unexpected behaviors and provide more information about the regulatory strategies employed by microorganisms.
+Such new understandings could be of great help for biotechnological purposes, in particular when using microorganisms to produce valuable elements for the health and energy industries.
 \end{abstract}
 
+%TC:ignore
+
+%%% MAX 1000 CHARACTERS %%%
+\begin{author-summary}
+Microbial cells sustain and replicate themselves from the matter and energy they extract from their environment.
+Strategies enabling microorganisms to optimize their replication rate have been extensively studied, but mostly in stable environments.
+Here, we build a simple model of microbial growth and use mathematical methods from optimal control theory to determine how the cells should behave when the environment around them is abruptly changed.
+Interestingly, we find that microbial cells should adopt an \textit{on-off} strategy: they should alternatively allocate all their resources to the production of each cell component, a prediction that we also tried to confirm experimentally.
+In other words, the regulations inside the bacterial cells could profit from functioning more like a switch, rather than a dim, in changing environments.
+These results might suggest new ways to optimize the biotechnological production of molecules of interest.
+\end{author-summary}
+
 \selectlanguage{french}
+%%% MAX 4000 CARACTERES %%%%
 \begin{abstract}
-La croissance et la reproduction sont des mécanismes fondamentaux du vivant.
-Généralement couplés chez les micro-organismes, ces deux processus reposent sur la transformation des ressources de l'environnement (matière et énergie) en nouvelles structures organiques.
-Malgré l'extrême diversité des micro-organismes, certaines caractéristiques clés du contrôle de la croissance semblent être largement répandues, au point que des lois de croissance ont pu être érigées.
-Ces dernières montrent que des mécanismes encore plus fondamentaux dictent la répartition des ressources dans les différents sous-systèmes de la cellule, généralement dans un souci d'optimisation de son taux de croissance.
-Mais ces lois ont été établies uniquement lors d'une croissance à l'état stationnaire, c'est-à-dire lorsque l'environnement et le taux de croissance ne varient pas dans le temps.
+Croissance et reproduction sont des mécanismes fondamentaux du vivant.
+Chez les micro-organismes, ces processus sont couplés dans la transformation des ressources de l'environnement (matière et énergie) en nouvelles structures organiques.
+Étonnamment, malgré l'extrême diversité des micro-organismes, certaines caractéristiques de leur physiologie suivent des lois de croissance universelles.
+Cela suggère l'existence de principes fondamentaux, ce qui a été en effet récemment confirmé en montrant que ces lois s'expliquent facilement si l'organisme maximise son taux de croissance dans chaque environnement.
+Cependant, ces lois ont seulement été étudiées lors de croissances à l'état stationnaire, c'est-à-dire lorsque l'environnement et donc le taux de croissance sont stables.
 Ces conditions, même si elles peuvent être reproduites en laboratoire, n'ont rien à voir avec les conditions de vie dans lesquelles les organismes ont évolué durant des milliards d'années.
+Le but de cette thèse est d'étendre dans un contexte d'environnement purement dynamique, l'étude à la fois théorique et expérimentale de ces stratégies de croissance microbienne.
 
-Ces lois sont-elles toujours valable en environnement dynamique ?
-Comment traduire les concepts d'optimalité dans de tels environnements ?
+En modélisant la cellule comme un auto-réplicateur, nous cherchons à savoir quelles sont les meilleures stratégies d'allocation des ressources lors de l'adaptation à un nouvel environnement.
+Ce problème se formule très bien comme un problème de contrôle optimal : la cellule choisit en temps réel comment allouer ses ressources entre la machinerie d'expression génique et le métabolisme.
+Le meilleur comportement possible, comme le révèle l'application du Principe de Maximisation de Pontryagin, est de successivement orienter toutes les ressources dans chacun des deux secteurs, une stratégie communément appelée \textit{bang-bang}.
+Mais s'approcher d'un tel contrôle requiert pour la cellule des stratégies de régulation bien plus complexes que celles qui étaient suffisantes pour maximiser le taux de croissance à l'état stationnaire.
+De manière intéressante, la régulation de la synthèse des ribosomes par le ppGpp chez la bactérie \textit{Escherichia coli} s'avère présenter la structure adéquate.
+Nous montrons en effet qu'elle permet de détecter rapidement toute incompatibilité entre la concentration de précurseurs et celle des ribosomes, et d'ajuster en conséquence la synthèse de ces derniers pour obtenir un comportement proche de l'optimum mathématique prédit.
 
-Le but de ce projet est de poser les bases de nouvelles lois de croissance étendues en environnement dynamique, c'est-à-dire plus proche du milieu réel de vie, et tenant mieux compte des pressions de sélection qui ont donné lieu à ces organismes.
-Notre cadre d'étude s'articule autour d'un modèle d'auto-réplicateur qui prend en compte les caractéristiques principales de la croissance par assimilation de nutriments.
-De ce cadre émerge un problème d'allocation des ressources entre la machinerie d'expression génique et le métabolisme, que nous formulons comme un problème de contrôle optimal.
-Le modèle prédit ainsi via l'application du Principe de Maximisation de Pontryagin que le meilleur comportement possible lors d'une transition est une allocation de type bang-bang, c'est-à-dire lors de laquelle les ressources vont par alternance soit uniquement dans la machinerie, soit uniquement dans le métabolisme.
-Malgré l'absence d'observation claire d'une tel phénomène, nous montrons qu'un tel comportement est conceptuellement possible à partir des régulations connues de l'enterobactérie \textit{Escherichia coli}, notamment le système ppGpp qui régule l'abondance des ribosomes.
+Même si de vieilles données le suggèrent, un tel comportement \textit{bang-bang} n'a jamais été totalement confirmé expérimentalement pour la synthèse des ribosomes.
+Nous mesurons donc l'abondance des ribosomes chez \textit{Escherichia coli} au niveau d'une cellule unique lors d'un changement brutal de milieu de culture.
+En particulier, nous créons une souche de \textit{E.coli} sur laquelle un rapporteur fluorescent est attaché à l'une des sous-unités ribosomales, permettant ainsi leur quantification \textit{in vivo}.
+Un appareillage micro-fluidique nous permet ensuite de contrôler en temps réel le milieu de croissance tout en observant individuellement chaque cellule fluorescente.
+Nous développons une méthode basée sur le lissage de Kalman qui est capable de reconstruire la façon dont les ressources sont aiguillées vers la synthèse des ribosomes.
+Même si ces résultats sont préliminaires, ils suggèrent que la concentration des ribosomes oscille après le changement d'environnement, ce qui rappelle une stratégie de type \textit{bang-bang}.
 
-Afin d'observer de tester ces prédictions, nous avons ingéniéré une souche de \textit{E. coli} arborant des ribosomes fluorescents.
-Nous avons observé ces cellules dans un système de microfluidique et enregistré leur fluorescence au cours de transitions de croissance.
-En utilisant les dernières méthodes d'identification paramétrique, nous pouvons reconstruire l'état du contrôleur d'allocation des ressources avec une très bonne précision temporelle.
-Même s'ils ne sont que préliminaires, ces résultats expérimentaux semblent confirmer la nature bang-bang de l'allocation lors des transitions de croissance chez \textit{E. coli}.
+Nos résultats montrent que la capacité des systèmes de régulations à intégrer l'état de différentes variables physiologiques est crucial dans l'optimisation de la croissance en environnement variable.
+Au final, nous démontrons que les principes utilisés à l'état stationnaire peuvent, lorsqu'ils sont appliqués en dynamique, générer des comportements inattendus et expliquer plus en détails les stratégies de régulations employées par les micro-organismes.
+\end{abstract}
 
-Notre travail peut jouer un rôle clé dans la formalution de nouvelles lois de croissances, étendues en environnement dynamique.
-Il a également de nombreuses applications en biotechnologie, dans la mesure où les micro-organismes et plus particulièrement \textit{E.
-coli} sont très largement utilisés par de nombreuses industries pour produire des molécules valorisables.
-\end{abstract}
+\begin{resume-vulgaire}
+Bactéries et autres micro-organismes se multiplient en utilisant la matière et l'énergie présentent dans leur environnement.
+Les stratégies qui leur permettent d'optimiser ce processus ont été longuement étudiées, mais uniquement dans des environnements stables.
+Ici, nous construisons un modèle simple de la croissance microbienne et utilisons la théorie du contrôle optimal pour déterminer la façon dont les cellules devrait agir lorsque leur environnement est soudainement modifié.
+Nous mettons en évidence que la meilleure stratégie consiste à aiguiller toutes les ressources disponibles vers la production d'un seul composant de manière alternative, une prédiction que nous tentons également de confirmer de manière expérimentale.
+En d'autres termes, dans un contexte dynamique, mieux vaut agir tel un interrupteur à bascule plutôt qu'un variateur de lumière.
+Ces résultats pourraient suggérer de nouvelles façons d'optimiser les processus biotechnologiques de production de molécules d'intérêt.
+\end{resume-vulgaire}
+
 \selectlanguage{english}
+%TC:endignore

### 2017-01-02 titlepage.tex
Finalize chap 3 suppinfo, finalize abstracts, remove eps files, version sent to Hidde and Hans

@@
 \EcoleDoctorale{Ecole doctorale Chimie et Science du Vivant}  
-\Titre{Microbial growth control in shifting environments}
+\Titre{Microbial growth control in changing environments}
 \Soustitre{Theoretical and experimental study of resource allocation in \textit{Escherichia coli}}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
@@
 
-\UGTRapporteur{M, Matthew SCOTT}{Assoc. Prof.}{University of Waterloo}      %% 1er rapporteur
-\UGTRapporteur{M, Guy-Bart STAN}{Reader}{Imperial College London}      %% second rapporteur
+\UGTRapporteur{M, Matthew SCOTT}{Assoc. Prof.}{Univ. of Waterloo (CA)}      %% 1er rapporteur
+\UGTRapporteur{M, Guy-Bart STAN}{Reader}{Imp. College London (UK)}      %% second rapporteur
 
-\UGTExaminateur{M, Olivier BERNARD}{DR}{Inria, Nice}     %% 2ème examinateur
-\UGTExaminateur{M, Matthieu JULES}{CR}{Inra, Grignon}     %% 1er examinateur
+\UGTExaminateur{M, Olivier BERNARD}{DR}{Inria Nice}     %% 2ème examinateur
+\UGTExaminateur{M, Matthieu JULES}{CR}{Inra Grignon}     %% 1er examinateur
 \UGTExaminatrice{Mme, Irina MIHALCESCU}{Prof.}{Univ. Grenoble Alpes}    %% 3ème examinateur
@@
 \UGTDirecteur{M, Johannes GEISELMANN}{Prof.}{Univ. Grenoble Alpes}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde DE JONG}{DR}{Inria, Grenoble}     %% Co-Directeur de thèse s'il y en a
-\UGTInvite{M, Jean-Luc GOUZÉ}{DR}{Inria, Nice}
+\UGTCoDirecteur{M, Hidde DE JONG}{DR}{Inria Grenoble}     %% Co-Directeur de thèse s'il y en a
+\UGTInvite{M, Jean-Luc GOUZÉ}{DR}{Inria Nice}
 %\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}

### 2017-01-02 Chapters/discussion.tex
Hidde's corrections of intro and discussion

--- 2016-06-01 Chapters/discussion.tex
@@
 \chapter{Discussion}
+\label{chap:discussion}
 
-- in praise of simple models
-- an intrinsic link between complexity and environment dynamic
-- some perspectives for industrial production
+\textit{"Apples fall onto the Earth because natural selection eliminated apples falling towards the sky."} -- @tomroud~\cite{tomroud_tom_2016}, original source unknown.
+
+\selectlanguage{french}
+\section*{Résumé en français}
+
+Dans cette section... (env. 1 page)
+\selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
+Physics has yet to be reduced to a formula that will fit on a piece of clothing~\cite{falk_universe_2005}.
+Biology is even more behind, but the extensive work on the mechanisms of evolution are helping to close this gap (the Price equation being the closest we have come to such an elegant formula~\cite{frank_natural_2012}).
+["even more behind"...
+NOT SURE IF THIS IS THE RIGHT WAY TO PHRASE IT] [YOU NEED TO EXPLAIN WHAT IS THE PRICE EQUATION, OTHERWISE THE PHRASE IS EMPTY]
+But evolution by itself is not sufficient to define life.
+Even the less restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
+Self-sustainment appears to be as fundamental as evolution, and relies on the transformation of matter and energy from the environment into LIVING SYSTEMS?, in other words, on growth.
+But the mathematical framing of questions regarding the self-sustaining character of living systems has not advanced as much as the mathematical formulation of evolutionary theory OR the mathematical analysis of evolution.
+
+Nevertheless, fundamental growth laws have been established for microorganisms.
+These growth laws show that, regardless of the molecular mechanisms ensuring groth control, microorganisms tend to follow the same empirical regularities when grown at steady state in different environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
+In particular, they adjust their internal molecular composition after a change in the environment, in such a way as to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
+Growth laws are a strong support for the theory of a modular organization of microorganisms~\cite{scott_emergence_2014,hartwell_molecular_1999,arkin_fast_2006,guido_bottom-up_2006}.
+They represent a big step forward towards gaining a general understanding of the physiology of microorganisms.
+They have been established at steady state, however, a state in which most microorganisms spend very little time~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,
+hobbie_microbes_2013,savageau_escherichia_1983,
+savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
+Why would microorganisms be optimal for a situation they rarely encounter?
+Would known growth laws be specific cases of more general laws allowing microorganisms to respond to changes in the environment?
+
+Our aim in this manuscript has been to establish a theoretical and experimental framework extending growth laws to a dynamical context.
+We focused on the growth law describing how ribosome abundance adapts to a change in environment so as to maximize steady-state growth in different media.
+Applying the same criterion of growth rate maximization, what is the optimal way to allocate resources during a growth transition between two different environments?
+In Chapter~\ref{chap:theory}, we used a simple self-replicator model of resource allocation, and showed that the steady state is invariant over a number of regulatory schemes: the growth rate can be maximized by measuring either the environment or the internal state of the cell.
+In order to achieve this steady-state growth rate, the expression of genes encoding the metabolic and gene expression machineries has to be set to a specific value.
+By contrast, biomass maximization during a growth transition requires information about the internal state rather than the environment.
+Moreover, expression of the metabolic and gene expression machineries is not set to one specific value, but varies with time in an on-off manner.
+What is optimal at steady state and during a growth transition is thus not the same, but this does not mean that a single regulatory system cannot meet both demands.
+
+The model developed here is an instance of proof-of-concept model~\cite{servedio_not_2014}.
+It does not necessarily aim at quantitatively predicting or controlling the behavior of microorganisms. [YOU WOULD LIKE TO HAVE AT LEAST QUALITATIVE CORRESPONDENCE...]
+By using a simple and abstract representation, it enables a convenient and tractable way of evaluating the implications of dynamical optimality, in particular its divergence from steady-state optimality.
+Is a dynamical perspective on growth laws necessary?
+Do regulatory mechanisms need to differ in a dynamical context?
+Those are the questions that were investigated by means of this model. 
+To our surprise, it additionally provided a new way of looking at the regulatory systems controlling ribosomal abundance in many bacteria.
+
+Some readers might be skeptical about the interest of using such a simple model for addressing the above questions.
+Given the profusion of knowledge and data available on biological systems, one might be tempted to introduce every known detail about the molecular implementation of a system of interest.
+One of the main drawbacks of such mechanistic models is that we quickly loose track of the big picture and the underlying assumptions.
+When including much molecular details, the models also tend to become untractable, prohibiting the use of mathematical tools like optimal control theory that are quickly overwhelmed by the sheer number of variables and parameters in the model.
+Besides, simple, abstract models make it easy for the modeler to state the critical, exploratory, and logistical assumptions that are made~\cite{servedio_not_2014}.
+[WHAT ARE "LOGISTICAL" ASSUMPTIONS?]
+For instance, an exploratory assumption of our model was to consider that the proposed strategies had to be optimal at steady state as well.
+This was used as a necessary step to reduce the space of possible solutions, and make the comparison between different strategies time-independent.
+Relaxing this assumption would require the exploration of other environmental changes beyond a simple nutrient upshift, but would probably raise interesting considerations~\cite{geisel_constitutive_2011,lopez-maury_tuning_2008,lambert_memory_2014,kussell_phenotypic_2005}.
+
+But the assumptions that are most in need of discussion are the critical assumptions, \textit{i.e.} the ones that would invalidate the model if they were falsified.
+In our case, a critical assumption is that biomass production is the main fitness factor in stable or changing environments.
+Throughout the manuscript, we did not really challenge this hypothesis, but only provided arguments about why this assumption is reasonable in some experimental conditions.
+The arguments given, however, are mostly based on results obtained with laboratory strains in laboratory conditions.
+While it is clear that microorganisms can be cultivated and are even naturally found in conditions where maximizing the growth rate ensures their persistence~\cite{edwards_silico_2001,ibarra_escherichia_2002,lewis_omic_2010,molenaar_shifts_2009}, we have no direct proof that biomass production is the fitness factor that allowed them to naturally persist on evolutionary time-scales.
+In addition, as illustrated by the quote at the beginning of this chapter, [THE QUOTE IS MORE OF A JOKE THAN A SERIOUS ARGUMENT...] one must be careful when explaining the behavior of living systems solely as a consequence of natural selection.
+Such systems are embedded in the physical world, and evolutionary arguments are not always relevant to explain an empirically observed behavior.
+[I DO NOT UNDERSTAND YOUR POINT, YOU SHOULD DEVELOP THIS PHRASE]
+
+The second part of this manuscript was dedicated to setting up an experimental framework for studying resource allocation during growth transitions.
+Much emphasis has been given to the experimental issues encountered when studying growth transitions and to come up with possible solutions.
+Fluorescent reporter proteins were used to cope with the need for \textit{in-vivo} measurements at high temporal resolution.
+Advances in microfluidics were exploited to perform steady-state-to-steady-state transitions following a change in medium, in order to buffer the effect of the pre-culture history~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}.
+Finally, the signal of interest was reconstructed through the use of Kalman smoothing, a powerful signal processing algorithm that has, as we showed, many features that make it suitable for the analysis of microscopy time-series data~\cite{kailath_linear_2000,jazwinski_stochastic_2007,kalman_new_1960}.
+Overall, the set-up still needs many improvements, though most of them are experimental issues that are just here to remind us that lab work is excessively harsh and merciless. [BOF]
+
+Chapter~\ref{chap:experiments} must not be taken as a test of the model in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see box~2 in~\cite{servedio_not_2014}).
+[WHAT IS THE DIFFERENCE?!]
+We apply the experimental set-up to empirically test the prediction of a bang-bang resource allocation profile following a nutrient upshift.
+This prediction follows directly from the critical assumption that microorganisms maximize their biomass production at steady state and during growth transitions.
+While we were not able to arrive at an unambiguous conclusion about the correctness of this prediction in this manuscript, we can draw up two possible scenarios in the wake of future improvements of the experimental set-up.
+
+In one scenario, the bang-bang or at least oscillatory nature of resource allocation during a growth transition is confirmed.
+This would be a strong argument in favor of an adaptation of microorganisms to changing environments, hence to the existence of more general dynamical growth laws.
+[THE OPPOSITE IS NOT AN ARGUMENT AGAINST ADAPTATION AND DYNAMICAL GROWTH LAWS, ONLY AGAINST THE ONE THE MODEL PREDICTS]
+In order to further establish these results, we would need additional data for a broader range of environmental conditions.
+The most straightforward extension would be to test other upshift and downshift schemes (different carbon sources, amino-acids supplementation, nitrogen instead of carbon limitation, ...).
+The experimental set-up could also be slightly modified to test a broader range of variables.
+For instance, the strain could be modified so as to express an additional fluorescent protein that would report the abundance of a key enzyme of metabolism.
+Along with the ribosomal reporter, this would allow us to test the prediction that the metabolic and gene expression machineries are expressed in anti-phase.
+
+In the other scenario, we do not observe any trace of a bang-bang resource allocation scheme, which would falsify the model.
+In that case, the questions would arise which critical assumptions are not valid.
+We also cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might in fact be critical for the predictions of the model. [LOGISTICAL?]
+Although unlikely \textit{a priori}, this might be the case for the expressions used for the macroreactions in the model (Eqs~\ref{eq:metaflux}-\ref{eq:machflux}), the definition of the volume as invariably proportional to the total mass of macromolecules (Eq.~\ref{eq:voldef}), or the fact that degradation is assumed negligible with respect to the rates of other reactions in the system.
+More probably, the assumptions regarding the cost of the regulatory system could be more critical than initially thought.
+By only comparing the regulatory schemes with respect to their benefits, we implicitly made the assumption that their costs are negligible or at least comparable.
+The costs of regulatory schemes are actually difficult to model~\cite{shachrai_cost_2010,dong_gratuitous_1995,dekel_environmental_2005}, since the cellular variables are not equally costly to measure and, as we showed for the ppGpp system, one mechanism may measure several variables at once.
+Nevertheless, relaxation of this hypothesis would definitively need to be explored before concluding that microorganisms do not optimize their biomass production during growth transitions.
+
+In any case, exploring the costs of regulatory mechanisms might uncover a fundamental link between the complexity of the regulatory schemes and the dynamics of the environment.
+As we illustrated in Chapter~\ref{chap:theory}, complex regulatory schemes only prove beneficial in a dynamical context, away from steady state.
+At the opposite extreme, one can expect the benefits of these regulatory schemes to saturate in a constantly-changing, fluctuating environment, whereas the cost of complexity would probably not saturate, since acquiring more information is expected to require additional sensing systems, diverting away an increasingly larger part of cellular resources.
+Overall, these two tendencies are expected to cross at a point that would represent an evolutionary bottleneck, beyond which the costs exceed the benefits (see~\cite{short_flows_2006} for another example of such a bottleneck).
+Interestingly, the position of the bottleneck might be dependent on the environment, resulting in a dynamic law that would link the complexity of regulatory systems to the dynamics of the environment.
+In other words, while the current growth laws consider how the physiology of the cell varies with the nutrient quality of the medium, we could add another dimension representing the time-varying availability of this nutrient in the natural environment.

### 2017-01-02 Chapters/introduction.tex
Hidde's corrections of intro and discussion

--- 2016-12-23 Chapters/introduction.tex
@@
 At every moment of our lives, we are surrounded by billions of microscopic organisms.
-They sustain themselves by collecting energy from their environment, in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
+They sustain themselves by harvesting energy from their environment, in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
 While most of our cells are kept in a stable and friendly internal environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to strongly variable conditions~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013,
@@
 Such ubiquity translates a strong potential for adaptation.
-Even the model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal track, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes~\cite{zimmer_microcosm:_2009}.
+The model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal tract, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes~\cite{zimmer_microcosm:_2009}.
 
-Microorganisms, along with all living cells, are mainly constituted of polymers (DNA, RNA, Proteins) made of many repeated sub-units (nucleotides, amino acids).
-The information about their sequences is stored in the form of genes on the DNA.
+Microorganisms, along with all living cells, are mainly constituted of polymers (DNA, RNA, proteins) consisting of many repeated subunits (nucleotides, amino acids).
+The information about their sequences is stored in the form of genes physically located on the DNA.
 Gene expression is the process by which this information is used in the synthesis of a functional final product (protein or RNA).
-The gene expression machinery (itself constituted of RNA and proteins) synthesizes the gene products after binding to gene promoters, which are target sequences on the DNA.
-In a given cell, they are roughly as many different target sequences as genes, which ensures that not all products are synthesized at the same rate.
-Furthermore, some of these products specifically bind some promoters and stimulates or represses the promoter activity, hence, the gene expression.
-They are called transcription factors and allow, in addition to other mechanisms, the regulation of cell composition.
+The gene expression machinery (itself constituted of RNA and proteins) synthesizes the gene products after binding to so-called promoters, target sequences on the DNA.
+[THIS IS NOT SUFFICIENTLY PRECISE: PROMOTERS IS ONLY FOR RNA, RBS FOR PROTEINS; GENERALIZE OR FOCUS ON TRANSCRIPTION]
+In a given cell, there are roughly as many different target sequences as genes [REF MISSING], which ensures that not all products are synthesized at the same rate.
+Furthermore, some of these products specifically bind some promoters and stimulate or repress promoter activity and hence gene expression.
+These products are called transcription factors and allow, in addition to other mechanisms, the regulation of cell composition.
 
-The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyzes the biochemical reactions allowing the cell to perform different functions.
+The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyze the biochemical reactions allowing the cell to perform different functions.
 For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reductive power, nucleotides and amino acids~\cite{schaechter_microbe_2006}.
@@
 
-Which resource allocations are optimal for a living cell?
-Systems capable of Darwinian evolution sustain themselves on the long term by \textit{fitting} to their environment~\cite{dawkins_selfish_1976}.
+Which allocation of resources is optimal for a living cell?
+Systems capable of Darwinian evolution sustain themselves in the long run by \textit{fitting} to their environment~\cite{dawkins_selfish_1976}.
+[A BIT STRANGE THAT YOU ONLY USE ITALICS FOR KEY TERMS HERE AND NOT IN REST OF THESIS]
 Living cells with the best \textit{fitness} are more likely to survive, replicate, and through long-term competition, eliminate their siblings.
@@
 Whether they apply at the species, the population, the organism, or even the gene level has long been debated~\cite{dawkins_selfish_1976}.
-An optimal resource allocation would be a gene expression scheme that maximizes the fitness factors.
+An optimal resource allocation would be a gene expression scheme maximizing the fitness factors.
 But since fitness factors have to take into account the ability to survive, reproduce, and compete with other organisms in a given (sometimes changing) environment, they are often context-dependent and highly difficult to identify.
 
-For this reason, there is no consensus about what constitutes the main fitness factors of microorganisms.
+For this reason, there is no consensus about what constitute the main fitness factors of microorganisms.
 Competition for common resources seems to favor the maximization of offspring, in the sense that a replicator with more descendants will outnumber its competitors in the long term~\cite{dawkins_selfish_1976}.
-It is unclear, however, how this general principle translates into their physiological behaviors.
-For instance, the use of genome-scale metabolic network models have shown that the metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental conditions~\cite{schuetz_systematic_2007}.
-But we can also found different strategies when the biomass production is at play.
-Microbes can either optimize their production yield (taking time to make the most of the nutrient) or their production rate (growing quickly while wasting part of the nutrient potential), the later being favored in fluctuating or poor environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuster_maximization_2008}.
+It is unclear, however, how this general principle translates into specific features of microbial physiology.
+For instance, the use of genome-scale metabolic network models has shown that metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental conditions~\cite{schuetz_systematic_2007}.
+But we can also find other strategies for biomass production. [I CORRECTED THE ENGLISH, BUT DO NOT UNDERSTAND HOW THIS FITS IN WITH THE PREVIOUS PHRASE]
+Microbes can either optimize their production yield (making the most of available nutrients at the expense of a lower growth rate) or their production rate (growing quickly while wasting part of the available nutrients), the latter being favored in fluctuating or poor [POOR, IS THAT TRUE?
+IT CONTRADICTS WHAT YOU SAY ON OVERFLOW METABOLISM BELOW...
+DATA OR THEORETICAL PREDICTION?] environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuster_maximization_2008}.
 Overall, it seems that cell optimality drifts in a multidimensional space and involves making several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
@@
 This result is not general though, and depends on the carbon source and strain used~\cite{ibarra_escherichia_2002}.
-But even in those cases, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
+But even in other situation, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
 The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not only the efficiency of the enzymes is affected, but also their abundance and thus the cellular composition.
-Overall, with the currently available knowledge and its limitations (discussed in section~\ref{chap:discussion}), growth-rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
+Overall, with the currently available knowledge and its limitations (discussed in Section~\ref{chap:discussion}) [SEC.
+1.4 or CHAP 4?], focusing on growth-rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
 
@@
 
-One of the first attempt to formalize microorganism physiology into elegant, fundamental laws was the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
-This empirical relationship states that the growth rate $\mu$ [\texttt{div.h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol.L\textsuperscript{-1}}], such that
+One of the first attempt to formalize the functioning of microbial physiology into elegant, fundamental laws was the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
+This empirical relationship states that the growth rate $\mu$ [\texttt{div h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol L\textsuperscript{-1}}], such that
 \begin{equation}
 \label{eq:monod_law}
-\mu = \mu_K \frac{C}{K_C + C}
+\mu = \mu_K \frac{C}{K_C + C},
 \end{equation}
-with $\mu_K$ and $K_C$ two constants depending on the quality of the nutrient (Fig~\ref{fig:monod_law}).
+with $\mu_K$ and $K_C$ two constants depending on the quality of the nutrient (Fig.~\ref{fig:monod_law}).
 It is remarkable that the growth rate, a parameter that depends on the use and production of thousands of proteins, can be so easily predicted by a simple relation.
@@
 This figure displays the steady-state growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
-As the concentration of glucose increases, the growth rate of \textit{E. coli} does so by following the hyperbolic relationship described in Eq~\ref{eq:monod_law}.
-Solid line is drawn with $\mu_K = 1.35$ div.h\textsuperscript{-1}, and $K_C = 0.22 \cdot 10^{-4}$ mol.L\textsuperscript{-1}.
+As the concentration of glucose increases, the growth rate of \textit{E. coli} also does by following the hyperbolic relationship described in Eq.~\ref{eq:monod_law}.
+The solid curve is obtained for $\mu_K = 1.35$ div h\textsuperscript{-1}, and $K_C = 0.22 \cdot 10^{-4}$ mol L\textsuperscript{-1}.
 }
@@
 But an obvious constraint is that at steady state, the synthesis rate of all individual components must be proportional to the growth rate in order to compensate for growth dilution~\cite{monod_growth_1949}.
-Many physiological parameters (like mass of DNA, RNA and proteins) are thus functions of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
+Many physiological parameters (like the mass of DNA, RNA and protein) are thus functions of the growth rate alone, regardless of the environmental conditions~\cite{schaechter_dependency_1958,bremer_modulation_1996}.
 These so-called growth laws were carefully measured~\cite{bremer_modulation_1996} and are still used today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
@@
 By measuring it under different environmental perturbations of the protein synthesis machinery, they built a coarse-grained model describing proteome resource allocation~\cite{scott_emergence_2014}.
-It has allowed to show that, whatever the details of the molecular implementation (differing between organisms), this growth law emerges from the underlying principles of robustness and optimization imposed by natural selection, especially growth-rate maximization in every conditions~\cite{scott_emergence_2014}.
+It has allowed to show that, whatever the details of the molecular implementation (differing between organisms), this growth law emerges from the underlying principles of robustness and optimization imposed by natural selection, especially growth-rate maximization, in all conditions~\cite{scott_emergence_2014}.
 
@@
 \textbf{The growth law of ribosomal abundance (figure reproduced from Fig~S1 in~\cite{scott_interdependence_2010}).}
-For a variety of carbon nutrients and their corresponding steady-state growth rate, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
+For a variety of carbon sources and their corresponding steady-state growth rate, the RNA/protein ratio is linearly correlated with the growth rate, a relation that holds for many species of microorganisms.
 This ratio is correlated with the fraction of ribosome-affiliated proteins, and therefore with the relative abundance of ribosomes in the cell.
@@
 It means that the properties that are independent of the cell volume or the cell mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
-It requires that the components of the cell "increases by the same factor over a time interval", which have motivated the creation of the term balanced growth~\cite{campbell_synchronization_1957}.
-Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that samplings are no longer time-sensitive~\cite{schaechter_microbe_2006}.
+It requires that the components of the cell "increase by the same factor over a time interval", which has motivated the use of the term balanced growth~\cite{campbell_synchronization_1957}.
+Experimentally, this growth scenario has been used as a standard because it improves reproducibility, in that the results do no longer depend on the precise timing of the samples~\cite{schaechter_microbe_2006}.
 It can be easily achieved in the laboratory either in continuous culture, where the substrate is continually supplied~\cite{borirak_molecular_2014}, or in batch conditions if the substrate is in high excess (Fig~\ref{fig:growth_curve}).
-This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical systems, allowing to build and analyze huge genome-scale models that encapsulate the enzymatic diversity of living cells~\cite{orth_what_2010}.
+This approach has been beneficial to mathematical modeling because considering the system at steady state reduces the complexity of the underlying dynamical system governing microbial growth, allowing genome-scale models encapsulating the enzymatic diversity of living cells to be built and analyzed~\cite{orth_what_2010}.
 
@@
 Its characteristics are time-invariant and quite robust accross conditions, which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
-This phase can be extended for hundreds of generations in continuous cultures, by the constant renewable of the medium~\cite{borirak_molecular_2014,herbert_continuous_1956,wang_robust_2010}.
+This phase can be extended for hundreds of generations in continuous cultures, by the constant renewal of the medium~\cite{borirak_molecular_2014,herbert_continuous_1956,wang_robust_2010}.
 As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
@@
 Steady-state growth requires stable conditions over a long period of time, but microorganisms live in
-environments where the key elements, such as carbon, nitrogen or phosphorus, are quickly depleted by the competitors as soon as they become available~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, a consideration that also holds for the lab strains on which growth laws are studied~\cite{savageau_escherichia_1983,savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
+environments where the key elements, such as carbon, nitrogen or phosphorus, are quickly depleted by the competitors as soon as they become available~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, a consideration that also holds for the lab strains for which growth laws have been established~\cite{savageau_escherichia_1983,savageau_demand_1998,blount_unexhausted_2015,vanelsas_survival_2011}.
 Why would microorganisms be optimal for a state they have barely encountered during their evolution?
-Are we missing something by studying in stable, unchanging conditions, systems that were in fact selected to cope with environmental variations?
+Are we missing something by studying in stable, unchanging conditions systems that were in fact selected to cope with environmental variations?
 
 \section{Problem statement}
+\label{sec:problemstatement}
 
 Microbial growth is essentially a resource allocation problem that can be summarized in simple, fundamental growth laws.
-Though established empirically, those laws are elegantly explained if we consider that microorganisms behave in way that optimizes their biomass production.
-However, these laws describe growth in stable environments, but it is hard to imagine how natural selection could have made microorganisms optimal for conditions they rarely encounter.
-This motivates the study of growth laws through a dynamical perspective, that is, in a changing environment, and leads us to the following problem statement: \textbf{Which strategies microorganisms follow to dynamically reallocate their resources after a change in the environment?}
+Though established empirically, those laws are elegantly explained if we consider that microorganisms behave in way that optimizes biomass production.
+However, these laws describe growth in stable environments and it is hard to imagine how natural selection could have resulted in microorganisms that are optimal for conditions they rarely encounter outside the laboratory.
+This motivates the study of growth laws from a dynamical perspective, that is, in a changing environment, and leads us to the following problem statement: \textbf{Which strategies microorganisms follow to dynamically reallocate their resources after a change in the environment?}
 The development of this question prompts the following investigations.
 
-What are the best strategies of resource allocation if, like for the steady-state growth laws, the biomass production is kept as the only optimization criteria?
-Through answering this question, we could assess to which extend the features of dynamical optimization differ from the steady-state optimization.
-In other words, it would test our verbal hypothesis that motivated a dynamical perspective on growth: the fact that we expect evolutionary pressure to apply differently between dynamic and stable conditions.
-This requires to translate the problem in the mathematical world, through the development of a proof-of-concept model of resource allocation in bacterial cells, and the theoretical determination of the optimal dynamical allocation of resources following a change in the environment.
+What are the best strategies of resource allocation if, like for the steady-state growth laws, we continue to assume that biomass production is the only optimization criterion?
+An answer to this question would allow us to assess to which extent the features of dynamical optimization differ from those of steady-state optimization.
+In other words, it would test our verbal hypothesis that motivated a dynamical perspective on growth: the expectation that evolutionary pressure applies differently in a changing or stable environment.
+In order to achieve this, we need to formulate our problem in mathematical terms, through the development of a proof-of-concept model of resource allocation in bacterial cells, and the theoretical determination of the optimal dynamical allocation of resources following a change in the environment.
 
-How could one measure empirically the resource allocation during a growth transition?
 The model will help us to establish actual resource allocation strategies that optimize biomass production during an environmental change.
-However, it will not be able to tell us if microorganism actually care to optimize such a criterion.
-We need though the possibility to measure biomass accumulation and changes in resource allocation when a shift is made in the environment.
-This would require to dynamically track the abundance of key macromolecules in living cells, while establishing an experimental set-up that temper the inherent variability of dynamical experimental studies of microbial growth.
+However, it will not be able to tell us if microorganism do actually optimize such a criterion.
+We therefore need to measure biomass accumulation and changes in resource allocation following a change in the environment.
+How could one measure resource allocation during a growth transition?
+In the light of the discussion in the previous section, tt requires to monitor the abundance of key macromolecules in living cells over time, while establishing an experimental set-up that allows to control the inherent variability of dynamical experimental studies of microbial growth.
 
@@
 What makes mathematical modeling so useful in microbial growth studies?
-Evolutionary biology is a good instance of field that has been resorting for a long time on mathematical modeling~\cite{servedio_not_2014}.
-The large time and population scales at which evolutionary processes occur truly impede most of the experimental work, despite some attempts on organisms with a short generation time~\cite{elena_evolution_2003}.
+Evolutionary biology is a good example of a field that has relied on mathematical modeling for a long time~\cite{servedio_not_2014}.
+The large time and population scales at which evolutionary processes occur complicate experimental work, despite some attempts on organisms with a short generation time~\cite{elena_evolution_2003}.
+[YOU MAY WISH TO CITE A RECENT REVIEW ON EXPERIMENTAL EVOLUTION, WHICH IS NO LONGER RESTRICTED TO COLI]
 Besides, microbiology is a field particularly prone to experimental study.
@@
 The last decades have strongly challenged this viewpoint: as knowledge and data on microbes accumulated, the need for a systemic understanding of biology was more and more pressing~\cite{alon_introduction_2006,kremling_systems_2013}.
+[I HAVE NOT CORRECTED THE PARAGRAPH ABOVE, BECAUSE I DO NOT UNDERSTAND VERY WELL WHAT YOU MEAN AND HOW IT RELATES TO THE OPENING QUESTION]
 
 Microbial physiology results from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation~\cite{schaechter_microbe_2006}.
-These reactions occurs on short time scales, and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations.
+These reactions occurs on short time scales [WHY SHORT?
+PROTEIN CONCENTRATION IS A SLOW VARIABLE?
+AS HALF6LIVES ARE OF THE ORDER OF 24H], and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations.
 As a consequence, even the simplest verbal hypothesis can resist direct empirical testing.
 But abstracting away complexity is the purpose of mathematical models.
-The overwhelming number of variables and time scales can be dealt with by choosing the correct framework.
-When working on a model, the microbe behaviors are abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014,mcgill_calm_2013}.
+The overwhelming number of variables and the co-existence of several different time scales can be dealt with by choosing the correct framework.
+When working on a model, microbial behavior is abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014,mcgill_calm_2013}.
+[THIS IS AMBIGUOUS, AS ONE CAN REASON IN A "LOGICAL" WAY WITHOUT MODELS; MAYBE/ "where it is easier to spell out the logical consequences of the assumptions underlying the model"]
 Predictions can be made, "unpacked" into our real world, and confronted with experimental testing.
 
-Modeling was particularly helpful to unveil how metabolic networks operate.
+Modeling has proven to be particularly helpful in unveiling how metabolic networks operate.
 For an increasing number of microorganisms, we are now able to draw quasi-exhaustive maps of their metabolic reactions.
-For instance, the most recent genome-scale reconstruction of \textit{E. coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
-Such a number of variables might seem overwhelming, and making sense of such knowledge is rarely straightforward.
+For instance, a much-used [NOT MOST RECENT] genome-scale reconstruction of \textit{E.
+coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
+The number of variables may seem overwhelming, and making sense of this information is not straightforward.
 Which reactions are important, and in which environmental conditions?
@@
 
-This apparent complexity does not impede a mathematical study of metabolic networks.
-Constraint-based modeling is a framework that abstracts away the unknown kinetics and represents the metabolic reactions by steady-state fluxes to which physico-chemical constraints can be applied, e.g.
-compartmentalization, mass conservation, molecular crowding, and thermodynamic directionality~\cite{ebrahim_cobrapy_2013}.
-In other words, through constraints imposed to the metabolic networks, it provides a space of admissible flux distributions, eliminating reactions that are unlikely to occur in the environmental condition of interest.
-This space is often further reduced by selecting the flux distributions optimizing a specified objective function (e.g., the growth rate of the cell).
-Such an approach, initially developed through flux balance analysis (FBA)~\cite{orth_comprehensive_2011,palsson_systems_2011}, has been shown to correctly predict many behaviors of the metabolic network of \textit{E.
+The complexity of the reconstructed metabolic networks does not impede their mathematical study.
+Constraint-based modeling is a framework that abstracts away the unknown kinetics and represents the metabolic reactions by steady-state fluxes to which physico-chemical constraints can be applied, e.g., compartmentalization, mass conservation, molecular crowding, and thermodynamic directionality~\cite{ebrahim_cobrapy_2013}.
+From a mathematical point of view, the model consists of a system of linear equations defining a space of admissible flux distributions that is further reduced by the above-mentioned constraints, which eliminate flux distributions that are unlikely to occur in an environmental condition of interest.
+The solutions space is often further reduced by selecting the flux distributions optimizing a specified objective function, such as the growth rate of the cell.
+This approach, called flux balance analysis (FBA)~\cite{orth_comprehensive_2011,palsson_systems_2011}, has been shown to correctly predict many behaviors of the metabolic network of \textit{E.
 coli}~\cite{varma_stoichiometric_1994,edwards_silico_2001}.
-It also proved capable of predicting its long-term recovery through genetic evolution after a gene deletion~\cite{fong_metabolic_2004}.
+It has also proved capable of predicting its long-term adaptation through evolution of the network after a gene deletion~\cite{fong_metabolic_2004}.
 
 More direct modeling approaches, in the form of large kinetic models, have also been helpful to understand growth-related processes.
-Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} showed how metabolic fluxes are sensed at different locations of the central carbon metabolism, and are integrated in a global cell response by the coupling of enzymatic and transcriptional regulations~\cite{kotte_bacterial_2010}.
-Such models are often sufficiently detailed to be used as \textit{in silico} testbed to investigate and uncover specific molecular mechanisms~\cite{peskov_kinetic_2012}.
-The most emblematic instance of that kind of approach is probably the recent whole-cell model of \textit{Mycoplasma genitalium}~\cite{karr_whole-cell_2012}.
+Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} have shown how metabolic fluxes are sensed at different locations in central carbon metabolism, and are integrated in a global cellular response by the coupling of enzymatic and transcriptional regulation~\cite{kotte_bacterial_2010}.
+Kinetic models of this type are sufficiently detailed to be used as an \textit{in silico} testbed to investigate specific molecular mechanisms and uncover their function in bringing about a cellular response~\cite{peskov_kinetic_2012}.
+The most emblematic instance of this approach is probably the recent whole-cell model of \textit{Mycoplasma genitalium}~\cite{karr_whole-cell_2012}.
 By aggregating all the available knowledge from more than 900 publications, Karr \textit{et al} constructed a dynamical model accounting for all the annotated gene functions of this pathogenic organism, and successfully used it to investigate unobserved molecular mechanisms and guide novel experimental analysis.
-While such approach is not yet applicable to all kind of organisms, it genuinely demonstrates how the increase in computational power could one day make at hand \textit{in silico} experimentation.
+While such approach is not yet applicable to many organisms, it genuinely demonstrates how the increase in computational power could one day guide \textit{in silico} experimentation.
 
-But while it represents their strength, the granularity of the models cited above is also their main weakness.
-Despite this apparently rich available knowledge, precise measurements of \textit{in-vivo} kinetic parameters still represents a technological bottleneck~\cite{park_metabolite_2016,bennett_absolute_2009,buscher_cross-platform_2009}.
+The granularity of the kinetic models cited above is their strength, but also their main weakness.
+Despite the available knowledge of molecular mechanisms, precisely measuring \textit{in-vivo} kinetic parameters still represents a technological bottleneck~\cite{park_metabolite_2016,bennett_absolute_2009,buscher_cross-platform_2009}.
+While parameter values can be collectively fitted to available data~\cite{jaqaman_linking_2006,mendes_non-linear_1998}, this is known to often produce large parameter uncertainty~\cite{cho_experimental_2003,brodersen_characterization_1987,rodriguez-fernandez_hybrid_2006}.
+This is problematic, because the predictions are often particularly sensitive to the parameter values~\cite{gutenkunst_universally_2007,ingram_network_2006,mayo_plasticity_2006}.
 Constraint-based modeling overcomes this limitation by using a mathematical approach that does not require any knowledge about the reaction kinetics.
-The drawback is that they highly depend on the constraints used to reduce the space of admissible solutions, e.g.
-the choice of the objective function that could sometimes be tricky~\cite{kauffman_advances_2003}.
-Kinetic models however are particularly sensible to missing parameters.
-While the so-called free parameters are often collectively fitted to available data~\cite{jaqaman_linking_2006,mendes_non-linear_1998}, this is known to produce large parameter uncertainty~\cite{cho_experimental_2003,brodersen_characterization_1987,rodriguez-fernandez_hybrid_2006}, which can significantly affect the predictions of the model~\cite{gutenkunst_universally_2007,ingram_network_2006,mayo_plasticity_2006}.
+The drawback is that this approach strongly depends on the constraints used to reduce the space of admissible solutions, e.g.
+the choice of the objective function that may be tricky, as discussed above~\cite{kauffman_advances_2003}.
 
-Such limitations often motivate the construction of simpler, coarse-grained models.
-The approach is dramatically different: instead of aggregating the available knowledge into a highly detailed model, the modeler is concerned about carefully filtering this information to keep the model as simple as possible.
-They generally consider a high level of abstraction of the cell functions, and are particularly valuable when looking for universal laws or principles~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}.
+The limitations of constraint-based and detailed kinetic models have motivated the construction of simpler, coarse-grained models.
+The philosophy is quite different: instead of aggregating all available knowledge into a detailed model, the modeler is concerned about carefully filtering this information to keep the model as simple as possible.
+The resulting models generally describe cellular functioning on a high level of abstraction, and are particularly valuable when looking for universal laws or principles~\cite{scott_bacterial_2011,scott_interdependence_2010,scott_emergence_2014}.
 They can take the form of minimal core models that focus on a given aspect of growth, while still abstracting away the molecular details (see \textit{e.g.}, ~\cite{spiesser_size_2012}).
-Such models can also be used as proof-of-concept to submit verbal hypothesis to the logic and rigor of mathematical reasoning~\cite{servedio_not_2014}.
+They can also be used as proof-of-concepts to submit verbal hypotheses to the logic and rigor of mathematical reasoning~\cite{servedio_not_2014}.
 For instance, a simple proof-of-concept model was used to uncover the principles leading to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
-This paradoxal and widely spread mechanism~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the trade-off between the efficiency of a pathway and the cost of synthesis of its enzymes.
-Overall, these models have several advantages for the purpose of our study: they clearly state and do not lose track of their assumptions, and are sufficiently tractable to allow access to complex mathematical tools (as will be required for dynamical optimization).
+This paradoxal and widespread phenomenon~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained as a trade-off between the efficiency of a pathway and the cost of synthesis of its enzymes.
+[THIS IS NOT SUFFICIENTLY PRECISE, YOU NEED TO MAKE THE TRADE-OFF EXPLICIT]
+Overall, these models have several advantages for the purpose of our study: they clearly state the underlying assumptions, and are sufficiently tractable to apply a variety of mathematical tools.
 
@@
 
-Either to formulate hypothesis or test predictions, data on microbial growth need to be acquired.
+Either to formulate hypotheses or test predictions, data on microbial growth need to be acquired.
 The most straightforward method is to work at the population level.
 When working with microorganisms, a clonal population of genetically identical cells can easily be obtained by inseminating a single colony into a growth medium~\cite{schaechter_microbe_2006}.
-Over time, this culture can be subject to many kind of measurements: for instance, biomass can be estimated  directly through measurements of the dry weight of samples~\cite{monod_growth_1949}, or indirectly through measurements of transmitted or scattered light across the culture~\cite{volkmer_condition-dependent_2011}, allowing to construct growth curves as presented in Fig.~\ref{fig:growth_curve}.
+Over time, this culture can be subjected to different types of measurements: for instance, biomass can be estimated  directly through measurements of the dry weight of samples~\cite{monod_growth_1949}, or indirectly through measurements of transmitted or scattered light across the culture~\cite{volkmer_condition-dependent_2011}, allowing to construct growth curves as presented in Fig.~\ref{fig:growth_curve}.
 Other population-wide parameters like the macromolecular composition~\cite{scott_interdependence_2010,scott_bacterial_2011} or the concentration of metabolites~\cite{bennett_absolute_2009} can also be evaluated.
@@
 
-Nevertheless, some questions are not solvable at the population level~\cite{davey_flow_1996,bakshi_superresolution_2012,balaban_bacterial_2004}.
-This can be the case of the structure of the inner cell~\cite{bakshi_superresolution_2012}.
+Nevertheless, some questions cannot be solved at the population level~\cite{davey_flow_1996,bakshi_superresolution_2012,balaban_bacterial_2004}.
+One example are questions about the internal structure of the cell~\cite{bakshi_superresolution_2012}.
 Bakshi \textit{et al} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerazes and DNA in living \textit{E.
 coli} cells~\cite{bakshi_superresolution_2012}.
-They showed that, despite what was previously thought, most of the translation occur far from the DNA, on free mRNA that migrate to ribosome-rich regions~\cite{bakshi_superresolution_2012}.
-But single-cell variability has also been shown to be exploited by the microbial cells themselves~\cite{balaban_bacterial_2004,booth_stress_2002,sumner_phenotypic_2002}.
+They showed that despite what was previously thought, most of the translation occurs far from the DNA, on free mRNA molecules that migrate to ribosome-rich regions~\cite{bakshi_superresolution_2012}.
+Another example concerns questions about single-cell variability~\cite{balaban_bacterial_2004,booth_stress_2002,sumner_phenotypic_2002}.
 Through growth and imaging of single cells in a microfluidic device, Balaban \textit{et al} showed that a clonal bacterial population can profit from a preexisting heterogeneity to persist when challenged by the presence of an antibiotic, a short-term mechanism that does not involve any genetic mutation~\cite{balaban_bacterial_2004}.
-On the long-term, heterogeneity has also been shown to help subpopulations to resist letal stresses in nature, letting them enough time to adapt and conquer new ecological niches~\cite{booth_stress_2002,sumner_phenotypic_2002}.
+In the long term, heterogeneity has also been shown to help subpopulations to resist lethal stresses, leaving them with enough time to adapt and conquer new ecological niches~\cite{booth_stress_2002,sumner_phenotypic_2002}.
 In fact, cell heterogeneity seems so crucial for the fitness of microorganisms that diversity-generating mechanisms have been identified~\cite{true_yeast_2000,fraser_noise_2004,raser_control_2004}.
 
-In addition to the level of measurement, the cultivation method plays a significant role in the kind of question that an experiment can answer.
-For a long time, batch growing condition has been the preferred choice of most microbiologists~[citation?].
+In addition to the level of measurement, the cultivation method plays a significant role in the kinds of question that an experiment can answer.
+For a long time, batch growth conditions have been the preferred choice of most microbiologists~[citation?].
 As illustrated in Fig.~\ref{fig:growth_curve}, the organism is inseminated in a closed-system and grows until the nutrient is depleted.
@@
 The golden age of batch culturing was attained in the last decade with the democratization of culture in microplates.
+[I DON'T THINK THIS ASSERTION IS CORRECT FROM A HISTORICAL POINT OF VIEW, MICROPLATES JUST PARALLELIZED EXISTING PRACTICE.
+THIS EXISTING PRACTICE WAS NOT ALWAYS BAD PRACTICE, SINCE ESSENTIAL FOR MANY QUESTIONS, E.G.
+DIAUXIC GROWTH]
 In parallel, dozens to hundreds of growth experiments can be performed each in less than one milliliter, while the optical density or the fluorescence of each culture is easily automatically monitored.
 This has proven extremely helpful for screening purposes and the establishment of standard libraries of gene labeling and modification~\cite{baba_construction_2006,zaslaver_comprehensive_2006}.
-Parallel growing has notably been exploited to construct roughly 4,000 single-cell knock-out mutants of \textit{Escherichia coli}, forming the Keio collection~\cite{baba_construction_2006}.
-It was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples had been distributed worldwide~\cite{yamamoto_update_2009}).
-Another approach was to build a library that allow for the quantitative estimation of gene expression.
-Here again, microplates helped building a massive library of transcription fusion of the green fluorescent protein to each of about 2,000 different promoters in E.~coli~\cite{zaslaver_comprehensive_2006}.
-Interestingly, microplate batch growing conditions are also used when applying the say constructions in the inference and analysis of gene regulatory networks~\cite{gerosa_dissecting_2013,berthoumieux_shared_2013,keren_promoters_2013,
+Parallel cultivation has notably been exploited to construct roughly 4,000 single-cell knock-out mutants of \textit{Escherichia coli}, forming the Keio collection~\cite{baba_construction_2006}.
+This mutant collection was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples [WHAT DO YOU MEAN?
+THEY CAN'T HAVE SOLD 4 M COPIES OF THE COLLECTION...] had been distributed worldwide~\cite{yamamoto_update_2009}).
+Microplates have also helped building a massive library of transcriptional fusions of the green fluorescent protein to each of about 2,000 different promoters in \textit{E.~coli}~\cite{zaslaver_comprehensive_2006}.
+Interestingly, microplate batch growing conditions have also been used when applying the above constructions in the inference and analysis of gene regulatory networks~\cite{gerosa_dissecting_2013,berthoumieux_shared_2013,keren_promoters_2013,
 ronen_assigning_2002,stefan_inference_2015}.
@@
 This could be dealt with by focusing on the beginning of the culture, but this part has been shown to strongly depend on the preculture history~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, and one is never sure when an internal steady state has been reached~\cite{myers_culture_1944}.
+[NB THE FACT THAT CULTURE CONDITIONS CHANGE IS NOT NECESSARILY "HAZARDOUS", IF ALL MEASUREMENTS USED FOR THE INFERENCE PROCESS HAVE BEEN OBTAINED UNDER THE SAME CONDITIONS.
+THE CHANGE IN CONDITIONS IS ALSO A PERTURBATION OF THE NETWORK AND POSSIBLY INFORMATIVE...]
 
 For this reason, we early observed development of continuous cultures for microbial studies~\cite{myers_culture_1944,novick_experiments_1950,herbert_continuous_1956}.
-Through permanent renewal of the growth medium, continuous cultures allow for maintenance of a chemically well controlled environment, allowing for the acquisition of reproducible and reliable data~\cite{borirak_molecular_2014,hoskisson_continuous_2005}.
-This showed particularly valuable for comparative omic analysis, for instance the analysis of changes in protein expressions~\cite{kolkman_comparative_2005} or genome-wide transcriptomic changes~\cite{boer_genome-wide_2003} that occur when \textit{Saccharomyces cerevisiae} is grown in different media.
-Continuous culture are however more difficult to set-up~\cite{novick_experiments_1950,borirak_molecular_2014}.
-They involve large quantity of medium and are, by design, expected to be more prone to contamination~\cite{novick_experiments_1950}.
-But the recent advances in microfluidic culturing has improved the former point~\cite{wang_robust_2010,balaban_bacterial_2004}.
+Through permanent renewal of the growth medium, continuous cultures allow for maintenance of a chemically well-controlled environment, allowing for the acquisition of reproducible and reliable data~\cite{borirak_molecular_2014,hoskisson_continuous_2005}.
+This has shown to be particularly valuable for comparative omic analysis, for instance the analysis of changes in protein levels~\cite{kolkman_comparative_2005} or genome-wide transcriptomic changes~\cite{boer_genome-wide_2003} that occur when \textit{Saccharomyces cerevisiae} is grown in different media.
+Continuous cultures are more difficult to set-up, however~\cite{novick_experiments_1950,borirak_molecular_2014}.
+They involve large quantities of medium and are, by design, expected to be more prone to contamination~\cite{novick_experiments_1950}.
+But the recent advances in microfluidics technology have made it possible to set up robust continuous cultures experiments on the microliter scale~\cite{wang_robust_2010,balaban_bacterial_2004}.
 For instance, the mother machine~\cite{wang_robust_2010} was developed to perform long-term growth of microorganisms while allowing for single-cell tracking of a large population.
-It showed that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
-Overall, such a device is able to set up a robust continuous culture while consuming less than 100~mL of fresh medium per day.
+It has enabled to show that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
+[NB TWO THINGS ARE A BIT MIXED UP HERE: CONTINUOUS CULTIVATION AND OBSERVATIONS ON THE SINGLE-CELL LEVELS.
+MINIBIOREACTORS ALREADY ADDRESS THE ISSUES YOU MENTION, WHILE MICROFLUIDICS ALSO COMPLICATES THINGS, FOR EXAMPLE IMAGE ANALYSIS]
 
-Finally, all these conditions can be applied to study microbial growth through dynamical scenarii~\cite{levy_strategy_2007,levy_coordination_2009,madar_promoter_2013,ehrenberg_mediumdependent_2012}.
-For instance, Levy \textit{et al}~\cite{levy_strategy_2007,levy_coordination_2009} measured the transcriptomic reorganization occuring after an environmental perturbation in yeasts grown at steady state in a continuous culture.
-They observed that the transcription regulation is affected before the growth rate of the organism, suggesting a significant role of feed-forward sensing from the environment.
-Madar \textit{et al}~\cite{madar_promoter_2013} analyzed the transcriptional reorganization occuring in \textit{E.~coli} during the lag phase through the coupling of culture in microplates and single-cell measurements \textit{via} flow cytometry of time samples.
+The experimental approaches briefly reviewed above can be applied to study microbial growth in a dynamical setting~\cite{levy_strategy_2007,levy_coordination_2009,madar_promoter_2013,ehrenberg_mediumdependent_2012}.
+For instance, Levy \textit{et al}~\cite{levy_strategy_2007,levy_coordination_2009} measured the transcriptomic reorganization occuring after an environmental perturbation in yeasts grown at steady state in continuous culture.
+[YOU NEED TO BE MORE PRECISE ON THE CULTIVATION CONDITIONS OF THE EXPERIMENT, SINCE CONTINUOUS CULTURE AND DYNAMIC ADAPTATION ARE NOT COMPATIBLE]
+They observed that the transcription levels are affected before the growth rate of the organism, suggesting a significant role of feed-forward sensing from the environment.
+Madar \textit{et al}~\cite{madar_promoter_2013} analyzed the transcriptional reorganization occuring in \textit{E.~coli} during the lag phase through the coupling of batch experiments in microplates and single-cell measurements \textit{via} flow cytometry.
 They showed that \textit{bottleneck enzymes} were produced early in the lag phase, before the cells actually switch to the production of ribosomes and general metabolic enzymes.
-As in here, the studies cited above investigated questions that could only be solved in a dynamical context.
+As in this PhD project, the studies cited above investigate questions that can only be answered in a dynamical context and they inspired the search for the most appropriate experimental approaches to answer the problem set out in Section~\ref{sec:problemstatement}
 There is still however no consensus about the growing conditions that are the more suitable for dynamical studies.
+[I WOULD LEAVE OUT THIS PHRASE, BECAUSE THERE IS NO ONE BEST CULTIVATION METHOD.
+EVERYTHING DEPENDS ON THE PRECISE QUESTIONS, JUST LIKE MODELLING...]
 By taking the best of the works presented above, a part of our study will focus on establishing such experimental conditions.
@@
 
-Thorough this study, we focus on a specific dynamical growth scenario, in particular the case of a nutrient upshift~\cite{ehrenberg_mediumdependent_2012,kjeldgaard_kinetics_1961,schaechter_patterns_1961,johnsen_control_1977}.
-Contrary to steady state growth, nutrient upshifts are frequently encountered in the life cycle of a microorganism~\cite{schaechter_microbe_2006,mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
-If combined with downshifts, they are also a good starting point if one wants to approximate more complex environments.
-However, note that in nature, nutrient upshifts generally start from stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, which involve complex mechanisms that impede most theoretical and experimental studies~\cite{stragier_molecular_1996,gonzalez-pastor_cannibalism:_2011,ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
-For this reason, we confined the problem to the study of resource allocation along a so called steady-state-to-steady-state transition, which could be seen as investigating what happens when the cells move along the growth law presented in Fig.~\ref{fig:scott_rnaprot}.
+Throughout this study, we focus on a specific dynamical growth scenario, namely the case of a nutrient upshift~\cite{ehrenberg_mediumdependent_2012,kjeldgaard_kinetics_1961,schaechter_patterns_1961,johnsen_control_1977}.
+Contrary to steady-state growth, nutrient upshifts and downshifts are frequently encountered in the life cycle of microorganisms~\cite{schaechter_microbe_2006,mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
+In combination, they also provide a good approximation of more complex environments.
+In nature, nutrient upshifts generally start from stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, a state in which complex adaptive mechanisms are at work that we would like to sidestep for this study~\cite{stragier_molecular_1996,gonzalez-pastor_cannibalism:_2011,ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
+For this reason, we confine the problem to the study of resource allocation during a so-called steady-state-to-steady-state transition, which could be seen as investigating what happens when the cells move along the growth law presented in Fig.~\ref{fig:scott_rnaprot}.
+[EXPLAIN BETTER OR LEAVE OUT THE PHRASE STARTING WITH "which could be seen..."]
 
-We start by developing in chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation which evaluates how biomass production can be maximized during such a growth transition.
-The model is an instance of self-replicator model~\cite{molenaar_shifts_2009}, but only focus on the allocation between two main sectors in the microbial cell: the metabolism that uptakes and converts nutrients in precursor metabolites, and the gene expression machinery that performs the macromolecular synthesis from these precursors.
-Though being kept as simple as possible, we show how this model is sufficient to account for the known growth laws of resource allocation at steady state~\cite{scott_interdependence_2010,scott_bacterial_2011}.
+We start by developing in Chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation which evaluates how biomass production can be maximized during an upshift from a medium with low nutrient content to a medium with high nutrient content.
+The model is an instance of a self-replicator model~\cite{molenaar_shifts_2009}, but focuses on the allocation of resources to only two sectors of the microbial cell: metabolism, taking up and converting nutrients to precursor metabolites, and gene expression, producing macromolecular from the precursors.
+Though the model is kept as simple as possible, we show how it is sufficient to account for two known growth laws of resource allocation at steady state~\cite{scott_interdependence_2010,scott_bacterial_2011}.
 
 Using this model, we pose the problem of dynamical optimization as an optimal control problem~\cite{stengel_optimal_1994}.
-Like for steady-state optimization, the cell objective is kept as being the biomass production, but we look for the dynamical resource allocation that maximizes it over an upshift scenario.
-Through the combination of Pontryagin Maximum Principle~\cite{carlson_infinite_1991} and numerical optimization~\cite{bonnans_bocop_2012}, we identify a solution involving bang-bang control~\cite{stengel_optimal_1994}.
-The mathematical gold standard for resources allocation is thus a scenario in which resources are always fully allocated to a single sector in a switchable manner.
+Like for steady-state optimization, the microbial cells are assumed to maximize biomass production, but we adapt the criterion to the dynamical context of an upshift scenario.
+Through the combination of the Pontryagin Maximum Principle~\cite{carlson_infinite_1991} and numerical optimization~\cite{bonnans_bocop_2012}, we identify a solution involving bang-bang control~\cite{stengel_optimal_1994}.
+The mathematical gold standard for resources allocation during a nutrient upshift is thus a scenario in which resources are always fully allocated to a single sector in a switchable manner.
 
@@
 We identify the system variables that need to be sensed in order to optimize resource allocation.
-We show that regulatory schemes that are equivalent at steady state significantly differ during an upshift scenario.
-In particular, feedback from the inner state of the cell is an information extremely more valuable than sensing directly the environmental variations (contrary to what is usually assumed, see e.g.
-\cite{levy_coordination_2009}).
-Furthermore, by just sensing at the same time the precursor and the gene expression machinery concentrations, the system is theoretically able to perform a near-perfect transition.
-We investigate if such a strategy could be implemented in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} to be a likely candidate.
-Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should allocate their resources towards the gene expression machinery in an on-off manner.
+We show that regulatory schemes that are equivalent at steady state yield significantly different outcomes during an upshift scenario.
+In particular, feedback from the intracellular state turns out to be much more valuable than information from the environment (contrary to what has been argued by some authors, e.g., see \cite{levy_coordination_2009}).
+Furthermore, by sensing at the same time the concentrations of precursor and gene expression machinery, the system is theoretically able to perform a near-optimal transition.
+We investigate if such a strategy could be active in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} as a possible molecular implementation of the strategy.
+Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should dynamically allocate their resources to the gene expression machinery in an on-off manner.
 
-In chapter~\ref{chap:experiments}, we address the challenging problem of verifying this prediction in \textit{Escherichia coli}.
-It requires the development of a set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
-The time scales identified in chapter~\ref{chap:theory} motivates the use of \textit{in vivo}, high-frequency, single cell measurements of the gene expression machinery concentration.
-Inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we construct a strain that displays a gfp-tagged S2 ribosomal subunit.
-We place this strain in a microfluidic device called the mother machine~\cite{wang_robust_2010}.
-While it was originally developed for studies of long-term steady-state growth, we actually use it to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source).
+In Chapter~\ref{chap:experiments}, we address the challenging problem of experimentally verifying the above prediction in \textit{Escherichia coli}.
+It requires the development of an experimental set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
+The time scales identified in Chapter~\ref{chap:theory} as well as the predicted on-off pattern motivate the use of \textit{in-vivo}, high-frequency, single-cell measurements of the concentration of the gene expression machinery.
+Inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we construct a strain with a GFP-tagged S2 ribosomal subunit, grow this strain in the mother machine, and monitor expression of the reporter gene using fluorescence microscopy~\cite{wang_robust_2010}.
+While originally developed for studies of long-term steady-state growth, we use the mother machine here to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source).
 
-The set-up is applied to the reconstruction of the resource allocation control during the upshift scenario.
-A preliminary analysis of the microscopy acquisition provided time series of the bacterial length and the single-cell concentration of fluorescence.
-We show that these measurements are sufficient for the intended signal reconstructions, and use Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007} to reconstruct the growth rate and the resource allocation over time.
-By exploiting the Bayesian logic behind this well-known technique (though rarely encountered in biology), we show it to be particularly suitable for microscopic data analysis and the reconstruction of complex gene expression signals.
-While the data acquired did not allow to conclude on the tested prediction, we present several realistic improvements that should allow to conclude in the near future.
+This experimental set-up is applied to test the theoretical predictions obtained from the optimal control analysis of the resource allocation model.
+A preliminary analysis of the acquired microscopy images provides time-series data of the bacterial length and the fluorescence density in around 60 individual cells.
+We show that these measurements are sufficient for the intended signal reconstructions, and use Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007} to reconstruct the growth rate and the resource allocation profile over time.
+Kalman smoothing, while well-known in engineering has been rarely used in quantitative biology, but turns out to be particularly suitable for the purpose of the model-based analysis of gene expression data in single cells.
+While the available preliminary data did not allow to unambiguously conclude if the model prediction is true or false, we present several realistic improvements for further work in the near future.
+
+[IN THIS SECTION ONE EXPECTS TO FIND THE APPROACH YOU FOLLOW AND YOUR MOTIVATION FOR FOLLOWING THIS APPROACH, RATHER THAN THE RESULTS. BUT AT THIS STAGE, YOU SHOULD LEAVE THE SECTION AS IS]

### 2017-01-02 Chapters/introduction.tex
Review hidde modifications of Introduction

@@
 Gene expression is the process by which this information is used in the synthesis of a functional final product (protein or RNA).
-The gene expression machinery (itself constituted of RNA and proteins) synthesizes the gene products after binding to so-called promoters, target sequences on the DNA.
-[THIS IS NOT SUFFICIENTLY PRECISE: PROMOTERS IS ONLY FOR RNA, RBS FOR PROTEINS; GENERALIZE OR FOCUS ON TRANSCRIPTION]
-In a given cell, there are roughly as many different target sequences as genes [REF MISSING], which ensures that not all products are synthesized at the same rate.
+The gene expression machinery (itself constituted of RNA and proteins) synthesizes gene products after binding to target sequences either at the state of transcription (promoter on the DNA), or translation (ribosome binding site on the messaging RNA).
+In a given cell, there are roughly as many different target sequences as genes~\cite{keseler_ecocyc:_2013}, which ensures that not all products are synthesized at the same rate.
 Furthermore, some of these products specifically bind some promoters and stimulate or repress promoter activity and hence gene expression.
@@
 Which allocation of resources is optimal for a living cell?
-Systems capable of Darwinian evolution sustain themselves in the long run by \textit{fitting} to their environment~\cite{dawkins_selfish_1976}.
-[A BIT STRANGE THAT YOU ONLY USE ITALICS FOR KEY TERMS HERE AND NOT IN REST OF THESIS]
-Living cells with the best \textit{fitness} are more likely to survive, replicate, and through long-term competition, eliminate their siblings.
-The optimization criteria that quantitatively measure \textit{fitness} are called \textit{fitness factors}.
+Systems capable of Darwinian evolution sustain themselves in the long run by fitting to their environment~\cite{dawkins_selfish_1976}.
+Living cells with the best fitness are more likely to survive, replicate, and through long-term competition, eliminate their siblings.
+The optimization criteria that quantitatively measure fitness are called fitness factors.
 Whether they apply at the species, the population, the organism, or even the gene level has long been debated~\cite{dawkins_selfish_1976}.
@@
 For instance, the use of genome-scale metabolic network models has shown that metabolism maximizes the production either of biomass or available energy (ATP) depending on the environmental conditions~\cite{schuetz_systematic_2007}.
-But we can also find other strategies for biomass production. [I CORRECTED THE ENGLISH, BUT DO NOT UNDERSTAND HOW THIS FITS IN WITH THE PREVIOUS PHRASE]
-Microbes can either optimize their production yield (making the most of available nutrients at the expense of a lower growth rate) or their production rate (growing quickly while wasting part of the available nutrients), the latter being favored in fluctuating or poor [POOR, IS THAT TRUE?
-IT CONTRADICTS WHAT YOU SAY ON OVERFLOW METABOLISM BELOW...
-DATA OR THEORETICAL PREDICTION?] environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuster_maximization_2008}.
+Microbes can also either optimize their production yield (making the most of available nutrients at the expense of a lower growth rate) or their production rate (growing quickly while wasting part of the available nutrients), the latter being favored in fluctuating environments~\cite{frank_tradeoff_2010,maclean_tragedy_2008,schuster_maximization_2008}.
 Overall, it seems that cell optimality drifts in a multidimensional space and involves making several trade-offs that strongly depend on the environmental context~\cite{schuetz_multidimensional_2012}.
@@
 This result is not general though, and depends on the carbon source and strain used~\cite{ibarra_escherichia_2002}.
-But even in other situation, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
+But even in other situations, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
 The mutants obtained present modications in the regulation of the synthesis of metabolic enzymes~\cite{lewis_omic_2010}, indicating that not only the efficiency of the enzymes is affected, but also their abundance and thus the cellular composition.
-Overall, with the currently available knowledge and its limitations (discussed in Section~\ref{chap:discussion}) [SEC.
-1.4 or CHAP 4?], focusing on growth-rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
+Overall, with the currently available knowledge and its limitations (discussed in Chapter~\ref{chap:discussion}), focusing on growth-rate maximization sounds like a rational choice to understand the growth strategies of microorganisms~\cite{molenaar_shifts_2009}.
 
@@
 One of the first attempt to formalize the functioning of microbial physiology into elegant, fundamental laws was the discovery of Monod's law more than sixty years ago~\cite{monod_growth_1949}.
-This empirical relationship states that the growth rate $\mu$ [\texttt{div h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol L\textsuperscript{-1}}], such that
+This empirical relationship states that the growth rate $\mu$ [\texttt{div~h\textsuperscript{-1}}] of \textit{Escherichia coli} is a hyperbolic function of the concentration of the limiting nutrient $C$ [\texttt{mol~L\textsuperscript{-1}}], such that
 \begin{equation}
@@
 These so-called growth laws were carefully measured~\cite{bremer_modulation_1996} and are still used today in the quantitative understanding of growth control in microorganisms~\cite{ehrenberg_mediumdependent_2012}.
-Recently, the topic was revitalized through the work of Scott \textit{et al}~\cite{scott_bacterial_2011}.
+Recently, the topic was revitalized through the work of Scott \textit{et al.}~\cite{scott_bacterial_2011}.
 They focused on the ribosome concentration, a physiological parameter that was long known to vary linearly with the growth rate in microorganisms (Fig~\ref{fig:scott_rnaprot}).
@@
 How could one measure resource allocation during a growth transition?
-In the light of the discussion in the previous section, tt requires to monitor the abundance of key macromolecules in living cells over time, while establishing an experimental set-up that allows to control the inherent variability of dynamical experimental studies of microbial growth.
+In the light of the discussion in the previous section, it requires to monitor the abundance of key macromolecules in living cells over time, while establishing an experimental set-up that allows to control the inherent variability of dynamical experimental studies of microbial growth.
 
@@
 What makes mathematical modeling so useful in microbial growth studies?
-Evolutionary biology is a good example of a field that has relied on mathematical modeling for a long time~\cite{servedio_not_2014}.
-The large time and population scales at which evolutionary processes occur complicate experimental work, despite some attempts on organisms with a short generation time~\cite{elena_evolution_2003}.
-[YOU MAY WISH TO CITE A RECENT REVIEW ON EXPERIMENTAL EVOLUTION, WHICH IS NO LONGER RESTRICTED TO COLI]
-Besides, microbiology is a field particularly prone to experimental study.
-A growth curve as presented in Fig.~\ref{fig:growth_curve} can be made in a day, so the limitations of evolutionary biology should not apply to microbial growth studies.
-Why care to mimic reality on a computer when we can just directly bend it to our will~\cite{hillis_why_1993}?
-The last decades have strongly challenged this viewpoint: as knowledge and data on microbes accumulated, the need for a systemic understanding of biology was more and more pressing~\cite{alon_introduction_2006,kremling_systems_2013}.
-[I HAVE NOT CORRECTED THE PARAGRAPH ABOVE, BECAUSE I DO NOT UNDERSTAND VERY WELL WHAT YOU MEAN AND HOW IT RELATES TO THE OPENING QUESTION]
-
+%Evolutionary biology is a good example of a field that has relied on mathematical modeling for a long time~\cite{servedio_not_2014}.
+%The large time and population scales at which evolutionary processes occur complicate experimental work, despite several attempts on organisms with a short generation time~\cite{barrick_genome_2013,elena_evolution_2003,kassen_experimental_2002,schneider_dynamics_2004}.
+%Besides, microbiology is a field particularly prone to experimental study.
+%A growth curve as presented in Fig.~\ref{fig:growth_curve} can be made in a day, so the limitations of evolutionary biology should not apply to microbial growth studies.
+%But as knowledge and experimental data on microbes accumulates, the need for a systemic understanding of biology is more and more pressing~\cite{alon_introduction_2006,kremling_systems_2013,hillis_why_1993}.
 Microbial physiology results from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation~\cite{schaechter_microbe_2006}.
-These reactions occurs on short time scales [WHY SHORT?
-PROTEIN CONCENTRATION IS A SLOW VARIABLE?
-AS HALF6LIVES ARE OF THE ORDER OF 24H], and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations.
+These reactions occurs on a wide range of time scales (shorter than 1~second for the metabolism, to more than 1~day for the degradation of stable proteins~\cite{schaechter_microbe_2006}), and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations~\cite{dawkins_selfish_1976}.
 As a consequence, even the simplest verbal hypothesis can resist direct empirical testing.
@@
 The overwhelming number of variables and the co-existence of several different time scales can be dealt with by choosing the correct framework.
-When working on a model, microbial behavior is abstracted into a world of clearly stated rules, where logical reasoning can be applied~\cite{servedio_not_2014,mcgill_calm_2013}.
-[THIS IS AMBIGUOUS, AS ONE CAN REASON IN A "LOGICAL" WAY WITHOUT MODELS; MAYBE/ "where it is easier to spell out the logical consequences of the assumptions underlying the model"]
+When working on a model, microbial behavior is abstracted into a world of clearly stated rules, where it is easier to spell out the logical consequences of the underlying assumptions~\cite{servedio_not_2014,mcgill_calm_2013}.
 Predictions can be made, "unpacked" into our real world, and confronted with experimental testing.
@@
 For an increasing number of microorganisms, we are now able to draw quasi-exhaustive maps of their metabolic reactions.
-For instance, a much-used [NOT MOST RECENT] genome-scale reconstruction of \textit{E.
-coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
+For instance, a much-used genome-scale reconstruction of \textit{E. coli} metabolism contains 1366 genes, 2251 metabolic reactions, and 1136 unique metabolites~\cite{orth_comprehensive_2011}.
 The number of variables may seem overwhelming, and making sense of this information is not straightforward.
@@
 From a mathematical point of view, the model consists of a system of linear equations defining a space of admissible flux distributions that is further reduced by the above-mentioned constraints, which eliminate flux distributions that are unlikely to occur in an environmental condition of interest.
-The solutions space is often further reduced by selecting the flux distributions optimizing a specified objective function, such as the growth rate of the cell.
+The solution space is often further reduced by selecting the flux distributions optimizing a specified objective function, such as the growth rate of the cell.
 This approach, called flux balance analysis (FBA)~\cite{orth_comprehensive_2011,palsson_systems_2011}, has been shown to correctly predict many behaviors of the metabolic network of \textit{E.
@@
 More direct modeling approaches, in the form of large kinetic models, have also been helpful to understand growth-related processes.
-Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al} have shown how metabolic fluxes are sensed at different locations in central carbon metabolism, and are integrated in a global cellular response by the coupling of enzymatic and transcriptional regulation~\cite{kotte_bacterial_2010}.
+Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al.} have shown how metabolic fluxes are sensed at different locations in central carbon metabolism, and are integrated in a global cellular response by the coupling of enzymatic and transcriptional regulation~\cite{kotte_bacterial_2010}.
 Kinetic models of this type are sufficiently detailed to be used as an \textit{in silico} testbed to investigate specific molecular mechanisms and uncover their function in bringing about a cellular response~\cite{peskov_kinetic_2012}.
 The most emblematic instance of this approach is probably the recent whole-cell model of \textit{Mycoplasma genitalium}~\cite{karr_whole-cell_2012}.
-By aggregating all the available knowledge from more than 900 publications, Karr \textit{et al} constructed a dynamical model accounting for all the annotated gene functions of this pathogenic organism, and successfully used it to investigate unobserved molecular mechanisms and guide novel experimental analysis.
+By aggregating all the available knowledge from more than 900 publications, Karr \textit{et al.} constructed a dynamical model accounting for all the annotated gene functions of this pathogenic organism, and successfully used it to investigate unobserved molecular mechanisms and guide novel experimental analysis.
 While such approach is not yet applicable to many organisms, it genuinely demonstrates how the increase in computational power could one day guide \textit{in silico} experimentation.
@@
 For instance, a simple proof-of-concept model was used to uncover the principles leading to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
-This paradoxal and widespread phenomenon~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained as a trade-off between the efficiency of a pathway and the cost of synthesis of its enzymes.
-[THIS IS NOT SUFFICIENTLY PRECISE, YOU NEED TO MAKE THE TRADE-OFF EXPLICIT]
+This paradoxal and widespread phenomenon~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the fact that high yield pathways also require the synthesis of more enzymes, revealing the occurrence of a cost-benefit trade-off producing the switch when nutrients are no longer the limiting factor.
 Overall, these models have several advantages for the purpose of our study: they clearly state the underlying assumptions, and are sufficiently tractable to apply a variety of mathematical tools.
@@
 One example are questions about the internal structure of the cell~\cite{bakshi_superresolution_2012}.
-Bakshi \textit{et al} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerazes and DNA in living \textit{E.
+Bakshi \textit{et al.} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerazes and DNA in living \textit{E.
 coli} cells~\cite{bakshi_superresolution_2012}.
@@
 Another example concerns questions about single-cell variability~\cite{balaban_bacterial_2004,booth_stress_2002,sumner_phenotypic_2002}.
-Through growth and imaging of single cells in a microfluidic device, Balaban \textit{et al} showed that a clonal bacterial population can profit from a preexisting heterogeneity to persist when challenged by the presence of an antibiotic, a short-term mechanism that does not involve any genetic mutation~\cite{balaban_bacterial_2004}.
+Through growth and imaging of single cells in a microfluidic device, Balaban \textit{et al.} showed that a clonal bacterial population can profit from a preexisting heterogeneity to persist when challenged by the presence of an antibiotic, a short-term mechanism that does not involve any genetic mutation~\cite{balaban_bacterial_2004}.
 In the long term, heterogeneity has also been shown to help subpopulations to resist lethal stresses, leaving them with enough time to adapt and conquer new ecological niches~\cite{booth_stress_2002,sumner_phenotypic_2002}.
@@
 In addition to the level of measurement, the cultivation method plays a significant role in the kinds of question that an experiment can answer.
-For a long time, batch growth conditions have been the preferred choice of most microbiologists~[citation?].
+With the advent of molecular genetics in the 1970s and 1980s, batch growth conditions have been the preferred choice of most microbiologists~\cite{hoskisson_continuous_2005}.
 As illustrated in Fig.~\ref{fig:growth_curve}, the organism is inseminated in a closed-system and grows until the nutrient is depleted.
-In a sense, such a condition is close to what microbes encounter in nature, where key nutrients are only available for a short time and quickly depleted~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
-The golden age of batch culturing was attained in the last decade with the democratization of culture in microplates.
-[I DON'T THINK THIS ASSERTION IS CORRECT FROM A HISTORICAL POINT OF VIEW, MICROPLATES JUST PARALLELIZED EXISTING PRACTICE.
-THIS EXISTING PRACTICE WAS NOT ALWAYS BAD PRACTICE, SINCE ESSENTIAL FOR MANY QUESTIONS, E.G.
-DIAUXIC GROWTH]
-In parallel, dozens to hundreds of growth experiments can be performed each in less than one milliliter, while the optical density or the fluorescence of each culture is easily automatically monitored.
+In a sense, such a condition is close to what microbes encounter in nature, where key nutrients are only available for a short time and quickly depleted~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, and is essential to study many questions, \textit{e.g.} diauxic growth~\cite{kremling_understanding_2015}.
+Currently, a strong advantage of batch culturing is the intense parallelization that can be attained using microplates.
+Indeed, microplates allow to perform dozens to hundreds of growth experiments, each in less than one milliliter, while the optical density or the fluorescence of each culture is automatically monitored.
 This has proven extremely helpful for screening purposes and the establishment of standard libraries of gene labeling and modification~\cite{baba_construction_2006,zaslaver_comprehensive_2006}.
 Parallel cultivation has notably been exploited to construct roughly 4,000 single-cell knock-out mutants of \textit{Escherichia coli}, forming the Keio collection~\cite{baba_construction_2006}.
-This mutant collection was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples [WHAT DO YOU MEAN?
-THEY CAN'T HAVE SOLD 4 M COPIES OF THE COLLECTION...] had been distributed worldwide~\cite{yamamoto_update_2009}).
+This mutant collection was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples issued from this collection had been shared worldwide~\cite{yamamoto_update_2009}).
 Microplates have also helped building a massive library of transcriptional fusions of the green fluorescent protein to each of about 2,000 different promoters in \textit{E.~coli}~\cite{zaslaver_comprehensive_2006}.
@@
 ronen_assigning_2002,stefan_inference_2015}.
-But the later use is hazardous, because the main drawback of the batch condition is that the culturing system is poorly controlled.
+But the later use can be hazardous, because the main drawback of the batch condition is that the culturing system is poorly controlled.
 Changes of key chemical parameters (\textit{e.g.} pH, pO\textsubscript{2}) have been shown to occur during the whole growth curve~\cite{bekker_changes_2007}. 
@@
 This could be dealt with by focusing on the beginning of the culture, but this part has been shown to strongly depend on the preculture history~\cite{ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967}, and one is never sure when an internal steady state has been reached~\cite{myers_culture_1944}.
-[NB THE FACT THAT CULTURE CONDITIONS CHANGE IS NOT NECESSARILY "HAZARDOUS", IF ALL MEASUREMENTS USED FOR THE INFERENCE PROCESS HAVE BEEN OBTAINED UNDER THE SAME CONDITIONS.
+%[NB THE FACT THAT CULTURE CONDITIONS CHANGE IS NOT NECESSARILY "HAZARDOUS", IF ALL MEASUREMENTS USED FOR THE INFERENCE PROCESS HAVE BEEN OBTAINED UNDER THE SAME CONDITIONS.
 THE CHANGE IN CONDITIONS IS ALSO A PERTURBATION OF THE NETWORK AND POSSIBLY INFORMATIVE...]
+% Nils: Indeed, but I need to make a point here in order to motivate why we used complicated culture.
 
@@
 Continuous cultures are more difficult to set-up, however~\cite{novick_experiments_1950,borirak_molecular_2014}.
-They involve large quantities of medium and are, by design, expected to be more prone to contamination~\cite{novick_experiments_1950}.
-But the recent advances in microfluidics technology have made it possible to set up robust continuous cultures experiments on the microliter scale~\cite{wang_robust_2010,balaban_bacterial_2004}.
-For instance, the mother machine~\cite{wang_robust_2010} was developed to perform long-term growth of microorganisms while allowing for single-cell tracking of a large population.
+Even if the working volume can be reduced~\cite{betts_miniature_2006}, they usually consume large quantities of medium and are, by design, expected to be more prone to contamination~\cite{novick_experiments_1950}.
+But recent advances in microfluidic technology have made it possible to set up robust continuous cultures experiments on the microliter scale~\cite{wang_robust_2010,balaban_bacterial_2004}.
+For instance, the mother machine~\cite{wang_robust_2010} allows to perform long-term growth of microorganisms
+in continuous culture using only a few milliliters of fresh medium per hour.
 It has enabled to show that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
-[NB TWO THINGS ARE A BIT MIXED UP HERE: CONTINUOUS CULTIVATION AND OBSERVATIONS ON THE SINGLE-CELL LEVELS.
-MINIBIOREACTORS ALREADY ADDRESS THE ISSUES YOU MENTION, WHILE MICROFLUIDICS ALSO COMPLICATES THINGS, FOR EXAMPLE IMAGE ANALYSIS]
+It is however important to note that at those scales, growth can only be monitored through microscopy analysis, which could unnecessarily complicate studies that do not rely on single-cell measurements.
 
 The experimental approaches briefly reviewed above can be applied to study microbial growth in a dynamical setting~\cite{levy_strategy_2007,levy_coordination_2009,madar_promoter_2013,ehrenberg_mediumdependent_2012}.
-For instance, Levy \textit{et al}~\cite{levy_strategy_2007,levy_coordination_2009} measured the transcriptomic reorganization occuring after an environmental perturbation in yeasts grown at steady state in continuous culture.
-[YOU NEED TO BE MORE PRECISE ON THE CULTIVATION CONDITIONS OF THE EXPERIMENT, SINCE CONTINUOUS CULTURE AND DYNAMIC ADAPTATION ARE NOT COMPATIBLE]
+For instance, Levy \textit{et al.}~\cite{levy_strategy_2007,levy_coordination_2009} applied pulse-like environmental perturbations to a continuous culture of yeasts, and measured how transcriptomic reorganization occurs.
 They observed that the transcription levels are affected before the growth rate of the organism, suggesting a significant role of feed-forward sensing from the environment.
-Madar \textit{et al}~\cite{madar_promoter_2013} analyzed the transcriptional reorganization occuring in \textit{E.~coli} during the lag phase through the coupling of batch experiments in microplates and single-cell measurements \textit{via} flow cytometry.
+Madar \textit{et al.}~\cite{madar_promoter_2013} analyzed the transcriptional reorganization occuring in \textit{E.~coli} during the lag phase through the coupling of batch experiments in microplates and single-cell measurements \textit{via} flow cytometry.
 They showed that \textit{bottleneck enzymes} were produced early in the lag phase, before the cells actually switch to the production of ribosomes and general metabolic enzymes.
-As in this PhD project, the studies cited above investigate questions that can only be answered in a dynamical context and they inspired the search for the most appropriate experimental approaches to answer the problem set out in Section~\ref{sec:problemstatement}
-There is still however no consensus about the growing conditions that are the more suitable for dynamical studies.
-[I WOULD LEAVE OUT THIS PHRASE, BECAUSE THERE IS NO ONE BEST CULTIVATION METHOD.
-EVERYTHING DEPENDS ON THE PRECISE QUESTIONS, JUST LIKE MODELLING...]
+As in this PhD project, the studies cited above investigate questions that can only be answered in a dynamical context and they inspired the search for the most appropriate experimental approaches to answer the problem set out in Section~\ref{sec:problemstatement}.
 By taking the best of the works presented above, a part of our study will focus on establishing such experimental conditions.
@@
 In nature, nutrient upshifts generally start from stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, a state in which complex adaptive mechanisms are at work that we would like to sidestep for this study~\cite{stragier_molecular_1996,gonzalez-pastor_cannibalism:_2011,ng_damage_1962,dufrenne_effect_1997,shaw_effect_1967,mcmeekin_predictive_2002,cheroutre-vialette_application_2002}.
-For this reason, we confine the problem to the study of resource allocation during a so-called steady-state-to-steady-state transition, which could be seen as investigating what happens when the cells move along the growth law presented in Fig.~\ref{fig:scott_rnaprot}.
-[EXPLAIN BETTER OR LEAVE OUT THE PHRASE STARTING WITH "which could be seen..."]
+For this reason, we confine the problem to the study of resource allocation during a so-called steady-state-to-steady-state transition.
+While we expect the starting and ending conditions to be on the growth law presented in Fig.~\ref{fig:scott_rnaprot}, we currently have no idea of what happens in the time between.
 
 We start by developing in Chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation which evaluates how biomass production can be maximized during an upshift from a medium with low nutrient content to a medium with high nutrient content.
-The model is an instance of a self-replicator model~\cite{molenaar_shifts_2009}, but focuses on the allocation of resources to only two sectors of the microbial cell: metabolism, taking up and converting nutrients to precursor metabolites, and gene expression, producing macromolecular from the precursors.
-Though the model is kept as simple as possible, we show how it is sufficient to account for two known growth laws of resource allocation at steady state~\cite{scott_interdependence_2010,scott_bacterial_2011}.
+The model is an instance of a self-replicator model~\cite{molenaar_shifts_2009}, but focuses on the allocation of resources to only two sectors of the microbial cell: metabolism, taking up and converting nutrients to precursor metabolites, and gene expression, producing macromolecules from the precursors.
+The model is kept as simple as possible in order to make sure it stays mathematically tractable in a dynamical context.
+A necessary step is however to verify that, at steady-state, the model account for known growth laws of resource allocation~\cite{scott_interdependence_2010,scott_bacterial_2011}.
 
 Using this model, we pose the problem of dynamical optimization as an optimal control problem~\cite{stengel_optimal_1994}.
-Like for steady-state optimization, the microbial cells are assumed to maximize biomass production, but we adapt the criterion to the dynamical context of an upshift scenario.
-Through the combination of the Pontryagin Maximum Principle~\cite{carlson_infinite_1991} and numerical optimization~\cite{bonnans_bocop_2012}, we identify a solution involving bang-bang control~\cite{stengel_optimal_1994}.
-The mathematical gold standard for resources allocation during a nutrient upshift is thus a scenario in which resources are always fully allocated to a single sector in a switchable manner.
-
-This theoretical solution is then used to explore possible regulatory strategies.
-We identify the system variables that need to be sensed in order to optimize resource allocation.
-We show that regulatory schemes that are equivalent at steady state yield significantly different outcomes during an upshift scenario.
-In particular, feedback from the intracellular state turns out to be much more valuable than information from the environment (contrary to what has been argued by some authors, e.g., see \cite{levy_coordination_2009}).
-Furthermore, by sensing at the same time the concentrations of precursor and gene expression machinery, the system is theoretically able to perform a near-optimal transition.
-We investigate if such a strategy could be active in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} as a possible molecular implementation of the strategy.
-Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should dynamically allocate their resources to the gene expression machinery in an on-off manner.
+To comply with what we know at steady state, the microbial cells are assumed to maximize biomass production, but we adapt the criterion to the dynamical context of an upshift scenario.
+By using a combination of analytical~\cite{carlson_infinite_1991} and numerical optimization~\cite{bonnans_bocop_2012}, we aim to identify a mathematical upper bound for the biomass produced~\cite{stengel_optimal_1994}.
+This theoretical solution can then be used to explore possible regulatory strategies.
+In particular, we use the bang-bang~\cite{stengel_optimal_1994} control solution found as a benchmark to identify the system variables that need to be sensed in order to optimize resource allocation.
+This provides a common scale on which the outcomes of different regulatory schemes can be compared during an upshift scenario.
+This allows us to identify one or several strategies that clearly outperform all the others, giving us testable predictions.
+Specifically, we show that feedback from the intracellular state turns out to be much more valuable than information from the environment.
+We investigate if such a strategy could be active in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} as a possible molecular implementation of that should allow the synthesis of ribosomes in a switch-like manner during a nutrient upshift.
+Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should dynamically allocate their resources to the gene expression machinery in an on-off manner, making an experimentally testable prediction.
 
@@
 It requires the development of an experimental set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
-The time scales identified in Chapter~\ref{chap:theory} as well as the predicted on-off pattern motivate the use of \textit{in-vivo}, high-frequency, single-cell measurements of the concentration of the gene expression machinery.
-Inspired by the work of Bakshi \textit{et al}~\cite{bakshi_superresolution_2012}, we construct a strain with a GFP-tagged S2 ribosomal subunit, grow this strain in the mother machine, and monitor expression of the reporter gene using fluorescence microscopy~\cite{wang_robust_2010}.
-While originally developed for studies of long-term steady-state growth, we use the mother machine here to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source).
+From the model developed in Chapter~\ref{chap:theory}, we identified that the predicted on-off pattern should occur on a short time scale after the transition (a couple of generations).
+This motivates the use of \textit{in-vivo}, high-frequency, single-cell measurements of the concentration of the gene expression machinery.
+Inspired by the work of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, we construct a strain with a GFP-tagged S2 ribosomal subunit, grow this strain in the mother machine, and monitor expression of the reporter gene using fluorescence microscopy~\cite{wang_robust_2010}.
+While originally developed for studies of long-term steady-state growth, we use the mother machine here to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source), in order to get a significant difference in growth rates between the two media~\cite{andersen_are_1980}.
 
-This experimental set-up is applied to test the theoretical predictions obtained from the optimal control analysis of the resource allocation model.
-A preliminary analysis of the acquired microscopy images provides time-series data of the bacterial length and the fluorescence density in around 60 individual cells.
-We show that these measurements are sufficient for the intended signal reconstructions, and use Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007} to reconstruct the growth rate and the resource allocation profile over time.
-Kalman smoothing, while well-known in engineering has been rarely used in quantitative biology, but turns out to be particularly suitable for the purpose of the model-based analysis of gene expression data in single cells.
-While the available preliminary data did not allow to unambiguously conclude if the model prediction is true or false, we present several realistic improvements for further work in the near future.
-
-[IN THIS SECTION ONE EXPECTS TO FIND THE APPROACH YOU FOLLOW AND YOUR MOTIVATION FOR FOLLOWING THIS APPROACH, RATHER THAN THE RESULTS. BUT AT THIS STAGE, YOU SHOULD LEAVE THE SECTION AS IS]
+After segmentation and cell tracking, this set-up is able to generate fluorescence and size time series for dozens of cells.
+Using again the self-replicator model as a framework, we show that these measurements are sufficient for the intended signal reconstructions, in particular the growth rate and the resource allocation profile over time.
+The method we use is called Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+While well-known in engineering, Kalman smoothing has been rarely used in quantitative biology, but turns out to be particularly suitable for the purpose of the model-based analysis of gene expression data in single cells, as we argue and show on synthetic data through the chapter.
+Finally, the experimental verification or falsification of the optimal control prediction, should allow to conclude if the biomass maximization, as considered in Chapter~\ref{chap:theory}, is indeed a cell objective in a dynamical context.

### 2017-01-03 Chapters/discussion.tex
Review of Hidde's corrections of discussion

@@
 Physics has yet to be reduced to a formula that will fit on a piece of clothing~\cite{falk_universe_2005}.
-Biology is even more behind, but the extensive work on the mechanisms of evolution are helping to close this gap (the Price equation being the closest we have come to such an elegant formula~\cite{frank_natural_2012}).
-["even more behind"...
-NOT SURE IF THIS IS THE RIGHT WAY TO PHRASE IT] [YOU NEED TO EXPLAIN WHAT IS THE PRICE EQUATION, OTHERWISE THE PHRASE IS EMPTY]
+This is even more true for Biology, but the extensive work on the mechanisms of evolution are helping to close this gap (see for instance the Price equation, which summarize in a short and elegant formula the mechanisms of evolution and natural selection~\cite{frank_natural_2012}).
 But evolution by itself is not sufficient to define life.
 Even the less restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
-Self-sustainment appears to be as fundamental as evolution, and relies on the transformation of matter and energy from the environment into LIVING SYSTEMS?, in other words, on growth.
-But the mathematical framing of questions regarding the self-sustaining character of living systems has not advanced as much as the mathematical formulation of evolutionary theory OR the mathematical analysis of evolution.
+Self-sustainment appears to be as fundamental as evolution, and relies on the transformation of matter and energy from the environment into organic matter, in other words, on growth.
+But the mathematical framing of questions regarding the self-sustaining character of living systems has not advanced as much as the mathematical analysis of evolution.
 
 Nevertheless, fundamental growth laws have been established for microorganisms.
-These growth laws show that, regardless of the molecular mechanisms ensuring groth control, microorganisms tend to follow the same empirical regularities when grown at steady state in different environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
+These growth laws show that, regardless of the molecular mechanisms ensuring growth control, microorganisms tend to follow the same empirical regularities when grown at steady state in different environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
 In particular, they adjust their internal molecular composition after a change in the environment, in such a way as to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
@@
 In Chapter~\ref{chap:theory}, we used a simple self-replicator model of resource allocation, and showed that the steady state is invariant over a number of regulatory schemes: the growth rate can be maximized by measuring either the environment or the internal state of the cell.
-In order to achieve this steady-state growth rate, the expression of genes encoding the metabolic and gene expression machineries has to be set to a specific value.
+In both cases, the expression of genes encoding the metabolic and gene expression machineries has to be set to a specific value.
 By contrast, biomass maximization during a growth transition requires information about the internal state rather than the environment.
@@
 The model developed here is an instance of proof-of-concept model~\cite{servedio_not_2014}.
-It does not necessarily aim at quantitatively predicting or controlling the behavior of microorganisms. [YOU WOULD LIKE TO HAVE AT LEAST QUALITATIVE CORRESPONDENCE...]
+It does not necessarily aim at quantitatively predicting or controlling the behavior of microorganisms.
 By using a simple and abstract representation, it enables a convenient and tractable way of evaluating the implications of dynamical optimality, in particular its divergence from steady-state optimality.
@@
 When including much molecular details, the models also tend to become untractable, prohibiting the use of mathematical tools like optimal control theory that are quickly overwhelmed by the sheer number of variables and parameters in the model.
-Besides, simple, abstract models make it easy for the modeler to state the critical, exploratory, and logistical assumptions that are made~\cite{servedio_not_2014}.
-[WHAT ARE "LOGISTICAL" ASSUMPTIONS?]
+Besides, simple, abstract models make it easy for the modeler to state the assumptions that are made, and more importantly to identify which ones are logistical, exploratory, or critical (see~\cite{servedio_not_2014}, in particular Box~1).
+Logistical assumptions do not affect the conclusions of the model, and are only necessary for tractability.
+A good example in our case is the Michaelis-Menten kinetics that were considered in Chapter~\ref{chap:theory}.
+Exploratory assumptions, however, might be important to vary and test.
 For instance, an exploratory assumption of our model was to consider that the proposed strategies had to be optimal at steady state as well.
@@
 
-But the assumptions that are most in need of discussion are the critical assumptions, \textit{i.e.} the ones that would invalidate the model if they were falsified.
+But the assumptions that are most in need of discussion are the critical assumptions, \textit{i.e.} the ones that are invalidated if the model predictions are not met in nature.
 In our case, a critical assumption is that biomass production is the main fitness factor in stable or changing environments.
@@
 While it is clear that microorganisms can be cultivated and are even naturally found in conditions where maximizing the growth rate ensures their persistence~\cite{edwards_silico_2001,ibarra_escherichia_2002,lewis_omic_2010,molenaar_shifts_2009}, we have no direct proof that biomass production is the fitness factor that allowed them to naturally persist on evolutionary time-scales.
-In addition, as illustrated by the quote at the beginning of this chapter, [THE QUOTE IS MORE OF A JOKE THAN A SERIOUS ARGUMENT...] one must be careful when explaining the behavior of living systems solely as a consequence of natural selection.
-Such systems are embedded in the physical world, and evolutionary arguments are not always relevant to explain an empirically observed behavior.
-[I DO NOT UNDERSTAND YOUR POINT, YOU SHOULD DEVELOP THIS PHRASE]
+In addition, one must be careful when explaining the behavior of living systems solely as a consequence of natural selection.
+Such systems are embedded in the physical world, and sometimes universal laws just result from the physical limitations that apply to the system (\textit{e.g.}, Monod's law described in Chapter~\ref{chap:introduction}).
 
@@
 Finally, the signal of interest was reconstructed through the use of Kalman smoothing, a powerful signal processing algorithm that has, as we showed, many features that make it suitable for the analysis of microscopy time-series data~\cite{kailath_linear_2000,jazwinski_stochastic_2007,kalman_new_1960}.
-Overall, the set-up still needs many improvements, though most of them are experimental issues that are just here to remind us that lab work is excessively harsh and merciless. [BOF]
+Overall, the set-up still needs many improvements, though most of them are experimental issues that should be easily fixable.
 
-Chapter~\ref{chap:experiments} must not be taken as a test of the model in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see box~2 in~\cite{servedio_not_2014}).
-[WHAT IS THE DIFFERENCE?!]
-We apply the experimental set-up to empirically test the prediction of a bang-bang resource allocation profile following a nutrient upshift.
-This prediction follows directly from the critical assumption that microorganisms maximize their biomass production at steady state and during growth transitions.
-While we were not able to arrive at an unambiguous conclusion about the correctness of this prediction in this manuscript, we can draw up two possible scenarios in the wake of future improvements of the experimental set-up.
+Chapter~\ref{chap:experiments} must not be taken as a "test of the model" in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see box~2 in~\cite{servedio_not_2014}).
+The model is correct in that bang-bang resource allocation follows mathematically from the assumption that microorganisms maximize their biomass production at steady state and during growth transitions.
+The model predictions can be used, however, as a tool to test whether this critical assumption is met in nature.
+In Chapter~\ref{chap:experiments}, we thus apply the experimental set-up to observe the resource allocation profile following a nutrient upshift in \textit{E.~coli}.
+While we were not able to arrive at an unambiguous conclusion about the correctness of the model prediction in this manuscript, we can draw up two possible scenarios in the wake of future improvements of the experimental set-up.
 
 In one scenario, the bang-bang or at least oscillatory nature of resource allocation during a growth transition is confirmed.
-This would be a strong argument in favor of an adaptation of microorganisms to changing environments, hence to the existence of more general dynamical growth laws.
-[THE OPPOSITE IS NOT AN ARGUMENT AGAINST ADAPTATION AND DYNAMICAL GROWTH LAWS, ONLY AGAINST THE ONE THE MODEL PREDICTS]
+This would suggest that the regulatory mechanisms of resource allocation do maximize biomass production during growth transitions, hence that microorganisms are adapted for changing environments.
 In order to further establish these results, we would need additional data for a broader range of environmental conditions.
@@
 
-In the other scenario, we do not observe any trace of a bang-bang resource allocation scheme, which would falsify the model.
-In that case, the questions would arise which critical assumptions are not valid.
-We also cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might in fact be critical for the predictions of the model. [LOGISTICAL?]
+In the other scenario, we do not observe any trace of a bang-bang resource allocation scheme.
+In that case, the questions that would arise are which critical assumptions are not valid.
+We also cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might in fact be critical for the predictions of the model.
 Although unlikely \textit{a priori}, this might be the case for the expressions used for the macroreactions in the model (Eqs~\ref{eq:metaflux}-\ref{eq:machflux}), the definition of the volume as invariably proportional to the total mass of macromolecules (Eq.~\ref{eq:voldef}), or the fact that degradation is assumed negligible with respect to the rates of other reactions in the system.
@@
 As we illustrated in Chapter~\ref{chap:theory}, complex regulatory schemes only prove beneficial in a dynamical context, away from steady state.
-At the opposite extreme, one can expect the benefits of these regulatory schemes to saturate in a constantly-changing, fluctuating environment, whereas the cost of complexity would probably not saturate, since acquiring more information is expected to require additional sensing systems, diverting away an increasingly larger part of cellular resources.
+At the opposite extreme, one can expect the benefits of these regulatory schemes to saturate, whereas the cost of complexity would probably not saturate, since acquiring more information is expected to require additional sensing systems, diverting away an increasingly larger part of cellular resources.
 Overall, these two tendencies are expected to cross at a point that would represent an evolutionary bottleneck, beyond which the costs exceed the benefits (see~\cite{short_flows_2006} for another example of such a bottleneck).

### 2017-01-03 Chapters/discussion.tex
Minor text fix in discussion

@@
 
-Chapter~\ref{chap:experiments} must not be taken as a "test of the model" in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see box~2 in~\cite{servedio_not_2014}).
+Chapter~\ref{chap:experiments} must not be taken as a "test of the model" in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see Box~2 in~\cite{servedio_not_2014}).
 The model is correct in that bang-bang resource allocation follows mathematically from the assumption that microorganisms maximize their biomass production at steady state and during growth transitions.
-The model predictions can be used, however, as a tool to test whether this critical assumption is met in nature.
+The model predictions can be used, however, as a tool to test whether this critical assumption is valid.
 In Chapter~\ref{chap:experiments}, we thus apply the experimental set-up to observe the resource allocation profile following a nutrient upshift in \textit{E.~coli}.

### 2017-01-03 Chapters/theory.tex
Change titlepage (new format) and write Suppinfo S5 about strains
Correct some figures and add the suppinfo of chapter 3 except the part about the results obtained in microplate reader
First draft of summary of chapter 2

@@
-\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies }
+\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
 \chaptermark{Dynamical allocation of cell resources}
@@
 
-%\section*{Summary of chapter \thechapter}
-%Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
-%Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
-%While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
-%The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
-%We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
-%We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
-%We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
-%A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
-%Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
-%It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
-%Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
 
 \selectlanguage{french}
-\section*{Résumé en français du chapître \thechapter : [titre français]}
-dans cette section blablabla
-
+\begin{chapter_summary}{Modélisation de l'allocation dynamique des ressources cellulaires comme un problème de contrôle optimal}
+
+Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
+En Section \ref{sec:model}, mous modélisons la cellule comme un auto-réplicateur (Fig.~\ref{fig:self_replicator}) devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules.
+Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
+$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle est produite à la proportion $1-\alpha$).
+Comme on l'a déjà vu avec les lois de croissance établie à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant plus que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
+Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaire, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?
+
+Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissances à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
+En Section~\ref{sec:growthlaws}, nous voyons en effet que pour chaque environnement, un $\alpha$ unique existe qui maximise la croissance de la cellule, mais surtout que ce $\alpha^*_{opt}$ est d'autant plus important que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
+De fait, les lois de croissances à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
+Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?
+
+Dans ce contexte, $\alpha$ est désormais une fonction du temps.
+Nous cherchons donc quelle fonction du temps maximise un objectif donné.
+Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif que l'on se fixe est une fonction d'une fonction, plus communément appelée fonctionnelle.
+Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant une changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
+Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
+La solution optimale obtenue montre que le contrôle $\alpha$ doit prendre alternativement les valeurs 0 ou 1 jusqu'à ce que le système soit stabilisé sur un nouvel équilibre (Fig.~\ref{fig:optimalcontrol} et Eq.~\ref{eq:optcontrol}).
+En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
+Ce type de solution se rencontre souvent en contrôle optimal, est répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
+Mais ce contrôle est-il pertinent d'un point de vue biologique ?
+La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des moments clés de la transition ?
+
+En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elle différentes stratégies de régulations (Fig.~\ref{fig:strategies_overview}).
+En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaires, nous montrons que deux stratégies de régulations simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
+De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc sont exactement équivalentes.
+Cependant, dans le contexte dynamique de notre banc d'essai, la stratégie qui consiste à mesurer les précurseurs est bien plus efficace, même si elle reste évidemment inférieure à la solution optimale (Fig.~\ref{fig:comparison_nutrients_precursor}).
+Mais il s'agit là de régulations simples qui exploitent l'information ne venant que d'une seule variable.
+
+Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
+Mais surtout, en réutilisant un modèle du système ppGpp qui contrôle la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
+Cela procure un nouveau point de vue sur ce système largement répandu chez de nombreux microorganismes: en intégrant l'information provenant à la fois de la quantité de ribosomes et celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule en environnement variable.
+\end{chapter_summary}
 \selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
+\section*{Beginning of Chapter \thechapter}
+
+\noindent \textbf{Important note:} Text and figures in this chapter have been published in Plos Computational Biology~\cite{giordano_dynamical_2016} under the terms of the \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution License} (CC-BY 4.0).
 
@@
 \centering
-\includegraphics[scale=0.9]{./Fig/Fig1.eps}
+\includegraphics[scale=0.7]{./Fig/Fig1}
 \caption{
@@
 \begin{figure}[tb]
-\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig2}
 \caption{\textbf{Analysis of self-replicator model of bacterial growth.}
@@
 \begin{figure}[p]
-\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig3}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig4}
 \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
+\includegraphics[width=\textwidth]{./Fig/Fig5}
 \caption{
@@
 \centering
-\includegraphics[scale=1]{./Fig/Fig6.eps}
+\includegraphics[scale=1]{./Fig/Fig6}
 \caption{
@@
 \centering
-\includegraphics[scale=1]{./Fig/Fig7.eps}
+\includegraphics[scale=1]{./Fig/Fig7}
 \caption{
@@
 \centering
-\includegraphics[scale=1]{./Fig/Fig8.eps}
+\includegraphics[scale=1]{./Fig/Fig8}
 \caption{
@@
 \clearpage
-
-%%%%%% FOR MANUAL REF%%%%%
-\makeatletter
-\newcommand{\manuallabel}[2]{\def\@currentlabel{#2}\label{#1}}
-\makeatother
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
+\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding}
 \caption
@@
 \centering
-\includegraphics[height=4cm]{Fig/adj2ab.eps} 
-\includegraphics[height=4cm]{Fig/adj2.eps} 
+\includegraphics[height=4cm]{Fig/adj2ab} 
+\includegraphics[height=4cm]{Fig/adj2} 
 \caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis.
@@
 
-\begin{figure}[tb]~Text
+\begin{figure}[p]
 \centering
-\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1.eps}
+\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1}
 \caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
@@
 
-\begin{table}[tb]
+\begin{table}[p]
 \scriptsize
@@
 \centering
-\includegraphics[width=\textwidth]{./Fig/plot_strategies.eps}
+\includegraphics[width=\textwidth]{./Fig/plot_strategies}
 \caption{

### 2017-01-03 Chapters/experiments.tex
First draft of chapter 3 summary

@@
 
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
 \selectlanguage{french}
-\section*{Résumé en français du Chapître \thechapter : Dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose.}
-
-Résumé d'environ 6-8 paragraphes (2 pages).
+\begin{chapter_summary}{Mesure dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose}
+
+Dans ce chapitre, nous nous intéressons à la validation expérimentale des prédictions faites au cours du Chapitre~\ref{chap:theory}.
+Nous avons en effet pu voir que, si l'on suppose qu'un microorganisme optimise sa production de biomasse lors d'une transition de croissance, on peut s'attendre à ce que ce dernier distribue ses ressources cellulaires entre la production de machinerie d'expression génique et de machinerie métabolique en suivant un schéma tout-ou-rien (\textit{bang-bang}).
+Le but est donc ici de mesurer précisément la dynamique de $\alpha$ \textit{in vivo} lors d'une transition de croissance contrôlée.
+
+La Section~\ref{sec:exp_design} s'attache à définir les conditions expérimentales dans lesquelles le comportement prédit lors du Chapitre~\ref{chap:theory} pourra être observé.
+En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que $\alpha (\cdot)$ peut être reconstruit au cours du temps si l'on dispose de mesure dynamiques lors de la transition du taux de croissance et de la concentration en machinerie d'expression génique.
+Cependant, le comportement tout-ou-rien présente une dynamique bien particulière qui peut facilement être offusquée au niveau d'une population de cellules si ces dernières ne sont pas synchrones.
+De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de la cellule unique, ce qui nous a conduit à utiliser des rapporteurs fluorescents de l'abondance de la machinerie.
+En particulier, nous construisons une souche de \textit{Escherichia coli} chez laquelle la sous-unité S2 du ribosome a été fusionnée avec une protéine fluorescente (GFP) (voir Section~\ref{sec:methods_strain} des matériels et méthodes).
+Couplé à de la vidéomicroscopie, cette solution permet de quantifier en temps réel la quantité de ribosomes au niveau de cellules uniques, ce qui procure un excellent moyen d'obtenir la concentration en machinerie d'expression génique (le ribosome étant son composant principal).
+Le taux de croissance quant à lui, peut-être facilement reconstruit en mesurant simplement la taille des cellules au cours du temps.
+
+Nous définissons également dans la Section~\ref{sec:exp_design} les conditions de croissance adéquates.
+Nous nous focalisons sur le cas particulier d'une transition d'un milieu de croissance pauvre vers un milieu riche.
+Cependant, aussi pauvre que soit le premier milieu, il est important que les bactéries soient en phase de croissance, pour la raison que leur physiologie change complètement lorsqu'elles s'arrêtent de croître, ce qui crée des effets expérimentaux indésirables qui sont durs à contrôler.
+Les cellules doivent donc croître pendant suffisamment longtemps sur le premier milieu, sans épuisé celui-ci, puis une transition rapide doit être effectuée vers le milieu riche.
+Nous avons utilisé pour cela la \textit{mother machine}, un dispositif microfluidique à la base inventé pour maintenir et observer les cellules bactériennes en état de croissance exponentielle pendant de longues périodes.
+Au sein de ce dispositif, les cellules sont constamment en contact avec du milieu frais, qui circule et apporte de nouveaux nutriments à leur portée durant toute la durée de l'expérience (voir Section~\ref{sec:microflu}).
+En changeant simplement le type de milieu injecté dans le dispositif, nous pouvons l'utiliser pour changer quasi-instantanément les nutriments auxquels les bactéries ont accès, et donc réaliser la transition de manière robuste et contrôlée.
+En particulier, le milieu pauvre utilisé contient de l'acétate comme seule source de carbone, tandis que le milieu riche contient essentiellement du glucose (Fig.~\ref{fig:experiment_schema}).
+
+De telles expériences s'avèrent malheureusement pointues et, dans le temps disponible pour le doctorat, nous n'avons pu en réaliser qu'un petit nombre, dont seule celle présentée ici s'est avérée exploitable.
+Celle-ci a cependant subit de nombreux problèmes expérimentaux qui, s'ils n'ont pas empêché complètement son exploitation, ont posé des problèmes pour l'analyse et l'interprétation des résultats.
+Ces problèmes expérimentaux sont abordés dans la Section~\ref{sec:data_acquisition}, et sont largement discutés en Section~\ref{sec:chap3_discussion}.
+En effet, suite à un problème logiciel et à ce qui semble être une contamination par un bacteriophage, nous n'avons pas pu obtenir des séries temporelles aussi longues qu'initialement prévu.
+Même si nous disposons néanmoins des données se situant autour de la transition, le fait de ne pas avoir atteint un nouvel état de croissance stationnaire sur glucose va s'avérer être un inconvénient majeur pour l'analyse des résultats.
+De plus, nous n'avons pu analyser qu'une partie des données, en nous focalisant sur une seule cellule par canal de la \textit{mother machine} (Fig.~\ref{fig:data_acquisition}).
+Au final, il faut retenir de ces sections que cette expérience et son analyse sont largement préliminaires, et seront consolidés dans le futur par de nouveaux travaux.
+
+La Section~\ref{sec:res_kalman} présente la façon dont nous pouvons reconstruire $\alpha (\cdot)$ lors de la transition à partir des données obtenues.
+Pour cela, nous développons une méthode basée sur le lissage de Kalman, qui repose sur une approche probabiliste et bayésienne pour reconstruire de manière robuste un signal à partir de mesures bruitées.
+Nous montrons comment l'information disponible sur le déroulement de l'expérience peut être injectées sous la forme d'un \textit{prior} pour faciliter la reconstruction, ce qui rend cette technique particulièrement efficace sur des données de microscopie.
+De plus, à partir de données synthétiques générée \textit{via} le modèle développé lors du Chapitre~\ref{chap:theory}, et constatons que la méthode est capable de capturer le profil oscillatoire généré par le comportement tout-ou-rien attendu (Fig.~\ref{fig:synthetic_upshift}).
+
+Finalement, nous appliquons cette méthode à la reconstruction du taux de croissance puis de $\alpha (\cdot)$, et présentons les résultats dans la Section~\ref{sec:estim_gr_ra}.
+En moyenne, le taux de croissance présente le comportement attendu (Figs~\ref{fig:growth_rate_estimation} et \ref{fig:growth_rate_estimation_median}), avec des valeurs à l'état stationnaire de croissance qui sont proches de celles connues pour \textit{E.~coli}, et une augmentation rapide au moment de la transition vers le milieu riche.
+Les résultats pour la reconstruction de $\alpha (\cdot)$ sont cependant plus mitigés.
+Même si un comportement oscillatoire est observé pour chaque cellule (Figs~\ref{fig:gene_activity} et \ref{fig:gene_activity_heatmap}), il n'est pas certain que ce dernier ne soit pas un artefact du à une mauvaise calibration de la méthode.
+Il aurait fallu en effet obtenir de longs état stationnaires de croissance avant et après la transition pour pouvoir réaliser cette calibration dans de bonnes conditions.
+Néanmoins, si l'on s'intéresse au comportement médian de la population (Fig.~\ref{fig:gene_activity_median}), on constate que la première oscillation survenant après la transition semble conservée (Fig.~\ref{fig:gene_activity_heatmap}).
+Ces résultats s'avèrent finalement encourageants, et incitent à pousser plus en avant cette étude, notamment via l'acquisition de nouvelles données pour lesquelles les problèmes cités auront pu être corrigé (voir discussion en Section~\ref{sec:chap3_discussion}).
+
+\end{chapter_summary}
 \selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
+\section*{Beginning of Chapter \thechapter}
 
@@
 The acquired data consisted of phase contrast and fluorescence images of 60 wells in total, in four different fields.
-While quite a few image analysis programs have been reported in the literature~\cite{klein_tlm-tracker:_2012,wang_image_2010,young_measuring_2012,paintdakhi_oufti:_2016}, and some of these specifically address the mother machine design of the microfluidic device~\cite{wang_robust_2010,sachs_image-based_2016}, they all presented limitations when applied to our data.
+While quite a few image analysis programs have been reported in the literature~\cite{klein_tlm-tracker_2012,wang_image_2010,young_measuring_2011,paintdakhi_oufti_2016}, and some of these specifically address the mother machine design of the microfluidic device~\cite{wang_robust_2010,sachs_image-based_2016}, they all presented limitations when applied to our data.
 For instance, the segmentation algorithms experienced difficulties due to the low resolution of the camera, and the phase contrast images had a superposed reflection band due to the microfluidic device.

### 2017-01-04 Chapters/introduction.tex
Review of Hidde's corrections of discussion
First draft of summary of Chapter 1

@@
 \chapter{Introduction}
+\label{chap:introduction}
 
@@
 
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
 \selectlanguage{french}
-\section*{Résumé en français}
+\begin{chapter_summary}{Introduction}
+Malgré leur apparente simplicité, les microorganismes sont des êtres vivants complexes capable de survivre dans les milieux les plus extrêmes.
+Une telle ubiquité repose en partie sur leur excellente capacité d'adaptation.
+En effet, \textit{via} l'information stockée dans son ADN, une cellule microbienne est capable, par la production des protéines nécessaires, d'assimiler et de se répliquer à partir d'une grande variété de nutriments.
+Ce mécanisme, que l'on appelle l'expression génétique, permet donc à la cellule de réguler sa composition en fonction des conditions environnementales.
+Chaque composant cellulaire est initialement issu des ressources trouvées dans l'environnement.
+\textit{Via} l'expression génétique, la machinerie cellulaire résout donc un problème de distribution des ressources: un bien commun (les nutriments) est partagé entre plusieurs milliers de bénéficiaires potentiels (les protéines et autre macromolécules).
 
-Dans cette section... (env. 1-2 pages)
+Selon quels critères les ressources sont-elles distribuées dans la cellule ?
+Le comportement de la plupart des systèmes vivants peut s'expliquer par les principes de la sélection naturelle.
+En définissant des facteurs de \textit{fitness} (adaptation au milieu), on peut généralement identifier des objectifs permettant à l'organisme de maximiser sa descendance, et donc de persister sur le long terme dans un milieu donné.
+Cependant, ces facteurs et surtout leurs importances relatives, peuvent s'avérer extrêmement complexes à identifier.
+C'est notamment le cas pour les microorganismes, où de nombreux facteurs peuvent entrer en compte.
+Cependant, de nombreuses études ont pu montrer que la maximisation du taux de croissance (ou taux de réplication) est un facteur qui semble commun à de nombreux microorganismes, et qui permet d'expliquer à lui seul une grande variété de comportements.
+
+Peut-on, à l'aide d'un tel principe universel, formaliser la croissance des microorganismes en lois simples, analogues à celle que l'on peut trouver dans d'autres domaines des sciences comme la physique ?
+Historiquement, la première tentative de formalisation de la physiologie microbienne en lois fondamentales fut la découverte de la loi de Monod il y a plus de 60 ans (Fig.~\ref{fig:monod_law} et Eq.~\ref{eq:monod_law}).
+Elle montre notamment que le taux de croissance d'un microorganisme est une loi hyperbolique de la concentration du nutriment limitant dans le milieu.
+Depuis, d'autres lois fondamentales ont pu être identifiées.
+L'une d'entre elle est une loi de croissance qui décrit la relation linéaire existant entre la concentration en ribosomes dans la cellule, et la richesse du milieu qui se traduit par le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
+De manière étonnante, quel que soient les détails moléculaires qui permettent à l'organisme cette adaptation, cette loi émerge s'il cherche à maximiser son taux de croissance dans chacun des milieux.
+
+Bien que quasi-universelle, ces lois se limitent cependant à un état bien particulier de croissance : la croissance à l'état stationnaire.
+Cet état est atteint lorsque l'organisme se multiplie pendant suffisamment longtemps dans un environnement stable (Fig.~\ref{fig:growth_curve}).
+Pour de nombreuses raisons, il s'avère très pratique pour les études théoriques et expérimentales car les composants de la cellule ne changent plus au cours de temps, ce qui lui vaut aussi l'appellation de croissance équilibrée (\textit{balanced growth}).
+Mais cet état a beau être pratique, il est très loin des conditions rencontrées par les microorganismes dans leur milieu de vie naturel.
+Ces derniers sont en effet plutôt soumis à des environnements très changeants, où les éléments chimiques indispensables à la croissance sont rares, et sont donc les objets d'une rude compétition.
+Pour quelles raisons alors les microorganismes seraient-ils optimisés pour une état de croissance qu'ils n'ont quasiment jamais rencontré au cours de leur évolution ?
+Si comme on s'y attend ces systèmes sont plutôt adaptés à des environnements variables, ratons-nous quelque chose en les étudiant dans des conditions stables ?
+
+Dans cette thèse, nous nous proposons ainsi d'étendre les études des lois de croissance en adoptant une perspective dynamique, c'est-à-dire, en environnements variables : quelles stratégies dynamiques les microorganismes suivent-ils pour redistribuer leurs ressources lors d'un changement environnemental ?
+Cette question sera développée ici en deux grands axes d'étude.
+
+Quelles sont les meilleures stratégies de distribution des ressources si, comme pour les lois de croissance à l'état stationnaire, la production de biomasse est le seul critère optimisé par la cellule ?
+Répondre à cette question est une étape nécessaire pour évaluer dans quelle mesure les critères d'optimisation dynamique peuvent différer de ceux à l'état stationnaire, et donc, de tester si les pressions évolutives s'appliquent effectivement différemment entre environnements stables et dynamiques.
+Cette preuve de concept reposera sur un modèle de la distribution des ressources dans une cellule microbienne, qui sera utilisé pour déterminer quelle est la manière dynamique optimale pour la cellule de distribuer ses ressources lors d'un changement d'environnement (Chapitre~\ref{chap:theory}).
+
+Pour tester cette prédiction expérimentalement, nous chercherons ensuite à mesurer cette distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
+Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse de série temporelle, permettent d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu.
+\end{chapter_summary}
 \selectlanguage{english}
@@
 The gene expression machinery (itself constituted of RNA and proteins) synthesizes gene products after binding to target sequences either at the state of transcription (promoter on the DNA), or translation (ribosome binding site on the messaging RNA).
-In a given cell, there are roughly as many different target sequences as genes~\cite{keseler_ecocyc:_2013}, which ensures that not all products are synthesized at the same rate.
+In a given cell, there are roughly as many different target sequences as genes~\cite{keseler_ecocyc_2013}, which ensures that not all products are synthesized at the same rate.
 Furthermore, some of these products specifically bind some promoters and stimulate or repress promoter activity and hence gene expression.

### 2017-01-04 Chapters/discussion.tex
Summary of chapter 4, first draft

@@
 
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
 \selectlanguage{french}
 \section*{Résumé en français}
+Contrairement à la physique, la biologie manque encore de théories quantitatives à fort pouvoir prédictif.
+Même si de grosses avancées ont eu lieu en biologie évolutive, d'autres domaines sont loin d'être aussi avancés d'un point de mathématique.
+C'est notamment le cas de la croissance, qui malgré son rôle fondamental dans la définition du vivant, repose sur des mécanismes qui semblent varier énormément d'un organisme à l'autre.
 
-Dans cette section... (env. 1 page)
+Chez les microorganismes, nous disposons néanmoins de lois de croissance.
+Elles montrent de manière empirique que malgré une grande diversité des mécanismes moléculaires qui assurent le contrôle de la croissance, la composition des microorganismes obéit à des règles universelles lorsqu'ils se multiplient à l'état stationnaire dans différents environnements.
+En particulier, l'abondance de leurs ribosomes s'ajuste linéairement avec la richesse du milieu, d'une manière que maximise leur taux de croissance.
+Mais comme décrit au cours du Chapitre~\ref{chap:introduction}, ces lois ont été essentiellement établies en croissance stationnaire, un état très rarement rencontré par ces organismes dans leur milieu naturel.
+Pour quelles raisons les microorganismes seraient-ils optimisés pour un état qu'ils n'ont que très peu rencontré au cours de leur évolution ?
+Les lois de croissance à l'état stationnaire ne seraient-elles pas une application particulière de lois plus générales qui s'appliquent en environnements variables ?
+
+Notre but dans ce manuscrit a été d'établir un cadre de travail à la fois théorique et expérimental dans lequel une perspective dynamique peut être adoptée sur les lois de croissance.
+Au cours du Chapitre~\ref{chap:theory}, nous avons utilisé un modèle d'auto-réplicateur pour évaluer la redistribution optimale des ressources lors d'un changement environnemental.
+En élargissant les principes d'optimalité établis dans les lois de croissance stationnaire, nous avons montré qu'une distribution de type tout-ou-rien (bang-bang en anglais) optimise la biomasse générée au cours de la transition.
+Cela nous a servi de point de référence pour comparer entre elles différentes stratégies de régulation.
+Ainsi, nous avons montré que les stratégies mesurant l'état interne de la cellule s'avèrent plus efficace que celles tirant leur information de l'environnement, contrairement à l'état stationnaire de croissance où toutes ces stratégies sont strictement équivalentes.
+De plus, une stratégie étonnamment proche du système ppGpp chez \textit{Escherichia coli}, est théoriquement capable de s'approcher de la distribution tout-ou-rien identifiée comme optimale.
+
+La simplicité du modèle, même si elle peut rendre certains lecteurs sceptiques, s'est avérée cruciale dans notre étude.
+En effet, cela nous a permis de clairement spécifier les hypothèses mise en jeu.
+La plus critique s'avère être celle qui considère la production de biomasse comme étant le facteur de \textit{fitness} principal que l'on soit en environnement stable ou dynamique.
+Bien que largement répandue, cette hypothèse n'est finalement basée que sur des arguments indirects, ou bien établis dans des conditions de laboratoire, sur des souches de laboratoire (décrit dans le Chapitre~\ref{chap:introduction}).
+De fait, en conditions naturelles et à l'échelle des temps évolutifs, il n'est pas certain que ce critère soit vraiment si universel que ça.
+Une étape importante a donc été de vérifier expérimentalement les prédictions que nous avons tirées de cette hypothèse.
+
+C'est ce qui a été abordé au cours du Chapitre~\ref{chap:experiments} de ce manuscrit.
+Nous avons mis en place une cadre d'étude expérimental dans lequel la distribution des ressources peut être observée lors d'une transition de croissance.
+Tout d'abord, une souche d'\textit{Escherichia coli} a été modifiée en attachant une protéine fluorescente (GFP) à l'une des sous-unités du ribosome.
+Cette souche a ensuite été observée en utilisant une technique avancée de microfluidique permettant de réaliser des transitions robustes et contrôlées du milieu de culture.
+Enfin, les mesures obtenues ont pu être exploitées \textit{via} le développement d'une technique de reconstruction du signal utilisant le lissage de Kalman.
+Plusieurs améliorations du dispositif seront cependant nécessaires pour conclure définitivement sur l'existence d'une production tout-ou-rien des ribosomes après une transition.
+Deux scénarios sont cependant envisageables.
+
+Dans le premier cas, le caractère tout-ou-rien ou du moins oscillatoire de la synthèse des ribosomes pourrait être confirmé.
+Cela suggèreraient que les mécanismes de régulation de la distribution des ressources maximisent la biomasse produite lors d'une transition, et donc que les microorganismes sont effectivement adaptés à des environnements changeants.
+Pour confirmer ces résultats, de nouvelles données devront nécessairement être acquises, notamment en mesurant de telle transition dans une variété plus large de conditions environnementales (changement de source de carbone, ajout d'acides aminés dans le milieu pour contourner le métabolisme, utilisation d'une restriction en azote au lieu du carbone, ...).
+Le dispositif expérimental pourrait également être modifié pour contrôler un plus grand nombre de paramètres.
+Par exemple, en attachant un autre rapporteur fluorescent à une enzyme clé du métabolisme, nous pourrions vérifier si les synthèses de la machinerie d'expression génique et des enzymes du métabolisme sont effectivement en anti-phase lors des oscillations.
+
+Dans le second cas, le comportement observé serait différent d'une production tout-ou-rien.
+Par construction, cela signifierait qu'une hypothèse du modèle n'est pas respectée.
+Par exemple, la cellule ne maximiserait pas la biomasse produite lors d'une transition.
+Une autre possibilité serait que les coûts d'une telle régulation dépasse pour la cellule les éventuels bénéfices qu'elle peut en tirer.
+En effet, en ne comparant les schémas de régulation que vis-à-vis de leur bénéfices, nous avons fait l'hypothèse implicite que leurs coûts pour la cellule sont négligeables, sinon au moins comparables.
+Il est possible que même si la production de biomasse s'avère être un critère déterminant, les limites physiques imposées par le coût des régulations empêchent la cellule de faire mieux que le comportement observé.
+La prise en compte de ces coûts dans le modèle devra donc être explorée.
+
+Dans tous les cas, cette exploration des coûts des schémas de régulation pourrait s'avérer cruciale dans l'établissement d'un lien fondamental entre la complexité des schémas de régulation d'une espèce microbienne, et la dynamique de son environnement.
+En effet, comme montré dans le Chapitre~\ref{chap:theory}, la cellule ne tire avantage d'un schéma de régulation complexe que lorsque son environnement varie.
+On peut élargir ce résultat en supposant que le bénéfice apporté par des régulations toujours plus complexes devrait saturer.
+En revanche, leurs coûts devraient continuer à croitre, dans la mesure où des régulations plus élaborées impliquent pour la cellule la synthèse de davantage de système de mesure, et donc détournent une part toujours plus grande des ressources cellulaires.
+On peut donc conjecturer que ces deux tendances vont finir par se croiser au niveau d'un goulot d'étranglement évolutif, au delà duquel les coûts excèdent les bénéfices.
+On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin vers la complexité en fonction de l'intensité et la fréquence des variations environnementales auxquels le microorganisme est soumis.
+En d'autres termes, là où les lois de croissances actuelles décrivent uniquement des variations physiologiques avec la qualité du milieu, nous pourrions ajouter une autre dimension à ces lois qui représenterait la variabilité temporelle de la disponibilité des nutriments dans l'environnement naturel.
 \selectlanguage{english}

### 2017-01-04 Chapters/introduction.tex
Text fix in summary of chapter 1

@@
 Malgré leur apparente simplicité, les microorganismes sont des êtres vivants complexes capable de survivre dans les milieux les plus extrêmes.
-Une telle ubiquité repose en partie sur leur excellente capacité d'adaptation.
+Une telle ubiquité repose en partie sur leurs excellentes capacités d'adaptation.
 En effet, \textit{via} l'information stockée dans son ADN, une cellule microbienne est capable, par la production des protéines nécessaires, d'assimiler et de se répliquer à partir d'une grande variété de nutriments.
@@
 Chaque composant cellulaire est initialement issu des ressources trouvées dans l'environnement.
-\textit{Via} l'expression génétique, la machinerie cellulaire résout donc un problème de distribution des ressources: un bien commun (les nutriments) est partagé entre plusieurs milliers de bénéficiaires potentiels (les protéines et autre macromolécules).
+Par l'expression génétique, la machinerie cellulaire résout donc un problème de distribution des ressources: un bien commun (les nutriments) est partagé entre plusieurs milliers de bénéficiaires potentiels (les protéines et autres macromolécules).
 
@@
 En définissant des facteurs de \textit{fitness} (adaptation au milieu), on peut généralement identifier des objectifs permettant à l'organisme de maximiser sa descendance, et donc de persister sur le long terme dans un milieu donné.
-Cependant, ces facteurs et surtout leurs importances relatives, peuvent s'avérer extrêmement complexes à identifier.
-C'est notamment le cas pour les microorganismes, où de nombreux facteurs peuvent entrer en compte.
+Ces facteurs et surtout leurs importances relatives, peuvent s'avérer extrêmement complexes à identifier, 
+ce qui est notamment le cas chez les microorganismes.
 Cependant, de nombreuses études ont pu montrer que la maximisation du taux de croissance (ou taux de réplication) est un facteur qui semble commun à de nombreux microorganismes, et qui permet d'expliquer à lui seul une grande variété de comportements.
@@
 Peut-on, à l'aide d'un tel principe universel, formaliser la croissance des microorganismes en lois simples, analogues à celle que l'on peut trouver dans d'autres domaines des sciences comme la physique ?
-Historiquement, la première tentative de formalisation de la physiologie microbienne en lois fondamentales fut la découverte de la loi de Monod il y a plus de 60 ans (Fig.~\ref{fig:monod_law} et Eq.~\ref{eq:monod_law}).
-Elle montre notamment que le taux de croissance d'un microorganisme est une loi hyperbolique de la concentration du nutriment limitant dans le milieu.
+Historiquement, la première tentative d'établissement de lois fondamentales de la physiologie microbienne  fut la découverte de la loi de Monod il y a plus de 60 ans (Fig.~\ref{fig:monod_law} et Eq.~\ref{eq:monod_law}).
+Elle montre que le taux de croissance d'un microorganisme suit une loi hyperbolique de la concentration du nutriment limitant dans le milieu.
 Depuis, d'autres lois fondamentales ont pu être identifiées.
-L'une d'entre elle est une loi de croissance qui décrit la relation linéaire existant entre la concentration en ribosomes dans la cellule, et la richesse du milieu qui se traduit par le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
-De manière étonnante, quel que soient les détails moléculaires qui permettent à l'organisme cette adaptation, cette loi émerge s'il cherche à maximiser son taux de croissance dans chacun des milieux.
+L'une d'entre elles est une loi de croissance qui décrit la relation linéaire existant entre la concentration en ribosomes dans la cellule et le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
+De manière étonnante, quel que soient les détails moléculaires qui permettent cette adaptation, cette loi émerge lorsque la cellule cherche à maximiser son taux de croissance dans chaque environnement.
 
-Bien que quasi-universelle, ces lois se limitent cependant à un état bien particulier de croissance : la croissance à l'état stationnaire.
+Bien que largement répandues, ces lois se limitent cependant à un état bien particulier qu'on appelle croissance à l'état stationnaire.
 Cet état est atteint lorsque l'organisme se multiplie pendant suffisamment longtemps dans un environnement stable (Fig.~\ref{fig:growth_curve}).
-Pour de nombreuses raisons, il s'avère très pratique pour les études théoriques et expérimentales car les composants de la cellule ne changent plus au cours de temps, ce qui lui vaut aussi l'appellation de croissance équilibrée (\textit{balanced growth}).
+Il s'avère très pratique pour les études théoriques et expérimentales, notamment car les composants de la cellule ne changent plus au cours de temps, ce qui lui vaut aussi l'appellation de croissance équilibrée (\textit{balanced growth}).
 Mais cet état a beau être pratique, il est très loin des conditions rencontrées par les microorganismes dans leur milieu de vie naturel.
 Ces derniers sont en effet plutôt soumis à des environnements très changeants, où les éléments chimiques indispensables à la croissance sont rares, et sont donc les objets d'une rude compétition.
-Pour quelles raisons alors les microorganismes seraient-ils optimisés pour une état de croissance qu'ils n'ont quasiment jamais rencontré au cours de leur évolution ?
-Si comme on s'y attend ces systèmes sont plutôt adaptés à des environnements variables, ratons-nous quelque chose en les étudiant dans des conditions stables ?
+Pour quelles raisons les microorganismes seraient-ils optimisés pour un état de croissance qu'ils n'ont quasiment jamais rencontré au cours de leur évolution ?
+Si comme on s'y attend ces systèmes sont plutôt adaptés à des environnements variables, ratons-nous quelque chose en ne les étudiant qu'en conditions stables ?
 
 Dans cette thèse, nous nous proposons ainsi d'étendre les études des lois de croissance en adoptant une perspective dynamique, c'est-à-dire, en environnements variables : quelles stratégies dynamiques les microorganismes suivent-ils pour redistribuer leurs ressources lors d'un changement environnemental ?
-Cette question sera développée ici en deux grands axes d'étude.
+Cette question sera développée en deux grands axes d'étude.
 
 Quelles sont les meilleures stratégies de distribution des ressources si, comme pour les lois de croissance à l'état stationnaire, la production de biomasse est le seul critère optimisé par la cellule ?
-Répondre à cette question est une étape nécessaire pour évaluer dans quelle mesure les critères d'optimisation dynamique peuvent différer de ceux à l'état stationnaire, et donc, de tester si les pressions évolutives s'appliquent effectivement différemment entre environnements stables et dynamiques.
+Répondre à cette question est une étape nécessaire pour évaluer dans quelles mesures les critères d'optimisation dynamique peuvent différer de ceux à l'état stationnaire, et donc, de tester si les pressions évolutives s'appliquent effectivement différemment entre environnements stables et dynamiques.
 Cette preuve de concept reposera sur un modèle de la distribution des ressources dans une cellule microbienne, qui sera utilisé pour déterminer quelle est la manière dynamique optimale pour la cellule de distribuer ses ressources lors d'un changement d'environnement (Chapitre~\ref{chap:theory}).
 
-Pour tester cette prédiction expérimentalement, nous chercherons ensuite à mesurer cette distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
-Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse de série temporelle, permettent d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu.
+Pour tester les prédictions faites expérimentalement, nous chercherons à mesurer cette distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
+Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse des séries temporelles, permettent d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu, notamment la transition d'un milieu pauvre vers un milieu riche (\textit{nutrient upshift}).
 \end{chapter_summary}
@@
 
+\section*{Beginning of Chapter \thechapter}
 \section{Context}

### 2017-01-04 Chapters/theory.tex
Text fix in summary of chapter 2

@@
 Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
-En Section \ref{sec:model}, mous modélisons la cellule comme un auto-réplicateur (Fig.~\ref{fig:self_replicator}) devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules.
+En Section \ref{sec:model}, nous modélisons la cellule comme un auto-réplicateur devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules (Fig.~\ref{fig:self_replicator}).
 Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
-$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle est produite à la proportion $1-\alpha$).
-Comme on l'a déjà vu avec les lois de croissance établie à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant plus que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
+$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle, est produite à la proportion $1-\alpha$).
+Comme on l'a déjà vu avec les lois de croissance établie à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
 Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaire, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?
@@
 Nous cherchons donc quelle fonction du temps maximise un objectif donné.
-Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif que l'on se fixe est une fonction d'une fonction, plus communément appelée fonctionnelle.
-Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant une changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
+Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif à optimiser est une fonction d'une autre fonction, plus communément appelée fonctionnelle.
+Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
 Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
@@
 En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
-Ce type de solution se rencontre souvent en contrôle optimal, est répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
+Ce type de solution se rencontre souvent en contrôle optimal, et répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
 Mais ce contrôle est-il pertinent d'un point de vue biologique ?
-La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des moments clés de la transition ?
-
-En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elle différentes stratégies de régulations (Fig.~\ref{fig:strategies_overview}).
-En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaires, nous montrons que deux stratégies de régulations simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
-De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc sont exactement équivalentes.
+La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des instants bien précis ?
+
+En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elles différentes stratégies de régulations (Fig.~\ref{fig:strategies_overview}).
+En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaire, nous montrons que deux stratégies de régulations simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
+De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc strictement équivalentes.
 Cependant, dans le contexte dynamique de notre banc d'essai, la stratégie qui consiste à mesurer les précurseurs est bien plus efficace, même si elle reste évidemment inférieure à la solution optimale (Fig.~\ref{fig:comparison_nutrients_precursor}).
-Mais il s'agit là de régulations simples qui exploitent l'information ne venant que d'une seule variable.
-
-Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
-Mais surtout, en réutilisant un modèle du système ppGpp qui contrôle la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
-Cela procure un nouveau point de vue sur ce système largement répandu chez de nombreux microorganismes: en intégrant l'information provenant à la fois de la quantité de ribosomes et celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule en environnement variable.
+
+Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie plus complexe mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
+Mais surtout, en réutilisant un modèle du système ppGpp, connu pour réguler la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
+Cela procure un nouveau point de vue sur ce système, d'ailleurs largement répandu chez de nombreux microorganismes.
+En intégrant l'information provenant à la fois de la quantité de ribosomes et de celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule lorsque celle-ci doit s'adapter rapidement à un nouvel environnement.
 \end{chapter_summary}

### 2017-01-04 Chapters/experiments.tex
Text fix in summary of chapter 3

--- 2017-01-02 Chapters/experiments.tex
@@
 
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
 \selectlanguage{french}
-\section*{Résumé en français du Chapître \thechapter : Dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose.}
-
-Résumé d'environ 6-8 paragraphes (2 pages).
+\begin{chapter_summary}{Mesure dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose}
+
+Dans ce chapitre, nous nous intéressons à la validation expérimentale des prédictions faites au cours du Chapitre~\ref{chap:theory}.
+Nous avons en effet pu voir que, si l'on suppose qu'un microorganisme optimise sa production de biomasse lors d'une transition de croissance, on peut s'attendre à ce que ce dernier distribue ses ressources cellulaires entre la production de machinerie d'expression génique et de machinerie métabolique en suivant un schéma tout-ou-rien (\textit{bang-bang}).
+Le but est donc ici de mesurer précisément la dynamique de $\alpha$ lors d'une transition de croissance contrôlée.
+
+La Section~\ref{sec:exp_design} s'attache à définir les conditions expérimentales dans lesquelles le comportement prédit lors du Chapitre~\ref{chap:theory} pourra être observé.
+En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que $\alpha (\cdot)$ lors de la transition peut être reconstruit au cours du temps si l'on dispose de mesures dynamiques du taux de croissance et de la concentration en machinerie d'expression génique.
+Cependant, le comportement tout-ou-rien a une dynamique bien particulière qui peut facilement être offusquée au niveau d'une population de cellules si ces dernières ne sont pas synchrones.
+De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de la cellule unique, ce qui nous a conduit à utiliser des rapporteurs fluorescents de l'abondance d'une des machineries.
+En particulier, nous construisons une souche d'\textit{Escherichia coli} chez laquelle la sous-unité S2 du ribosome a été fusionnée avec une protéine fluorescente (GFP) (voir Section~\ref{sec:methods_strain} des matériels et méthodes).
+Couplée à de la vidéomicroscopie, cette souche nous permet de quantifier en temps réel la quantité de ribosomes au niveau de cellules uniques, ce qui procure un excellent moyen d'obtenir la concentration en machinerie d'expression génique (le ribosome étant son composant principal).
+Le taux de croissance quant à lui, peut-être facilement reconstruit en mesurant la taille des cellules au cours du temps.
+
+Nous définissons également dans la Section~\ref{sec:exp_design} les conditions de croissance adéquates.
+Nous nous focalisons sur le cas particulier d'une transition d'un milieu de croissance pauvre vers un milieu riche.
+Cependant, aussi pauvre que soit le premier milieu, il est important que les bactéries soient en phase de croissance, pour la raison que leur physiologie change complètement lorsqu'elles s'arrêtent de croître, ce qui crée des effets indésirables durs à contrôler expérimentalement.
+Les cellules doivent donc croître pendant suffisamment longtemps sur le premier milieu, sans épuisé celui-ci, puis une transition rapide doit être effectuée vers le milieu riche.
+Nous avons utilisé pour cela la \textit{Mother Machine}, un dispositif microfluidique à la base inventé pour maintenir et observer les cellules bactériennes en état de croissance stationnaire pendant de longues périodes.
+Au sein de ce dispositif, les cellules sont constamment en contact avec du milieu frais, qui circule et leur procure de nouveaux nutriments durant toute la durée de l'expérience (voir Section~\ref{sec:microflu}).
+En changeant simplement le type de milieu injecté dans le dispositif, nous pouvons l'utiliser pour modifier en quelques secondes les nutriments auxquels les bactéries ont accès, et donc réaliser la transition de manière robuste et contrôlée.
+Plus particulièrement, le milieu pauvre utilisé contient de l'acétate comme seule source de carbone, tandis que le milieu riche contient essentiellement du glucose (Fig.~\ref{fig:experiment_schema}).
+
+De telles expériences s'avèrent malheureusement pointues et, dans le temps disponible pour le doctorat, nous n'avons pu en réaliser qu'un petit nombre, dont seule celle présentée ici s'est avérée exploitable.
+Celle-ci a cependant subit de nombreux problèmes expérimentaux qui, s'ils n'ont pas empêché complètement son exploitation, ont posé des problèmes pour l'analyse et l'interprétation des résultats.
+Ces problèmes expérimentaux sont abordés dans la Section~\ref{sec:data_acquisition}, et sont largement discutés en Section~\ref{sec:chap3_discussion}.
+En effet, suite à un problème logiciel et à ce qui semble être une contamination par un bacteriophage, nous n'avons pas pu obtenir des séries temporelles aussi longues qu'initialement prévu.
+Même si nous disposons néanmoins des données se situant autour de la transition, le fait de ne pas avoir atteint un nouvel état de croissance stationnaire sur glucose va s'avérer être un inconvénient majeur pour l'analyse des résultats.
+De plus, nous n'avons pu analyser qu'une partie des données, en nous focalisant sur une seule cellule par canal de la \textit{Mother Machine} (Fig.~\ref{fig:data_acquisition}).
+Au final, il faut retenir de ces sections que cette expérience et son analyse sont largement préliminaires, et seront consolidées dans le futur par de nouveaux travaux.
+
+La Section~\ref{sec:res_kalman} présente la façon dont nous pouvons reconstruire $\alpha (\cdot)$ lors de la transition à partir des données obtenues.
+Pour cela, nous développons une méthode basée sur le lissage de Kalman, qui utilise une approche probabiliste et bayésienne pour reconstruire de manière robuste un signal à partir de mesures bruitées.
+Nous montrons comment l'information dont nous disposons sur le déroulement de l'expérience peut être injectée dans l'analyse sous la forme d'un \textit{prior} bayésien afin de faciliter la reconstruction, ce qui s'avère particulièrement utile sur nos données.
+De plus, à partir de données synthétiques générée grâce au modèle développé lors du Chapitre~\ref{chap:theory}, nous constatons que la méthode est capable de capturer le profil oscillatoire généré par le comportement tout-ou-rien que nous attendons (Fig.~\ref{fig:synthetic_upshift}).
+
+Finalement, nous appliquons cette méthode à la reconstruction du taux de croissance puis de $\alpha (\cdot)$, et présentons les résultats dans la Section~\ref{sec:estim_gr_ra}.
+En moyenne, le taux de croissance présente le comportement attendu (Figs~\ref{fig:growth_rate_estimation} et \ref{fig:growth_rate_estimation_median}), avec des valeurs à l'état de croissance stationnaire qui sont proches de celles connues pour \textit{E.~coli}, et une augmentation rapide au moment de la transition vers le milieu riche.
+Les résultats pour la reconstruction de $\alpha (\cdot)$ sont cependant plus mitigés.
+Même si un comportement oscillatoire est observé pour chaque cellule (Figs~\ref{fig:gene_activity} et \ref{fig:gene_activity_heatmap}), il n'est pas certain que ce dernier ne soit pas un artefact dû à une mauvaise calibration de la méthode.
+Il aurait fallu en effet obtenir de longs états stationnaires de croissance avant et après la transition pour pouvoir réaliser cette calibration dans de bonnes conditions.
+Néanmoins, si l'on s'intéresse au comportement médian des cellules (Fig.~\ref{fig:gene_activity_median}), on constate que la première oscillation survenant après la transition semble conservée dans la population (Fig.~\ref{fig:gene_activity_heatmap}).
+Ces résultats s'avèrent finalement encourageants, et incitent à pousser plus en avant cette étude, notamment via l'acquisition de nouvelles données pour lesquelles les problèmes cités auront pu être corrigés (voir discussion en Section~\ref{sec:chap3_discussion}).
+
+\end{chapter_summary}
 \selectlanguage{english}
+
+\begin{center}
+\noindent\rule{4cm}{0.1pt}
+\end{center}
+
+\section*{Beginning of Chapter \thechapter}
 

### 2017-01-04 Chapters/discussion.tex
Text fix in summary of chapter 4

@@
 \selectlanguage{french}
-\section*{Résumé en français}
-Contrairement à la physique, la biologie manque encore de théories quantitatives à fort pouvoir prédictif.
-Même si de grosses avancées ont eu lieu en biologie évolutive, d'autres domaines sont loin d'être aussi avancés d'un point de mathématique.
-C'est notamment le cas de la croissance, qui malgré son rôle fondamental dans la définition du vivant, repose sur des mécanismes qui semblent varier énormément d'un organisme à l'autre.
+\begin{chapter_summary}{Introduction}
+Contrairement à sa cousine la physique, la biologie manque encore de théories quantitatives à fort pouvoir prédictif.
+Même si de grosses avancées ont eu lieu en biologie évolutive, d'autres domaines sont loin d'être aussi développés d'un point de mathématique.
+C'est notamment le cas de la croissance, qui malgré la place fondamentale qu'elle occupe dans ce qui définit un être-vivant, repose sur des mécanismes qui semblent varier énormément d'un organisme à l'autre, et pour lesquels des lois fondamentales sont dures à identifier.
 
 Chez les microorganismes, nous disposons néanmoins de lois de croissance.
-Elles montrent de manière empirique que malgré une grande diversité des mécanismes moléculaires qui assurent le contrôle de la croissance, la composition des microorganismes obéit à des règles universelles lorsqu'ils se multiplient à l'état stationnaire dans différents environnements.
-En particulier, l'abondance de leurs ribosomes s'ajuste linéairement avec la richesse du milieu, d'une manière que maximise leur taux de croissance.
+Elles montrent de manière empirique que malgré la grande diversité des mécanismes moléculaires qui assurent le contrôle de la croissance, la composition des microorganismes obéit à des règles universelles lorsqu'ils se multiplient à l'état stationnaire dans différents environnements.
+En particulier, l'abondance de leurs ribosomes s'ajuste linéairement avec la richesse du milieu, d'une manière qui maximise leur taux de croissance.
 Mais comme décrit au cours du Chapitre~\ref{chap:introduction}, ces lois ont été essentiellement établies en croissance stationnaire, un état très rarement rencontré par ces organismes dans leur milieu naturel.
@@
 Au cours du Chapitre~\ref{chap:theory}, nous avons utilisé un modèle d'auto-réplicateur pour évaluer la redistribution optimale des ressources lors d'un changement environnemental.
-En élargissant les principes d'optimalité établis dans les lois de croissance stationnaire, nous avons montré qu'une distribution de type tout-ou-rien (bang-bang en anglais) optimise la biomasse générée au cours de la transition.
+En élargissant les principes d'optimalité établis dans les lois de croissance stationnaire, nous avons montré qu'une distribution des ressources de type tout-ou-rien (\textit{bang-bang} en anglais) optimise la biomasse générée tant que la composition cellulaire n'est pas à l'équilibre.
 Cela nous a servi de point de référence pour comparer entre elles différentes stratégies de régulation.
-Ainsi, nous avons montré que les stratégies mesurant l'état interne de la cellule s'avèrent plus efficace que celles tirant leur information de l'environnement, contrairement à l'état stationnaire de croissance où toutes ces stratégies sont strictement équivalentes.
-De plus, une stratégie étonnamment proche du système ppGpp chez \textit{Escherichia coli}, est théoriquement capable de s'approcher de la distribution tout-ou-rien identifiée comme optimale.
+Ainsi, nous avons montré que les stratégies mesurant l'état interne de la cellule s'avèrent plus efficaces que celles tirant leur information de l'environnement, contrairement à l'état stationnaire de croissance où toutes ces stratégies sont strictement équivalentes.
+De plus, une stratégie légèrement plus complexe mais proche du système ppGpp chez \textit{Escherichia coli}, est théoriquement capable de s'approcher de la distribution tout-ou-rien identifiée comme optimale.
 
@@
 La plus critique s'avère être celle qui considère la production de biomasse comme étant le facteur de \textit{fitness} principal que l'on soit en environnement stable ou dynamique.
-Bien que largement répandue, cette hypothèse n'est finalement basée que sur des arguments indirects, ou bien établis dans des conditions de laboratoire, sur des souches de laboratoire (décrit dans le Chapitre~\ref{chap:introduction}).
-De fait, en conditions naturelles et à l'échelle des temps évolutifs, il n'est pas certain que ce critère soit vraiment si universel que ça.
-Une étape importante a donc été de vérifier expérimentalement les prédictions que nous avons tirées de cette hypothèse.
+Bien que largement répandue, cette hypothèse n'est finalement basée que sur des arguments indirects, ou bien établis dans des conditions de laboratoire, et sur des souches de laboratoire (Chapitre~\ref{chap:introduction}).
+De fait, en conditions naturelles et à l'échelle des temps évolutifs, il n'est pas certain que ce critère soit vraiment universel.
+Une étape importante a donc été de vérifier expérimentalement les prédictions que nous avons pu tirer de cette hypothèse.
 
 C'est ce qui a été abordé au cours du Chapitre~\ref{chap:experiments} de ce manuscrit.
-Nous avons mis en place une cadre d'étude expérimental dans lequel la distribution des ressources peut être observée lors d'une transition de croissance.
+Nous avons mis en place un cadre d'étude expérimentale dans lequel la distribution des ressources peut être observée lors d'une transition de croissance.
 Tout d'abord, une souche d'\textit{Escherichia coli} a été modifiée en attachant une protéine fluorescente (GFP) à l'une des sous-unités du ribosome.
@@
 Cela suggèreraient que les mécanismes de régulation de la distribution des ressources maximisent la biomasse produite lors d'une transition, et donc que les microorganismes sont effectivement adaptés à des environnements changeants.
-Pour confirmer ces résultats, de nouvelles données devront nécessairement être acquises, notamment en mesurant de telle transition dans une variété plus large de conditions environnementales (changement de source de carbone, ajout d'acides aminés dans le milieu pour contourner le métabolisme, utilisation d'une restriction en azote au lieu du carbone, ...).
+Pour confirmer ces résultats, de nouvelles données devront nécessairement être acquises, notamment en mesurant des transitions similaires dans une variété plus large de conditions environnementales (changement de source de carbone, ajout d'acides aminés dans le milieu pour contourner le métabolisme, utilisation d'une restriction en azote au lieu du carbone, ...).
 Le dispositif expérimental pourrait également être modifié pour contrôler un plus grand nombre de paramètres.
-Par exemple, en attachant un autre rapporteur fluorescent à une enzyme clé du métabolisme, nous pourrions vérifier si les synthèses de la machinerie d'expression génique et des enzymes du métabolisme sont effectivement en anti-phase lors des oscillations.
+Par exemple, en attachant un autre rapporteur fluorescent à une enzyme clé du métabolisme, nous pourrions vérifier si les synthèses de la machinerie d'expression génique et de la machinerie métaboliques sont effectivement en anti-phase lors des oscillations.
 
@@
 Par construction, cela signifierait qu'une hypothèse du modèle n'est pas respectée.
-Par exemple, la cellule ne maximiserait pas la biomasse produite lors d'une transition.
-Une autre possibilité serait que les coûts d'une telle régulation dépasse pour la cellule les éventuels bénéfices qu'elle peut en tirer.
-En effet, en ne comparant les schémas de régulation que vis-à-vis de leur bénéfices, nous avons fait l'hypothèse implicite que leurs coûts pour la cellule sont négligeables, sinon au moins comparables.
+Par exemple, cela pourrait signifier l'optimisation de la biomasse produite lors d'une transition n'est pas un objectif pour la cellule.
+Une autre possibilité serait que les coûts pour la cellule d'une telle régulation dépassent les éventuels bénéfices qu'elle peut en tirer.
+En effet, en ne comparant les schémas de régulation qu'en terme de bénéfices, nous avons fait l'hypothèse implicite que leurs coûts pour la cellule sont négligeables, sinon au moins comparables.
 Il est possible que même si la production de biomasse s'avère être un critère déterminant, les limites physiques imposées par le coût des régulations empêchent la cellule de faire mieux que le comportement observé.
-La prise en compte de ces coûts dans le modèle devra donc être explorée.
+La prise en compte de ces coûts dans le modèle devra donc être explorée avant toute conclusion.
 
@@
 En effet, comme montré dans le Chapitre~\ref{chap:theory}, la cellule ne tire avantage d'un schéma de régulation complexe que lorsque son environnement varie.
-On peut élargir ce résultat en supposant que le bénéfice apporté par des régulations toujours plus complexes devrait saturer.
-En revanche, leurs coûts devraient continuer à croitre, dans la mesure où des régulations plus élaborées impliquent pour la cellule la synthèse de davantage de système de mesure, et donc détournent une part toujours plus grande des ressources cellulaires.
+On peut élargir ce résultat en supposant que le bénéfice apporté par des régulations toujours plus complexes finit par saturer.
+En revanche, le coût de ces régulations devrait continuer à croitre, dans la mesure où des régulations plus élaborées impliquent pour la cellule la synthèse de davantage de systèmes de mesure, et donc détournent une part toujours plus grande des ressources cellulaires.
 On peut donc conjecturer que ces deux tendances vont finir par se croiser au niveau d'un goulot d'étranglement évolutif, au delà duquel les coûts excèdent les bénéfices.
-On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin vers la complexité en fonction de l'intensité et la fréquence des variations environnementales auxquels le microorganisme est soumis.
+On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin en fonction de l'intensité et la fréquence des variations environnementales auxquels les microorganismes sont soumis.
 En d'autres termes, là où les lois de croissances actuelles décrivent uniquement des variations physiologiques avec la qualité du milieu, nous pourrions ajouter une autre dimension à ces lois qui représenterait la variabilité temporelle de la disponibilité des nutriments dans l'environnement naturel.
+\end{chapter_summary}
 \selectlanguage{english}
@@
 \end{center}
+
+\section*{Beginning of Chapter \thechapter}
 

### 2017-01-04 abstract.tex
Hans modifications of english abstract

@@
 \begin{abstract}
-Growth is one of the most fundamental processes of life.
-In microbial cells, it relies on the transformation of matter and energy from the environment into diverse organic structures.
-Interestingly when comparing growth in different environments, microbial physiology exhibits general growth laws that relate the macromolecular composition of the cell to the growth rate.
+Growth is the most fundamental property of life.
+Growth consists in the transformation of matter and energy from the environment into diverse organic structures.
+Interestingly, general growth laws relate the macromolecular composition of the cell to growth rate.
 These laws are widespread and conserved in different microbial species, suggesting a fundamental principle of design.
-Indeed, recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation, and explained by the principles of natural selection.
-However, while these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
-The aim of this thesis is to extend the theoretical and experimental studies of microbial growth strategies to dynamical environments.
+Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation and explained by the principles of natural selection.
+However, the vast majority of these studies focus on steady-state growth.
+Such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
+The aim of this thesis is to extend the theoretical and experimental studies of microbial growth strategies to changing environments.
 
-Using a self-replicator model, we develop an intuitive framework that encapsulates the main features of growth through nutrient assimilation.
-We formulate dynamical growth maximization as an optimal control problem that the microbial cell must solve to allocate its resources between gene expression machinery and metabolism.
-We find a general solution using Pontryagin's Maximum Principle, and compare it with different possible implementations of growth control in bacterial cells.
-It shows that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
-A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
-Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
-It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
+Using a self-replicator model, we developed a theoretical framework that encapsulates the main features of growth.
+We formulate dynamical growth maximization as an optimal control problem that the microbial cell must solve in order to allocate the available resources to the gene expression machinery or to metabolism.
+Using Pontryagin's Maximum Principle, we have derived a general solution to the optimization problem and we have compared the optimal strategy with possible implementations of growth control in bacterial cells.
+Our results show that simple control strategies that maximize the growth-rate at steady state are suboptimal for transitions from one growth regime to another.
+We show that a near-optimal control strategy in dynamical conditions requires information about several, rather than a single, physiological variable.
+Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by the signaling molecule ppGpp in the enterobacterium \textit{Escherichia coli}.
+The strategy involves sensing a discrepancy between the concentrations of precursor metabolites and ribosomes, and the control of the rate of ribosome synthesis in a switch-like manner.
 
-While suggested by old literature data, this switch-like ribosome synthesis after a growth transition has never been experimentally confirmed.
-For this reason, we measure ribosomal abundance of \textit{Escherichia coli} at the single-cell level during a nutrient upshift.
-More precisely, we have constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing \textit{in-vivo} monitoring of the abundance of ribosomes.
-We monitor this strain in a microfluidic device designed for long-term imaging of individual cells in a continuous culture, and use it to simulate a nutrient upshift by changing the input medium.
-We develop a Kalman smoothing method that is able from these data to reconstruct how resources are allocated towards the synthesis of ribosomes.
-The results obtained are preliminary and do not allow a decisive validation, but they suggest the presence of oscillatory patterns after an upshift that could be reminiscent of the expected behavior.
+Even though this switch-like ribosome synthesis has been suggested by published data, the phenomenon has never been experimentally confirmed.
+We therefore measured ribosomal abundance in \textit{Escherichia coli} at the single-cell level during a nutrient upshift.
+More precisely, we constructed a strain in which a fluorescent marker has been attached to a ribosomal subunit, thus allowing \textit{in-vivo} monitoring of the abundance of ribosomes.
+We monitored this strain in a microfluidics device designed for long-term imaging of individual cells in a continuous culture, and used this experimental setup to simulate a nutrient upshift by changing the input medium.
+We developed a Kalman smoothing method for extracting quantitative information about resource allocation to ribosome synthesis from the raw data.
+Even though our preliminary results do not allow to reach a final conclusion, they do suggest the presence of oscillatory patterns after an upshift that are reminiscent of the expected behavior.
 
-Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
-All over, they demonstrate that when applied in dynamical instead of steady-state conditions, the same principles can generate unexpected behaviors and provide more information about the regulatory strategies employed by microorganisms.
-Such new understandings could be of great help for biotechnological purposes, in particular when using microorganisms to produce valuable elements for the health and energy industries.
+Our results demonstrate that the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.
+The proposed control scheme correctly reproduces the observed growth laws at steady state, but also predicts novel and unexpected behaviors when applied to a dynamical environment.
+Our improved understanding of the principles that govern the control of bacterial growth could be used for improving biotechnological processes, in particular those that use microorganisms to produce high valuable-added products for the chemical or biomedical industry.
 \end{abstract}
@@
 \begin{author-summary}
-Microbial cells sustain and replicate themselves from the matter and energy they extract from their environment.
-Strategies enabling microorganisms to optimize their replication rate have been extensively studied, but mostly in stable environments.
-Here, we build a simple model of microbial growth and use mathematical methods from optimal control theory to determine how the cells should behave when the environment around them is abruptly changed.
-Interestingly, we find that microbial cells should adopt an \textit{on-off} strategy: they should alternatively allocate all their resources to the production of each cell component, a prediction that we also tried to confirm experimentally.
-In other words, the regulations inside the bacterial cells could profit from functioning more like a switch, rather than a dim, in changing environments.
-These results might suggest new ways to optimize the biotechnological production of molecules of interest.
+Bacteria grow by exploiting matter and energy extracted from their environment.
+The strategies that enable bacteria to optimize their growth rate have been extensively studied.
+However, most of these studies were carried out in a constant environment.
+Here, we construct a simple model of microbial growth and use mathematical methods from optimal control theory to determine how the cells should behave when the environment changes abruptly.
+We find that microbial cells should adopt an \textit{on-off} strategy:
+they should alternatively allocate all of their resources to the production of two categories of cellular components, the elements needed for gene expression and the ones needed for the general metabolism of the cell.
+Our preliminary experiments confirm this prediction.
+In other words, theory tells us that switch-like mechanisms would optimize the growth of bacterial cells and our experiments validate the theory.
+This suggests new ways for optimizing biotechnological processes.
 \end{author-summary}

### 2017-01-04 Chapters/theory.tex
Layout fix, changing size of some figures

@@
 \centering
-\includegraphics[width=\textwidth]{./Fig/Fig4}
+\includegraphics[clip,trim={0.4cm 0.5cm 0.54cm 0.45cm}]{./Fig/Fig4}
 \caption{
@@
 \includegraphics[scale=1]{./Fig/Fig6}
-\caption{
+\caption{\small
 \textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
@@
 \includegraphics[scale=1]{./Fig/Fig7}
-\caption{
+\caption{\small
 \textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
@@
 As stated in the section \textit{Biomass maximization as an optimal control problem} in the main text, the objective of this study is to maximize the growth rate on an interval $[0,\tau]$ after a nutrient upshift.
-With Eq.~\ref{S1.eq:supp_growthrate_adim} in \ref{S1_Text}, we have
+With Eq.~\ref{eq:supp_growthrate_adim} in \ref{S1_Text}, we have
 
@@
 $\lambda_0=0$).
-In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.\\
+In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.
 
@@
 We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
-$$
-(-1)^q\frac{\partial}{\partial\alpha}\frac{d^{2q}}{d\hat{t}^{2q}}\phi(\hat{t})<0,
-$$
+\[
+(-1)^q \; \frac{\partial}{\partial\alpha} \; \frac{d^{2q}}{d\hat{t}^{2q}} \; \phi(\hat{t})<0,
+\]
 satisfying the Kelley condition necessary for optimality.
 Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (also called the Fuller's phenomenon, \textit{i.e.}, an arc with an infinite number of switches~\cite{borisov_fullers_2000,marchal_chattering_2013}).
-\\
 
@@
 We explain below how we use the model to compare the action of the ppGpp system with the on-off control strategy.
-The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} , and is reproduced in Table~\ref{S4.tab:notations} in order to make the text self-contained.
+The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} , and is reproduced in Table~\ref{tab:notations} in order to make the text self-contained.
 
@@
 \end{equation}
-where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{S4.tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:
+where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:
 \begin{equation}
@@
 \end{equation}
-From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{S4.eq:rttot-f}:
+From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{eq:rttot-f}:
 \begin{equation}
@@
 \begin{equation}
-\footnotesize
 \texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
 \end{equation}
-This function is represented in Fig.~\ref{S4.fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.
+This function is represented in Fig.~\ref{fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.
 

### 2017-01-04 Chapters/experiments.tex
Layout fix, changing size of some figures

@@
 \chapter{Dynamics of the gene expression machinery during an acetate-glucose upshift}
-\chaptermark{Gene expr. machinery during an upshift}
+\chaptermark{Measurements during an upshift}
 \label{chap:experiments}
@@
 Ces résultats s'avèrent finalement encourageants, et incitent à pousser plus en avant cette étude, notamment via l'acquisition de nouvelles données pour lesquelles les problèmes cités auront pu être corrigés (voir discussion en Section~\ref{sec:chap3_discussion}).
-
 \end{chapter_summary}
@@
 \includegraphics[scale=1]{./Fig/data_acquisition}
-\caption{
+\caption{\small
 \textbf{Results of data acquisition.}
@@
 \includegraphics[scale=1]{./Fig/synthetic_upshift}
-\caption{
+\caption{\small
 \textbf{Performance of the Kalman smoothing procedure on synthetic data simulating an acetate-glucose upshift.}
@@
 \centering
-\includegraphics[scale=1]{./Fig/suppinfo_caract}
-\caption{
+\includegraphics[scale=0.985]{./Fig/suppinfo_caract}
+\caption{\small
 \textbf{Growth curves in M9~0.2\% glucose for the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
-Growth and measurements were performed in a 96-well microplate placed in a Tecan infinite 200 pro reader thermostated at 37$^\circ$C.
-We monitored 4 wells inoculated with the WT strain, the \textit{rpsB-gfp} strain, and the \textit{rpsB-mCherry} strain, for a total of 12 wells each containing 150~$\mu$L of growth medium.
-Curves were shifted in time to correct for pipetting variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
-\textit{(A)}~The light absorbance as measured by the Tecan microplate reader, corrected for background by removing M9 absorbance.
-The superposition of the curves indicates similar growth rates between the WT and the modified strains.
-\textit{(B)}~The fluorescence measured in the wells, expressed in Relative Fluorescence Units (RFU).
+Growth and measurements were performed at 37$^\circ$C on a 96-well microplate in a Tecan infinite 200 pro reader.
+We monitored 4 wells for each of the WT, \textit{rpsB-gfp}, and \textit{rpsB-mCherry} strains.
+Curves were time shifted to correct for variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
+\textit{(A)}~600~nm absorbance as measured by the Tecan microplate reader, corrected for background by removing M9 absorbance.
+Curve superposition indicates similar growth rates between the WT and the modified strains.
+\textit{(B)}~Fluorescence measured in each well, expressed in Relative Fluorescence Units (RFU).
 Green fluorescence (485~nm excitation, 535~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
-Red fluorescence (560~nm excitation, 635~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
-As can be seen, fluorescence levels of the modified strains are far above the autofluorescence measured on the WT strain.
-\textit{(C)}~The ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
-Autofluorescence background was corrected by removing the autofluorescence measured on the WT wells.
-We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something does not occur on the \textit{rpsB-gfp} strain.
-Before each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
+Red fluorescence (560~nm exc., 635~nm em.) is measured for the WT (dashed lines) and the \textit{rpsB-mCherry} (solid lines) strains.
+Fluorescence levels of the modified strains are far above the autofluorescences measured on the WT strain.
+\textit{(C)}~Ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
+Autofluorescence background was corrected by removing the fluorescence measured on the WT.
+We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something that does not occur on the \textit{rpsB-gfp} strain.
+At each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
 }
@@
 \includegraphics[scale=1]{./Fig/cell1}
-\caption{
+\caption{\small
 \textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y1, W11".}
@@
 \includegraphics[scale=1]{./Fig/cell2}
-\caption{
+\caption{\small
 \textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y1, W7".}
@@
 \centering
-\includegraphics[scale=1]{./Fig/cell3}
-\caption{
+\includegraphics[scale=0.99]{./Fig/cell3}
+\caption{\small
 \textbf{Complete analysis for the cell located at the bottom of the well labelled "X3Y2, W5".}
@@
 \includegraphics[scale=1]{./Fig/cell4}
-\caption{
+\caption{\small
 \textbf{Complete analysis for the cell located at the bottom of the well labelled "X2Y2, W3".}
@@
 \includegraphics[scale=1]{./Fig/cell5}
-\caption{
+\caption{\small
 \textbf{Complete analysis for the cell located at the bottom of the well labelled "X1Y2, W6".}

### 2017-01-04 Chapters/introduction.tex
Hans' corrections of section 1.1 to 1.3 of the introduction'

@@
 At every moment of our lives, we are surrounded by billions of microscopic organisms.
-They sustain themselves by harvesting energy from their environment, in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
+They sustain themselves by harvesting energy from their environment in the form of organic matter, highly reactive chemicals, or light~\cite{madigan_biology_2006,schaechter_microbe_2006}.
 While most of our cells are kept in a stable and friendly internal environment (temperature, oxygenation, nutrient abundance, ...), microorganisms have evolved to constantly adapt their physiology to strongly variable conditions~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013,
@@
 Such ubiquity translates a strong potential for adaptation.
-The model organism \textit{Escherichia coli}, a bacteria first isolated from our intestinal tract, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes~\cite{zimmer_microcosm:_2009}.
+The model organism \textit{Escherichia coli}, a bacterium first isolated from our intestinal tract, is capable of growing on dozens of different carbon sources through changes in the expression of thousands of genes~\cite{zimmer_microcosm:_2009}.
 
 Microorganisms, along with all living cells, are mainly constituted of polymers (DNA, RNA, proteins) consisting of many repeated subunits (nucleotides, amino acids).
-The information about their sequences is stored in the form of genes physically located on the DNA.
+The information about their sequences is stored in the form of genes, physically located on the DNA.
 Gene expression is the process by which this information is used in the synthesis of a functional final product (protein or RNA).
-The gene expression machinery (itself constituted of RNA and proteins) synthesizes gene products after binding to target sequences either at the state of transcription (promoter on the DNA), or translation (ribosome binding site on the messaging RNA).
+The gene expression machinery (itself constituted of RNA and proteins) synthesizes gene products after binding to target sequences either by transcription (promoter on the DNA), or translation (ribosome binding site on the messaging RNA).
 In a given cell, there are roughly as many different target sequences as genes~\cite{keseler_ecocyc_2013}, which ensures that not all products are synthesized at the same rate.
@@
 The reorganization of gene expression controls the abundance of enzymes and ribozymes that catalyze the biochemical reactions allowing the cell to perform different functions.
-For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reductive power, nucleotides and amino acids~\cite{schaechter_microbe_2006}.
+For instance, assimilating a given nutrient involves the synthesis of the corresponding uptake protein and of the metabolic enzymes converting the nutrients into reducing power, nucleotides and amino acids~\cite{schaechter_microbe_2006}.
 These precursors are then used to produce new macromolecules, closing the loop of self-replication.
 In order to optimize the overall process in different environmental conditions, the cell must adjust the abundance of every enzyme involved in this network of biochemical reactions.
-By doing so, the cell actually resolves a resource allocation problem, where a common wealth (precursors) has to be distributed over thousands of potential beneficiaries (proteins and other macromolecules).
+By doing so, the cell actually solves a resource allocation problem, where a common wealth (precursors) has to be distributed over thousands of potential beneficiaries (proteins and other macromolecules).
 
@@
 
-For this reason, there is no consensus about what constitute the main fitness factors of microorganisms.
+For this reason, there is no consensus about what constitutes the main fitness factors of microorganisms.
 Competition for common resources seems to favor the maximization of offspring, in the sense that a replicator with more descendants will outnumber its competitors in the long term~\cite{dawkins_selfish_1976}.
@@
 When cultivated in minimal medium with acetate or succinate, \textit{Escherichia coli} uptake rates are correctly predicted by assuming that the metabolic network operates in a mode that maximizes growth rate~\cite{edwards_silico_2001}.
-This result is not general though, and depends on the carbon source and strain used~\cite{ibarra_escherichia_2002}.
+This result is not general though and depends on the carbon source and strain used~\cite{ibarra_escherichia_2002}.
 But even in other situations, the metabolic networks evolve rapidly towards growth-rate maximization if a constant nutrient-providing environment is maintained~\cite{ibarra_escherichia_2002}.
@@
 This figure displays the steady-state growth rate (in number of divisions per hour) of \textit{Escherichia coli} in a synthetic minimal medium at 37$^{\circ}$C containing different concentrations of glucose (a common carbon source).
-As the concentration of glucose increases, the growth rate of \textit{E. coli} also does by following the hyperbolic relationship described in Eq.~\ref{eq:monod_law}.
+As the concentration of glucose increases, so does the growth rate of \textit{E.~coli} by following the hyperbolic relationship described in Eq.~\ref{eq:monod_law}.
 The solid curve is obtained for $\mu_K = 1.35$ div h\textsuperscript{-1}, and $K_C = 0.22 \cdot 10^{-4}$ mol L\textsuperscript{-1}.
@@
 The growth laws cited above apply at steady state, where all intensive properties of the cell are time-invariant~\cite{schaechter_microbe_2006,fishov_microbial_1995}.
-It means that the properties that are independent of the cell volume or the cell mass (temperature, concentrations, ...) are constant over time, even though the cell is still growing.
+This means that the properties that are independent of the cell volume or the cell mass (temperature, concentrations, ...) are constant over time, even though the cell is growing.
 It requires that the components of the cell "increase by the same factor over a time interval", which has motivated the use of the term balanced growth~\cite{campbell_synchronization_1957}.
@@
 Its characteristics are time-invariant and quite robust accross conditions, which has made it a standard for microbial growth studies~\cite{schaechter_microbe_2006}.
-This phase can be extended for hundreds of generations in continuous cultures, by the constant renewal of the medium~\cite{borirak_molecular_2014,herbert_continuous_1956,wang_robust_2010}.
+This phase can be extended for hundreds of generations in continuous cultures by the constant renewal of the medium~\cite{borirak_molecular_2014,herbert_continuous_1956,wang_robust_2010}.
 As represented here, nutrients are quickly depleted in batch conditions, which does not allow to maintain steady-state growth for a long period of time.
@@
 In natural conditions, microorganisms usually encounter poor media and spend most of their time in stationary phase~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}.
-Some species which have evolved long-term resistance mechanisms, like sporulation~\cite{stragier_molecular_1996} or cannibalism~\cite{gonzalez-pastor_cannibalism:_2011}, can survive particularly long stationary phases.
+Some species that have evolved long-term resistance mechanisms, like sporulation~\cite{stragier_molecular_1996} or cannibalism~\cite{gonzalez-pastor_cannibalism:_2011}, can survive particularly long stationary phases.
 }
@@
 Microbial growth is essentially a resource allocation problem that can be summarized in simple, fundamental growth laws.
-Though established empirically, those laws are elegantly explained if we consider that microorganisms behave in way that optimizes biomass production.
+Though established empirically, those laws are elegantly explained if we consider that microorganisms behave in ways that optimize biomass production.
 However, these laws describe growth in stable environments and it is hard to imagine how natural selection could have resulted in microorganisms that are optimal for conditions they rarely encounter outside the laboratory.
-This motivates the study of growth laws from a dynamical perspective, that is, in a changing environment, and leads us to the following problem statement: \textbf{Which strategies microorganisms follow to dynamically reallocate their resources after a change in the environment?}
+This motivates the study of growth laws from a dynamical perspective, that is, in a changing environment, and leads us to the following problem statement: \textbf{Which strategies do microorganisms follow in order to dynamically reallocate their resources after a change in the environment?}
 The development of this question prompts the following investigations.
@@
 How could one measure resource allocation during a growth transition?
-In the light of the discussion in the previous section, it requires to monitor the abundance of key macromolecules in living cells over time, while establishing an experimental set-up that allows to control the inherent variability of dynamical experimental studies of microbial growth.
+In the light of the discussion in the previous section, this requires monitoring the abundance of key macromolecules in living cells over time, while establishing an experimental set-up that allows to control the inherent variability of dynamical experimental studies of microbial growth.
 
@@
 
-My recent experience as a teacher taught me that, as of today, a lot of students in biology do not like mathematics very much.
+My recent experience as a teacher taught me that, as of today, many students in biology do not like mathematics very much.
 Since mathematics is the main language of modeling, this has often made it difficult for me to convince them that modeling is of key importance for biologists.
@@
 Microbial physiology results from the interplay of thousands of chemical reactions that are not necessarily relevant in any given situation~\cite{schaechter_microbe_2006}.
-These reactions occurs on a wide range of time scales (shorter than 1~second for the metabolism, to more than 1~day for the degradation of stable proteins~\cite{schaechter_microbe_2006}), and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations~\cite{dawkins_selfish_1976}.
+These reactions occurs on a wide range of time scales (shorter than 1~second for metabolism, to more than 1~day for the degradation of stable proteins~\cite{schaechter_microbe_2006}), and are controlled by several layers of regulatory mechanisms that can only be understood through evolutionary considerations~\cite{dawkins_selfish_1976}.
 As a consequence, even the simplest verbal hypothesis can resist direct empirical testing.
@@
 
-More direct modeling approaches, in the form of large kinetic models, have also been helpful to understand growth-related processes.
+More direct modeling approaches, in the form of large kinetic models, have also been helpful for understanding growth-related processes.
 Through the construction of a model of 47 differential equations and 193 parameters, Kotte \textit{et al.} have shown how metabolic fluxes are sensed at different locations in central carbon metabolism, and are integrated in a global cellular response by the coupling of enzymatic and transcriptional regulation~\cite{kotte_bacterial_2010}.
@@
 For instance, a simple proof-of-concept model was used to uncover the principles leading to overflow metabolism, a mechanism by which microorganisms switch to inefficient metabolic pathways when growing at high nutrient availability~\cite{molenaar_shifts_2009}.
-This paradoxal and widespread phenomenon~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the fact that high yield pathways also require the synthesis of more enzymes, revealing the occurrence of a cost-benefit trade-off producing the switch when nutrients are no longer the limiting factor.
-Overall, these models have several advantages for the purpose of our study: they clearly state the underlying assumptions, and are sufficiently tractable to apply a variety of mathematical tools.
+This paradoxical and widespread phenomenon~\cite{dijken_kinetics_1993,vemuri_overflow_2006,mckeehan_glycolysis_1982,hsu_cancer_2008} was shown to be easily explained by the fact that high yield pathways also require the synthesis of more enzymes, revealing the occurrence of a cost-benefit trade-off producing the switch when nutrients are no longer the limiting factor.
+Overall, these models have several advantages for the purpose of our study: they clearly state the underlying assumptions, and are sufficiently tractable to be analyzed by a variety of mathematical tools.
 
@@
 The most straightforward method is to work at the population level.
-When working with microorganisms, a clonal population of genetically identical cells can easily be obtained by inseminating a single colony into a growth medium~\cite{schaechter_microbe_2006}.
+When working with microorganisms, a clonal population of genetically identical cells can easily be obtained by inoculating a single colony into a growth medium~\cite{schaechter_microbe_2006}.
 Over time, this culture can be subjected to different types of measurements: for instance, biomass can be estimated  directly through measurements of the dry weight of samples~\cite{monod_growth_1949}, or indirectly through measurements of transmitted or scattered light across the culture~\cite{volkmer_condition-dependent_2011}, allowing to construct growth curves as presented in Fig.~\ref{fig:growth_curve}.
-Other population-wide parameters like the macromolecular composition~\cite{scott_interdependence_2010,scott_bacterial_2011} or the concentration of metabolites~\cite{bennett_absolute_2009} can also be evaluated.
-They represent averaged estimates from billions of genetically identical cells, and are thus usually robust and reliable measurements.
+Other population-wide parameters, like the macromolecular composition~\cite{scott_interdependence_2010,scott_bacterial_2011} or the concentration of metabolites~\cite{bennett_absolute_2009} can also be evaluated.
+They represent averaged estimates from billions of genetically identical cells and are thus usually robust and reliable measurements.
 In fact, most of \textit{E.~coli} parameters that are currently used in today's growth models have been measured at the population level, notably through the emblematic work of Bremer and Dennis~\cite{churchward_macromolecular_1982,bremer_modulation_1996,bremer_free_2003,bremer_feedback_2008}.
@@
 One example are questions about the internal structure of the cell~\cite{bakshi_superresolution_2012}.
-Bakshi \textit{et al.} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerazes and DNA in living \textit{E.
-coli} cells~\cite{bakshi_superresolution_2012}.
+Bakshi \textit{et al.} used a combination of fluorescent labeling and superresolution imaging to localize the ribosomes, RNA polymerases and DNA in living \textit{E.~coli} cells~\cite{bakshi_superresolution_2012}.
 They showed that despite what was previously thought, most of the translation occurs far from the DNA, on free mRNA molecules that migrate to ribosome-rich regions~\cite{bakshi_superresolution_2012}.
@@
 With the advent of molecular genetics in the 1970s and 1980s, batch growth conditions have been the preferred choice of most microbiologists~\cite{hoskisson_continuous_2005}.
-As illustrated in Fig.~\ref{fig:growth_curve}, the organism is inseminated in a closed-system and grows until the nutrient is depleted.
+As illustrated in Fig.~\ref{fig:growth_curve}, the organism is inoculated in a closed-system and grows until the nutrient is depleted.
 In a sense, such a condition is close to what microbes encounter in nature, where key nutrients are only available for a short time and quickly depleted~\cite{mcarthur_microbial_2006,menge_nitrogen_2012,hobbie_microbes_2013}, and is essential to study many questions, \textit{e.g.} diauxic growth~\cite{kremling_understanding_2015}.
@@
 Parallel cultivation has notably been exploited to construct roughly 4,000 single-cell knock-out mutants of \textit{Escherichia coli}, forming the Keio collection~\cite{baba_construction_2006}.
-This mutant collection was intensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4 millions samples issued from this collection had been shared worldwide~\cite{yamamoto_update_2009}).
+This mutant collection was extensively used during the last decade to unveil unknown gene functions and test genome-wide effects of gene deletions (as of 2009, more than 4~millions samples issued from this collection had been shared worldwide~\cite{yamamoto_update_2009}).
 Microplates have also helped building a massive library of transcriptional fusions of the green fluorescent protein to each of about 2,000 different promoters in \textit{E.~coli}~\cite{zaslaver_comprehensive_2006}.
@@
 ronen_assigning_2002,stefan_inference_2015}.
-But the later use can be hazardous, because the main drawback of the batch condition is that the culturing system is poorly controlled.
+But the latter use can be hazardous, because the main drawback of the batch condition is that the culturing system is poorly controlled.
 Changes of key chemical parameters (\textit{e.g.} pH, pO\textsubscript{2}) have been shown to occur during the whole growth curve~\cite{bekker_changes_2007}. 
@@
 
-For this reason, we early observed development of continuous cultures for microbial studies~\cite{myers_culture_1944,novick_experiments_1950,herbert_continuous_1956}.
-Through permanent renewal of the growth medium, continuous cultures allow for maintenance of a chemically well-controlled environment, allowing for the acquisition of reproducible and reliable data~\cite{borirak_molecular_2014,hoskisson_continuous_2005}.
+For this reason, continuous cultures were developed for microbial studies~\cite{myers_culture_1944,novick_experiments_1950,herbert_continuous_1956}.
+Through the permanent renewal of the growth medium, continuous cultures allow for maintenance of a chemically well-controlled environment, allowing for the acquisition of reproducible and reliable data~\cite{borirak_molecular_2014,hoskisson_continuous_2005}.
 This has shown to be particularly valuable for comparative omic analysis, for instance the analysis of changes in protein levels~\cite{kolkman_comparative_2005} or genome-wide transcriptomic changes~\cite{boer_genome-wide_2003} that occur when \textit{Saccharomyces cerevisiae} is grown in different media.
@@
 in continuous culture using only a few milliliters of fresh medium per hour.
-It has enabled to show that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
+This has enabled to show that \textit{E.~coli} growth is remarkably stable on the long term and rather immune to the aging mechanisms that normally affect mother cells in other microorganisms.
 It is however important to note that at those scales, growth can only be monitored through microscopy analysis, which could unnecessarily complicate studies that do not rely on single-cell measurements.
@@
 The experimental approaches briefly reviewed above can be applied to study microbial growth in a dynamical setting~\cite{levy_strategy_2007,levy_coordination_2009,madar_promoter_2013,ehrenberg_mediumdependent_2012}.
-For instance, Levy \textit{et al.}~\cite{levy_strategy_2007,levy_coordination_2009} applied pulse-like environmental perturbations to a continuous culture of yeasts, and measured how transcriptomic reorganization occurs.
+For instance, Levy \textit{et al.}~\cite{levy_strategy_2007,levy_coordination_2009} applied pulse-like environmental perturbations to a continuous culture of yeast, and measured how transcriptomic reorganization occurs.
 They observed that the transcription levels are affected before the growth rate of the organism, suggesting a significant role of feed-forward sensing from the environment.

### 2017-01-05 Chapters/introduction.tex
Hans corrections of 1.4

@@
 
-We start by developing in Chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation which evaluates how biomass production can be maximized during an upshift from a medium with low nutrient content to a medium with high nutrient content.
+We start by developing in Chapter~\ref{chap:theory} a simple proof-of-concept model of resource allocation that evaluates how biomass production can be maximized during an upshift from a medium with low nutrient content to a medium with high nutrient content.
 The model is an instance of a self-replicator model~\cite{molenaar_shifts_2009}, but focuses on the allocation of resources to only two sectors of the microbial cell: metabolism, taking up and converting nutrients to precursor metabolites, and gene expression, producing macromolecules from the precursors.
@@
 Specifically, we show that feedback from the intracellular state turns out to be much more valuable than information from the environment.
-We investigate if such a strategy could be active in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} as a possible molecular implementation of that should allow the synthesis of ribosomes in a switch-like manner during a nutrient upshift.
-Overall, this chapter demonstrates that if microbial cells actually optimize their biomass production during an upshift, they should dynamically allocate their resources to the gene expression machinery in an on-off manner, making an experimentally testable prediction.
+We investigate if such a strategy could be active in real cells, and identify the widespread ppGpp system~\cite{bosdriesz_how_2015} as a possible molecular implementation that could control the synthesis of ribosomes in a switch-like manner during a nutrient upshift.
+Overall, this chapter demonstrates that, if microbial cells actually optimize their biomass production during an upshift, they should dynamically allocate their resources to the gene expression machinery in an on-off manner, making an experimentally testable prediction.
 
-In Chapter~\ref{chap:experiments}, we address the challenging problem of experimentally verifying the above prediction in \textit{Escherichia coli}.
-It requires the development of an experimental set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
+In Chapter~\ref{chap:experiments}, we address the challenging problem of experimentally verifying this prediction in \textit{Escherichia coli}.
+This task requires the development of an experimental set-up allowing the measurement of resource allocation in our dynamical growth scenario, \textit{i.e.} a steady-state-to-steady-state nutrient upshift.
 From the model developed in Chapter~\ref{chap:theory}, we identified that the predicted on-off pattern should occur on a short time scale after the transition (a couple of generations).
 This motivates the use of \textit{in-vivo}, high-frequency, single-cell measurements of the concentration of the gene expression machinery.
-Inspired by the work of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, we construct a strain with a GFP-tagged S2 ribosomal subunit, grow this strain in the mother machine, and monitor expression of the reporter gene using fluorescence microscopy~\cite{wang_robust_2010}.
-While originally developed for studies of long-term steady-state growth, we use the mother machine here to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source), in order to get a significant difference in growth rates between the two media~\cite{andersen_are_1980}.
+Inspired by the work of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, we constructed a strain with a GFP-tagged S2 ribosomal subunit, grew this strain in the mother machine, and monitored expression of the reporter gene using fluorescence microscopy~\cite{wang_robust_2010}.
+While originally developed for studies of long-term steady-state growth, we used the mother machine here to perform an instantaneous medium transition in a controlled manner from acetate (a poor carbon source) to glucose (a rich carbon source), in order to obtain a significant difference in growth rates between the two media~\cite{andersen_are_1980}.
 
-After segmentation and cell tracking, this set-up is able to generate fluorescence and size time series for dozens of cells.
-Using again the self-replicator model as a framework, we show that these measurements are sufficient for the intended signal reconstructions, in particular the growth rate and the resource allocation profile over time.
-The method we use is called Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
-While well-known in engineering, Kalman smoothing has been rarely used in quantitative biology, but turns out to be particularly suitable for the purpose of the model-based analysis of gene expression data in single cells, as we argue and show on synthetic data through the chapter.
-Finally, the experimental verification or falsification of the optimal control prediction, should allow to conclude if the biomass maximization, as considered in Chapter~\ref{chap:theory}, is indeed a cell objective in a dynamical context.
+After segmentation and cell tracking, this set-up was able to generate fluorescence and size time series for dozens of cells.
+Using again the self-replicator model as a framework, we showed that these measurements are sufficient for the intended signal reconstructions, in particular the growth rate and the resource allocation profile over time.
+The method we used is called Kalman smoothing~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+While well-known in engineering, Kalman smoothing has been rarely used in quantitative biology, but turned out to be particularly suitable for the purpose of the model-based analysis of gene expression data in single cells, as we argue and show on synthetic data in Chapter~\ref{chap:experiments}.
+Finally, the experimental verification or falsification of the optimal control prediction, should allow to conclude whether the biomass maximization, as considered in Chapter~\ref{chap:theory}, is indeed a cell objective in a dynamical context.

### 2017-01-05 Chapters/discussion.tex
Last corrections from hans, also add list of figures and tables

@@
 \selectlanguage{french}
-\begin{chapter_summary}{Introduction}
+\begin{chapter_summary}{Discussion}
 Contrairement à sa cousine la physique, la biologie manque encore de théories quantitatives à fort pouvoir prédictif.
@@
 But evolution by itself is not sufficient to define life.
-Even the less restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
+Even the least restrictive definition we have -- the one we use to look for life in the universe -- defines a living system as "a self-sustaining chemical system capable of Darwinian evolution"~\cite{deamer_origins_1994,benner_defining_2010}.
 Self-sustainment appears to be as fundamental as evolution, and relies on the transformation of matter and energy from the environment into organic matter, in other words, on growth.
-But the mathematical framing of questions regarding the self-sustaining character of living systems has not advanced as much as the mathematical analysis of evolution.
+But the mathematical formulation of questions regarding the self-sustaining character of living systems has not advanced as much as the mathematical analysis of evolution.
 
 Nevertheless, fundamental growth laws have been established for microorganisms.
-These growth laws show that, regardless of the molecular mechanisms ensuring growth control, microorganisms tend to follow the same empirical regularities when grown at steady state in different environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
-In particular, they adjust their internal molecular composition after a change in the environment, in such a way as to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
+These growth laws show that, regardless of the molecular mechanisms underlying growth control, microorganisms tend to follow the same empirical regularities when grown at steady state in different environmental conditions~\cite{molenaar_shifts_2009,scott_emergence_2014,scott_interdependence_2010,scott_bacterial_2011}.
+In particular, they adjust their internal molecular composition after a change in the environment in such a way as to maximize their growth rate~(see in particular~\cite{molenaar_shifts_2009,scott_emergence_2014}).
 Growth laws are a strong support for the theory of a modular organization of microorganisms~\cite{scott_emergence_2014,hartwell_molecular_1999,arkin_fast_2006,guido_bottom-up_2006}.
@@
 It does not necessarily aim at quantitatively predicting or controlling the behavior of microorganisms.
-By using a simple and abstract representation, it enables a convenient and tractable way of evaluating the implications of dynamical optimality, in particular its divergence from steady-state optimality.
+By using a simple and abstract representation, it provides a convenient and tractable way of evaluating the implications of dynamical optimality, in particular its divergence from steady-state optimality.
 Is a dynamical perspective on growth laws necessary?
@@
 
-Some readers might be skeptical about the interest of using such a simple model for addressing the above questions.
-Given the profusion of knowledge and data available on biological systems, one might be tempted to introduce every known detail about the molecular implementation of a system of interest.
+Some readers may be skeptical about the interest of using such a simple model for addressing the above questions.
+Given the profusion of knowledge and data available on biological systems, one might be tempted to include in the model every known detail about the molecular implementation of the system of interest.
 One of the main drawbacks of such mechanistic models is that we quickly loose track of the big picture and the underlying assumptions.
-When including much molecular details, the models also tend to become untractable, prohibiting the use of mathematical tools like optimal control theory that are quickly overwhelmed by the sheer number of variables and parameters in the model.
+When including many molecular details, the models also tend to become untractable, prohibiting the use of mathematical tools like optimal control theory that are quickly overwhelmed by the sheer number of variables and parameters in the model.
 Besides, simple, abstract models make it easy for the modeler to state the assumptions that are made, and more importantly to identify which ones are logistical, exploratory, or critical (see~\cite{servedio_not_2014}, in particular Box~1).
@@
 
-Chapter~\ref{chap:experiments} must not be taken as a "test of the model" in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see Box~2 in~\cite{servedio_not_2014}).
+Chapter~\ref{chap:experiments} must not be taken as a "test of the model" presented in Chapter~\ref{chap:theory}, but rather as a test of its assumptions (see Box~2 in~\cite{servedio_not_2014}).
 The model is correct in that bang-bang resource allocation follows mathematically from the assumption that microorganisms maximize their biomass production at steady state and during growth transitions.
@@
 In the other scenario, we do not observe any trace of a bang-bang resource allocation scheme.
-In that case, the questions that would arise are which critical assumptions are not valid.
-We also cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might in fact be critical for the predictions of the model.
+In that case, the crucial questions is: "Which critical assumptions are not valid?"
+We cannot discard the possibility that an assumption that was originally identified as exploratory or logistical might in fact be critical for the predictions of the model.
 Although unlikely \textit{a priori}, this might be the case for the expressions used for the macroreactions in the model (Eqs~\ref{eq:metaflux}-\ref{eq:machflux}), the definition of the volume as invariably proportional to the total mass of macromolecules (Eq.~\ref{eq:voldef}), or the fact that degradation is assumed negligible with respect to the rates of other reactions in the system.
-More probably, the assumptions regarding the cost of the regulatory system could be more critical than initially thought.
+More likely, the assumptions regarding the cost of the regulatory system could be more critical than initially thought.
 By only comparing the regulatory schemes with respect to their benefits, we implicitly made the assumption that their costs are negligible or at least comparable.
-The costs of regulatory schemes are actually difficult to model~\cite{shachrai_cost_2010,dong_gratuitous_1995,dekel_environmental_2005}, since the cellular variables are not equally costly to measure and, as we showed for the ppGpp system, one mechanism may measure several variables at once.
+The costs of regulatory schemes are difficult to model~\cite{shachrai_cost_2010,dong_gratuitous_1995,
+dekel_environmental_2005} since the cellular variables are not equally costly to measure and, as we showed for the ppGpp system, one mechanism may measure several variables at once.
 Nevertheless, relaxation of this hypothesis would definitively need to be explored before concluding that microorganisms do not optimize their biomass production during growth transitions.
@@
 Overall, these two tendencies are expected to cross at a point that would represent an evolutionary bottleneck, beyond which the costs exceed the benefits (see~\cite{short_flows_2006} for another example of such a bottleneck).
-Interestingly, the position of the bottleneck might be dependent on the environment, resulting in a dynamic law that would link the complexity of regulatory systems to the dynamics of the environment.
+Interestingly, the position of the bottleneck might be dependent on the environment, resulting in a law that would link the complexity of regulatory systems to the dynamics of the environment.
 In other words, while the current growth laws consider how the physiology of the cell varies with the nutrient quality of the medium, we could add another dimension representing the time-varying availability of this nutrient in the natural environment.

### 2017-01-05 Chapters/experiments.tex
Last corrections from hans, also add list of figures and tables

@@
 
-The following pages contain the full annotated sequences of the region downstream \textit{rpsB} after the insertion of \textit{gfp\_mut2-cassette} and \textit{mCherry-cassette}, respectively (third line in Fig.~\ref{fig:Cloning}).
-The primers annealing positions are indicated using the same color code as in the description of the primers above.
+The following pages contain the full annotated sequences of the region downstream of \textit{rpsB} after the insertion of the \textit{gfp\_mut2-cassette} and the \textit{mCherry-cassette}, respectively (third line in Fig.~\ref{fig:Cloning}).
+The positions where the primers anneal to the template are indicated using the same color code as in the description of the primers above.
 The primers reported in the publication of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012} are indicated for information.
@@
 
-After the strains were made using the protocol described in Material and Methods~\ref{sec:methods_strain}, the downstream region of the \textit{rpsB} gene was amplified and sent for sequencing.
-Alignments of the results showed they exhibit the expected sequences, except for a couple of single-base mutations.
+After the strains were constructed using the protocol described in Material and Methods~\ref{sec:methods_strain}, the downstream region of the \textit{rpsB} gene was amplified and sequenced.
+Alignments of the results confirmed the expected sequences, except for a few single-base mutations.
 The raw results of the sequencing are reproduced below.
@@
 As for the construction of Bakshi \textit{et al.}~\cite{bakshi_superresolution_2012}, our strains possess a wild-type growth rate (Fig.~\ref{fig:suppinfo_caract}A), indicating that the gfp-tagging or mCherry-tagging of the ribosomal S2 subunit does not impede the functioning of the ribosome.
-It is known that \textit{E.~coli} cultures exhibit a high autofluorescence in the GFP bandwidth~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
-Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the corrected signal level for mCherry is even more stronger.
-Overall, both strains possess enough fluorescence to allow quantification at the population level, all the more so at the cell level because most part of the autofluorescence is concentrated in the medium outside the cell~\cite{mihalcescu_green_2015}.
-
-In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance (corrected means that we substracted the autofluorescence of the wild-type strain).
+It is known that \textit{E.~coli} cultures exhibit a high autofluorescence overlapping the GFP spectrum~\cite{mihalcescu_green_2015}, but the fluorescence of the \textit{rpsB-gfp} strain is at least 4-time higher than the \textit{wild-type} (Fig.~\ref{fig:suppinfo_caract}B).
+Contrary to green autofluorescence, red autofluorescence of \textit{E.~coli} cultures is very low (Fig.~\ref{fig:suppinfo_caract}B) so the corrected signal level for mCherry is even stronger.
+Overall, both strains possess enough fluorescence to allow quantification at the population level, all the more so at the cell level because most of the autofluorescence is concentrated in the medium outside the cell~\cite{mihalcescu_green_2015}.
+
+In Fig.~\ref{fig:suppinfo_caract}C, we display the ratio between the (corrected) fluorescence of the strain and the absorbance ("corrected" meaning that we substracted the autofluorescence of the wild-type strain).
 As a first approximation, this ratio can be used as a proxy for the fluorescence concentration in the cells, therefore for the ribosome concentration.
-The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty to obtain robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
-The RFU/Abs ratio is stable in the interval [400,600]~min for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, and can be interpreted as exponential steady-state growth.
-Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problem preventing the use of batch conditions to obtain robust data about growth transitions.
-Indeed, most of our attempt to perform robust growth transitions in this region were unsuccessful.
+The instability at the beginning of the experiment (before 400~min) is characteristic of the difficulty of obtaining robust estimations when the absorbance of the culture is low~\cite{zulkower_robust_2015}.
+The RFU/Abs ratio is stable in the interval [400,600]~min for both \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains, indicative of exponential steady-state growth.
+Note that this short time interval (a couple of generations) of readability before approaching the stationary phase is one of the main problems preventing the use of batch conditions for obtaining robust data about growth transitions.
+Indeed, most of our attempts to perform robust growth transitions in this region were unsuccessful.
 
@@
 As would be expected, the fluorescence of the \textit{rpsB-gfp} strain starts to slowly decrease as GFP proteins are degraded or photobleached.
-On the contrary, the fluorescence of the \textit{rpsB-mCherry} strain quickly triples when entering stationary phase, at a rate that is way higher than the physical limit imposed by the protein synthesis rate~\cite{scott_interdependence_2010}.
+On the contrary, the fluorescence of the \textit{rpsB-mCherry} strain quickly triples when entering stationary phase, at a rate that is much higher than the physical limit imposed by the protein synthesis rate~\cite{scott_interdependence_2010}.
 
@@
 \textbf{Growth curves in M9~0.2\% glucose for the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
-Growth and measurements were performed at 37$^\circ$C on a 96-well microplate in a Tecan infinite 200 pro reader.
+Growth and measurements were performed at 37$^\circ$C on a 96-well microplate in a Tecan infinite 200 pro.
 We monitored 4 wells for each of the WT, \textit{rpsB-gfp}, and \textit{rpsB-mCherry} strains.
 Curves were time shifted to correct for variability in the inoculation process (-50 min for WT, and -20 min for \textit{rpsB-mCherry}).
-\textit{(A)}~600~nm absorbance as measured by the Tecan microplate reader, corrected for background by removing M9 absorbance.
+\textit{(A)}~Absorbance at 600~nm, corrected for background by subtracting the absorbance of the M9 medium.
 Curve superposition indicates similar growth rates between the WT and the modified strains.
-\textit{(B)}~Fluorescence measured in each well, expressed in Relative Fluorescence Units (RFU).
+\textit{(B)}~Fluorescence measured in each well in Relative Fluorescence Units (RFU).
 Green fluorescence (485~nm excitation, 535~nm emission) is measured for the WT (dashed lines) and the \textit{rpsB-gfp} (solid lines) strains.
@@
 Fluorescence levels of the modified strains are far above the autofluorescences measured on the WT strain.
-\textit{(C)}~Ratio of fluorescence over absorbance, a proxy for the fluorescence concentration in the cells.
+\textit{(C)}~Ratio of fluorescence over absorbance (proxy for the fluorescence concentration in the cells).
 Autofluorescence background was corrected by removing the fluorescence measured on the WT.
-We observe a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain at stationary phase, something that does not occur on the \textit{rpsB-gfp} strain.
+We observed a strange increase in fluorescence concentration for the \textit{rpsB-mCherry} strain after entry into stationary phase, something that does not occur on the \textit{rpsB-gfp} strain.
 At each measurement cycle, the following procedure was applied: shaking (Orbital 6mm) for 30~s, shaking (Linear 6mm) for 30~s, waiting for 5~s.
@@
 \end{eqnarray}
-This system of differential equations can be analytically solved into:
+This system of differential equations can be analytically solved to yield:
 \[
@@
 \end{equation}
-Since growth is stopped, the bacterial volume is constant and then $\frac{P_m(t)}{P_m(0)}$ can be easily measured by directly taking the fluorescence normalized by its initial value, while the maturation rate $K$, the degradation rate $D$, and the initial fraction of immature proteins $\frac{P(0)}{P_m(0)}$ are free parameters that can be fitted on fluorescence time series.
+Since growth is stopped, the bacterial volume is constant and $\frac{P_m(t)}{P_m(0)}$ can now be easily measured by directly taking the fluorescence normalized by its initial value, while the maturation rate $K$, the degradation rate $D$, and the initial fraction of immature proteins $\frac{P(0)}{P_m(0)}$ are free parameters that can be fitted on fluorescence time series.
 The curves used for the estimation are represented in Fig.~\ref{fig:suppinfo_matur} and the results are reported in Tab.~\ref{tab:suppinfo_matur}.
 As expected, both reporter proteins are stable and exhibit a long half-life in exponential phase (>24~h).
-The maturation rate of GFP was too fast to obtain numerical value.
-However, the maturation rate of mCherry is rather slow, with a half-maturation time in the order of 30~min (time needed to mature 50\% of a pool of mCherry).
+The maturation rate of GFP was too fast for obtaining a numerical value.
+However, the maturation rate of mCherry is rather slow, with a half-maturation time on the order of 30~min (time needed to mature 50\% of a pool of mCherry).
 Even though degradation and maturation rate can be slightly different in stationary phase, the estimated value are not sufficient to explain the rapid increase in fluorescence observed in Fig.~\ref{fig:suppinfo_caract}.
@@
 \textbf{Estimation of the maturation and degradation rates of the reporter proteins in the \textit{rpsB-gfp} and \textit{rpsB-mCherry} strains.}
-The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except that a massive dose of Chloramphenicol was added in mid-exponential phase ($t=0$, vertical dashed grey line).
+The strains were grown in the same condition as in Fig.~\ref{fig:suppinfo_caract}, except that a high concentration of Chloramphenicol was added in mid-exponential phase ($t=0$, vertical dashed grey line).
 The fluorescence was normalized by the value at $t=0$ and the degradation and maturation rates were estimated by using the model in Eq.~\ref{eq:matur_deg}.
@@
 In particular, mCherry has been reported to be extremely sensible to the presence of oxygen for its maturation~\cite{doherty_stage-specific_2010}.
-We however reproduced the following results in conditions were the oxygen is not limiting.
+However, we reproduced the following results in conditions were the oxygen is not limiting.
 In particular, the abrupt increase of mCherry fluorescence is conserved when the stationary phase is attained at low bacterial density, in a larger volume (20-mL flask), and on several type of carbon sources (acetate, xylose, glycerol, maltose).
-We have so far no explanation for this strange dynamic, and therefore concentrated our effort on the \textit{rpsB-gfp} strain in the microfluidic experiments.
+We have so far no explanation for these strange dynamics, and we therefore concentrated our efforts on the \textit{rpsB-gfp} strain in the microfluidic experiments.
 
@@
 We estimated the noise in the RFU/pixel/cell measurements by using the data points just before the upshift (Fig.~\ref{fig:noise_poi}).
-Since they have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
-For this reason, ribosome concentration, and so fluorescence concentration, are expected constant.
-Fig.~\ref{fig:noise_poi} shows the distribution of points in this region, that can be approximated by a Gaussian distribution, for a particular cell.
+Since the cells have been growing for 20~h on acetate in the microfluidic device, bacteria are assumed to be in balanced growth in this region of $\sim$2~hours (roughly 23-24 points depending on the well of interest).
+For this reason, ribosome concentration, and therefore the fluorescence concentration, are expected to be constant.
+Fig.~\ref{fig:noise_poi} shows the distribution of points in this region, which can be approximated by a Gaussian distribution, for a particular cell.
 
@@
 \textit{(Top graph)} RFU/pixel/cell data points for the cell at the bottom of the well labeled X3Y2,W2.
-Noise estimation was performed on the points just before the upshift (red points), were the bacteria is assumed to grow at steady state.
+Noise estimation was performed on the points just before the upshift (red points), were the bacteria are assumed to grow at steady state.
 \textit{(Bottom graph)} Cumulative density function (CDF, in red) and probability density function (PDF, in gray) of the points highlighted in red on the top graph.
@@
 We did not investigate if this heterogeneity between cells is the result of true biological variations or a bias due to the microfluidic device.
-By looking at the points of Fig.~\ref{fig:noise_distrib_mean_std}, there are however no strong correlation of the mean and standard deviation with the acquisition field (XY), or the position of the well on the image (W).
+By looking at the points of Fig.~\ref{fig:noise_distrib_mean_std}, there are however no strong correlations of the mean and standard deviation with the acquisition field (XY), or the position of the well on the image (W).
 
@@
 Noise characteristics were computed by normalizing and aggregating data for the 45 normal cells presented in Fig.~\ref{fig:noise_distrib_mean_std}.
-Two scenarii were considered.
-First, the noise model can be considered additive, which means we have the measurement model presented in Section~\ref{sec:estim_gr_ra}:
+Two scenarios were considered.
+First, the noise is additive, which means that we have the measurement model presented in Section~\ref{sec:estim_gr_ra}:
 \begin{equation}
@@
 \end{equation}
-In other words, we can normalize the data points of each cell by removing their mean in order to aggregate the noise residues and get an estimation of the noise level (Fig.~\ref{fig:noise_estim_add_mul}, top panel).
-
-Alternatively, the noise model can be considered multiplicative.
-The noise model is than different than presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
+In other words, we can normalize the data points of each cell by removing their mean in order to aggregate the noise residues and thus obtain an estimation of the noise level (Fig.~\ref{fig:noise_estim_add_mul}, top panel).
+
+Alternatively, the noise could be multiplicative.
+The noise model is than different of the one presented in Section~\ref{sec:estim_gr_ra}, and can be re-written:
 \begin{equation}
@@
 For this reason, we decided in the main analysis to stick with an additive noise, the main reason being that the implementation used for the Kalman smoothing procedure are faster and more stable with an additive noise model.
-We assumed an additive white Gaussian noise with mean 0 and standard deviation 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
-As discussed in Section~\ref{sec:chap3_discussion}, the only way to decisively choose a noise model over the other would have been to obtain two different steady-state growths for each cell.
-Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future instances of the experiment.
+We assumed an additive white Gaussian noise with mean 0 and standard deviation of 28.29 for the simulation of the synthetic data and the signal reconstruction through the Kalman smoothing procedure.
+As discussed in Section~\ref{sec:chap3_discussion}, the only way to definitively choose a noise model over the other would have been to obtain two different steady-state growths for each cell.
+Unfortunately, the presented time series were not long enough to obtain a new steady state on glucose, something that will be corrected in future experiments.
 
@@
 Two noise models were considered for the normalization: either the temporal mean $m_k (\cdot)$ was removed from the measurements independently for each cell (additive noise model in Eq.~\ref{eq:noise_add}), or $1$ was removed from the ratio of the measurements with the temporal mean $m_k (\cdot)$ (multiplicative noise model in Eq.~\ref{eq:noise_mul}).
-For the additive noise model, the fitted Gaussian distribution as a mean of $1.256\cdot 10^{-15}$ and a standard deviation of 28.29.
-For the multiplicative noise model, the fitted Gaussian distribution as a mean of $2.666\cdot 10^{-18}$ and a standard deviation of 0.05383.
+For the additive noise model, the fitted Gaussian distribution has a mean of $1.256\cdot 10^{-15}$ and a standard deviation of 28.29.
+For the multiplicative noise model, the fitted Gaussian distribution has a mean of $2.666\cdot 10^{-18}$ and a standard deviation of 0.05383.
 }
@@
 \clearpage
-\subsection{S8 Text -- Examples of complete analysis for 5 normal cells}
+\subsection{S8 Text -- Examples of complete analysis for 5 cells}
 \manuallabel{S8_Text}{S8~Text}
 
-In order to get a more visual idea of the results of the analysis, we display below the complete signal reconstructions for 5 cells.
+In order to obtain a more visual idea of the results of the analysis, we display below the complete signal reconstructions for 5 cells.
 In Figs~\ref{fig:cell1}-\ref{fig:cell5}, the image on the left is the last image analyzed for the corresponding well.
-The bacteria either at the top or the bottom (depending on the orientation in the device) was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
+The cell either at the top or the bottom (depending on the orientation in the device) was manually segmented by selecting two pixels at the poles on the fluorescence images (red cross).
 A 6-pixel-wide rectangular mask was computed for each image, resulting in the masked image on the right.
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seems to have stabilized before the upshift.
-Oscillations in the RFU/pixel/cell signal are clearly visible, and result in oscillations in the resource allocation signal reconstruction.
+Growth rate and resource allocation are particularly unstable at the beginning of the experiment, but seem to stabilize before the upshift.
+Oscillations in the RFU/pixel/cell signal are clearly visible and result in oscillations in the resource allocation signal reconstruction.
 }
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-The reconstruction on this cell is a good instance of what was expected: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
+The reconstruction on this cell is a good example of what was expected: the growth rate is stable on acetate, than quickly increases after the upshift on glucose.
 The resource allocation is stable on acetate (indicating a steady state) than starts to oscillate after the upshift on glucose, even though the smoothing factor seems a little too high for this particular cell (some oscillations are poorly predicted after the upshift, in particular in the interval [200,400]~min).
@@
 (Figure description available in the introduction of \ref{S8_Text}.)\newline
-The results on this cell are particularly terrible: we do not seem to get a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating before the upshift.
+The results on this cell are particularly questionable: we do not seem to obtain a good steady state on acetate since the reconstructed resource allocation appears either negative or oscillating before the upshift.
 It is thus impossible to conclude if the oscillations observed after the upshift are real or due to overfitting.
-This cell highlights why robust steady states before and after the transition are important to conclude on the signal reconstruction.
+This cell highlights why robust steady states before and after the transition are important for a reliable signal reconstruction.
 }
@@
 \clearpage
-\subsection{S2 Figure -- Cell categories identified in the microscopy analysis}
+\subsection{S2 Figure -- Cell categories identified in the analysis}
 \manuallabel{S2_Figure}{S2~Fig}
@@
 \textbf{Cell categories identified in the microscopy analysis.}
-As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: dying cells stop growing (lowest graphs) , pausing cells present growth rate oscillations (middle graph), and normal cells does not present any of the above (highest graph).}
+As stated in Section~\ref{sec:estim_gr_ra}, the reconstruction of the growth rate for the 68 available cells motivated their classification in three categories: dying cells stop growing (lowest graphs) , pausing cells present growth rate oscillations (middle graph), and normal cells that do not present any of the above (highest graph).}
 \label{fig:subcat_cells}

### 2017-01-05 Chapters/theory.tex
Last corrections from hans, also add list of figures and tables

@@
 \includegraphics[width=0.7\textwidth]{./Fig/FigS4-1}
-\caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
-{
-{\bf ppGpp concentration is a function of total ribosome and amino acid concentrations.}\newline
+\caption{
+\textbf{ppGpp concentration is a function of total ribosome and amino acid concentrations.}
 We assume the dynamics of ppGpp to be fast on the time-scale of changes in the ribosome and amino acid concentrations.
@@
 \caption{
-\noindent\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
+\textbf{Simple control strategies for the self-replicator of bacterial growth.}
+\textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.
 The dashed, black curve is the (unique) strategy driving the system exactly to the optimal steady state, that is, the state in which growth occurs at the maximal rate supported by $E_M$.

### 2017-01-19 titlepage.tex
Fixing the title page and the template (du laboratoire and not de)
Final version, printed
Update affiliations, sent to reviewers instead of the other

@@
 \Laboratoire{Laboratoire Interdisciplinaire de Physique}
-\EcoleDoctorale{Ecole doctorale Chimie et Science du Vivant}  
+\EcoleDoctorale{Chimie et Sciences du Vivant}  
 \Titre{Microbial growth control in changing environments}
 \Soustitre{Theoretical and experimental study of resource allocation in \textit{Escherichia coli}}      %%optionnel : à décommenter et à renseigenr si présence d'un sous-titre de thèse
-\Depot{ (fin Mars)} %% DATE DE LA SOUTENANCE
+\Depot{23 mars 2017} %% DATE DE LA SOUTENANCE
 
@@
 
-\UGTExaminateur{M, Olivier BERNARD}{DR}{Inria Nice}     %% 2ème examinateur
+\UGTExaminateur{M, Olivier BERNARD}{DR}{Inria Sophia-Antipolis}     %% 2ème examinateur
 \UGTExaminateur{M, Matthieu JULES}{CR}{Inra Grignon}     %% 1er examinateur
@@
 
-\UGTDirecteur{M, Johannes GEISELMANN}{Prof.}{Univ. Grenoble Alpes}       %% Directeur de thèse
-\UGTCoDirecteur{M, Hidde DE JONG}{DR}{Inria Grenoble}     %% Co-Directeur de thèse s'il y en a
-\UGTInvite{M, Jean-Luc GOUZÉ}{DR}{Inria Nice}
+%\UGTDirecteur{M, Johannes GEISELMANN}{Prof.}{Univ. Grenoble Alpes}       %% Directeur de thèse
+%\UGTCoDirecteur{M, Hidde DE JONG}{DR}{Inria Grenoble}     %% Co-Directeur de thèse s'il y en a
+\UGTInvite{M, Jean-Luc GOUZÉ}{DR}{Inria Sophia-Antipolis}
 %\UGTInvitee{...Civilité, Prénom-et-Nom...}{...titre-et-affiliation...}

### 2017-04-04 Chapters/discussion.tex
french proofreading abstract chap 4

@@
 La simplicité du modèle, même si elle peut rendre certains lecteurs sceptiques, s'est avérée cruciale dans notre étude.
-En effet, cela nous a permis de clairement spécifier les hypothèses mise en jeu.
+En effet, cela nous a permis de clairement spécifier les hypothèses mises en jeu.
 La plus critique s'avère être celle qui considère la production de biomasse comme étant le facteur de \textit{fitness} principal que l'on soit en environnement stable ou dynamique.
@@
 C'est ce qui a été abordé au cours du Chapitre~\ref{chap:experiments} de ce manuscrit.
-Nous avons mis en place un cadre d'étude expérimentale dans lequel la distribution des ressources peut être observée lors d'une transition de croissance.
+Nous avons mis en place un cadre d'étude expérimental dans lequel la distribution des ressources peut être observée lors d'une transition de croissance.
 Tout d'abord, une souche d'\textit{Escherichia coli} a été modifiée en attachant une protéine fluorescente (GFP) à l'une des sous-unités du ribosome.
@@
 Dans le premier cas, le caractère tout-ou-rien ou du moins oscillatoire de la synthèse des ribosomes pourrait être confirmé.
-Cela suggèreraient que les mécanismes de régulation de la distribution des ressources maximisent la biomasse produite lors d'une transition, et donc que les microorganismes sont effectivement adaptés à des environnements changeants.
+Cela suggèrerait que les mécanismes de régulation de la distribution des ressources maximisent la biomasse produite lors d'une transition, et donc que les microorganismes sont effectivement adaptés à des environnements changeants.
 Pour confirmer ces résultats, de nouvelles données devront nécessairement être acquises, notamment en mesurant des transitions similaires dans une variété plus large de conditions environnementales (changement de source de carbone, ajout d'acides aminés dans le milieu pour contourner le métabolisme, utilisation d'une restriction en azote au lieu du carbone, ...).
@@
 Par construction, cela signifierait qu'une hypothèse du modèle n'est pas respectée.
-Par exemple, cela pourrait signifier l'optimisation de la biomasse produite lors d'une transition n'est pas un objectif pour la cellule.
+Par exemple, cela pourrait signifier que l'optimisation de la biomasse produite lors d'une transition n'est pas un objectif pour la cellule.
 Une autre possibilité serait que les coûts pour la cellule d'une telle régulation dépassent les éventuels bénéfices qu'elle peut en tirer.
@@
 On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin en fonction de l'intensité et la fréquence des variations environnementales auxquels les microorganismes sont soumis.
-En d'autres termes, là où les lois de croissances actuelles décrivent uniquement des variations physiologiques avec la qualité du milieu, nous pourrions ajouter une autre dimension à ces lois qui représenterait la variabilité temporelle de la disponibilité des nutriments dans l'environnement naturel.
+En d'autres termes, là où les lois de croissance actuelles décrivent uniquement des variations physiologiques avec la qualité du milieu, nous pourrions ajouter une autre dimension à ces lois qui représenterait la variabilité temporelle de la disponibilité des nutriments dans l'environnement naturel.
 \end{chapter_summary}

### 2017-04-05 Chapters/experiments.tex

@@
 De telles expériences s'avèrent malheureusement pointues et, dans le temps disponible pour le doctorat, nous n'avons pu en réaliser qu'un petit nombre, dont seule celle présentée ici s'est avérée exploitable.
-Celle-ci a cependant subit de nombreux problèmes expérimentaux qui, s'ils n'ont pas empêché complètement son exploitation, ont posé des problèmes pour l'analyse et l'interprétation des résultats.
+Celle-ci a cependant subi de nombreux problèmes expérimentaux qui, s'ils n'ont pas empêché complètement son exploitation, ont posé des problèmes pour l'analyse et l'interprétation des résultats.
 Ces problèmes expérimentaux sont abordés dans la Section~\ref{sec:data_acquisition}, et sont largement discutés en Section~\ref{sec:chap3_discussion}.
@@
 Nous montrons comment l'information dont nous disposons sur le déroulement de l'expérience peut être injectée dans l'analyse sous la forme d'un \textit{prior} bayésien afin de faciliter la reconstruction, ce qui s'avère particulièrement utile sur nos données.
-De plus, à partir de données synthétiques générée grâce au modèle développé lors du Chapitre~\ref{chap:theory}, nous constatons que la méthode est capable de capturer le profil oscillatoire généré par le comportement tout-ou-rien que nous attendons (Fig.~\ref{fig:synthetic_upshift}).
+De plus, à partir de données synthétiques générées grâce au modèle développé lors du Chapitre~\ref{chap:theory}, nous constatons que la méthode est capable de capturer le profil oscillatoire généré par le comportement tout-ou-rien que nous attendons (Fig.~\ref{fig:synthetic_upshift}).
 

### 2017-04-05 Chapters/introduction.tex

@@
 \begin{chapter_summary}{Introduction}
-Malgré leur apparente simplicité, les microorganismes sont des êtres vivants complexes capable de survivre dans les milieux les plus extrêmes.
+Malgré leur apparente simplicité, les microorganismes sont des êtres vivants complexes capables de survivre dans les milieux les plus extrêmes.
 Une telle ubiquité repose en partie sur leurs excellentes capacités d'adaptation.
@@
 L'une d'entre elles est une loi de croissance qui décrit la relation linéaire existant entre la concentration en ribosomes dans la cellule et le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
-De manière étonnante, quel que soient les détails moléculaires qui permettent cette adaptation, cette loi émerge lorsque la cellule cherche à maximiser son taux de croissance dans chaque environnement.
+De manière étonnante, quels que soient les détails moléculaires qui permettent cette adaptation, cette loi émerge lorsque la cellule cherche à maximiser son taux de croissance dans chaque environnement.
 
@@
 Pour tester les prédictions faites expérimentalement, nous chercherons à mesurer cette distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
-Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse des séries temporelles, permettent d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu, notamment la transition d'un milieu pauvre vers un milieu riche (\textit{nutrient upshift}).
+Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse des séries temporelles, permet d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu, notamment la transition d'un milieu pauvre vers un milieu riche (\textit{nutrient upshift}).
 \end{chapter_summary}

### 2017-04-05 Chapters/theory.tex

@@
 $\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle, est produite à la proportion $1-\alpha$).
-Comme on l'a déjà vu avec les lois de croissance établie à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
-Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaire, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?
-
-Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissances à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
+Comme on l'a déjà vu avec les lois de croissance établies à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
+Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaires, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?
+
+Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissance à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
 En Section~\ref{sec:growthlaws}, nous voyons en effet que pour chaque environnement, un $\alpha$ unique existe qui maximise la croissance de la cellule, mais surtout que ce $\alpha^*_{opt}$ est d'autant plus important que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
-De fait, les lois de croissances à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
+De fait, les lois de croissance à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
 Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?
@@
 Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
-Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
+Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant l'utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
 La solution optimale obtenue montre que le contrôle $\alpha$ doit prendre alternativement les valeurs 0 ou 1 jusqu'à ce que le système soit stabilisé sur un nouvel équilibre (Fig.~\ref{fig:optimalcontrol} et Eq.~\ref{eq:optcontrol}).
@@
 
-En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elles différentes stratégies de régulations (Fig.~\ref{fig:strategies_overview}).
-En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaire, nous montrons que deux stratégies de régulations simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
+En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elles différentes stratégies de régulation (Fig.~\ref{fig:strategies_overview}).
+En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaire, nous montrons que deux stratégies de régulation simple sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
 De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc strictement équivalentes.

### 2017-04-05 abstract.tex
Layout fix, changing size of some figures

@@
+%\newgeometry{left=2.5cm,right=2.5cm,top=1.5cm,bottom=2cm}
+
 \selectlanguage{english}
@@
 \begin{resume-vulgaire}
-Bactéries et autres micro-organismes se multiplient en utilisant la matière et l'énergie présentent dans leur environnement.
+Bactéries et autres micro-organismes se multiplient en utilisant la matière et l'énergie présentes dans leur environnement.
 Les stratégies qui leur permettent d'optimiser ce processus ont été longuement étudiées, mais uniquement dans des environnements stables.
-Ici, nous construisons un modèle simple de la croissance microbienne et utilisons la théorie du contrôle optimal pour déterminer la façon dont les cellules devrait agir lorsque leur environnement est soudainement modifié.
+Ici, nous construisons un modèle simple de la croissance microbienne et utilisons la théorie du contrôle optimal pour déterminer la façon dont les cellules devraient agir lorsque leur environnement est soudainement modifié.
 Nous mettons en évidence que la meilleure stratégie consiste à aiguiller toutes les ressources disponibles vers la production d'un seul composant de manière alternative, une prédiction que nous tentons également de confirmer de manière expérimentale.
@@
 %TC:endignore 
+%\restoregeometry

### 2017-04-05 abstract.tex
Correction for abstract

@@
 
-Nos résultats montrent que la capacité des systèmes de régulations à intégrer l'état de différentes variables physiologiques est crucial dans l'optimisation de la croissance en environnement variable.
-Au final, nous démontrons que les principes utilisés à l'état stationnaire peuvent, lorsqu'ils sont appliqués en dynamique, générer des comportements inattendus et expliquer plus en détails les stratégies de régulations employées par les micro-organismes.
+Nos résultats montrent que la capacité des systèmes de régulation à intégrer l'état de différentes variables physiologiques est cruciale dans l'optimisation de la croissance en environnement variable.
+Au final, nous démontrons que les principes utilisés à l'état stationnaire peuvent, lorsqu'ils sont appliqués en dynamique, générer des comportements inattendus et expliquer plus en détails les stratégies de régulation employées par les micro-organismes.
 \end{abstract}

### 2017-04-05 Chapters/introduction.tex
Corrections chap 1

@@
 Depuis, d'autres lois fondamentales ont pu être identifiées.
-L'une d'entre elles est une loi de croissance qui décrit la relation linéaire existant entre la concentration en ribosomes dans la cellule et le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
+L'une d'entre elles est une loi de croissance qui décrit la relation linéaire existante entre la concentration en ribosomes dans la cellule et le taux de croissance de l'organisme (Fig.~\ref{fig:scott_rnaprot}).
 De manière étonnante, quels que soient les détails moléculaires qui permettent cette adaptation, cette loi émerge lorsque la cellule cherche à maximiser son taux de croissance dans chaque environnement.
@@
 Cet état est atteint lorsque l'organisme se multiplie pendant suffisamment longtemps dans un environnement stable (Fig.~\ref{fig:growth_curve}).
-Il s'avère très pratique pour les études théoriques et expérimentales, notamment car les composants de la cellule ne changent plus au cours de temps, ce qui lui vaut aussi l'appellation de croissance équilibrée (\textit{balanced growth}).
+Il s'avère très pratique pour les études théoriques et expérimentales, notamment car les composants de la cellule ne changent plus au cours du temps, ce qui lui vaut aussi l'appellation de croissance équilibrée (\textit{balanced growth}).
 Mais cet état a beau être pratique, il est très loin des conditions rencontrées par les microorganismes dans leur milieu de vie naturel.

### 2017-04-05 Chapters/theory.tex
Corrections chap 2

@@
 En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elles différentes stratégies de régulation (Fig.~\ref{fig:strategies_overview}).
-En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaire, nous montrons que deux stratégies de régulation simple sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
+En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaire, nous montrons que deux stratégies de régulation simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
 De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc strictement équivalentes.

### 2017-04-05 Chapters/experiments.tex
Corrections chap 3

@@
 Couplée à de la vidéomicroscopie, cette souche nous permet de quantifier en temps réel la quantité de ribosomes au niveau de cellules uniques, ce qui procure un excellent moyen d'obtenir la concentration en machinerie d'expression génique (le ribosome étant son composant principal).
-Le taux de croissance quant à lui, peut-être facilement reconstruit en mesurant la taille des cellules au cours du temps.
+Le taux de croissance quant à lui, peut être facilement reconstruit en mesurant la taille des cellules au cours du temps.
 
@@
 Cependant, aussi pauvre que soit le premier milieu, il est important que les bactéries soient en phase de croissance, pour la raison que leur physiologie change complètement lorsqu'elles s'arrêtent de croître, ce qui crée des effets indésirables durs à contrôler expérimentalement.
-Les cellules doivent donc croître pendant suffisamment longtemps sur le premier milieu, sans épuisé celui-ci, puis une transition rapide doit être effectuée vers le milieu riche.
+Les cellules doivent donc croître pendant suffisamment longtemps sur le premier milieu, sans épuiser celui-ci, puis une transition rapide doit être effectuée vers le milieu riche.
 Nous avons utilisé pour cela la \textit{Mother Machine}, un dispositif microfluidique à la base inventé pour maintenir et observer les cellules bactériennes en état de croissance stationnaire pendant de longues périodes.

### 2017-04-05 Chapters/discussion.tex
Corrections chap 4

@@
 Contrairement à sa cousine la physique, la biologie manque encore de théories quantitatives à fort pouvoir prédictif.
-Même si de grosses avancées ont eu lieu en biologie évolutive, d'autres domaines sont loin d'être aussi développés d'un point de mathématique.
+Même si de grosses avancées ont eu lieu en biologie évolutive, d'autres domaines sont loin d'être aussi développés d'un point de vue mathématique.
 C'est notamment le cas de la croissance, qui malgré la place fondamentale qu'elle occupe dans ce qui définit un être-vivant, repose sur des mécanismes qui semblent varier énormément d'un organisme à l'autre, et pour lesquels des lois fondamentales sont dures à identifier.
@@
 Le dispositif expérimental pourrait également être modifié pour contrôler un plus grand nombre de paramètres.
-Par exemple, en attachant un autre rapporteur fluorescent à une enzyme clé du métabolisme, nous pourrions vérifier si les synthèses de la machinerie d'expression génique et de la machinerie métaboliques sont effectivement en anti-phase lors des oscillations.
+Par exemple, en attachant un autre rapporteur fluorescent à une enzyme clé du métabolisme, nous pourrions vérifier si les synthèses de la machinerie d'expression génique et de la machinerie métabolique sont effectivement en anti-phase lors des oscillations.
 
@@
 Une autre possibilité serait que les coûts pour la cellule d'une telle régulation dépassent les éventuels bénéfices qu'elle peut en tirer.
-En effet, en ne comparant les schémas de régulation qu'en terme de bénéfices, nous avons fait l'hypothèse implicite que leurs coûts pour la cellule sont négligeables, sinon au moins comparables.
+En effet, en ne comparant les schémas de régulation que par leurs bénéfices, nous avons fait l'hypothèse implicite que leurs coûts pour la cellule sont négligeables, sinon au moins comparables.
 Il est possible que même si la production de biomasse s'avère être un critère déterminant, les limites physiques imposées par le coût des régulations empêchent la cellule de faire mieux que le comportement observé.
@@
 On peut élargir ce résultat en supposant que le bénéfice apporté par des régulations toujours plus complexes finit par saturer.
-En revanche, le coût de ces régulations devrait continuer à croitre, dans la mesure où des régulations plus élaborées impliquent pour la cellule la synthèse de davantage de systèmes de mesure, et donc détournent une part toujours plus grande des ressources cellulaires.
+En revanche, le coût de ces régulations devrait continuer à croître, dans la mesure où des régulations plus élaborées impliquent pour la cellule la synthèse de davantage de systèmes de mesure, et donc détournent une part toujours plus grande des ressources cellulaires.
 On peut donc conjecturer que ces deux tendances vont finir par se croiser au niveau d'un goulot d'étranglement évolutif, au delà duquel les coûts excèdent les bénéfices.
-On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin en fonction de l'intensité et la fréquence des variations environnementales auxquels les microorganismes sont soumis.
+On peut de plus s'attendre à ce que ce goulot d'étranglement survienne plus ou moins loin en fonction de l'intensité et la fréquence des variations environnementales auxquelles les microorganismes sont soumis.
 En d'autres termes, là où les lois de croissance actuelles décrivent uniquement des variations physiologiques avec la qualité du milieu, nous pourrions ajouter une autre dimension à ces lois qui représenterait la variabilité temporelle de la disponibilité des nutriments dans l'environnement naturel.

### 2017-04-05 Chapters/experiments.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

@@
-\chapter{Dynamics of the gene expression machinery during an acetate-glucose upshift}
+\chapter{Dynamics of resource allocation during an acetate-glucose upshift}
 \chaptermark{Measurements during an upshift}
@@
 \selectlanguage{french}
-\begin{chapter_summary}{Mesure dynamique de la machinerie d'expression génique lors d'une transition acétate vers glucose}
+\begin{chapter_summary}{Mesures dynamiques de l'allocation des ressources lors d'une transition acétate vers glucose}
 
@@
 La Section~\ref{sec:exp_design} s'attache à définir les conditions expérimentales dans lesquelles le comportement prédit lors du Chapitre~\ref{chap:theory} pourra être observé.
-En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que $\alpha (\cdot)$ lors de la transition peut être reconstruit au cours du temps si l'on dispose de mesures dynamiques du taux de croissance et de la concentration en machinerie d'expression génique.
+En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que $\alpha (\cdot)$ peut être reconstruit au cours du temps de la transition si l'on dispose de mesures dynamiques du taux de croissance et de la concentration en machinerie d'expression génique.
 Cependant, le comportement tout-ou-rien a une dynamique bien particulière qui peut facilement être offusquée au niveau d'une population de cellules si ces dernières ne sont pas synchrones.
-De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de la cellule unique, ce qui nous a conduit à utiliser des rapporteurs fluorescents de l'abondance d'une des machineries.
+De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de cellules individuelles, ce qui nous a conduit à utiliser des rapporteurs fluorescents de l'abondance d'une des machineries.
 En particulier, nous construisons une souche d'\textit{Escherichia coli} chez laquelle la sous-unité S2 du ribosome a été fusionnée avec une protéine fluorescente (GFP) (voir Section~\ref{sec:methods_strain} des matériels et méthodes).
 Couplée à de la vidéomicroscopie, cette souche nous permet de quantifier en temps réel la quantité de ribosomes au niveau de cellules uniques, ce qui procure un excellent moyen d'obtenir la concentration en machinerie d'expression génique (le ribosome étant son composant principal).
-Le taux de croissance quant à lui, peut être facilement reconstruit en mesurant la taille des cellules au cours du temps.
+Le taux de croissance quant à lui, peut être facilement reconstruit en mesurant l'évolution de la taille des cellules dans le temps.
 
@@
 Nous nous focalisons sur le cas particulier d'une transition d'un milieu de croissance pauvre vers un milieu riche.
-Cependant, aussi pauvre que soit le premier milieu, il est important que les bactéries soient en phase de croissance, pour la raison que leur physiologie change complètement lorsqu'elles s'arrêtent de croître, ce qui crée des effets indésirables durs à contrôler expérimentalement.
+Il est important que les bactéries soient en phase de croissance dans chaque milieu du fait de l'existence d'une physiologie bien particulière propre à la non-croissance.
 Les cellules doivent donc croître pendant suffisamment longtemps sur le premier milieu, sans épuiser celui-ci, puis une transition rapide doit être effectuée vers le milieu riche.
-Nous avons utilisé pour cela la \textit{Mother Machine}, un dispositif microfluidique à la base inventé pour maintenir et observer les cellules bactériennes en état de croissance stationnaire pendant de longues périodes.
-Au sein de ce dispositif, les cellules sont constamment en contact avec du milieu frais, qui circule et leur procure de nouveaux nutriments durant toute la durée de l'expérience (voir Section~\ref{sec:microflu}).
+Nous avons utilisé pour cela la \textit{Mother Machine}, un dispositif microfluidique à la base mis au point pour maintenir et observer les cellules bactériennes en état de croissance stationnaire pendant de longues périodes.
+Au sein de ce dispositif, les cellules sont en contact avec du milieu frais qui circule et leur procure de nouveaux nutriments durant toute la durée de l'expérience (voir Section~\ref{sec:microflu}).
 En changeant simplement le type de milieu injecté dans le dispositif, nous pouvons l'utiliser pour modifier en quelques secondes les nutriments auxquels les bactéries ont accès, et donc réaliser la transition de manière robuste et contrôlée.
@@
 
-De telles expériences s'avèrent malheureusement pointues et, dans le temps disponible pour le doctorat, nous n'avons pu en réaliser qu'un petit nombre, dont seule celle présentée ici s'est avérée exploitable.
+De telles expériences s'avèrent malheureusement pointues et, au cours du déroulement de cette thèse, nous n'avons pu en réaliser qu'un petit nombre, dont seule celle présentée ici s'est avérée exploitable.
 Celle-ci a cependant subi de nombreux problèmes expérimentaux qui, s'ils n'ont pas empêché complètement son exploitation, ont posé des problèmes pour l'analyse et l'interprétation des résultats.
@@
 Are such measurements feasible in practice?
-The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, or using radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spectrometry techniques~\cite{hui_quantitative_2015}.
-However, only the use of fluorescent proteins complies with our need for real-time single-cell quantification.
+The abundance of ribosomes can be quantified, for instance, by measuring the total RNA of the cell~\cite{scott_interdependence_2010}, or using radioactive markers \cite{gausing_regulation_1980,zengel_transcription_1986}, fluorescent labels~\cite{bakshi_superresolution_2012}, or mass spec\-trom\-e\-try techniques~\cite{hui_quantitative_2015}.
+However, only the use of fluo\-res\-cent pro\-teins com\-plies with our need for real-time single-cell quantification.
 Obtaining an estimate of precursor abundance is even more challenging, since there is no clear proxy for the totality of precursors in the cell and, while absolute quantification of all internal metabolites has been achieved in \textit{E.
@@
 \end{eqnarray}
-where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in units RFU/pixel measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$ the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
+where $L(t_k)$ and $F(t_k)$ are the length and fluorescence density in Relative Fluorescence Units per pixels measured at time $t_k$, respectively,  and $V(t_k)$  and $r(t_k)$ the corresponding actual volume and ribosome concentration at time $t_k$, respectively.
 ($\lambda$,$\gamma$) are unknown proportionality constants, and ($\epsilon_k$,$\eta_k$) uncorrelated sequence of measurement noise assumed normally distributed with mean 0.
@@
 \caption{
-\textbf{Estimation of the resource allocation profile using Kalman smoothing based on the fluorescence density measurements and the estimated growth rates from Fig.~\ref{fig:growth_rate_estimation}.}
+\textbf{Estimation of the resource allocation profile using Kal\-man smooth\-ing based on the fluorescence density measurements and the estimated growth rates from Fig.~\ref{fig:growth_rate_estimation}.}
 \textit{(A-B)} Gray lines represent the estimation of the fluorescence density (RFU/pixel) $F(\cdot)$ (in A) and the resource allocation profile $\gamma \alpha / \beta$ (in B) by the Kalman smoothing procedure for the 45 normal cells.
@@
 containing the pSIM5 plasmid for lambda-red recombinaison~\cite{sharan_recombineering_2009}.
-A kanamycin-resistant colony was selected on LA-glucose medium and
+A kana\-my\-cin-resistant colony was selected on LA-glucose medium and
 verified to show green fluorescence (485~nm excitation, 535~nm emission) in a microplate reader (Tecan Infinite 200 pro).
@@
 \textit{The 20 first bp anneals with the end of mCherry.}
-\item[] \includegraphics[scale=1]{Fig/cassette_R} \seqsplit{GAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTAGAAGAACTCGTCAAGAAGG}
+\item[] \includegraphics[scale=1]{Fig/cassette_R} \seqsplit{GAGCTTGCCGCCTTTCTGCAACTCGAACTATTTTGGGGGAGTTATCAAGCTTAGAAGAACTCGTCAAGAAGG}\\
 (Length: 72, used for both)\\
@@
 \hline
-\textbf{$D$~[min\textsuperscript{-1}]} & 0.0002064 $\pm$ 2.066e-06 & 0.0003027 $\pm$ 2.645e-06\\
+\textbf{$D$~[min\textsuperscript{-1}]} & 0.0002064 $\pm$ 2.1$\cdot 10^{-6}$ & 0.0003027 $\pm$ 2.7$\cdot 10^{-6}$\\
 \hline

### 2017-04-05 Chapters/introduction.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

@@
 
-Bien que largement répandues, ces lois se limitent cependant à un état bien particulier qu'on appelle croissance à l'état stationnaire.
+Bien que répandues chez de nombreux microorganismes, ces lois se limitent à un état bien particulier qu'on appelle croissance à l'état stationnaire.
 Cet état est atteint lorsque l'organisme se multiplie pendant suffisamment longtemps dans un environnement stable (Fig.~\ref{fig:growth_curve}).
@@
 Mais cet état a beau être pratique, il est très loin des conditions rencontrées par les microorganismes dans leur milieu de vie naturel.
-Ces derniers sont en effet plutôt soumis à des environnements très changeants, où les éléments chimiques indispensables à la croissance sont rares, et sont donc les objets d'une rude compétition.
+Ces derniers sont en effet plutôt soumis à des environnements changeants, où les éléments chimiques indispensables à la croissance sont rares, et sont donc les objets d'une rude compétition.
 Pour quelles raisons les microorganismes seraient-ils optimisés pour un état de croissance qu'ils n'ont quasiment jamais rencontré au cours de leur évolution ?
@@
 
-Pour tester les prédictions faites expérimentalement, nous chercherons à mesurer cette distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
-Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse des séries temporelles, permet d'observer et de reconstruire dans le temps l'allocation des ressources lors d'une transition brutale du milieu, notamment la transition d'un milieu pauvre vers un milieu riche (\textit{nutrient upshift}).
+Les prédictions faites seront ensuite testées expérimentalement.
+Nous chercherons donc à mesurer la distribution des ressources chez \textit{Escherichia coli} lors d'une transition de croissance (Chapitre~\ref{chap:experiments}).
+Nous montrerons comment l'utilisation de marqueurs fluorescents des ribosomes, d'un dispositif microfluidique de culture continue, et de méthodes originales d'analyse des séries temporelles, permet d'observer et de reconstruire l'allocation des ressources au cours d'une transition brutale du milieu, notamment la transition d'un milieu pauvre vers un milieu riche (\textit{nutrient upshift}).
 \end{chapter_summary}

### 2017-04-05 Chapters/theory.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

@@
-\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
+\chapter[Dynamical Allocation of Cellular Resources as an Optimal Control Problem]{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
 \chaptermark{Dynamical allocation of cell resources}
@@
 Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
-En Section \ref{sec:model}, nous modélisons la cellule comme un auto-réplicateur devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules (Fig.~\ref{fig:self_replicator}).
+En Section \ref{sec:model}, nous modélisons la cellule comme un auto-réplicateur devant allouer une ressource (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules (Fig.~\ref{fig:self_replicator}).
 Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
@@
 Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissance à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
-En Section~\ref{sec:growthlaws}, nous voyons en effet que pour chaque environnement, un $\alpha$ unique existe qui maximise la croissance de la cellule, mais surtout que ce $\alpha^*_{opt}$ est d'autant plus important que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
-De fait, les lois de croissance à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
+C'est aussi le cas dans notre modèle, et en Section~\ref{sec:growthlaws} nous voyons que pour chaque environnement, la croissance de la cellule est maximale pour une valeur unique de $\alpha$.
+Mais surtout, cette valeur que l'on nomme $\alpha^*_{opt}$ est d'autant plus grande que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
+C'est finalement toutes les lois de croissance à l'état stationnaire qui sont prédites par notre modèle simple, sous la simple hypothèse qu'à tout moment, la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
 Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?
@@
 Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif à optimiser est une fonction d'une autre fonction, plus communément appelée fonctionnelle.
-Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
+Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance, la maximisation de la production de biomasse sur un intervalle de temps comprenant un changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
 Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant l'utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
@@
 En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
-Ce type de solution se rencontre souvent en contrôle optimal, et répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
+Ce type de solution se rencontre souvent en contrôle optimal, et répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang} en anglais).
 Mais ce contrôle est-il pertinent d'un point de vue biologique ?
@@
 Mais surtout, en réutilisant un modèle du système ppGpp, connu pour réguler la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
-Cela procure un nouveau point de vue sur ce système, d'ailleurs largement répandu chez de nombreux microorganismes.
-En intégrant l'information provenant à la fois de la quantité de ribosomes et de celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule lorsque celle-ci doit s'adapter rapidement à un nouvel environnement.
+Cela éclaire d'un jour nouveau le fonctionnement et l'origine de ce système de régulation largement répandu chez de nombreux microorganismes.
+En intégrant l'information provenant à la fois de la quantité de ribosomes et de celle des précurseurs, ce dernier est capable d'allouer efficacement les ressources de la cellule lorsque celle-ci doit s'adapter rapidement à un nouvel environnement.
 \end{chapter_summary}
@@
 
-For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-grained models of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
+For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-\linebreak grained mod\-els of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
 The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
@@
 
-Bearing in mind that the major constituents of the cell are macromolecules (DNA, RNA, proteins), produced from precursor metabolites, a fundamental resource allocation question is the following: How much of the cellular resources are invested in the making of new macromolecules (gene expression machinery) and how much in performing other functions, in particular producing metabolic enzymes involved in the uptake of nutrients and their conversion to precursor metabolites (metabolic machinery)?
+Bearing in mind that the major constituents of the cell are mac\-ro\-mol\-e\-cules (DNA, RNA, proteins), produced from precursor metabolites, a fundamental resource allocation question is the following: How much of the cellular resources are invested in the making of new macromolecules (gene expression machinery) and how much in performing other functions, in particular producing metabolic enzymes involved in the uptake of nutrients and their conversion to precursor metabolites (metabolic machinery)?
 In order to address this question, we consider a self-replicating system composed of the gene expression machinery ($R$) and the metabolic machinery ($M$).
@@
 
-\subsection{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
+\subsection{A near-optimal feedback control strategy that exploits information on the imbalance between precursors and the gene expression machinery}
 \label{sec:strategies}
@@
 \item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
-\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
-\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
+\item[(C2)] For any given constant environment $E_M$, they drive the self-rep\-li\-cat\-or system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
+\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \linebreak \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
 \end{description}
@@
 \end{equation}
-As shown in \ref{S1_Text} by computing the Jacobian, the system given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
+As shown in \ref{S1_Text} by computing the Jacobian, the system given by \linebreak Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and \ref{eq:meth_g} has a single nontrivial stable steady state for any environment $E_M$ (Condition~C2).
 Moreover, Eq.~\ref{eq:meth_g} guarantees this steady state to be optimal (Condition~C3).
@@
 \end{equation}
-This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in Fig.~\ref{fig:onoffresults}\textit{A}.
+This strategy drives the system to a single steady state, because the $\hat{p}$-nullcline crosses the function $g(\hat{p})$ only once, as shown graphically in \linebreak Fig.~\ref{fig:onoffresults}\textit{A}.
 In \ref{S1_Text} we argue that this steady state is stable, by taking into account so-called sliding modes on the switching curve~\cite{filippov_differential_1988} (Condition~C2).
@@
 
-Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glutaminyl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second.
+Ref.~\cite{uter_longrange_2004} provides a typical value for such a reaction, catalyzed by glu\-ta\-min\-yl-tRNA synthetase: $k_{\textit{cat,GlnRS}} = 3.2$~s\textsuperscript{-1}, indicating that on average 3.2 glutaminyl-tRNA molecules are produced per glutaminyl-tRNA synthetase molecule per second.
 After conversion to mass units using molar weight from~\cite{freist_glutaminyltrna_1997}, this yields
@@
 
-An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}.
+An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying \linebreak Eqs~\ref{eq:model}-\ref{eq:PMP}.
 The extremal is said to be normal (resp.
@@
 
-\subsection{S1 Table -- Parameter values of self-replicator model}
+\subsection{S1 Table -- Parameter values of the self-rep\-li\-cat\-or model}
 \manuallabel{S1_Table}{S1~Table}
@@
 \caption{
-\textbf{Simple control strategies for the self-replicator of bacterial growth.}
+\textbf{Simple control strategies for the self-rep\-li\-cat\-or of bacterial growth.}
 \textit{A:} Nutrient-only strategy: $\alpha = f(E_M)$.

### 2017-04-05 abstract.tex
Some small frenchproofreading and correct overfull/underfull hbox when possible

@@
 Même si de vieilles données le suggèrent, un tel comportement \textit{bang-bang} n'a jamais été totalement confirmé expérimentalement pour la synthèse des ribosomes.
-Nous mesurons donc l'abondance des ribosomes chez \textit{Escherichia coli} au niveau d'une cellule unique lors d'un changement brutal de milieu de culture.
+Nous mesurons donc l'abondance des ribosomes chez \textit{Escherichia coli} au niveau de cellules individuelles lors d'un changement brutal de milieu de culture.
 En particulier, nous créons une souche de \textit{E.coli} sur laquelle un rapporteur fluorescent est attaché à l'une des sous-unités ribosomales, permettant ainsi leur quantification \textit{in vivo}.
@@
 Nous mettons en évidence que la meilleure stratégie consiste à aiguiller toutes les ressources disponibles vers la production d'un seul composant de manière alternative, une prédiction que nous tentons également de confirmer de manière expérimentale.
-En d'autres termes, dans un contexte dynamique, mieux vaut agir tel un interrupteur à bascule plutôt qu'un variateur de lumière.
 Ces résultats pourraient suggérer de nouvelles façons d'optimiser les processus biotechnologiques de production de molécules d'intérêt.

### 2017-04-09 Chapters/introduction.tex
Correct quotes, add quote to chapter 3

@@
 
-\textit{There's an infinity of things that have never been done before, and most of those things are not worth doing.} -- Jeremy Fox, \textit{Dynamic Ecology (2016)}~\cite{fox_how_2016}
+\textit{"There's an infinity of things that have never been done before, and most of those things are not worth doing."} -- Jeremy Fox, \textit{Dynamic Ecology}~\cite{fox_how_2016}
 

### 2017-04-09 Chapters/theory.tex
Correct quotes, add quote to chapter 3

@@
 
-\textit{Rather than propose a new theory or unearth a new fact, often the most important contribution a scientist can make is to discover a new way of seeing old theories or facts.} -- Richard Dawkins, \textit{The Selfish Gene}~\cite{dawkins_selfish_1976}, preface to 1989 edition.
+\textit{"Rather than propose a new theory or unearth a new fact, often the most important contribution a scientist can make is to discover a new way of seeing old theories or facts."} -- Richard Dawkins, \textit{The Selfish Gene}~\cite{dawkins_selfish_1976} (preface to 1989 edition).
 

### 2017-04-09 Chapters/experiments.tex
Correct quotes, add quote to chapter 3
Patricia's proofreading of french parts

@@
 
+\textit{"Oh you like beetles? Wonderful! Then that's what you'll be killing for the next 50 years."} -- Zach Weinersmith, \textit{Biologists are weird}~\cite{weinersmith_biologists_2016}.
+
 \begin{center}
@@
 La Section~\ref{sec:exp_design} s'attache à définir les conditions expérimentales dans lesquelles le comportement prédit lors du Chapitre~\ref{chap:theory} pourra être observé.
-En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que $\alpha (\cdot)$ peut être reconstruit au cours du temps de la transition si l'on dispose de mesures dynamiques du taux de croissance et de la concentration en machinerie d'expression génique.
+En utilisant le modèle d'auto-réplicateur (Eqs~\ref{eq:pdef-exp} et \ref{eq:rdef-exp}), nous montrons que l'allocation des ressources $\alpha (\cdot)$ peut être reconstruite au cours de la transition si l'on dispose de mesures dynamiques du taux de croissance et de la concentration en machinerie d'expression génique.
 Cependant, le comportement tout-ou-rien a une dynamique bien particulière qui peut facilement être offusquée au niveau d'une population de cellules si ces dernières ne sont pas synchrones.
-De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de cellules individuelles, ce qui nous a conduit à utiliser des rapporteurs fluorescents de l'abondance d'une des machineries.
+De fait, cette question ne peut être résolue qu'en obtenant des mesures à haute fréquence et au niveau de cellules individuelles, ce qui nous a conduits à utiliser des rapporteurs fluorescents de l'abondance d'une des machineries.
 En particulier, nous construisons une souche d'\textit{Escherichia coli} chez laquelle la sous-unité S2 du ribosome a été fusionnée avec une protéine fluorescente (GFP) (voir Section~\ref{sec:methods_strain} des matériels et méthodes).
@@
 Ces problèmes expérimentaux sont abordés dans la Section~\ref{sec:data_acquisition}, et sont largement discutés en Section~\ref{sec:chap3_discussion}.
-En effet, suite à un problème logiciel et à ce qui semble être une contamination par un bacteriophage, nous n'avons pas pu obtenir des séries temporelles aussi longues qu'initialement prévu.
+En effet, suite à un problème logiciel et à ce qui semble être une contamination par un bactériophage, nous n'avons pas pu obtenir des séries temporelles aussi longues qu'initialement prévu.
 Même si nous disposons néanmoins des données se situant autour de la transition, le fait de ne pas avoir atteint un nouvel état de croissance stationnaire sur glucose va s'avérer être un inconvénient majeur pour l'analyse des résultats.

### 2017-04-09 Chapters/introduction.tex
Patricia's proofreading of french parts

@@
 
-Peut-on, à l'aide d'un tel principe universel, formaliser la croissance des microorganismes en lois simples, analogues à celle que l'on peut trouver dans d'autres domaines des sciences comme la physique ?
+Peut-on, à l'aide d'un tel principe universel, formaliser la croissance des microorganismes en lois simples, analogues à celles que l'on peut trouver dans d'autres domaines des sciences comme la physique ?
 Historiquement, la première tentative d'établissement de lois fondamentales de la physiologie microbienne  fut la découverte de la loi de Monod il y a plus de 60 ans (Fig.~\ref{fig:monod_law} et Eq.~\ref{eq:monod_law}).

### 2017-04-09 Chapters/experiments.tex
Add Eugenio's citation on Kalman smoothing

@@
 While this is suitable when bacteria are at steady state, it is not applicable during a growth transition, where we expect the growth rate to vary between successive divisions.
-Other techniques are less sensible to the above problem~\cite{zulkower_robust_2015}, and we decided to use Kalman smoothing for our purpose~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+Other techniques are less sensible to the above problem~\cite{zulkower_robust_2015}, and we decided to use Kalman smoothing for our purpose~\cite{kailath_linear_2000,jazwinski_stochastic_2007,cinquemani_estimation_2017}.
 
@@
 It has been extensively used in engineering applications (guidance, tracking, control, ...) requiring the real-time estimation of hidden variables in a dynamical system from present and past measurements.
-Kalman smoothing is an extension of Kalman filtering using information about past and present but also future measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007}.
+Kalman smoothing is an extension of Kalman filtering using information about past and present but also future measurements of the state of the system, and is widely applied to estimate unknown inputs in time series analysis~\cite{kailath_linear_2000,jazwinski_stochastic_2007,cinquemani_estimation_2017}.
 The advantage of Kalman filtering with respect to Kalman smoothing is that it uses all available information to estimate the hidden variables of a dynamical system, in applications where a real-time response is not required.
@@
 This provides constraints on the variation of $\mu$ that have the effect of a regularization.
-In particular, in the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process.
+In particular, in the Kalman smoothing procedure, we model the variations of $\mu$ as the outcome of a stochastic process~\cite{cinquemani_estimation_2017}.
 The laws describing this process play the role of a prior on the expected regularity of $\mu$ (the more noisy the process, the less constrained the variations of $\mu$).
@@
 that may themselves be Gaussian random variables with some given statistics.
-The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable.
+The advantages of this regularization method are that the reconstructed $\mu$ is guaranteed to be second-order differentiable~\cite{cinquemani_estimation_2017}.
 It is also equivalent to other methods like Tikhonov regularization~\cite{tikhonov1977solutions,tikhonov1963solution,de_nicolao_nonparametric_1997}, which minimizes least-square differences between predictions and observations along with penalizing the variations of the input.

### 2017-04-09 acknowledgements.tex
Finalize chap 3 suppinfo, finalize abstracts, remove eps files, version sent to Hidde and Hans
Add acknowledgements

@@
 \cleardoublepage
+\selectlanguage{french}
 \begin{acknowledgements}
-(remerciements ici)
+
+Il est probable que cette page, même si elle s'adresse au public restreint des francophones, sera la plus consultée de cette thèse.
+Cela fait effectivement longtemps que le manuscrit de thèse a perdu le monopole de la communication scientifique.
+On exige cependant qu'il soit traité avec le même respect qu'il y a un siècle, en faisant comme s'il n'avait rien perdu de sa gloire passée, non sans une certaine forme d'hypocrisie administrative.
+Ce paradoxe rend généralement cet exercice difficile, mais par sa longueur, le déroulement du doctorat lui même l'est bien davantage.
+À ce jour, j'ai par exemple investi plus de 15\% de mon existence dans ce projet de recherche.
+C'est long, surtout pour quelqu'un comme moi qui aime toucher à tout.
+Inévitablement, de nombreuses rencontres qui ont grandement influencé ce travail se sont succédées, et c'est le rôle de cette page que de leur manifester ma gratitude.
+
+Tout d'abord, je remercie évidemment Marion Joulié, qui même si son engagement date de bien avant ce doctorat, a su rester jusqu'au bout et restera encore je l'espère pour les nombreuses années à venir.
+
+Je remercie particulièrement Valentin Zulkower et Edith Grac qui étaient présents pendant la meilleure période de ce doctorat.
+Sans nos nombreuses discussions, ce travail n'aurait probablement pas été le même.
+Viennent ensuite tous les jeunes membres des équipes Ibis et BIOP, notamment les passages remarqués de Çiğdem Ak, Cindy Barillot, Diana Stefan, Bernard Chelli, Ludowic Lancelot, Julien Sauvage, et tous les autres dont les noms ne me viennent pas au moment d'écrire ces lignes.
+Je n'oublie pas de remercier les (nombreux) jeunes chercheurs de l'équipe Mistis qui ont su m'accueillir lorsque la plupart des gens sus-cités se sont progressivement évaporés.
+
+D'un point de vue plus scientifique, je remercie évidemment mes directeurs Hans Geiselmann et Hidde de Jong, mais aussi Corinne Pinel, Irina Mihalcescu, Delphine Ropers, Eugenio Cinquemani, Jean-Luc Gouzé, et Francis Mairet, qui ont apporté des contributions plus que concrètes dans les étapes clés de ce travail.
+Je remercie pour les mêmes raisons tous les membres des équipes Ibis et BIOP avec qui j'ai pu échanger lors de nos nombreuses réunions hebdomadaires.
+
+Enfin, je tenais également à remercier les communautés de Bioinfo-fr.net et JeBiF -- RSG France, en particulier jnsll pour ses corrections des parties françaises de ce manuscrit, mais aussi Chopopope, Kumquatum, Norore, Hautbit, m4rsu et Will qui ont su virtuellement me supporter dans les meilleures et les pires périodes de ce doctorat.
 \end{acknowledgements}
+\selectlanguage{english}
